### 2022-11-29 manuscript/manuscript.Rmd
so it begins :seedling:

---
title: Title here
subtitle: Do you have a subtitle? If so, write it here
titlerunning: Short form of title (if too long for head)
authorrunning: Short form of author list if too long for running head
thanks: | 
    Grants or other notes about the article that should go on the front 
    page should be placed here. General acknowledgments should be placed at the
    end of the article.

authors: 
- name: Äüthör 1
  address: Department of YYY, University of XXX
  email: abc@def

- name: Âuthóř 2
  address: Department of ZZZ, University of WWW
  email: djf@wef

keywords:
- key
- dictionary
- word

#PACS: 
#- PAC1
#- superPAC

MSC:
- MSC code 1
- MSC code 2

abstract: |
  The text of your abstract.  150 -- 250 words.

bibliography: bibliography.bib
biblio-style: spphys
# bibstyle options spbasic(default), spphys, spmpsci
output: rticles::springer_article
---

# Introduction {#intro}

Your text comes here. Separate text sections with \cite{Mislevy06Cog}.

# Section title {#sec:1}

Text with citations by \cite{Galyardt14mmm}.

## Subsection title {#sec:2}

as required. Don't forget to give each section
and subsection a unique label (see Sect. \ref{sec:1}).

#### Paragraph headings

Use paragraph headings as needed.

\begin{align}
a^2+b^2=c^2
\end{align}

### 2022-11-29 manuscript/manuscript.tex
so it begins :seedling:

% !TeX program = pdfLaTeX
\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage[utf8]{inputenc}

\usepackage[hyphens]{url} % not crucial - just used below for the URL
\usepackage{hyperref}

%
% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
%
% insert here the call for the packages your document requires
%\usepackage{latexsym}
% etc.
%
% please place your own definitions here and don't use \def but
% \newcommand{}{}
%
% Insert the name of "your journal" with
% \journalname{myjournal}
%

%% load any required packages here

% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\begin{document}

\title{Title here \thanks{Grants or other notes about the article that
should go on the front page should be placed here. General
acknowledgments should be placed at the end of the article.} }
 \subtitle{Do you have a subtitle? If so, write it here}

    \titlerunning{Short form of title (if too long for head)}

\author{  Äüthör 1 \and  Âuthóř 2 \and  }

    \authorrunning{ Short form of author list if too long for running
head }

\institute{
        Äüthör 1 \at
     Department of YYY, University of XXX \\
     \email{\href{mailto:abc@def}{\nolinkurl{abc@def}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        Âuthóř 2 \at
     Department of ZZZ, University of WWW \\
     \email{\href{mailto:djf@wef}{\nolinkurl{djf@wef}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
    }

\date{Received: date / Accepted: date}
% The correct dates will be entered by the editor

\maketitle

\begin{abstract}
The text of your abstract. 150 -- 250 words.
\\
\keywords{
        key \and
        dictionary \and
        word \and
    }

    \subclass{
                    MSC code 1 \and
                    MSC code 2 \and
            }

\end{abstract}

\def\spacingset#1{\renewcommand{\baselinestretch}%
{#1}\small\normalsize} \spacingset{1}

\hypertarget{intro}{%
\section{Introduction}\label{intro}}

Your text comes here. Separate text sections with \cite{Mislevy06Cog}.

\hypertarget{sec:1}{%
\section{Section title}\label{sec:1}}

Text with citations by \cite{Galyardt14mmm}.

\hypertarget{sec:2}{%
\subsection{Subsection title}\label{sec:2}}

as required. Don't forget to give each section and subsection a unique
label (see Sect. \ref{sec:1}).

\hypertarget{paragraph-headings}{%
\paragraph{Paragraph headings}\label{paragraph-headings}}

Use paragraph headings as needed.

\begin{align}
a^2+b^2=c^2
\end{align}

\bibliographystyle{spphys}
\bibliography{bibliography.bib}

\end{document}

### 2022-11-29 manuscript/manuscript.Rmd
a little historical hat-tip :tophat:

@@
 ---
-title: Title here
-subtitle: Do you have a subtitle? If so, write it here
-titlerunning: Short form of title (if too long for head)
-authorrunning: Short form of author list if too long for running head
+title: "Pretty darn good control: when are approximate solutions better than approximate models"
+titlerunning: Pretty darn good control
 thanks: | 
-    Grants or other notes about the article that should go on the front 
-    page should be placed here. General acknowledgments should be placed at the
-    end of the article.
+    This material is based upon work supported by the National Science
+    Foundation under Grant No. DBI-1942280.
 
 authors: 
-- name: Äüthör 1
-  address: Department of YYY, University of XXX
-  email: abc@def
-
-- name: Âuthóř 2
-  address: Department of ZZZ, University of WWW
-  email: djf@wef
+- name: Felipe Montealegre-Mora
+  address: Department of Environmental Science, Policy, and Management, University of California, Berkeley
+- name: Marcus Laperolerie
+  address: Department of Environmental Science, Policy, and Management, University of California, Berkeley
+- name: Melissa Chapman
+  address: Department of Environmental Science, Policy, and Management, University of California, Berkeley
+- name: Carl Boettiger
+  address: Department of Environmental Science, Policy, and Management, University of California, Berkeley
+  email: cboettig@berkele.edu
 
 keywords:
-- key
-- dictionary
-- word
+- Optimal Control
+- Reinforcement Learning
+- Uncertainty
+- Decision Theory
 
@@
 biblio-style: spphys
+
 # bibstyle options spbasic(default), spphys, spmpsci
 output: rticles::springer_article
+
 ---
@@
 \end{align}
+
+## Acknowledgements
+
+The title of this piece references a mathematical biology workshop at NIMBioS
+organized by Paul Armsworth, Alan Hastings, Megan Donahue, and Carl Towes in
+2011 that first posed the question addressed here.
+
+# References

### 2022-12-13 manuscript/manuscript.tex
added in some section names, couple of sentences

@@
 \hypertarget{sec:1}{%
-\section{Section title}\label{sec:1}}
+\section{Background}\label{sec:background}}
 
-Text with citations by \cite{Galyardt14mmm}.
+% Text with citations by \cite{Galyardt14mmm}.
+In this section we first review two well-established techniques commonly used in sustainable fishery management. 
+These are the maximum sustainable yield (MSY) and the constant escapement (CE) approaches. 
+After this, reinforcement learning is briefly reviewed
 
 \hypertarget{sec:2}{%
-\subsection{Subsection title}\label{sec:2}}
+\subsection{Fishery management}\label{sec:fishery}}
 
-as required. Don't forget to give each section and subsection a unique
-label (see Sect. \ref{sec:1}).
+\hypertarget{sec:3}{%
+\subsection{Reinforcement learning}\label{sec:rl}}
+
+\hypertarget{sec:4}{%
+\section{A three-species dynamical model}\label{sec:model}}
+
+\hypertarget{sec:5}{%
+\section{Results}\label{sec:results}}
+
+\hypertarget{sec:6}{%
+\section{Discussion}\label{sec:discussion}}
 
@@
 
+Use paragraph headings as needed. 
+Use paragraph headings as needed. 
+Use paragraph headings as needed. 
+Use paragraph headings as needed. 
+Use paragraph headings as needed. 
+Use paragraph headings as needed. 
 Use paragraph headings as needed.

### 2022-12-13 manuscript/manuscript.tex
started writing a bit about the models

@@
 %
-\usepackage{amsmath}
+\usepackage{amsmath,amssymb}
 \usepackage{graphicx}
@@
 \usepackage{hyperref}
+
+\def \der #1 #2 {\frac{\mathrm{d}#1}{\mathrm{d}#2}}
 
@@
 These are the maximum sustainable yield (MSY) and the constant escapement (CE) approaches. 
-After this, reinforcement learning is briefly reviewed
+After this, deep reinforcement learning is briefly reviewed
 
@@
 \hypertarget{sec:3}{%
-\subsection{Reinforcement learning}\label{sec:rl}}
+\subsection{Deep reinforcement learning}\label{sec:rl}}
 
 \hypertarget{sec:4}{%
-\section{A three-species dynamical model}\label{sec:model}}
+\section{A hierarchy of dynamical models}\label{sec:model}}
+
+In this section we present three models of increasing complexity which plausibly describe the population dynamics of a marine ecosystem.
+These models will form the test beds for the comparison between classical fishery management strategies and DRL.
+
+\hypertarget{paragraph-headings}{%
+\paragraph{A one-dimensional tipping point model.}\label{p:may}}
+Our first model has been used in~\cite{may77} to study the possible dangers of driving a dynamical ecosystem past a tipping point.
+Its dynamics is given by
+\begin{align}
+  \label{eq:may77}
+  \der{V} t = rV\left(
+    1-V/K
+    \right) 
+    -
+    \frac{\beta H V^2}{V_0^2 + V^2}. 
+\end{align}
+Here the only dynamic degree of freedom is $V$, while $r,\ K,\ \beta,\ V_0,$ and $H$ are fixed parameters.
+This model describes a population $V$ which grows logistically up to a carrying capacity of $K$, and whose growth is controlled by a constant population $H$ of predators.
+For fixed values of $r,\ K,\ V_0$ and $\beta$, varying $H$ may produce a \emph{catastrophe} (see Fig.~\ref{fig:tbd}).
 

### 2022-12-13 manuscript/manuscript.tex
more on 1D model

--- 2022-12-13 manuscript/manuscript.tex
@@
 \usepackage{hyperref}
+
+% common operators
+\newcommand*{\der}[2]{\frac{\mathrm{d}#1}{\mathrm{d}#2}}
+
+% structural markers
+\def\div{\;|\;}
+\def\bigdiv{\;\big|\;}
+\def\Bigdiv{\;\Big|\;}
+\def\Biggdiv{\;\Bigg|\;}
 
@@
 \hypertarget{sec:3}{%
-\subsection{Reinforcement learning}\label{sec:rl}}
+\subsection{Deep reinforcement learning}\label{sec:rl}}
 
 \hypertarget{sec:4}{%
-\section{A three-species dynamical model}\label{sec:model}}
+\section{Dynamical models used}\label{sec:model}}
+
+In this section we present three models of increasing complexity which plausibly describe the population dynamics of a marine ecosystem.
+These models will form the test beds for the comparison between classical fishery management strategies and DRL.
+
+\hypertarget{paragraph-headings}{%
+\paragraph{A one-dimensional tipping point model.}\label{p:may}}
+
+Consider a population $V$ whose dynamics is given by
+\begin{align}
+  \label{eq:may77}
+  \der{V}{t} = rV\left(
+    1-V/K
+    \right) 
+    -
+    \frac{\beta H V^2}{V_0^2 + V^2}. 
+\end{align}
+This model has been used in~\cite{may77} to study the possible dangers of driving a 
+Here the only dynamic degree of freedom is $V$, while $r,\ K,\ \beta,\ V_0,$ and $H > 0$ are fixed positive parameters.
+
+In~\eqref{eq:may77}, a population $V$ grows logistically with rate $r$ up to carrying capacity $K$.
+This is expressed by the first term in the equation,
+\begin{align*}
+  L(V \div r,\ K) := rV\left(1 - V / K\right).
+\end{align*}
+Moreover, $V$ is predated on by a (constant) population $H$, as can be seen from the negative term
+\begin{align*}
+  F(V,\ H \div \beta,\ V_0) := \frac{\beta H V^2}{V_0^2 + V^2}
+\end{align*}
+which saturates to $F\to\beta H$ as $V\to\infty$, and whose half maximum is $V_0$, i.e.\ $F(V=V_0,\ H; \beta,\ V_0)=\beta H/2$.
+
+For fixed values of $r,\ K,\ V_0$ and $\beta$, varying $H$ may produce a \emph{catastrophe} (see Fig.~\ref{fig:tbd})---a sudden disappearance of a stable fixed point of the system which .
 

### 2022-12-13 manuscript/manuscript.tex
more may77

@@
 
-\hypertarget{paragraph-headings}{%
-\paragraph{A one-dimensional tipping point model.}\label{p:may}}
+\hypertarget{sec:4.1}{%
+\subsection{A one-dimensional tipping point model.}\label{p:may}}
 
@@
 \end{align}
-This model has been used in~\cite{may77} to study the possible dangers of driving a 
-Here the only dynamic degree of freedom is $V$, while $r,\ K,\ \beta,\ V_0,$ and $H > 0$ are fixed positive parameters.
+This model has been used in~\cite{may77} to describe a grazing ecosystem, where a species $V$ of vegetation is harvested by a constant herbivore population $H$. 
+% Here the only dynamic degree of freedom is $V$, while $r,\ K,\ \beta,\ V_0,$ and $H > 0$ are fixed positive constants.
 
@@
 \end{align*}
-which saturates to $F\to\beta H$ as $V\to\infty$, and whose half maximum is $V_0$, i.e.\ $F(V=V_0,\ H; \beta,\ V_0)=\beta H/2$.
+which saturates to $\beta H$ as $V\to\infty$, and whose half maximum is $V_0$, i.e.\ $F(V=V_0,\ H; \beta,\ V_0)=\beta H/2$.
 
-For fixed values of $r,\ K,\ V_0$ and $\beta$, varying $H$ may produce a \emph{catastrophe} (see Fig.~\ref{fig:tbd})---a sudden disappearance of a stable fixed point of the system which .
+Ref.~\cite{may77} studies the fixed points of~\eqref{eq:may77} in order to show that in certain parameter regimes, its dynamics can undergo a \emph{catastrophe}.
+A catastrophe is a sudden change in the state of the system from one stable state to another---often, the final state is ecologically detrimental, possibly associated with extinction or near-extinction events.
+
+Fig.~\ref{fig:tbd} shows the stable $V$ populations for differing values of $H$. 
+Here one sees that as $H\to T_2$, the top stable state is annihilated with the unstable fixed point.
+This way, if, e.g.\ $H$ were to slowly drift until $H=T_2$, the system would collapse to the low stable state, leading to a near extinction of $V$.
+
+% For fixed values of $r,\ K,\ V_0$ and $\beta$, varying $H$ may produce a \emph{catastrophe} (see Fig.~\ref{fig:tbd})---a sudden disappearance of a stable fixed point of the system which .
 

### 2022-12-13 manuscript/manuscript.tex
1d model

@@
 \usepackage[utf8]{inputenc}
+\usepackage{xcolor}
 
@@
 \def\Biggdiv{\;\Bigg|\;}
+\def\darg{\;\cdot\;}
+
+% comments / colors
+\definecolor{Fabulous}{RGB}{220,0,100}
+\newcommand{\fm}[1]{{\color{Fabulous}\bf [FMM: #1]}}
 
@@
 Here one sees that as $H\to T_2$, the top stable state is annihilated with the unstable fixed point.
-This way, if, e.g.\ $H$ were to slowly drift until $H=T_2$, the system would collapse to the low stable state, leading to a near extinction of $V$.
+A possible situation of concern here is the following:
+The system begins in the high stable point for some $H<T_2$.
+Slowly, $H$ slowly drifts upward, bringing the system up to $H=T_2$.
+At this point, the system suddenly finds itself in the lower stable point's ‘‘basin’’---this leads to a quick collapse of $V$ down to the stable point.
+Along a similar vein, if the system's evolution is noisy, then this catastrophe may happen for lower $H$ values: for $H$ close enough to $T_2$, there is a large probability that the $V$ crosses the unstable state and collapses towards the lower stable state.
+\fm{Maybe this can be explained in the figure better?}
+
+This way, if, e.g.\ the system is at the high stable state for low $H$, and $H$ were to slowly drift until $H=T_2$, the system would collapse to the low stable state, leading to a near extinction of $V$.
+
+\hypertarget{paragraph-headings}{%
+\paragraph{System~\eqref{eq:may77} as a simple fishery model.}\label{p:may as fishery}}
+We reinterpret the model~\eqref{eq:may77} as a simplified model for a fishery.
+There is some precedent for this already in~\cite{may77}, however our approach here is not in 1-to-1 correspondence to the approach used in that reference.
+\fm{Refers to the fact that in the reference $F(\darg)$ is our harvest quota, using constant effort.
+But here, it is a predation term, we compete with the predator in this case.}
+
+% Ref.~\cite{may77} points out that the system presented above can be reinterpreted as a simple model for a fishery, where quotas are set using a constant effort rule.
+% Here, the natural dynamics of $V$ is purely logistic growth, which is hampered by human harvesting according to the function $F$.
+% This way, $H$ is the effort allowed for the sake of quotas.
+% \footnote{%
+%   The non-linearity of the $F$ function is justified in the fact that, for low very populations $V$, the effort necessary to obtain a constant-size catch grows faster than $H\sim 1/V$.
+%   Here, for our case, $H\sim 1/V^2$ is used.
+% }
+
+We use this
 

### 2022-12-15 manuscript/manuscript.tex
removed legacy comments

@@
 \paragraph{System~\eqref{eq:may77} as a simple fishery model.}\label{p:may as fishery}}
-We reinterpret the model~\eqref{eq:may77} as a simplified model for a fishery.
-There is some precedent for this already in~\cite{may77}, however our approach here is not in 1-to-1 correspondence to the approach used in that reference.
-\fm{Refers to the fact that in the reference $F(\darg)$ is our harvest quota, using constant effort.
-But here, it is a predation term, we compete with the predator in this case.}
+We now reinterpret the model~\eqref{eq:may77} as a simplified model for a fishery.
+The population $V$ is taken to be the a fish species which we are interested in harvesting and $H$ is some predator which feeds on $V$.
+
+ Already~\cite{may77} has used~\eqref{eq:may77} to describe a fishery.
+ However, our approach here is not in 1-to-1 correspondence to the approach used in that reference.
+\footnote{
+  In~\cite{may77}, the term $F(V,\ H\div \beta,\ V_0)$ is reinterpreted as a constant-effort harvest quota by humans, where $H$ is the effort expended, while the natural dynamics of $V$ is logistic growth.
+  The non-linearity of the $V$-dependence on the harvest quota is justified in the fact that, for very low populations, the necessary effort to obtain a constant-size catch scales faster than $H\sim 1/V$.
+
+  In contrast, as will be discussed in the following, our approach assumes that $V$ is predated on by some population $H$.
+}
 

### 2023-01-04 manuscript/manuscript.Rmd
figure brainstorm

@@
 
-# Section title {#sec:1}
+# Figures brainstorm
 
-Text with citations by \cite{Galyardt14mmm}.
+Figure 1: 1-D and 3-D model conceptual figure. something about the objective / decision
+Figure 2: Stability / multistability in the 1D and 3D models.  state space + time views
 
-## Subsection title {#sec:2}
+Figure 3: The 1-D optimal management solution.  'constant escapement' intuition etc
 
-as required. Don't forget to give each section
-and subsection a unique label (see Sect. \ref{sec:1}).
-
-#### Paragraph headings
-
-Use paragraph headings as needed.
-
-\begin{align}
-a^2+b^2=c^2
-\end{align}
+Results figures:
+- timeseries example of management under the RL policy.  probably compare to managing under 1-D solution / rule-of-thumb methods
+- state-space view of management doughnut 
+- visualization / encapsulation of the policy heatmaps, slices, policy vs position along ellipse 
+- reward plot over time (comparing methods)
 

### 2023-01-06 manuscript/manuscript.Rmd
added a bit RL

--- 2022-11-29 manuscript/manuscript.Rmd
@@
 
-Your text comes here. Separate text sections with \cite{Mislevy06Cog}.
+<!-- Your text comes here. Separate text sections with \cite{Mislevy06Cog}. -->
 
-# Section title {#sec:1}
+# Reinforcement learning {#sec:DRL}
 
-Text with citations by \cite{Galyardt14mmm}.
+Reinforcement learning (RL) is a way of approaching *control problems* through machine learning.
+An RL application can be conceptually separated into two parts: an *agent*, and an *environment*.
+The agent may *act* on the environment.
+This produces two effects: on the one hand, the state of the environment is changed, and on the other, the agent receives a *reward* (see Fig. ).
+The reward depends on what action the agent chose, and on the initial and final states of the environment.
+The reward encodes the goal which we desire the agent to acheive---in other words, it encodes how we wish the agent controls the environment.
+The main part of any RL algorithm is then to progressively improve the agent's *policy*, the function the agent uses to choose which action to take given the state of the environment.
+This is done by aggregating experience and learning from it.
+
+# Sustainable fishery management {#sec:fishery}
+
+# Results
+
+# Discussion
+
+<!-- Text with citations by \cite{Galyardt14mmm}.
 
@@
 a^2+b^2=c^2
-\end{align}
+\end{align} -->
 

### 2023-01-06 manuscript/manuscript.Rmd

@@
 
+# Figures brainstorm
+
+Figure 1: 1-D and 3-D model conceptual figure. something about the objective / decision
+Figure 2: Stability / multistability in the 1D and 3D models.  state space + time views
+
+Figure 3: The 1-D optimal management solution.  'constant escapement' intuition etc
+
+Results figures:
+- timeseries example of management under the RL policy.  probably compare to managing under 1-D solution / rule-of-thumb methods
+- state-space view of management doughnut 
+- visualization / encapsulation of the policy heatmaps, slices, policy vs position along ellipse 
+- reward plot over time (comparing methods)
+
 # Reinforcement learning {#sec:DRL}
@@
 This is done by aggregating experience and learning from it.
-
-# Sustainable fishery management {#sec:fishery}
-
-# Results
-
-# Discussion
-
-<!-- Text with citations by \cite{Galyardt14mmm}.
-
-## Subsection title {#sec:2}
-
-as required. Don't forget to give each section
-and subsection a unique label (see Sect. \ref{sec:1}).
-
-#### Paragraph headings
-
-Use paragraph headings as needed.
 

### 2023-01-06 manuscript/manuscript.Rmd
rewrote a bit

--- 2023-01-04 manuscript/manuscript.Rmd
@@
 
-Your text comes here. Separate text sections with \cite{Mislevy06Cog}.
+<!-- Your text comes here. Separate text sections with \cite{Mislevy06Cog}. -->
 
@@
 
+# Reinforcement learning {#sec:DRL}
+
+Reinforcement learning (RL) is a way of approaching *control problems* through machine learning.
+An RL application can be conceptually separated into two parts: an *agent*, and an *environment*.
+The *environment* is commonly a computer simulation.
+The *agent*, on the other hand, is a computer program which interacts with the environment.
+That is, the agent may act on the environment and change its state.
+Such an action and its consequence, moreover, give the agent a *reward* which encodes our goal for the agent.
+The main part of an RL algorithm is then to progressively improve the agent's *policy*, in order to maximize the cumulative reward received. 
+This is done by aggregating experience and learning from it.
+
+In our case, the environment will be a computational model of the population dynamics of a fishery.
+The agent chooses how much to fish from a given species in the environment, thereby changing the populations in the environment's state.
+
+\begin{align}
+a^2+b^2=c^2
+\end{align} -->
+
 ## Acknowledgements

### 2023-01-06 manuscript/manuscript.Rmd
a bit more RL

@@
 
+# Fisheries: approximate models and approximate solutions
+
+<!-- Introduce the 1D and 3D models, and have Figure 1 already here. -->
+
 # Reinforcement learning {#sec:DRL}
@@
 An RL application can be conceptually separated into two parts: an *agent*, and an *environment*.
-The *environment* is commonly a computer simulation.
+The *environment* is commonly a computer simulation, although it sometimes can be a real world system.
 The *agent*, on the other hand, is a computer program which interacts with the environment.
 That is, the agent may act on the environment and change its state.
-Such an action and its consequence, moreover, give the agent a *reward* which encodes our goal for the agent.
+This, moreover, gives the agent a *reward* which encodes the control goal.
 The main part of an RL algorithm is then to progressively improve the agent's *policy*, in order to maximize the cumulative reward received. 
 This is done by aggregating experience and learning from it.
+See Fig. \ref{tbd}.
 
-In our case, the environment will be a computational model of the population dynamics of a fishery.
-The agent chooses how much to fish from a given species in the environment, thereby changing the populations in the environment's state.
-
-\begin{align}
-a^2+b^2=c^2
-\end{align} -->
+In our case, the environment will be a computational model of the population dynamics of a fishery, with the environment state being a vector of all the fish populations, $S=(V_1, V_2, H)$.
+The system evolves naturally according to \eqref{eq:3d model}.
+At each time step, the agent harvests one of the populations, say $A$. 
+Specifically, it chooses a quota $q$: the fraction of $A$ that will be fished.
+This changes the state as
+\begin{align*}
+  (V_1,\ V_2,\ H) \mapsto ((1-q)V_1,\ V_2,\ H)
+\end{align*}
+and secures a reward of $qA$ to the agent.
+Afterwards, the environment undergoes its natural dynamics.
 

### 2023-01-06 manuscript/manuscript.Rmd
typo
bit more rl

@@
 \end{align*}
-and secures a reward of $qA$ to the agent.
+and secures a reward of $qV_1$ to the agent.
 Afterwards, the environment undergoes its natural dynamics.
+
+## Mathematical framework for RL
+
+Mathematically, RL may be formulated using a discrete time *partially observable Markov decision process (POMDP)*.
+This formalization is rather flexible and allows one, e.g., to account for situations where the agent may not fully observe the environment state, or where the only observations available to the agent are certain functions of the underlying state.
+For the sake of clarity, we will present here only the class of POMDPs which are relevant to our work: fully observed POMDPs with trivial emission function (FMDPs for short).
+An FMDP may be defined by the following data:
+- $\mathcal{S}$: *state space*, the set of states of the environment,
+- $\mathcal{A}$: *action space*, the set of actions which the agent may choose from,
+- $T(s_{t+1}|s_t, a_t)$: *transition operator*, a conditional distribution which describes the dynamics of the system,
+- $r(s_t, a_t)$: *reward function*, the reward obtained after performing action $a_t\in\mathcal{A}$ in state $s_t$,
+- $d(s_0)$: *initial state distribution*, the initial state of the environment is sampled from this distribution,
+- $\gamma\in[0,1]$: *discount factor*.
 

### 2023-01-07 manuscript/manuscript.Rmd
mini change

@@
 An FMDP may be defined by the following data:
+
 - $\mathcal{S}$: *state space*, the set of states of the environment,
+
 - $\mathcal{A}$: *action space*, the set of actions which the agent may choose from,
+
 - $T(s_{t+1}|s_t, a_t)$: *transition operator*, a conditional distribution which describes the dynamics of the system,
+
 - $r(s_t, a_t)$: *reward function*, the reward obtained after performing action $a_t\in\mathcal{A}$ in state $s_t$,
+
 - $d(s_0)$: *initial state distribution*, the initial state of the environment is sampled from this distribution,
+
 - $\gamma\in[0,1]$: *discount factor*.

### 2023-01-07 manuscript/manuscript.Rmd
bit more rl

@@
 This formalization is rather flexible and allows one, e.g., to account for situations where the agent may not fully observe the environment state, or where the only observations available to the agent are certain functions of the underlying state.
-For the sake of clarity, we will present here only the class of POMDPs which are relevant to our work: fully observed POMDPs with trivial emission function (FMDPs for short).
+For the sake of clarity, we will present here only the class of POMDPs which are relevant to our work: *fully observable MDPs with a trivial emission function* (FMDPs for short).
 An FMDP may be defined by the following data:
@@
 
+As the name implies, the agent observes the full state of the environment of an FMDP, and chooses its next action based on this observation according to a *policy function* $\pi(a_t | s_t)$.
+In return, the agent receives a discounted reward $\gamma^t r(a_t, s_t)$.
+The discount factor helps regularize the agent, helping the optimization algorithm find solutions which pay off within a timescale of $t \sim \log(\gamma^{-1})$.
+
 ## Acknowledgements

### 2023-01-09 manuscript/manuscript.Rmd
:pen:

--- 2023-01-07 manuscript/manuscript.Rmd
@@
 abstract: |
-  The text of your abstract.  150 -- 250 words.
+  Existing methods for optimal control methods struggle to deal with the complexity commonly encountered in real-world systems, including dimensionality, process error, model bias and heterogeneity. 
+  Instead of tackling these complexities directly, researchers have typically sought to find exact optimal solutions to simplified models of the processes in question. 
+  When is the optimal solution to a very approximate, stylized model better than an approximate solution to a more accurate model? 
+  This question has largely gone unanswered owing to the difficulty of finding even approximate solutions in the case of complex models.  
+  Our approach draws on recent algorithmic and computational advances in deep reinforcement learning. 
+  These methods have hitherto focused on problems in games or robotic mechanics, which operate under precisely known rules. 
+  We demonstrate the ability for novel algorithms using deep neural networks to successfully approximate such solutions (the "policy function" or control rule) without knowing or ever attempting to infer a model for the process itself.
+  This powerful new technique lets us finally begin to answer the question.
+  We show that in many but not all cases, the optimal policy for a carefully chosen over-simplified model can still out-perform these novel algorithms trained to find approximate solutions to simulations of a realistically complex system.
+  This illustrates the promise of combining complex, realistic simulations with emerging machine-learning techniques to improve our understanding and management of ecological systems.
 
@@
 
+<!--
+ Special Issue Description:
+
+This special issue highlights the development of novel data-driven methods, including statistics, machine learning, parameter estimation, and uncertainty quantification, and combinations thereof, towards modeling biological systems.
+These newly developed methods will tackle challenges that are commonly encountered when modeling real-world experimental, field, pre-clinical, or clinical data.
+Examples of such challenges include high dimensionality, computational complexity, observation or process error, model bias, and intra- or inter-individual heterogeneity.
+Contributions to this special issue require validation of new methods with real-world data or simulated data sets that contain features of real-world data that exemplify an outlined modeling challenge.
+Papers should include a discussion justifying why the developed method is novel and not an application of previously developed methods, as well as how the method may be broadly applicable across different areas of biology, including medical, ecological, genetics, and epidemiological applications.
+-->
+
 # Introduction {#intro}
 
-<!-- Your text comes here. Separate text sections with \cite{Mislevy06Cog}. -->
+Despite the complexity inherent in natural ecosystems, analytically tractable models have long played an essential role in their understanding and management.
+Ecological management may be understood as an optimal control problem - a manager must determine a sequence of actions (a "policy") to optimize some desired objective or combination of objectives (conservation goals, natural resource extraction, economic and social constraints.)
 
@@
 
-# Fisheries: approximate models and approximate solutions
-
-<!-- Introduce the 1D and 3D models, and have Figure 1 already here. -->
-
 # Reinforcement learning {#sec:DRL}
@@
 An RL application can be conceptually separated into two parts: an *agent*, and an *environment*.
-The *environment* is commonly a computer simulation, although it sometimes can be a real world system.
+The *environment* is commonly a computer simulation.
 The *agent*, on the other hand, is a computer program which interacts with the environment.
 That is, the agent may act on the environment and change its state.
-This, moreover, gives the agent a *reward* which encodes the control goal.
+Such an action and its consequence, moreover, give the agent a *reward* which encodes our goal for the agent.
 The main part of an RL algorithm is then to progressively improve the agent's *policy*, in order to maximize the cumulative reward received. 
 This is done by aggregating experience and learning from it.
-See Fig. \ref{tbd}.
 
-In our case, the environment will be a computational model of the population dynamics of a fishery, with the environment state being a vector of all the fish populations, $S=(V_1, V_2, H)$.
-The system evolves naturally according to \eqref{eq:3d model}.
-At each time step, the agent harvests one of the populations, say $A$. 
-Specifically, it chooses a quota $q$: the fraction of $A$ that will be fished.
-This changes the state as
-\begin{align*}
-  (V_1,\ V_2,\ H) \mapsto ((1-q)V_1,\ V_2,\ H)
-\end{align*}
-and secures a reward of $qV_1$ to the agent.
-Afterwards, the environment undergoes its natural dynamics.
+In our case, the environment will be a computational model of the population dynamics of a fishery.
+The agent chooses how much to fish from a given species in the environment, thereby changing the populations in the environment's state.
 
-## Mathematical framework for RL
-
-Mathematically, RL may be formulated using a discrete time *partially observable Markov decision process (POMDP)*.
-This formalization is rather flexible and allows one, e.g., to account for situations where the agent may not fully observe the environment state, or where the only observations available to the agent are certain functions of the underlying state.
-For the sake of clarity, we will present here only the class of POMDPs which are relevant to our work: fully observed POMDPs with trivial emission function (FMDPs for short).
-An FMDP may be defined by the following data:
-
-- $\mathcal{S}$: *state space*, the set of states of the environment,
-
-- $\mathcal{A}$: *action space*, the set of actions which the agent may choose from,
-
-- $T(s_{t+1}|s_t, a_t)$: *transition operator*, a conditional distribution which describes the dynamics of the system,
-
-- $r(s_t, a_t)$: *reward function*, the reward obtained after performing action $a_t\in\mathcal{A}$ in state $s_t$,
-
-- $d(s_0)$: *initial state distribution*, the initial state of the environment is sampled from this distribution,
-
-- $\gamma\in[0,1]$: *discount factor*.
+\begin{align}
+a^2+b^2=c^2
+\end{align} -->
 

### 2023-01-09 code/train-agent.py

@@ -1,0 +1,117 @@
+from gym_fishing.envs import FishingCtsEnv
+from gym_fishing.envs import forageVVHcont
+from ray.rllib.agents.ppo import PPOTrainer
+from ray.tune import register_env
+
+
+## We could call env directly without this if only  our envs took a env_config dict argument
+register_env("fish-3sp", lambda config: forageVVHcont())
+
+# Configure the algorithm.
+config = {
+    "env": "fish-3sp",
+    "num_workers": 1,
+    "num_envs_per_worker": 20,
+    "resources": {
+      "num_cpus_per_worker": 12,
+    },
+    "framework": "torch",
+    "num_gpus": 1,
+    "log_level": "ERROR",
+    "create_env_on_local_worker": True  # need this to restore env from checkpoint
+}
+
+trainer = PPOTrainer(config=config)
+iterations = 250
+
+for _ in range(iterations):
+    trainer.train()
+    
+checkpoint = trainer.save("cache")
+
+
+# Evaluate a saved agent:
+checkpoint = ("cache/checkpoint_000{}".format(iterations))
+model = PPOTrainer(config=config)
+model.restore(checkpoint)
+model.evaluate()
+
+# simulate from the saved agent:
+env_config = model.evaluation_config.env_config
+env = model.env_creator(env_config)
+
+import pandas as pd
+import numpy as np
+df = []
+episode_reward = 0
+observation = env.reset()
+for rep in range(10):
+  for t in range(200):
+    action = model.compute_single_action(observation)
+    df.append(np.append([t, rep, action[0], episode_reward], observation))
+    observation, reward, terminated, info = env.step(action)
+    episode_reward += reward
+
+cols = ["t", "rep", "action", "reward", "sp1", "sp2", "sp3"]
+df = pd.DataFrame(df, columns = cols)
+#df.to_csv("msy.csv")
+
+from plotnine import ggplot, geom_point, aes, geom_line, facet_wrap, geom_path
+df2 = (df
+       .melt(id_vars=["t", "action", "reward", "rep"])
+       .groupby(['t', "variable"], as_index=False)
+       .agg({'reward': 'mean',
+             'value': 'mean',,
+             'action': 'mean'})) 
+
+
+(ggplot(df2, aes("t", "value", color="variable")) +
+ geom_line())
+(ggplot(df2, aes("t", "action", color="variable")) + geom_line())
[…]

### 2023-01-09 manuscript/manuscript.Rmd

@@
 An RL application can be conceptually separated into two parts: an *agent*, and an *environment*.
-The *environment* is commonly a computer simulation.
+The *environment* is commonly a computer simulation, although it sometimes can be a real world system.
 The *agent*, on the other hand, is a computer program which interacts with the environment.
 That is, the agent may act on the environment and change its state.
-Such an action and its consequence, moreover, give the agent a *reward* which encodes our goal for the agent.
+This, moreover, gives the agent a *reward* which encodes the control goal.
 The main part of an RL algorithm is then to progressively improve the agent's *policy*, in order to maximize the cumulative reward received. 
 This is done by aggregating experience and learning from it.
+See Fig. \ref{tbd}.
 
-In our case, the environment will be a computational model of the population dynamics of a fishery.
-The agent chooses how much to fish from a given species in the environment, thereby changing the populations in the environment's state.
+In our case, the environment will be a computational model of the population dynamics of a fishery, with the environment state being a vector of all the fish populations, $S=(V_1, V_2, H)$.
+The system evolves naturally according to \eqref{eq:3d model}.
+At each time step, the agent harvests one of the populations, say $A$. 
+Specifically, it chooses a quota $q$: the fraction of $A$ that will be fished.
+This changes the state as
+\begin{align*}
+  (V_1,\ V_2,\ H) \mapsto ((1-q)V_1,\ V_2,\ H)
+\end{align*}
+and secures a reward of $qV_1$ to the agent.
+Afterwards, the environment undergoes its natural dynamics.
 
-\begin{align}
-a^2+b^2=c^2
-\end{align} -->
+## Mathematical framework for RL
+
+Mathematically, RL may be formulated using a discrete time *partially observable Markov decision process (POMDP)*.
+This formalization is rather flexible and allows one, e.g., to account for situations where the agent may not fully observe the environment state, or where the only observations available to the agent are certain functions of the underlying state.
+For the sake of clarity, we will present here only the class of POMDPs which are relevant to our work: *fully observable MDPs with a trivial emission function* (FMDPs for short).
+An FMDP may be defined by the following data:
+
+- $\mathcal{S}$: *state space*, the set of states of the environment,
+
+- $\mathcal{A}$: *action space*, the set of actions which the agent may choose from,
+
+- $T(s_{t+1}|s_t, a_t)$: *transition operator*, a conditional distribution which describes the dynamics of the system,
+
+- $r(s_t, a_t)$: *reward function*, the reward obtained after performing action $a_t\in\mathcal{A}$ in state $s_t$,
+
+- $d(s_0)$: *initial state distribution*, the initial state of the environment is sampled from this distribution,
+
+- $\gamma\in[0,1]$: *discount factor*.
+
+As the name implies, the agent observes the full state of the environment of an FMDP, and chooses its next action based on this observation according to a *policy function* $\pi(a_t | s_t)$.
+In return, the agent receives a discounted reward $\gamma^t r(a_t, s_t)$.
+The discount factor helps regularize the agent, helping the optimization algorithm find solutions which pay off within a timescale of $t \sim \log(\gamma^{-1})$.
 

### 2023-01-09 manuscript/manuscript.Rmd
a bit of rewording

--- 2023-01-07 manuscript/manuscript.Rmd
@@
 
-Reinforcement learning (RL) is a way of approaching *control problems* through machine learning.
-An RL application can be conceptually separated into two parts: an *agent*, and an *environment*.
-The *environment* is commonly a computer simulation, although it sometimes can be a real world system.
-The *agent*, on the other hand, is a computer program which interacts with the environment.
-That is, the agent may act on the environment and change its state.
-This, moreover, gives the agent a *reward* which encodes the control goal.
+Reinforcement learning (RL) is, in a nutshell, a way of approaching *control problems* through machine learning.
+An RL algorithm can be conceptually separated into two parts: an *agent*, and an *environment* which the agent can interact with.
+That is, the agent may act on the environment and thus change its state, while the environment gives a *reward* to the agent in return (see Fig. \ref{tbd}).
+The rewards encode the agent's goal.
 The main part of an RL algorithm is then to progressively improve the agent's *policy*, in order to maximize the cumulative reward received. 
 This is done by aggregating experience and learning from it.
-See Fig. \ref{tbd}.
 
-In our case, the environment will be a computational model of the population dynamics of a fishery, with the environment state being a vector of all the fish populations, $S=(V_1, V_2, H)$.
-The system evolves naturally according to \eqref{eq:3d model}.
-At each time step, the agent harvests one of the populations, say $A$. 
-Specifically, it chooses a quota $q$: the fraction of $A$ that will be fished.
+<!-- The *environment* is commonly a computer simulation, although it sometimes can be a real world system.
+The *agent*, on the other hand, is a computer program which interacts with the environment. -->
+
+For our use case, the environment will be a computational model of the population dynamics of a fishery, with the environment state being a vector of all the fish populations, $S=(V_1, V_2, H)$.
+At each time step, the agent harvests one of the populations, $V_1$. 
 This changes the state as
 \begin{align*}
-  (V_1,\ V_2,\ H) \mapsto ((1-q)V_1,\ V_2,\ H)
+  (V_1,\ V_2,\ H) \mapsto ((1-q)V_1,\ V_2,\ H),
 \end{align*}
-and secures a reward of $qV_1$ to the agent.
-Afterwards, the environment undergoes its natural dynamics.
+where $q$ is a fishing *quota* set by the agent. 
+This secures a reward of $qV_1$.
+Afterwards, the environment undergoes a timestep under its natural dynamics given by \eqref{eq:3d model}.
 
@@
 
-As the name implies, the agent observes the full state of the environment of an FMDP, and chooses its next action based on this observation according to a *policy function* $\pi(a_t | s_t)$.
-In return, the agent receives a discounted reward $\gamma^t r(a_t, s_t)$.
-The discount factor helps regularize the agent, helping the optimization algorithm find solutions which pay off within a timescale of $t \sim \log(\gamma^{-1})$.
+At a time $t$, the FMDP agent observes the full state $s_t$ of the environment and chooses an action based on this observation according to a *policy function* $\pi(a_t | s_t)$.
+In return, it receives a discounted reward $\gamma^t r(a_t, s_t)$.
+The discount factor helps regularize the agent, helping the optimization algorithm find solutions which pay off within a timescale of $t \sim \log(\gamma^{-1})^{-1}$.
 

### 2023-01-09 manuscript/manuscript.Rmd
value function, etc

@@
 
+With any fixed policy function, the agent will traverse a path 
+$\tau=(s_0,\ a_0,\ s_1,\ a_1\ \dots,\ s_{t_{\text{fin.}}})$ 
+sampled randomly from the distribution
+\begin{align*}
+  p_\pi(\tau) = 
+  d(s_0) \prod_{t=0}^{ t_{\text{fin.}}-1 }
+  \pi( a_t | s_t ) T( s_{t+1} | s_t, a_t ).
+\end{align*}
+Reinforcement learning seeks to optimize $\pi$ such that the expected rewards are maximal,
+$$  
+  \pi^* = \mathrm{argmax}\ \mathbb{E}_{\tau\sim p_\pi}[R(\tau)],
+$$
+where
+$$
+  R(\tau) = \sum_{t=0}^{ t_{\text{fin.}}-1 } \gamma^t r(a_t, s_t)
+$$
+is the cumulative reward of path $\tau$.
+The function $J(\pi):=\mathbb{E}_{\tau\sim p_\pi}[R(\tau)]$ is called the *value function*.
+
 ## Acknowledgements

### 2023-01-09 manuscript/manuscript.Rmd
drl

@@
 
+## Deep Reinforcement Learning
+
+The policy function $\pi$ often lives in a high- or infinite-dimensional space.
+This makes it unfeasible to directly optimize $\pi$.
+In practice, an alternative approach is used: $\pi$ is optimized over a much lower-dimensional parametrized family of functions.
+Deep reinforcement learning uses this strategy, focusing on function families parametrized by neural networks. (See Fig. \ref{tbd}.)
+
 ## Acknowledgements

### 2023-01-09 manuscript/manuscript.Rmd
more drl

@@
 Figure 3: The 1-D optimal management solution.  'constant escapement' intuition etc
+
+RL figures:
+- schematic of RL (a-la previous marcus paper Fig 1)
+- Neural network optimization figure
 
@@
 $$
-where
+where,
 $$
-  R(\tau) = \sum_{t=0}^{ t_{\text{fin.}}-1 } \gamma^t r(a_t, s_t)
+  R(\tau) = \sum_{t=0}^{ t_{\text{fin.}}-1 } \gamma^t r(a_t, s_t),
 $$
@@
 
-The policy function $\pi$ often lives in a high- or infinite-dimensional space.
+The policy function $\pi$ often lives in a high- or even infinite-dimensional space.
 This makes it unfeasible to directly optimize $\pi$.
@@
 Deep reinforcement learning uses this strategy, focusing on function families parametrized by neural networks. (See Fig. \ref{tbd}.)
+<!-- The figure should kinda explain how the parametrization arises, and how gradient ascent is used. Maybe namedrop backpropagataion for the sake of usefulness to the reader? -->
+
+Since our state and observation spaces are continuous, we will focus on deep reinforcement learning throughout this paper.
+Specifically, we parametrize $\pi$ using a XXXX neural network and use the *proximal policy optimization (PPO)* algorithm to optimize $\pi$.
+Within the RL literature there is a wealth of algorithms from which to choose from, each with its pros and cons.
+Here we have focused on a single one of these for the sake of particularity---to draw a clear comparison between the RL-based and the classical fishery management approaches.
+In practice, further improvements can be expected by a careful selection of the optimization algorithm.
 

### 2023-01-09 manuscript/manuscript.Rmd
mini

@@
 Since our state and observation spaces are continuous, we will focus on deep reinforcement learning throughout this paper.
-Specifically, we parametrize $\pi$ using a XXXX neural network and use the *proximal policy optimization (PPO)* algorithm to optimize $\pi$.
+Specifically, we parametrize $\pi$ using a neural network with two hidden layers of 64 neurons.
+
+We use the *proximal policy optimization (PPO)* algorithm to optimize $\pi$.
 Within the RL literature there is a wealth of algorithms from which to choose from, each with its pros and cons.

### 2023-01-09 manuscript/manuscript.Rmd
added figure thought

@@
 
+Figure 0: contrast 1D vs 3D strategies (see commented out diagram).
+
+<!-- Reality __
+  |        \
+  v         \
+1-D Model   More expressive model
+  |                  |
+  v                  v
+Optimal        Suboptimal, but
+solution   well-performing solution -->
+
 Figure 1: 1-D and 3-D model conceptual figure. something about the objective / decision
 Figure 2: Stability / multistability in the 1D and 3D models.  state space + time views
-
 Figure 3: The 1-D optimal management solution.  'constant escapement' intuition etc
@@
 <!-- Introduce the 1D and 3D models, and have Figure 1 already here. -->
+
+Here we review classical strategies for sustainable fishery management and we provide a birds-eye view of the alternative approach we propose.
+A summary of the contrast between the two strategies is given in Fig. \ref{tbd}.
+
+## Classical strategies: MSY, escapement, total allowable catch
 

### 2023-01-09 manuscript/manuscript.Rmd
model-based

@@
 
-## Classical strategies: MSY, escapement, total allowable catch
+## Classical approaches to sustainable fisheries {#sec:classical}
 
-# Reinforcement learning {#sec:DRL}
+There are several strategies that have been used to manage fisheries: *escapement*, *maximum sustainable yield (MSY)*, *total allowable catch (TAC)*, among others.
+We collectively refere to these as *classical*, and will compare their performance to RL-based management strategies.
+As shown in Fig.
+\ref{fig:tbd}, classical strategies have the common aspect of reducing the complex dynamics of the fishery ecosystem to a single equation governing the harvested population (say, $F$).
+A common example is using a logistic growth equation,
+$$
+  F_{t+1} - F_t = r F_t( 1 - F_t / K),
+$$
+where the interaction between $F$ and its environment is summarized to two parameters, the growth rate $r$, and the carrying capacity $K$.
+
+# Reinforcement learning {#sec:RL}
 
@@
 
+## Model-free reinforcement learning
+
+Within control theory, the usual setup is one where we use as much information from the model as possible in order to derive an optimal solution.
+**(Wax poetic a bit about Bellmann eq. approaches here?)**
+
+The classical sustainable fishery management approaches summarized in Sec. \ref{sec:classical} are model-based controls.
+As we saw in that section, these controls may run into trouble in the case where there are inaccuracies in the model parameter estimates.
+
+More generally, there are many situations in which the exact model of the system is not known or not tractable.
+This is a standard situation in ecology: mathematical models capture the most prominent aspects of the ecosystem's dynamics, while ignoring or summarizing most of its complexity.
+In this case, it is clear, model-based controls run a grave danger of mismanaging the system.
+
+Reinforcement learning is, on the other hand, typically a model-free approach to control theory.
+\footnote{There are, however, approaches to perform model-based reinforcement learning. While we will not focus on these in this paper they are discussed in \cite{tbd1, tbd2}.}
+While the model is certainly used to generate training data, it is not directly used by the optimization algorithm.
+This provides more flexibility to use model-free RL in instances where the model of the system is not accurately known.
+In fact, it has been shown to generally be preferable to use model-free RL in such instances.
+
+**Mention CL here? Where solutions to other related problems in the curriculum are used to build solutions to the target problem.**
+
+This context provides a motivation for this paper. 
+Indeed, models for ecosystem dynamics are only ever approximate and incomplete descriptions of reality. 
+This way, it is plausible that model-free RL controls outperform currently used model-based controls in ecological management problems.
+
 ## Acknowledgements

### 2023-01-09 manuscript/git_script.sh
added nifty git script

@@ -1,0 +1,4 @@
+pandoc -o manuscript.pdf manuscript.Rmd
+git commit -am "$1"
+git pull
+git push

### 2023-01-09 manuscript/manuscript.Rmd
mini

@@
 
-Reinforcement learning is, on the other hand, typically a model-free approach to control theory.
-\footnote{There are, however, approaches to perform model-based reinforcement learning. While we will not focus on these in this paper they are discussed in \cite{tbd1, tbd2}.}
+Reinforcement learning is, on the other hand, typically a model-free approach to control theory.\footnote{There are, however, approaches to perform model-based reinforcement learning.
+While we will not focus on these in this paper they are discussed in \cite{tbd1, tbd2}.}
 While the model is certainly used to generate training data, it is not directly used by the optimization algorithm.
@@
 
+# Results
+
 ## Acknowledgements

### 2023-01-09 manuscript/manuscript.Rmd
msy, etc

@@
 $$
-  F_{t+1} - F_t = r F_t( 1 - F_t / K),
+  F_{t+1} - F_t = r F_t( 1 - F_t / K) - h_t =: L(F_t) - h_t,
 $$
 where the interaction between $F$ and its environment is summarized to two parameters, the growth rate $r$, and the carrying capacity $K$.
+In the equation above, $h_t$ is the *harvest* at timestep $t$.
+The goal is to choose the harvest policy $h:F_t\mapsto h_t$ such that long-term profits are maximized.
+
+An advantage of one dimensional approaches is that the optimal policy is often known exactly, and, moreover, is intuitive.
+For example, in the logistic equation pointed out above, the maximal sustainable yield of the system is attained at $F = F_{MSY} := K/2$.
+The optimizer is an *escapement* policy:
+$$
+  h_t = \begin{cases}
+    F_t - K/2, \qquad &\text{if, } F_t > K/2\\
+    0, \qquad &\text{else.}
+  \end{cases}
+$$
+This corresponds to keeping the system at its optimal growth rate as much as possible.
+
+Escapement policies, or more generally *bang bang* policies, tend to be the optimal solution for these types of control problems.
+A drawback of these solutions, in the fishery context, is the presence of several timesteps with zero harvest which can arise.
+To mend this, certain suboptimal solutions have been constructed for fishery management.
+
+One ubiquitous solution is simply called maximum sustainable yield (MSY).
+It consists on letting $h(F) = rF/2$, so that
+$$
+  h(F_{MSY}) = L(F_{MSY}) = r K/4.
+$$
+That is, at the MSY biomass, the logistic growth of $F$ is cancelled exactly by the harvest.
+
+The MSY rule fixes the drawback in the escapement policy by having $h(F)>0$ for all $F>0$.
 

### 2023-01-09 manuscript/manuscript.Rmd
more msy

@@
 The MSY rule fixes the drawback in the escapement policy by having $h(F)>0$ for all $F>0$.
+It, however, has its own drawbacks.
+It is particularly sensitive to misestimates of the parameter $r$, as we will discuss in Sec. \ref{sec:tbd}.
+Due to this, similar but more conservative policies have been used.
+
+*Total allowable catch (TAC)* is one such policy.
+It consists on reducing the inclination of the line defined by $h(F)$, for example using $h(F) = 0.8\times r/2$.
 

### 2023-01-09 manuscript/manuscript.Rmd
TAC, variable prefactor, etc

@@
 *Total allowable catch (TAC)* is one such policy.
-It consists on reducing the inclination of the line defined by $h(F)$, for example using $h(F) = 0.8\times r/2$.
+It consists on reducing the inclination of the line defined by $h(F)$ using a prefactor $\alpha$ in $h(F) = \alpha\times rF/2$. 
+Plausible examples are $\alpha = 0.8$ or $0.9$.
+Another common alternative is to have a prefactor $\alpha = \alpha(F)$ which decreases from one to zero as $F$ decreases below a threshold:
+$$
+  h(F) = \alpha(F)\times rF/2,
+$$
+where,
+$$
+  \alpha(F) = \begin{cases}
+  1, \qquad &\text{if } F > F_{\text{thresh.}},\\
+  F / F_{\text{thresh.}}, \qquad & \text{else}.
+  \end{cases}
+$$
 

### 2023-01-10 manuscript/manuscript.Rmd
mini

@@
 $$
+We call this policy *variable total allowable catch (VTAC)*.
+
+## Dangers: Noisy parameters, noisy observations and model inaccuracy
 

### 2023-01-11 manuscript/manuscript.Rmd
models

--- 2023-01-09 manuscript/manuscript.Rmd
@@
 
-# Fisheries: approximate models and approximate solutions
+# Mathematical models of fisheries
+
+In this section we introduce some models describing the population dynamics of fisheries.
+In general, the class of models that appear in this context are *first order finite difference equations*.
+For $n$ species, these models have the general form
+\begin{align}
+  \label{eq:general model}
+  N_{t+1} - N_t = f(N_t) - h_t,
+\end{align}
+where $N(t) = (A_t,\ B_t,\ \dots) \in \mathbb{R}^{n}_+$ is a vector of populations, $f:\mathbb{R}^{n}\to\mathbb{R}^{n}$ is an arbitrary function, and $h_t$ is the harvest size at timestep $t$.
+
+Most commonly, fisheries use 1-dimensional models, $n=1$.
+One such model appearing in a variety of ecological contexts is *logistic growth*, for which
+$$
+  f(N_t) = r N_t\big(1 - N_t / K \big) =: L(N_t).
+$$
+More relevant to our current work, is the following single-species model studied in \cite{may77},
+\begin{align}
+  \label{eq:may}
+  N_{t+1} - N_t
+  = L(N_t) - \frac{\beta H N_t^2}{c^2 + N_t^2}.
+\end{align}
+The model has five parameters: a constant population $H$ of a species which preys on $N$, the maximal rate of predation $\beta$, and the predation half-maximum biomass $c$.
+
+Part of what makes eq. \eqref{eq:may} an interesting study case is the presence of a *tipping point*.
+
+# Fishery management approaches
 

### 2023-01-12 code/escapement.py
move msy example to seperate file

@@ -1,0 +1,45 @@
+import gym_fishing
+import gym
+import pandas as pd
+import numpy as np
+
+env = gym.make("threeFishing-v2")
+
+df = []
+
+effort * pop = harvest
+escapement = pop - harvest
+harvest = pop - escapement
+effort * pop = pop - escapement
+effort = (pop - escapement)/pop
+
+actions = np.linspace(0,1,101)
+for escapement in actions:
+  episode_reward = 0
+  observation = env.reset()
+  for rep in range(10):
+    for t in range(200):
+      df.append(np.append([t, rep, action, episode_reward], observation))
+      effort = np.min(1 - escapement / observation[0], 0)
+      observation, reward, terminated, info = env.step(effort)
+      episode_reward += reward
+
+cols = ["t", "rep", "action", "reward", "sp1", "sp2", "sp3"]
+df = pd.DataFrame(df, columns = cols)
+df.to_csv("msy.csv")
+
+
+
+# identify action at associated with highest cumulative mean reward by last timestep
+df2 = ( df
+        .groupby(['t','action'], as_index=False)
+        .agg({'reward': 'mean',
+              'sp1': 'mean',
+              'sp2': 'mean',
+              'sp3': 'mean'})
+        .melt(id_vars=["t", "action", "reward"]) 
+        )
+best_action = df2.query("t==199")[df2.reward == max(df2.reward)].action.values[0]
+df3 = df.query("action == " + str(best_action))
+                                
+

### 2023-01-12 code/msy.py

@@ -1,0 +1,37 @@
+import gym_fishing
+import gym
+import pandas as pd
+import numpy as np
+
+env = gym.make("threeFishing-v2")
+
+df = []
+actions = np.linspace(0,.1,101)
+for action in actions:
+  episode_reward = 0
+  observation = env.reset()
+  for rep in range(10):
+    for t in range(200):
+      df.append(np.append([t, rep, action, episode_reward], observation))
+      observation, reward, terminated, info = env.step(action)
+      episode_reward += reward
+
+cols = ["t", "rep", "action", "reward", "sp1", "sp2", "sp3"]
+df = pd.DataFrame(df, columns = cols)
+df.to_csv("msy.csv")
+
+
+
+# identify action at associated with highest cumulative mean reward by last timestep
+df2 = ( df
+        .groupby(['t','action'], as_index=False)
+        .agg({'reward': 'mean',
+              'sp1': 'mean',
+              'sp2': 'mean',
+              'sp3': 'mean'})
+        .melt(id_vars=["t", "action", "reward"]) 
+        )
+best_action = df2.query("t==199")[df2.reward == max(df2.reward)].action.values[0]
+df3 = df.query("action == " + str(best_action))
+                                
+

### 2023-01-12 code/train-agent.py

@@ -1,8 +1,7 @@
-from gym_fishing.envs import FishingCtsEnv
 from gym_fishing.envs import forageVVHcont
 from ray.rllib.agents.ppo import PPOTrainer
 from ray.tune import register_env
-
+import os
 
 ## We could call env directly without this if only  our envs took a env_config dict argument
 register_env("fish-3sp", lambda config: forageVVHcont())
@@ -18,27 +17,27 @@ config = {
     "framework": "torch",
     "num_gpus": 1,
     "log_level": "ERROR",
-    "create_env_on_local_worker": True  # need this to restore env from checkpoint
+    "create_env_on_local_worker": True  # needed to restore env from checkpoint
 }
 
-trainer = PPOTrainer(config=config)
+agent = PPOTrainer(config=config)
 iterations = 250
-
-for _ in range(iterations):
-    trainer.train()
-    
-checkpoint = trainer.save("cache")
-
-
-# Evaluate a saved agent:
 checkpoint = ("cache/checkpoint_000{}".format(iterations))
-model = PPOTrainer(config=config)
-model.restore(checkpoint)
-model.evaluate()
 
-# simulate from the saved agent:
-env_config = model.evaluation_config.env_config
-env = model.env_creator(env_config)
+if not os.path.exists(checkpoint): # train only if no trained agent saved
+  for _ in range(iterations):
+      agent.train()
+  checkpoint = agent.save("cache")
+
+
+# Restore saved agent:
+agent = PPOTrainer(config=config)
+agent.restore(checkpoint)
+
+# agent.evaluate() # built-in method to evaluate agent on eval env
+
+# Initialize saved copy of eval environment:
+env = agent.env_creator(agent.evaluation_config.env_config)
 
 import pandas as pd
 import numpy as np
@@ -47,14 +46,16 @@ episode_reward = 0
 observation = env.reset()
 for rep in range(10):
   for t in range(200):
-    action = model.compute_single_action(observation)
+    action = agent.compute_single_action(observation)
     df.append(np.append([t, rep, action[0], episode_reward], observation))
     observation, reward, terminated, info = env.step(action)
     episode_reward += reward
 
 cols = ["t", "rep", "action", "reward", "sp1", "sp2", "sp3"]
 df = pd.DataFrame(df, columns = cols)
-#df.to_csv("msy.csv")
+df.to_csv("PPO.csv")
+
+
 
[…]

### 2023-01-12 code/msy.py
allow renv to handle python

@@ -11,6 +11,7 @@ for action in actions:
   episode_reward = 0
   observation = env.reset()
   for rep in range(10):
+    _ = env.reset()
     for t in range(200):
       df.append(np.append([t, rep, action, episode_reward], observation))
       observation, reward, terminated, info = env.step(action)

### 2023-01-12 code/escapement.py
reset each rep

@@ -16,8 +16,8 @@ effort = (pop - escapement)/pop
 actions = np.linspace(0,1,101)
 for escapement in actions:
   episode_reward = 0
-  observation = env.reset()
   for rep in range(10):
+    observation = env.reset()
     for t in range(200):
       df.append(np.append([t, rep, action, episode_reward], observation))
       effort = np.min(1 - escapement / observation[0], 0)

### 2023-01-12 code/msy.py

@@ -8,10 +8,9 @@ env = gym.make("threeFishing-v2")
 df = []
 actions = np.linspace(0,.1,101)
 for action in actions:
-  episode_reward = 0
-  observation = env.reset()
   for rep in range(10):
-    _ = env.reset()
+    episode_reward = 0
+    observation = env.reset()
     for t in range(200):
       df.append(np.append([t, rep, action, episode_reward], observation))
       observation, reward, terminated, info = env.step(action)

### 2023-01-12 code/train-agent.py

@@ -42,9 +42,10 @@ env = agent.env_creator(agent.evaluation_config.env_config)
 import pandas as pd
 import numpy as np
 df = []
-episode_reward = 0
-observation = env.reset()
+
 for rep in range(10):
+  episode_reward = 0
+  observation = env.reset()
   for t in range(200):
     action = agent.compute_single_action(observation)
     df.append(np.append([t, rep, action[0], episode_reward], observation))
@@ -54,8 +55,6 @@ for rep in range(10):
 cols = ["t", "rep", "action", "reward", "sp1", "sp2", "sp3"]
 df = pd.DataFrame(df, columns = cols)
 df.to_csv("PPO.csv")
-
-
 
 from plotnine import ggplot, geom_point, aes, geom_line, facet_wrap, geom_path
 df2 = (df

### 2023-01-12 code/plots.Rmd
paths

---
---

```{r setup}
library(tidyverse)
```

```{r}
## reticulate::source_python("msy.py")
```

```{r}
escapement <- read_csv("../data/escapement.csv")
escapement |> 
  filter(t == max(t)) |> 
  group_by(action) |> summarise(reward = mean(reward)) |> 
  filter(reward == max(reward))
```

### 2023-01-12 code/escapement.py

@@ -3,43 +3,28 @@ import gym
 import pandas as pd
 import numpy as np
 
+# Note that state space is mapped to [-1,1]
+#
+# effort corresponding to constant escapement is
+# effort = np.min(1 - escapement / observation[0], 0)
+
+actions = np.linspace(-1,1,201)
 env = gym.make("threeFishing-v2")
-
 df = []
-
-effort * pop = harvest
-escapement = pop - harvest
-harvest = pop - escapement
-effort * pop = pop - escapement
-effort = (pop - escapement)/pop
-
-actions = np.linspace(0,1,101)
-for escapement in actions:
-  episode_reward = 0
+for action in actions:
   for rep in range(10):
+    episode_reward = 0
     observation = env.reset()
     for t in range(200):
       df.append(np.append([t, rep, action, episode_reward], observation))
-      effort = np.min(1 - escapement / observation[0], 0)
+      effort = np.min(1 - action / observation[0], 0)
       observation, reward, terminated, info = env.step(effort)
       episode_reward += reward
 
 cols = ["t", "rep", "action", "reward", "sp1", "sp2", "sp3"]
 df = pd.DataFrame(df, columns = cols)
-df.to_csv("msy.csv")
+df.to_csv("data/escapement.csv.gz", index=False)
 
 
 
-# identify action at associated with highest cumulative mean reward by last timestep
-df2 = ( df
-        .groupby(['t','action'], as_index=False)
-        .agg({'reward': 'mean',
-              'sp1': 'mean',
-              'sp2': 'mean',
-              'sp3': 'mean'})
-        .melt(id_vars=["t", "action", "reward"]) 
-        )
-best_action = df2.query("t==199")[df2.reward == max(df2.reward)].action.values[0]
-df3 = df.query("action == " + str(best_action))
-                                
 

### 2023-01-12 code/msy.py

@@ -4,9 +4,8 @@ import pandas as pd
 import numpy as np
 
 env = gym.make("threeFishing-v2")
-
-df = []
 actions = np.linspace(0,.1,101)
+df = []
 for action in actions:
   for rep in range(10):
     episode_reward = 0
@@ -18,10 +17,13 @@ for action in actions:
 
 cols = ["t", "rep", "action", "reward", "sp1", "sp2", "sp3"]
 df = pd.DataFrame(df, columns = cols)
-df.to_csv("msy.csv")
+df.to_csv("data/msy.csv", index=False)
 
 
 
+
+
+# ignore me, we'll transform and plot in R
 # identify action at associated with highest cumulative mean reward by last timestep
 df2 = ( df
         .groupby(['t','action'], as_index=False)

### 2023-01-12 code/train-agent.py

@@ -54,7 +54,7 @@ for rep in range(10):
 
 cols = ["t", "rep", "action", "reward", "sp1", "sp2", "sp3"]
 df = pd.DataFrame(df, columns = cols)
-df.to_csv("PPO.csv")
+df.to_csv("data/PPO.csv")
 
 from plotnine import ggplot, geom_point, aes, geom_line, facet_wrap, geom_path
 df2 = (df

### 2023-01-12 code/train-agent.py
escapement

@@ -61,7 +61,7 @@ df2 = (df
        .melt(id_vars=["t", "action", "reward", "rep"])
        .groupby(['t', "variable"], as_index=False)
        .agg({'reward': 'mean',
-             'value': 'mean',,
+             'value': 'mean',
              'action': 'mean'})) 
 
 

### 2023-01-12 code/plots.Rmd
PPO has learned Reed '79

@@
 ```{r}
-## reticulate::source_python("msy.py")
+ppo <- read_csv("../data/PPO.csv")
 ```
@@
 ```{r}
-escapement <- read_csv("../data/escapement.csv")
-escapement |> 
+ppo |> 
+  mutate(pop = (sp1+1)/2, 
+         escapement = pop - action * pop) |> 
+  ggplot(aes(t, escapement, group=rep)) + geom_line() + ylim(0,1)
+```
+
+```{r}
+msy <- read_csv("../data/msy.csv")
+best <- msy |> 
   filter(t == max(t)) |> 
@@
   filter(reward == max(reward))
+best
 ```
+```{r}
+msy_sim <- msy |>
+  filter(action == best$action) |> 
+  pivot_longer(c("sp1", "sp2", "sp3"), names_to = "species", values_to = "abundance")
+
+msy_sim |>
+  ggplot(aes(t, abundance, group = interaction(species, rep), col=species)) + 
+    geom_line(alpha=0.5)
+
+```
+
+```{r}
+escapement <- read_csv("../data/escapement.csv.gz")
+best_e <- escapement |> 
+  filter(t == max(t)) |> 
+  group_by(action) |> summarise(reward = mean(reward)) |> 
+  filter(reward == max(reward))
+best_e
+```
+
+```{r}
+esc_sims <- escapement |>
+  filter(action == best_e$action) |> 
+  mutate(effort = pmax(1 - action / sp1, 0)) |>
+  pivot_longer(c("sp1", "sp2", "sp3"), names_to = "species", values_to = "abundance")
+
+esc_sims |>
+  ggplot(aes(t, abundance, group = interaction(species, rep), col=species)) + geom_line(alpha=0.2)
+
+esc_sims |> ggplot(aes(t,action, group=rep)) + geom_line(alpha=.2)
+
+esc_sims |> ggplot(aes(t,effort, group=rep)) + geom_line(alpha=.2)
+```

### 2023-01-12 code/escapement.py

@@ -6,10 +6,11 @@ import numpy as np
 # Note that state space is mapped to [-1,1]
 #
 # effort corresponding to constant escapement is
-# effort = np.min(1 - escapement / observation[0], 0)
+# effort = max(1 - escapement / sp1_pop, 0)
 
-actions = np.linspace(-1,1,201)
+actions = np.linspace(0,1,101)
 env = gym.make("threeFishing-v2")
+env.reset()
 df = []
 for action in actions:
   for rep in range(10):
@@ -17,8 +18,11 @@ for action in actions:
     observation = env.reset()
     for t in range(200):
       df.append(np.append([t, rep, action, episode_reward], observation))
-      effort = np.min(1 - action / observation[0], 0)
+      sp1_pop = (observation[0] + 1 ) / 2 # natural state-space
+      effort = np.max([1 - action / sp1_pop, 0])
       observation, reward, terminated, info = env.step(effort)
+      if terminated:
+        break
       episode_reward += reward
 
 cols = ["t", "rep", "action", "reward", "sp1", "sp2", "sp3"]

### 2023-01-12 code/train-agent.py

@@ -54,7 +54,7 @@ for rep in range(10):
 
 cols = ["t", "rep", "action", "reward", "sp1", "sp2", "sp3"]
 df = pd.DataFrame(df, columns = cols)
-df.to_csv("data/PPO.csv")
+df.to_csv("data/PPO.csv", index = False)
 
 from plotnine import ggplot, geom_point, aes, geom_line, facet_wrap, geom_path
 df2 = (df

### 2023-01-12 code/plots.Rmd
tidy up

@@
 ```{r}
-ppo <- read_csv("../data/PPO.csv")
+ppo <- read_csv("../data/PPO.csv.gz")
 ```
@@
 ```{r}
-msy <- read_csv("../data/msy.csv")
+msy <- read_csv("../data/msy.csv.gz")
 best <- msy |> 

### 2023-01-12 code/msy.py

@@ -17,7 +17,7 @@ for action in actions:
 
 cols = ["t", "rep", "action", "reward", "sp1", "sp2", "sp3"]
 df = pd.DataFrame(df, columns = cols)
-df.to_csv("data/msy.csv", index=False)
+df.to_csv("data/msy.csv.gz", index=False)
 
 
 

### 2023-01-12 code/train-agent.py

@@ -2,6 +2,8 @@ from gym_fishing.envs import forageVVHcont
 from ray.rllib.agents.ppo import PPOTrainer
 from ray.tune import register_env
 import os
+import pandas as pd
+import numpy as np
 
 ## We could call env directly without this if only  our envs took a env_config dict argument
 register_env("fish-3sp", lambda config: forageVVHcont())
@@ -29,7 +31,6 @@ if not os.path.exists(checkpoint): # train only if no trained agent saved
       agent.train()
   checkpoint = agent.save("cache")
 
-
 # Restore saved agent:
 agent = PPOTrainer(config=config)
 agent.restore(checkpoint)
@@ -39,10 +40,7 @@ agent.restore(checkpoint)
 # Initialize saved copy of eval environment:
 env = agent.env_creator(agent.evaluation_config.env_config)
 
-import pandas as pd
-import numpy as np
 df = []
-
 for rep in range(10):
   episode_reward = 0
   observation = env.reset()
@@ -54,26 +52,4 @@ for rep in range(10):
 
 cols = ["t", "rep", "action", "reward", "sp1", "sp2", "sp3"]
 df = pd.DataFrame(df, columns = cols)
-df.to_csv("data/PPO.csv", index = False)
-
-from plotnine import ggplot, geom_point, aes, geom_line, facet_wrap, geom_path
-df2 = (df
-       .melt(id_vars=["t", "action", "reward", "rep"])
-       .groupby(['t', "variable"], as_index=False)
-       .agg({'reward': 'mean',
-             'value': 'mean',
-             'action': 'mean'})) 
-
-
-(ggplot(df2, aes("t", "value", color="variable")) +
- geom_line())
-(ggplot(df2, aes("t", "action", color="variable")) + geom_line())
-(ggplot(df2, aes("t", "reward", color="variable")) + geom_line())
-
-(ggplot(df, aes("sp2", "sp3", color="action", size="sp1")) + geom_point())
-(ggplot(df, aes("sp1", "sp2", color="action", size="sp3")) + geom_point())
-(ggplot(df, aes("sp1", "sp3", color="action", size="sp2")) + geom_point())
-
-
-
-
+df.to_csv("data/PPO.csv.gz", index = False)

### 2023-01-12 manuscript/manuscript.Rmd
finished 1d section, starting 3d

--- 2023-01-11 manuscript/manuscript.Rmd
@@
 
+## Single species models
+
 Most commonly, fisheries use 1-dimensional models, $n=1$.
@@
 
-Part of what makes eq. \eqref{eq:may} an interesting study case is the presence of a *tipping point*.
+Part of what makes eq. \eqref{eq:may} an interesting study case is the presence of a *tipping point* (see Fig. \ref{tbd}).
+Holding the value of $\beta$ fixed, for intermediate values of $H$ there exist two stable fixed points for the state $V_1$ of the system, these two attractors separated by an unstable fixed point.
+At a certain threshold value of $H$, however, the top stable fixed point collides with the unstable fixed point and both are annihilated.
+For this value of $H$, and for higher values, only the lower fixed point remains.
+
+This structure implies two things.
+First, that a drift in $H$ could lead to catastrophic consequences, with the population $V_1$ plummeting to the lower fixed stable point.
+Second, that if the evolution of $V_1$ is *stochastic*, then, even at values of $H$ below the threshold point, the system runs a sizeable danger of tipping over towards the lower stable point.
+
+## A three species model
 

### 2023-01-13 code/plots.Rmd
plots of fixed action

@@
 ---
+output: github_document
 ---
+
+This notebook generates plots from the `.csv` results produced by the corresponding python scripts.
+We compare policies across constant effort (MSY-like policies), constant escapement, and the RL agent (PPO).
 
@@
 
-```{r}
-ppo <- read_csv("../data/PPO.csv.gz")
-```
-
-```{r}
-ppo |> 
-  mutate(pop = (sp1+1)/2, 
-         escapement = pop - action * pop) |> 
-  ggplot(aes(t, escapement, group=rep)) + geom_line() + ylim(0,1)
-```
+# Constant Effort
 
@@
 ```
+
+# Constant Escapement
 
@@
 ```
+
+# PPO Agent
+
+```{r}
+ppo <- read_csv("../data/PPO-long.csv.gz")
+```
+
+```{r}
+ppo |> ggplot(aes(t, sp1, group=rep)) + geom_line()
+
+ppo |> ggplot(aes(t, reward, group=rep)) + geom_line()
+
+ppo |> 
+  mutate(pop = (sp1+1), 
+         escapement = pop - action * pop) |> 
+  ggplot(aes(t, escapement, group=rep)) + geom_line() + ylim(0,1)
+
+ppo |> 
+  mutate(effort = pmax(1 - action / sp1, 0)) |>
+  pivot_longer(c("sp1", "sp2", "sp3"), names_to = "species", values_to = "abundance") |>
+  ggplot(aes(t, abundance, group = interaction(species, rep), col=species)) +
+  geom_line(alpha=0.8)
+
+```
+
+```{r}
+ppo |> filter(t == max(t)) |> pull(reward) |> mean()
+ppo |> filter(t == max(t)) |> pull(reward) |> sd()
+
+```

### 2023-01-13 code/train-agent.py

@@ -6,11 +6,11 @@ import pandas as pd
 import numpy as np
 
 ## We could call env directly without this if only  our envs took a env_config dict argument
-register_env("fish-3sp", lambda config: forageVVHcont())
+register_env("threeFishing-v2", lambda config: forageVVHcont())
 
 # Configure the algorithm.
 config = {
-    "env": "fish-3sp",
+    "env": "threeFishing-v2",
     "num_workers": 1,
     "num_envs_per_worker": 20,
     "resources": {
@@ -23,7 +23,7 @@ config = {
 }
 
 agent = PPOTrainer(config=config)
-iterations = 250
+iterations = 100
 checkpoint = ("cache/checkpoint_000{}".format(iterations))
 
 if not os.path.exists(checkpoint): # train only if no trained agent saved
@@ -49,7 +49,44 @@ for rep in range(10):
     df.append(np.append([t, rep, action[0], episode_reward], observation))
     observation, reward, terminated, info = env.step(action)
     episode_reward += reward
-
+    
+episode_reward
 cols = ["t", "rep", "action", "reward", "sp1", "sp2", "sp3"]
 df = pd.DataFrame(df, columns = cols)
-df.to_csv("data/PPO.csv.gz", index = False)
+df.to_csv("data/PPO" + str(iterations) + ".csv.gz", index = False)
+
+
+
+
+## Plots exploring agent action submanifolds
+from plotnine import ggplot, geom_point, aes, geom_line, facet_wrap, geom_path
+
+## agent action vs H
+dd = []
+for p in np.linspace(-1, 1, 101):
+  observation = np.append(observation[range(2)], p)
+  dd.append(np.append(p, agent.compute_single_action(observation) ))
+
+df2 = pd.DataFrame(dd, columns = ['H', 'action'])
+(ggplot(df2, aes("H", "action")) +geom_line())
+
+
+dd = []
+observation = env.reset()
+for p in np.linspace(-1, 1, 101):
+  observation = np.array([observation[0], p, observation[2]])
+  dd.append(np.append(p, agent.compute_single_action(observation) ))
+
+df2 = pd.DataFrame(dd, columns = ['V2', 'action'])
+(ggplot(df2, aes("V2", "action")) +geom_line())
+
+dd = []
+observation = env.reset()
+for p in np.linspace(-1, 1, 101):
+  observation = np.array([p, observation[1], observation[2]])
[…]

### 2023-01-14 code/train-agent.py
tidy up

@@ -23,7 +23,7 @@ config = {
 }
 
 agent = PPOTrainer(config=config)
-iterations = 100
+iterations = 250
 checkpoint = ("cache/checkpoint_000{}".format(iterations))
 
 if not os.path.exists(checkpoint): # train only if no trained agent saved
@@ -35,7 +35,7 @@ if not os.path.exists(checkpoint): # train only if no trained agent saved
 agent = PPOTrainer(config=config)
 agent.restore(checkpoint)
 
-# agent.evaluate() # built-in method to evaluate agent on eval env
+stats = agent.evaluate() # built-in method to evaluate agent on eval env
 
 # Initialize saved copy of eval environment:
 env = agent.env_creator(agent.evaluation_config.env_config)
@@ -50,7 +50,6 @@ for rep in range(10):
     observation, reward, terminated, info = env.step(action)
     episode_reward += reward
     
-episode_reward
 cols = ["t", "rep", "action", "reward", "sp1", "sp2", "sp3"]
 df = pd.DataFrame(df, columns = cols)
 df.to_csv("data/PPO" + str(iterations) + ".csv.gz", index = False)
@@ -58,33 +57,52 @@ df.to_csv("data/PPO" + str(iterations) + ".csv.gz", index = False)
 
 
 
-## Plots exploring agent action submanifolds
+## Plots ## 
 from plotnine import ggplot, geom_point, aes, geom_line, facet_wrap, geom_path
+
+## Timeseries
+df2 = (df
+       .melt(id_vars=["t", "action", "reward", "rep"])
+       .groupby(['t', "variable"], as_index=False)
+       .agg({'reward': 'mean',
+             'value': 'mean',
+             'action': 'mean'})) 
+(ggplot(df2, aes("t", "value", color="variable")) +
+ geom_line())
+
+# overall average state of each sp across reps and time
+aves = (df
+       .melt(id_vars=["t", "action", "reward", "rep"])
+       .groupby(["variable"], as_index=False)
+       .agg({'reward': 'mean',
+             'value': 'mean',
+             'action': 'mean'})) 
+init = aves.value.values
 
 ## agent action vs H
 dd = []
-for p in np.linspace(-1, 1, 101):
-  observation = np.append(observation[range(2)], p)
-  dd.append(np.append(p, agent.compute_single_action(observation) ))
+for p in np.linspace(-1, 0, 101):
+  observation = [init[0]+.1, init[1], p]
[…]

### 2023-01-14 code/plots.Rmd
fix selection of best escapement / msy action

--- 2023-01-12 code/plots.Rmd
@@
 
-```{r}
-ppo <- read_csv("../data/PPO.csv")
-```
+# Constant Effort
 
 ```{r}
-ppo |> 
-  mutate(pop = (sp1+1)/2, 
-         escapement = pop - action * pop) |> 
-  ggplot(aes(t, escapement, group=rep)) + geom_line() + ylim(0,1)
-```
-
-```{r}
-msy <- read_csv("../data/msy.csv")
+msy <- read_csv("../data/msy.csv.gz")
 best <- msy |> 
+  group_by(action, rep) |>
   filter(t == max(t)) |> 
-  group_by(action) |> summarise(reward = mean(reward)) |> 
+  group_by(action) |>
+  summarise(reward = mean(reward)) |> 
   filter(reward == max(reward))
@@
 ```
+
 ```{r}
@@
 
+# Constant Escapement
+
 ```{r}
 escapement <- read_csv("../data/escapement.csv.gz")
-best_e <- escapement |> 
+
+escapement |> 
+  group_by(action, rep) |>
   filter(t == max(t)) |> 
-  group_by(action) |> summarise(reward = mean(reward)) |> 
+  group_by(action) |>
+  summarise(reward = mean(reward)) |> 
   filter(reward == max(reward))
-best_e
+
+esc_sims <- escapement |> 
+  filter(action == 0.58) |>
+  rename(escapement = action) |>
+  mutate(effort = pmax(1 - escapement / (sp1+1), 0))
+
+esc_sims |> group_by(rep) |> filter(t == max(t)) |> pull(reward) |> mean()
 ```
@@
 ```{r}
-esc_sims <- escapement |>
-  filter(action == best_e$action) |> 
-  mutate(effort = pmax(1 - action / sp1, 0)) |>
-  pivot_longer(c("sp1", "sp2", "sp3"), names_to = "species", values_to = "abundance")
+esc_sims |>
+  pivot_longer(c("sp1", "sp2", "sp3"),
+               names_to = "species", values_to = "abundance") |>
+    filter(rep < 3) |>
+  ggplot(aes(t, abundance, group = interaction(species, rep), col=species)) +
+  geom_line(alpha=0.5) + facet_wrap(~rep, ncol=1)
+```
+
+```{r}
+ppo |> 
+  mutate(escapement = (sp1+1) - action * (sp1+1) -1,
+         harvest = action - 1) |>
+  pivot_longer(c("sp1", "sp2", "sp3", "escapement", "harvest"),
+               names_to = "species", values_to = "abundance") |>
+  ggplot(aes(t, abundance, col=species, group=interaction(rep, species))) +
+  geom_line(alpha=0.6) + facet_wrap(~rep, ncol=1)
 
 esc_sims |>
-  ggplot(aes(t, abundance, group = interaction(species, rep), col=species)) + geom_line(alpha=0.2)
 
-esc_sims |> ggplot(aes(t,action, group=rep)) + geom_line(alpha=.2)
+esc_sims |> ggplot(aes(t,action, group=rep)) + geom_line(alpha=.6)
 
-esc_sims |> ggplot(aes(t,effort, group=rep)) + geom_line(alpha=.2)
+esc_sims |> ggplot(aes(t,effort, group=rep)) + geom_line(alpha=.6)
 ```
+
+# PPO Agent
+
+```{r}
+ppo <- read_csv("../data/PPO250.csv.gz")
+
+```
+
+```{r}
+#ppo |> ggplot(aes(t, sp1, group=rep)) + geom_line()
+
+ppo |> ggplot(aes(t, reward, group=rep)) + geom_line()
+
+ppo |> 
+  mutate(pop = (sp1+1), 
+         escapement = pop - action * pop) |> 
+  ggplot(aes(t)) +
+  #geom_line(aes(t, pop, group=rep), col="black",alpha=0.1) +
+  geom_line(aes(t, escapement, group=rep),  col="black",alpha=0.2)
+
+ppo |> 
+  mutate(escapement = (sp1+1) - action * (sp1+1) -1,
+         harvest = action - 1) |>
+  pivot_longer(c("sp1", "sp2", "sp3", "escapement", "harvest"),
+               names_to = "species", values_to = "abundance") |>
+  group_by(t,species) |>
+  summarise(abundance = mean(abundance)+1) |>
+  ggplot(aes(t, abundance, col=species)) +
+  geom_line(alpha=0.6)
+
+```
+
+```{r}
+ppo |> 
+  mutate(escapement = (sp1+1) - action * (sp1+1) -1,
+         harvest = action - 1) |>
+  pivot_longer(c("sp1", "sp2", "sp3", "escapement", "harvest"),
+               names_to = "species", values_to = "abundance") |>
+  filter(rep < 3) |>
+  ggplot(aes(t, abundance, col=species, group=interaction(rep, species))) +
+  geom_line(alpha=0.6) + facet_wrap(~rep, ncol=1)
+```
+
+```{r}
+ppo |> filter(t == max(t)) |> pull(reward) |> mean()
+
+ppo |> filter(t == max(t)) |> pull(reward) |> sd()
+
+```

### 2023-01-14 code/python-plot.py

@@ -1,24 +1,0 @@
-
-## plot locally for fun
-import pandas as pd
-from plotnine import ggplot, geom_point, aes, geom_line, facet_wrap, geom_path
-
-df = pd.read_csv("data/PPO.csv.gz")
-
-df2 = (df
-       .melt(id_vars=["t", "action", "reward", "rep"])
-       .groupby(['t', "variable"], as_index=False)
-       .agg({'reward': 'mean',
-             'value': 'mean',
-             'action': 'mean'})) 
-
-
-(ggplot(df2, aes("t", "value", color="variable")) +
- geom_line())
-(ggplot(df2, aes("t", "action", color="variable")) + geom_line())
-(ggplot(df2, aes("t", "reward", color="variable")) + geom_line())
-
-(ggplot(df, aes("sp2", "sp3", color="action", size="sp1")) + geom_point())
-(ggplot(df, aes("sp1", "sp2", color="action", size="sp3")) + geom_point())
-(ggplot(df, aes("sp1", "sp3", color="action", size="sp2")) + geom_point())
-

### 2023-01-14 code/escapement.py
use env.Tmax

@@ -16,9 +16,9 @@ for action in actions:
   for rep in range(10):
     episode_reward = 0
     observation = env.reset()
-    for t in range(200):
+    for t in range(env.Tmax):
       df.append(np.append([t, rep, action, episode_reward], observation))
-      sp1_pop = (observation[0] + 1 ) / 2 # natural state-space
+      sp1_pop = (observation[0] + 1 ) # natural state-space
       effort = np.max([1 - action / sp1_pop, 0])
       observation, reward, terminated, info = env.step(effort)
       if terminated:
@@ -31,4 +31,12 @@ df.to_csv("data/escapement.csv.gz", index=False)
 
 
 
+## spot check max reward
+df2 = (df
+       .melt(id_vars=["t", "action", "reward", "rep"])
+       .groupby(['action', 't', "variable"], as_index=False)
+       .agg({'reward': 'mean',
+             'value': 'mean',
+             'action': 'mean'}))
+df2[df2.reward == np.max(df2.reward)]
 

### 2023-01-14 code/msy.py

@@ -10,7 +10,7 @@ for action in actions:
   for rep in range(10):
     episode_reward = 0
     observation = env.reset()
-    for t in range(200):
+    for t in range(env.Tmax):
       df.append(np.append([t, rep, action, episode_reward], observation))
       observation, reward, terminated, info = env.step(action)
       episode_reward += reward

### 2023-01-14 code/plots.Rmd
auto select best_e

@@
 
-escapement |> 
+best_e <- escapement |> 
   group_by(action, rep) |>
@@
 esc_sims <- escapement |> 
-  filter(action == 0.58) |>
+  filter(action == best_e$action) |>
   rename(escapement = action) |>

### 2023-01-14 code/escapement.py

@@ -30,13 +30,3 @@ df = pd.DataFrame(df, columns = cols)
 df.to_csv("data/escapement.csv.gz", index=False)
 
 
-
-## spot check max reward
-df2 = (df
-       .melt(id_vars=["t", "action", "reward", "rep"])
-       .groupby(['action', 't', "variable"], as_index=False)
-       .agg({'reward': 'mean',
-             'value': 'mean',
-             'action': 'mean'}))
-df2[df2.reward == np.max(df2.reward)]
-

### 2023-01-14 code/escapement.py
ray parallel

@@ -2,17 +2,12 @@ import gym_fishing
 import gym
 import pandas as pd
 import numpy as np
+import ray
+from timebudget import timebudget
 
-# Note that state space is mapped to [-1,1]
-#
-# effort corresponding to constant escapement is
-# effort = max(1 - escapement / sp1_pop, 0)
-
-actions = np.linspace(0,1,101)
-env = gym.make("threeFishing-v2")
-env.reset()
-df = []
-for action in actions:
+@ray.remote
+def simulate(env, action):
+  df = []
   for rep in range(10):
     episode_reward = 0
     observation = env.reset()
@@ -21,12 +16,23 @@ for action in actions:
       sp1_pop = (observation[0] + 1 ) # natural state-space
       effort = np.max([1 - action / sp1_pop, 0])
       observation, reward, terminated, info = env.step(effort)
+      episode_reward += reward
       if terminated:
         break
-      episode_reward += reward
+  return(df)
+
+
+# parallelize over actions
+@timebudget
+def parallel(operation, env, input):
+  df = ray.get([operation.remote(env, i) for i in input])
+  return(df)
+
+
+env = gym.make("threeFishing-v2")
+actions = np.linspace(0,1,101)
+df = parallel(simulate, env, actions)
 
 cols = ["t", "rep", "action", "reward", "sp1", "sp2", "sp3"]
-df = pd.DataFrame(df, columns = cols)
-df.to_csv("data/escapement.csv.gz", index=False)
-
-
+df2 = pd.DataFrame(np.vstack(df), columns = cols)
+df2.to_csv("data/escapement.csv.gz", index=False)

### 2023-01-14 code/msy.py

@@ -2,11 +2,12 @@ import gym_fishing
 import gym
 import pandas as pd
 import numpy as np
+import ray
+from timebudget import timebudget
 
-env = gym.make("threeFishing-v2")
-actions = np.linspace(0,.1,101)
-df = []
-for action in actions:
+@ray.remote
+def simulate(env, action):
+  df = []
   for rep in range(10):
     episode_reward = 0
     observation = env.reset()
@@ -14,26 +15,22 @@ for action in actions:
       df.append(np.append([t, rep, action, episode_reward], observation))
       observation, reward, terminated, info = env.step(action)
       episode_reward += reward
+  return(df)
+
+# parallelize over actions
+@timebudget
+def parallel(operation, env, input):
+  df = ray.get([operation.remote(env, i) for i in input])
+  return(df)
+
+
+env = gym.make("threeFishing-v2")
+actions = np.linspace(0,.1,101)
+df = parallel(simulate, env, actions)
 
 cols = ["t", "rep", "action", "reward", "sp1", "sp2", "sp3"]
-df = pd.DataFrame(df, columns = cols)
-df.to_csv("data/msy.csv.gz", index=False)
+df2 = pd.DataFrame(np.vstack(df), columns = cols)
+df2.to_csv("data/msy.csv.gz", index=False)
 
 
-
-
-
-# ignore me, we'll transform and plot in R
-# identify action at associated with highest cumulative mean reward by last timestep
-df2 = ( df
-        .groupby(['t','action'], as_index=False)
-        .agg({'reward': 'mean',
-              'sp1': 'mean',
-              'sp2': 'mean',
-              'sp3': 'mean'})
-        .melt(id_vars=["t", "action", "reward"]) 
-        )
-best_action = df2.query("t==199")[df2.reward == max(df2.reward)].action.values[0]
-df3 = df.query("action == " + str(best_action))
-                                
 

### 2023-01-17 manuscript/manuscript.Rmd
mini
changes to how models are presented

@@
   \label{eq:general model}
-  N_{t+1} - N_t = f(N_t) - h_t,
+  N_{t+1} - N_t = f(N_t) - h_t := N_t,
 \end{align}
-where $N(t) = (A_t,\ B_t,\ \dots) \in \mathbb{R}^{n}_+$ is a vector of populations, $f:\mathbb{R}^{n}\to\mathbb{R}^{n}$ is an arbitrary function, and $h_t$ is the harvest size at timestep $t$.
+where $N_t = (X_t,\ Y_t,\ \dots) \in \mathbb{R}^{n}_+$ is a vector of populations, $f:\mathbb{R}^{n}\to\mathbb{R}^{n}$ is an arbitrary function, and $h_t$ is the harvest size at timestep $t$.
 
-## Single species models
+## A single species model
 
@@
 $$
-  f(N_t) = r N_t\big(1 - N_t / K \big) =: L(N_t).
+  f(N_t) = r N_t\big(1 - N_t / K \big) =: L(N_t;\ r, K).
 $$
@@
   N_{t+1} - N_t
-  = L(N_t) - \frac{\beta H N_t^2}{c^2 + N_t^2}.
+  = L(N_t) - F(N_t, H; \beta,c),
 \end{align}
+where,
+\begin{align*}
+  F(N,H;\ \beta,c) := \frac{\beta H N^2}{c^2 + N^2}.
+\end{align*}
 The model has five parameters: a constant population $H$ of a species which preys on $N$, the maximal rate of predation $\beta$, and the predation half-maximum biomass $c$.
 
-Part of what makes eq. \eqref{eq:may} an interesting study case is the presence of a *tipping point* (see Fig. \ref{tbd}).
+Eq. \eqref{eq:may} is an interesting study case of a *tipping point* (see Fig. \ref{tbd}).
 Holding the value of $\beta$ fixed, for intermediate values of $H$ there exist two stable fixed points for the state $V_1$ of the system, these two attractors separated by an unstable fixed point.
@@
 ## A three species model
+
+Consider the following three species *generalization* of eq. \eqref{eq:may}:
+\begin{align}
+  \label{eq:may 3d}
+  \Delta X_t = L(X_t; r_X, K_X)
+\end{align}
 

### 2023-01-17 manuscript/manuscript.Rmd
3d model
bit more 3d model

@@
 $$
-  f(N_t) = r N_t\big(1 - N_t / K \big) =: L(N_t;\ r, K).
+  f(X_t) = r X_t\big(1 - X_t / K \big) =: L(X_t;\ r, K).
 $$
@@
   \label{eq:may}
-  N_{t+1} - N_t
-  = L(N_t) - F(N_t, H; \beta,c),
+  X_{t+1} - X_t
+  = L(X_t;\ r, K) - F(X_t, H; \beta,c),
 \end{align}
@@
 \begin{align*}
-  F(N,H;\ \beta,c) := \frac{\beta H N^2}{c^2 + N^2}.
+  F(X,H;\ \beta,c) := \frac{\beta H X^2}{c^2 + X^2}.
 \end{align*}
-The model has five parameters: a constant population $H$ of a species which preys on $N$, the maximal rate of predation $\beta$, and the predation half-maximum biomass $c$.
+The model has five parameters: the growth rate $r$ and carrying capacity $K$ for $X$, a constant population $H$ of a species which preys on $X$, the maximal rate of predation $\beta$, and the predation half-maximum biomass $c$.
 
@@
 \begin{align}
-  \label{eq:may 3d}
-  \Delta X_t = L(X_t; r_X, K_X)
+\label{eq:3d model}
+\begin{split}
+  \Delta X_t &= L(X_t;\ r_X, K_X) - F(X_t, Z_t;\ \beta, c) - c_{XY}X_tY_t - h_t,\\
+  \Delta Y_t &= L(Y_t;\ r_Y, K_Y) - D F(Y_t, Z_t;\ \beta, c) - c_{XY}X_tY_t,\\
+  \Delta Z_t &= \left(f(X_t + DY_t\right) - d_Z)Z_t.
+\end{split}
 \end{align}
+The model is conceptualized in Fig. \ref{fig:tbd}.
+It contains three populations, the state at time $t$ is $(X_t, Y_t, Z_t)$. 
+Species $Z$ preys on both $X$ and $Y$, while the latter two compete for resources.
+There are ten parameters in this model: 
+The growth rate and carrying capacity, $r_X$, $K_X$, $r_Y$ and $K_Y$, of $X$ and $Y$.
+A parameter $c_{XY}$ mediating a Lotka-Volterra competition between $X$ and $Y$.
+A maximum hunting intensity $\beta$ and half-maximum $c$ specifying how $Z$ forages on $X$ and $Y$.
+A parameter $D$ regulating a relative preference of $Z$ to prey on $Y$.
+Finally, a death rate $d_Z$ and a parameter $f$ related to the birth rate of $Z$.
+
+This model generalizes eq. \eqref{eq:may} in the following sense:
+If $f=d_Z=c_{XY}=0$, and we let $Z_{t=0}=H$, then eq. \eqref{eq:3d model} becomes
+\begin{align*}
+  \Delta X_t &= L(X_t;\ r_X, K_X) - F(X_t, H;\ \beta, c) - h_t,\\
+  \Delta Y_t &= L(Y_t;\ r_Y, K_Y) - D F(Y_t, H;\ \beta, c),\\
+  Z_t &= H.
+\end{align*}
+The equations for $X_t$ and $Y_t$ thus become decoupled, with $X_t$ evolving according to \eqref{eq:may}.
 

### 2023-01-17 code/escapement.py
updated plotting

@@ -2,18 +2,12 @@ import gym_fishing
 import gym
 import pandas as pd
 import numpy as np
+import ray
 
-# Note that state space is mapped to [-1,1]
-#
-# effort corresponding to constant escapement is
-# effort = max(1 - escapement / sp1_pop, 0)
-
-actions = np.linspace(0,1,101)
-env = gym.make("threeFishing-v2")
-env.reset()
-df = []
-for action in actions:
-  for rep in range(10):
+@ray.remote
+def simulate(env, action):
+  df = []
+  for rep in range(50):
     episode_reward = 0
     observation = env.reset()
     for t in range(env.Tmax):
@@ -21,12 +15,18 @@ for action in actions:
       sp1_pop = (observation[0] + 1 ) # natural state-space
       effort = np.max([1 - action / sp1_pop, 0])
       observation, reward, terminated, info = env.step(effort)
+      episode_reward += reward
       if terminated:
         break
-      episode_reward += reward
+  return(df)
+
+env = gym.make("threeFishing-v2")
+actions = np.linspace(0,1,101)
+
+# define parllel loop and execute
+parallel = [simulate.remote(env, i) for i in actions]
+df = ray.get(parallel)
 
 cols = ["t", "rep", "action", "reward", "sp1", "sp2", "sp3"]
-df = pd.DataFrame(df, columns = cols)
-df.to_csv("data/escapement.csv.gz", index=False)
-
-
+df2 = pd.DataFrame(np.vstack(df), columns = cols)
+df2.to_csv("data/escapement.csv.gz", index=False)

### 2023-01-17 code/msy.py

@@ -2,38 +2,35 @@ import gym_fishing
 import gym
 import pandas as pd
 import numpy as np
+import ray
 
-env = gym.make("threeFishing-v2")
-actions = np.linspace(0,.1,101)
-df = []
-for action in actions:
-  for rep in range(10):
+@ray.remote
+def simulate(env, action):
+  df = []
+  for rep in range(50):
     episode_reward = 0
     observation = env.reset()
     for t in range(env.Tmax):
       df.append(np.append([t, rep, action, episode_reward], observation))
       observation, reward, terminated, info = env.step(action)
       episode_reward += reward
+      if terminated:
+        break
+  return(df)
 
+
+env = gym.make("threeFishing-v2")
+actions = np.linspace(0,0.1,201)
+
+# define parllel loop and execute
+parallel = [simulate.remote(env, i) for i in actions]
+df = ray.get(parallel)
+
+# convert to data.frame & write to csv
 cols = ["t", "rep", "action", "reward", "sp1", "sp2", "sp3"]
-df = pd.DataFrame(df, columns = cols)
-df.to_csv("data/msy.csv.gz", index=False)
+df2 = pd.DataFrame(np.vstack(df), columns = cols)
+df2.to_csv("data/msy.csv.gz", index=False)
 
 
 
-
-
-# ignore me, we'll transform and plot in R
-# identify action at associated with highest cumulative mean reward by last timestep
-df2 = ( df
-        .groupby(['t','action'], as_index=False)
-        .agg({'reward': 'mean',
-              'sp1': 'mean',
-              'sp2': 'mean',
-              'sp3': 'mean'})
-        .melt(id_vars=["t", "action", "reward"]) 
-        )
-best_action = df2.query("t==199")[df2.reward == max(df2.reward)].action.values[0]
-df3 = df.query("action == " + str(best_action))
-                                
 

### 2023-01-17 code/train-agent.py

@@ -1,5 +1,5 @@
 from gym_fishing.envs import forageVVHcont
-from ray.rllib.agents.ppo import PPOTrainer
+from ray.rllib.algorithms import ppo
 from ray.tune import register_env
 import os
 import pandas as pd
@@ -8,31 +8,24 @@ import numpy as np
 ## We could call env directly without this if only  our envs took a env_config dict argument
 register_env("threeFishing-v2", lambda config: forageVVHcont())
 
-# Configure the algorithm.
-config = {
-    "env": "threeFishing-v2",
-    "num_workers": 1,
-    "num_envs_per_worker": 20,
-    "resources": {
-      "num_cpus_per_worker": 12,
-    },
-    "framework": "torch",
-    "num_gpus": 1,
-    "log_level": "ERROR",
-    "create_env_on_local_worker": True  # needed to restore env from checkpoint
-}
+config = ppo.PPOConfig()
+config.num_envs_per_worker=20
+config = config.resources(num_gpus=1)
+config.framework_str="torch"
+config.create_env_on_local_worker = True
+agent = config.build(env="threeFishing-v2")
 
-agent = PPOTrainer(config=config)
+
 iterations = 250
 checkpoint = ("cache/checkpoint_000{}".format(iterations))
 
 if not os.path.exists(checkpoint): # train only if no trained agent saved
   for _ in range(iterations):
       agent.train()
-  checkpoint = agent.save("cache")
+  checkpoint = agent.save("ppo")
 
-# Restore saved agent:
-agent = PPOTrainer(config=config)
+#agent_restored = config.build(env="threeFishing-v2")
+#agent_restored.evaluate()
 agent.restore(checkpoint)
 
 stats = agent.evaluate() # built-in method to evaluate agent on eval env
@@ -41,10 +34,10 @@ stats = agent.evaluate() # built-in method to evaluate agent on eval env
 env = agent.env_creator(agent.evaluation_config.env_config)
 
 df = []
-for rep in range(10):
+for rep in range(30):
   episode_reward = 0
   observation = env.reset()
-  for t in range(200):
+  for t in range(env.Tmax):
     action = agent.compute_single_action(observation)
     df.append(np.append([t, rep, action[0], episode_reward], observation))
     observation, reward, terminated, info = env.step(action)

### 2023-01-17 code/plots.Rmd

--- 2023-01-13 code/plots.Rmd
@@
 library(tidyverse)
+
+best_reward <- function(df) {
+  df |> 
+  group_by(action, rep) |>
+  filter(t == max(t)) |> 
+  group_by(action) |>
+  summarise(reward = mean(reward)) |> 
+  filter(reward == max(reward))
+}
 ```
@@
 msy <- read_csv("../data/msy.csv.gz")
-best <- msy |> 
-  filter(t == max(t)) |> 
-  group_by(action) |> summarise(reward = mean(reward)) |> 
-  filter(reward == max(reward))
+best <- best_reward(msy)
 best
 ```
+
+```{r}
+rewards <- msy |>
+  group_by(action, rep) |>
+  filter(t == max(t)) |> 
+  group_by(action) |>
+  summarise(mean_reward = mean(reward),
+            sd = sd(reward))
+
+rewards |>
+  filter(action < 0.1) |>
+  ggplot(aes(action, mean_reward, 
+             ymin = mean_reward - 2*sd,
+             ymax=mean_reward+2*sd)) +
+  geom_point() + geom_ribbon(alpha=0.4)
+```
+
 ```{r}
@@
   filter(action == best$action) |> 
-  pivot_longer(c("sp1", "sp2", "sp3"), names_to = "species", values_to = "abundance")
+  pivot_longer(c("sp1", "sp2", "sp3"),
+               names_to = "species", values_to = "abundance") |>
+  mutate(abundance = abundance + 1) # natural units
 
 msy_sim |>
-  ggplot(aes(t, abundance, group = interaction(species, rep), col=species)) + 
+  filter(rep < 4) |>
+  ggplot(aes(t, abundance,  col=species)) + 
+    geom_line() + facet_wrap(~rep)
+
+msy_sim |> 
+  filter(rep < 30, species == "sp1") |>
+  ggplot(aes(t, abundance, group = interaction(rep,species),  col=species)) + 
     geom_line(alpha=0.5)
 
+```
+```{r}
+msy_sim |> 
+  group_by(rep) |>
+  filter(t==max(t), species=="sp1") |>
+  ggplot(aes(rep, reward)) + geom_col()
+```
+# TAC
+
+80% of  MSY
+
+```{r}
+actions <- unique(msy$action)
+i <- which.min(abs(actions - best$action* 0.8))
+tac_sim <- msy |> 
+  filter(action == actions[[i]]) |>
+  pivot_longer(c("sp1", "sp2", "sp3"),
+               names_to = "species", values_to = "abundance")
+tac_sim |>
+  filter(rep < 4) |>
+  ggplot(aes(t, abundance,  col=species)) + 
+    geom_line() + facet_wrap(~rep)
+
+```
+```{r}
+tac_sim |> 
+  group_by(rep) |>
+  filter(t==max(t), species=="sp1") |>
+  ggplot(aes(rep, reward)) + geom_col()
+```
+
+# PPO Agent
+
+```{r}
+ppo <- read_csv("../data/PPO250.csv.gz")
+
+```
+
+```{r}
+ppo_sims <- ppo |> 
+  mutate(escapement = (sp1+1) - action * (sp1+1),
+         effort = action) |>
+  pivot_longer(c("sp1", "sp2", "sp3"),
+               names_to = "species", values_to = "abundance") |>
+  mutate(abundance = abundance + 1)
+
+ppo_sims |>
+  filter(rep < 4) |>
+  ggplot(aes(t, abundance, col=species)) +
+  geom_line(alpha=0.6) + facet_wrap(~rep)
+```
+
+```{r}
+ppo_sims |>
+  filter(species == "sp1") |>
+  group_by(t) |>
+  summarise(abundance = mean(abundance), 
+            effort = mean(effort),
+            escapement = mean(escapement)) |> 
+  pivot_longer(-t, names_to = "metric", values_to = "population") |>
+  ggplot(aes(t, population, col=metric)) + geom_line()
 ```
@@
 ```{r}
-escapement <- read_csv("../data/escapement.csv.gz")
-best_e <- escapement |> 
-  filter(t == max(t)) |> 
-  group_by(action) |> summarise(reward = mean(reward)) |> 
-  filter(reward == max(reward))
+escapement <- read_csv("../data/escapement.csv.gz", show_col_types = FALSE)
+
+## Determine which level of constant escapement gives highest net reward
+best_e <- best_reward(escapement)
 best_e
@@
   filter(action == best_e$action) |>
-  mutate(effort = pmax(1 - action / sp1, 0)) |>
-  pivot_longer(c("sp1", "sp2", "sp3"), names_to = "species", values_to = "abundance")
+  pivot_longer(c("sp1", "sp2", "sp3"),
+               names_to = "species", values_to = "abundance") |>
+  mutate(abundance = abundance + 1) |>
+  rename(escapement = action)
 
 esc_sims |>
-  ggplot(aes(t, abundance, group = interaction(species, rep), col=species)) + geom_line(alpha=0.2)
-
-esc_sims |> ggplot(aes(t,action, group=rep)) + geom_line(alpha=.2)
-
-esc_sims |> ggplot(aes(t,effort, group=rep)) + geom_line(alpha=.2)
-```
-
-# PPO Agent
-
-```{r}
-ppo <- read_csv("../data/PPO-long.csv.gz")
+  filter(rep < 4) |>
+  ggplot(aes(t, abundance, col=species)) + geom_line() +
+  geom_line(aes(t, escapement), col="grey50") +
+  facet_wrap(~rep)
 ```
@@
 ```{r}
-ppo |> ggplot(aes(t, sp1, group=rep)) + geom_line()
+rewards <- escapement |>
+  group_by(action, rep) |>
+  filter(t == max(t)) |> 
+  group_by(action) |>
+  summarise(mean_reward = mean(reward),
+            sd = sd(reward))
 
-ppo |> ggplot(aes(t, reward, group=rep)) + geom_line()
-
-ppo |> 
-  mutate(pop = (sp1+1), 
-         escapement = pop - action * pop) |> 
-  ggplot(aes(t, escapement, group=rep)) + geom_line() + ylim(0,1)
-
-ppo |> 
-  mutate(effort = pmax(1 - action / sp1, 0)) |>
-  pivot_longer(c("sp1", "sp2", "sp3"), names_to = "species", values_to = "abundance") |>
-  ggplot(aes(t, abundance, group = interaction(species, rep), col=species)) +
-  geom_line(alpha=0.8)
-
+rewards |>
+  ggplot(aes(action, mean_reward, 
+             ymin = mean_reward - 2*sd,
+             ymax=mean_reward+2*sd)) +
+  geom_point() + geom_ribbon(alpha=0.4)
 ```
-
-```{r}
-ppo |> filter(t == max(t)) |> pull(reward) |> mean()
-ppo |> filter(t == max(t)) |> pull(reward) |> sd()
-
-```

### 2023-01-17 manuscript/manuscript.Rmd
mini

@@
   X_{t+1} - X_t
-  = L(X_t;\ r, K) - F(X_t, H; \beta,c),
+  = L(X_t;\ r, K) - F(X_t, H; \beta,c) - h_t,
 \end{align}
@@
 The equations for $X_t$ and $Y_t$ thus become decoupled, with $X_t$ evolving according to \eqref{eq:may}.
+
+As can be expected, the dynamics of eq. \eqref{eq:3d model} are significantly more complex than those of eq. \eqref{eq:may}.
 

### 2023-01-17 manuscript/manuscript.Rmd
3d model section done (up to figures)

--- 2023-01-17 manuscript/manuscript.Rmd
@@
 The equations for $X_t$ and $Y_t$ thus become decoupled, with $X_t$ evolving according to \eqref{eq:may}.
+
+As can be expected, the dynamics of eq. \eqref{eq:3d model} are significantly more complex than those of eq. \eqref{eq:may}.
+There are a variety of fixed points for the equations---around 7 or 8 for wide ranges of parameter values.
+Fig. \ref{fig:tbd} provides a rough idea of the fixed point structure by plotting the bifurcation diagram obtained from varying $D$ and fixing the other parameters.
+
+Throughout most of this paper we will fix the parameters of the model according to Table \ref{tab:tbd}.
+At these values, the fixed points of the model are:
+\begin{align*}
+  TBD
+\end{align*}
+The classification of fixed points for multi-dimensional systems is generally more complicated than in one dimension.
+A fixed point might be stable against perturbations in one direction, but unstable against perturbations along another.
+While a detailed classification of these fixed points is beyond the scope of this work, we point out that the point $(TBD)$ is stable against perturbations in all directions.
+We will thus use it as a starting point for our simulations.
 

### 2023-01-17 manuscript/manuscript.Rmd

--- 2023-01-17 manuscript/manuscript.Rmd
@@
 As can be expected, the dynamics of eq. \eqref{eq:3d model} are significantly more complex than those of eq. \eqref{eq:may}.
+There are a variety of fixed points for the equations---around 7 or 8 for wide ranges of parameter values.
+Fig. \ref{fig:tbd} provides a rough idea of the fixed point structure by plotting the bifurcation diagram obtained from varying $D$ and fixing the other parameters.
+
+Throughout most of this paper we will fix the parameters of the model according to Table \ref{tab:tbd}.
+At these values, the fixed points of the model are:
+\begin{align*}
+  TBD
+\end{align*}
+The classification of fixed points for multi-dimensional systems is generally more complicated than in one dimension.
+A fixed point might be stable against perturbations in one direction, but unstable against perturbations along another.
+While a detailed classification of these fixed points is beyond the scope of this work, we point out that the point $(TBD)$ is stable against perturbations in all directions.
+We will thus use it as a starting point for our simulations.
 

### 2023-01-17 code/train-agent.py
small bugfix: terminate ppo sim if necessary

@@ -17,7 +17,7 @@ agent = config.build(env="threeFishing-v2")
 
 
 iterations = 250
-checkpoint = ("cache/checkpoint_000{}".format(iterations))
+checkpoint = ("ppo/checkpoint_000{}".format(iterations))
 
 if not os.path.exists(checkpoint): # train only if no trained agent saved
   for _ in range(iterations):
@@ -42,6 +42,8 @@ for rep in range(30):
     df.append(np.append([t, rep, action[0], episode_reward], observation))
     observation, reward, terminated, info = env.step(action)
     episode_reward += reward
+    if terminated:
+      break
     
 cols = ["t", "rep", "action", "reward", "sp1", "sp2", "sp3"]
 df = pd.DataFrame(df, columns = cols)

### 2023-01-17 manuscript/manuscript.Rmd
results

@@
   \label{eq:general model}
-  N_{t+1} - N_t = f(N_t) - h_t := N_t,
+  N_{t+1} - N_t = f(N_t) - h_t := \Delta N_t,
 \end{align}
@@
   \label{eq:may}
-  X_{t+1} - X_t
+  \Delta X_t
   = L(X_t;\ r, K) - F(X_t, H; \beta,c) - h_t,
@@
 We will thus use it as a starting point for our simulations.
+
+## Stochastic dynamics
+
+\textbf{Probably just merge on previous sections.}
+We consider a stochastic contribution to the system's dynamics.
+This way, eq. \eqref{eq:may} becomes
+\begin{align}
+  \label{eq:may stoch}
+  \Delta X_t
+  = L(X_t;\ r, K) - F(X_t, H; \beta,c) - h_t + \eta_t,
+\end{align}
+where $\{\eta_t\}_t$ are independent Gaussian variables with zero mean.
+The variance $\sigma^2$ of $\eta_t$ is an additional parameter of the model.
+
+Similarly, eq. \eqref{eq:3d model} becomes
+\begin{align*}
+  \label{eq:3d model stoch}
+  \begin{split}
+    \Delta X_t &= L(X_t;\ r_X, K_X) - F(X_t, Z_t;\ \beta, c) - c_{XY}X_tY_t - h_t + \eta_{X,t},\\
+    \Delta Y_t &= L(Y_t;\ r_Y, K_Y) - D F(Y_t, Z_t;\ \beta, c) - c_{XY}X_tY_t + \eta_{Y,t},\\
+    \Delta Z_t &= \left(f(X_t + DY_t\right) - d_Z)Z_t + \eta_{Z,t}.
+\end{split}
+\end{align*}
 
@@
 
+We trained a DRL agent to obtain a candidate policy
+\begin{align*}
+  \pi: (X_t,Y_t,Z_t) \mapsto h_t
+\end{align*}
+for eq. \eqref{eq:3d model}.
+The agent could fully and precisely observe the state of the system at each time step and decide on its harvest based on this information. 
+The agent was trained using the PPO algorithm, using a neural network with two hidden layers of 64 neurons to express $\pi$.
+Details of the training process may be found in App. \ref{app:tbd}.
+Our code is available at TBD.
+
 ## Acknowledgements

### 2023-01-17 manuscript/manuscript.Rmd
arxiv style, and add headers

@@
 abstract: |
-  Existing methods for optimal control methods struggle to deal with the complexity commonly encountered in real-world systems, including dimensionality, process error, model bias and heterogeneity. 
-  Instead of tackling these complexities directly, researchers have typically sought to find exact optimal solutions to simplified models of the processes in question. 
-  When is the optimal solution to a very approximate, stylized model better than an approximate solution to a more accurate model? 
-  This question has largely gone unanswered owing to the difficulty of finding even approximate solutions in the case of complex models.  
-  Our approach draws on recent algorithmic and computational advances in deep reinforcement learning. 
-  These methods have hitherto focused on problems in games or robotic mechanics, which operate under precisely known rules. 
-  We demonstrate the ability for novel algorithms using deep neural networks to successfully approximate such solutions (the "policy function" or control rule) without knowing or ever attempting to infer a model for the process itself.
-  This powerful new technique lets us finally begin to answer the question.
-  We show that in many but not all cases, the optimal policy for a carefully chosen over-simplified model can still out-perform these novel algorithms trained to find approximate solutions to simulations of a realistically complex system.
-  This illustrates the promise of combining complex, realistic simulations with emerging machine-learning techniques to improve our understanding and management of ecological systems.
+  Write
 
 bibliography: bibliography.bib
-biblio-style: spphys
+#biblio-style: spphys
+
+header-includes:
+  - \usepackage{amsmath}
+  - \usepackage{amssymb}
+  - \usepackage{lineno}
+  - \usepackage{setspace}
+  - \usepackage{bookmark}
+  - \linenumbers
+  - \doublespacing
 
 # bibstyle options spbasic(default), spphys, spmpsci
-output: rticles::springer_article
+# output: rticles::springer_article
+
+output: 
+  rticles::arxiv_article:
+    keep_tex: true
+
+editor_options: 
+  markdown: 
+    wrap: sentence
 
@@
 
+<!--
+Existing methods for optimal control methods struggle to deal with the complexity commonly encountered in real-world systems, including dimensionality, process error, model bias and heterogeneity. 
+Instead of tackling these complexities directly, researchers have typically sought to find exact optimal solutions to simplified models of the processes in question. 
+When is the optimal solution to a very approximate, stylized model better than an approximate solution to a more accurate model? 
+This question has largely gone unanswered owing to the difficulty of finding even approximate solutions in the case of complex models.  
+Our approach draws on recent algorithmic and computational advances in deep reinforcement learning. 
+These methods have hitherto focused on problems in games or robotic mechanics, which operate under precisely known rules. 
+We demonstrate the ability for novel algorithms using deep neural networks to successfully approximate such solutions (the "policy function" or control rule) without knowing or ever attempting to infer a model for the process itself.
+This powerful new technique lets us finally begin to answer the question.
+We show that in many but not all cases, the optimal policy for a carefully chosen over-simplified model can still out-perform these novel algorithms trained to find approximate solutions to simulations of a realistically complex system.
+This illustrates the promise of combining complex, realistic simulations with emerging machine-learning techniques to improve our understanding and management of ecological systems.
+-->
+
 Despite the complexity inherent in natural ecosystems, analytically tractable models have long played an essential role in their understanding and management.
-Ecological management may be understood as an optimal control problem - a manager must determine a sequence of actions (a "policy") to optimize some desired objective or combination of objectives (conservation goals, natural resource extraction, economic and social constraints.)
+Ecological management may be understood as an optimal control problem - a manager must determine a sequence of actions (a "policy") to optimize some desired objective(s) and constraints (e.g.
+conservation goals, natural resource extraction, economic and social constraints.)
 
@@
 The agent could fully and precisely observe the state of the system at each time step and decide on its harvest based on this information.
-The agent was trained using the PPO algorithm, using a neural network with two hidden layers of 64 neurons to express $\pi$.
-Details of the training process may be found in App. \ref{app:tbd}.
-Our code is available at TBD.
 

### 2023-01-17 code/plots.Rmd

@@
   group_by(action) |>
-  summarise(reward = mean(reward)) |> 
-  filter(reward == max(reward))
+  summarise(mean_reward = mean(reward), sd = sd(reward)) |> 
+  filter(mean_reward == max(mean_reward))
 }
@@
 ```{r}
+overall_ppo <- 
+ppo |> 
+  group_by(rep) |>
+  filter(t==max(t)) |> ungroup() |>
+  filter(reward > 0) |>
+  summarise(reward = mean(reward),
+            t = mean(t),
+            effort = mean(action)) |>
+  mutate(policy="ppo")
+overall_ppo
+
+ppo |> 
+  group_by(rep) |>
+  filter(t==max(t)) |> 
+  ggplot(aes(rep, reward)) + geom_col()
+```
+
+```{r}
 ppo_sims <- ppo |> 
@@
 ppo_sims |>
-  filter(rep < 4) |>
+  filter(rep > 20, rep < 25) |>
   ggplot(aes(t, abundance, col=species)) +

### 2023-01-18 manuscript/manuscript.Rmd
a bit of actual writing :pen:

@@
 
-Despite the complexity inherent in natural ecosystems, analytically tractable models have long played an essential role in their understanding and management.
 Ecological management may be understood as an optimal control problem - a manager must determine a sequence of actions (a "policy") to optimize some desired objective(s) and constraints (e.g.
 conservation goals, natural resource extraction, economic and social constraints.)
+Despite the complexity inherent in natural ecosystems, analytically tractable models have long played an essential role in their understanding and management.
+
+<!-- Why fisheries-focused examples -->
+Global fisheries management has long been a crucible for both the development of theory and best practices in actual management for many reasons.
+Fisheries are an essential natural resource that provide the primary source of protein for one in every four humans, and have faced widely documented declines due to overfishing [cite, cite].
+Because of this importance, fisheries also represent some of the longest and best studied timeseries of population dynamics available to ecologists, as well as a correspondingly rich area for the development of theory and practice.
+Methods developed in the study and management of fisheries are frequently applied to other areas of ecological management, from concepts such as maximum sustainable yield [], optimal control under observation uncertainty [], to adaptive management [], management strategy evaluation [], to the studies of conservation policy and international negotiations [].
+
+<!-- Real-world application-->
+To understand the importance of a potential application of novel data-driven methods to the real world, it is most instructive to have a clearer picture of how models are currently used to inform ecological management in fisheries.
+In any issue as large and contentious as fisheries management, it may be far more difficult to provide a generalization  about how things actually work than to make sweeping recommendations about how it ought to.
+All the same, it is our firm belief that any researcher seeking to do the latter should first attempt to understand the former, so we beg the forbearance of any specialists in the practice as we seek to provide a concise overview of fisheries management today.
+Here, we focus specifically on practices in data rich, 'well-managed' fisheries, which, despite some issues, may be among the best examples of ecological management at continental scale that is largely driven by scientific processes, models and data.
+
+Despite such rich development, fundamental concepts and approaches widely considered best practices remain rooted in the analysis of analytically tractable one-dimensional models.
+This fact is not always obvious at cursory inspection, and can be met with cynicism even among professional fisheries managers -- after all, typical stock assessment models have over 100 parameters (e.g.
+see the stock assessments referenced from in @RAMLegacy) and best practices such as management strategy evaluation (MSE) evaluate and select potential policies by comparing across large and diverse simulations [e.g.
+@Punt2016].
+However, the space of policies considered under such approaches is necessarily far smaller than the set of all possible action sequences.
+Instead, fisheries management approaches such as MSE rely on theory based on more tractable models to restrict the search space of possible policies.  
+Specifically, most fisheries policies are framed in terms of a *constant mortality* policy, (sometimes a simple piece-wise linear function, i.e.
+a single linear segment, is introduced as a precaution a very low stock sizes), where fishing 'mortality' is defined as the ratio of harvest (catch) to stock size.
+(This definition sidesteps much complexity -- fisheries regulation is typically implemented in terms of catch quotas -- i.e.
+limit to metric tons of a given species that can be harvested in a given season.
+Thus to determine the quota, fisheries scientists must fist estimate the current biomass, and then multiply by this dimensionless ratio to determine the corresponding harvest quota, $H = F x_t$.
+Thus the process of measuring the stock size, a 'stock assessment', gives rise to the harvest quota).
+Constant mortality policies have a strong basis in theory as well as practical justifications.
+
+The best known such policy is that of "Maximum Sustainable Yield" (MSY), i.e. the largest possible constant mortality that can be sustained indefinitely without depleting the stock.
+MSY remains the basis of much fisheries management, including international treaties such as the 1982 UN Law of the Sea [@UNCLOS1982].
+MSY in 1D models [@Schaefer1954]
+critique and actual practice, ABC and TAC, (but go lighter on acronyms)
+
+The management of any ecological system can be framed in terms of a Sequential Decision Process - 
+the manager observes the ecosystem state ($x_t$) and decides on action ($a_t$) to take (including potentially no action), potentially receiving any costs or benefits associated with that action and and state (reward $r_t$  $r_t = U(x_t,a_t)$),
+while the system evolves into a new state in response $x_{t+1} = f(x_t, a_t)$.
+When the probability of the future state can be determined by the current state and action, we say the system is Markovian and the process a Markov Decision Process (MDP).
+As in a game of chess, the space of possible sequences of actions a manager might choose over any realization of the Markov process is enormous.
+When the action and state space are discrete and sufficiently small, computational approaches such as dynamic programming may be used to exhaustively search
+
+Effort $Q = E X_t$
+
+Fisheries policies are typically implemented in terms of *catch quotas*, i.e. a limit to the number of metric tonnes of fish that can be caught during a given season.
+
+and/or *effort controls*, restrictions on the fishing effort -- in numbers of boats, time of year, or days at sea
 

### 2023-01-18 manuscript/manuscript.Rmd
more results

--- 2023-01-17 manuscript/manuscript.Rmd
@@
 - name: Melissa Chapman
+  address: Department of Environmental Science, Policy, and Management, University of California, Berkeley
+- name: Abigail G. Keller
   address: Department of Environmental Science, Policy, and Management, University of California, Berkeley
@@
 The agent could fully and precisely observe the state of the system at each time step and decide on its harvest based on this information. 
+The agent was trained using the PPO algorithm, using a neural network with two hidden layers of 64 neurons to express $\pi$.
+Details of the training process may be found in App. \ref{app:tbd}.
+Our code is available at TBD.
+
+We compare the performance of this agent to the performance of MSY and of an 80% TAC.
+
+Our results suggest that DRL can find solutions that are more sustainable *and* more profitable than both MSY and TAC.
+Indeed, DRL finds solutions in which, for certain periods of time, no harvest is performed on the system.
+These periods effectively allow the system to, in a sense, ‘‘reset’’ itself.
+They lead to a distinctive oscillating behavior of the system state over time, with periods of relative depletion followed by periods of growth. 
+Figure \ref{fig:tbd} exemplifies this point by comparing the harvest effort and state times series obtained from these policies.
+
+It is well known that following MSY policies can lead to over-fishing.
+Our results reproduce this behavior and, furthermore, warn that using a TAC policy instead might not be enough to gain sustainability.
+Figure TBD-A shows the mean episode length for
 

### 2023-01-18 manuscript/manuscript.Rmd
more results

@@
 
+Rewards were defined to be a compound of two contributions.
+The first is economical: if at time $t$ a harvest of $h_t$ is secured, then the agents total reward is increased by $h_t$.
+The second regards sustainability: if an extinction happens at time step $t$, then a penalty of $-50/t$ is incurred by the agent.
+The choice of the constant $50$ in this rule was rather arbitrary: we set it to be a fourth of the maximal episode length ($200$ time steps), so as to encourage the agent to find policies that do not lead to extinctions below the time scale of around 100 time steps.
+\textbf{Too long of a ‘‘sorry’’ message?}
+
 We compare the performance of this agent to the performance of MSY and of an 80% TAC.
@@
 Our results reproduce this behavior and, furthermore, warn that using a TAC policy instead might not be enough to gain sustainability.
-Figure TBD-A shows the mean episode length for
+Figure TBD-A summarizes episode length distribution for MSY, TAC and DRL.
+While the DRL-derived policy does not lead to extinctions, both MSY and TAC do so, and thus their episode mean length is lower than the maximal episode length (in our case set at 200 years).
+This difference is reflected on the profits earned by the agent, as shown in Fig. TBD-B.
 

### 2023-01-19 manuscript/manuscript.Rmd
:pen: not much progress here

--- 2023-01-18 manuscript/manuscript.Rmd
@@
   address: Department of Environmental Science, Policy, and Management, University of California, Berkeley
+- name: Abigail Keller
+  address: Department of Environmental Science, Policy, and Management, University of California, Berkeley
 - name: Carl Boettiger
   address: Department of Environmental Science, Policy, and Management, University of California, Berkeley
-  email: cboettig@berkele.edu
+  email: cboettig@berkeley.edu
 
@@
 abstract: |
-  Write
+  Write something here
 
@@
 Despite the complexity inherent in natural ecosystems, analytically tractable models have long played an essential role in their understanding and management.
+...
 
@@
 Global fisheries management has long been a crucible for both the development of theory and best practices in actual management for many reasons.
-Fisheries are an essential natural resource that provide the primary source of protein for one in every four humans, and have faced widely documented declines due to overfishing [cite, cite].
+Fisheries are an essential natural resource that provide the primary source of protein for one in every four humans, and have faced widely documented declines due to overfishing [@Worm2006, @Costello2016].
 Because of this importance, fisheries also represent some of the longest and best studied timeseries of population dynamics available to ecologists, as well as a correspondingly rich area for the development of theory and practice.
-Methods developed in the study and management of fisheries are frequently applied to other areas of ecological management, from concepts such as maximum sustainable yield [], optimal control under observation uncertainty [], to adaptive management [], management strategy evaluation [], to the studies of conservation policy and international negotiations [].
+Methods developed in the study and management of fisheries are frequently applied to other areas of ecological management, from concepts such as maximum sustainable yield [], optimal control under observation uncertainty [@Memarzadeh2019], to adaptive management [@Walters], management strategy evaluation [@Punt], to the studies of conservation policy and international negotiations [@].
 
-<!-- Real-world application-->
+<!-- A sweeping summary of fisheries decision theory, don't bite me-->
 To understand the importance of a potential application of novel data-driven methods to the real world, it is most instructive to have a clearer picture of how models are currently used to inform ecological management in fisheries.
@@
 
-Despite such rich development, fundamental concepts and approaches widely considered best practices remain rooted in the analysis of analytically tractable one-dimensional models.
+<!-- Fisheries population models are rich but decisions aren't, mostly constant mortality -->
+While fisheries stock assessment models are quite rich, the choice of possible strategies remains motivated by intution from much simpler, analytically tractable one-dimensional models.
 This fact is not always obvious at cursory inspection, and can be met with cynicism even among professional fisheries managers -- after all, typical stock assessment models have over 100 parameters (e.g.
@@
 Specifically, most fisheries policies are framed in terms of a *constant mortality* policy, (sometimes a simple piece-wise linear function, i.e.
-a single linear segment, is introduced as a precaution a very low stock sizes), where fishing 'mortality' is defined as the ratio of harvest (catch) to stock size.
-(This definition sidesteps much complexity -- fisheries regulation is typically implemented in terms of catch quotas -- i.e.
-limit to metric tons of a given species that can be harvested in a given season.
-Thus to determine the quota, fisheries scientists must fist estimate the current biomass, and then multiply by this dimensionless ratio to determine the corresponding harvest quota, $H = F x_t$.
-Thus the process of measuring the stock size, a 'stock assessment', gives rise to the harvest quota).
+a single linear segment, is introduced as a precaution a very low stock sizes), where fishing 'mortality' is defined as the ratio of harvest (catch) to stock size, $F = H_t/x_t$.
+
+The intuition for a constant mortality strategy is simple and well established
+
 Constant mortality policies have a strong basis in theory as well as practical justifications.
 
-The best known such policy is that of "Maximum Sustainable Yield" (MSY), i.e. the largest possible constant mortality that can be sustained indefinitely without depleting the stock.
-MSY remains the basis of much fisheries management, including international treaties such as the 1982 UN Law of the Sea [@UNCLOS1982].
-MSY in 1D models [@Schaefer1954]
-critique and actual practice, ABC and TAC, (but go lighter on acronyms)
-
+<!-- In contrast to constant mortality, the possible decision space is huge-->
 The management of any ecological system can be framed in terms of a Sequential Decision Process - 
@@
 As in a game of chess, the space of possible sequences of actions a manager might choose over any realization of the Markov process is enormous.
-When the action and state space are discrete and sufficiently small, computational approaches such as dynamic programming may be used to exhaustively search
+When the action and state space are discrete and sufficiently small, computational approaches such as dynamic programming may be used to guarantee an exhaustive search for a single optimal policy.
+Dynamic programming scales very poorly to larger state and action spaces.
+
+Deep reinforcement learning has opened up increasingly compelling alternative
 

### 2023-01-19 code/escapement_parallel.py

@@ -1,41 +1,0 @@
-import gym_fishing
-import gym
-import pandas as pd
-import numpy as np
-import ray
-from timebudget import timebudget
-
-@ray.remote
-def simulate(env, action, rep):    
-  df = []
-  episode_reward = 0
-  observation = env.reset()
-  for t in range(200):
-    df.append(np.append([t, rep, action, episode_reward], observation))
-    effort = np.min(1 - action / observation[0], 0)
-    observation, reward, terminated, info = env.step(effort)
-    episode_reward += reward
-  return(df)
-
-@ray.remote
-def replicate_sim(env, action):
-  simulate(env, action)
-
-actions = np.linspace(-1,0,11)
-env = gym.make("threeFishing-v2")
-
-
-@timebudget
-def parallel(operation, env, input):
-  df = ray.get([operation.remote(env, i) for i in input])
-  return(df)
-
-#ray.init()
-action = 0.1
-
-input = range(10)
-df = parallel(simulate, env, input)
-
-cols = ["t", "rep", "action", "reward", "sp1", "sp2", "sp3"]
-df2 = pd.DataFrame(np.vstack(df), columns = cols)
-

### 2023-01-19 manuscript/manuscript.Rmd

@@
   address: Department of Environmental Science, Policy, and Management, University of California, Berkeley
-- name: Abigail Keller
+- name: Abigail G. Keller
   address: Department of Environmental Science, Policy, and Management, University of California, Berkeley
@@
 The agent could fully and precisely observe the state of the system at each time step and decide on its harvest based on this information. 
+The agent was trained using the PPO algorithm, using a neural network with two hidden layers of 64 neurons to express $\pi$.
+Details of the training process may be found in App. \ref{app:tbd}.
+Our code is available at TBD.
+
+Rewards were defined to be a compound of two contributions.
+The first is economical: if at time $t$ a harvest of $h_t$ is secured, then the agents total reward is increased by $h_t$.
+The second regards sustainability: if an extinction happens at time step $t$, then a penalty of $-50/t$ is incurred by the agent.
+The choice of the constant $50$ in this rule was rather arbitrary: we set it to be a fourth of the maximal episode length ($200$ time steps), so as to encourage the agent to find policies that do not lead to extinctions below the time scale of around 100 time steps.
+\textbf{Too long of a ‘‘sorry’’ message?}
+
+We compare the performance of this agent to the performance of MSY and of an 80% TAC.
+
+Our results suggest that DRL can find solutions that are more sustainable *and* more profitable than both MSY and TAC.
+Indeed, DRL finds solutions in which, for certain periods of time, no harvest is performed on the system.
+These periods effectively allow the system to, in a sense, ‘‘reset’’ itself.
+They lead to a distinctive oscillating behavior of the system state over time, with periods of relative depletion followed by periods of growth. 
+Figure \ref{fig:tbd} exemplifies this point by comparing the harvest effort and state times series obtained from these policies.
+
+It is well known that following MSY policies can lead to over-fishing.
+Our results reproduce this behavior and, furthermore, warn that using a TAC policy instead might not be enough to gain sustainability.
+Figure TBD-A summarizes episode length distribution for MSY, TAC and DRL.
+While the DRL-derived policy does not lead to extinctions, both MSY and TAC do so, and thus their episode mean length is lower than the maximal episode length (in our case set at 200 years).
+This difference is reflected on the profits earned by the agent, as shown in Fig. TBD-B.
 

### 2023-01-19 manuscript/manuscript.Rmd
...

@@
 The choice of the constant $50$ in this rule was rather arbitrary: we set it to be a fourth of the maximal episode length ($200$ time steps), so as to encourage the agent to find policies that do not lead to extinctions below the time scale of around 100 time steps.
-\textbf{Too long of a ‘‘sorry’’ message?}
+\textbf{Too long of a "sorry" message?}
 
@@
 Indeed, DRL finds solutions in which, for certain periods of time, no harvest is performed on the system.
-These periods effectively allow the system to, in a sense, ‘‘reset’’ itself.
+These periods effectively allow the system to, in a sense, "reset" itself.
 They lead to a distinctive oscillating behavior of the system state over time, with periods of relative depletion followed by periods of growth. 
@@
 organized by Paul Armsworth, Alan Hastings, Megan Donahue, and Carl Towes in
-2011 that first posed the question addressed here.
+2011 seeking to pivot control from 'optimal' to 'pretty darn good'.
 

### 2023-01-19 manuscript/manuscript.Rmd
conceptual figure and caption draft

--- 2023-01-19 manuscript/manuscript.Rmd
@@
   address: Department of Environmental Science, Policy, and Management, University of California, Berkeley
-  email: cboettig@berkeley.edu
+  email: cboettig@berkele.edu
 
@@
 abstract: |
-  Write something here
+  Write
 
@@
 
+Despite the complexity inherent in natural ecosystems, analytically tractable models have long played an essential role in their understanding and management.
 Ecological management may be understood as an optimal control problem - a manager must determine a sequence of actions (a "policy") to optimize some desired objective(s) and constraints (e.g.
 conservation goals, natural resource extraction, economic and social constraints.)
-Despite the complexity inherent in natural ecosystems, analytically tractable models have long played an essential role in their understanding and management.
-...
-
-<!-- Why fisheries-focused examples -->
-Global fisheries management has long been a crucible for both the development of theory and best practices in actual management for many reasons.
-Fisheries are an essential natural resource that provide the primary source of protein for one in every four humans, and have faced widely documented declines due to overfishing [@Worm2006, @Costello2016].
-Because of this importance, fisheries also represent some of the longest and best studied timeseries of population dynamics available to ecologists, as well as a correspondingly rich area for the development of theory and practice.
-Methods developed in the study and management of fisheries are frequently applied to other areas of ecological management, from concepts such as maximum sustainable yield [], optimal control under observation uncertainty [@Memarzadeh2019], to adaptive management [@Walters], management strategy evaluation [@Punt], to the studies of conservation policy and international negotiations [@].
-
-<!-- A sweeping summary of fisheries decision theory, don't bite me-->
-To understand the importance of a potential application of novel data-driven methods to the real world, it is most instructive to have a clearer picture of how models are currently used to inform ecological management in fisheries.
-In any issue as large and contentious as fisheries management, it may be far more difficult to provide a generalization  about how things actually work than to make sweeping recommendations about how it ought to.
-All the same, it is our firm belief that any researcher seeking to do the latter should first attempt to understand the former, so we beg the forbearance of any specialists in the practice as we seek to provide a concise overview of fisheries management today.
-Here, we focus specifically on practices in data rich, 'well-managed' fisheries, which, despite some issues, may be among the best examples of ecological management at continental scale that is largely driven by scientific processes, models and data.
-
-<!-- Fisheries population models are rich but decisions aren't, mostly constant mortality -->
-While fisheries stock assessment models are quite rich, the choice of possible strategies remains motivated by intution from much simpler, analytically tractable one-dimensional models.
-This fact is not always obvious at cursory inspection, and can be met with cynicism even among professional fisheries managers -- after all, typical stock assessment models have over 100 parameters (e.g.
-see the stock assessments referenced from in @RAMLegacy) and best practices such as management strategy evaluation (MSE) evaluate and select potential policies by comparing across large and diverse simulations [e.g.
-@Punt2016].
-However, the space of policies considered under such approaches is necessarily far smaller than the set of all possible action sequences.
-Instead, fisheries management approaches such as MSE rely on theory based on more tractable models to restrict the search space of possible policies.  
-Specifically, most fisheries policies are framed in terms of a *constant mortality* policy, (sometimes a simple piece-wise linear function, i.e.
-a single linear segment, is introduced as a precaution a very low stock sizes), where fishing 'mortality' is defined as the ratio of harvest (catch) to stock size, $F = H_t/x_t$.
-
-The intuition for a constant mortality strategy is simple and well established
-
-Constant mortality policies have a strong basis in theory as well as practical justifications.
-
-<!-- In contrast to constant mortality, the possible decision space is huge-->
-The management of any ecological system can be framed in terms of a Sequential Decision Process - 
-the manager observes the ecosystem state ($x_t$) and decides on action ($a_t$) to take (including potentially no action), potentially receiving any costs or benefits associated with that action and and state (reward $r_t$  $r_t = U(x_t,a_t)$),
-while the system evolves into a new state in response $x_{t+1} = f(x_t, a_t)$.
-When the probability of the future state can be determined by the current state and action, we say the system is Markovian and the process a Markov Decision Process (MDP).
-As in a game of chess, the space of possible sequences of actions a manager might choose over any realization of the Markov process is enormous.
-When the action and state space are discrete and sufficiently small, computational approaches such as dynamic programming may be used to guarantee an exhaustive search for a single optimal policy.
-Dynamic programming scales very poorly to larger state and action spaces.
-
-Deep reinforcement learning has opened up increasingly compelling alternative
-
-Effort $Q = E X_t$
-
-Fisheries policies are typically implemented in terms of *catch quotas*, i.e. a limit to the number of metric tonnes of fish that can be caught during a given season.
-
-and/or *effort controls*, restrictions on the fishing effort -- in numbers of boats, time of year, or days at sea
 
@@
 Figure 0: contrast 1D vs 3D strategies (see commented out diagram).
+
+```{r conceptual, echo = FALSE, message=FALSE, fig.align='center', fig.cap='(A) Classical management strategies often have the common aspect of reducing the complex dynamics of the fishery ecosystem to a single equation governing the harvested population (say, $V1$).
+where the interaction between $V1$ and its environment is summarized to a set of parameters, such as the growth rate $r$, and the carrying capacity $K$.
+An advantage of one-dimensional approaches is that the optimal policy is often known exactly, and, moreover, is intuitive.
+For example, in the logistic equation shown, the maximal sustainable yield of the system is attained at K/2.
+(B) RL allows us to solve for quantitative decision strategies considering more complex ecosystem dynamics.
+For example, in this paper, we find harvest strategies for a three-species model using a model-free RL algorithm, PPO.
+(C) We show how the RL-based solution that considers all three species finds harvest policies with higher average reward and with (D) lower probability of harvest species collapse than the classical approaches, only considering the dynamics of V1.', out.width='1\\linewidth', fig.pos='H'}
+knitr::include_graphics("figures/conceptual-figure.png")
+```
 

### 2023-01-19 code/may.Rmd
:pen:

---
  output: github_document
---

  ```{r}
knitr::opts_chunk$set(fig.width = 7)
```

```{r message=FALSE}
#library(MDPtoolbox)
library(tidyverse) # for plotting
```

```{r}
states <- seq(0,2, length=100)
actions <- seq(0,1, length.out = 10)
observations <- states
sigma_g <- 0.2
sigma_m <- 0.0

Q <- 2
V <- 15
C <- 10
reward_fn <- function(x,h) h
discount <- 0.999

r <- 0.8
K <- 1.53
q <- 2
b <- .2
# a <- 0.27 # only one (high) stable state, but creates nonlinear policy
a <-   0.28 # .27 #0.29

f <- function(x, h){ # May
  # h is conservation effort

  x + x * r * (1 - x / K) -  a * x ^ q / (x ^ q + b ^ q) - h * x
}
```

```{r}
# B_MSY - biomass max growth rate
X <- seq(0,1.2, length.out=200)
V <- map_dbl(X, \(x) f(x,0) - x)

tibble(X, V) |> ggplot(aes(X, V)) + geom_point()

i <- which.max(V)
B_MSY <- X[i]

F_MSY <- (f(X[i],0) - X[i]) / X[i]

```

```{r}

df <- map_dfr(seq(0, 1, length.out = 31),
function(init) {
  df <- tibble(init, time = 1:50, x = init)
  for(t in 1:max(df$time)) {
    df[t+1,] = tibble(init, time = t+1, x = f(df$x[t], F)) 
  }
  df
})
df2 <- df |> mutate(h = F_MSY * x)
df2 |> ggplot(aes(time,x, col=init, group=init)) +
  geom_line() + ylim(0,1) + 
  geom_hline(aes(yintercept=X[i]))
```
```{r}

  df <- map_dfr(seq(0, 1, length.out = 31),
  function(init) {
    df <- tibble(init, time = 1:50, x = init)
    for(t in 1:max(df$time)) {
      effort = pmax( (df$x[t] - B_MSY) /  df$x[t], 0)
      df[t+1,] = tibble(init, time = t+1, x = f(df$x[t], effort)) 
    }
    df
  })
  df |> ggplot(aes(time,x, col=init, group=init)) +
    geom_line() + ylim(0,1)
```

```{r}
df <- 
bind_rows(
  tibble(policy="const mortality", x = seq(0,1.2, length.out=100), delta = f(x,F_MSY) - x),
  tibble(policy="const escapement", x = seq(0,1.2, length.out=100), delta = f(x,0) - x),
  )
df %>% 
  ggplot(aes(x, delta, col=policy, group=policy)) + 
  geom_line() + geom_hline(aes(yintercept=0))
```

### 2023-01-19 manuscript/manuscript.Rmd
:pen:

--- 2023-01-19 manuscript/manuscript.Rmd
@@
 In this section we introduce some models describing the population dynamics of fisheries.
-In general, the class of models that appear in this context are *first order finite difference equations*.
+In general, the class of models that appear in this context are *first order finite difference equations* (potentially stochastic difference equations).
 For $n$ species, these models have the general form
+
 \begin{align}
   \label{eq:general model}
-  N_{t+1} - N_t = f(N_t) - h_t := \Delta N_t,
+  N_{t+1} - N_t = f(N_t) - F N_t := \Delta N_t,
 \end{align}
-where $N_t = (X_t,\ Y_t,\ \dots) \in \mathbb{R}^{n}_+$ is a vector of populations, $f:\mathbb{R}^{n}\to\mathbb{R}^{n}$ is an arbitrary function, and $h_t$ is the harvest size at timestep $t$.
+where $N_t = (X_t,\ Y_t,\ \dots) \in \mathbb{R}^{n}_+$ is a vector of populations, $f:\mathbb{R}^{n}\to\mathbb{R}^{n}$ is an arbitrary function, where the product of the mortality and the stock size, $F N_t$ is the harvest size at timestep $t$.
 
@@
 
-Most commonly, fisheries use 1-dimensional models, $n=1$.
-One such model appearing in a variety of ecological contexts is *logistic growth*, for which
+Optimal control policies for fisheries are frequently based on 1-dimensional models, $n=1$.
+The most familiar model of $f(X)$ is that of *logistic growth*, for which
+
 $$
@@
 $$
-More relevant to our current work, is the following single-species model studied in \cite{may77},
+Which combined with the harvest term from \eqref(eq:general model) is known as the Gordon-Schaefer model, after foundational work which first developed the concept of maximum sustainable yield, MSY, [@Gordon1954; @Schaefer1954], which remains a central concept in fisheries management.
+In this model, note that the maximum growth rate occurs at population size of $N = K/2$, which is approached asymptotically from any positive initial state $N_0 \in (0,K)$ under a constant mortality of $F = r/2$, which thus produces a harvest of $rK/4$ at equilibrium -- the maximum yield sustainable indefinitely.
+
+Real world ecological systems are obviously far more complicated than this simple model suggests.  
+One particularly important aspect that has garnered much attention is the potential for the kind of highly non-linear functions that can support dynamics such as alternative stable states and hysteresis.
+A seminal example of such dynamics was introduced in \cite{may77}, using a one-dimensional model of a prey (resource) species under the pressure of a (fixed) predator:
+
 \begin{align}
   \label{eq:may}
-  \Delta X_t
-  = L(X_t;\ r, K) - F(X_t, H; \beta,c) - h_t,
+  f(X_t)
+  = L(X_t;\ r, K) - F(X_t, H; \beta,c),
 \end{align}
@@
 \begin{align*}
-  F(X,H;\ \beta,c) := \frac{\beta H X^2}{c^2 + X^2}.
+  F(X_t,H;\ \beta,c) := \frac{\beta H X_t^2}{c^2 + X_t^2}.
 \end{align*}
+
 The model has five parameters: the growth rate $r$ and carrying capacity $K$ for $X$, a constant population $H$ of a species which preys on $X$, the maximal rate of predation $\beta$, and the predation half-maximum biomass $c$.
 
-Eq. \eqref{eq:may} is an interesting study case of a *tipping point* (see Fig. \ref{tbd}).
-Holding the value of $\beta$ fixed, for intermediate values of $H$ there exist two stable fixed points for the state $V_1$ of the system, these two attractors separated by an unstable fixed point.
+Eq. \eqref{eq:may} is an interesting study case of a *tipping point* (saddle-node bifurcation) (see Fig. \ref{tbd}).
+Holding the value of $\beta$ fixed, for intermediate values of $H$ there exist two stable fixed points for the state $\hat X_t$ of the system, these two attractors separated by an unstable fixed point.
 At a certain threshold value of $H$, however, the top stable fixed point collides with the unstable fixed point and both are annihilated.
 For this value of $H$, and for higher values, only the lower fixed point remains.
+This also creates the phenomenon of *hysteresis*, where returning $H$ to its original value is not sufficient to restore $\hat X_t$ to the original stable state.
 
 This structure implies two things.
-First, that a drift in $H$ could lead to catastrophic consequences, with the population $V_1$ plummeting to the lower fixed stable point.
-Second, that if the evolution of $V_1$ is *stochastic*, then, even at values of $H$ below the threshold point, the system runs a sizeable danger of tipping over towards the lower stable point.
+First, that a drift in $H$ could lead to catastrophic consequences, with the population $X_t$ plummeting to the lower fixed stable point.
+Second, that if the evolution of $X_t$ is *stochastic*, then, even at values of $H$ below the threshold point, the system runs a sizeable danger of tipping over towards the lower stable point.
+
+<!-- Discuss control solutions of this model(?) -->
+<!-- Briefly motivate reason to consider 3d model, adding competition and predation processes -->
 
@@
 Consider the following three species *generalization* of eq. \eqref{eq:may}:
+
 \begin{align}
@@
 organized by Paul Armsworth, Alan Hastings, Megan Donahue, and Carl Towes in
-2011 seeking to pivot control from 'optimal' to 'pretty darn good'.
+2011 which first sought to emphasize 'pretty darn good' control solutions to
+more realistic problems over optimal control to idealized ones. 
+This material is based upon work supported by the National Science Foundation under Grant No. DBI-1942280.
 

### 2023-01-19 code/plots.Rmd
...
:pen:
added figs

@@
 best <- best_reward(msy)
+colnames(msy) <- c('t', 'rep', 'action', 'reward', 'X', 'Y', 'Z')
+msy
 best
@@
 ```{r}
+colnames(msy)
 msy_sim <- msy |>
     filter(action == best$action) |> 
-  pivot_longer(c("sp1", "sp2", "sp3"),
+    group_by(rep) |>
+    pivot_longer(c("X", "Y", "Z"),
                 names_to = "species", values_to = "abundance") |>
@@
 
+msy_sim
+
 msy_sim |>
-  filter(rep < 4) |>
-  ggplot(aes(t, abundance,  col=species)) + 
+    filter(rep < 5) |> # filter(rep > 8) |>
+    # group_by(rep) |>
+    ggplot(aes(t, abundance,  col=species, group =interaction(rep,species)))+ 
         geom_line() + facet_wrap(~rep)
@@
 msy_sim |> 
-  filter(rep < 30, species == "sp1") |>
+  filter(rep < 4, species == "X") |>
   ggplot(aes(t, abundance, group = interaction(rep,species),  col=species)) + 
@@
   group_by(rep) |>
-  filter(t==max(t), species=="sp1") |>
+  filter(t==max(t), species=="X") |>
   ggplot(aes(rep, reward)) + geom_col()
@@
   filter(action == actions[[i]]) |>
-  pivot_longer(c("sp1", "sp2", "sp3"),
+  pivot_longer(c("X", "Y", "Z"),
                names_to = "species", values_to = "abundance")
 tac_sim |>
-  filter(rep < 4) |>
+  filter(rep < 20) |>
   ggplot(aes(t, abundance,  col=species)) + 
@@
 ```{r}
-ppo <- read_csv("../data/PPO250.csv.gz")
-
+ppo <- read_csv("../data/ppo-200.csv")
+colnames(ppo) <- c("1", "t", "X", "Y", "Z", "action", "reward", "rep")
+head(ppo)
 ```
@@
 
-ppo |> 
-  group_by(rep) |>
+ppo_cumulative <- ppo |> group_by(rep) |> mutate(cumulative_reward = cumsum(reward))
+
+ppo_cumulative |> 
   filter(t==max(t)) |> 
-  ggplot(aes(rep, reward)) + geom_col()
+  ggplot(aes(rep, cumulative_reward)) + geom_col()
 ```
@@
 ```{r}
-ppo_sims <- ppo |> 
-  mutate(escapement = (sp1+1) - action * (sp1+1),
+ppo_sim <- ppo |> 
+  mutate(escapement = (X+1) - action * (X+1),
          effort = action) |>
-  pivot_longer(c("sp1", "sp2", "sp3"),
+  pivot_longer(c("X", "Y", "Z"),
                names_to = "species", values_to = "abundance") |>
-  mutate(abundance = abundance + 1)
+  mutate(abundance = abundance)
 
-ppo_sims |>
-  filter(rep > 20, rep < 25) |>
+ppo_sim
+
+ppo_sim |>
+  filter(rep < 4) |>
   ggplot(aes(t, abundance, col=species)) +
@@
 ```{r}
-ppo_sims |>
+ppo_sim |>
   filter(species == "sp1") |>
@@
 escapement <- read_csv("../data/escapement.csv.gz", show_col_types = FALSE)
+colnames(escapement) <- c('t', 'rep', 'action', 'reward', 'X', 'Y', 'Z')
+escapement
 
@@
   filter(action == best_e$action) |>
-  pivot_longer(c("sp1", "sp2", "sp3"),
+  pivot_longer(c("X", "Y", "Z"),
                names_to = "species", values_to = "abundance") |>
@@
 ```
+
+# Millie's suggestion for plots:
+
+```{r}
+msy_sim <- msy_sim |> mutate(type = "Opt const mortality", cumulative_reward = reward)
+tac_sim <- tac_sim |>
+    mutate(abundance = abundance + 1) |>
+    mutate(type = "80% opt const mortality", cumulative_reward = reward)
+ppo_sim <- ppo_sim |> mutate(type = "DRL")
+ppo_sim <- within(ppo_sim, rm("1", "effort", "escapement"))
+ppo_sim <- ppo_sim |>
+    group_by(rep, species) |>
+    mutate(cumulative_reward = cumsum(reward))
+```
+
+```{r}
+ppo_sim
+```
+
+```{r}
+ppo_sim <- within(ppo_sim, rm("reward"))
+msy_sim <- within(msy_sim, rm("reward"))
+tac_sim <- within(tac_sim, rm("reward"))
+all_sim <- bind_rows(msy_sim, tac_sim, ppo_sim)
+ppo_sim
+msy_sim
+tac_sim
+all_sim
+```
+
+```{r}
+all_sim |> #filter(rep < 10, species == "X") |> 
+    group_by(rep) |>
+    filter(t==max(t), rep < 10, species == "X") |>
+    ggplot(aes(rep, cumulative_reward)) + geom_col() +
+    facet_wrap(~type)
+
+msy_sim |> 
+  group_by(rep) |>
+  filter(t==max(t), species=="X", rep < 10) |>
+  ggplot(aes(rep, cumulative_reward)) + geom_col()
+
+tac_sim |> 
+  group_by(rep) |>
+  filter(t==max(t), species=="X", rep < 10) |>
+  ggplot(aes(rep, cumulative_reward)) + geom_col()
+
+all_sim |> 
+  group_by(rep, type) |>
+  filter(t==max(t), species=="X", rep < 10) |>
+  ggplot(aes(rep, cumulative_reward)) + 
+    geom_col()  + 
+    facet_wrap(~type) + 
+    labs(y = "reward", x = "repetition")
+
+```
+
+```{r}
+p1 <- all_sim |> 
+  group_by(rep, type) |>
+  filter(rep == 0) |>
+  ggplot(aes(t, abundance, col = species)) + 
+    geom_line()  + 
+    facet_wrap(~type) + 
+    labs(y = "reward", x = "repetition")
+
+#all_sim |> group_by(rep) |> filter(t==max(t), species == "X") |> group_by(t)
+```

### 2023-01-19 manuscript/manuscript.Rmd
added figs

--- 2023-01-19 manuscript/manuscript.Rmd
@@
   - \usepackage{bookmark}
+  - \usepackage{float}
   - \linenumbers
@@
 (C) We show how the RL-based solution that considers all three species finds harvest policies with higher average reward and with (D) lower probability of harvest species collapse than the classical approaches, only considering the dynamics of V1.', out.width='1\\linewidth', fig.pos='H'}
-knitr::include_graphics("figures/conceptual-figure.png")
+knitr::include_graphics("Figures/conceptual-figure.png")
 ```

### 2023-01-20 code/plots.Rmd
plots.Rmd modified to create some figures, added a figure top manuscript

@@
 ```{r}
-p1 <- all_sim |> 
+all_sim |> 
   group_by(rep, type) |>
@@
     labs(y = "reward", x = "repetition")
+all_sim |>
+    group_by(rep,type) |>
+    filter(rep<10) |>
+    ggplot(aes(t, action)) + 
+    geom_line()  + 
+    facet_wrap(~type) + 
+    labs(y = "action", x = "repetition")
 

### 2023-01-20 manuscript/manuscript.Rmd
plots.Rmd modified to create some figures, added a figure top manuscript

@@
   - \usepackage{float}
+  - \usepackage{graphicx}
   - \linenumbers
@@
 They lead to a distinctive oscillating behavior of the system state over time, with periods of relative depletion followed by periods of growth. 
-Figure \ref{fig:tbd} exemplifies this point by comparing the harvest effort and state times series obtained from these policies.
+Figure \ref{fig:tbd} exemplifies this point by comparing one times series of the system state obtained using each of these policies.
+
+\begin{figure}
+	\includegraphics[width=\linewidth]{Figures/timeseries.png}
+\caption{
+\label{fig:one timeseries}
+tbd
+}
+\end{figure}
 

### 2023-01-20 code/plots.Rmd
replaced figures

--- 2023-01-19 code/plots.Rmd
@@
 all_sim |> 
+    mutate(across(type, factor, levels = c("Opt const mortality", "80% opt const mortality", 
+                                           "DRL"))) |>
     group_by(rep, type) |>
@@
 ```{r}
-p1 <- all_sim |> 
+all_sim |> 
+    mutate(across(type, factor, levels = c("Opt const mortality", "80% opt const mortality", 
+                                           "DRL"))) |>
     group_by(rep, type) |>
@@
         labs(y = "reward", x = "repetition")
+all_sim |>
+    group_by(rep,type) |>
+    filter(rep<10) |>
+    ggplot(aes(t, action)) + 
+    geom_line()  + 
+    facet_wrap(~type) + 
+    labs(y = "action", x = "repetition")
 

### 2023-01-20 code/plots.Rmd
added code for figs

@@
 ```
+
+```{r}
+all_sim |> 
+    mutate(across(type, factor, levels = c("Opt const mortality", "80% opt const mortality", 
+                                           "DRL"))) |>
+    group_by(rep, type) |>
+    filter(t==max(t), species=="X", rep < 10) |>
+    ggplot(aes(rep,t)) + 
+        geom_col()  + 
+        facet_wrap(~type) + 
+        labs(y = "episode length", x = "repetition") +
+        theme(axis.text.x = element_blank())
+```

### 2023-01-20 manuscript/manuscript.Rmd
added code for figs

@@
   - \usepackage{graphicx}
+  - \usepackage{caption}
+  - \usepackage{subcaption}
   - \linenumbers
@@
 
+```{r}
+library(tidyverse)
+msy <- read_csv("../data/msy.csv.gz")
+ppo <- read_csv("../data/ppo-200.csv")
+escapement <- read_csv("../data/escapement.csv.gz", show_col_types = FALSE)
+
+# rename some columns for figure purposes:
+colnames(msy) <- c('t', 'rep', 'action', 'reward', 'X', 'Y', 'Z')
+colnames(ppo) <- c('1', 't', 'X', 'Y', 'Z', 'action', 'reward', 'rep')
+
+# ppo dataset uses step rewards rather than cumulative rewards, which is what is useful here.
+# so, switch stuff around:
+ppo |> 
+    group_by(rep) |> 
+    mutate(cumulative_reward = cumsum(reward)) |>
+    within(rm("reward", "1")) |>
+    rename(reward = cumulative_reward)
+
+# set up for plots:
+
+# best constant effort based on the data on ../data/msy.csv.gz
+best_action <- 0.0465
+msy_sim <- msy |>
+    filter(action == best_action) |> 
+    group_by(rep) |>
+    pivot_longer(c("X", "Y", "Z"),
+                names_to = "species", values_to = "abundance") |>
+    mutate(abundance = abundance + 1, policy = "msy") # natural units, policy name
+
+i <- which.min(abs(actions - best$action* 0.8))
+tac_sim <- msy |> 
+    filter(action == actions[[i]]) |>
+    pivot_longer(c("X", "Y", "Z"),
+        names_to = "species", values_to = "abundance") |>
+    mutate(abundance = abundance + 1, policy = "tac")
+
+ppo_sim <- ppo |> 
+    mutate(escapement = (X+1) - action * (X+1),
+        effort = action) |>
+    pivot_longer(c("X", "Y", "Z"),
+        names_to = "species", values_to = "abundance") |>
+    mutate(policy = "ppo") # already in natural units
+
+best_e <- 0.62 # optimal constant escapement, computed from ../data/escapement.csv.gz
+esc_sim <- escapement |> 
+    rename(escapement = action) |>
+    filter(escapement == best_e) |> # data had a naming problem
+    pivot_longer(c("X", "Y", "Z"),
+        names_to = "species", values_to = "abundance") |>
+    mutate(abundance = abundance + 1)
+```
+
 Our results suggest that DRL can find solutions that are more sustainable *and* more profitable than both MSY and TAC.
@@
 
+```{r fig.cap="TBD."}
+
+```
+
 \begin{figure}
-	\includegraphics[width=\linewidth]{Figures/timeseries.png}
+	\includegraphics[width=\linewidth, height = 0.2\textheight]{Figures/timeseries.png}
 \caption{
@@
 
+\begin{figure}
+    \begin{subfigure}[b]{0.5\textwidth}
+        \includegraphics[width=\linewidth]{Figures/ep_length.png}
+        \caption{Episode lengths}
+    \end{subfigure}
+    \hfill
+    \begin{subfigure}[b]{0.5\textwidth}
+        \includegraphics[width=\linewidth]{Figures/rewards.png}
+        \caption{Episode rewards}
+    \end{subfigure}
+    \caption{
+        TBD
+    }
+    \label{fig:lengths and rewards}
+\end{figure}
+
 ## Acknowledgements

### 2023-01-20 manuscript/manuscript.Rmd

@@
   address: Department of Environmental Science, Policy, and Management, University of California, Berkeley
-  email: cboettig@berkele.edu
+  email: cboettig@berkeley.edu
 
@@
 
-Despite the complexity inherent in natural ecosystems, analytically tractable models have long played an essential role in their understanding and management.
 Ecological management may be understood as an optimal control problem - a manager must determine a sequence of actions (a "policy") to optimize some desired objective(s) and constraints (e.g.
 conservation goals, natural resource extraction, economic and social constraints.)
+Despite the complexity inherent in natural ecosystems, analytically tractable models have long played an essential role in their understanding and management.
+...
+
+<!-- Why fisheries-focused examples -->
+Global fisheries management has long been a crucible for both the development of theory and best practices in actual management for many reasons.
+Fisheries are an essential natural resource that provide the primary source of protein for one in every four humans, and have faced widely documented declines due to overfishing [@Worm2006, @Costello2016].
+Because of this importance, fisheries also represent some of the longest and best studied timeseries of population dynamics available to ecologists, as well as a correspondingly rich area for the development of theory and practice.
+Methods developed in the study and management of fisheries are frequently applied to other areas of ecological management, from concepts such as maximum sustainable yield [], optimal control under observation uncertainty [@Memarzadeh2019], to adaptive management [@Walters], management strategy evaluation [@Punt], to the studies of conservation policy and international negotiations [@].
+
+<!-- A sweeping summary of fisheries decision theory, don't bite me-->
+To understand the importance of a potential application of novel data-driven methods to the real world, it is most instructive to have a clearer picture of how models are currently used to inform ecological management in fisheries.
+In any issue as large and contentious as fisheries management, it may be far more difficult to provide a generalization  about how things actually work than to make sweeping recommendations about how it ought to.
+All the same, it is our firm belief that any researcher seeking to do the latter should first attempt to understand the former, so we beg the forbearance of any specialists in the practice as we seek to provide a concise overview of fisheries management today.
+Here, we focus specifically on practices in data rich, 'well-managed' fisheries, which, despite some issues, may be among the best examples of ecological management at continental scale that is largely driven by scientific processes, models and data.
+
+<!-- Fisheries population models are rich but decisions aren't, mostly constant mortality -->
+While fisheries stock assessment models are quite rich, the choice of possible strategies remains motivated by intution from much simpler, analytically tractable one-dimensional models.
+This fact is not always obvious at cursory inspection, and can be met with cynicism even among professional fisheries managers -- after all, typical stock assessment models have over 100 parameters (e.g.
+see the stock assessments referenced from in @RAMLegacy) and best practices such as management strategy evaluation (MSE) evaluate and select potential policies by comparing across large and diverse simulations [e.g.
+@Punt2016].
+However, the space of policies considered under such approaches is necessarily far smaller than the set of all possible action sequences.
+Instead, fisheries management approaches such as MSE rely on theory based on more tractable models to restrict the search space of possible policies.  
+Specifically, most fisheries policies are framed in terms of a *constant mortality* policy, (sometimes a simple piece-wise linear function, i.e.
+a single linear segment, is introduced as a precaution a very low stock sizes), where fishing 'mortality' is defined as the ratio of harvest (catch) to stock size, $F = H_t/x_t$.
+
+The intuition for a constant mortality strategy is simple and well established
+
+Constant mortality policies have a strong basis in theory as well as practical justifications.
+
+<!-- In contrast to constant mortality, the possible decision space is huge-->
+The management of any ecological system can be framed in terms of a Sequential Decision Process - 
+the manager observes the ecosystem state ($x_t$) and decides on action ($a_t$) to take (including potentially no action), potentially receiving any costs or benefits associated with that action and and state (reward $r_t$  $r_t = U(x_t,a_t)$),
+while the system evolves into a new state in response $x_{t+1} = f(x_t, a_t)$.
+When the probability of the future state can be determined by the current state and action, we say the system is Markovian and the process a Markov Decision Process (MDP).
+As in a game of chess, the space of possible sequences of actions a manager might choose over any realization of the Markov process is enormous.
+When the action and state space are discrete and sufficiently small, computational approaches such as dynamic programming may be used to guarantee an exhaustive search for a single optimal policy.
+Dynamic programming scales very poorly to larger state and action spaces.
+
+Deep reinforcement learning has opened up increasingly compelling alternative
+
+Effort $Q = E X_t$
+
+Fisheries policies are typically implemented in terms of *catch quotas*, i.e. a limit to the number of metric tonnes of fish that can be caught during a given season.
+
+and/or *effort controls*, restrictions on the fishing effort -- in numbers of boats, time of year, or days at sea
 
@@
 organized by Paul Armsworth, Alan Hastings, Megan Donahue, and Carl Towes in
-2011 that first posed the question addressed here.
+2011 which first sought to emphasize 'pretty darn good' control solutions to
+more realistic problems over optimal control to idealized ones. 
+This material is based upon work supported by the National Science Foundation under Grant No. DBI-1942280.
 

### 2023-01-20 code/plots.Rmd
manuscript figures, plots

@@
 }
+```
+
+```{r}
+best_reward(msy)
 ```
@@
 actions <- unique(msy$action)
-i <- which.min(abs(actions - best$action* 0.8))
+
 tac_sim <- msy |> 
@@
 ppo_sim
+ppo_sim |> within(rm("reward"))
+ppo_sim
 ```
@@
         facet_wrap(~type) + 
-        labs(y = "reward", x = "repetition")
+        labs(y = "reward", x = "repetition") +
+        theme(aspect.ratio = 2/5)
 all_sim |>

### 2023-01-20 manuscript/manuscript.Rmd
manuscript figures, plots

--- 2023-01-20 manuscript/manuscript.Rmd
@@
 colnames(ppo) <- c('1', 't', 'X', 'Y', 'Z', 'action', 'reward', 'rep')
+colnames(escapement) <- c('t', 'rep', 'action', 'reward', 'X', 'Y', 'Z')
 
@@
 
-i <- which.min(abs(actions - best$action* 0.8))
+i <- which.min(abs(actions - best_action* 0.8))
 tac_sim <- msy |> 
@@
 ppo_sim <- ppo |> 
-    mutate(escapement = (X+1) - action * (X+1),
-        effort = action) |>
     pivot_longer(c("X", "Y", "Z"),
         names_to = "species", values_to = "abundance") |>
+    within(rm("1")) |>
     mutate(policy = "ppo") # already in natural units
@@
     mutate(abundance = abundance + 1)
+
+msy_tac_ppo <- bind_rows(msy_sim, tac_sim, ppo_sim)
+#head(msy_tac_ppo)
+#head(ppo_sim)
 ```

### 2023-01-20 code/may.Rmd

@@
   for(t in 1:max(df$time)) {
-    df[t+1,] = tibble(init, time = t+1, x = f(df$x[t], F)) 
+    df[t+1,] = tibble(init, time = t+1, x = f(df$x[t], F_MSY)) 
   }

### 2023-01-21 manuscript/manuscript.Rmd
partially got the first figure to work as r code

--- 2023-01-20 manuscript/manuscript.Rmd
@@
 ---
+
+```{r}
+knitr::opts_chunk$set(cache = TRUE, echo = FALSE, message = FALSE, 
+                      warning = FALSE, fig.height=5, fig.width=7.5)
+library(tidyverse)
+```
 
@@
 ```{r}
-library(tidyverse)
+knitr::opts_chunk$set(cache = TRUE, echo = FALSE, message = FALSE, 
+                      warning = FALSE, fig.height=5, fig.width=7.5)
 msy <- read_csv("../data/msy.csv.gz")
-ppo <- read_csv("../data/ppo-200.csv")
+ppo <- read_csv("../data/PPO250.csv.gz")
 escapement <- read_csv("../data/escapement.csv.gz", show_col_types = FALSE)
@@
 colnames(msy) <- c('t', 'rep', 'action', 'reward', 'X', 'Y', 'Z')
-colnames(ppo) <- c('1', 't', 'X', 'Y', 'Z', 'action', 'reward', 'rep')
+colnames(ppo) <- c('t', 'rep', 'action', 'reward', 'X', 'Y', 'Z') 
+colnames(escapement) <- c('t', 'rep', 'action', 'reward', 'X', 'Y', 'Z')
 
@@
 # so, switch stuff around:
-ppo |> 
-    group_by(rep) |> 
-    mutate(cumulative_reward = cumsum(reward)) |>
-    within(rm("reward", "1")) |>
-    rename(reward = cumulative_reward)
+#ppo |> 
+#    group_by(rep) |> 
+#    mutate(cumulative_reward = cumsum(reward)) |>
+#    within(rm("reward", "1")) |>
+#    rename(reward = cumulative_reward)
 
@@
 
-i <- which.min(abs(actions - best$action* 0.8))
+actions <- unique(msy$action)
+i <- which.min(abs(actions - best_action* 0.8))
 tac_sim <- msy |> 
@@
 ppo_sim <- ppo |> 
-    mutate(escapement = (X+1) - action * (X+1),
-        effort = action) |>
     pivot_longer(c("X", "Y", "Z"),
         names_to = "species", values_to = "abundance") |>
-    mutate(policy = "ppo") # already in natural units
+    mutate(abundance = abundance + 1, policy = "ppo")
 
@@
 esc_sim <- escapement |> 
-    rename(escapement = action) |>
-    filter(escapement == best_e) |> # data had a naming problem
+    rename(escapement = action) |> # data had a naming problem
+    filter(escapement == best_e) |> 
     pivot_longer(c("X", "Y", "Z"),
@@
     mutate(abundance = abundance + 1)
+
+msy_tac_ppo <- bind_rows(msy_sim, tac_sim, ppo_sim)
 ```
 
-Our results suggest that DRL can find solutions that are more sustainable *and* more profitable than both MSY and TAC.
+Our results suggest that DRL can find solutions that are more sustainable *and* more profitable than any constant mortality strategy.
 Indeed, DRL finds solutions in which, for certain periods of time, no harvest is performed on the system.
@@
 ```{r fig.cap="TBD."}
-
+knitr::opts_chunk$set(cache = TRUE, echo = FALSE, message = FALSE, 
+                      warning = FALSE, fig.height=5, fig.width=7.5)
+policy_labels <- list("msy" = "Opt. Const. Mortality", "tac"= "80% Opt. Const. Mortality", "ppo"= "DRL")
+policy_labeller <- function(variable,value){
+  return(policy_labels[value])
+}
+msy_tac_ppo |> 
+    group_by(rep, policy) |>
+    filter(rep == 0) |>
+    ggplot(aes(t, abundance, col = species)) + 
+        geom_line()  + 
+        facet_wrap(~factor(policy, levels = c("msy", "tac", "ppo")), labeller = policy_labeller ) + 
+        labs(y = "abundance", x = "time") +
+        theme(aspect.ratio = 2/5)
 ```

### 2023-01-23 manuscript/manuscript.Rmd
fixed the figure thingy, erased old figure versions

@@
 
-```{r}
+```{r knit_global, include=FALSE}
 knitr::opts_chunk$set(cache = TRUE, echo = FALSE, message = FALSE, 
@@
 
-\begin{figure}
-	\includegraphics[width=\linewidth, height = 0.2\textheight]{Figures/timeseries.png}
-\caption{
-\label{fig:one timeseries}
-tbd
-}
-\end{figure}
-
 It is well known that following MSY policies can lead to over-fishing.
@@
 This difference is reflected on the profits earned by the agent, as shown in Fig. TBD-B.
-
-\begin{figure}
-    \begin{subfigure}[b]{0.5\textwidth}
-        \includegraphics[width=\linewidth]{Figures/ep_length.png}
-        \caption{Episode lengths}
-    \end{subfigure}
-    \hfill
-    \begin{subfigure}[b]{0.5\textwidth}
-        \includegraphics[width=\linewidth]{Figures/rewards.png}
-        \caption{Episode rewards}
-    \end{subfigure}
-    \caption{
-        TBD
-    }
-    \label{fig:lengths and rewards}
-\end{figure}
 

### 2023-01-23 manuscript/manuscript.Rmd
figures working nicely :)

@@
 library(tidyverse)
+library(patchwork)
 ```
@@
 
-```{r fig.cap="TBD."}
-knitr::opts_chunk$set(cache = TRUE, echo = FALSE, message = FALSE, 
-                      warning = FALSE, fig.height=5, fig.width=7.5)
+```{r timeseries_fig, fig.cap="TBD.", out.width = "6.5in"}
 policy_labels <- list("msy" = "Opt. Const. Mortality", "tac"= "80% Opt. Const. Mortality", "ppo"= "DRL")
@@
         geom_line()  + 
+        scale_color_manual(values=c('palegreen3', 'coral1', 'deepskyblue4')) +
         facet_wrap(~factor(policy, levels = c("msy", "tac", "ppo")), labeller = policy_labeller ) +
         labs(y = "abundance", x = "time") +
-        theme(aspect.ratio = 2/5)
+        theme(aspect.ratio = 2/3)
+```
+
+```{r rew_len_fig, fig.cap="TBD.", out.width = "6.5in"}
+library(patchwork)
+
+p1 <- msy_tac_ppo |> 
+    group_by(rep, policy) |>
+    filter(t==max(t), rep < 30) |>
+    ggplot(aes(rep, t)) + 
+      geom_col() +
+      facet_wrap(~factor(policy, levels = c("msy", "tac", "ppo")), labeller = policy_labeller ) +
+      labs(y = "episode length", x = "repetition") +
+      theme(axis.text.x = element_blank())
+
+p2 <- msy_tac_ppo |> 
+    group_by(rep, policy) |>
+    filter(t==max(t), rep < 30) |>
+    ggplot(aes(rep, reward)) + 
+      geom_col() +
+      facet_wrap(~factor(policy, levels = c("msy", "tac", "ppo")), labeller = policy_labeller ) +
+      labs(y = "episode reward", x = "repetition") +
+      theme(axis.text.x = element_blank())
+
+(p1 / p2)
 ```

### 2023-01-23 manuscript/manuscript.Rmd
mini
state space figure

@@
 colnames(ppo) <- c('t', 'rep', 'action', 'reward', 'X', 'Y', 'Z') 
-colnames(escapement) <- c('t', 'rep', 'action', 'reward', 'X', 'Y', 'Z')
+colnames(escapement) <- c('t', 'rep', 'esc_level', 'reward', 'X', 'Y', 'Z') 
+opt_escapement <- escapement |> filter(esc_level == 0.62)
+head(opt_escapement)
 
@@
 esc_sim <- escapement |> 
-    rename(escapement = action) |> # data had a naming problem
-    filter(escapement == best_e) |> 
     pivot_longer(c("X", "Y", "Z"),
@@
 They lead to a distinctive oscillating behavior of the system state over time, with periods of relative depletion followed by periods of growth. 
-Figure \ref{fig:tbd} exemplifies this point by comparing one times series of the system state obtained using each of these policies.
+Figure TBD exemplifies this point by comparing one times series of the system state obtained using each of these policies.
 
-```{r timeseries_fig, fig.cap="TBD.", out.width = "6.5in"}
+```{r timeseries, fig.cap="TBD.", out.width = "6.5in"}
 policy_labels <- list("msy" = "Opt. Const. Mortality", "tac"= "80% Opt. Const. Mortality", "ppo"= "DRL")
@@
 
-(p1 / p2)
+(p1 / p2) + plot_annotation(tag_levels = 'A')
 ```
 
+```{r state-space}
+i_tac <- which.min(abs(actions - best_action * 0.8))
+
+# color pallette limits
+library(RColorBrewer)
+myPalette <- colorRampPalette(rev(brewer.pal(11, "Spectral")))
+sc <- scale_colour_gradientn(colours = myPalette(100), limits=c(0,1))
+
+pmsy <- msy |>
+  mutate(X = X+1, Y = Y+1, Z = Z+1, harvest = action * X) |>
+  filter(action == best_action) |>
+  ggplot(aes(X, Z, color = harvest)) +
+  geom_point() + sc +
+  theme(aspect.ratio = 1/1)
+
+ptac <- msy |>
+  mutate(X = X+1, Y = Y+1, Z = Z+1, harvest = action * X) |>
+  filter(action == actions[[i_tac]]) |>
+  ggplot(aes(X, Z, color = harvest)) +
+  geom_point() + sc +
+  theme(aspect.ratio = 1/1)
+
+pppo <- ppo |>
+  mutate(X = X+1, Y = Y+1, Z = Z+1, harvest = action * X) |>
+  ggplot(aes(X, Z, color = harvest)) +
+  geom_point() + sc +
+  theme(aspect.ratio = 1/1)
+
+pesc <- opt_escapement |>
+  mutate(X = X+1, Y = Y+1, Z = Z+1, 
+        harvest = case_when(
+          X - esc_level > 0 ~ X - esc_level,
+          X - esc_level <= 0 ~  0)
+         ) |> 
+  ggplot(aes(X, Z, color = harvest)) +
+  geom_point() + sc +
+  theme(aspect.ratio = 1/1)
+
+(
+  (pmsy | ptac) / (pppo | pesc) +
+  plot_layout(guides = "collect") & theme(legend.position = 'right')
+) + 
+  plot_annotation(tag_levels = c('A'))
+```
+
+**Probably more for the discussion**
 It is well known that following MSY policies can lead to over-fishing.
 Our results reproduce this behavior and, furthermore, warn that using a TAC policy instead might not be enough to gain sustainability.
-Figure TBD-A summarizes episode length distribution for MSY, TAC and DRL.
+Figure \ref{rew_len_fig}-A summarizes episode length distribution for MSY, TAC and DRL.
 While the DRL-derived policy does not lead to extinctions, both MSY and TAC do so, and thus their episode mean length is lower than the maximal episode length (in our case set at 200 years).

### 2023-01-23 manuscript/manuscript.Rmd
nicer figure

@@
 
-```{r state-space}
+```{r state-space, fig.cap="TBD.", out.width = "6.5in"}
 i_tac <- which.min(abs(actions - best_action * 0.8))
@@
 myPalette <- colorRampPalette(rev(brewer.pal(11, "Spectral")))
-sc <- scale_colour_gradientn(colours = myPalette(100), limits=c(0,1))
+sc <- scale_colour_gradientn(colours = myPalette(100), limits=c(0,0.25))
 
@@
   geom_point() + sc +
-  theme(aspect.ratio = 1/1)
+  ggtitle("Opt. Const. Mortality")
 
@@
   geom_point() + sc +
-  theme(aspect.ratio = 1/1)
+  ggtitle("80% Opt. Const. Mortality")
 
@@
   geom_point() + sc +
-  theme(aspect.ratio = 1/1)
+  ggtitle("DRL")
 
@@
   geom_point() + sc +
-  theme(aspect.ratio = 1/1)
+  ggtitle("Opt. Const. Escapement")
 

### 2023-01-23 manuscript/manuscript.Rmd
added escapement-msy tuning fig

@@
 library(patchwork)
+```
+
+```{r}
+# Collect data (outsource to separate R script later TBD)
+msy <- read_csv("../data/msy.csv.gz")
+ppo <- read_csv("../data/PPO250.csv.gz")
+escapement <- read_csv("../data/escapement.csv.gz", show_col_types = FALSE)
+
+# rename some columns for figure purposes:
+colnames(msy) <- c('t', 'rep', 'action', 'reward', 'X', 'Y', 'Z')
+colnames(ppo) <- c('t', 'rep', 'action', 'reward', 'X', 'Y', 'Z') 
+colnames(escapement) <- c('t', 'rep', 'esc_level', 'reward', 'X', 'Y', 'Z') 
+opt_escapement <- escapement |> filter(esc_level == 0.62)
+
+# ppo dataset uses step rewards rather than cumulative rewards, which is what is useful here.
+# so, switch stuff around:
+#ppo |> 
+#    group_by(rep) |> 
+#    mutate(cumulative_reward = cumsum(reward)) |>
+#    within(rm("reward", "1")) |>
+#    rename(reward = cumulative_reward)
+
+# set up for plots:
+
+# best constant effort based on the data on ../data/msy.csv.gz
+best_action <- 0.0465
+msy_sim <- msy |>
+    filter(action == best_action) |> 
+    group_by(rep) |>
+    pivot_longer(c("X", "Y", "Z"),
+                names_to = "species", values_to = "abundance") |>
+    mutate(abundance = abundance + 1, policy = "msy") # natural units, policy name
+
+actions <- unique(msy$action)
+i <- which.min(abs(actions - best_action* 0.8))
+tac_sim <- msy |> 
+    filter(action == actions[[i]]) |>
+    pivot_longer(c("X", "Y", "Z"),
+        names_to = "species", values_to = "abundance") |>
+    mutate(abundance = abundance + 1, policy = "tac")
+
+ppo_sim <- ppo |> 
+    pivot_longer(c("X", "Y", "Z"),
+        names_to = "species", values_to = "abundance") |>
+    mutate(abundance = abundance + 1, policy = "ppo")
+
+best_e <- 0.62 # optimal constant escapement, computed from ../data/escapement.csv.gz
+esc_sim <- escapement |> 
+    pivot_longer(c("X", "Y", "Z"),
+        names_to = "species", values_to = "abundance") |>
+    mutate(abundance = abundance + 1)
+
+msy_tac_ppo <- bind_rows(msy_sim, tac_sim, ppo_sim)
 ```
@@
 
+# Fishery management figures
+
+```{r escapement_msy, fig.cap="TBD.", out.width = "6.5in"}
+library(patchwork)
+msy_rewards <- msy |>
+  group_by(action, rep) |>
+  filter(t == max(t)) |> 
+  group_by(action) |>
+  summarise(mean_reward = mean(reward),
+            sd = sd(reward))
+
+esc_rewards <- escapement |>
+  group_by(esc_level, rep) |>
+  filter(t == max(t)) |> 
+  group_by(esc_level) |>
+  summarise(mean_reward = mean(reward),
+            sd = sd(reward))
+
+rmsy <- msy_rewards |>
+  ggplot(aes(action, mean_reward, 
+             ymin = mean_reward - 2*sd,
+             ymax=mean_reward+2*sd)) +
+  geom_point() + geom_ribbon(alpha=0.4) +
+  labs(y = "episode reward", x = "constant mortality") +
+  theme(aspect.ratio = 1/1)
+
+resc <- esc_rewards |>
+  ggplot(aes(esc_level, mean_reward, 
+             ymin = mean_reward - 2*sd,
+             ymax=mean_reward+2*sd)) +
+  geom_point() + geom_ribbon(alpha=0.4) +
+  labs(y = "episode reward", x = "constant escapement") +
+  theme(aspect.ratio = 1/1)
+
+(rmsy | resc)
+```
+
 # Fishery management approaches
@@
 
-```{r}
-knitr::opts_chunk$set(cache = TRUE, echo = FALSE, message = FALSE, 
-                      warning = FALSE, fig.height=5, fig.width=7.5)
-msy <- read_csv("../data/msy.csv.gz")
-ppo <- read_csv("../data/PPO250.csv.gz")
-escapement <- read_csv("../data/escapement.csv.gz", show_col_types = FALSE)
-
-# rename some columns for figure purposes:
-colnames(msy) <- c('t', 'rep', 'action', 'reward', 'X', 'Y', 'Z')
-colnames(ppo) <- c('t', 'rep', 'action', 'reward', 'X', 'Y', 'Z') 
-colnames(escapement) <- c('t', 'rep', 'esc_level', 'reward', 'X', 'Y', 'Z') 
-opt_escapement <- escapement |> filter(esc_level == 0.62)
-head(opt_escapement)
-
-# ppo dataset uses step rewards rather than cumulative rewards, which is what is useful here.
-# so, switch stuff around:
-#ppo |> 
-#    group_by(rep) |> 
-#    mutate(cumulative_reward = cumsum(reward)) |>
-#    within(rm("reward", "1")) |>
-#    rename(reward = cumulative_reward)
-
-# set up for plots:
-
-# best constant effort based on the data on ../data/msy.csv.gz
-best_action <- 0.0465
-msy_sim <- msy |>
-    filter(action == best_action) |> 
-    group_by(rep) |>
-    pivot_longer(c("X", "Y", "Z"),
-                names_to = "species", values_to = "abundance") |>
-    mutate(abundance = abundance + 1, policy = "msy") # natural units, policy name
-
-actions <- unique(msy$action)
-i <- which.min(abs(actions - best_action* 0.8))
-tac_sim <- msy |> 
-    filter(action == actions[[i]]) |>
-    pivot_longer(c("X", "Y", "Z"),
-        names_to = "species", values_to = "abundance") |>
-    mutate(abundance = abundance + 1, policy = "tac")
-
-ppo_sim <- ppo |> 
-    pivot_longer(c("X", "Y", "Z"),
-        names_to = "species", values_to = "abundance") |>
-    mutate(abundance = abundance + 1, policy = "ppo")
-
-best_e <- 0.62 # optimal constant escapement, computed from ../data/escapement.csv.gz
-esc_sim <- escapement |> 
-    pivot_longer(c("X", "Y", "Z"),
-        names_to = "species", values_to = "abundance") |>
-    mutate(abundance = abundance + 1)
-
-msy_tac_ppo <- bind_rows(msy_sim, tac_sim, ppo_sim)
-```
-
 Our results suggest that DRL can find solutions that are more sustainable *and* more profitable than any constant mortality strategy.

### 2023-01-23 manuscript/manuscript.Rmd
cant get the msy-escapement tuning figure to print out correctly

@@
 
-# Fishery management figures
-
-```{r escapement_msy, fig.cap="TBD.", out.width = "6.5in"}
+```{r classical_tuning, fig.cap = "TBD. Red = MSY, Green = 80% TAC", out.width = "6.5in"}
 library(patchwork)
@@
   labs(y = "episode reward", x = "constant mortality") +
-  theme(aspect.ratio = 1/1)
+  theme(aspect.ratio = 1/1) + 
+  geom_vline(aes(xintercept=0.0465), color="red") + 
+  geom_vline(aes(xintercept=0.0370), color="green")
 
@@
 i_tac <- which.min(abs(actions - best_action * 0.8))
+actions[[i_tac]]
 

### 2023-01-23 manuscript/manuscript.Rmd
it was because of the % sign in the caption

@@
 
-```{r classical_tuning, fig.cap = "TBD. Red = MSY, Green = 80% TAC", out.width = "6.5in"}
-library(patchwork)
+```{r classical_tuning, fig.cap = "TBD.  Red = MSY, Green = 0.8 TAC", out.width = "6.5in"}
 msy_rewards <- msy |>

### 2023-01-23 manuscript/manuscript.Rmd
just getting git to commit a folder name change
added may stable point fig, and a caption

@@
 (C) We show how the RL-based solution that considers all three species finds harvest policies with higher average reward and with (D) lower probability of harvest species collapse than the classical approaches, only considering the dynamics of V1.', out.width='1\\linewidth', fig.pos='H'}
-knitr::include_graphics("Figures/conceptual-figure.png")
+knitr::include_graphics("Figures/conceptual-figure-old.png")
 ```
@@
   \label{eq:general model}
-  N_{t+1} - N_t = f(N_t) - F N_t := \Delta N_t,
+  \Delta N_t := N_{t+1} - N_t = f(N_t) - F^T N_t,
 \end{align}
-where $N_t = (X_t,\ Y_t,\ \dots) \in \mathbb{R}^{n}_+$ is a vector of populations, $f:\mathbb{R}^{n}\to\mathbb{R}^{n}$ is an arbitrary function, where the product of the mortality and the stock size, $F N_t$ is the harvest size at timestep $t$.
+where $N_t = (X_t,\ Y_t,\ \dots) \in \mathbb{R}^{n}_+$ is a vector of populations, $f:\mathbb{R}^{n}\to\mathbb{R}^{n}$ is an arbitrary function, where the product of the mortality and the stock size, $F^T N_t$ is the harvest size at timestep $t$.
+Here, $F\in\mathbb{R}^n$ is a vector of fishing intensities, where $F_i$ is the intensity with which the $i$-th species is fished.
+We will focus on the case where only the first species is fished, $F= (F_1,\ 0,\ \dots,\ 0)$.
 
@@
 $$
-Which combined with the harvest term from \eqref(eq:general model) is known as the Gordon-Schaefer model, after foundational work which first developed the concept of maximum sustainable yield, MSY, [@Gordon1954; @Schaefer1954], which remains a central concept in fisheries management.
-In this model, note that the maximum growth rate occurs at population size of $N = K/2$, which is approached asymptotically from any positive initial state $N_0 \in (0,K)$ under a constant mortality of $F = r/2$, which thus produces a harvest of $rK/4$ at equilibrium -- the maximum yield sustainable indefinitely.
+The model this $f$ gives rise to is known as the Gordon-Schaefer model, after foundational work which first developed the concept of maximum sustainable yield, MSY, [@Gordon1954; @Schaefer1954], which remains a central concept in fisheries management.
+In this model, the maximum growth rate occurs at population size of $N = K/2$.
+This value is approached asymptotically, from any positive initial state $N_0 \in (0,K)$, under a constant mortality $F = r/2$ policy.
+At equilibrium this produces an asymptotic harvest of $rK/4$ per time step---the maximum yield sustainable indefinitely.
 
@@
   \label{eq:may}
-  f(X_t)
-  = L(X_t;\ r, K) - F(X_t, H; \beta,c),
+  f_{\text{May}}(X_t)
+  := L(X_t;\ r, K) - F(X_t, H; \beta,c),
 \end{align}
@@
 
-Eq. \eqref{eq:may} is an interesting study case of a *tipping point* (saddle-node bifurcation) (see Fig. \ref{tbd}).
+Eq. \eqref{eq:may} is an interesting study case of a *tipping point* (saddle-node bifurcation) (see Fig. \ref{fig:may stable}).
 Holding the value of $\beta$ fixed, for intermediate values of $H$ there exist two stable fixed points for the state $\hat X_t$ of the system, these two attractors separated by an unstable fixed point.
@@
 This also creates the phenomenon of *hysteresis*, where returning $H$ to its original value is not sufficient to restore $\hat X_t$ to the original stable state.
+
+\begin{figure}
+  \centering
+  \includegraphics[width = 0.8\linewidth]{figures/MayStablePoints.pdf}
+  \caption{
+    \label{fig:may stable}
+    The fixed point diagram for the model \cite{may77} associated to $f_{\text{May}}$ as a function of varying the parameter $\beta H$.
+    That is, these are the zeroes of the function $f_{\text{May}}$. 
+    Stable fixed points (also known as *attractors*) are plotted using a solid line, while the unstable fixed point is shown as a dotted line.
+  }
+\end{figure}
 
@@
 i_tac <- which.min(abs(actions - best_action * 0.8))
-actions[[i_tac]]
 

### 2023-01-23 manuscript/manuscript.Rmd
latex style conceptual figure for cross referencing

@@
 
+\begin{figure}
+  \centering
+  \includegraphics[width = 0.8\linewidth]{figures/conceptual-figure-old.png}
+  \caption{
+    \label{fig:conceptual}
+    (A) Classical management strategies often have the common aspect of reducing the complex dynamics of the fishery ecosystem to a single equation governing the harvested population (say, $V1$).
+where the interaction between $V1$ and its environment is summarized to a set of parameters, such as the growth rate $r$, and the carrying capacity $K$.
+An advantage of one-dimensional approaches is that the optimal policy is often known exactly, and, moreover, is intuitive.
+For example, in the logistic equation shown, the maximal sustainable yield of the system is attained at K/2.
+(B) RL allows us to solve for quantitative decision strategies considering more complex ecosystem dynamics.
+For example, in this paper, we find harvest strategies for a three-species model using a model-free RL algorithm, PPO.
+(C) We show how the RL-based solution that considers all three species finds harvest policies with higher average reward and with (D) lower probability of harvest species collapse than the classical approaches, only considering the dynamics of V1.
+  }
+\end{figure}
+
+<!--
 ```{r conceptual, echo = FALSE, message=FALSE, fig.align='center', fig.cap='(A) Classical management strategies often have the common aspect of reducing the complex dynamics of the fishery ecosystem to a single equation governing the harvested population (say, $V1$).
@@
 ```
+-->
 
@@
 \end{align}
-The model is conceptualized in Fig. \ref{fig:tbd}.
+The model is conceptualized in Fig. \ref{fig:conceptual}, panel B.
 It contains three populations, the state at time $t$ is $(X_t, Y_t, Z_t)$. 

### 2023-01-24 manuscript/manuscript.Rmd
added flow chart

@@
   - \usepackage{subcaption}
+  - \usepackage{tikz-cd}
   - \linenumbers
@@
 
-$$
+\begin{align}
+  \label{eq:logistic}
   f(X_t) = r X_t\big(1 - X_t / K \big) =: L(X_t;\ r, K).
-$$
+\end{align}
+
 The model this $f$ gives rise to is known as the Gordon-Schaefer model, after foundational work which first developed the concept of maximum sustainable yield, MSY, [@Gordon1954; @Schaefer1954], which remains a central concept in fisheries management.
@@
 
-\textbf{Probably just merge on previous sections.}
-We consider a stochastic contribution to the system's dynamics.
-This way, eq. \eqref{eq:may} becomes
+We have, up to now, presented eqs. \eqref{eq:may} and \eqref{eq:3d model} as *deterministic* dynamical models for the sake of clarity.
+In practice, we will consider a stochastic contribution to these dynamics.
+This way, instead of eq. \eqref{eq:may}, we consider the following stochastic single-species system
 \begin{align}
@@
 
-Similarly, eq. \eqref{eq:3d model} becomes
+Similarly, we use the following dynamical model in place of eq. \eqref{eq:3d model}
 \begin{align*}
@@
 \end{align*}
+Here, again, $\{\eta_{i,t}\}$ is a set of independent Gaussian variables centered at zero with variances, respectively, $\sigma_X^2$, $\sigma_Y^2$ and $\sigma_Z^2$.
 
@@
 Here we review classical strategies for sustainable fishery management and we provide a birds-eye view of the alternative approach we propose.
-A summary of the contrast between the two strategies is given in Fig. \ref{tbd}.
+A summary of the contrast between the two strategies is given in Fig. \ref{fig:decision approaches}.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.8\linewidth]{figures/DecisionApproaches.pdf}
+  \caption{
+  \label{fig:decision approaches}
+  TBD.
+  }
+\end{figure}
 
@@
 We collectively refere to these as *classical*, and will compare their performance to RL-based management strategies.
-As shown in Fig.
-\ref{fig:tbd}, classical strategies have the common aspect of reducing the complex dynamics of the fishery ecosystem to a single equation governing the harvested population (say, $F$).
+Classical strategies derive an optimal harvest policy using a simple model for the system dynamics.
+While often complex models of the ecosystem are used to estimate the fish population of interest (i.e. $X_t$), the quota is set only taking into account the estimate $X_t$.
+The function determining this quota is derived from a simple dynamical model for the population (see text box 1).
+
+\textbf{Text box 1 (TBD formatting)}
+Let us consider a ubiquitous simple example: an MSY policy based on a logistic growth model for $X_t$ (see eqs. \eqref{eq:general model} and \eqref{eq:logistic}).
+
+Classical strategies have the common aspect of reducing the complex dynamics of the fishery ecosystem to a single equation governing the harvested population (say, $F$). 
 A common example is using a logistic growth equation,

### 2023-01-24 manuscript/manuscript.Rmd
added escapement timeseries

@@
 
+```{r escapement_timeseries, fig.cap = "TBD", out.width = "6.5in"}
+
+ppo_mod <- ppo |> 
+  mutate(X = X+1, Y = Y+1, Z = Z+1, escapement = (1-action)*X, time = t)
+
+avg_esc <- mean(ppo_mod$escapement)
+# = 0.6201131
+
+ppo_esc <- ppo_mod |>
+  filter(rep==0) |>
+  group_by(rep) |>
+  ggplot(aes(time, escapement), alpha = rep) +
+  geom_line() +
+  geom_hline(aes(yintercept=avg_esc), color="red")+
+  theme(aspect.ratio = 1/5)
+
+ppo_esc
+```
+
 **Probably more for the discussion**

### 2023-01-24 manuscript/manuscript.Rmd
added a bit of discussion on methods

@@
 
-\begin{figure}
-  \centering
-  \includegraphics[width=0.8\linewidth]{figures/DecisionApproaches.pdf}
-  \caption{
-  \label{fig:decision approaches}
-  TBD.
-  }
-\end{figure}
-
 ## Classical approaches to sustainable fisheries {#sec:classical}
@@
 
-# Results
+# Methods and Results
 
+## Methods
+
+Model-based controls, such as MSY, are known to have serious deficiencies when the model used to find the optimal control does not match reality.
+As mentioned, for the particular case of fisheries this can lead to depletion of resources and affect severely the surrounding ecosystems.
+Because of this, we have opted to compare the performance of several data-based approaches to fishery control (see Fig. \ref{fig:decision approaches}).
+
+As shown in Fig.
+\ref{fig:decision approaches}, these data-based strategies rely on choosing a family of *approximating functions* for the policy, $\mathcal{P}:=\{\pi_\theta\}_\theta$, parametrized by some vector $\theta$.
+These functions are deterministic policies, i.e. they map states to actions, or, specifically,
+$$
+  \pi_\theta : \mathbb{R}^n \to \mathbb{R}_+,
+$$
+where, as before, $n$ is the number of species.
+The aim of the optimization process is to find a parameter choice $\theta^*$ for which $\pi_{\theta^*}$ is optimal within $\mathcal{P}$.
+
+Our focus here will be comparing two data-based approaches: *constant mortality* and *DRL*.
+Constant mortality is, as the name suggests, based on MSY.
+
+For the constant mortality approach, we choose a linear harvest $\pi_F(N_t)= F X_t$, where the parameter $F$ is scalar, and where $N_t = (X_t,\ \dots\ )\in \mathbb{R}^n$.
+This first control is intended to delineate how one might apply MSY to a real-world fishery.
+We will also compare a TAC-inspired rule of thumb which uses, instead, the parameter value $F_{\text{TAC}} = 0.8 F^*$.
+In App. \ref{app:tbd}, we discuss several other MSY-based approaches which yield similar results to the strategies just discussed.
+
+In the DRL approach, we instead parametrize $\mathcal{P}$ using a neural network with two hidden layers of 64 neurons.
+Our hypothesis is that the additional flexibility yielded by neural networks will allow the agent solutions that are better than MSY based approaches.
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=0.8\linewidth]{figures/DecisionApproaches.pdf}
+  \caption{
+  \label{fig:decision approaches}
+  TBD.
+  }
+\end{figure}
+
+## Results
+
+\textbf{Reword a bit.}
 We trained a DRL agent to obtain a candidate policy

### 2023-01-24 manuscript/manuscript.Rmd
readabilility, bump renv

--- 2023-01-24 manuscript/manuscript.Rmd
@@
   - \usepackage{subcaption}
-  - \usepackage{tikz-cd}
   - \linenumbers
@@
 library(patchwork)
-```
-
-```{r}
-# Collect data (outsource to separate R script later TBD)
-msy <- read_csv("../data/msy.csv.gz")
-ppo <- read_csv("../data/PPO250.csv.gz")
-escapement <- read_csv("../data/escapement.csv.gz", show_col_types = FALSE)
-
-# rename some columns for figure purposes:
-colnames(msy) <- c('t', 'rep', 'action', 'reward', 'X', 'Y', 'Z')
-colnames(ppo) <- c('t', 'rep', 'action', 'reward', 'X', 'Y', 'Z') 
-colnames(escapement) <- c('t', 'rep', 'esc_level', 'reward', 'X', 'Y', 'Z') 
-opt_escapement <- escapement |> filter(esc_level == 0.62)
-
-# ppo dataset uses step rewards rather than cumulative rewards, which is what is useful here.
-# so, switch stuff around:
-#ppo |> 
-#    group_by(rep) |> 
-#    mutate(cumulative_reward = cumsum(reward)) |>
-#    within(rm("reward", "1")) |>
-#    rename(reward = cumulative_reward)
-
-# set up for plots:
-
-# best constant effort based on the data on ../data/msy.csv.gz
-best_action <- 0.0465
-msy_sim <- msy |>
-    filter(action == best_action) |> 
-    group_by(rep) |>
-    pivot_longer(c("X", "Y", "Z"),
-                names_to = "species", values_to = "abundance") |>
-    mutate(abundance = abundance + 1, policy = "msy") # natural units, policy name
-
-actions <- unique(msy$action)
-i <- which.min(abs(actions - best_action* 0.8))
-tac_sim <- msy |> 
-    filter(action == actions[[i]]) |>
-    pivot_longer(c("X", "Y", "Z"),
-        names_to = "species", values_to = "abundance") |>
-    mutate(abundance = abundance + 1, policy = "tac")
-
-ppo_sim <- ppo |> 
-    pivot_longer(c("X", "Y", "Z"),
-        names_to = "species", values_to = "abundance") |>
-    mutate(abundance = abundance + 1, policy = "ppo")
-
-best_e <- 0.62 # optimal constant escapement, computed from ../data/escapement.csv.gz
-esc_sim <- escapement |> 
-    pivot_longer(c("X", "Y", "Z"),
-        names_to = "species", values_to = "abundance") |>
-    mutate(abundance = abundance + 1)
-
-msy_tac_ppo <- bind_rows(msy_sim, tac_sim, ppo_sim)
 ```
@@
 
-\begin{figure}
-  \centering
-  \includegraphics[width = 0.8\linewidth]{figures/conceptual-figure-old.png}
-  \caption{
-    \label{fig:conceptual}
-    (A) Classical management strategies often have the common aspect of reducing the complex dynamics of the fishery ecosystem to a single equation governing the harvested population (say, $V1$).
+```{r}
+
+#| label: conceptual
+#| echo: FALSE
+#| message" FALSE
+#| fig.align: 'center'
+#| fig.cap: '(A) Classical management strategies often have the common aspect of reducing the complex dynamics of the fishery ecosystem to a single equation governing the harvested population (say, $V1$).
 where the interaction between $V1$ and its environment is summarized to a set of parameters, such as the growth rate $r$, and the carrying capacity $K$.
@@
 For example, in the logistic equation shown, the maximal sustainable yield of the system is attained at K/2.
-(B) RL allows us to solve for quantitative decision strategies considering more complex ecosystem dynamics.
+#| (B) RL allows us to solve for quantitative decision strategies considering more complex ecosystem dynamics.
 For example, in this paper, we find harvest strategies for a three-species model using a model-free RL algorithm, PPO.
-(C) We show how the RL-based solution that considers all three species finds harvest policies with higher average reward and with (D) lower probability of harvest species collapse than the classical approaches, only considering the dynamics of V1.
-  }
-\end{figure}
+#| (C) We show how the RL-based solution that considers all three species finds harvest policies with higher average reward and with (D) lower probability of harvest species collapse than the classical approaches, only considering the dynamics of V1.',
+#| out.width: '1\\linewidth', 
+#| fig.pos: 'H'
 
-<!--
-```{r conceptual, echo = FALSE, message=FALSE, fig.align='center', fig.cap='(A) Classical management strategies often have the common aspect of reducing the complex dynamics of the fishery ecosystem to a single equation governing the harvested population (say, $V1$).
-where the interaction between $V1$ and its environment is summarized to a set of parameters, such as the growth rate $r$, and the carrying capacity $K$.
-An advantage of one-dimensional approaches is that the optimal policy is often known exactly, and, moreover, is intuitive.
-For example, in the logistic equation shown, the maximal sustainable yield of the system is attained at K/2.
-(B) RL allows us to solve for quantitative decision strategies considering more complex ecosystem dynamics.
-For example, in this paper, we find harvest strategies for a three-species model using a model-free RL algorithm, PPO.
-(C) We show how the RL-based solution that considers all three species finds harvest policies with higher average reward and with (D) lower probability of harvest species collapse than the classical approaches, only considering the dynamics of V1.', out.width='1\\linewidth', fig.pos='H'}
-knitr::include_graphics("Figures/conceptual-figure-old.png")
+knitr::include_graphics("Figures/conceptual-figure.png")
 ```
--->
 
@@
   \label{eq:general model}
-  \Delta N_t := N_{t+1} - N_t = f(N_t) - F^T N_t,
+  N_{t+1} - N_t = f(N_t) - F N_t := \Delta N_t,
 \end{align}
-where $N_t = (X_t,\ Y_t,\ \dots) \in \mathbb{R}^{n}_+$ is a vector of populations, $f:\mathbb{R}^{n}\to\mathbb{R}^{n}$ is an arbitrary function, where the product of the mortality and the stock size, $F^T N_t$ is the harvest size at timestep $t$.
-Here, $F\in\mathbb{R}^n$ is a vector of fishing intensities, where $F_i$ is the intensity with which the $i$-th species is fished.
-We will focus on the case where only the first species is fished, $F= (F_1,\ 0,\ \dots,\ 0)$.
+where $N_t = (X_t,\ Y_t,\ \dots) \in \mathbb{R}^{n}_+$ is a vector of populations, $f:\mathbb{R}^{n}\to\mathbb{R}^{n}$ is an arbitrary function, where the product of the mortality and the stock size, $F N_t$ is the harvest size at timestep $t$.
 
@@
 
-\begin{align}
-  \label{eq:logistic}
+$$
   f(X_t) = r X_t\big(1 - X_t / K \big) =: L(X_t;\ r, K).
-\end{align}
-
-The model this $f$ gives rise to is known as the Gordon-Schaefer model, after foundational work which first developed the concept of maximum sustainable yield, MSY, [@Gordon1954; @Schaefer1954], which remains a central concept in fisheries management.
-In this model, the maximum growth rate occurs at population size of $N = K/2$.
-This value is approached asymptotically, from any positive initial state $N_0 \in (0,K)$, under a constant mortality $F = r/2$ policy.
-At equilibrium this produces an asymptotic harvest of $rK/4$ per time step---the maximum yield sustainable indefinitely.
+$$
+Which combined with the harvest term from \eqref(eq:general model) is known as the Gordon-Schaefer model, after foundational work which first developed the concept of maximum sustainable yield, MSY, [@Gordon1954; @Schaefer1954], which remains a central concept in fisheries management.
+In this model, note that the maximum growth rate occurs at population size of $N = K/2$, which is approached asymptotically from any positive initial state $N_0 \in (0,K)$ under a constant mortality of $F = r/2$, which thus produces a harvest of $rK/4$ at equilibrium -- the maximum yield sustainable indefinitely.
 
@@
   \label{eq:may}
-  f_{\text{May}}(X_t)
-  := L(X_t;\ r, K) - F(X_t, H; \beta,c),
+  f(X_t)
+  = L(X_t;\ r, K) - F(X_t, H; \beta,c),
 \end{align}
@@
 
-Eq. \eqref{eq:may} is an interesting study case of a *tipping point* (saddle-node bifurcation) (see Fig. \ref{fig:may stable}).
+Eq. \eqref{eq:may} is an interesting study case of a *tipping point* (saddle-node bifurcation) (see Fig. \ref{tbd}).
 Holding the value of $\beta$ fixed, for intermediate values of $H$ there exist two stable fixed points for the state $\hat X_t$ of the system, these two attractors separated by an unstable fixed point.
@@
 This also creates the phenomenon of *hysteresis*, where returning $H$ to its original value is not sufficient to restore $\hat X_t$ to the original stable state.
-
-\begin{figure}
-  \centering
-  \includegraphics[width = 0.8\linewidth]{figures/MayStablePoints.pdf}
-  \caption{
-    \label{fig:may stable}
-    The fixed point diagram for the model \cite{may77} associated to $f_{\text{May}}$ as a function of varying the parameter $\beta H$.
-    That is, these are the zeroes of the function $f_{\text{May}}$. 
-    Stable fixed points (also known as *attractors*) are plotted using a solid line, while the unstable fixed point is shown as a dotted line.
-  }
-\end{figure}
 
@@
 \end{align}
-The model is conceptualized in Fig. \ref{fig:conceptual}, panel B.
+The model is conceptualized in Fig. \ref{fig:tbd}.
 It contains three populations, the state at time $t$ is $(X_t, Y_t, Z_t)$. 
@@
 
-We have, up to now, presented eqs. \eqref{eq:may} and \eqref{eq:3d model} as *deterministic* dynamical models for the sake of clarity.
-In practice, we will consider a stochastic contribution to these dynamics.
-This way, instead of eq. \eqref{eq:may}, we consider the following stochastic single-species system
+\textbf{Probably just merge on previous sections.}
+We consider a stochastic contribution to the system's dynamics.
+This way, eq. \eqref{eq:may} becomes
 \begin{align}
@@
 
-Similarly, we use the following dynamical model in place of eq. \eqref{eq:3d model}
+Similarly, eq. \eqref{eq:3d model} becomes
 \begin{align*}
@@
 \end{align*}
-Here, again, $\{\eta_{i,t}\}$ is a set of independent Gaussian variables centered at zero with variances, respectively, $\sigma_X^2$, $\sigma_Y^2$ and $\sigma_Z^2$.
-
-```{r classical_tuning, fig.cap = "TBD.  Red = MSY, Green = 0.8 TAC", out.width = "6.5in"}
-msy_rewards <- msy |>
-  group_by(action, rep) |>
-  filter(t == max(t)) |> 
-  group_by(action) |>
-  summarise(mean_reward = mean(reward),
-            sd = sd(reward))
-
-esc_rewards <- escapement |>
-  group_by(esc_level, rep) |>
-  filter(t == max(t)) |> 
-  group_by(esc_level) |>
-  summarise(mean_reward = mean(reward),
-            sd = sd(reward))
-
-rmsy <- msy_rewards |>
-  ggplot(aes(action, mean_reward, 
-             ymin = mean_reward - 2*sd,
-             ymax=mean_reward+2*sd)) +
-  geom_point() + geom_ribbon(alpha=0.4) +
-  labs(y = "episode reward", x = "constant mortality") +
-  theme(aspect.ratio = 1/1) + 
-  geom_vline(aes(xintercept=0.0465), color="red") + 
-  geom_vline(aes(xintercept=0.0370), color="green")
-
-resc <- esc_rewards |>
-  ggplot(aes(esc_level, mean_reward, 
-             ymin = mean_reward - 2*sd,
-             ymax=mean_reward+2*sd)) +
-  geom_point() + geom_ribbon(alpha=0.4) +
-  labs(y = "episode reward", x = "constant escapement") +
-  theme(aspect.ratio = 1/1)
-
-(rmsy | resc)
-```
 
@@
 Here we review classical strategies for sustainable fishery management and we provide a birds-eye view of the alternative approach we propose.
-A summary of the contrast between the two strategies is given in Fig. \ref{fig:decision approaches}.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=0.8\linewidth]{figures/DecisionApproaches.pdf}
-  \caption{
-  \label{fig:decision approaches}
-  TBD.
-  }
-\end{figure}
+A summary of the contrast between the two strategies is given in Fig. \ref{tbd}.
 
@@
 We collectively refere to these as *classical*, and will compare their performance to RL-based management strategies.
-Classical strategies derive an optimal harvest policy using a simple model for the system dynamics.
-While often complex models of the ecosystem are used to estimate the fish population of interest (i.e. $X_t$), the quota is set only taking into account the estimate $X_t$.
-The function determining this quota is derived from a simple dynamical model for the population (see text box 1).
-
-\textbf{Text box 1 (TBD formatting)}
-Let us consider a ubiquitous simple example: an MSY policy based on a logistic growth model for $X_t$ (see eqs. \eqref{eq:general model} and \eqref{eq:logistic}).
-
-Classical strategies have the common aspect of reducing the complex dynamics of the fishery ecosystem to a single equation governing the harvested population (say, $F$). 
+As shown in Fig.
+\ref{fig:tbd}, classical strategies have the common aspect of reducing the complex dynamics of the fishery ecosystem to a single equation governing the harvested population (say, $F$).
 A common example is using a logistic growth equation,
@@
 
+```{r}
+knitr::opts_chunk$set(cache = TRUE, echo = FALSE, message = FALSE, 
+                      warning = FALSE, fig.height=5, fig.width=7.5)
+msy <- read_csv("../data/msy.csv.gz")
+ppo <- read_csv("../data/PPO250.csv.gz")
+escapement <- read_csv("../data/escapement.csv.gz", show_col_types = FALSE)
+
+# rename some columns for figure purposes:
+colnames(msy) <- c('t', 'rep', 'action', 'reward', 'X', 'Y', 'Z')
+colnames(ppo) <- c('t', 'rep', 'action', 'reward', 'X', 'Y', 'Z') 
+colnames(escapement) <- c('t', 'rep', 'action', 'reward', 'X', 'Y', 'Z')
+
+# ppo dataset uses step rewards rather than cumulative rewards, which is what is useful here.
+# so, switch stuff around:
+#ppo |> 
+#    group_by(rep) |> 
+#    mutate(cumulative_reward = cumsum(reward)) |>
+#    within(rm("reward", "1")) |>
+#    rename(reward = cumulative_reward)
+
+# set up for plots:
+
+# best constant effort based on the data on ../data/msy.csv.gz
+best_action <- 0.0465
+msy_sim <- msy |>
+    filter(action == best_action) |> 
+    group_by(rep) |>
+    pivot_longer(c("X", "Y", "Z"),
+                names_to = "species", values_to = "abundance") |>
+    mutate(abundance = abundance + 1, policy = "msy") # natural units, policy name
+
+actions <- unique(msy$action)
+i <- which.min(abs(actions - best_action* 0.8))
+tac_sim <- msy |> 
+    filter(action == actions[[i]]) |>
+    pivot_longer(c("X", "Y", "Z"),
+        names_to = "species", values_to = "abundance") |>
+    mutate(abundance = abundance + 1, policy = "tac")
+
+ppo_sim <- ppo |> 
+    pivot_longer(c("X", "Y", "Z"),
+        names_to = "species", values_to = "abundance") |>
+    mutate(abundance = abundance + 1, policy = "ppo")
+
+best_e <- 0.62 # optimal constant escapement, computed from ../data/escapement.csv.gz
+esc_sim <- escapement |> 
+    rename(escapement = action) |> # data had a naming problem
+    filter(escapement == best_e) |> 
+    pivot_longer(c("X", "Y", "Z"),
+        names_to = "species", values_to = "abundance") |>
+    mutate(abundance = abundance + 1)
+
+msy_tac_ppo <- bind_rows(msy_sim, tac_sim, ppo_sim)
+```
+
 Our results suggest that DRL can find solutions that are more sustainable *and* more profitable than any constant mortality strategy.
@@
 They lead to a distinctive oscillating behavior of the system state over time, with periods of relative depletion followed by periods of growth. 
-Figure TBD exemplifies this point by comparing one times series of the system state obtained using each of these policies.
+Figure \ref{fig:tbd} exemplifies this point by comparing one times series of the system state obtained using each of these policies.
 
-```{r timeseries, fig.cap="TBD.", out.width = "6.5in"}
+```{r timeseries_fig, fig.cap="TBD.", out.width = "6.5in"}
 policy_labels <- list("msy" = "Opt. Const. Mortality", "tac"= "80% Opt. Const. Mortality", "ppo"= "DRL")
@@
 
-(p1 / p2) + plot_annotation(tag_levels = 'A')
-```
-
-```{r state-space, fig.cap="TBD.", out.width = "6.5in"}
-i_tac <- which.min(abs(actions - best_action * 0.8))
-
-# color pallette limits
-library(RColorBrewer)
-myPalette <- colorRampPalette(rev(brewer.pal(11, "Spectral")))
-sc <- scale_colour_gradientn(colours = myPalette(100), limits=c(0,0.25))
-
-pmsy <- msy |>
-  mutate(X = X+1, Y = Y+1, Z = Z+1, harvest = action * X) |>
-  filter(action == best_action) |>
-  ggplot(aes(X, Z, color = harvest)) +
-  geom_point() + sc +
-  ggtitle("Opt. Const. Mortality")
-
-ptac <- msy |>
-  mutate(X = X+1, Y = Y+1, Z = Z+1, harvest = action * X) |>
-  filter(action == actions[[i_tac]]) |>
-  ggplot(aes(X, Z, color = harvest)) +
-  geom_point() + sc +
-  ggtitle("80% Opt. Const. Mortality")
-
-pppo <- ppo |>
-  mutate(X = X+1, Y = Y+1, Z = Z+1, harvest = action * X) |>
-  ggplot(aes(X, Z, color = harvest)) +
-  geom_point() + sc +
-  ggtitle("DRL")
-
-pesc <- opt_escapement |>
-  mutate(X = X+1, Y = Y+1, Z = Z+1, 
-        harvest = case_when(
-          X - esc_level > 0 ~ X - esc_level,
-          X - esc_level <= 0 ~  0)
-         ) |> 
-  ggplot(aes(X, Z, color = harvest)) +
-  geom_point() + sc +
-  ggtitle("Opt. Const. Escapement")
-
-(
-  (pmsy | ptac) / (pppo | pesc) +
-  plot_layout(guides = "collect") & theme(legend.position = 'right')
-) + 
-  plot_annotation(tag_levels = c('A'))
+(p1 / p2)
 ```
@@
 Our results reproduce this behavior and, furthermore, warn that using a TAC policy instead might not be enough to gain sustainability.
-Figure \ref{rew_len_fig}-A summarizes episode length distribution for MSY, TAC and DRL.
+Figure TBD-A summarizes episode length distribution for MSY, TAC and DRL.
 While the DRL-derived policy does not lead to extinctions, both MSY and TAC do so, and thus their episode mean length is lower than the maximal episode length (in our case set at 200 years).

### 2023-01-24 code/plots.Rmd

@@
 
-# Constant Effort
+# Constant Mortality
 
@@
 esc_sims <- escapement |> 
-  filter(action == best_e$action) |>
+  filter(action == best_e) |>
   pivot_longer(c("X", "Y", "Z"),

### 2023-01-24 code/plots.Rmd
lock in data files for now

--- 2023-01-20 code/plots.Rmd
@@
 
-# Constant Effort
+# Constant Mortality
 
 ```{r}
-msy <- read_csv("../data/msy.csv.gz")
+msy <- read_csv("../data/msy.csv.xz")
 best <- best_reward(msy)
-colnames(msy) <- c('t', 'rep', 'action', 'reward', 'X', 'Y', 'Z')
-msy
-best
+```
+
+```{r}
+# we don't realize all states at each possible action during simulation
+msy |> mutate(X = X+1) |>
+ # distinct(action, X) |>
+  filter(action == best$action) |>
+  mutate(harvest = action * X) |>
+  ggplot(aes(X, harvest, col=action, group=action)) + geom_point()
+
+expand_grid(mortality = seq(0,.1, length.out = 20),
+            state = seq(0,1, length.out=100)) |> 
+  mutate(case = "grid") |> 
+  bind_rows(tibble(mortality = best$action, 
+                   state= seq(0,1, length.out=100),
+                   case = "optimal")) |>
+  mutate(harvest = mortality * state) |>
+  ggplot(aes(state, harvest, col=case, group=mortality)) +
+  geom_line(show.legend = FALSE, lwd=1.5, alpa=0.8) + 
+  scale_color_manual(values=c("grey60", "darkblue"))
 ```
@@
 rewards |>
-  filter(action < 0.1) |>
   ggplot(aes(action, mean_reward, 
@@
 ```{r}
-colnames(msy)
 msy_sim <- msy |>
@@
 msy_sim
+```
 
+```{r}
 msy_sim |>
-    filter(rep < 5) |> # filter(rep > 8) |>
-    # group_by(rep) |>
+    filter(rep < 5) |>
     ggplot(aes(t, abundance,  col=species, group =interaction(rep,species)))+ 
         geom_line() + facet_wrap(~rep)
-
-msy_sim |> 
-  filter(rep < 4, species == "X") |>
-  ggplot(aes(t, abundance, group = interaction(rep,species),  col=species)) + 
-    geom_line(alpha=0.5)
 
@@
 ```{r}
+# determine which action is 80% of MSY
 actions <- unique(msy$action)
 i <- which.min(abs(actions - best$action* 0.8))
+```
+
+```{r}
 tac_sim <- msy |> 
@@
 ```{r}
-ppo <- read_csv("../data/ppo-200.csv")
-colnames(ppo) <- c("1", "t", "X", "Y", "Z", "action", "reward", "rep")
-head(ppo)
+ppo <- read_csv("../data/PPO250.csv.gz")
+```
+```{r}
+ppo_sim <- ppo |> 
+  mutate(escapement = (X+1) - action * (X+1),
+         effort = action) |>
+  pivot_longer(c("X", "Y", "Z"),
+               names_to = "species", values_to = "abundance") |>
+  mutate(abundance = abundance + 1)
+```
+
+```{r}
+ppo_sim |> 
+  group_by(rep) |>
+  filter(t==max(t), species=="X") |>
+  ggplot(aes(rep, reward)) + geom_col()
+
+ppo_sim |> 
+  group_by(rep) |>
+  filter(t==max(t), species=="X") |> 
+  ungroup() |>
+  summarise(mean = mean(reward))
+```
+
+```{r}
+ppo_sim |>
+  filter(rep < 4) |>
+  ggplot(aes(t, abundance, col=species)) +
+  geom_line(alpha=0.6) + facet_wrap(~rep)
+```
+
+```{r}
+ppo |>
+  mutate(X = X+1, Z = Z+1, Z = round(Z, 1)) |>
+  distinct(action, X, Z) |> 
+  mutate(harvest = action * X) |>
+  ggplot(aes(X, harvest, group=Z, color=Z)) +
+  geom_smooth(se=FALSE, method=lm,  formula = y ~ splines::ns(x, 6)) +
+  geom_point(alpha=0.6, shape="+")
+
+```
+
+```{r}
+ppo |>
+  mutate(X = X+1, Z = Z+1, Z = round(Z, 1)) |>
+  distinct(action, X, Z) |> 
+  mutate(harvest = action * X,
+         escapement = X - harvest) |>
+  ggplot(aes(X, escapement, group=Z, color=Z)) +
+  #geom_smooth(se=FALSE, method=lm,  formula = y ~ splines::ns(x, 6)) +
+  geom_point(alpha=0.6, shape="+")
+
 ```
@@
 overall_ppo
-
-ppo_cumulative <- ppo |> group_by(rep) |> mutate(cumulative_reward = cumsum(reward))
-
-ppo_cumulative |> 
-  filter(t==max(t)) |> 
-  ggplot(aes(rep, cumulative_reward)) + geom_col()
-```
-
-```{r}
-ppo_sim <- ppo |> 
-  mutate(escapement = (X+1) - action * (X+1),
-         effort = action) |>
-  pivot_longer(c("X", "Y", "Z"),
-               names_to = "species", values_to = "abundance") |>
-  mutate(abundance = abundance)
-
-ppo_sim
-
-ppo_sim |>
-  filter(rep < 4) |>
-  ggplot(aes(t, abundance, col=species)) +
-  geom_line(alpha=0.6) + facet_wrap(~rep)
-```
-
-```{r}
-ppo_sim |>
-  filter(species == "sp1") |>
-  group_by(t) |>
-  summarise(abundance = mean(abundance), 
-            effort = mean(effort),
-            escapement = mean(escapement)) |> 
-  pivot_longer(-t, names_to = "metric", values_to = "population") |>
-  ggplot(aes(t, population, col=metric)) + geom_line()
 ```
@@
 ```{r}
-escapement <- read_csv("../data/escapement.csv.gz", show_col_types = FALSE)
-colnames(escapement) <- c('t', 'rep', 'action', 'reward', 'X', 'Y', 'Z')
-escapement
-
+escapement <- read_csv("../data/escapement.csv.xz", show_col_types = FALSE)
 ## Determine which level of constant escapement gives highest net reward
@@
 ```
-
-# Millie's suggestion for plots:
-
-```{r}
-msy_sim <- msy_sim |> mutate(type = "Opt const mortality", cumulative_reward = reward)
-tac_sim <- tac_sim |>
-    mutate(abundance = abundance + 1) |>
-    mutate(type = "80% opt const mortality", cumulative_reward = reward)
-ppo_sim <- ppo_sim |> mutate(type = "DRL")
-ppo_sim <- within(ppo_sim, rm("1", "effort", "escapement"))
-ppo_sim <- ppo_sim |>
-    group_by(rep, species) |>
-    mutate(cumulative_reward = cumsum(reward))
-```
-
-```{r}
-ppo_sim
-```
-
-```{r}
-ppo_sim <- within(ppo_sim, rm("reward"))
-msy_sim <- within(msy_sim, rm("reward"))
-tac_sim <- within(tac_sim, rm("reward"))
-all_sim <- bind_rows(msy_sim, tac_sim, ppo_sim)
-ppo_sim
-msy_sim
-tac_sim
-all_sim
-```
-
-```{r}
-all_sim |> #filter(rep < 10, species == "X") |> 
-    group_by(rep) |>
-    filter(t==max(t), rep < 10, species == "X") |>
-    ggplot(aes(rep, cumulative_reward)) + geom_col() +
-    facet_wrap(~type)
-
-msy_sim |> 
-  group_by(rep) |>
-  filter(t==max(t), species=="X", rep < 10) |>
-  ggplot(aes(rep, cumulative_reward)) + geom_col()
-
-tac_sim |> 
-  group_by(rep) |>
-  filter(t==max(t), species=="X", rep < 10) |>
-  ggplot(aes(rep, cumulative_reward)) + geom_col()
-
-all_sim |> 
-    mutate(across(type, factor, levels = c("Opt const mortality", "80% opt const mortality", 
-                                           "DRL"))) |>
-    group_by(rep, type) |>
-    filter(t==max(t), species=="X", rep < 10) |>
-    ggplot(aes(rep, cumulative_reward)) + 
-        geom_col()  + 
-        facet_wrap(~type) + 
-        labs(y = "reward", x = "repetition")
-
-```
-
-```{r}
-all_sim |> 
-    mutate(across(type, factor, levels = c("Opt const mortality", "80% opt const mortality", 
-                                           "DRL"))) |>
-    group_by(rep, type) |>
-    filter(rep == 0) |>
-    ggplot(aes(t, abundance, col = species)) + 
-        geom_line()  + 
-        facet_wrap(~type) + 
-        labs(y = "reward", x = "repetition")
-all_sim |>
-    group_by(rep,type) |>
-    filter(rep<10) |>
-    ggplot(aes(t, action)) + 
-    geom_line()  + 
-    facet_wrap(~type) + 
-    labs(y = "action", x = "repetition")
-
-#all_sim |> group_by(rep) |> filter(t==max(t), species == "X") |> group_by(t)
-```
-
-```{r}
-all_sim |> 
-    mutate(across(type, factor, levels = c("Opt const mortality", "80% opt const mortality", 
-                                           "DRL"))) |>
-    group_by(rep, type) |>
-    filter(t==max(t), species=="X", rep < 10) |>
-    ggplot(aes(rep,t)) + 
-        geom_col()  + 
-        facet_wrap(~type) + 
-        labs(y = "episode length", x = "repetition") +
-        theme(axis.text.x = element_blank())
-```

### 2023-01-24 code/escapement.py

@@ -27,6 +27,6 @@ actions = np.linspace(0,1,101)
 parallel = [simulate.remote(env, i) for i in actions]
 df = ray.get(parallel)
 
-cols = ["t", "rep", "action", "reward", "sp1", "sp2", "sp3"]
+cols = ["t", "rep", "action", "reward", "X", "Y", "Z"]
 df2 = pd.DataFrame(np.vstack(df), columns = cols)
-df2.to_csv("data/escapement.csv.gz", index=False)
+df2.to_csv("data/escapement.csv.xz", index=False)

### 2023-01-24 code/msy.py

@@ -20,16 +20,16 @@ def simulate(env, action):
 
 
 env = gym.make("threeFishing-v2")
-actions = np.linspace(0,0.1,201)
+actions = np.linspace(0,0.1,101)
 
 # define parllel loop and execute
 parallel = [simulate.remote(env, i) for i in actions]
 df = ray.get(parallel)
 
 # convert to data.frame & write to csv
-cols = ["t", "rep", "action", "reward", "sp1", "sp2", "sp3"]
+cols = ["t", "rep", "action", "reward", "X", "Y", "Z"]
 df2 = pd.DataFrame(np.vstack(df), columns = cols)
-df2.to_csv("data/msy.csv.gz", index=False)
+df2.to_csv("data/msy.csv.xz", index=False)
 
 
 

### 2023-01-24 code/train-agent.py

@@ -4,13 +4,14 @@ from ray.tune import register_env
 import os
 import pandas as pd
 import numpy as np
+import torch
 
 ## We could call env directly without this if only  our envs took a env_config dict argument
 register_env("threeFishing-v2", lambda config: forageVVHcont())
 
 config = ppo.PPOConfig()
 config.num_envs_per_worker=20
-config = config.resources(num_gpus=1)
+config = config.resources(num_gpus=torch.cuda.device_count())
 config.framework_str="torch"
 config.create_env_on_local_worker = True
 agent = config.build(env="threeFishing-v2")
@@ -31,10 +32,12 @@ agent.restore(checkpoint)
 stats = agent.evaluate() # built-in method to evaluate agent on eval env
 
 # Initialize saved copy of eval environment:
-env = agent.env_creator(agent.evaluation_config.env_config)
+config = agent.evaluation_config.env_config
+config.update({'seed': 42})
+env = agent.env_creator(config)
 
 df = []
-for rep in range(30):
+for rep in range(50):
   episode_reward = 0
   observation = env.reset()
   for t in range(env.Tmax):
@@ -45,16 +48,14 @@ for rep in range(30):
     if terminated:
       break
     
-cols = ["t", "rep", "action", "reward", "sp1", "sp2", "sp3"]
+cols = ["t", "rep", "action", "reward", "X", "Y", "Z"]
 df = pd.DataFrame(df, columns = cols)
 df.to_csv("data/PPO" + str(iterations) + ".csv.gz", index = False)
 
 
 
-
 ## Plots ## 
 from plotnine import ggplot, geom_point, aes, geom_line, facet_wrap, geom_path
-
 ## Timeseries
 df2 = (df
        .melt(id_vars=["t", "action", "reward", "rep"])
@@ -64,42 +65,5 @@ df2 = (df
              'action': 'mean'})) 
 (ggplot(df2, aes("t", "value", color="variable")) +
  geom_line())
-
-# overall average state of each sp across reps and time
-aves = (df
-       .melt(id_vars=["t", "action", "reward", "rep"])
-       .groupby(["variable"], as_index=False)
-       .agg({'reward': 'mean',
-             'value': 'mean',
-             'action': 'mean'})) 
-init = aves.value.values
-
-## agent action vs H
-dd = []
-for p in np.linspace(-1, 0, 101):
[…]

### 2023-01-24 manuscript/manuscript.Rmd
lock in data files for now
started new results

--- 2023-01-24 manuscript/manuscript.Rmd
@@
 
-# Methods and Results
+# Results
+
+\textbf{Training a DRL agent.}
+We trained a DRL agent parametrized by a neural network $\theta$ with two hidden, 64-neuron, layers.
+The state space used was a cube $[0,2]^{\times 3}$, while the policies explored mapped states $N_t$ to a single number, $h_t = \pi_\theta(N_t)$, the harvest quota on species $X$ at that timestep.
+Each episode in the training and in the evaluation stages was at most 600 time steps long.
+Episodes were ended early in case of a ‘‘near-extinction event,’’ a moment in which any of the populations dipped below a threshold value, i.e.,
+\begin{align}
+  \label{eq:thresh}
+  X_t \leq X_{\text{thresh.}},\quad
+  Y_t \leq Y_{\text{thresh.}},\quad
+  \text{or,}\quad
+  Z \leq Z_{\text{thresh.}}.
+\end{align}
+The reward function had two componenets:
+First, at each time step the harvest $h_t$ was received as a reward.
+Second, if the episode ended early (say, at time-step $t$) because of a near-extinction event, a negative reward of $-50/t$ was awarded.
+The constant 50 was chosen so as to select against highly extractive policies in the early stages of training, and thus guide the agent towards sustainable policies.
+
+We used the Ray framework [@ray_package] for training, in particular we used a Ray PPOConfig object to build the algorithm using the default values for all hyper-parameters (cf.
+[@repo] for the code used).
+The agent was trained for 250 iterations of the PPO optimization step.
+
+\textbf{Tuning the constant mortality strategy.}
+A grid of 200 mortality rates was laid in the interval $[0, 0.1]$.
+For each one of these rates, say $F$, we simulated 50 episodes based on \eqref{eq:3d model stoch}:
+At each time step the state $(X_t, Y_t, Z_t)$ was observed, a harvest of $FX_t$ was collected and then the system evolved according to its natural dynamics \eqref{eq:3d model stoch}.
+The optimal mortality rate was the value $F^*$ for which the mean episode reward was maximal.
+
+\textbf{Parameter values used.}
+
+# Old Methods and Results
 

### 2023-01-24 manuscript/manuscript.Rmd
added a TAC sentence

@@
 Similarly, we use the following dynamical model in place of eq. \eqref{eq:3d model}
-\begin{align*}
+\begin{align}
   \label{eq:3d model stoch}
@@
 \end{split}
-\end{align*}
+\end{align}
 Here, again, $\{\eta_{i,t}\}$ is a set of independent Gaussian variables centered at zero with variances, respectively, $\sigma_X^2$, $\sigma_Y^2$ and $\sigma_Z^2$.
@@
 The optimal mortality rate was the value $F^*$ for which the mean episode reward was maximal.
+We additionally evaluated a more conservative strategy where a mortality of $0.8 F^*$ instead is used.
 
 \textbf{Parameter values used.}
+The model's dynamic parameters were chosen as follows
+\begin{align*}
+  r_X = K_X = r_Y = K_Y = 1, \quad \beta = 0.3, \quad c = 0.1\\
+  c_{XY} = 0.5,\quad f = 0.15,\quad D = 1.1, \quad d_Z = 0.135.
+\end{align*}
+Moreover, the variances for the stochastic terms were chosen as
+$$
+  \sigma^2(\eta_{X,t}) = \sigma^2(\eta_{Y,t}) = \sigma^2(\eta_{Z,t}) = 0.05.
+$$
+The threshold values used to determine an early episode end were chosen as
+$$
+  X_{\text{thresh.}} = 0.08,\quad Y_{\text{thresh.}} = Z_{\text{thresh.}} = 10^{-4},
+$$
+where the higher value of $X_{\text{thresh.}}$ helped select for more sustainable fishing strategies (for both the DRL and constant mortality approaches).
 

### 2023-01-25 manuscript/manuscript.Rmd
added results, moved graphs around, added line graphs to the const mortality tuning figure

@@
 
-# Results
+# Methods
 
@@
 
+The result of the tuning procedure is shown in Fig. \ref{fig:linear_tuning}. 
+There, we see that the optimal value $F^*$ lies in an area with high episode reward variance.
+We will see in Fig. \ref{fig:rew_len_fig} that a likely candidate for this variance is the presence of episodes with early endings.
+Thus, the value $0.8 F^*$ supports a smaller variance, and correspondingly -- as Fig. \ref{fig:rew_len_fig} shows -- this policy leads to a much smaller probability of a near-extinction event.
+
+```{r linear_tuning, fig.cap = "Result of the tuning procedure for the constant mortality strategy.
+(A) Mean episode reward (averaged over 50 episodes) shown as scatter data, $2\times$ standard deviation around each scatter point shown as a grey shade.
+Vertical lines correspond to the optimal constant mortality rate $F^* = 0.465$ (red) and $0.8 F^*=0.0372$ (green).
+(B) Several constant mortality policies shown for illustrative purposes.
+Slopes $0.02$, $0.03$, $0.055$ and $0.065$ shown in black; the optimal linear policy $F^*X$ shown in red, while $0.8F^*X$ shown in green.", out.width = "6.5in"}
+msy_rewards <- msy |>
+  group_by(action, rep) |>
+  filter(t == max(t)) |> 
+  group_by(action) |>
+  summarise(mean_reward = mean(reward),
+            sd = sd(reward))
+
+rmsy <- msy_rewards |>
+  ggplot(aes(action, mean_reward, 
+             ymin = mean_reward - 2*sd,
+             ymax=mean_reward+2*sd)) +
+  geom_point() + geom_ribbon(alpha=0.4) +
+  labs(y = "episode reward", x = "constant mortality") +
+  theme(aspect.ratio = 1/1) + 
+  geom_vline(aes(xintercept=0.0465), color="red") + 
+  geom_vline(aes(xintercept=0.0372), color="green") +
+  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
+
+# linear function blueprint
+linear_f <- function(slope, x) slope*x
+# generates linear function with given slope
+linear_f_gen <- function(slope) (function(x) slope*x)
+
+# dummy data set to keep ggplot happy
+p <- ggplot(data = data.frame(x = 0), mapping = aes(x = x))
+
+# p + stat_function(fun = linear_f_gen(0.5)) + xlim(0,2)
+lines <- p + stat_function(fun = linear_f_gen(0.01)) + 
+  stat_function(fun = linear_f_gen(0.02)) +
+  stat_function(fun = linear_f_gen(0.03)) +
+  stat_function(fun = linear_f_gen(0.8*best_action), col = "green") +
+  stat_function(fun = linear_f_gen(best_action), col = "red") +
+  stat_function(fun = linear_f_gen(0.055)) + 
+  stat_function(fun = linear_f_gen(0.065)) +
+  xlim(0,2) +
+  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
+
+(rmsy | lines) + plot_annotation(tag_levels = c('A'))
+```
+
 \textbf{Parameter values used.}
@@
 where the higher value of $X_{\text{thresh.}}$ helped select for more sustainable fishing strategies (for both the DRL and constant mortality approaches).
+
+# Results
+
+In Fig. \ref{fig:timeseries} we show one example time series for each of the three controls we analyzed: 
+the left panel uses an optimalconstant mortality control $F^*X$, the middle panel uses $0.8 F^* X$, while the right panel is controlled by the trained DRL agent.
+We see that that optimal constant mortality control leads to a near-extinction event.
+The constant mortality policy at $0.8 F^*$ is already conservative enough to support a sustainable path.
+There, the population levels are maintained roughly in a stationary state.
+Finally, the DRL time series shows an oscillatory behavior, with a slight dephase between the $Y$ and $Z$ curves, while the $X$ oscillations are roughly anticorrelated with $Z$.
+
+```{r timeseries, fig.cap="TBD.", out.width = "6.5in"}
+policy_labels <- list("msy" = "Opt. Const. Mortality", "tac"= "80% Opt. Const. Mortality", "ppo"= "DRL")
+policy_labeller <- function(variable,value){
+  return(policy_labels[value])
+}
+msy_tac_ppo |> 
+    group_by(rep, policy) |>
+    filter(rep == 0) |>
+    ggplot(aes(t, abundance, col = species)) + 
+        geom_line()  + 
+        scale_color_manual(values=c('palegreen3', 'coral1', 'deepskyblue4')) +
+        facet_wrap(~factor(policy, levels = c("msy", "tac", "ppo")), labeller = policy_labeller ) +
+        labs(y = "abundance", x = "time") +
+        theme(aspect.ratio = 2/3)
+```
 
@@
 
-```{r timeseries, fig.cap="TBD.", out.width = "6.5in"}
+```{r timeseries_old, fig.cap="TBD.", out.width = "6.5in"}
 policy_labels <- list("msy" = "Opt. Const. Mortality", "tac"= "80% Opt. Const. Mortality", "ppo"= "DRL")

### 2023-01-25 manuscript/manuscript.Rmd
graph mini changes

@@
 ```{r knit_global, include=FALSE}
-knitr::opts_chunk$set(cache = TRUE, echo = FALSE, message = FALSE, 
+knitr::opts_chunk$set(cache = FALSE, echo = FALSE, message = FALSE, 
                       warning = FALSE, fig.height=5, fig.width=7.5)
@@
 escapement <- read_csv("../data/escapement.csv.gz", show_col_types = FALSE)
+
+ppo |> filter(t==max(t))
 
@@
 # best constant effort based on the data on ../data/msy.csv.gz
-best_action <- 0.0465
+best_action <- 0.036
 msy_sim <- msy |>
@@
 
-best_e <- 0.62 # optimal constant escapement, computed from ../data/escapement.csv.gz
+best_e <- 0.61 # optimal constant escapement, computed from ../data/escapement.csv.gz
 esc_sim <- escapement |> 
@@
   theme(aspect.ratio = 1/1) + 
-  geom_vline(aes(xintercept=0.0465), color="red") + 
-  geom_vline(aes(xintercept=0.0370), color="green")
+  geom_vline(aes(xintercept=0.0360), color="red") + 
+  geom_vline(aes(xintercept=0.0288), color="green")
 
@@
 (A) Mean episode reward (averaged over 50 episodes) shown as scatter data, $2\times$ standard deviation around each scatter point shown as a grey shade.
-Vertical lines correspond to the optimal constant mortality rate $F^* = 0.465$ (red) and $0.8 F^*=0.0372$ (green).
+Vertical lines correspond to the optimal constant mortality rate $F^* = 0.036$ (red) and $0.8 F^*=0.0288$ (green).
 (B) Several constant mortality policies shown for illustrative purposes.
@@
   theme(aspect.ratio = 1/1) + 
-  geom_vline(aes(xintercept=0.0465), color="red") + 
-  geom_vline(aes(xintercept=0.0372), color="green") +
+  geom_vline(aes(xintercept=best_action), color="red") + 
+  geom_vline(aes(xintercept=0.8*best_action), color="green") +
   theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
@@
 the left panel uses an optimalconstant mortality control $F^*X$, the middle panel uses $0.8 F^* X$, while the right panel is controlled by the trained DRL agent.
-We see that that optimal constant mortality control leads to a near-extinction event.
+We see that that optimal constant mortality control leads to a near-extinction event at around $t=100$.
 The constant mortality policy at $0.8 F^*$ is already conservative enough to support a sustainable path.
@@
     group_by(rep, policy) |>
-    filter(rep == 0) |>
+    filter(rep == 1) |>
     ggplot(aes(t, abundance, col = species)) + 
@@
     group_by(rep, policy) |>
-    filter(t==max(t), rep < 30) |>
+    filter(t==max(t), species == "X", rep < 30) |>
     ggplot(aes(rep, t)) + 
@@
     group_by(rep, policy) |>
-    filter(t==max(t), rep < 30) |>
+    filter(t==max(t), species == "X", rep < 30) |>
     ggplot(aes(rep, reward)) + 

### 2023-01-25 code/train-agent.py
moved checkpoint folder to include its tmax (will now save a checkpoint at tmax = 200)

@@ -21,7 +21,8 @@ checkpoint = ("cache/checkpoint_000{}".format(iterations))
 
 if not os.path.exists(checkpoint): # train only if no trained agent saved
   for _ in range(iterations):
-      agent.train()
+    print(f"iteration {_}", end = "\r")
+    agent.train()
   checkpoint = agent.save("cache")
 
 #agent_restored = config.build(env="threeFishing-v2")

### 2023-01-25 code/msy.py
env.training = false on the msy and escapement scripts

@@ -20,6 +20,7 @@ def simulate(env, action):
 
 
 env = gym.make("threeFishing-v2")
+env.training = False
 actions = np.linspace(0,0.1,201)
 
 # define parllel loop and execute

### 2023-01-25 manuscript/manuscript.Rmd
still tinkering
mini

@@
 # best constant effort based on the data on ../data/msy.csv.gz
-best_action <- 0.036
+best_action <- 0.045
 msy_sim <- msy |>

### 2023-01-25 code/train-agent.py
200 iterations training

@@ -16,7 +16,7 @@ config.create_env_on_local_worker = True
 agent = config.build(env="threeFishing-v2")
 
 
-iterations = 250
+iterations = 200
 checkpoint = ("cache/checkpoint_000{}".format(iterations))
 
 if not os.path.exists(checkpoint): # train only if no trained agent saved

### 2023-01-25 manuscript/manuscript.Rmd
focus on rep == 2 for time series

@@
 msy <- read_csv("../data/msy.csv.gz")
-ppo <- read_csv("../data/PPO250.csv.gz")
+ppo <- read_csv("../data/PPO200.csv.gz")
 escapement <- read_csv("../data/escapement.csv.gz", show_col_types = FALSE)
@@
     group_by(rep, policy) |>
-    filter(rep == 1) |>
+    filter(rep == 2) |>
     ggplot(aes(t, abundance, col = species)) + 

### 2023-01-25 manuscript/manuscript.Rmd
mini changes

@@
 escapement <- read_csv("../data/escapement.csv.gz", show_col_types = FALSE)
-
-ppo |> filter(t==max(t))
 
@@
 
-```{r classical_tuning, fig.cap = "TBD.  Red = MSY, Green = 0.8 TAC", out.width = "6.5in"}
+```{r}
 msy_rewards <- msy |>
@@
 
-(rmsy | resc)
+# (rmsy | resc)
 ```

### 2023-01-25 manuscript/manuscript.Rmd
mini

@@
 ```{r linear_tuning, fig.cap = "Result of the tuning procedure for the constant mortality strategy.
-(A) Mean episode reward (averaged over 50 episodes) shown as scatter data, $2\times$ standard deviation around each scatter point shown as a grey shade.
+(A) Mean episode reward (averaged over 50 episodes) shown as scatter data, and shaded in gray are two standard deviations around each scatter point.
 Vertical lines correspond to the optimal constant mortality rate $F^* = 0.036$ (red) and $0.8 F^*=0.0288$ (green).
@@
 the left panel uses an optimalconstant mortality control $F^*X$, the middle panel uses $0.8 F^* X$, while the right panel is controlled by the trained DRL agent.
-We see that that optimal constant mortality control leads to a near-extinction event at around $t=100$.
+We see that that optimal constant mortality control leads to a near-extinction event at around $t=150$.
 The constant mortality policy at $0.8 F^*$ is already conservative enough to support a sustainable path.
@@
     group_by(rep, policy) |>
-    filter(rep == 2) |>
+    filter(rep == 5) |>
     ggplot(aes(t, abundance, col = species)) + 

### 2023-01-25 manuscript/manuscript.Rmd
mini
discussed the rewards/ep len figure

@@
 (A) Mean episode reward (averaged over 50 episodes) shown as scatter data, and shaded in gray are two standard deviations around each scatter point.
-Vertical lines correspond to the optimal constant mortality rate $F^* = 0.036$ (red) and $0.8 F^*=0.0288$ (green).
+Vertical lines correspond to the optimal constant mortality rate $F^* = $ TBD (red) and $0.8 F^*=$ TBD (green).
 (B) Several constant mortality policies shown for illustrative purposes.
@@
 ```
+
+This time series is representative of the patterns observed in the data (see App. \ref{app:tbd}).
+The episode reward and episode length statistics are shown in Fig. \ref{fig:rew_len_fig}.
+There, we highlight two things:
+First, that the tendency towards early episode ends is fixed by the conservative approach of fishing at 80% of the optimal mortality rate.
+In App. \ref{app:tbd} we discuss a warning about this approach, however: the arbitrarily chosen percentange itself lives on a ‘‘knife's edge.’’
+More concretely, choosing a higher percentage, say 90%, can switch the behaviour of the controlled system and give rise to the proliferation of early episode ends.
+Second, that while the conservative 80% approach has a stable behavior, it consistently produces smaller harvests than a DRL generated policy.
 

### 2023-01-25 manuscript/manuscript.Rmd
reward-> harvest in plot

@@
       facet_wrap(~factor(policy, levels = c("msy", "tac", "ppo")), labeller = policy_labeller ) +
-      labs(y = "episode reward", x = "repetition") +
+      labs(y = "episode harvest", x = "repetition") +
       theme(axis.text.x = element_blank())

### 2023-01-25 manuscript/manuscript.Rmd

@@
 For example, in the logistic equation shown, the maximal sustainable yield of the system is attained at K/2.
-(B) RL allows us to solve for quantitative decision strategies considering more complex ecosystem dynamics.
+#| (B) RL allows us to solve for quantitative decision strategies considering more complex ecosystem dynamics.
 For example, in this paper, we find harvest strategies for a three-species model using a model-free RL algorithm, PPO.
-(C) We show how the RL-based solution that considers all three species finds harvest policies with higher average reward and with (D) lower probability of harvest species collapse than the classical approaches, only considering the dynamics of V1.
-  }
-\end{figure}
+#| (C) We show how the RL-based solution that considers all three species finds harvest policies with higher average reward and with (D) lower probability of harvest species collapse than the classical approaches, only considering the dynamics of V1.',
+#| out.width: '1\\linewidth', 
+#| fig.pos: 'H'
 
-<!--
-```{r conceptual, echo = FALSE, message=FALSE, fig.align='center', fig.cap='(A) Classical management strategies often have the common aspect of reducing the complex dynamics of the fishery ecosystem to a single equation governing the harvested population (say, $V1$).
-where the interaction between $V1$ and its environment is summarized to a set of parameters, such as the growth rate $r$, and the carrying capacity $K$.
-An advantage of one-dimensional approaches is that the optimal policy is often known exactly, and, moreover, is intuitive.
-For example, in the logistic equation shown, the maximal sustainable yield of the system is attained at K/2.
-(B) RL allows us to solve for quantitative decision strategies considering more complex ecosystem dynamics.
-For example, in this paper, we find harvest strategies for a three-species model using a model-free RL algorithm, PPO.
-(C) We show how the RL-based solution that considers all three species finds harvest policies with higher average reward and with (D) lower probability of harvest species collapse than the classical approaches, only considering the dynamics of V1.', out.width='1\\linewidth', fig.pos='H'}
-knitr::include_graphics("Figures/conceptual-figure-old.png")
+#knitr::include_graphics("Figures/conceptual-figure.png")
 ```
--->
 

### 2023-01-25 code/train-agent.py
env.training = False everywhere

@@ -37,6 +37,7 @@ config = agent.evaluation_config.env_config
 config.update({'seed': 42})
 env = agent.env_creator(config)
 
+env.training = False
 df = []
 for rep in range(50):
   episode_reward = 0

### 2023-01-25 manuscript/manuscript.Rmd
datafreeze: modified code/*.py files to create data in a standardized form. created final data files data/*.xz, removed previous data/*.gz files (still probably need to remove them from history with git filter-branch (TBD). Modified code blocks within manuscript.rmd to fit the new data form, things are stable here.

@@
 ```{r}
+# for best const mortality and best const escapement
+msy_optimizer <- function(df) {
+  df |> 
+  group_by(action, rep) |>
+  filter(t == max(t)) |> 
+  group_by(action) |>
+  summarise(mean_reward = mean(reward), sd = sd(reward)) |> 
+  filter(mean_reward == max(mean_reward))
+}
+escapement_optimizer <- function(df) {
+  df |> 
+  group_by(escapement, rep) |>
+  filter(t == max(t)) |> 
+  group_by(escapement) |>
+  summarise(mean_reward = mean(reward), sd = sd(reward)) |> 
+  filter(mean_reward == max(mean_reward))
+}
+
 # Collect data (outsource to separate R script later TBD)
-msy <- read_csv("../data/msy.csv.gz")
-ppo <- read_csv("../data/PPO200.csv.gz")
-escapement <- read_csv("../data/escapement.csv.gz", show_col_types = FALSE)
+msy_df <- read_csv("../data/msy.csv.xz")
+best_action <- msy_optimizer(msy_df)$action
+escapement_df <- read_csv("../data/escapement.csv.xz", show_col_types = FALSE)
+best_e <- escapement_optimizer(escapement_df)$escapement
+ppo_df <- read_csv("../data/PPO200.csv.xz")
 
-# rename some columns for figure purposes:
-colnames(msy) <- c('t', 'rep', 'action', 'reward', 'X', 'Y', 'Z')
-colnames(ppo) <- c('t', 'rep', 'action', 'reward', 'X', 'Y', 'Z') 
-colnames(escapement) <- c('t', 'rep', 'esc_level', 'reward', 'X', 'Y', 'Z') 
-opt_escapement <- escapement |> filter(esc_level == 0.62)
-
-# ppo dataset uses step rewards rather than cumulative rewards, which is what is useful here.
-# so, switch stuff around:
-#ppo |> 
-#    group_by(rep) |> 
-#    mutate(cumulative_reward = cumsum(reward)) |>
-#    within(rm("reward", "1")) |>
-#    rename(reward = cumulative_reward)
+opt_escapement <- escapement_df |> filter(escapement == best_e)
 
@@
 
-# best constant effort based on the data on ../data/msy.csv.gz
-best_action <- 0.045
-msy_sim <- msy |>
+msy_sim <- msy_df |>
     filter(action == best_action) |> 
@@
 i <- which.min(abs(actions - best_action* 0.8))
-tac_sim <- msy |> 
+tac_sim <- msy_df |> 
     filter(action == actions[[i]]) |>
@@
 
-ppo_sim <- ppo |> 
+ppo_sim <- ppo_df |> 
     pivot_longer(c("X", "Y", "Z"),
@@
 best_e <- 0.61 # optimal constant escapement, computed from ../data/escapement.csv.gz
-esc_sim <- escapement |> 
+esc_sim_df <- escapement |> 
     pivot_longer(c("X", "Y", "Z"),
@@
 ```{r}
-msy_rewards <- msy |>
+msy_rewards <- msy_df |>
   group_by(action, rep) |>
@@
 
-esc_rewards <- escapement |>
-  group_by(esc_level, rep) |>
+esc_rewards <- escapement_df |>
+  group_by(escapement, rep) |>
   filter(t == max(t)) |> 
-  group_by(esc_level) |>
+  group_by(escapement) |>
   summarise(mean_reward = mean(reward),
@@
 ```{r timeseries, fig.cap="TBD.", out.width = "6.5in"}
-policy_labels <- list("msy" = "Opt. Const. Mortality", "tac"= "80% Opt. Const. Mortality", "ppo"= "DRL")
+policy_labels <- list("msy" = "MSY Const. Mortality", "tac"= "80% MSY Const. Mortality", "ppo"= "DRL")
 policy_labeller <- function(variable,value){
@@
 ```{r timeseries_old, fig.cap="TBD.", out.width = "6.5in"}
-policy_labels <- list("msy" = "Opt. Const. Mortality", "tac"= "80% Opt. Const. Mortality", "ppo"= "DRL")
 policy_labeller <- function(variable,value){
@@
         harvest = case_when(
-          X - esc_level > 0 ~ X - esc_level,
-          X - esc_level <= 0 ~  0)
+          X - escapement > 0 ~ X - escapement,
+          X - escapement <= 0 ~  0)
          ) |> 
@@
 
-ppo_mod <- ppo |> 
+ppo_mod <- ppo_df |> 
   mutate(X = X+1, Y = Y+1, Z = Z+1, escapement = (1-action)*X, time = t)

### 2023-01-25 code/escapement.py

@@ -5,16 +5,15 @@ import numpy as np
 import ray
 
 @ray.remote
-def simulate(env, action):
+def simulate(env, esc_level):
   df = []
   for rep in range(50):
     episode_reward = 0
     observation = env.reset()
     for t in range(env.Tmax):
-      df.append(np.append([t, rep, action, episode_reward], observation))
+      df.append(np.append([t, rep, esc_level, episode_reward], observation))
       sp1_pop = (observation[0] + 1 ) # natural state-space
-      effort = np.max([1 - action / sp1_pop, 0])
-      observation, reward, terminated, info = env.step(effort)
+      action = np.max([1 - esc_level / sp1_pop, 0]) # action is a mortality rate. 
       episode_reward += reward
       if terminated:
         break
@@ -22,12 +21,12 @@ def simulate(env, action):
 
 env = gym.make("threeFishing-v2")
 env.training = False
-actions = np.linspace(0,1,101)
+esc_choices = np.linspace(0,1,101)
 
 # define parllel loop and execute
-parallel = [simulate.remote(env, i) for i in actions]
+parallel = [simulate.remote(env, i) for i in esc_choices]
 df = ray.get(parallel)
 
-cols = ["t", "rep", "action", "reward", "X", "Y", "Z"]
+cols = ["t", "rep", "escapement", "reward", "X", "Y", "Z"]
 df2 = pd.DataFrame(np.vstack(df), columns = cols)
 df2.to_csv("data/escapement.csv.xz", index=False)

### 2023-01-25 code/train-agent.py

@@ -52,7 +52,7 @@ for rep in range(50):
     
 cols = ["t", "rep", "action", "reward", "X", "Y", "Z"]
 df = pd.DataFrame(df, columns = cols)
-df.to_csv("data/PPO" + str(iterations) + ".csv.gz", index = False)
+df.to_csv(f"data/PPO{iterations}.csv.xz", index = False)
 
 
 

### 2023-01-25 manuscript/manuscript.Rmd
things are stable now

@@
 
-actions <- unique(msy$action)
+actions <- unique(msy_df$action)
 i <- which.min(abs(actions - best_action* 0.8))
@@
 best_e <- 0.61 # optimal constant escapement, computed from ../data/escapement.csv.gz
-esc_sim_df <- escapement |> 
+esc_sim_df <- escapement_df |> 
     pivot_longer(c("X", "Y", "Z"),
@@
 Slopes $0.02$, $0.03$, $0.055$ and $0.065$ shown in black; the optimal linear policy $F^*X$ shown in red, while $0.8F^*X$ shown in green.", out.width = "6.5in"}
-msy_rewards <- msy |>
+msy_rewards <- msy_df |>
   group_by(action, rep) |>
@@
     group_by(rep, policy) |>
-    filter(rep == 5) |>
+    filter(rep == 6) |>
     ggplot(aes(t, abundance, col = species)) + 
@@
 
-pmsy <- msy |>
+pmsy <- msy_df |>
   mutate(X = X+1, Y = Y+1, Z = Z+1, harvest = action * X) |>
@@
 
-ptac <- msy |>
+ptac <- msy_df |>
   mutate(X = X+1, Y = Y+1, Z = Z+1, harvest = action * X) |>
@@
 
-pppo <- ppo |>
+pppo <- ppo_df |>
   mutate(X = X+1, Y = Y+1, Z = Z+1, harvest = action * X) |>

### 2023-01-25 manuscript/manuscript.Rmd
added all 50 data points to rew len graph

@@
     group_by(rep, policy) |>
-    filter(rep == 6) |>
+    filter(rep == 49) |>
     ggplot(aes(t, abundance, col = species)) + 
@@
     group_by(rep, policy) |>
-    filter(t==max(t), species == "X", rep < 30) |>
+    filter(t==max(t), species == "X", rep < 50) |>
     ggplot(aes(rep, t)) + 
@@
     group_by(rep, policy) |>
-    filter(t==max(t), species == "X", rep < 30) |>
+    filter(t==max(t), species == "X", rep < 50) |>
     ggplot(aes(rep, reward)) + 

### 2023-01-25 code/train-agent.py
used new seed for RL data file (previous data file had a low probability event)

@@ -34,7 +34,7 @@ stats = agent.evaluate() # built-in method to evaluate agent on eval env
 
 # Initialize saved copy of eval environment:
 config = agent.evaluation_config.env_config
-config.update({'seed': 42})
+config.update({'seed': 41})
 env = agent.env_creator(config)
 
 env.training = False

### 2023-01-25 manuscript/manuscript.Rmd
included #| version of fig:linear_tuning, some minor plot changes

@@
 
-```{r linear_tuning, fig.cap = "Result of the tuning procedure for the constant mortality strategy.
-(A) Mean episode reward (averaged over 50 episodes) shown as scatter data, and shaded in gray are two standard deviations around each scatter point.
+```{r}
+#| label: linear_tuning, 
+#| fig.cap : "Result of the tuning procedure for the constant mortality strategy. 
+#| (A) Mean episode reward (averaged over 50 episodes) shown as scatter data, and shaded in gray are two standard deviations around each scatter point.
 Vertical lines correspond to the optimal constant mortality rate $F^* = $ TBD (red) and $0.8 F^*=$ TBD (green).
-(B) Several constant mortality policies shown for illustrative purposes.
-Slopes $0.02$, $0.03$, $0.055$ and $0.065$ shown in black; the optimal linear policy $F^*X$ shown in red, while $0.8F^*X$ shown in green.", out.width = "6.5in"}
+#| (B) Several constant mortality policies shown for illustrative purposes.
+Slopes $0.02$, $0.03$, $0.055$ and $0.065$ shown in black; the optimal linear policy $F^*X$ shown in red, while $0.8F^*X$ shown in green."
+#| out.width : "6.5in"
 msy_rewards <- msy_df |>
@@
     group_by(rep, policy) |>
-    filter(rep == 49) |>
+    filter(rep == 6) |>
     ggplot(aes(t, abundance, col = species)) + 
@@
 ppo_esc <- ppo_mod |>
-  filter(rep==0) |>
+  filter(rep == 6) |>
   group_by(rep) |>
@@
   geom_line() +
+  ylim(0,1) +
   geom_hline(aes(yintercept=avg_esc), color="red")+

### 2023-01-26 manuscript/manuscript.Rmd
another fig caption

@@
 ```{r knit_global, include=FALSE}
-knitr::opts_chunk$set(cache = FALSE, echo = FALSE, message = FALSE, 
+knitr::opts_chunk$set(cache = TRUE, echo = FALSE, message = FALSE, 
                       warning = FALSE, fig.height=5, fig.width=7.5)
@@
 
-```{r timeseries, fig.cap="TBD.", out.width = "6.5in"}
+```{r} 
+#| label: timeseries 
+#| fig.cap: "A sample time series for the three fish populations, with the system being controlled, respectively, by the MSY optimal mortality, by $80$ percent of the MSY optimal mortality, and by DRL."
+#| out.width: "6.5in"
 policy_labels <- list("msy" = "MSY Const. Mortality", "tac"= "80% MSY Const. Mortality", "ppo"= "DRL")

### 2023-01-26 manuscript/manuscript.Rmd
reward shaping, normalizing, hyperparameter tuning

@@
 The constant 50 was chosen so as to select against highly extractive policies in the early stages of training, and thus guide the agent towards sustainable policies.
+This technique, falling under the ‘‘umbrella’’ category of *reward shaping*, is commonly used -- alongside hyper-parameter tuning and state space normalization -- to improve the speed of training by avoiding badly-performing local optima in the gradient ascent algorithm.
+
+Thus, we normalized state space (to be contained in a cube of length 2) and engaged in reward shaping in order to optimize our agent's learning.
+We did not need to tune hyper-parameters for this goal, however.
 

### 2023-01-26 manuscript/manuscript.Rmd
began figure action-X and the discussion leading up to it

@@
 
-In Fig. \ref{fig:timeseries} we show one example time series for each of the three controls we analyzed: 
-the left panel uses an optimalconstant mortality control $F^*X$, the middle panel uses $0.8 F^* X$, while the right panel is controlled by the trained DRL agent.
+We evaluated each of the three control approaches (MSY constant mortality, 80% MSY constant mortality and DRL) by running 50 episodes using the respective control rule.
+In Fig. \ref{fig:timeseries} we show the time series of one such repetition.
+The left panel uses an MSY constant mortality control $F^*X$, the middle panel uses $0.8 F^* X$, while the right panel is controlled by the trained DRL agent.
 We see that that optimal constant mortality control leads to a near-extinction event at around $t=150$.
@@
 First, that the tendency towards early episode ends is fixed by the conservative approach of fishing at 80% of the optimal mortality rate.
-In App. \ref{app:tbd} we discuss a warning about this approach, however: the arbitrarily chosen percentange itself lives on a ‘‘knife's edge.’’
-More concretely, choosing a higher percentage, say 90%, can switch the behaviour of the controlled system and give rise to the proliferation of early episode ends.
+(In App. \ref{app:tbd} we discuss a warning about this approach, however: the arbitrarily chosen percentage itself lives on a ‘‘knife's edge.’’
+More concretely, choosing a higher percentage, say 90%, can switch the behaviour of the controlled system and give rise to the proliferation of early episode ends.)
 Second, that while the conservative 80% approach has a stable behavior, it consistently produces smaller harvests than a DRL generated policy.
+
+```{r} 
+#| label: rew_len_fig
+#| fig.cap: "Bar charts for the episode lengths (top panel) and episode total harvests (bottom panel) for 50 episodes." 
+#| out.width: "6.5in"
+p1 <- msy_tac_ppo |> 
+    group_by(rep, policy) |>
+    filter(t==max(t), species == "X", rep < 50) |>
+    ggplot(aes(rep, t)) + 
+      geom_col() +
+      facet_wrap(~factor(policy, levels = c("msy", "tac", "ppo")), labeller = policy_labeller ) +
+      labs(y = "episode length", x = "repetition") +
+      theme(axis.text.x = element_blank())
+
+p2 <- msy_tac_ppo |> 
+    group_by(rep, policy) |>
+    filter(t==max(t), species == "X", rep < 50) |>
+    ggplot(aes(rep, reward)) + 
+      geom_col() +
+      facet_wrap(~factor(policy, levels = c("msy", "tac", "ppo")), labeller = policy_labeller ) +
+      labs(y = "episode harvest", x = "repetition") +
+      theme(axis.text.x = element_blank())
+
+(p1 / p2) + plot_annotation(tag_levels = 'A')
+```
+
+To understand the DRL-derived policy, Fig. \ref{fig:action-X} depicts the harvest as a function of $X$.
+Specifically, for each point $(x,y,z)$ visited by one of the aforementioned 50 evaluation episodes, we evaluate the harvest collected by the DRL agent and we subsequently plot the scatter data as a function of the projection $x$ onto the $X$ axis.
+
+```{r}
+#| label: action-X
+#| fig.cap: "TBD"
+#| out.width: "6.5in"
+p <- ppo_df |>
+  mutate(X = X+1, Y = Y+1, Z = Z+1, harvest = action*X)
+
+```
 
@@
 
-```{r rew_len_fig, fig.cap="TBD.", out.width = "6.5in"}
+```{r rew_len_fig_old, fig.cap="TBD.", out.width = "6.5in"}
 library(patchwork)

### 2023-01-26 manuscript/manuscript.Rmd
ended action-X and its discussion, started discussion leading to fig:state-space

@@
 Specifically, for each point $(x,y,z)$ visited by one of the aforementioned 50 evaluation episodes, we evaluate the harvest collected by the DRL agent and we subsequently plot the scatter data as a function of the projection $x$ onto the $X$ axis.
+In panel A we have used color to add the a $Y$-dependence, while in panel B we did the same with $Z$-dependence.
+This shows that the $Z$ and $Y$ dependence of the policy is rather weak and in fact it has roughly the same shape as an escapement policy.
+
+Fig. \ref{action-X} additionally shows (as a red line) the optimal escapement policy.
+This policy was computed in a manner similar to how the MSY constant mortality was computed: for each escapement in a grid of 100 values between 0 and 1, 50 episodes were run using the corresponding escapement policy.
+The value with the best mean episode reward was chosen as the optimal escapement value.
 
@@
 #| out.width: "6.5in"
-p <- ppo_df |>
-  mutate(X = X+1, Y = Y+1, Z = Z+1, harvest = action*X)
+opt_escapement_harvest <- function(X) case_when(
+  X > best_e ~ X - best_e,
+  X <= best_e ~ 0
+)
 
+py <- ppo_df |>
+  mutate(X = X+1, Y = Y+1, Z = Z+1, harvest = action*X) |>
+  ggplot(aes(X, harvest, color = Y)) +
+  geom_point() +
+  theme(aspect.ratio = 1/1) +
+  stat_function(fun = opt_escapement_harvest, col = 'red')
+
+pz <- ppo_df |>
+  mutate(X = X+1, Y = Y+1, Z = Z+1, harvest = action*X) |>
+  ggplot(aes(X, harvest, color = Z)) +
+  geom_point() +
+  theme(aspect.ratio = 1/1) +
+  stat_function(fun = opt_escapement_harvest, col = 'red')
+
+(py | pz) + plot_annotation(tag_levels = c('A'))
 ```
+
+In Fig. \ref{fig:state-space}, we perform a back to back visualization of all the policies evaluated up to now---MSY constant mortality, 80% MSY constant mortality, DRL and escapement.
+This highlights the similarity between the policy found by the DRL agent and the optimal escapement policy.
 

### 2023-01-26 manuscript/figures/classical_fisheries_textbox.Rmd
Add files via upload

---
output: 
  pdf_document:
    extra_dependencies: ["tcolorbox", "tikz", "bm"]
---

\begin{tcolorbox}[colback=red!5!white,colframe=red!75!black,title=\large Classical approaches to sustainable fisheries]

There are several existing approaches that have been used to manage fisheries, including \textbf{escapement}, control rules associated with \textbf{maximum sustainable yield (MSY)}, among others. We collectively refer to these as $classical$, and will compare their performance to RL-based management strategies. While often complex models of the ecosystem are used to estimate the size of fish population of interest, these classical strategies derive an optimal harvest policy using a simple model for the system dynamics. Across these strategies, setting the harvest quota has the shared aspect of reducing the complex dynamics of the fishery ecosystem to a single equation governing the harvested population, $X$.

A common example is the surplus production model that assumes logistic population growth in the absence of exploitation [@Schaefer1954]:

\begin{equation*}
X_{t+1} - X_t = r X_t( 1 - X_t / K) - h_t = L(X_t) - h_t
\end{equation*}

The interaction between $X$ and its environment is summarized to two parameters, the maximum intrinsic growth rate $r$, and the carrying capacity $K$. In the equation above, $h_t$ is the $harvest$ at time step $t$. The goal is to choose the harvest policy $h:X_t\mapsto h_t$, such that long-term profits are maximized.

An advantage of one dimensional approaches is that the optimal policy is intuitive and often known exactly. For example, in the logistic equation pointed out above, the biomass associated with maximum sustainable yield of the system is attained at $X = X_{MSY} = K/2$. The optimizer is an \textbf{escapement} policy, which corresponds to a harvest, $h_t$, that keeps the system at its optimal growth rate as much as possible:

\begin{equation*}
  h_t = \begin{cases}
    X_t - K/2, \qquad &\text{if } X_t > K/2\\
    0, \qquad &\text{else.}
  \end{cases}
\end{equation*}

Escapement policies, or more generally $bang bang$ policies, tend to be the optimal solution for these types of control problems. A drawback of these solutions in the fishery context, is the possible presence of several time steps with zero harvest. To mend this, certain suboptimal solutions have been constructed for fishery management.

One ubiquitous solution, often called \textbf{maximum sustainable yield (MSY)}, provides biological reference points that can be simply calculated from the surplus production model parameters. The solution consists of letting $h(X) = rX/2$, so that

\begin{equation*}
MSY = h(X_{MSY}) = rK/4.
\end{equation*}

That is, at the MSY biomass, $X_{MSY}$, the logistic growth of $X$ is cancelled exactly by the harvest. Additionally, the constant rate of fishing mortality, $F_{MSY}$, that results in $X_{MSY}$ is attained at $F_{MSY}=r/2$. This \textbf{MSY constant mortality} rule fixes the drawback in the escapement policy by having $h(X)>0$ for all $X>0$. It, however, has its own drawbacks, as it is particularly sensitive to misestimates of the parameter $r$.

Due to this, similar but more conservative policies are often applied where the constant rate of fishing mortality is $< F_{MSY}$. This control rule consists on reducing the inclination of the line defined by $h(X)$ using a prefactor $\alpha$ in $h(X) = \alpha\cdot rX/2$. Plausible examples are $\alpha = 0.8$ or $0.9$; here we examine an \textbf{80\% MSY constant mortality} policy (ref. supplemental here?).

\end{tcolorbox}

### 2023-01-26 manuscript/manuscript.Rmd
mini
added textbox attepmt one

@@
   - \usepackage{tikz-cd}
+  - \usepackage{tcolorbox}
+  - \usepackage{etoolbox}
   - \linenumbers
@@
 
+\begin{tcolorbox}[colback=red!5!white,colframe=red!75!black,title=\large Classical approaches to sustainable fisheries]
+\scriptsize
+
+There are several existing approaches that have been used to manage fisheries, including \textbf{escapement}, control rules associated with \textbf{maximum sustainable yield (MSY)}, among others.
+We collectively refer to these as $classical$, and will compare their performance to RL-based management strategies.
+While often complex models of the ecosystem are used to estimate the size of fish population of interest, these classical strategies derive an optimal harvest policy using a simple model for the system dynamics.
+Across these strategies, setting the harvest quota has the shared aspect of reducing the complex dynamics of the fishery ecosystem to a single equation governing the harvested population, $X$.
+
+A common example is the surplus production model that assumes logistic population growth in the absence of exploitation [@Schaefer1954]:
+
+\begin{equation*}
+X_{t+1} - X_t = r X_t( 1 - X_t / K) - h_t = L(X_t) - h_t
+\end{equation*}
+
+The interaction between $X$ and its environment is summarized to two parameters, the maximum intrinsic growth rate $r$, and the carrying capacity $K$.
+In the equation above, $h_t$ is the $harvest$ at time step $t$.
+The goal is to choose the harvest policy $h:X_t\mapsto h_t$, such that long-term profits are maximized.
+
+An advantage of one dimensional approaches is that the optimal policy is intuitive and often known exactly.
+For example, in the logistic equation pointed out above, the biomass associated with maximum sustainable yield of the system is attained at $X = X_{MSY} = K/2$.
+The optimizer is an \textbf{escapement} policy, which corresponds to a harvest, $h_t$, that keeps the system at its optimal growth rate as much as possible:
+
+\begin{equation*}
+  h_t = \begin{cases}
+    X_t - K/2, \qquad &\text{if } X_t > K/2\\
+    0, \qquad &\text{else.}
+  \end{cases}
+\end{equation*}
+
+Escapement policies, or more generally $bang bang$ policies, tend to be the optimal solution for these types of control problems.
+A drawback of these solutions in the fishery context, is the possible presence of several time steps with zero harvest.
+To mend this, certain suboptimal solutions have been constructed for fishery management.
+
+One ubiquitous solution, often called \textbf{maximum sustainable yield (MSY)}, provides biological reference points that can be simply calculated from the surplus production model parameters.
+The solution consists of letting $h(X) = rX/2$, so that
+
+\begin{equation*}
+MSY = h(X_{MSY}) = rK/4.
+\end{equation*}
+
+That is, at the MSY biomass, $X_{MSY}$, the logistic growth of $X$ is cancelled exactly by the harvest.
+Additionally, the constant rate of fishing mortality, $F_{MSY}$, that results in $X_{MSY}$ is attained at $F_{MSY}=r/2$.
+This \textbf{MSY constant mortality} rule fixes the drawback in the escapement policy by having $h(X)>0$ for all $X>0$.
+It, however, has its own drawbacks, as it is particularly sensitive to misestimates of the parameter $r$.
+
+Due to this, similar but more conservative policies are often applied where the constant rate of fishing mortality is $< F_{MSY}$.
+This control rule consists on reducing the inclination of the line defined by $h(X)$ using a prefactor $\alpha$ in $h(X) = \alpha\cdot rX/2$.
+Plausible examples are $\alpha = 0.8$ or $0.9$; here we examine an \textbf{80\% MSY constant mortality} policy (ref.
+supplemental here?).
+
+\end{tcolorbox}
+
 The intuition for a constant mortality strategy is simple and well established
@@
 The state space used was a cube $[0,2]^{\times 3}$, while the policies explored mapped states $N_t$ to a single number, $h_t = \pi_\theta(N_t)$, the harvest quota on species $X$ at that timestep.
-Each episode in the training and in the evaluation stages was at most 600 time steps long.
+Each episode in the training and in the evaluation stages was at most 200 time steps long.
 Episodes were ended early in case of a ‘‘near-extinction event,’’ a moment in which any of the populations dipped below a threshold value, i.e.,

### 2023-01-26 manuscript/manuscript.Rmd
reworded math models a bit

@@
 For $n$ species, these models have the general form
-
 \begin{align}
   \label{eq:general model}
-  \Delta N_t := N_{t+1} - N_t = f(N_t) - F^T N_t,
+  \begin{split}
+    \Delta X_t &:= X_{t+1} - X_t = f_X(N_t) - M_X X_t\\
+    \Delta Y_t &:= X_{t+1} - Y_t = f_Y(N_t) - M_Y Y_t\\
+    &\dots
+  \end{split}
 \end{align}
-where $N_t = (X_t,\ Y_t,\ \dots) \in \mathbb{R}^{n}_+$ is a vector of populations, $f:\mathbb{R}^{n}\to\mathbb{R}^{n}$ is an arbitrary function, where the product of the mortality and the stock size, $F^T N_t$ is the harvest size at timestep $t$.
-Here, $F\in\mathbb{R}^n$ is a vector of fishing intensities, where $F_i$ is the intensity with which the $i$-th species is fished.
-We will focus on the case where only the first species is fished, $F= (F_1,\ 0,\ \dots,\ 0)$.
+where $N_t = (X_t,\ Y_t,\ \dots) \in \mathbb{R}^{n}_+$ is a vector of populations and $f_i:\mathbb{R}^{n}\to\mathbb{R}$ are arbitrary functions.
+Here, $M_i=M_i(N_t)\in\mathbb{R}_+$ is a state-dependent fish mortality arising from harvesting (sometimes referred to as *fishing effort*).
+Thus, the term $M_X X_t$ is the total $X$ harvest at time $t$.
+We will focus on systems where a single species, $X$ is harvested, i.e. $M_Y = M_Z = \dots = 0$.
 

### 2023-01-26 code/plots.Rmd
datafreeze: modified code/*.py files to create data in a standardized form. created final data files data/*.xz, removed previous data/*.gz files (still probably need to remove them from history with git filter-branch (TBD). Modified code blocks within manuscript.rmd to fit the new data form, things are stable here.
filename

@@
 best <- best_reward(msy)
+best
 ```
@@
 ```{r}
-ppo <- read_csv("../data/PPO250.csv.gz")
+ppo <- read_csv("../data/PPO200.csv.xz")
 ```

### 2023-01-26 manuscript/manuscript.Rmd

--- 2023-01-26 manuscript/manuscript.Rmd
@@
 
+```{r}
+#| label: textbox
+#| out.width: "1\\linewidth"
+
+knitr::include_graphics("figures/textbox.png")
+```
+
 The intuition for a constant mortality strategy is simple and well established
@@
 In Fig. \ref{fig:state-space}, we perform a back to back visualization of all the policies evaluated up to now---MSY constant mortality, 80% MSY constant mortality, DRL and escapement.
-This highlights the similarity between the policy found by the DRL agent and the optimal escapement policy.
+The policies are computed at each state along the paths they traverse in the 50 evaluation rounds, and the values are projected on the $X$-$Z$ plane for visualization.
+This figure highlights the similarity between the policy found by the DRL agent and the optimal escapement policy.
+
+Notice the anti-correlation between $X$ and $Y$ along the paths, this is especially prominent for the constant mortality plots.
+This anti-correlation is, furthermore, observed in the time series plots (see Fig. \ref{fig:timeseries} and App. \ref{app:tbd}).
+
+```{r}
+#| label: state-space
+#| fig.cap: "Policy heatmaps.
+Each policy, say, $\\pi$, was evaluated at points $(x,y,z)$ visited during any of the 50 evaluation rounds.
+We projected this data down to the $X$-$Y$ plane---i.e.
+we plotted the heatmap $\\pi(x,y,z)$ as a function of $x$ and $y$."
+#| out.width: "6.5in"
+i_tac <- which.min(abs(actions - best_action * 0.8))
+
+# color pallette limits
+library(RColorBrewer)
+myPalette <- colorRampPalette(rev(brewer.pal(11, "Spectral")))
+sc <- scale_colour_gradientn(colours = myPalette(100), limits=c(0,0.25))
+
+pmsy <- msy_df |>
+  mutate(X = X+1, Y = Y+1, Z = Z+1, harvest = action * X) |>
+  filter(action == best_action) |>
+  ggplot(aes(X, Y, color = harvest)) +
+  geom_point() + sc +
+  xlim(0,1) +
+  ylim(0,0.85) +
+  ggtitle("Opt. Const. Mortality")
+
+ptac <- msy_df |>
+  mutate(X = X+1, Y = Y+1, Z = Z+1, harvest = action * X) |>
+  filter(action == actions[[i_tac]]) |>
+  ggplot(aes(X, Y, color = harvest)) +
+  geom_point() + sc +
+  xlim(0,1) +
+  ylim(0,0.85) +
+  ggtitle("80% Opt. Const. Mortality")
+
+pppo <- ppo_df |>
+  mutate(X = X+1, Y = Y+1, Z = Z+1, harvest = action * X) |>
+  ggplot(aes(X, Y, color = harvest)) +
+  geom_point() + sc +
+  xlim(0,1) +
+  ylim(0,0.85) +
+  ggtitle("DRL")
+
+pesc <- opt_escapement |>
+  mutate(X = X+1, Y = Y+1, Z = Z+1, 
+        harvest = case_when(
+          X - escapement > 0 ~ X - escapement,
+          X - escapement <= 0 ~  0)
+         ) |> 
+  ggplot(aes(X, Y, color = harvest)) +
+  geom_point() + sc +
+  xlim(0,1) +
+  ylim(0,0.85) +
+  ggtitle("Opt. Const. Escapement")
+
+(
+  (pmsy | ptac) / (pppo | pesc) +
+  plot_layout(guides = "collect") & theme(legend.position = 'right')
+) + 
+  plot_annotation(tag_levels = c('A'))
+```
 
@@
 
-```{r state-space, fig.cap="TBD.", out.width = "6.5in"}
+```{r state-space_old, fig.cap="TBD.", out.width = "6.5in"}
 i_tac <- which.min(abs(actions - best_action * 0.8))

### 2023-01-26 manuscript/manuscript.Rmd
erased old methods and results section

@@
 
-# Old Methods and Results
-
-## Methods
-
-Model-based controls, such as MSY, are known to have serious deficiencies when the model used to find the optimal control does not match reality.
-As mentioned, for the particular case of fisheries this can lead to depletion of resources and affect severely the surrounding ecosystems.
-Because of this, we have opted to compare the performance of several data-based approaches to fishery control (see Fig. \ref{fig:decision approaches}).
-
-As shown in Fig.
-\ref{fig:decision approaches}, these data-based strategies rely on choosing a family of *approximating functions* for the policy, $\mathcal{P}:=\{\pi_\theta\}_\theta$, parametrized by some vector $\theta$.
-These functions are deterministic policies, i.e. they map states to actions, or, specifically,
-$$
-  \pi_\theta : \mathbb{R}^n \to \mathbb{R}_+,
-$$
-where, as before, $n$ is the number of species.
-The aim of the optimization process is to find a parameter choice $\theta^*$ for which $\pi_{\theta^*}$ is optimal within $\mathcal{P}$.
-
-Our focus here will be comparing two data-based approaches: *constant mortality* and *DRL*.
-Constant mortality is, as the name suggests, based on MSY.
-
-For the constant mortality approach, we choose a linear harvest $\pi_F(N_t)= F X_t$, where the parameter $F$ is scalar, and where $N_t = (X_t,\ \dots\ )\in \mathbb{R}^n$.
-This first control is intended to delineate how one might apply MSY to a real-world fishery.
-We will also compare a TAC-inspired rule of thumb which uses, instead, the parameter value $F_{\text{TAC}} = 0.8 F^*$.
-In App. \ref{app:tbd}, we discuss several other MSY-based approaches which yield similar results to the strategies just discussed.
-
-In the DRL approach, we instead parametrize $\mathcal{P}$ using a neural network with two hidden layers of 64 neurons.
-Our hypothesis is that the additional flexibility yielded by neural networks will allow the agent solutions that are better than MSY based approaches.
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=0.8\linewidth]{figures/DecisionApproaches.pdf}
-  \caption{
-  \label{fig:decision approaches}
-  TBD.
-  }
-\end{figure}
-
-## Results
-
-\textbf{Reword a bit.}
-We trained a DRL agent to obtain a candidate policy
-\begin{align*}
-  \pi: (X_t,Y_t,Z_t) \mapsto h_t
-\end{align*}
-for eq. \eqref{eq:3d model}.
-The agent could fully and precisely observe the state of the system at each time step and decide on its harvest based on this information. 
-The agent was trained using the PPO algorithm, using a neural network with two hidden layers of 64 neurons to express $\pi$.
-Details of the training process may be found in App. \ref{app:tbd}.
-Our code is available at TBD.
-
-Rewards were defined to be a compound of two contributions.
-The first is economical: if at time $t$ a harvest of $h_t$ is secured, then the agents total reward is increased by $h_t$.
-The second regards sustainability: if an extinction happens at time step $t$, then a penalty of $-50/t$ is incurred by the agent.
-The choice of the constant $50$ in this rule was rather arbitrary: we set it to be a fourth of the maximal episode length ($200$ time steps), so as to encourage the agent to find policies that do not lead to extinctions below the time scale of around 100 time steps.
-\textbf{Too long of a "sorry" message?}
-
-We compare the performance of this agent to the performance of MSY and of an 80% TAC.
-
-Our results suggest that DRL can find solutions that are more sustainable *and* more profitable than any constant mortality strategy.
-Indeed, DRL finds solutions in which, for certain periods of time, no harvest is performed on the system.
-These periods effectively allow the system to, in a sense, "reset" itself.
-They lead to a distinctive oscillating behavior of the system state over time, with periods of relative depletion followed by periods of growth. 
-Figure TBD exemplifies this point by comparing one times series of the system state obtained using each of these policies.
-
-```{r timeseries_old, fig.cap="TBD.", out.width = "6.5in"}
-policy_labeller <- function(variable,value){
-  return(policy_labels[value])
-}
-msy_tac_ppo |> 
-    group_by(rep, policy) |>
-    filter(rep == 6) |>
-    ggplot(aes(t, abundance, col = species)) + 
-        geom_line()  + 
-        scale_color_manual(values=c('palegreen3', 'coral1', 'deepskyblue4')) +
-        facet_wrap(~factor(policy, levels = c("msy", "tac", "ppo")), labeller = policy_labeller ) +
-        labs(y = "abundance", x = "time") +
-        theme(aspect.ratio = 2/3)
-```
-
-```{r rew_len_fig_old, fig.cap="TBD.", out.width = "6.5in"}
-library(patchwork)
-
-p1 <- msy_tac_ppo |> 
-    group_by(rep, policy) |>
-    filter(t==max(t), species == "X", rep < 50) |>
-    ggplot(aes(rep, t)) + 
-      geom_col() +
-      facet_wrap(~factor(policy, levels = c("msy", "tac", "ppo")), labeller = policy_labeller ) +
-      labs(y = "episode length", x = "repetition") +
-      theme(axis.text.x = element_blank())
-
-p2 <- msy_tac_ppo |> 
-    group_by(rep, policy) |>
-    filter(t==max(t), species == "X", rep < 50) |>
-    ggplot(aes(rep, reward)) + 
-      geom_col() +
-      facet_wrap(~factor(policy, levels = c("msy", "tac", "ppo")), labeller = policy_labeller ) +
-      labs(y = "episode harvest", x = "repetition") +
-      theme(axis.text.x = element_blank())
-
-(p1 / p2) + plot_annotation(tag_levels = 'A')
-```
-
-```{r state-space_old, fig.cap="TBD.", out.width = "6.5in"}
-i_tac <- which.min(abs(actions - best_action * 0.8))
-
-# color pallette limits
-library(RColorBrewer)
-myPalette <- colorRampPalette(rev(brewer.pal(11, "Spectral")))
-sc <- scale_colour_gradientn(colours = myPalette(100), limits=c(0,0.25))
-
-pmsy <- msy_df |>
-  mutate(X = X+1, Y = Y+1, Z = Z+1, harvest = action * X) |>
-  filter(action == best_action) |>
-  ggplot(aes(X, Z, color = harvest)) +
-  geom_point() + sc +
-  ggtitle("Opt. Const. Mortality")
-
-ptac <- msy_df |>
-  mutate(X = X+1, Y = Y+1, Z = Z+1, harvest = action * X) |>
-  filter(action == actions[[i_tac]]) |>
-  ggplot(aes(X, Z, color = harvest)) +
-  geom_point() + sc +
-  ggtitle("80% Opt. Const. Mortality")
-
-pppo <- ppo_df |>
-  mutate(X = X+1, Y = Y+1, Z = Z+1, harvest = action * X) |>
-  ggplot(aes(X, Z, color = harvest)) +
-  geom_point() + sc +
-  ggtitle("DRL")
-
-pesc <- opt_escapement |>
-  mutate(X = X+1, Y = Y+1, Z = Z+1, 
-        harvest = case_when(
-          X - escapement > 0 ~ X - escapement,
-          X - escapement <= 0 ~  0)
-         ) |> 
-  ggplot(aes(X, Z, color = harvest)) +
-  geom_point() + sc +
-  ggtitle("Opt. Const. Escapement")
-
-(
-  (pmsy | ptac) / (pppo | pesc) +
-  plot_layout(guides = "collect") & theme(legend.position = 'right')
-) + 
-  plot_annotation(tag_levels = c('A'))
-```
-
-```{r escapement_timeseries, fig.cap = "TBD", out.width = "6.5in"}
-
-ppo_mod <- ppo_df |> 
-  mutate(X = X+1, Y = Y+1, Z = Z+1, escapement = (1-action)*X, time = t)
-
-avg_esc <- mean(ppo_mod$escapement)
-# = 0.6201131
-
-ppo_esc <- ppo_mod |>
-  filter(rep == 6) |>
-  group_by(rep) |>
-  ggplot(aes(time, escapement), alpha = rep) +
-  geom_line() +
-  ylim(0,1) +
-  geom_hline(aes(yintercept=avg_esc), color="red")+
-  theme(aspect.ratio = 1/5)
-
-ppo_esc
-```
-
-**Probably more for the discussion**
-It is well known that following MSY policies can lead to over-fishing.
-Our results reproduce this behavior and, furthermore, warn that using a TAC policy instead might not be enough to gain sustainability.
-Figure \ref{rew_len_fig}-A summarizes episode length distribution for MSY, TAC and DRL.
-While the DRL-derived policy does not lead to extinctions, both MSY and TAC do so, and thus their episode mean length is lower than the maximal episode length (in our case set at 200 years).
-This difference is reflected on the profits earned by the agent, as shown in Fig. TBD-B.
-
 ## Acknowledgements

### 2023-01-26 manuscript/manuscript.Rmd
added new version of Millies conceptual figure

@@
 ```{r}
-
 #| label: conceptual
 #| echo: FALSE
-#| message" FALSE
+#| message: FALSE
 #| fig.align: 'center'
@@
 For example, in this paper, we find harvest strategies for a three-species model using a model-free RL algorithm, PPO.
-#| (C) We show how the RL-based solution that considers all three species finds harvest policies with higher average reward and with (D) lower probability of harvest species collapse than the classical approaches, only considering the dynamics of V1.',
-#| out.width: '1\\linewidth', 
+#| (C) We show how the RL-based solution that considers all three species finds harvest policies with higher average reward and with (D) lower probability of harvest species collapse than the classical approaches, only considering the dynamics of V1.'
+#| out.width: '6.5in' 
 #| fig.pos: 'H'
 
-#knitr::include_graphics("Figures/conceptual-figure.png")
+knitr::include_graphics("figures/conceptual.png")
 ```

### 2023-01-27 manuscript/manuscript.Rmd
added some discussion

@@
 
+# Discussion
+
+Fisheries are complex ecosystems, with interactions between species leading to highly non-linear dynamics.
+While current models for population estimation include many important aspects of this complexity, it is still common to use simplified dynamical models in order to guide decision making.
+This provides very interpretable solutions, such as constant mortality policies.
+There is a drawback here, however: due to the simplicity of these dynamical models, the policies might not respond effectively in many situations---situations where the predictions of these simple models deviates considerably from reality.
+Because of this, policies such as MSY have faced pushback and are believed to have contributed to the depletion of fish stocks.
+
+We propose an alternative approach to the problem of fishery control: to use a more expressive -- albeit more complicated -- dynamical model to guide decision making.
+Furthermore, rather than computing the optimal control policy for the model -- something that is impossible in practice for complex systems -- we use deep reinforcement learning to obtain a ‘‘pretty darn good’’ policy.
+
+In this paper we have exemplified this approach by applying it to the control of a three-species fishery model.
+In this example, we have shown that the DRL agent finds an approximately *bang-bang* type of solution.
+While bang-bang solutions are known to be optimal for a wide variety of single-species fishery control problems [@alexandru], there is no such proof of optimality for general multi-species models.
+In this sense this example is rather surprising and reassuring: while no general guarantee points at the optimality of bang-bang solutions for multi-species model, the existence of one such example provides a benchmark for the use of DRL in fishery control.
+
+One intuition for why a constant escapement policy performs so well in such a model comes from the anticorrelation between $X$ and $Y$ noted in the results section.
+We have observed in a variety of policies, both classical as well as some obtained by training DRL agents.
+In all of these, the pattern has been present.
+In fact, ‘‘crashes’’ in the $X$ population are almost invariably shortly preceeded by dramatic increases in $Y$.
+We hypothesize that this effectively leads to a dimensionality reduction: once the value of $X$ is known, along a path traced by the control system, then $Y$ may be roughly predicted.
+A similar, although weaker, anticorrelation is also observed between $X$ and $Z$.
+These two informal observations point at an effective simplification of the prediction process: the values of $Y$ and $Z$ are much less impactful to policy decisions than the value of $X$ itself.
+We have not yet pushed further to establish a more precise picture of this dynamical simplification.
+
+This way, this manuscript functions as a first step in the process of incorporating DRL into fishery decision making.
+Importantly, DRL is equipped with tools to deal with more complicated models which might require a more refined approach to control them.
+For example *long short-term memories (LSTMs)* may be used to incorporate not only $N_t$ but also $\mathrm{d}N_t/\mathrm{d}t$ into policy decisions---while this additional degree of freedom would make analytical control theory approaches unwieldy, it may be easily and cost-effectively implemented in DRL.
+Similarly, *curriculum learning* could be incorporated to deal with phenomena such as parametric noise or parametric drift [@openai].
+
+This way, DRL provides a framework flexible enough to capture the intricate population dynamics that are relevant in fishery management.
+We have shown that highly non-linear systems may give rise to simplified and good policies and that these may be found via reinforcement learning.
+Moreover, the framework allows for the possibility that these simplifications do not in fact happen, and that good harvest policies are highly non-linear, multi-parameter functions.
+
 ## Acknowledgements

### 2023-01-27 manuscript/manuscript.Rmd
bit more discussion

@@
 
+Conversely, our analysis suggests that, for high dimensional systems, bang-bang policies lead to increased sustainability and long-term profits as compared to constant mortality strategies.
+Similarly, in App. \ref{app:tbd} we investigate several generalizations of the constant effort approach. 
+For example, we examine strategies with a ramp-down from constant effort to zero effort at low populations.
+These, too, yield subpar results as compared to escapement.
+This seems to indicate that the ‘‘rest periods’’ so prominent in both the DRL solution and in the optimal bang-bang policy -- periods of several time steps with no fishing at all -- are important to maintain the sustainability of the fishery and, additionally, increase long-term profits.
+
 ## Acknowledgements

### 2023-01-27 manuscript/manuscript.Rmd
mini changes around msy tuning

@@
 \textbf{Tuning the constant mortality strategy.}
-A grid of 200 mortality rates was laid in the interval $[0, 0.1]$.
+A grid of 100 mortality rates was laid in the interval $[0, 0.1]$.
 For each one of these rates, say $F$, we simulated 50 episodes based on \eqref{eq:3d model stoch}:
@@
 
-The result of the tuning procedure is shown in Fig. \ref{fig:linear_tuning}. 
+The result of the tuning procedure is shown in Fig. \ref{fig:tuning}. 
 There, we see that the optimal value $F^*$ lies in an area with high episode reward variance.
@@
 ```{r}
-#| label: linear_tuning, 
+#| label: tuning
 #| fig.cap : "Result of the tuning procedure for the constant mortality strategy. 
@@
   xlim(0,2) +
-  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
+  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
+  xlab("X") + ylab("harvest")
 

### 2023-01-27 manuscript/manuscript.Rmd
:pen:

--- 2023-01-27 manuscript/manuscript.Rmd
@@
 
-MSC:
-- MSC code 1
-- MSC code 2
+#MSC:
+#- MSC code 1
+#- MSC code 2
 
@@
 
-Ecological management may be understood as an optimal control problem - a manager must determine a sequence of actions (a "policy") to optimize some desired objective(s) and constraints (e.g.
-conservation goals, natural resource extraction, economic and social constraints.)
-Despite the complexity inherent in natural ecosystems, analytically tractable models have long played an essential role in their understanding and management.
-...
+Much effort has been spent grappling with the complexity of our natural world in contrast to relative simplicity of the models we use to understand it.
+Now, as we grapple with the manifold consequences of human impacts on that world, we are faced with difficult questions of figuring out how best to respond when our knowledge and models fall so far short of the complexity our actions have disrupted.
+Heroic amounts of data and computation are being brought to bear on developing better, more complex, more realistic models of our environments and ecosystems, yet these models remain disconnected or oversimplify the space of actions they seek to inform.
+In this paper we explore a different tack. 
+Here we explore the potential of a emerging data-driven machine learning technique, not to better capture the complexity of the natural processes, but rather, to embrace the complexity of the potential _policies_ that might be used to manage them.
+Even as our research community develops simulations of the natural world that fit only in supercomputers, we analyze a space of policies that would fit on index cards.
 
-<!-- Why fisheries-focused examples -->
-Global fisheries management has long been a crucible for both the development of theory and best practices in actual management for many reasons.
-Fisheries are an essential natural resource that provide the primary source of protein for one in every four humans, and have faced widely documented declines due to overfishing [@Worm2006, @Costello2016].
+<!-- Why fisheries-focused examples -- maybe we can cut/merge this with paragraph above -->
+There are myriad examples from climate scenarios to protected areas, we will illustrate this with a specific example in the management of marine fisheries, which is both an important issue to itself and a frequent crucible in which methods of ecological management have methods for ecological management have been fired before extending to other fields, and where the space of possible policies is arguably already more richly developed than most.
+Fisheries are an essential natural resource that provide the primary source of protein for one in every four humans, and have faced widely documented declines due to over-fishing [@Worm2006, @Costello2016].
 Because of this importance, fisheries also represent some of the longest and best studied timeseries of population dynamics available to ecologists, as well as a correspondingly rich area for the development of theory and practice.
 Methods developed in the study and management of fisheries are frequently applied to other areas of ecological management, from concepts such as maximum sustainable yield [], optimal control under observation uncertainty [@Memarzadeh2019], to adaptive management [@Walters], management strategy evaluation [@Punt], to the studies of conservation policy and international negotiations [@].
+
+Mathematically, the ecological management may be understood as an optimal control problem - a manager must determine a sequence of actions (a "policy") to optimize some desired objective(s) and constraints (e.g.
+conservation goals, natural resource extraction, economic and social constraints.)
+Given an appropriate representation of the underlying mechanistic processes involved, such sequential decision problems may be posed more precisely as Markov Decision Process [MDP, @Marescot2013; @Memarzadeh2019].
+Such problems are notoriously difficult owing not only to the complexity of those underlying mechanisms alone, but to the exponentially branching consequences the decision-maker's actions have upon that system -- a game of chess is not won by thinking only one move ahead at a time.
+Thus, while MDP problems have been studied in a conservation context since they were first introduced more than six decades ago, their application has been highly constrained to simple scenarios.
+Meanwhile, a class of machine learning techniques designed for MDP problems known as reinforcement learning (RL), has cracked open the door for exploring complex strategies in more realistic settings [@rl_intro].
+
+- This is not a tutorial or a sales pitch for RL - which brings at least as many difficulties as advantages -- including being hard to use, easy to get stuck on far from optimal performance, which itself may be hard to diagnose, as we have already discussed in our tutorial.
+The method
+
+has through the use of clever algorithms, deep neural networks, and massive amounts of data,
 
@@
 ```{r}
+
 #| label: conceptual
 #| echo: FALSE
-#| message: FALSE
+#| message" FALSE
 #| fig.align: 'center'
@@
 For example, in this paper, we find harvest strategies for a three-species model using a model-free RL algorithm, PPO.
-#| (C) We show how the RL-based solution that considers all three species finds harvest policies with higher average reward and with (D) lower probability of harvest species collapse than the classical approaches, only considering the dynamics of V1.'
-#| out.width: '6.5in' 
+#| (C) We show how the RL-based solution that considers all three species finds harvest policies with higher average reward and with (D) lower probability of harvest species collapse than the classical approaches, only considering the dynamics of V1.',
+#| out.width: '1\\linewidth', 
 #| fig.pos: 'H'
 
-knitr::include_graphics("figures/conceptual.png")
+#knitr::include_graphics("Figures/conceptual-figure.png")
 ```
@@
 
-# Discussion
+# Old Methods and Results
 
-Fisheries are complex ecosystems, with interactions between species leading to highly non-linear dynamics.
-While current models for population estimation include many important aspects of this complexity, it is still common to use simplified dynamical models in order to guide decision making.
-This provides very interpretable solutions, such as constant mortality policies.
-There is a drawback here, however: due to the simplicity of these dynamical models, the policies might not respond effectively in many situations---situations where the predictions of these simple models deviates considerably from reality.
-Because of this, policies such as MSY have faced pushback and are believed to have contributed to the depletion of fish stocks.
+## Methods
 
-We propose an alternative approach to the problem of fishery control: to use a more expressive -- albeit more complicated -- dynamical model to guide decision making.
-Furthermore, rather than computing the optimal control policy for the model -- something that is impossible in practice for complex systems -- we use deep reinforcement learning to obtain a ‘‘pretty darn good’’ policy.
+Model-based controls, such as MSY, are known to have serious deficiencies when the model used to find the optimal control does not match reality.
+As mentioned, for the particular case of fisheries this can lead to depletion of resources and affect severely the surrounding ecosystems.
+Because of this, we have opted to compare the performance of several data-based approaches to fishery control (see Fig. \ref{fig:decision approaches}).
 
-In this paper we have exemplified this approach by applying it to the control of a three-species fishery model.
-In this example, we have shown that the DRL agent finds an approximately *bang-bang* type of solution.
-While bang-bang solutions are known to be optimal for a wide variety of single-species fishery control problems [@alexandru], there is no such proof of optimality for general multi-species models.
-In this sense this example is rather surprising and reassuring: while no general guarantee points at the optimality of bang-bang solutions for multi-species model, the existence of one such example provides a benchmark for the use of DRL in fishery control.
+As shown in Fig.
+\ref{fig:decision approaches}, these data-based strategies rely on choosing a family of *approximating functions* for the policy, $\mathcal{P}:=\{\pi_\theta\}_\theta$, parametrized by some vector $\theta$.
+These functions are deterministic policies, i.e. they map states to actions, or, specifically,
+$$
+  \pi_\theta : \mathbb{R}^n \to \mathbb{R}_+,
+$$
+where, as before, $n$ is the number of species.
+The aim of the optimization process is to find a parameter choice $\theta^*$ for which $\pi_{\theta^*}$ is optimal within $\mathcal{P}$.
 
-One intuition for why a constant escapement policy performs so well in such a model comes from the anticorrelation between $X$ and $Y$ noted in the results section.
-We have observed in a variety of policies, both classical as well as some obtained by training DRL agents.
-In all of these, the pattern has been present.
-In fact, ‘‘crashes’’ in the $X$ population are almost invariably shortly preceeded by dramatic increases in $Y$.
-We hypothesize that this effectively leads to a dimensionality reduction: once the value of $X$ is known, along a path traced by the control system, then $Y$ may be roughly predicted.
-A similar, although weaker, anticorrelation is also observed between $X$ and $Z$.
-These two informal observations point at an effective simplification of the prediction process: the values of $Y$ and $Z$ are much less impactful to policy decisions than the value of $X$ itself.
-We have not yet pushed further to establish a more precise picture of this dynamical simplification.
+Our focus here will be comparing two data-based approaches: *constant mortality* and *DRL*.
+Constant mortality is, as the name suggests, based on MSY.
 
-This way, this manuscript functions as a first step in the process of incorporating DRL into fishery decision making.
-Importantly, DRL is equipped with tools to deal with more complicated models which might require a more refined approach to control them.
-For example *long short-term memories (LSTMs)* may be used to incorporate not only $N_t$ but also $\mathrm{d}N_t/\mathrm{d}t$ into policy decisions---while this additional degree of freedom would make analytical control theory approaches unwieldy, it may be easily and cost-effectively implemented in DRL.
-Similarly, *curriculum learning* could be incorporated to deal with phenomena such as parametric noise or parametric drift [@openai].
+For the constant mortality approach, we choose a linear harvest $\pi_F(N_t)= F X_t$, where the parameter $F$ is scalar, and where $N_t = (X_t,\ \dots\ )\in \mathbb{R}^n$.
+This first control is intended to delineate how one might apply MSY to a real-world fishery.
+We will also compare a TAC-inspired rule of thumb which uses, instead, the parameter value $F_{\text{TAC}} = 0.8 F^*$.
+In App. \ref{app:tbd}, we discuss several other MSY-based approaches which yield similar results to the strategies just discussed.
 
-This way, DRL provides a framework flexible enough to capture the intricate population dynamics that are relevant in fishery management.
-We have shown that highly non-linear systems may give rise to simplified and good policies and that these may be found via reinforcement learning.
-Moreover, the framework allows for the possibility that these simplifications do not in fact happen, and that good harvest policies are highly non-linear, multi-parameter functions.
+In the DRL approach, we instead parametrize $\mathcal{P}$ using a neural network with two hidden layers of 64 neurons.
+Our hypothesis is that the additional flexibility yielded by neural networks will allow the agent solutions that are better than MSY based approaches.
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=0.8\linewidth]{figures/DecisionApproaches.pdf}
+  \caption{
+  \label{fig:decision approaches}
+  TBD.
+  }
+\end{figure}
+
+## Results
+
+\textbf{Reword a bit.}
+We trained a DRL agent to obtain a candidate policy
+\begin{align*}
+  \pi: (X_t,Y_t,Z_t) \mapsto h_t
+\end{align*}
+for eq. \eqref{eq:3d model}.
+The agent could fully and precisely observe the state of the system at each time step and decide on its harvest based on this information. 
+The agent was trained using the PPO algorithm, using a neural network with two hidden layers of 64 neurons to express $\pi$.
+Details of the training process may be found in App. \ref{app:tbd}.
+Our code is available at TBD.
+
+Rewards were defined to be a compound of two contributions.
+The first is economical: if at time $t$ a harvest of $h_t$ is secured, then the agents total reward is increased by $h_t$.
+The second regards sustainability: if an extinction happens at time step $t$, then a penalty of $-50/t$ is incurred by the agent.
+The choice of the constant $50$ in this rule was rather arbitrary: we set it to be a fourth of the maximal episode length ($200$ time steps), so as to encourage the agent to find policies that do not lead to extinctions below the time scale of around 100 time steps.
+\textbf{Too long of a "sorry" message?}
+
+We compare the performance of this agent to the performance of MSY and of an 80% TAC.
+
+Our results suggest that DRL can find solutions that are more sustainable *and* more profitable than any constant mortality strategy.
+Indeed, DRL finds solutions in which, for certain periods of time, no harvest is performed on the system.
+These periods effectively allow the system to, in a sense, "reset" itself.
+They lead to a distinctive oscillating behavior of the system state over time, with periods of relative depletion followed by periods of growth. 
+Figure TBD exemplifies this point by comparing one times series of the system state obtained using each of these policies.
+
+```{r timeseries_old, fig.cap="TBD.", out.width = "6.5in"}
+policy_labeller <- function(variable,value){
+  return(policy_labels[value])
+}
+msy_tac_ppo |> 
+    group_by(rep, policy) |>
+    filter(rep == 6) |>
+    ggplot(aes(t, abundance, col = species)) + 
+        geom_line()  + 
+        scale_color_manual(values=c('palegreen3', 'coral1', 'deepskyblue4')) +
+        facet_wrap(~factor(policy, levels = c("msy", "tac", "ppo")), labeller = policy_labeller ) +
+        labs(y = "abundance", x = "time") +
+        theme(aspect.ratio = 2/3)
+```
+
+```{r rew_len_fig_old, fig.cap="TBD.", out.width = "6.5in"}
+library(patchwork)
+
+p1 <- msy_tac_ppo |> 
+    group_by(rep, policy) |>
+    filter(t==max(t), species == "X", rep < 50) |>
+    ggplot(aes(rep, t)) + 
+      geom_col() +
+      facet_wrap(~factor(policy, levels = c("msy", "tac", "ppo")), labeller = policy_labeller ) +
+      labs(y = "episode length", x = "repetition") +
+      theme(axis.text.x = element_blank())
+
+p2 <- msy_tac_ppo |> 
+    group_by(rep, policy) |>
+    filter(t==max(t), species == "X", rep < 50) |>
+    ggplot(aes(rep, reward)) + 
+      geom_col() +
+      facet_wrap(~factor(policy, levels = c("msy", "tac", "ppo")), labeller = policy_labeller ) +
+      labs(y = "episode harvest", x = "repetition") +
+      theme(axis.text.x = element_blank())
+
+(p1 / p2) + plot_annotation(tag_levels = 'A')
+```
+
+```{r state-space_old, fig.cap="TBD.", out.width = "6.5in"}
+i_tac <- which.min(abs(actions - best_action * 0.8))
+
+# color pallette limits
+library(RColorBrewer)
+myPalette <- colorRampPalette(rev(brewer.pal(11, "Spectral")))
+sc <- scale_colour_gradientn(colours = myPalette(100), limits=c(0,0.25))
+
+pmsy <- msy_df |>
+  mutate(X = X+1, Y = Y+1, Z = Z+1, harvest = action * X) |>
+  filter(action == best_action) |>
+  ggplot(aes(X, Z, color = harvest)) +
+  geom_point() + sc +
+  ggtitle("Opt. Const. Mortality")
+
+ptac <- msy_df |>
+  mutate(X = X+1, Y = Y+1, Z = Z+1, harvest = action * X) |>
+  filter(action == actions[[i_tac]]) |>
+  ggplot(aes(X, Z, color = harvest)) +
+  geom_point() + sc +
+  ggtitle("80% Opt. Const. Mortality")
+
+pppo <- ppo_df |>
+  mutate(X = X+1, Y = Y+1, Z = Z+1, harvest = action * X) |>
+  ggplot(aes(X, Z, color = harvest)) +
+  geom_point() + sc +
+  ggtitle("DRL")
+
+pesc <- opt_escapement |>
+  mutate(X = X+1, Y = Y+1, Z = Z+1, 
+        harvest = case_when(
+          X - escapement > 0 ~ X - escapement,
+          X - escapement <= 0 ~  0)
+         ) |> 
+  ggplot(aes(X, Z, color = harvest)) +
+  geom_point() + sc +
+  ggtitle("Opt. Const. Escapement")
+
+(
+  (pmsy | ptac) / (pppo | pesc) +
+  plot_layout(guides = "collect") & theme(legend.position = 'right')
+) + 
+  plot_annotation(tag_levels = c('A'))
+```
+
+```{r escapement_timeseries, fig.cap = "TBD", out.width = "6.5in"}
+
+ppo_mod <- ppo_df |> 
+  mutate(X = X+1, Y = Y+1, Z = Z+1, escapement = (1-action)*X, time = t)
+
+avg_esc <- mean(ppo_mod$escapement)
+# = 0.6201131
+
+ppo_esc <- ppo_mod |>
+  filter(rep == 6) |>
+  group_by(rep) |>
+  ggplot(aes(time, escapement), alpha = rep) +
+  geom_line() +
+  ylim(0,1) +
+  geom_hline(aes(yintercept=avg_esc), color="red")+
+  theme(aspect.ratio = 1/5)
+
+ppo_esc
+```
+
+**Probably more for the discussion**
+It is well known that following MSY policies can lead to over-fishing.
+Our results reproduce this behavior and, furthermore, warn that using a TAC policy instead might not be enough to gain sustainability.
+Figure \ref{rew_len_fig}-A summarizes episode length distribution for MSY, TAC and DRL.
+While the DRL-derived policy does not lead to extinctions, both MSY and TAC do so, and thus their episode mean length is lower than the maximal episode length (in our case set at 200 years).
+This difference is reflected on the profits earned by the agent, as shown in Fig. TBD-B.
 

### 2023-01-27 code/scratch.Rmd
carl :pen: intro
Note: some old text has been moved to code/scratch.Rmd while we decide we want to eliminate it.

--- 2023-01-27 manuscript/manuscript.Rmd
@@
----
-title: "Pretty darn good control: when are approximate solutions better than approximate models"
-titlerunning: Pretty darn good control
-thanks: | 
-    This material is based upon work supported by the National Science
-    Foundation under Grant No. DBI-1942280.
-
-authors: 
-- name: Felipe Montealegre-Mora
-  address: Department of Environmental Science, Policy, and Management, University of California, Berkeley
-- name: Marcus Laperolerie
-  address: Department of Environmental Science, Policy, and Management, University of California, Berkeley
-- name: Melissa Chapman
-  address: Department of Environmental Science, Policy, and Management, University of California, Berkeley
-- name: Abigail G. Keller
-  address: Department of Environmental Science, Policy, and Management, University of California, Berkeley
-- name: Carl Boettiger
-  address: Department of Environmental Science, Policy, and Management, University of California, Berkeley
-  email: cboettig@berkeley.edu
-
-keywords:
-- Optimal Control
-- Reinforcement Learning
-- Uncertainty
-- Decision Theory
-
-#PACS: 
-#- PAC1
-#- superPAC
-
-#MSC:
-#- MSC code 1
-#- MSC code 2
-
-abstract: |
-  Write something here
-
-bibliography: bibliography.bib
-#biblio-style: spphys
-
-header-includes:
-  - \usepackage{amsmath}
-  - \usepackage{amssymb}
-  - \usepackage{lineno}
-  - \usepackage{setspace}
-  - \usepackage{bookmark}
-  - \usepackage{float}
-  - \usepackage{graphicx}
-  - \usepackage{caption}
-  - \usepackage{subcaption}
-  - \usepackage{tikz-cd}
-  - \usepackage{tcolorbox}
-  - \usepackage{etoolbox}
-  - \linenumbers
-  - \doublespacing
-
-# bibstyle options spbasic(default), spphys, spmpsci
-# output: rticles::springer_article
-
-output: 
-  rticles::arxiv_article:
-    keep_tex: true
-
-editor_options: 
-  markdown: 
-    wrap: sentence
-
----
-
-```{r knit_global, include=FALSE}
-knitr::opts_chunk$set(cache = TRUE, echo = FALSE, message = FALSE, 
-                      warning = FALSE, fig.height=5, fig.width=7.5)
-library(tidyverse)
-library(patchwork)
-```
-
-```{r}
-# for best const mortality and best const escapement
-msy_optimizer <- function(df) {
-  df |> 
-  group_by(action, rep) |>
-  filter(t == max(t)) |> 
-  group_by(action) |>
-  summarise(mean_reward = mean(reward), sd = sd(reward)) |> 
-  filter(mean_reward == max(mean_reward))
-}
-escapement_optimizer <- function(df) {
-  df |> 
-  group_by(escapement, rep) |>
-  filter(t == max(t)) |> 
-  group_by(escapement) |>
-  summarise(mean_reward = mean(reward), sd = sd(reward)) |> 
-  filter(mean_reward == max(mean_reward))
-}
-
-# Collect data (outsource to separate R script later TBD)
-msy_df <- read_csv("../data/msy.csv.xz")
-best_action <- msy_optimizer(msy_df)$action
-escapement_df <- read_csv("../data/escapement.csv.xz", show_col_types = FALSE)
-best_e <- escapement_optimizer(escapement_df)$escapement
-ppo_df <- read_csv("../data/PPO200.csv.xz")
-
-opt_escapement <- escapement_df |> filter(escapement == best_e)
-
-# set up for plots:
-
-msy_sim <- msy_df |>
-    filter(action == best_action) |> 
-    group_by(rep) |>
-    pivot_longer(c("X", "Y", "Z"),
-                names_to = "species", values_to = "abundance") |>
-    mutate(abundance = abundance + 1, policy = "msy") # natural units, policy name
-
-actions <- unique(msy_df$action)
-i <- which.min(abs(actions - best_action* 0.8))
-tac_sim <- msy_df |> 
-    filter(action == actions[[i]]) |>
-    pivot_longer(c("X", "Y", "Z"),
-        names_to = "species", values_to = "abundance") |>
-    mutate(abundance = abundance + 1, policy = "tac")
-
-ppo_sim <- ppo_df |> 
-    pivot_longer(c("X", "Y", "Z"),
-        names_to = "species", values_to = "abundance") |>
-    mutate(abundance = abundance + 1, policy = "ppo")
-
-best_e <- 0.61 # optimal constant escapement, computed from ../data/escapement.csv.gz
-esc_sim_df <- escapement_df |> 
-    pivot_longer(c("X", "Y", "Z"),
-        names_to = "species", values_to = "abundance") |>
-    mutate(abundance = abundance + 1)
-
-msy_tac_ppo <- bind_rows(msy_sim, tac_sim, ppo_sim)
-```
-
-<!--
- Special Issue Description:
-
-This special issue highlights the development of novel data-driven methods, including statistics, machine learning, parameter estimation, and uncertainty quantification, and combinations thereof, towards modeling biological systems.
-These newly developed methods will tackle challenges that are commonly encountered when modeling real-world experimental, field, pre-clinical, or clinical data.
-Examples of such challenges include high dimensionality, computational complexity, observation or process error, model bias, and intra- or inter-individual heterogeneity.
-Contributions to this special issue require validation of new methods with real-world data or simulated data sets that contain features of real-world data that exemplify an outlined modeling challenge.
-Papers should include a discussion justifying why the developed method is novel and not an application of previously developed methods, as well as how the method may be broadly applicable across different areas of biology, including medical, ecological, genetics, and epidemiological applications.
--->
-
-# Introduction {#intro}
-
-<!--
-Existing methods for optimal control methods struggle to deal with the complexity commonly encountered in real-world systems, including dimensionality, process error, model bias and heterogeneity. 
-Instead of tackling these complexities directly, researchers have typically sought to find exact optimal solutions to simplified models of the processes in question. 
-When is the optimal solution to a very approximate, stylized model better than an approximate solution to a more accurate model? 
-This question has largely gone unanswered owing to the difficulty of finding even approximate solutions in the case of complex models.  
-Our approach draws on recent algorithmic and computational advances in deep reinforcement learning. 
-These methods have hitherto focused on problems in games or robotic mechanics, which operate under precisely known rules. 
-We demonstrate the ability for novel algorithms using deep neural networks to successfully approximate such solutions (the "policy function" or control rule) without knowing or ever attempting to infer a model for the process itself.
-This powerful new technique lets us finally begin to answer the question.
-We show that in many but not all cases, the optimal policy for a carefully chosen over-simplified model can still out-perform these novel algorithms trained to find approximate solutions to simulations of a realistically complex system.
-This illustrates the promise of combining complex, realistic simulations with emerging machine-learning techniques to improve our understanding and management of ecological systems.
--->
-
-Much effort has been spent grappling with the complexity of our natural world in contrast to relative simplicity of the models we use to understand it.
-Now, as we grapple with the manifold consequences of human impacts on that world, we are faced with difficult questions of figuring out how best to respond when our knowledge and models fall so far short of the complexity our actions have disrupted.
-Heroic amounts of data and computation are being brought to bear on developing better, more complex, more realistic models of our environments and ecosystems, yet these models remain disconnected or oversimplify the space of actions they seek to inform.
-In this paper we explore a different tack. 
-Here we explore the potential of a emerging data-driven machine learning technique, not to better capture the complexity of the natural processes, but rather, to embrace the complexity of the potential _policies_ that might be used to manage them.
-Even as our research community develops simulations of the natural world that fit only in supercomputers, we analyze a space of policies that would fit on index cards.
-
-<!-- Why fisheries-focused examples -- maybe we can cut/merge this with paragraph above -->
-There are myriad examples from climate scenarios to protected areas, we will illustrate this with a specific example in the management of marine fisheries, which is both an important issue to itself and a frequent crucible in which methods of ecological management have methods for ecological management have been fired before extending to other fields, and where the space of possible policies is arguably already more richly developed than most.
-Fisheries are an essential natural resource that provide the primary source of protein for one in every four humans, and have faced widely documented declines due to over-fishing [@Worm2006, @Costello2016].
-Because of this importance, fisheries also represent some of the longest and best studied timeseries of population dynamics available to ecologists, as well as a correspondingly rich area for the development of theory and practice.
-Methods developed in the study and management of fisheries are frequently applied to other areas of ecological management, from concepts such as maximum sustainable yield [], optimal control under observation uncertainty [@Memarzadeh2019], to adaptive management [@Walters], management strategy evaluation [@Punt], to the studies of conservation policy and international negotiations [@].
-
 Mathematically, the ecological management may be understood as an optimal control problem - a manager must determine a sequence of actions (a "policy") to optimize some desired objective(s) and constraints (e.g.
@@
 Instead, fisheries management approaches such as MSE rely on theory based on more tractable models to restrict the search space of possible policies.
-Specifically, most fisheries policies are framed in terms of a *constant mortality* policy, (sometimes a simple piece-wise linear function, i.e.
-a single linear segment, is introduced as a precaution a very low stock sizes), where fishing 'mortality' is defined as the ratio of harvest (catch) to stock size, $F = H_t/x_t$.
-
-```{r}
-#| label: textbox
-#| out.width: "1\\linewidth"
-
-knitr::include_graphics("figures/textbox.png")
-```
-
-The intuition for a constant mortality strategy is simple and well established
-
-Constant mortality policies have a strong basis in theory as well as practical justifications.
 
@@
 
-Deep reinforcement learning has opened up increasingly compelling alternative
-
 Effort $Q = E X_t$
@@
 and/or *effort controls*, restrictions on the fishing effort -- in numbers of boats, time of year, or days at sea
-
-# Figures brainstorm
-
-Figure 0: contrast 1D vs 3D strategies (see commented out diagram).
-
-```{r}
-#| label: conceptual
-#| echo: FALSE
-#| message: FALSE
-#| fig.align: 'center'
-#| fig.cap:  '(A) Classical management strategies often have the common aspect of reducing the complex dynamics of the fishery ecosystem to a single equation governing the harvested population (say, $V1$).
-where the interaction between $V1$ and its environment is summarized to a set of parameters, such as the growth rate $r$, and the carrying capacity $K$.
-An advantage of one-dimensional approaches is that the optimal policy is often known exactly, and, moreover, is intuitive.
-For example, in the logistic equation shown, the maximal sustainable yield of the system is attained at K/2.
-#| (B) RL allows us to solve for quantitative decision strategies considering more complex ecosystem dynamics.
-For example, in this paper, we find harvest strategies for a three-species model using a model-free RL algorithm, PPO.
-#| (C) We show how the RL-based solution that considers all three species finds harvest policies with higher average reward and with (D) lower probability of harvest species collapse than the classical approaches, only considering the dynamics of V1.'
-#| out.width: '6.5in' 
-#| fig.pos: 'H'
-
-knitr::include_graphics("figures/conceptual.png")
-```
 
@@
 
-Figure 1: 1-D and 3-D model conceptual figure. something about the objective / decision
-Figure 2: Stability / multistability in the 1D and 3D models.  state space + time views
-Figure 3: The 1-D optimal management solution.  'constant escapement' intuition etc
+## original draft of abstract?
 
-RL figures:
-- schematic of RL (a-la previous marcus paper Fig 1)
-- Neural network optimization figure
+Existing methods for optimal control methods struggle to deal with the complexity commonly encountered in real-world systems, including dimensionality, process error, model bias and heterogeneity. 
+Instead of tackling these complexities directly, researchers have typically sought to find exact optimal solutions to simplified models of the processes in question. 
+When is the optimal solution to a very approximate, stylized model better than an approximate solution to a more accurate model? 
+This question has largely gone unanswered owing to the difficulty of finding even approximate solutions in the case of complex models.  
+Our approach draws on recent algorithmic and computational advances in deep reinforcement learning. 
+We demonstrate the ability for novel algorithms using deep neural networks to successfully approximate such solutions (the "policy function" or control rule) without knowing or ever attempting to infer a model for the process itself.
+This powerful new technique lets us finally begin to answer the question.
+We show that in many but not all cases, the optimal policy for a carefully chosen over-simplified model can still out-perform these novel algorithms trained to find approximate solutions to simulations of a realistically complex system.
+This illustrates the promise of combining complex, realistic simulations with emerging machine-learning techniques to improve our understanding and management of ecological systems.
 
-Results figures:
-- timeseries example of management under the RL policy.  probably compare to managing under 1-D solution / rule-of-thumb methods
-- state-space view of management doughnut 
-- visualization / encapsulation of the policy heatmaps, slices, policy vs position along ellipse 
-- reward plot over time (comparing methods)
-
-# Mathematical models of fisheries
-
-In this section we introduce some models describing the population dynamics of fisheries.
-In general, the class of models that appear in this context are *first order finite difference equations* (potentially stochastic difference equations).
-For $n$ species, these models have the general form
-\begin{align}
-  \label{eq:general model}
-  \begin{split}
-    \Delta X_t &:= X_{t+1} - X_t = f_X(N_t) - M_X X_t\\
-    \Delta Y_t &:= X_{t+1} - Y_t = f_Y(N_t) - M_Y Y_t\\
-    &\dots
-  \end{split}
-\end{align}
-where $N_t = (X_t,\ Y_t,\ \dots) \in \mathbb{R}^{n}_+$ is a vector of populations and $f_i:\mathbb{R}^{n}\to\mathbb{R}$ are arbitrary functions.
-Here, $M_i=M_i(N_t)\in\mathbb{R}_+$ is a state-dependent fish mortality arising from harvesting (sometimes referred to as *fishing effort*).
-Thus, the term $M_X X_t$ is the total $X$ harvest at time $t$.
-We will focus on systems where a single species, $X$ is harvested, i.e. $M_Y = M_Z = \dots = 0$.
-
-## A single species model
-
-Optimal control policies for fisheries are frequently based on 1-dimensional models, $n=1$.
-The most familiar model of $f(X)$ is that of *logistic growth*, for which
-
-\begin{align}
-  \label{eq:logistic}
-  f(X_t) = r X_t\big(1 - X_t / K \big) =: L(X_t;\ r, K).
-\end{align}
-
-The model this $f$ gives rise to is known as the Gordon-Schaefer model, after foundational work which first developed the concept of maximum sustainable yield, MSY, [@Gordon1954; @Schaefer1954], which remains a central concept in fisheries management.
-In this model, the maximum growth rate occurs at population size of $N = K/2$.
-This value is approached asymptotically, from any positive initial state $N_0 \in (0,K)$, under a constant mortality $F = r/2$ policy.
-At equilibrium this produces an asymptotic harvest of $rK/4$ per time step---the maximum yield sustainable indefinitely.
-
-Real world ecological systems are obviously far more complicated than this simple model suggests.
-One particularly important aspect that has garnered much attention is the potential for the kind of highly non-linear functions that can support dynamics such as alternative stable states and hysteresis.
-A seminal example of such dynamics was introduced in \cite{may77}, using a one-dimensional model of a prey (resource) species under the pressure of a (fixed) predator:
-
-\begin{align}
-  \label{eq:may}
-  f_{\text{May}}(X_t)
-  := L(X_t;\ r, K) - F(X_t, H; \beta,c),
-\end{align}
-where,
-\begin{align*}
-  F(X_t,H;\ \beta,c) := \frac{\beta H X_t^2}{c^2 + X_t^2}.
-\end{align*}
-
-The model has five parameters: the growth rate $r$ and carrying capacity $K$ for $X$, a constant population $H$ of a species which preys on $X$, the maximal rate of predation $\beta$, and the predation half-maximum biomass $c$.
-
-Eq. \eqref{eq:may} is an interesting study case of a *tipping point* (saddle-node bifurcation) (see Fig. \ref{fig:may stable}).
-Holding the value of $\beta$ fixed, for intermediate values of $H$ there exist two stable fixed points for the state $\hat X_t$ of the system, these two attractors separated by an unstable fixed point.
-At a certain threshold value of $H$, however, the top stable fixed point collides with the unstable fixed point and both are annihilated.
-For this value of $H$, and for higher values, only the lower fixed point remains.
-This also creates the phenomenon of *hysteresis*, where returning $H$ to its original value is not sufficient to restore $\hat X_t$ to the original stable state.
-
-\begin{figure}
-  \centering
-  \includegraphics[width = 0.8\linewidth]{figures/MayStablePoints.pdf}
-  \caption{
-    \label{fig:may stable}
-    The fixed point diagram for the model \cite{may77} associated to $f_{\text{May}}$ as a function of varying the parameter $\beta H$.
-    That is, these are the zeroes of the function $f_{\text{May}}$. 
-    Stable fixed points (also known as *attractors*) are plotted using a solid line, while the unstable fixed point is shown as a dotted line.
-  }
-\end{figure}
-
-This structure implies two things.
-First, that a drift in $H$ could lead to catastrophic consequences, with the population $X_t$ plummeting to the lower fixed stable point.
-Second, that if the evolution of $X_t$ is *stochastic*, then, even at values of $H$ below the threshold point, the system runs a sizeable danger of tipping over towards the lower stable point.
-
-<!-- Discuss control solutions of this model(?) -->
-<!-- Briefly motivate reason to consider 3d model, adding competition and predation processes -->
-
-## A three species model
-
-Consider the following three species *generalization* of eq. \eqref{eq:may}:
-
-\begin{align}
-\label{eq:3d model}
-\begin{split}
-  \Delta X_t &= L(X_t;\ r_X, K_X) - F(X_t, Z_t;\ \beta, c) - c_{XY}X_tY_t - h_t,\\
-  \Delta Y_t &= L(Y_t;\ r_Y, K_Y) - D F(Y_t, Z_t;\ \beta, c) - c_{XY}X_tY_t,\\
-  \Delta Z_t &= \left(f(X_t + DY_t\right) - d_Z)Z_t.
-\end{split}
-\end{align}
-The model is conceptualized in Fig. \ref{fig:conceptual}, panel B.
-It contains three populations, the state at time $t$ is $(X_t, Y_t, Z_t)$. 
-Species $Z$ preys on both $X$ and $Y$, while the latter two compete for resources.
-There are ten parameters in this model: 
-The growth rate and carrying capacity, $r_X$, $K_X$, $r_Y$ and $K_Y$, of $X$ and $Y$.
-A parameter $c_{XY}$ mediating a Lotka-Volterra competition between $X$ and $Y$.
-A maximum hunting intensity $\beta$ and half-maximum $c$ specifying how $Z$ forages on $X$ and $Y$.
-A parameter $D$ regulating a relative preference of $Z$ to prey on $Y$.
-Finally, a death rate $d_Z$ and a parameter $f$ related to the birth rate of $Z$.
-
-This model generalizes eq. \eqref{eq:may} in the following sense:
-If $f=d_Z=c_{XY}=0$, and we let $Z_{t=0}=H$, then eq. \eqref{eq:3d model} becomes
-\begin{align*}
-  \Delta X_t &= L(X_t;\ r_X, K_X) - F(X_t, H;\ \beta, c) - h_t,\\
-  \Delta Y_t &= L(Y_t;\ r_Y, K_Y) - D F(Y_t, H;\ \beta, c),\\
-  Z_t &= H.
-\end{align*}
-The equations for $X_t$ and $Y_t$ thus become decoupled, with $X_t$ evolving according to \eqref{eq:may}.
-
-As can be expected, the dynamics of eq. \eqref{eq:3d model} are significantly more complex than those of eq. \eqref{eq:may}.
-There are a variety of fixed points for the equations---around 7 or 8 for wide ranges of parameter values.
-Fig. \ref{fig:tbd} provides a rough idea of the fixed point structure by plotting the bifurcation diagram obtained from varying $D$ and fixing the other parameters.
-
-Throughout most of this paper we will fix the parameters of the model according to Table \ref{tab:tbd}.
-At these values, the fixed points of the model are:
-\begin{align*}
-  TBD
-\end{align*}
-The classification of fixed points for multi-dimensional systems is generally more complicated than in one dimension.
-A fixed point might be stable against perturbations in one direction, but unstable against perturbations along another.
-While a detailed classification of these fixed points is beyond the scope of this work, we point out that the point $(TBD)$ is stable against perturbations in all directions.
-We will thus use it as a starting point for our simulations.
-
-## Stochastic dynamics
-
-We have, up to now, presented eqs. \eqref{eq:may} and \eqref{eq:3d model} as *deterministic* dynamical models for the sake of clarity.
-In practice, we will consider a stochastic contribution to these dynamics.
-This way, instead of eq. \eqref{eq:may}, we consider the following stochastic single-species system
-\begin{align}
-  \label{eq:may stoch}
-  \Delta X_t
-  = L(X_t;\ r, K) - F(X_t, H; \beta,c) - h_t + \eta_t,
-\end{align}
-where $\{\eta_t\}_t$ are independent Gaussian variables with zero mean.
-The variance $\sigma^2$ of $\eta_t$ is an additional parameter of the model.
-
-Similarly, we use the following dynamical model in place of eq. \eqref{eq:3d model}
-\begin{align}
-  \label{eq:3d model stoch}
-  \begin{split}
-    \Delta X_t &= L(X_t;\ r_X, K_X) - F(X_t, Z_t;\ \beta, c) - c_{XY}X_tY_t - h_t + \eta_{X,t},\\
-    \Delta Y_t &= L(Y_t;\ r_Y, K_Y) - D F(Y_t, Z_t;\ \beta, c) - c_{XY}X_tY_t + \eta_{Y,t},\\
-    \Delta Z_t &= \left(f(X_t + DY_t\right) - d_Z)Z_t + \eta_{Z,t}.
-\end{split}
-\end{align}
-Here, again, $\{\eta_{i,t}\}$ is a set of independent Gaussian variables centered at zero with variances, respectively, $\sigma_X^2$, $\sigma_Y^2$ and $\sigma_Z^2$.
+# Millie's suggestion for plots:
 
 ```{r}
-msy_rewards <- msy_df |>
-  group_by(action, rep) |>
-  filter(t == max(t)) |> 
-  group_by(action) |>
-  summarise(mean_reward = mean(reward),
-            sd = sd(reward))
-
-esc_rewards <- escapement_df |>
-  group_by(escapement, rep) |>
-  filter(t == max(t)) |> 
-  group_by(escapement) |>
-  summarise(mean_reward = mean(reward),
-            sd = sd(reward))
-
-rmsy <- msy_rewards |>
-  ggplot(aes(action, mean_reward, 
-             ymin = mean_reward - 2*sd,
-             ymax=mean_reward+2*sd)) +
-  geom_point() + geom_ribbon(alpha=0.4) +
-  labs(y = "episode reward", x = "constant mortality") +
-  theme(aspect.ratio = 1/1) + 
-  geom_vline(aes(xintercept=0.0360), color="red") + 
-  geom_vline(aes(xintercept=0.0288), color="green")
-
-resc <- esc_rewards |>
-  ggplot(aes(esc_level, mean_reward, 
-             ymin = mean_reward - 2*sd,
-             ymax=mean_reward+2*sd)) +
-  geom_point() + geom_ribbon(alpha=0.4) +
-  labs(y = "episode reward", x = "constant escapement") +
-  theme(aspect.ratio = 1/1)
-
-# (rmsy | resc)
+msy_sim <- msy_sim |> mutate(type = "Opt const mortality", cumulative_reward = reward)
+tac_sim <- tac_sim |>
+    mutate(abundance = abundance + 1) |>
+    mutate(type = "80% opt const mortality", 
+           cumulative_reward = reward)
+ppo_sim <- ppo_sim |> mutate(type = "DRL")
+all_sim <- bind_rows(msy_sim, tac_sim, ppo_sim)
 ```
 
-# Fishery management approaches
+```{r}
+all_sim |> #filter(rep < 10, species == "X") |> 
+    group_by(rep) |>
+    filter(t==max(t), rep < 10, species == "X") |>
+    ggplot(aes(rep, cumulative_reward)) + geom_col() +
+    facet_wrap(~type)
 
-<!-- Introduce the 1D and 3D models, and have Figure 1 already here. -->
+msy_sim |> 
+  group_by(rep) |>
+  filter(t==max(t), species=="X", rep < 10) |>
+  ggplot(aes(rep, cumulative_reward)) + geom_col()
 
-Here we review classical strategies for sustainable fishery management and we provide a birds-eye view of the alternative approach we propose.
-A summary of the contrast between the two strategies is given in Fig. \ref{fig:decision approaches}.
+tac_sim |> 
+  group_by(rep) |>
+  filter(t==max(t), species=="X", rep < 10) |>
+  ggplot(aes(rep, cumulative_reward)) + geom_col()
 
-## Classical approaches to sustainable fisheries {#sec:classical}
+all_sim |> 
+    mutate(across(type, factor, levels = c("Opt const mortality", "80% opt const mortality", 
+                                           "DRL"))) |>
+    group_by(rep, type) |>
+    filter(t==max(t), species=="X", rep < 10) |>
+    ggplot(aes(rep, cumulative_reward)) + 
+        geom_col()  + 
+        facet_wrap(~type) + 
+        labs(y = "reward", x = "repetition")
 
-There are several strategies that have been used to manage fisheries: *escapement*, *maximum sustainable yield (MSY)*, *total allowable catch (TAC)*, among others.
-We collectively refere to these as *classical*, and will compare their performance to RL-based management strategies.
-Classical strategies derive an optimal harvest policy using a simple model for the system dynamics.
-While often complex models of the ecosystem are used to estimate the fish population of interest (i.e. $X_t$), the quota is set only taking into account the estimate $X_t$.
-The function determining this quota is derived from a simple dynamical model for the population (see text box 1).
-
-\textbf{Text box 1 (TBD formatting)}
-Let us consider a ubiquitous simple example: an MSY policy based on a logistic growth model for $X_t$ (see eqs. \eqref{eq:general model} and \eqref{eq:logistic}).
-
-Classical strategies have the common aspect of reducing the complex dynamics of the fishery ecosystem to a single equation governing the harvested population (say, $F$). 
-A common example is using a logistic growth equation,
-$$
-  F_{t+1} - F_t = r F_t( 1 - F_t / K) - h_t =: L(F_t) - h_t,
-$$
-where the interaction between $F$ and its environment is summarized to two parameters, the growth rate $r$, and the carrying capacity $K$.
-In the equation above, $h_t$ is the *harvest* at timestep $t$.
-The goal is to choose the harvest policy $h:F_t\mapsto h_t$ such that long-term profits are maximized.
-
-An advantage of one dimensional approaches is that the optimal policy is often known exactly, and, moreover, is intuitive.
-For example, in the logistic equation pointed out above, the maximal sustainable yield of the system is attained at $F = F_{MSY} := K/2$.
-The optimizer is an *escapement* policy:
-$$
-  h_t = \begin{cases}
-    F_t - K/2, \qquad &\text{if, } F_t > K/2\\
-    0, \qquad &\text{else.}
-  \end{cases}
-$$
-This corresponds to keeping the system at its optimal growth rate as much as possible.
-
-Escapement policies, or more generally *bang bang* policies, tend to be the optimal solution for these types of control problems.
-A drawback of these solutions, in the fishery context, is the presence of several timesteps with zero harvest which can arise.
-To mend this, certain suboptimal solutions have been constructed for fishery management.
-
-One ubiquitous solution is simply called maximum sustainable yield (MSY).
-It consists on letting $h(F) = rF/2$, so that
-$$
-  h(F_{MSY}) = L(F_{MSY}) = r K/4.
-$$
-That is, at the MSY biomass, the logistic growth of $F$ is cancelled exactly by the harvest.
-
-The MSY rule fixes the drawback in the escapement policy by having $h(F)>0$ for all $F>0$.
-It, however, has its own drawbacks.
-It is particularly sensitive to misestimates of the parameter $r$, as we will discuss in Sec. \ref{sec:tbd}.
-Due to this, similar but more conservative policies have been used.
-
-*Total allowable catch (TAC)* is one such policy.
-It consists on reducing the inclination of the line defined by $h(F)$ using a prefactor $\alpha$ in $h(F) = \alpha\times rF/2$. 
-Plausible examples are $\alpha = 0.8$ or $0.9$.
-Another common alternative is to have a prefactor $\alpha = \alpha(F)$ which decreases from one to zero as $F$ decreases below a threshold:
-$$
-  h(F) = \alpha(F)\times rF/2,
-$$
-where,
-$$
-  \alpha(F) = \begin{cases}
-  1, \qquad &\text{if } F > F_{\text{thresh.}},\\
-  F / F_{\text{thresh.}}, \qquad & \text{else}.
-  \end{cases}
-$$
-We call this policy *variable total allowable catch (VTAC)*.
-
-## Dangers: Noisy parameters, noisy observations and model inaccuracy
-
-# Reinforcement learning {#sec:RL}
-
-Reinforcement learning (RL) is, in a nutshell, a way of approaching *control problems* through machine learning.
-An RL algorithm can be conceptually separated into two parts: an *agent*, and an *environment* which the agent can interact with.
-That is, the agent may act on the environment and thus change its state, while the environment gives a *reward* to the agent in return (see Fig. \ref{tbd}).
-The rewards encode the agent's goal.
-The main part of an RL algorithm is then to progressively improve the agent's *policy*, in order to maximize the cumulative reward received. 
-This is done by aggregating experience and learning from it.
-
-<!-- The *environment* is commonly a computer simulation, although it sometimes can be a real world system.
-The *agent*, on the other hand, is a computer program which interacts with the environment. -->
-
-For our use case, the environment will be a computational model of the population dynamics of a fishery, with the environment state being a vector of all the fish populations, $S=(V_1, V_2, H)$.
-At each time step, the agent harvests one of the populations, $V_1$. 
-This changes the state as
-\begin{align*}
-  (V_1,\ V_2,\ H) \mapsto ((1-q)V_1,\ V_2,\ H),
-\end{align*}
-where $q$ is a fishing *quota* set by the agent. 
-This secures a reward of $qV_1$.
-Afterwards, the environment undergoes a timestep under its natural dynamics given by \eqref{eq:3d model}.
-
-## Mathematical framework for RL
-
-Mathematically, RL may be formulated using a discrete time *partially observable Markov decision process (POMDP)*.
-This formalization is rather flexible and allows one, e.g., to account for situations where the agent may not fully observe the environment state, or where the only observations available to the agent are certain functions of the underlying state.
-For the sake of clarity, we will present here only the class of POMDPs which are relevant to our work: *fully observable MDPs with a trivial emission function* (FMDPs for short).
-An FMDP may be defined by the following data:
-
-- $\mathcal{S}$: *state space*, the set of states of the environment,
-
-- $\mathcal{A}$: *action space*, the set of actions which the agent may choose from,
-
-- $T(s_{t+1}|s_t, a_t)$: *transition operator*, a conditional distribution which describes the dynamics of the system,
-
-- $r(s_t, a_t)$: *reward function*, the reward obtained after performing action $a_t\in\mathcal{A}$ in state $s_t$,
-
-- $d(s_0)$: *initial state distribution*, the initial state of the environment is sampled from this distribution,
-
-- $\gamma\in[0,1]$: *discount factor*.
-
-At a time $t$, the FMDP agent observes the full state $s_t$ of the environment and chooses an action based on this observation according to a *policy function* $\pi(a_t | s_t)$.
-In return, it receives a discounted reward $\gamma^t r(a_t, s_t)$.
-The discount factor helps regularize the agent, helping the optimization algorithm find solutions which pay off within a timescale of $t \sim \log(\gamma^{-1})^{-1}$.
-
-With any fixed policy function, the agent will traverse a path 
-$\tau=(s_0,\ a_0,\ s_1,\ a_1\ \dots,\ s_{t_{\text{fin.}}})$ 
-sampled randomly from the distribution
-\begin{align*}
-  p_\pi(\tau) = 
-  d(s_0) \prod_{t=0}^{ t_{\text{fin.}}-1 }
-  \pi( a_t | s_t ) T( s_{t+1} | s_t, a_t ).
-\end{align*}
-Reinforcement learning seeks to optimize $\pi$ such that the expected rewards are maximal,
-$$  
-  \pi^* = \mathrm{argmax}\ \mathbb{E}_{\tau\sim p_\pi}[R(\tau)],
-$$
-where,
-$$
-  R(\tau) = \sum_{t=0}^{ t_{\text{fin.}}-1 } \gamma^t r(a_t, s_t),
-$$
-is the cumulative reward of path $\tau$.
-The function $J(\pi):=\mathbb{E}_{\tau\sim p_\pi}[R(\tau)]$ is called the *value function*.
-
-## Deep Reinforcement Learning
-
-The policy function $\pi$ often lives in a high- or even infinite-dimensional space.
-This makes it unfeasible to directly optimize $\pi$.
-In practice, an alternative approach is used: $\pi$ is optimized over a much lower-dimensional parametrized family of functions.
-Deep reinforcement learning uses this strategy, focusing on function families parametrized by neural networks. (See Fig. \ref{tbd}.)
-<!-- The figure should kinda explain how the parametrization arises, and how gradient ascent is used. Maybe namedrop backpropagataion for the sake of usefulness to the reader? -->
-
-Since our state and observation spaces are continuous, we will focus on deep reinforcement learning throughout this paper.
-Specifically, we parametrize $\pi$ using a neural network with two hidden layers of 64 neurons.
-
-We use the *proximal policy optimization (PPO)* algorithm to optimize $\pi$.
-Within the RL literature there is a wealth of algorithms from which to choose from, each with its pros and cons.
-Here we have focused on a single one of these for the sake of particularity---to draw a clear comparison between the RL-based and the classical fishery management approaches.
-In practice, further improvements can be expected by a careful selection of the optimization algorithm.
-
-## Model-free reinforcement learning
-
-Within control theory, the usual setup is one where we use as much information from the model as possible in order to derive an optimal solution.
-**(Wax poetic a bit about Bellmann eq. approaches here?)**
-
-The classical sustainable fishery management approaches summarized in Sec. \ref{sec:classical} are model-based controls.
-As we saw in that section, these controls may run into trouble in the case where there are inaccuracies in the model parameter estimates.
-
-More generally, there are many situations in which the exact model of the system is not known or not tractable.
-This is a standard situation in ecology: mathematical models capture the most prominent aspects of the ecosystem's dynamics, while ignoring or summarizing most of its complexity.
-In this case, it is clear, model-based controls run a grave danger of mismanaging the system.
-
-Reinforcement learning is, on the other hand, typically a model-free approach to control theory.\footnote{There are, however, approaches to perform model-based reinforcement learning.
-While we will not focus on these in this paper they are discussed in \cite{tbd1, tbd2}.}
-While the model is certainly used to generate training data, it is not directly used by the optimization algorithm.
-This provides more flexibility to use model-free RL in instances where the model of the system is not accurately known.
-In fact, it has been shown to generally be preferable to use model-free RL in such instances.
-
-**Mention CL here? Where solutions to other related problems in the curriculum are used to build solutions to the target problem.**
-
-This context provides a motivation for this paper. 
-Indeed, models for ecosystem dynamics are only ever approximate and incomplete descriptions of reality. 
-This way, it is plausible that model-free RL controls outperform currently used model-based controls in ecological management problems.
-
-# Methods
-
-\textbf{Training a DRL agent.}
-We trained a DRL agent parametrized by a neural network $\theta$ with two hidden, 64-neuron, layers.
-The state space used was a cube $[0,2]^{\times 3}$, while the policies explored mapped states $N_t$ to a single number, $h_t = \pi_\theta(N_t)$, the harvest quota on species $X$ at that timestep.
-Each episode in the training and in the evaluation stages was at most 200 time steps long.
-Episodes were ended early in case of a ‘‘near-extinction event,’’ a moment in which any of the populations dipped below a threshold value, i.e.,
-\begin{align}
-  \label{eq:thresh}
-  X_t \leq X_{\text{thresh.}},\quad
-  Y_t \leq Y_{\text{thresh.}},\quad
-  \text{or,}\quad
-  Z \leq Z_{\text{thresh.}}.
-\end{align}
-The reward function had two componenets:
-First, at each time step the harvest $h_t$ was received as a reward.
-Second, if the episode ended early (say, at time-step $t$) because of a near-extinction event, a negative reward of $-50/t$ was awarded.
-The constant 50 was chosen so as to select against highly extractive policies in the early stages of training, and thus guide the agent towards sustainable policies.
-This technique, falling under the ‘‘umbrella’’ category of *reward shaping*, is commonly used -- alongside hyper-parameter tuning and state space normalization -- to improve the speed of training by avoiding badly-performing local optima in the gradient ascent algorithm.
-
-Thus, we normalized state space (to be contained in a cube of length 2) and engaged in reward shaping in order to optimize our agent's learning.
-We did not need to tune hyper-parameters for this goal, however.
-
-We used the Ray framework [@ray_package] for training, in particular we used a Ray PPOConfig object to build the algorithm using the default values for all hyper-parameters (cf.
-[@repo] for the code used).
-The agent was trained for 250 iterations of the PPO optimization step.
-
-\textbf{Tuning the constant mortality strategy.}
-A grid of 100 mortality rates was laid in the interval $[0, 0.1]$.
-For each one of these rates, say $F$, we simulated 50 episodes based on \eqref{eq:3d model stoch}:
-At each time step the state $(X_t, Y_t, Z_t)$ was observed, a harvest of $FX_t$ was collected and then the system evolved according to its natural dynamics \eqref{eq:3d model stoch}.
-The optimal mortality rate was the value $F^*$ for which the mean episode reward was maximal.
-We additionally evaluated a more conservative strategy where a mortality of $0.8 F^*$ instead is used.
-
-The result of the tuning procedure is shown in Fig. \ref{fig:tuning}. 
-There, we see that the optimal value $F^*$ lies in an area with high episode reward variance.
-We will see in Fig. \ref{fig:rew_len_fig} that a likely candidate for this variance is the presence of episodes with early endings.
-Thus, the value $0.8 F^*$ supports a smaller variance, and correspondingly -- as Fig. \ref{fig:rew_len_fig} shows -- this policy leads to a much smaller probability of a near-extinction event.
+```
 
 ```{r}
-#| label: tuning
-#| fig.cap : "Result of the tuning procedure for the constant mortality strategy. 
-#| (A) Mean episode reward (averaged over 50 episodes) shown as scatter data, and shaded in gray are two standard deviations around each scatter point.
-Vertical lines correspond to the optimal constant mortality rate $F^* = $ TBD (red) and $0.8 F^*=$ TBD (green).
-#| (B) Several constant mortality policies shown for illustrative purposes.
-Slopes $0.02$, $0.03$, $0.055$ and $0.065$ shown in black; the optimal linear policy $F^*X$ shown in red, while $0.8F^*X$ shown in green."
-#| out.width : "6.5in"
-msy_rewards <- msy_df |>
-  group_by(action, rep) |>
-  filter(t == max(t)) |> 
-  group_by(action) |>
-  summarise(mean_reward = mean(reward),
-            sd = sd(reward))
+all_sim |> 
+    mutate(across(type, factor, levels = c("Opt const mortality", "80% opt const mortality", 
+                                           "DRL"))) |>
+    group_by(rep, type) |>
+    filter(rep == 0) |>
+    ggplot(aes(t, abundance, col = species)) + 
+        geom_line()  + 
+        facet_wrap(~type) + 
+        labs(y = "reward", x = "repetition") +
+        theme(aspect.ratio = 2/5)
+all_sim |>
+    group_by(rep,type) |>
+    filter(rep<10) |>
+    ggplot(aes(t, action)) + 
+    geom_line()  + 
+    facet_wrap(~type) + 
+    labs(y = "action", x = "repetition")
 
-rmsy <- msy_rewards |>
-  ggplot(aes(action, mean_reward, 
-             ymin = mean_reward - 2*sd,
-             ymax=mean_reward+2*sd)) +
-  geom_point() + geom_ribbon(alpha=0.4) +
-  labs(y = "episode reward", x = "constant mortality") +
-  theme(aspect.ratio = 1/1) + 
-  geom_vline(aes(xintercept=best_action), color="red") + 
-  geom_vline(aes(xintercept=0.8*best_action), color="green") +
-  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
-
-# linear function blueprint
-linear_f <- function(slope, x) slope*x
-# generates linear function with given slope
-linear_f_gen <- function(slope) (function(x) slope*x)
-
-# dummy data set to keep ggplot happy
-p <- ggplot(data = data.frame(x = 0), mapping = aes(x = x))
-
-# p + stat_function(fun = linear_f_gen(0.5)) + xlim(0,2)
-lines <- p + stat_function(fun = linear_f_gen(0.01)) + 
-  stat_function(fun = linear_f_gen(0.02)) +
-  stat_function(fun = linear_f_gen(0.03)) +
-  stat_function(fun = linear_f_gen(0.8*best_action), col = "green") +
-  stat_function(fun = linear_f_gen(best_action), col = "red") +
-  stat_function(fun = linear_f_gen(0.055)) + 
-  stat_function(fun = linear_f_gen(0.065)) +
-  xlim(0,2) +
-  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
-  xlab("X") + ylab("harvest")
-
-(rmsy | lines) + plot_annotation(tag_levels = c('A'))
+#all_sim |> group_by(rep) |> filter(t==max(t), species == "X") |> group_by(t)
 ```
 
-\textbf{Parameter values used.}
-The model's dynamic parameters were chosen as follows
-\begin{align*}
-  r_X = K_X = r_Y = K_Y = 1, \quad \beta = 0.3, \quad c = 0.1\\
-  c_{XY} = 0.5,\quad f = 0.15,\quad D = 1.1, \quad d_Z = 0.135.
-\end{align*}
-Moreover, the variances for the stochastic terms were chosen as
-$$
-  \sigma^2(\eta_{X,t}) = \sigma^2(\eta_{Y,t}) = \sigma^2(\eta_{Z,t}) = 0.05.
-$$
-The threshold values used to determine an early episode end were chosen as
-$$
-  X_{\text{thresh.}} = 0.08,\quad Y_{\text{thresh.}} = Z_{\text{thresh.}} = 10^{-4},
-$$
-where the higher value of $X_{\text{thresh.}}$ helped select for more sustainable fishing strategies (for both the DRL and constant mortality approaches).
-
-# Results
-
-We evaluated each of the three control approaches (MSY constant mortality, 80% MSY constant mortality and DRL) by running 50 episodes using the respective control rule.
-In Fig. \ref{fig:timeseries} we show the time series of one such repetition.
-The left panel uses an MSY constant mortality control $F^*X$, the middle panel uses $0.8 F^* X$, while the right panel is controlled by the trained DRL agent.
-We see that that optimal constant mortality control leads to a near-extinction event at around $t=150$.
-The constant mortality policy at $0.8 F^*$ is already conservative enough to support a sustainable path.
-There, the population levels are maintained roughly in a stationary state.
-Finally, the DRL time series shows an oscillatory behavior, with a slight dephase between the $Y$ and $Z$ curves, while the $X$ oscillations are roughly anticorrelated with $Z$.
-
 ```{r}
-#| label: timeseries 
-#| fig.cap: "A sample time series for the three fish populations, with the system being controlled, respectively, by the MSY optimal mortality, by $80$ percent of the MSY optimal mortality, and by DRL."
-#| out.width: "6.5in"
-policy_labels <- list("msy" = "MSY Const. Mortality", "tac"= "80% MSY Const. Mortality", "ppo"= "DRL")
-policy_labeller <- function(variable,value){
-  return(policy_labels[value])
-}
-msy_tac_ppo |> 
-    group_by(rep, policy) |>
-    filter(rep == 6) |>
-    ggplot(aes(t, abundance, col = species)) + 
-        geom_line()  + 
-        scale_color_manual(values=c('palegreen3', 'coral1', 'deepskyblue4')) +
-        facet_wrap(~factor(policy, levels = c("msy", "tac", "ppo")), labeller = policy_labeller ) +
-        labs(y = "abundance", x = "time") +
-        theme(aspect.ratio = 2/3)
-```
-
-This time series is representative of the patterns observed in the data (see App. \ref{app:tbd}).
-The episode reward and episode length statistics are shown in Fig. \ref{fig:rew_len_fig}.
-There, we highlight two things:
-First, that the tendency towards early episode ends is fixed by the conservative approach of fishing at 80% of the optimal mortality rate.
-(In App. \ref{app:tbd} we discuss a warning about this approach, however: the arbitrarily chosen percentage itself lives on a ‘‘knife's edge.’’
-More concretely, choosing a higher percentage, say 90%, can switch the behaviour of the controlled system and give rise to the proliferation of early episode ends.)
-Second, that while the conservative 80% approach has a stable behavior, it consistently produces smaller harvests than a DRL generated policy.
-
-```{r} 
-#| label: rew_len_fig
-#| fig.cap: "Bar charts for the episode lengths (top panel) and episode total harvests (bottom panel) for 50 episodes." 
-#| out.width: "6.5in"
-p1 <- msy_tac_ppo |> 
-    group_by(rep, policy) |>
-    filter(t==max(t), species == "X", rep < 50) |>
-    ggplot(aes(rep, t)) + 
+all_sim |> 
+    mutate(across(type, factor, levels = c("Opt const mortality", "80% opt const mortality", 
+                                           "DRL"))) |>
+    group_by(rep, type) |>
+    filter(t==max(t), species=="X", rep < 10) |>
+    ggplot(aes(rep,t)) + 
         geom_col()  + 
-      facet_wrap(~factor(policy, levels = c("msy", "tac", "ppo")), labeller = policy_labeller ) +
+        facet_wrap(~type) + 
         labs(y = "episode length", x = "repetition") +
         theme(axis.text.x = element_blank())
-
-p2 <- msy_tac_ppo |> 
-    group_by(rep, policy) |>
-    filter(t==max(t), species == "X", rep < 50) |>
-    ggplot(aes(rep, reward)) + 
-      geom_col() +
-      facet_wrap(~factor(policy, levels = c("msy", "tac", "ppo")), labeller = policy_labeller ) +
-      labs(y = "episode harvest", x = "repetition") +
-      theme(axis.text.x = element_blank())
-
-(p1 / p2) + plot_annotation(tag_levels = 'A')
 ```
-
-To understand the DRL-derived policy, Fig. \ref{fig:action-X} depicts the harvest as a function of $X$.
-Specifically, for each point $(x,y,z)$ visited by one of the aforementioned 50 evaluation episodes, we evaluate the harvest collected by the DRL agent and we subsequently plot the scatter data as a function of the projection $x$ onto the $X$ axis.
-In panel A we have used color to add the a $Y$-dependence, while in panel B we did the same with $Z$-dependence.
-This shows that the $Z$ and $Y$ dependence of the policy is rather weak and in fact it has roughly the same shape as an escapement policy.
-
-Fig. \ref{action-X} additionally shows (as a red line) the optimal escapement policy.
-This policy was computed in a manner similar to how the MSY constant mortality was computed: for each escapement in a grid of 100 values between 0 and 1, 50 episodes were run using the corresponding escapement policy.
-The value with the best mean episode reward was chosen as the optimal escapement value.
-
-```{r}
-#| label: action-X
-#| fig.cap: "TBD"
-#| out.width: "6.5in"
-opt_escapement_harvest <- function(X) case_when(
-  X > best_e ~ X - best_e,
-  X <= best_e ~ 0
-)
-
-py <- ppo_df |>
-  mutate(X = X+1, Y = Y+1, Z = Z+1, harvest = action*X) |>
-  ggplot(aes(X, harvest, color = Y)) +
-  geom_point() +
-  theme(aspect.ratio = 1/1) +
-  stat_function(fun = opt_escapement_harvest, col = 'red')
-
-pz <- ppo_df |>
-  mutate(X = X+1, Y = Y+1, Z = Z+1, harvest = action*X) |>
-  ggplot(aes(X, harvest, color = Z)) +
-  geom_point() +
-  theme(aspect.ratio = 1/1) +
-  stat_function(fun = opt_escapement_harvest, col = 'red')
-
-(py | pz) + plot_annotation(tag_levels = c('A'))
-```
-
-In Fig. \ref{fig:state-space}, we perform a back to back visualization of all the policies evaluated up to now---MSY constant mortality, 80% MSY constant mortality, DRL and escapement.
-The policies are computed at each state along the paths they traverse in the 50 evaluation rounds, and the values are projected on the $X$-$Z$ plane for visualization.
-This figure highlights the similarity between the policy found by the DRL agent and the optimal escapement policy.
-
-Notice the anti-correlation between $X$ and $Y$ along the paths, this is especially prominent for the constant mortality plots.
-This anti-correlation is, furthermore, observed in the time series plots (see Fig. \ref{fig:timeseries} and App. \ref{app:tbd}).
-
-```{r}
-#| label: state-space
-#| fig.cap: "Policy heatmaps.
-Each policy, say, $\\pi$, was evaluated at points $(x,y,z)$ visited during any of the 50 evaluation rounds.
-We projected this data down to the $X$-$Y$ plane---i.e.
-we plotted the heatmap $\\pi(x,y,z)$ as a function of $x$ and $y$."
-#| out.width: "6.5in"
-i_tac <- which.min(abs(actions - best_action * 0.8))
-
-# color pallette limits
-library(RColorBrewer)
-myPalette <- colorRampPalette(rev(brewer.pal(11, "Spectral")))
-sc <- scale_colour_gradientn(colours = myPalette(100), limits=c(0,0.25))
-
-pmsy <- msy_df |>
-  mutate(X = X+1, Y = Y+1, Z = Z+1, harvest = action * X) |>
-  filter(action == best_action) |>
-  ggplot(aes(X, Y, color = harvest)) +
-  geom_point() + sc +
-  xlim(0,1) +
-  ylim(0,0.85) +
-  ggtitle("Opt. Const. Mortality")
-
-ptac <- msy_df |>
-  mutate(X = X+1, Y = Y+1, Z = Z+1, harvest = action * X) |>
-  filter(action == actions[[i_tac]]) |>
-  ggplot(aes(X, Y, color = harvest)) +
-  geom_point() + sc +
-  xlim(0,1) +
-  ylim(0,0.85) +
-  ggtitle("80% Opt. Const. Mortality")
-
-pppo <- ppo_df |>
-  mutate(X = X+1, Y = Y+1, Z = Z+1, harvest = action * X) |>
-  ggplot(aes(X, Y, color = harvest)) +
-  geom_point() + sc +
-  xlim(0,1) +
-  ylim(0,0.85) +
-  ggtitle("DRL")
-
-pesc <- opt_escapement |>
-  mutate(X = X+1, Y = Y+1, Z = Z+1, 
-        harvest = case_when(
-          X - escapement > 0 ~ X - escapement,
-          X - escapement <= 0 ~  0)
-         ) |> 
-  ggplot(aes(X, Y, color = harvest)) +
-  geom_point() + sc +
-  xlim(0,1) +
-  ylim(0,0.85) +
-  ggtitle("Opt. Const. Escapement")
-
-(
-  (pmsy | ptac) / (pppo | pesc) +
-  plot_layout(guides = "collect") & theme(legend.position = 'right')
-) + 
-  plot_annotation(tag_levels = c('A'))
-```
-
-# Discussion
-
-Fisheries are complex ecosystems, with interactions between species leading to highly non-linear dynamics.
-While current models for population estimation include many important aspects of this complexity, it is still common to use simplified dynamical models in order to guide decision making.
-This provides very interpretable solutions, such as constant mortality policies.
-There is a drawback here, however: due to the simplicity of these dynamical models, the policies might not respond effectively in many situations---situations where the predictions of these simple models deviates considerably from reality.
-Because of this, policies such as MSY have faced pushback and are believed to have contributed to the depletion of fish stocks.
-
-We propose an alternative approach to the problem of fishery control: to use a more expressive -- albeit more complicated -- dynamical model to guide decision making.
-Furthermore, rather than computing the optimal control policy for the model -- something that is impossible in practice for complex systems -- we use deep reinforcement learning to obtain a ‘‘pretty darn good’’ policy.
-
-In this paper we have exemplified this approach by applying it to the control of a three-species fishery model.
-In this example, we have shown that the DRL agent finds an approximately *bang-bang* type of solution.
-While bang-bang solutions are known to be optimal for a wide variety of single-species fishery control problems [@alexandru], there is no such proof of optimality for general multi-species models.
-In this sense this example is rather surprising and reassuring: while no general guarantee points at the optimality of bang-bang solutions for multi-species model, the existence of one such example provides a benchmark for the use of DRL in fishery control.
-
-One intuition for why a constant escapement policy performs so well in such a model comes from the anticorrelation between $X$ and $Y$ noted in the results section.
-We have observed in a variety of policies, both classical as well as some obtained by training DRL agents.
-In all of these, the pattern has been present.
-In fact, ‘‘crashes’’ in the $X$ population are almost invariably shortly preceeded by dramatic increases in $Y$.
-We hypothesize that this effectively leads to a dimensionality reduction: once the value of $X$ is known, along a path traced by the control system, then $Y$ may be roughly predicted.
-A similar, although weaker, anticorrelation is also observed between $X$ and $Z$.
-These two informal observations point at an effective simplification of the prediction process: the values of $Y$ and $Z$ are much less impactful to policy decisions than the value of $X$ itself.
-We have not yet pushed further to establish a more precise picture of this dynamical simplification.
-
-This way, this manuscript functions as a first step in the process of incorporating DRL into fishery decision making.
-Importantly, DRL is equipped with tools to deal with more complicated models which might require a more refined approach to control them.
-For example *long short-term memories (LSTMs)* may be used to incorporate not only $N_t$ but also $\mathrm{d}N_t/\mathrm{d}t$ into policy decisions---while this additional degree of freedom would make analytical control theory approaches unwieldy, it may be easily and cost-effectively implemented in DRL.
-Similarly, *curriculum learning* could be incorporated to deal with phenomena such as parametric noise or parametric drift [@openai].
-
-This way, DRL provides a framework flexible enough to capture the intricate population dynamics that are relevant in fishery management.
-We have shown that highly non-linear systems may give rise to simplified and good policies and that these may be found via reinforcement learning.
-Moreover, the framework allows for the possibility that these simplifications do not in fact happen, and that good harvest policies are highly non-linear, multi-parameter functions.
-
-Conversely, our analysis suggests that, for high dimensional systems, bang-bang policies lead to increased sustainability and long-term profits as compared to constant mortality strategies.
-Similarly, in App. \ref{app:tbd} we investigate several generalizations of the constant effort approach. 
-For example, we examine strategies with a ramp-down from constant effort to zero effort at low populations.
-These, too, yield subpar results as compared to escapement.
-This seems to indicate that the ‘‘rest periods’’ so prominent in both the DRL solution and in the optimal bang-bang policy -- periods of several time steps with no fishing at all -- are important to maintain the sustainability of the fishery and, additionally, increase long-term profits.
-
-## Acknowledgements
-
-The title of this piece references a mathematical biology workshop at NIMBioS
-organized by Paul Armsworth, Alan Hastings, Megan Donahue, and Carl Towes in
-2011 which first sought to emphasize 'pretty darn good' control solutions to
-more realistic problems over optimal control to idealized ones. 
-This material is based upon work supported by the National Science Foundation under Grant No. DBI-1942280.
-
-# References

### 2023-01-27 manuscript/manuscript.Rmd
carl :pen: intro
Note: some old text has been moved to code/scratch.Rmd while we decide we want to eliminate it.

--- 2023-01-27 manuscript/manuscript.Rmd
@@
 
-<!--
-Existing methods for optimal control methods struggle to deal with the complexity commonly encountered in real-world systems, including dimensionality, process error, model bias and heterogeneity. 
-Instead of tackling these complexities directly, researchers have typically sought to find exact optimal solutions to simplified models of the processes in question. 
-When is the optimal solution to a very approximate, stylized model better than an approximate solution to a more accurate model? 
-This question has largely gone unanswered owing to the difficulty of finding even approximate solutions in the case of complex models.  
-Our approach draws on recent algorithmic and computational advances in deep reinforcement learning. 
-These methods have hitherto focused on problems in games or robotic mechanics, which operate under precisely known rules. 
-We demonstrate the ability for novel algorithms using deep neural networks to successfully approximate such solutions (the "policy function" or control rule) without knowing or ever attempting to infer a model for the process itself.
-This powerful new technique lets us finally begin to answer the question.
-We show that in many but not all cases, the optimal policy for a carefully chosen over-simplified model can still out-perform these novel algorithms trained to find approximate solutions to simulations of a realistically complex system.
-This illustrates the promise of combining complex, realistic simulations with emerging machine-learning techniques to improve our understanding and management of ecological systems.
--->
+Much effort has been spent grappling with the complexity of our natural world in contrast to relative simplicity of the models we use to understand it.
+Now, as we grapple with the manifold consequences of human impacts on that world, we are faced with difficult questions of figuring out how best to respond when our knowledge and models fall so far short of the complexity our actions have disrupted.
+Heroic amounts of data and computation are being brought to bear on developing better, more realistic models of our environments and ecosystems.
+However, these rich models are applied to highly oversimplified descriptions of potential actions they seek to inform.
+For instance, Global Circulation Models such as HadOM3 model earth's climate using 1.5M variables, while the comparably vast potential action space of is modeled as one of six emissions scenarios from the IPCC.
+Even as our research community develops simulations of the natural world that fit only in supercomputers, we analyze a space of policies that would fit on index cards.
+Similar combinations of rich process models and highly simplified decision models (often not even given the status of 'model') are common.
+Modeling the potential action space as one of a handful of discrete scenarios is sometimes a well justified acknowledgement of the constraints of real-world decision-makers -- 
+and may seem to reflect a natural division of responsibilities between 'scientists' modeling the 'natural processes' and policy-makers who must make the decisions.
+But often this simplification of decision choices is simply mathematically or conceptually convenient.
+<!-- there are questions of policy scale hiding here -- e.g.
+local regulators of a specific industry obviously face a different action space than national legislatures or international diplomats -->
+This reflects trade-offs between tractablity and complexity at the basis of any mathematical modeling -- if we make both the state space and action space too realistic, the problem of finding the best sequence of actions quickly becomes intractable.
+However, emerging data-driven methods from machine learning offer a new choice -- algorithms that can find good strategies in such problems, but at the cost of opacity.
 
-Ecological management may be understood as an optimal control problem - a manager must determine a sequence of actions (a "policy") to optimize some desired objective(s) and constraints (e.g.
-conservation goals, natural resource extraction, economic and social constraints.)
-Despite the complexity inherent in natural ecosystems, analytically tractable models have long played an essential role in their understanding and management.
-...
+In this paper, we focus on a well-developed application of model-based management of the natural world that has long illustrated the trade-offs between model complexity and policy complexity: the management of marine fisheries.
+Fisheries management is both an important issue to itself and a rich & frequent test-bed of ecological management more generally.
+Fisheries are an essential natural resource that provide the primary source of protein for one in every four humans, and have faced widely documented declines due to over-fishing [@Worm2006, @Costello2016].
+In the start of this literature we find the roots of both the fields of _ecosystem management_ and _natural resource economics_.
+Both fields might trace their origins to notion of maximum sustainable yield (MSY), introduced independently by a fisheries ecologist @Schaefer1954 and the economist @Gordon1954 in the same year.
+From this shared origin, each field would depart from the simplifying assumptions of the Gordon-Schaefer model in different ways, leading to different techniques for deriving policies from models.The heart of the management problem is easily understood: a manager seeks to set quotas on fishing that will ensure the long-term profitability and sustainability of the industry.
+Mathematical approaches developed over the past century may be roughly divided between two camps: (A), ecologists, focused on ever more realistic models of the biological processes of growth and recruitment of fish while considering relatively stylized suite of potential management strategies, and (B) economists, focused on far more stylized models of the ecology while exploring a far less constrained set of possible policies.
+The economist's approach can characterized by the mathematics of a Markov decision process [MDP @Clark1974; @Clark1990; @Marescot2013], in which the decision-maker must observe the stock each year and recommend a possible action.
+Note that the policy space that much be searched is exponentially large -- for a management horizon of T decisions and a space of N actions, the number of possible policies is N^T.
+In contrast, fisheries ecologists and ecosystem management typically search a space of policies that does not scale with the time horizon.
+Under methods such as "Management Strategy Evaluation" (MSE, [@Punt2016]) a manager identifies a candidate set of "strategies" a priori, and then compares the performance of each strategy over a suite of simulations to determine which strategy gives the best outcome (i.e.
+best expected utility).
+This approach is far more amenable to complex simulations of fisheries dynamics and more closely corresponds to how most marine fishing quotas are managed today in the United States [see stock assessments documented in @RAMLegacyDB].
+<!-- Do we say here that these strategies retain the 'constant mortality' intuition of @Schaefer? -->
 
-<!-- Why fisheries-focused examples -->
-Global fisheries management has long been a crucible for both the development of theory and best practices in actual management for many reasons.
-Fisheries are an essential natural resource that provide the primary source of protein for one in every four humans, and have faced widely documented declines due to overfishing [@Worm2006, @Costello2016].
-Because of this importance, fisheries also represent some of the longest and best studied timeseries of population dynamics available to ecologists, as well as a correspondingly rich area for the development of theory and practice.
-Methods developed in the study and management of fisheries are frequently applied to other areas of ecological management, from concepts such as maximum sustainable yield [], optimal control under observation uncertainty [@Memarzadeh2019], to adaptive management [@Walters], management strategy evaluation [@Punt], to the studies of conservation policy and international negotiations [@].
+Recent advances in machine learning may now once again bridge these approaches, while also bringing new challenges of their own. 
+Advances in data-driven methods have allowed these models to evolve into ever more complex and realistic simulations used in fisheries management, where models with over 100 parameters are not uncommon [@RamLegacyDB].
+Constrained by computational limits, MDP approaches have been intractable on suitably realistic models and largely confined to more academic applications [@Costello2016].
+However, advances _Deep Reinforcement Learning_, (RL) a sub-field of machine learning, have recently demonstrated remarkable performance in a range of such MDP problems, from video games [@atari] to fusion reactions [@fusion] to the remarkable dialog abilities of ChatGPT [@chatgpt].
 
-<!-- A sweeping summary of fisheries decision theory, don't bite me-->
-To understand the importance of a potential application of novel data-driven methods to the real world, it is most instructive to have a clearer picture of how models are currently used to inform ecological management in fisheries.
-In any issue as large and contentious as fisheries management, it may be far more difficult to provide a generalization  about how things actually work than to make sweeping recommendations about how it ought to.
-All the same, it is our firm belief that any researcher seeking to do the latter should first attempt to understand the former, so we beg the forbearance of any specialists in the practice as we seek to provide a concise overview of fisheries management today.
-Here, we focus specifically on practices in data rich, 'well-managed' fisheries, which, despite some issues, may be among the best examples of ecological management at continental scale that is largely driven by scientific processes, models and data.
-
-<!-- Fisheries population models are rich but decisions aren't, mostly constant mortality -->
-While fisheries stock assessment models are quite rich, the choice of possible strategies remains motivated by intution from much simpler, analytically tractable one-dimensional models.
-This fact is not always obvious at cursory inspection, and can be met with cynicism even among professional fisheries managers -- after all, typical stock assessment models have over 100 parameters (e.g.
-see the stock assessments referenced from in @RAMLegacy) and best practices such as management strategy evaluation (MSE) evaluate and select potential policies by comparing across large and diverse simulations [e.g.
-@Punt2016].
-However, the space of policies considered under such approaches is necessarily far smaller than the set of all possible action sequences.
-Instead, fisheries management approaches such as MSE rely on theory based on more tractable models to restrict the search space of possible policies.  
 Specifically, most fisheries policies are framed in terms of a *constant mortality* policy, (sometimes a simple piece-wise linear function, i.e.
@@
 ```
-
-The intuition for a constant mortality strategy is simple and well established
-
-Constant mortality policies have a strong basis in theory as well as practical justifications.
-
-<!-- In contrast to constant mortality, the possible decision space is huge-->
-The management of any ecological system can be framed in terms of a Sequential Decision Process - 
-the manager observes the ecosystem state ($x_t$) and decides on action ($a_t$) to take (including potentially no action), potentially receiving any costs or benefits associated with that action and and state (reward $r_t$  $r_t = U(x_t,a_t)$),
-while the system evolves into a new state in response $x_{t+1} = f(x_t, a_t)$.
-When the probability of the future state can be determined by the current state and action, we say the system is Markovian and the process a Markov Decision Process (MDP).
-As in a game of chess, the space of possible sequences of actions a manager might choose over any realization of the Markov process is enormous.
-When the action and state space are discrete and sufficiently small, computational approaches such as dynamic programming may be used to guarantee an exhaustive search for a single optimal policy.
-Dynamic programming scales very poorly to larger state and action spaces.
-
-Deep reinforcement learning has opened up increasingly compelling alternative
-
-Effort $Q = E X_t$
-
-Fisheries policies are typically implemented in terms of *catch quotas*, i.e. a limit to the number of metric tonnes of fish that can be caught during a given season.
-
-and/or *effort controls*, restrictions on the fishing effort -- in numbers of boats, time of year, or days at sea
-
-# Figures brainstorm
-
-Figure 0: contrast 1D vs 3D strategies (see commented out diagram).
 
@@
 
-<!-- Reality __
-  |        \
-  v         \
-1-D Model   More expressive model
-  |                  |
-  v                  v
-Optimal        Suboptimal, but
-solution   well-performing solution -->
-
+<!--
 Figure 1: 1-D and 3-D model conceptual figure. something about the objective / decision
@@
 - reward plot over time (comparing methods)
+-->
 
@@
 
+<!-- Tidy up names.  Let's call these: MSY-targeted Constant Mortality, 80%-MSY-targeted constant mortality, constant escapement -->
 There are several strategies that have been used to manage fisheries: *escapement*, *maximum sustainable yield (MSY)*, *total allowable catch (TAC)*, among others.
-We collectively refere to these as *classical*, and will compare their performance to RL-based management strategies.
+We collectively refer to these as *classical*, and will compare their performance to RL-based management strategies.
 Classical strategies derive an optimal harvest policy using a simple model for the system dynamics.

### 2023-01-27 manuscript/manuscript.Rmd
:pen:

@@
 Recent advances in machine learning may now once again bridge these approaches, while also bringing new challenges of their own. 
-Advances in data-driven methods have allowed these models to evolve into ever more complex and realistic simulations used in fisheries management, where models with over 100 parameters are not uncommon [@RamLegacyDB].
+Novel data-driven methods have allowed these models to evolve into ever more complex and realistic simulations used in fisheries management, where models with over 100 parameters are not uncommon [@RamLegacyDB].
 Constrained by computational limits, MDP approaches have been intractable on suitably realistic models and largely confined to more academic applications [@Costello2016].
 However, advances _Deep Reinforcement Learning_, (RL) a sub-field of machine learning, have recently demonstrated remarkable performance in a range of such MDP problems, from video games [@atari] to fusion reactions [@fusion] to the remarkable dialog abilities of ChatGPT [@chatgpt].
+RL methods also bring many challenges of their own: being notoriously difficult to train and evaluate, immense computational costs, and frequent challenges with transfer-ability or reproducibility. 
+A review of all these issues is beyond our scope but can be found elsewhere [@rl_intro; @millie_AM-RL].
+Here, though, we will focus on the issue of opacity and interpretability raised by these methods.
+In contrast with optimization algorithms currently used in either ecosystem management or resource economics, RL algorithms have no guarantees of or metrics for convergence to an optimal solution.
+We can only assess the performance of these black box methods relative to alternatives.
 

### 2023-01-27 manuscript/manuscript.Rmd
:pen:

@@
 Constrained by computational limits, MDP approaches have been intractable on suitably realistic models and largely confined to more academic applications [@Costello2016].
-However, advances _Deep Reinforcement Learning_, (RL) a sub-field of machine learning, have recently demonstrated remarkable performance in a range of such MDP problems, from video games [@atari] to fusion reactions [@fusion] to the remarkable dialog abilities of ChatGPT [@chatgpt].
+However, advances _Deep Reinforcement Learning_, (DRL) a sub-field of machine learning, have recently demonstrated remarkable performance in a range of such MDP problems, from video games [@atari] to fusion reactions [@fusion] to the remarkable dialog abilities of ChatGPT [@chatgpt].
 RL methods also bring many challenges of their own: being notoriously difficult to train and evaluate, immense computational costs, and frequent challenges with transfer-ability or reproducibility. 
@@
 In contrast with optimization algorithms currently used in either ecosystem management or resource economics, RL algorithms have no guarantees of or metrics for convergence to an optimal solution.
-We can only assess the performance of these black box methods relative to alternatives.
+In general, one can only assess the performance of these black box methods relative to alternatives.
 
+<!--NEED some kind of map paragraph here saying what's ahead, what we show  --> 
+We compare the performance of a DRL solution to that of a management strategy evaluation across potential 'constant mortality' policies.
+
+<!--
 Specifically, most fisheries policies are framed in terms of a *constant mortality* policy, (sometimes a simple piece-wise linear function, i.e.
 a single linear segment, is introduced as a precaution a very low stock sizes), where fishing 'mortality' is defined as the ratio of harvest (catch) to stock size, $F = H_t/x_t$.
+-->
 

### 2023-01-27 manuscript/manuscript.Rmd
Add files via upload
Shortened 1-d model section of manuscript and referenced textbox

--- 2023-01-27 manuscript/manuscript.Rmd
@@
 
+<!--
+Existing methods for optimal control methods struggle to deal with the complexity commonly encountered in real-world systems, including dimensionality, process error, model bias and heterogeneity. 
+Instead of tackling these complexities directly, researchers have typically sought to find exact optimal solutions to simplified models of the processes in question. 
+When is the optimal solution to a very approximate, stylized model better than an approximate solution to a more accurate model? 
+This question has largely gone unanswered owing to the difficulty of finding even approximate solutions in the case of complex models.  
+Our approach draws on recent algorithmic and computational advances in deep reinforcement learning. 
+These methods have hitherto focused on problems in games or robotic mechanics, which operate under precisely known rules. 
+We demonstrate the ability for novel algorithms using deep neural networks to successfully approximate such solutions (the "policy function" or control rule) without knowing or ever attempting to infer a model for the process itself.
+This powerful new technique lets us finally begin to answer the question.
+We show that in many but not all cases, the optimal policy for a carefully chosen over-simplified model can still out-perform these novel algorithms trained to find approximate solutions to simulations of a realistically complex system.
+This illustrates the promise of combining complex, realistic simulations with emerging machine-learning techniques to improve our understanding and management of ecological systems.
+-->
+
 Much effort has been spent grappling with the complexity of our natural world in contrast to relative simplicity of the models we use to understand it.
 Now, as we grapple with the manifold consequences of human impacts on that world, we are faced with difficult questions of figuring out how best to respond when our knowledge and models fall so far short of the complexity our actions have disrupted.
-Heroic amounts of data and computation are being brought to bear on developing better, more realistic models of our environments and ecosystems.
-However, these rich models are applied to highly oversimplified descriptions of potential actions they seek to inform.
-For instance, Global Circulation Models such as HadOM3 model earth's climate using 1.5M variables, while the comparably vast potential action space of is modeled as one of six emissions scenarios from the IPCC.
+Heroic amounts of data and computation are being brought to bear on developing better, more complex, more realistic models of our environments and ecosystems, yet these models remain disconnected or oversimplify the space of actions they seek to inform.
+In this paper we explore a different tack. 
+Here we explore the potential of a emerging data-driven machine learning technique, not to better capture the complexity of the natural processes, but rather, to embrace the complexity of the potential _policies_ that might be used to manage them.
 Even as our research community develops simulations of the natural world that fit only in supercomputers, we analyze a space of policies that would fit on index cards.
-Similar combinations of rich process models and highly simplified decision models (often not even given the status of 'model') are common.
-Modeling the potential action space as one of a handful of discrete scenarios is sometimes a well justified acknowledgement of the constraints of real-world decision-makers -- 
-and may seem to reflect a natural division of responsibilities between 'scientists' modeling the 'natural processes' and policy-makers who must make the decisions.
-But often this simplification of decision choices is simply mathematically or conceptually convenient.
-<!-- there are questions of policy scale hiding here -- e.g.
-local regulators of a specific industry obviously face a different action space than national legislatures or international diplomats -->
-This reflects trade-offs between tractablity and complexity at the basis of any mathematical modeling -- if we make both the state space and action space too realistic, the problem of finding the best sequence of actions quickly becomes intractable.
-However, emerging data-driven methods from machine learning offer a new choice -- algorithms that can find good strategies in such problems, but at the cost of opacity.
 
-In this paper, we focus on a well-developed application of model-based management of the natural world that has long illustrated the trade-offs between model complexity and policy complexity: the management of marine fisheries.
-Fisheries management is both an important issue to itself and a rich & frequent test-bed of ecological management more generally.
+<!-- Why fisheries-focused examples -- maybe we can cut/merge this with paragraph above -->
+There are myriad examples from climate scenarios to protected areas, we will illustrate this with a specific example in the management of marine fisheries, which is both an important issue to itself and a frequent crucible in which methods of ecological management have been fired before extending to other fields, and where the space of possible policies is arguably already more richly developed than most.
 Fisheries are an essential natural resource that provide the primary source of protein for one in every four humans, and have faced widely documented declines due to over-fishing [@Worm2006, @Costello2016].
-In the start of this literature we find the roots of both the fields of _ecosystem management_ and _natural resource economics_.
-Both fields might trace their origins to notion of maximum sustainable yield (MSY), introduced independently by a fisheries ecologist @Schaefer1954 and the economist @Gordon1954 in the same year.
-From this shared origin, each field would depart from the simplifying assumptions of the Gordon-Schaefer model in different ways, leading to different techniques for deriving policies from models.The heart of the management problem is easily understood: a manager seeks to set quotas on fishing that will ensure the long-term profitability and sustainability of the industry.
-Mathematical approaches developed over the past century may be roughly divided between two camps: (A), ecologists, focused on ever more realistic models of the biological processes of growth and recruitment of fish while considering relatively stylized suite of potential management strategies, and (B) economists, focused on far more stylized models of the ecology while exploring a far less constrained set of possible policies.
-The economist's approach can characterized by the mathematics of a Markov decision process [MDP @Clark1974; @Clark1990; @Marescot2013], in which the decision-maker must observe the stock each year and recommend a possible action.
-Note that the policy space that much be searched is exponentially large -- for a management horizon of T decisions and a space of N actions, the number of possible policies is N^T.
-In contrast, fisheries ecologists and ecosystem management typically search a space of policies that does not scale with the time horizon.
-Under methods such as "Management Strategy Evaluation" (MSE, [@Punt2016]) a manager identifies a candidate set of "strategies" a priori, and then compares the performance of each strategy over a suite of simulations to determine which strategy gives the best outcome (i.e.
-best expected utility).
-This approach is far more amenable to complex simulations of fisheries dynamics and more closely corresponds to how most marine fishing quotas are managed today in the United States [see stock assessments documented in @RAMLegacyDB].
-<!-- Do we say here that these strategies retain the 'constant mortality' intuition of @Schaefer? -->
+Because of this importance, fisheries also represent some of the longest and best studied timeseries of population dynamics available to ecologists, as well as a correspondingly rich area for the development of theory and practice.
+Methods developed in the study and management of fisheries are frequently applied to other areas of ecological management, from concepts such as maximum sustainable yield [], optimal control under observation uncertainty [@Memarzadeh2019], to adaptive management [@Walters], management strategy evaluation [@Punt], to the studies of conservation policy and international negotiations [@].
 
-Recent advances in machine learning may now once again bridge these approaches, while also bringing new challenges of their own. 
-Advances in data-driven methods have allowed these models to evolve into ever more complex and realistic simulations used in fisheries management, where models with over 100 parameters are not uncommon [@RamLegacyDB].
-Constrained by computational limits, MDP approaches have been intractable on suitably realistic models and largely confined to more academic applications [@Costello2016].
-However, advances _Deep Reinforcement Learning_, (RL) a sub-field of machine learning, have recently demonstrated remarkable performance in a range of such MDP problems, from video games [@atari] to fusion reactions [@fusion] to the remarkable dialog abilities of ChatGPT [@chatgpt].
+Mathematically, the ecological management may be understood as an optimal control problem - a manager must determine a sequence of actions (a "policy") to optimize some desired objective(s) and constraints (e.g.
+conservation goals, natural resource extraction, economic and social constraints.)
+Given an appropriate representation of the underlying mechanistic processes involved, such sequential decision problems may be posed more precisely as Markov Decision Process [MDP, @Marescot2013; @Memarzadeh2019].
+Such problems are notoriously difficult owing not only to the complexity of those underlying mechanisms alone, but to the exponentially branching consequences the decision-maker's actions have upon that system -- a game of chess is not won by thinking only one move ahead at a time.
+Thus, while MDP problems have been studied in a conservation context since they were first introduced more than six decades ago, their application has been highly constrained to simple scenarios.
+Meanwhile, a class of machine learning techniques designed for MDP problems known as reinforcement learning (RL), has cracked open the door for exploring complex strategies in more realistic settings [@rl_intro].
 
+- This is not a tutorial or a sales pitch for RL - which brings at least as many difficulties as advantages -- including being hard to use, easy to get stuck on far from optimal performance, which itself may be hard to diagnose, as we have already discussed in our tutorial.
+The method
+
+has through the use of clever algorithms, deep neural networks, and massive amounts of data,
+
+<!-- A sweeping summary of fisheries decision theory, don't bite me-->
+To understand the importance of a potential application of novel data-driven methods to the real world, it is most instructive to have a clearer picture of how models are currently used to inform ecological management in fisheries.
+In any issue as large and contentious as fisheries management, it may be far more difficult to provide a generalization  about how things actually work than to make sweeping recommendations about how it ought to.
+All the same, it is our firm belief that any researcher seeking to do the latter should first attempt to understand the former, so we beg the forbearance of any specialists in the practice as we seek to provide a concise overview of fisheries management today.
+Here, we focus specifically on practices in data rich, 'well-managed' fisheries, which, despite some issues, may be among the best examples of ecological management at continental scale that is largely driven by scientific processes, models and data.
+
+<!-- Fisheries population models are rich but decisions aren't, mostly constant mortality -->
+While fisheries stock assessment models are quite rich, the choice of possible strategies remains motivated by intution from much simpler, analytically tractable one-dimensional models.
+This fact is not always obvious at cursory inspection, and can be met with cynicism even among professional fisheries managers -- after all, typical stock assessment models have over 100 parameters (e.g.
+see the stock assessments referenced from in @RAMLegacy) and best practices such as management strategy evaluation (MSE) evaluate and select potential policies by comparing across large and diverse simulations [e.g.
+@Punt2016].
+However, the space of policies considered under such approaches is necessarily far smaller than the set of all possible action sequences.
+Instead, fisheries management approaches such as MSE rely on theory based on more tractable models to restrict the search space of possible policies.  
 Specifically, most fisheries policies are framed in terms of a *constant mortality* policy, (sometimes a simple piece-wise linear function, i.e.
@@
 ```
+
+The intuition for a constant mortality strategy is simple and well established
+
+Constant mortality policies have a strong basis in theory as well as practical justifications.
+
+<!-- In contrast to constant mortality, the possible decision space is huge-->
+The management of any ecological system can be framed in terms of a Sequential Decision Process - 
+the manager observes the ecosystem state ($x_t$) and decides on action ($a_t$) to take (including potentially no action), potentially receiving any costs or benefits associated with that action and and state (reward $r_t$  $r_t = U(x_t,a_t)$),
+while the system evolves into a new state in response $x_{t+1} = f(x_t, a_t)$.
+When the probability of the future state can be determined by the current state and action, we say the system is Markovian and the process a Markov Decision Process (MDP).
+As in a game of chess, the space of possible sequences of actions a manager might choose over any realization of the Markov process is enormous.
+When the action and state space are discrete and sufficiently small, computational approaches such as dynamic programming may be used to guarantee an exhaustive search for a single optimal policy.
+Dynamic programming scales very poorly to larger state and action spaces.
+
+Deep reinforcement learning has opened up increasingly compelling alternative
+
+Effort $Q = E X_t$
+
+Fisheries policies are typically implemented in terms of *catch quotas*, i.e. a limit to the number of metric tonnes of fish that can be caught during a given season.
+
+and/or *effort controls*, restrictions on the fishing effort -- in numbers of boats, time of year, or days at sea
+
+# Figures brainstorm
+
+Figure 0: contrast 1D vs 3D strategies (see commented out diagram).
 
@@
 
-<!--
+<!-- Reality __
+  |        \
+  v         \
+1-D Model   More expressive model
+  |                  |
+  v                  v
+Optimal        Suboptimal, but
+solution   well-performing solution -->
+
 Figure 1: 1-D and 3-D model conceptual figure. something about the objective / decision
@@
 - reward plot over time (comparing methods)
--->
 
@@
 
-Optimal control policies for fisheries are frequently based on 1-dimensional models, $n=1$.
+Optimal control policies for fisheries are frequently based on 1-dimensional models, $n=1$, as described in $Text$ $Box$ $1$.
 The most familiar model of $f(X)$ is that of *logistic growth*, for which
@@
 \end{align}
-
-The model this $f$ gives rise to is known as the Gordon-Schaefer model, after foundational work which first developed the concept of maximum sustainable yield, MSY, [@Gordon1954; @Schaefer1954], which remains a central concept in fisheries management.
-In this model, the maximum growth rate occurs at population size of $N = K/2$.
-This value is approached asymptotically, from any positive initial state $N_0 \in (0,K)$, under a constant mortality $F = r/2$ policy.
-At equilibrium this produces an asymptotic harvest of $rK/4$ per time step---the maximum yield sustainable indefinitely.
 
@@
 ```
-
-# Fishery management approaches
-
-<!-- Introduce the 1D and 3D models, and have Figure 1 already here. -->
-
-Here we review classical strategies for sustainable fishery management and we provide a birds-eye view of the alternative approach we propose.
-A summary of the contrast between the two strategies is given in Fig. \ref{fig:decision approaches}.
-
-## Classical approaches to sustainable fisheries {#sec:classical}
-
-<!-- Tidy up names.  Let's call these: MSY-targeted Constant Mortality, 80%-MSY-targeted constant mortality, constant escapement -->
-There are several strategies that have been used to manage fisheries: *escapement*, *maximum sustainable yield (MSY)*, *total allowable catch (TAC)*, among others.
-We collectively refer to these as *classical*, and will compare their performance to RL-based management strategies.
-Classical strategies derive an optimal harvest policy using a simple model for the system dynamics.
-While often complex models of the ecosystem are used to estimate the fish population of interest (i.e. $X_t$), the quota is set only taking into account the estimate $X_t$.
-The function determining this quota is derived from a simple dynamical model for the population (see text box 1).
-
-\textbf{Text box 1 (TBD formatting)}
-Let us consider a ubiquitous simple example: an MSY policy based on a logistic growth model for $X_t$ (see eqs. \eqref{eq:general model} and \eqref{eq:logistic}).
-
-Classical strategies have the common aspect of reducing the complex dynamics of the fishery ecosystem to a single equation governing the harvested population (say, $F$). 
-A common example is using a logistic growth equation,
-$$
-  F_{t+1} - F_t = r F_t( 1 - F_t / K) - h_t =: L(F_t) - h_t,
-$$
-where the interaction between $F$ and its environment is summarized to two parameters, the growth rate $r$, and the carrying capacity $K$.
-In the equation above, $h_t$ is the *harvest* at timestep $t$.
-The goal is to choose the harvest policy $h:F_t\mapsto h_t$ such that long-term profits are maximized.
-
-An advantage of one dimensional approaches is that the optimal policy is often known exactly, and, moreover, is intuitive.
-For example, in the logistic equation pointed out above, the maximal sustainable yield of the system is attained at $F = F_{MSY} := K/2$.
-The optimizer is an *escapement* policy:
-$$
-  h_t = \begin{cases}
-    F_t - K/2, \qquad &\text{if, } F_t > K/2\\
-    0, \qquad &\text{else.}
-  \end{cases}
-$$
-This corresponds to keeping the system at its optimal growth rate as much as possible.
-
-Escapement policies, or more generally *bang bang* policies, tend to be the optimal solution for these types of control problems.
-A drawback of these solutions, in the fishery context, is the presence of several timesteps with zero harvest which can arise.
-To mend this, certain suboptimal solutions have been constructed for fishery management.
-
-One ubiquitous solution is simply called maximum sustainable yield (MSY).
-It consists on letting $h(F) = rF/2$, so that
-$$
-  h(F_{MSY}) = L(F_{MSY}) = r K/4.
-$$
-That is, at the MSY biomass, the logistic growth of $F$ is cancelled exactly by the harvest.
-
-The MSY rule fixes the drawback in the escapement policy by having $h(F)>0$ for all $F>0$.
-It, however, has its own drawbacks.
-It is particularly sensitive to misestimates of the parameter $r$, as we will discuss in Sec. \ref{sec:tbd}.
-Due to this, similar but more conservative policies have been used.
-
-*Total allowable catch (TAC)* is one such policy.
-It consists on reducing the inclination of the line defined by $h(F)$ using a prefactor $\alpha$ in $h(F) = \alpha\times rF/2$. 
-Plausible examples are $\alpha = 0.8$ or $0.9$.
-Another common alternative is to have a prefactor $\alpha = \alpha(F)$ which decreases from one to zero as $F$ decreases below a threshold:
-$$
-  h(F) = \alpha(F)\times rF/2,
-$$
-where,
-$$
-  \alpha(F) = \begin{cases}
-  1, \qquad &\text{if } F > F_{\text{thresh.}},\\
-  F / F_{\text{thresh.}}, \qquad & \text{else}.
-  \end{cases}
-$$
-We call this policy *variable total allowable catch (VTAC)*.
-
-## Dangers: Noisy parameters, noisy observations and model inaccuracy
 

### 2023-01-27 manuscript/manuscript.Rmd
more discussion

--- 2023-01-27 manuscript/manuscript.Rmd
@@
 This provides very interpretable solutions, such as constant mortality policies.
-There is a drawback here, however: due to the simplicity of these dynamical models, the policies might not respond effectively in many situations---situations where the predictions of these simple models deviates considerably from reality.
-Because of this, policies such as MSY have faced pushback and are believed to have contributed to the depletion of fish stocks.
+There is a drawback here, however: due to the simplicity of these dynamical models, the policies might not respond effectively in many situations---situations where the predictions of these simple models deviate considerably from reality.
+Because of this, policies such as MSY have faced pushback and are believed to have contributed to the depletion of fish stocks [@tbd].
 
 We propose an alternative approach to the problem of fishery control: to use a more expressive -- albeit more complicated -- dynamical model to guide decision making.
-Furthermore, rather than computing the optimal control policy for the model -- something that is impossible in practice for complex systems -- we use deep reinforcement learning to obtain a ‘‘pretty darn good’’ policy.
+Furthermore, rather than computing the optimal control policy for the model (something that is impossible in practice for complex systems) we use deep reinforcement learning to obtain a ‘‘pretty darn good’’ policy.
+This policy is estimated in a *model-free* setting, i.e., the agent treats the dynamical model (e.g. eq. \eqref{eq:3d model stoch}) as a black box of input-output pairs.
+By not relying on precise knowledge of the model's parameter values, but rather just relying on input-output statistics, model-free approaches have gained traction in a variety of control theory settings [@rl_people].
 
-In this paper we have exemplified this approach by applying it to the control of a three-species fishery model.
-In this example, we have shown that the DRL agent finds an approximately *bang-bang* type of solution.
+We compare deep reinforcement learning-based policies against optimal constant mortality policies. 
+While the latter are MSY-inspired, they are optimized in a model-free way as well: the optimal constant mortality rate is computed empirically from simulated data.
+
+Because of the simplicity of the constant mortality ansatz, the optimal such policy may be easily estimated (see Figure \ref{fig:tuning}).
+In contrast, DRL optimizes over the more expressive -- and more complicated -- family of policies parametrized by a neural network.
+Because of this, our DRL approach uses a variant of *gradient ascent* (the proximal policy optimization algorithm) to obtain a good candidate policy.
+
+We have shown that the DRL agent finds an approximately *bang-bang* type of solution (see Textbox 1 for the definition of bang-bang solutions and Figs.
+\ref{fig:action-X} and \ref{fig:state-space} for the result).
 While bang-bang solutions are known to be optimal for a wide variety of single-species fishery control problems [@alexandru], there is no such proof of optimality for general multi-species models.
@@
 
-This way, this manuscript functions as a first step in the process of incorporating DRL into fishery decision making.
-Importantly, DRL is equipped with tools to deal with more complicated models which might require a more refined approach to control them.
+This manuscript functions as a first step in the process of incorporating DRL into fishery decision making.
+We have shown that highly non-linear systems may give rise to simplified and good policies and that these may be found via reinforcement learning.
+Moreover, the framework allows for the possibility that these simplifications do not in fact happen, and that good harvest policies are highly non-linear, multi-parameter functions.
+
+In this regard, DRL provides a framework flexible enough to capture the intricate population dynamics that are relevant in fishery management.
+Importantly, DRL is equipped with tools to deal with complex models for which simple policies might yield unsatisfactory results (and thus more refined techniques are needed to control them).
 For example *long short-term memories (LSTMs)* may be used to incorporate not only $N_t$ but also $\mathrm{d}N_t/\mathrm{d}t$ into policy decisions---while this additional degree of freedom would make analytical control theory approaches unwieldy, it may be easily and cost-effectively implemented in DRL.
@@
 
-This way, DRL provides a framework flexible enough to capture the intricate population dynamics that are relevant in fishery management.
-We have shown that highly non-linear systems may give rise to simplified and good policies and that these may be found via reinforcement learning.
-Moreover, the framework allows for the possibility that these simplifications do not in fact happen, and that good harvest policies are highly non-linear, multi-parameter functions.
-
-Conversely, our analysis suggests that, for high dimensional systems, bang-bang policies lead to increased sustainability and long-term profits as compared to constant mortality strategies.
-Similarly, in App. \ref{app:tbd} we investigate several generalizations of the constant effort approach. 
-For example, we examine strategies with a ramp-down from constant effort to zero effort at low populations.
-These, too, yield subpar results as compared to escapement.
-This seems to indicate that the ‘‘rest periods’’ so prominent in both the DRL solution and in the optimal bang-bang policy -- periods of several time steps with no fishing at all -- are important to maintain the sustainability of the fishery and, additionally, increase long-term profits.
+Finally, in our study case, we have shown that the ‘‘rest periods’’ so prominent in both the DRL solution and in the optimal bang-bang policy -- periods of several time steps with no fishing at all -- may be important to maintain the sustainability of the fishery and to increase its long-term profits.
+This hypothesis could be explored in more generality as it might apply to a wide variety of models with different optimal control strategies.
+In App. \ref{app:tbd} we provide some elementary support for this hypothesis by examining the performance of several ‘‘rule of thumb’’ policies based on conservative corrections to MSY.
 

### 2023-01-27 manuscript/figures/classical_fisheries_textbox.Rmd

@@
 
-\begin{tcolorbox}[colback=red!5!white,colframe=red!75!black,title=\large Classical approaches to sustainable fisheries]
+\begin{tcolorbox}[parbox=false,colback=red!5!white,colframe=red!75!black,title=\large Text Box 1: Classical approaches to sustainable fisheries]
 
@@
 
-A common example is the surplus production model that assumes logistic population growth in the absence of exploitation [@Schaefer1954]:
+A common example is the surplus production model that assumes logistic population growth in the absence of exploitation (Gordon, 1954; Schaefer, 1954):
 
@@
 An advantage of one dimensional approaches is that the optimal policy is intuitive and often known exactly.
-For example, in the logistic equation pointed out above, the biomass associated with maximum sustainable yield of the system is attained at $X = X_{MSY} = K/2$.
+For example, in the logistic equation pointed out above, the maximum growth rate occurs at a population size $X = K/2$.
 The optimizer is an \textbf{escapement} policy, which corresponds to a harvest, $h_t$, that keeps the system at its optimal growth rate as much as possible:
@@
 
-Escapement policies, or more generally $bang bang$ policies, tend to be the optimal solution for these types of control problems.
+Escapement policies, or more generally $bang$ $bang$ policies, tend to be the optimal solution for these types of control problems.
 A drawback of these solutions in the fishery context, is the possible presence of several time steps with zero harvest.
@@
 One ubiquitous solution, often called \textbf{maximum sustainable yield (MSY)}, provides biological reference points that can be simply calculated from the surplus production model parameters.
-The solution consists of letting $h(X) = rX/2$, so that
+The population size at the maximum growth rate, $X = K/2$, is approached asymptotically from any positive initial state $X_0 \in (0,K)$, under a constant mortality $M_{MSY}=r/2$ policy.
+At equilibrium, this produces the maximum sustainable yield:
 
@@
 That is, at the MSY biomass, $X_{MSY}$, the logistic growth of $X$ is cancelled exactly by the harvest.
-Additionally, the constant rate of fishing mortality, $F_{MSY}$, that results in $X_{MSY}$ is attained at $F_{MSY}=r/2$.
 This \textbf{MSY constant mortality} rule fixes the drawback in the escapement policy by having $h(X)>0$ for all $X>0$.
@@
 
-Due to this, similar but more conservative policies are often applied where the constant rate of fishing mortality is $< F_{MSY}$.
+Due to this, similar but more conservative policies are often applied where the constant rate of fishing mortality is $< M_{MSY}$.
 This control rule consists on reducing the inclination of the line defined by $h(X)$ using a prefactor $\alpha$ in $h(X) = \alpha\cdot rX/2$.
-Plausible examples are $\alpha = 0.8$ or $0.9$; here we examine an \textbf{80\% MSY constant mortality} policy (ref.
-supplemental here?).
+Plausible examples are $\alpha = 0.8$ or $0.9$; here we examine an \textbf{80\% MSY constant mortality} policy.
 

### 2023-01-27 manuscript/manuscript.Rmd
First pass at edits

--- 2023-01-27 manuscript/manuscript.Rmd
@@
 Recent advances in machine learning may now once again bridge these approaches, while also bringing new challenges of their own. 
-Novel data-driven methods have allowed these models to evolve into ever more complex and realistic simulations used in fisheries management, where models with over 100 parameters are not uncommon [@RamLegacyDB].
+Advances in data-driven methods have allowed these models to evolve into ever more complex and realistic simulations used in fisheries management, where models with over 100 parameters are not uncommon [@RamLegacyDB].
 Constrained by computational limits, MDP approaches have been intractable on suitably realistic models and largely confined to more academic applications [@Costello2016].
 However, advances _Deep Reinforcement Learning_, (RL) a sub-field of machine learning, have recently demonstrated remarkable performance in a range of such MDP problems, from video games [@atari] to fusion reactions [@fusion] to the remarkable dialog abilities of ChatGPT [@chatgpt].
-RL methods also bring many challenges of their own: being notoriously difficult to train and evaluate, immense computational costs, and frequent challenges with transfer-ability or reproducibility. 
-A review of all these issues is beyond our scope but can be found elsewhere [@rl_intro; @millie_AM-RL].
-Here, though, we will focus on the issue of opacity and interpretability raised by these methods.
-In contrast with optimization algorithms currently used in either ecosystem management or resource economics, RL algorithms have no guarantees of or metrics for convergence to an optimal solution.
-We can only assess the performance of these black box methods relative to alternatives.
 
@@
 
-Reinforcement learning (RL) is, in a nutshell, a way of approaching *control problems* through machine learning.
-An RL algorithm can be conceptually separated into two parts: an *agent*, and an *environment* which the agent can interact with.
-That is, the agent may act on the environment and thus change its state, while the environment gives a *reward* to the agent in return (see Fig. \ref{tbd}).
+Reinforcement learning (RL) is a way of approaching *sequential decision problems* through machine learning.
+All applications of RL can be conceptually separated into two parts: an *agent* and an *environment* which the agent interacts with.
+That is, the agent takes actions within the environment.
+As a result of an agent's action, the environment will transition to a new state and return a numerical *reward* to the agent (see Fig. \ref{tbd}).
 The rewards encode the agent's goal.
-The main part of an RL algorithm is then to progressively improve the agent's *policy*, in order to maximize the cumulative reward received. 
-This is done by aggregating experience and learning from it.
+The main task of any RL algorithm is then to maximize the cumulative reward received. 
+This objective is achieved by aggregating experience in what is called the *training* period and learning from such experience.
+<!-- ML: We could cite our deep RL paper here instead of going into further details. Not sure what we are expecting as the audience. -->.
 
@@
 
-Mathematically, RL may be formulated using a discrete time *partially observable Markov decision process (POMDP)*.
+The RL environment can be formally described as a discrete time *partially observable Markov decision process (POMDP)*.
 This formalization is rather flexible and allows one, e.g., to account for situations where the agent may not fully observe the environment state, or where the only observations available to the agent are certain functions of the underlying state.
-For the sake of clarity, we will present here only the class of POMDPs which are relevant to our work: *fully observable MDPs with a trivial emission function* (FMDPs for short).
+For the sake of clarity, we will only present here the class of POMDPs which are relevant to our work: *fully observable MDPs with a trivial emission function* (FMDPs for short).
 An FMDP may be defined by the following data:
@@
 is the cumulative reward of path $\tau$.
-The function $J(\pi):=\mathbb{E}_{\tau\sim p_\pi}[R(\tau)]$ is called the *value function*.
+The function $J(\pi):=\mathbb{E}_{\tau\sim p_\pi}[R(\tau)]$ is called the *expected return*.
 
@@
 
-The policy function $\pi$ often lives in a high- or even infinite-dimensional space.
+The policy function $\pi$ often lives in a high or even infinite-dimensional space.
 This makes it unfeasible to directly optimize $\pi$.
-In practice, an alternative approach is used: $\pi$ is optimized over a much lower-dimensional parametrized family of functions.
-Deep reinforcement learning uses this strategy, focusing on function families parametrized by neural networks. (See Fig. \ref{tbd}.)
+In practice, an alternative approach is used: $\pi$ is optimized over a much lower-dimensional parameterized family of functions.
+<!-- ML: I am not following this line of thought here.
+I see the main reason why NN's are used as being that NN's are universal function approximators, and we don't know what the policy and/or value function will look like.-->
+Deep reinforcement learning uses this strategy, focusing on function families parameterized by neural networks. (See Fig. \ref{tbd}.)
 <!-- The figure should kinda explain how the parametrization arises, and how gradient ascent is used. Maybe namedrop backpropagataion for the sake of usefulness to the reader? -->
@@
 
-Reinforcement learning is, on the other hand, typically a model-free approach to control theory.\footnote{There are, however, approaches to perform model-based reinforcement learning.
+Reinforcement learning, on the other hand, can provide a model-free approach to control theory.\footnote{There are approaches to perform model-based reinforcement learning.
 While we will not focus on these in this paper they are discussed in \cite{tbd1, tbd2}.}
-While the model is certainly used to generate training data, it is not directly used by the optimization algorithm.
-This provides more flexibility to use model-free RL in instances where the model of the system is not accurately known.
-In fact, it has been shown to generally be preferable to use model-free RL in such instances.
+While a model is often used to generate training data, this model is not directly used by model-free RL algorithms.
+This provides more flexibility to use RL in instances where the model of the system is not accurately known.
+In fact, it has been shown that model-free RL outperforms model-based alternatives in such instances \cite{Janner et al., 2019}.
 
@@
 This provides very interpretable solutions, such as constant mortality policies.
-There is a drawback here, however: due to the simplicity of these dynamical models, the policies might not respond effectively in many situations---situations where the predictions of these simple models deviate considerably from reality.
-Because of this, policies such as MSY have faced pushback and are believed to have contributed to the depletion of fish stocks [@tbd].
+There is a drawback here, however: due to the simplicity of these dynamical models, the policies might not respond effectively in many situations---situations where the predictions of these simple models deviates considerably from reality.
+Because of this, policies such as MSY have faced pushback and are believed to have contributed to the depletion of fish stocks.
 
 We propose an alternative approach to the problem of fishery control: to use a more expressive -- albeit more complicated -- dynamical model to guide decision making.
-Furthermore, rather than computing the optimal control policy for the model (something that is impossible in practice for complex systems) we use deep reinforcement learning to obtain a ‘‘pretty darn good’’ policy.
-This policy is estimated in a *model-free* setting, i.e., the agent treats the dynamical model (e.g. eq. \eqref{eq:3d model stoch}) as a black box of input-output pairs.
-By not relying on precise knowledge of the model's parameter values, but rather just relying on input-output statistics, model-free approaches have gained traction in a variety of control theory settings [@rl_people].
+Furthermore, rather than computing the optimal control policy for the model -- something that is impossible in practice for complex systems -- we use deep reinforcement learning to obtain a ‘‘pretty darn good’’ policy.
 
-We compare deep reinforcement learning-based policies against optimal constant mortality policies. 
-While the latter are MSY-inspired, they are optimized in a model-free way as well: the optimal constant mortality rate is computed empirically from simulated data.
-
-Because of the simplicity of the constant mortality ansatz, the optimal such policy may be easily estimated (see Figure \ref{fig:tuning}).
-In contrast, DRL optimizes over the more expressive -- and more complicated -- family of policies parametrized by a neural network.
-Because of this, our DRL approach uses a variant of *gradient ascent* (the proximal policy optimization algorithm) to obtain a good candidate policy.
-
-We have shown that the DRL agent finds an approximately *bang-bang* type of solution (see Textbox 1 for the definition of bang-bang solutions and Figs.
-\ref{fig:action-X} and \ref{fig:state-space} for the result).
+In this paper we have exemplified this approach by applying it to the control of a three-species fishery model.
+In this example, we have shown that the DRL agent finds an approximately *bang-bang* type of solution.
 While bang-bang solutions are known to be optimal for a wide variety of single-species fishery control problems [@alexandru], there is no such proof of optimality for general multi-species models.
@@
 
-This manuscript functions as a first step in the process of incorporating DRL into fishery decision making.
+This way, this manuscript functions as a first step in the process of incorporating DRL into fishery decision making.
+Importantly, DRL is equipped with tools to deal with more complicated models which might require a more refined approach to control them.
+For example *long short-term memories (LSTMs)* may be used to incorporate not only $N_t$ but also $\mathrm{d}N_t/\mathrm{d}t$ into policy decisions---while this additional degree of freedom would make analytical control theory approaches unwieldy, it may be easily and cost-effectively implemented in DRL.
+Similarly, *curriculum learning* could be incorporated to deal with phenomena such as parametric noise or parametric drift [@openai].
+<!-- ML: Could also use LSTMs for parametric drift.
+Might be better to use a different selling point for curriculum learning like if you want to have a bit more generalizability in your RL agent.
+-->
+
+This way, DRL provides a framework flexible enough to capture the intricate population dynamics that are relevant in fishery management.
 We have shown that highly non-linear systems may give rise to simplified and good policies and that these may be found via reinforcement learning.
@@
 
-In this regard, DRL provides a framework flexible enough to capture the intricate population dynamics that are relevant in fishery management.
-Importantly, DRL is equipped with tools to deal with complex models for which simple policies might yield unsatisfactory results (and thus more refined techniques are needed to control them).
-For example *long short-term memories (LSTMs)* may be used to incorporate not only $N_t$ but also $\mathrm{d}N_t/\mathrm{d}t$ into policy decisions---while this additional degree of freedom would make analytical control theory approaches unwieldy, it may be easily and cost-effectively implemented in DRL.
-Similarly, *curriculum learning* could be incorporated to deal with phenomena such as parametric noise or parametric drift [@openai].
-
-Finally, in our study case, we have shown that the ‘‘rest periods’’ so prominent in both the DRL solution and in the optimal bang-bang policy -- periods of several time steps with no fishing at all -- may be important to maintain the sustainability of the fishery and to increase its long-term profits.
-This hypothesis could be explored in more generality as it might apply to a wide variety of models with different optimal control strategies.
-In App. \ref{app:tbd} we provide some elementary support for this hypothesis by examining the performance of several ‘‘rule of thumb’’ policies based on conservative corrections to MSY.
+Conversely, our analysis suggests that, for high dimensional systems, bang-bang policies lead to increased sustainability and long-term profits as compared to constant mortality strategies.
+Similarly, in App. \ref{app:tbd} we investigate several generalizations of the constant effort approach. 
+For example, we examine strategies with a ramp-down from constant effort to zero effort at low populations.
+These, too, yield subpar results as compared to escapement.
+This seems to indicate that the ‘‘rest periods’’ so prominent in both the DRL solution and in the optimal bang-bang policy -- periods of several time steps with no fishing at all -- are important to maintain the sustainability of the fishery and, additionally, increase long-term profits.
 

### 2023-01-27 manuscript/manuscript.Rmd
mini changes discussion

--- 2023-01-27 manuscript/manuscript.Rmd
@@
 One intuition for why a constant escapement policy performs so well in such a model comes from the anticorrelation between $X$ and $Y$ noted in the results section.
-We have observed in a variety of policies, both classical as well as some obtained by training DRL agents.
-In all of these, the pattern has been present.
+We have observed this pattern arise in a variety of policies, both classical as well as some obtained by training DRL agents.
 In fact, ‘‘crashes’’ in the $X$ population are almost invariably shortly preceeded by dramatic increases in $Y$.
@@
 This manuscript functions as a first step in the process of incorporating DRL into fishery decision making.
-We have shown that highly non-linear systems may give rise to simplified and good policies and that these may be found via reinforcement learning.
+We have shown that highly non-linear systems may give rise to simple, good policies, and that these may be found via reinforcement learning.
 Moreover, the framework allows for the possibility that these simplifications do not in fact happen, and that good harvest policies are highly non-linear, multi-parameter functions.

### 2023-01-27 manuscript/manuscript.Rmd

@@
 Similarly, *curriculum learning* could be incorporated to deal with phenomena such as parametric noise or parametric drift [@openai].
+<!-- ML: Could also use LSTMs for parametric drift.
+Might be better to use a different selling point for curriculum learning like if you want to have a bit more generalizability in your RL agent.
+-->
 

### 2023-01-28 manuscript/manuscript.Rmd
mini

@@
 That is, the agent takes actions within the environment.
-As a result of an agent's action, the environment will transition to a new state and return a numerical *reward* to the agent (see Fig. \ref{tbd}).
+As a result of an agent's action, the environment will transition to a new state and return a numerical *reward* to the agent.\footnote{See Fig.
+TBD in [@rl-intro-marcus] for a conceptual description of reinforcement learning algorithms.}
 The rewards encode the agent's goal.
@@
 This formalization is rather flexible and allows one, e.g., to account for situations where the agent may not fully observe the environment state, or where the only observations available to the agent are certain functions of the underlying state.
-For the sake of clarity, we will only present here the class of POMDPs which are relevant to our work: *fully observable MDPs with a trivial emission function* (FMDPs for short).
+For the sake of clarity, we will only present here the class of POMDPs which are relevant to our work: *fully observable MDPs with a trivial emission function* (henceforth MDPs for short).
 An FMDP may be defined by the following data:
@@
 
-Fig. \ref{action-X} additionally shows (as a red line) the optimal escapement policy.
+Fig. \ref{fig:action-X} additionally shows (as a red line) the optimal escapement policy.
 This policy was computed in a manner similar to how the MSY constant mortality was computed: for each escapement in a grid of 100 values between 0 and 1, 50 episodes were run using the corresponding escapement policy.

### 2023-01-28 manuscript/manuscript.Rmd
added escapement vs X graphs

@@
 [@repo] for the code used).
-The agent was trained for 250 iterations of the PPO optimization step.
+The agent was trained for 200 iterations of the PPO optimization step.
 
@@
 
-(py | pz) + plot_annotation(tag_levels = c('A'))
+escxy <- ppo_df |>
+  mutate(X = X+1, Y = Y+1, Z = Z+1, escapement = (1-action)*X) |>
+  ggplot(aes(X, escapement, color = Y)) +
+  geom_point() +
+  theme(aspect.ratio = 1/1)
+
+escxz <- ppo_df |>
+  mutate(X = X+1, Y = Y+1, Z = Z+1, escapement = (1-action)*X) |>
+  ggplot(aes(X, escapement, color = Z)) +
+  geom_point() +
+  theme(aspect.ratio = 1/1)
+
+((py | pz)/ (escxy | escxz)) + plot_annotation(tag_levels = c('A'))
 ```

### 2023-01-28 manuscript/manuscript.Rmd
added msy to action-X

--- 2023-01-28 manuscript/manuscript.Rmd
@@
 
+opt_escapement_escapement <- function(X) case_when(
+  X <= best_e ~ X,
+  X > best_e ~ best_e
+)
+
+msy_harvest <- function(X) best_action*X
+tac_harvest <- function(X) 0.8*msy_harvest(X)
+
 py <- ppo_df |>
@@
   theme(aspect.ratio = 1/1) +
-  stat_function(fun = opt_escapement_harvest, col = 'red')
+  stat_function(fun = opt_escapement_harvest, col = 'deeppink') + 
+  stat_function(fun = msy_harvest, col = 'sienna4')
 
@@
   theme(aspect.ratio = 1/1) +
-  stat_function(fun = opt_escapement_harvest, col = 'red')
+  stat_function(fun = opt_escapement_harvest, col = 'deeppink') + 
+  stat_function(fun = msy_harvest, col = 'sienna4')
 
-(py | pz) + plot_annotation(tag_levels = c('A'))
+escxy <- ppo_df |>
+  mutate(X = X+1, Y = Y+1, Z = Z+1, escapement = (1-action)*X) |>
+  ggplot(aes(X, escapement, color = Y)) +
+  geom_point() +
+  theme(aspect.ratio = 1/1) +
+  stat_function(fun = opt_escapement_escapement, col = 'deeppink')
+
+escxz <- ppo_df |>
+  mutate(X = X+1, Y = Y+1, Z = Z+1, escapement = (1-action)*X) |>
+  ggplot(aes(X, escapement, color = Z)) +
+  geom_point() +
+  theme(aspect.ratio = 1/1) +
+  stat_function(fun = opt_escapement_escapement, col = 'deeppink')
+
+((py | pz)/ (escxy | escxz)) + plot_annotation(tag_levels = c('A'))
 ```

### 2023-01-28 manuscript/manuscript.Rmd
colors, caption

--- 2023-01-28 manuscript/manuscript.Rmd
@@
 #| label: action-X
-#| fig.cap: "TBD"
+#| fig.cap: "(A-B) Harvest and escapement scatter data for the DRL policy together with the optimal constant escapement (pink) and constant mortality (green) policies for comparison. 
+#| (A) includes $Y$ dependence through color, whereas (B) includes $Z$ dependence similarly.
+#| (B-C) escapement (i.e. $X_t - h_t$) scatter data for the DRL policy. 
+#| Optimal constant escapement policy shown in pink for comparison."
 #| out.width: "6.5in"
@@
   theme(aspect.ratio = 1/1) +
-  stat_function(fun = opt_escapement_harvest, col = 'red')
+  stat_function(fun = opt_escapement_harvest, col = 'deeppink', size = 0.8) + 
+  stat_function(fun = msy_harvest, col = 'seagreen2', size = 0.8)
 
@@
   theme(aspect.ratio = 1/1) +
-  stat_function(fun = opt_escapement_harvest, col = 'red')
+  stat_function(fun = opt_escapement_harvest, col = 'deeppink', size = 0.8) + 
+  stat_function(fun = msy_harvest, col = 'seagreen2', size = 0.8)
 
-(py | pz) + plot_annotation(tag_levels = c('A'))
+escxy <- ppo_df |>
+  mutate(X = X+1, Y = Y+1, Z = Z+1, escapement = (1-action)*X) |>
+  ggplot(aes(X, escapement, color = Y)) +
+  geom_point() +
+  theme(aspect.ratio = 1/1) +
+  stat_function(fun = opt_escapement_escapement, col = 'deeppink', size = 0.8)
+
+escxz <- ppo_df |>
+  mutate(X = X+1, Y = Y+1, Z = Z+1, escapement = (1-action)*X) |>
+  ggplot(aes(X, escapement, color = Z)) +
+  geom_point() +
+  theme(aspect.ratio = 1/1) +
+  stat_function(fun = opt_escapement_escapement, col = 'deeppink', size = 0.8)
+
+((py | pz)/ (escxy | escxz)) + plot_annotation(tag_levels = c('A'))
 ```

### 2023-01-28 manuscript/manuscript.Rmd
thickness

@@
   theme(aspect.ratio = 1/1) +
-  stat_function(fun = opt_escapement_harvest, col = 'deeppink', size = 0.8) + 
-  stat_function(fun = msy_harvest, col = 'seagreen2', size = 0.8)
+  stat_function(fun = opt_escapement_harvest, col = 'deeppink', size = 0.7) + 
+  stat_function(fun = msy_harvest, col = 'seagreen2', size = 0.7)
 
@@
   theme(aspect.ratio = 1/1) +
-  stat_function(fun = opt_escapement_escapement, col = 'deeppink', size = 0.8)
+  stat_function(fun = opt_escapement_escapement, col = 'deeppink', size = 0.7)
 

### 2023-01-28 manuscript/manuscript.Rmd
tuning fig, mortality now M

@@
 
-best_e <- 0.61 # optimal constant escapement, computed from ../data/escapement.csv.gz
 esc_sim_df <- escapement_df |> 
@@
 A grid of 100 mortality rates was laid in the interval $[0, 0.1]$.
-For each one of these rates, say $F$, we simulated 50 episodes based on \eqref{eq:3d model stoch}:
-At each time step the state $(X_t, Y_t, Z_t)$ was observed, a harvest of $FX_t$ was collected and then the system evolved according to its natural dynamics \eqref{eq:3d model stoch}.
-The optimal mortality rate was the value $F^*$ for which the mean episode reward was maximal.
-We additionally evaluated a more conservative strategy where a mortality of $0.8 F^*$ instead is used.
+For each one of these rates, say $M$, we simulated 50 episodes based on \eqref{eq:3d model stoch}:
+At each time step the state $(X_t, Y_t, Z_t)$ was observed, a harvest of $MX_t$ was collected and then the system evolved according to its natural dynamics \eqref{eq:3d model stoch}.
+The optimal mortality rate was the value $M^*$ for which the mean episode reward was maximal.
+We additionally evaluated a more conservative strategy where a mortality of $0.8 M^*$ instead is used.
 
 The result of the tuning procedure is shown in Fig. \ref{fig:tuning}. 
-There, we see that the optimal value $F^*$ lies in an area with high episode reward variance.
+There, we see that the optimal value $M^*$ lies in an area with high episode reward variance.
 We will see in Fig. \ref{fig:rew_len_fig} that a likely candidate for this variance is the presence of episodes with early endings.
-Thus, the value $0.8 F^*$ supports a smaller variance, and correspondingly -- as Fig. \ref{fig:rew_len_fig} shows -- this policy leads to a much smaller probability of a near-extinction event.
+Thus, the value $0.8 M^*$ supports a smaller variance, and correspondingly -- as Fig. \ref{fig:rew_len_fig} shows -- this policy leads to a much smaller probability of a near-extinction event.
 
@@
 #| (A) Mean episode reward (averaged over 50 episodes) shown as scatter data, and shaded in gray are two standard deviations around each scatter point.
-Vertical lines correspond to the optimal constant mortality rate $F^* = $ TBD (red) and $0.8 F^*=$ TBD (green).
+Vertical lines correspond to the optimal constant mortality rate $M^* = $ 0.045 (green) and $0.8 M^*=$ 0.036 (violet).
 #| (B) Several constant mortality policies shown for illustrative purposes.
-Slopes $0.02$, $0.03$, $0.055$ and $0.065$ shown in black; the optimal linear policy $F^*X$ shown in red, while $0.8F^*X$ shown in green."
+Slopes $0.02$, $0.03$, $0.055$ and $0.065$ shown in black; the optimal linear policy $M^*X$ shown in green, while $0.8M^*X$ shown in violet.
+(TBD choose nicer and plentier  slopes)"
 #| out.width : "6.5in"
@@
   theme(aspect.ratio = 1/1) + 
-  geom_vline(aes(xintercept=best_action), color="red") + 
-  geom_vline(aes(xintercept=0.8*best_action), color="green") +
+  geom_vline(aes(xintercept=best_action), color="seagreen2") + 
+  geom_vline(aes(xintercept=0.8*best_action), color="violetred1") +
   theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
@@
   stat_function(fun = linear_f_gen(0.03)) +
-  stat_function(fun = linear_f_gen(0.8*best_action), col = "green") +
-  stat_function(fun = linear_f_gen(best_action), col = "red") +
+  stat_function(fun = linear_f_gen(0.8*best_action), col = "violetred") +
+  stat_function(fun = linear_f_gen(best_action), col = "seagreen2") +
   stat_function(fun = linear_f_gen(0.055)) + 
@@
 In Fig. \ref{fig:timeseries} we show the time series of one such repetition.
-The left panel uses an MSY constant mortality control $F^*X$, the middle panel uses $0.8 F^* X$, while the right panel is controlled by the trained DRL agent.
+The left panel uses an MSY constant mortality control $M^*X$, the middle panel uses $0.8 M^* X$, while the right panel is controlled by the trained DRL agent.
 We see that that optimal constant mortality control leads to a near-extinction event at around $t=150$.
-The constant mortality policy at $0.8 F^*$ is already conservative enough to support a sustainable path.
+The constant mortality policy at $0.8 M^*$ is already conservative enough to support a sustainable path.
 There, the population levels are maintained roughly in a stationary state.

### 2023-01-28 manuscript/manuscript.Rmd
comment response

@@
 <!-- ML: I am not following this line of thought here.
-I see the main reason why NN's are used as being that NN's are universal function approximators, and we don't know what the policy and/or value function will look like.-->
+I see the main reason why NN's are used as being that NN's are universal function approximators, and we don't know what the policy and/or value function will look like.
+
+FM: my take here is that the space of functions from states to actions is in principle infinite-dimensional or at least very high dimensional.
+(e.g.
+exponentially high dimensional in terms of the number of degrees of freedom in state space, for the case of discretized state spaces).
+By restricting our ansatz to neural networks of a fixed depth/breadth, we're restricting ourselves to a finite dimensional (or, in the discrete case, potentially much lower dimensional) manifold of policies over which gradient descent may be performed much more efficiently.
+E.g.
+the dimensionality of the space we're searching for is determined by the size of the neural network, it's finite dimensional for continuous state spaces, and its dimension is not necessarily exponential in the number of species.
+-->
 Deep reinforcement learning uses this strategy, focusing on function families parameterized by neural networks. (See Fig. \ref{tbd}.)

### 2023-01-28 manuscript/manuscript.Rmd
discuss figure action-X

@@
 
-Fig. \ref{fig:action-X} additionally shows (as a red line) the optimal escapement policy.
-This policy was computed in a manner similar to how the MSY constant mortality was computed: for each escapement in a grid of 100 values between 0 and 1, 50 episodes were run using the corresponding escapement policy.
+Fig. \ref{fig:action-X} additionally shows (pink line) the optimal constant escapement policy and the MSY constant mortality policy (green line).
+The optimal constant escapement was computed in a manner similar to how the MSY constant mortality was computed: for each escapement in a grid of 100 values between 0 and 1, 50 episodes were run using the corresponding escapement policy.
 The value with the best mean episode reward was chosen as the optimal escapement value.
+We observe that the DRL policy approximates the optimal constant escapement policy, with the difference of being more conservative in its harvest at high $X$ values.
+This pattern can be seen in panels (C-D), where at high $X$ values, the DRL escapement mildly increases in a non-linear fashion rather than staying constant.
+Finally, relative to these two policies, we observe that MSY constant mortality underfishes at high $X$ values and overfishes at low $X$ values.
 

### 2023-01-28 manuscript/manuscript.Rmd
started populating bib file

@@
   address: Department of Environmental Science, Policy, and Management, University of California, Berkeley
-- name: Marcus Laperolerie
+- name: Marcus Lapeyrolerie
   address: Department of Environmental Science, Policy, and Management, University of California, Berkeley

### 2023-01-28 manuscript/manuscript.Rmd
more refs

@@
 There are myriad examples from climate scenarios to protected areas, we will illustrate this with a specific example in the management of marine fisheries, which is both an important issue to itself and a frequent crucible in which methods of ecological management have been fired before extending to other fields, and where the space of possible policies is arguably already more richly developed than most.
-Fisheries are an essential natural resource that provide the primary source of protein for one in every four humans, and have faced widely documented declines due to over-fishing [@Worm2006, @Costello2016].
+Fisheries are an essential natural resource that provide the primary source of protein for one in every four humans, and have faced widely documented declines due to over-fishing [@Worm2006; @Costello2016].
 Because of this importance, fisheries also represent some of the longest and best studied timeseries of population dynamics available to ecologists, as well as a correspondingly rich area for the development of theory and practice.
@@
 There is a drawback here, however: due to the simplicity of these dynamical models, the policies might not respond effectively in many situations---situations where the predictions of these simple models deviate considerably from reality.
-Because of this, policies such as MSY have faced pushback and are believed to have contributed to the depletion of fish stocks [@tbd].
+Because of this, policies such as MSY have faced pushback and are believed to have contributed to the depletion of fish stocks [@Worm2006; @Costello2016].
 
@@
 \ref{fig:action-X} and \ref{fig:state-space} for the result).
-While bang-bang solutions are known to be optimal for a wide variety of single-species fishery control problems [@alexandru], there is no such proof of optimality for general multi-species models.
+While bang-bang solutions are known to be optimal for a wide variety of single-species fishery control problems [@hening2019; @hening2021], there is no such proof of optimality for general multi-species models.
 In this sense this example is rather surprising and reassuring: while no general guarantee points at the optimality of bang-bang solutions for multi-species model, the existence of one such example provides a benchmark for the use of DRL in fishery control.
@@
 For example *long short-term memories (LSTMs)* may be used to incorporate not only $N_t$ but also $\mathrm{d}N_t/\mathrm{d}t$ into policy decisions---while this additional degree of freedom would make analytical control theory approaches unwieldy, it may be easily and cost-effectively implemented in DRL.
-Similarly, *curriculum learning* could be incorporated to deal with phenomena such as parametric noise or parametric drift [@openai].
+Similarly, *curriculum learning* [@curriculum-learning] or *open-ended learning* [@open-ended] could be incorporated to deal with phenomena such as parametric noise, parametric drift, and, more generally, for deriving generally well-performing agents *vis-à-vis* uncertainties in the system's dynamics.
 <!-- ML: Could also use LSTMs for parametric drift.

### 2023-01-28 manuscript/manuscript.Rmd
more refs

@@
 This policy is estimated in a *model-free* setting, i.e., the agent treats the dynamical model (e.g. eq. \eqref{eq:3d model stoch}) as a black box of input-output pairs.
-By not relying on precise knowledge of the model's parameter values, but rather just relying on input-output statistics, model-free approaches have gained traction in a variety of control theory settings [@rl_people].
+By not relying on precise knowledge of the model's parameter values, but rather just relying on input-output statistics, model-free approaches have gained traction in a variety of control theory settings (see, e.g., [@sato2019; @ramirez2022; @zeng2019]).
 

### 2023-01-28 manuscript/manuscript.Rmd
wording

@@
 Importantly, DRL is equipped with tools to deal with complex models for which simple policies might yield unsatisfactory results (and thus more refined techniques are needed to control them).
-For example *long short-term memories (LSTMs)* may be used to incorporate not only $N_t$ but also $\mathrm{d}N_t/\mathrm{d}t$ into policy decisions---while this additional degree of freedom would make analytical control theory approaches unwieldy, it may be easily and cost-effectively implemented in DRL.
-Similarly, *curriculum learning* [@curriculum-learning] or *open-ended learning* [@open-ended] could be incorporated to deal with phenomena such as parametric noise, parametric drift, and, more generally, for deriving generally well-performing agents *vis-à-vis* uncertainties in the system's dynamics.
+For example, tools such as *long short-term memories* [@lstm], *curriculum learning* [@curriculum-learning] and *open-ended learning* [@open-ended] can be incorporated when faced with parametric noise or drift, or generally uncertainties about the agent's environment.
+These techniques can be used to obtain agents who perform well over a variety of different environments [@open-ended].
 <!-- ML: Could also use LSTMs for parametric drift.

### 2023-01-28 code/variable_tac.py
var_tac.py for appendix

@@ -1,0 +1,37 @@
+import gym_fishing
+import gym
+import pandas as pd
+import numpy as np
+import ray
+
+@ray.remote
+def simulate(env, fixed_action, threshold_pop = 0.5):
+  df = []
+  for rep in range(50):
+    episode_reward = 0
+    observation = env.reset()
+    for t in range(env.Tmax):
+      action = fixed_action
+      if observation[0]+1 < threshold_pop:
+        action = action * (observation[0] + 1) / threshold_pop
+      df.append(np.append([t, rep, action, episode_reward], observation))
+      observation, reward, terminated, info = env.step(action)
+      episode_reward += reward
+      if terminated:
+        break
+  return(df)
+
+
+env = gym.make("threeFishing-v2")
+env.training = False
+actions = np.linspace(0,0.2,101)
+
+# define parllel loop and execute
+parallel = [simulate.remote(env, i) for i in actions]
+df = ray.get(parallel)
+
+# convert to data.frame & write to csv
+cols = ["t", "rep", "action", "reward", "X", "Y", "Z"]
+df2 = pd.DataFrame(np.vstack(df), columns = cols)
+df2.to_csv("data/var_tac.csv.xz", index=False)
+

### 2023-01-28 manuscript/manuscript.Rmd
added a TAC appendix

@@
 
+# Appendix A: collected time series data
+
+# Appendix B: variations on MSY
+
+Here we show the performance of some variations of the MSY constant mortality policy.
+
+First, consider constant mortality policies with a fraction $\alpha$ of the MSY mortality rate (in the main text we focused on $\alpha=0.8$).
+Notice the sharp transition in performance between $\alpha = 0.9$ and $0.95$.
+While the behaviour of the former closely mimics, e.g., the $\alpha=0.8$ policy, the performance of the latter is essentially the same as the optimal constant mortality ($\alpha=1$, see Fig.
+\ref{fig:rew_len_fig}).
+
+```{r}
+#| label: several-tac
+#| fig.cap: "Performance of constant mortality policies with mortality $\alpha M^*$ for several values of $\alpha$.
+#| The top panel displays the episode lengths for each repetition, while the bottom pannel displays the episode harvest.
+#| The labels in the plots are the value of $\alpha$ expressed as a percentage."
+#| out.width: "65.in"
+N <- 30
+actions <- unique(msy_df$action)
+i85 <- which.min(abs(actions - best_action* 0.85))
+i90 <- which.min(abs(actions - best_action* 0.90))
+i95 <- which.min(abs(actions - best_action* 0.95))
+tac_85_sim <- msy_df |> 
+    filter(action == actions[[i85]]) |>
+    pivot_longer(c("X", "Y", "Z"),
+        names_to = "species", values_to = "abundance") |>
+    mutate(abundance = abundance + 1, policy = "85%")
+tac_90_sim <- msy_df |> 
+    filter(action == actions[[i90]]) |>
+    pivot_longer(c("X", "Y", "Z"),
+        names_to = "species", values_to = "abundance") |>
+    mutate(abundance = abundance + 1, policy = "90%")
+tac_95_sim <- msy_df |> 
+    filter(action == actions[[i95]]) |>
+    pivot_longer(c("X", "Y", "Z"),
+        names_to = "species", values_to = "abundance") |>
+    mutate(abundance = abundance + 1, policy = "95%")
+
+tac_80_sim <- tac_sim |> mutate(policy = "80%")
+
+tac4 <- bind_rows(tac_80_sim, tac_85_sim, tac_90_sim, tac_95_sim)
+
+p1 <- tac4 |> 
+    group_by(rep, policy) |>
+    filter(t==max(t), species == "X", rep < N) |>
+    ggplot(aes(rep, t)) + 
+      geom_col() +
+      facet_wrap(~factor(policy), ncol = 4) +
+      labs(y = "episode length", x = "repetition") +
+      theme(axis.text.x = element_blank())
+
+p2 <- tac4 |> 
+    group_by(rep, policy) |>
+    filter(t==max(t), species == "X", rep < N) |>
+    ggplot(aes(rep, reward)) + 
+      geom_col() +
+      facet_wrap(~factor(policy), ncol=4) +
+      labs(y = "episode harvest", x = "repetition") +
+      theme(axis.text.x = element_blank())
+
+(p1 / p2)
+```
+
 ## Acknowledgements

### 2023-01-28 code/variable_tac.py
code mini change

@@ -14,7 +14,7 @@ def simulate(env, fixed_action, threshold_pop = 0.5):
       action = fixed_action
       if observation[0]+1 < threshold_pop:
         action = action * (observation[0] + 1) / threshold_pop
-      df.append(np.append([t, rep, action, episode_reward], observation))
+      df.append(np.append([t, rep, fixed_action, action, episode_reward], observation))
       observation, reward, terminated, info = env.step(action)
       episode_reward += reward
       if terminated:
@@ -31,7 +31,7 @@ parallel = [simulate.remote(env, i) for i in actions]
 df = ray.get(parallel)
 
 # convert to data.frame & write to csv
-cols = ["t", "rep", "action", "reward", "X", "Y", "Z"]
+cols = ["t", "rep", "policy_action", "action", "reward", "X", "Y", "Z"]
 df2 = pd.DataFrame(np.vstack(df), columns = cols)
 df2.to_csv("data/var_tac.csv.xz", index=False)
 

### 2023-01-28 manuscript/manuscript.Rmd
appendices A and B

--- 2023-01-28 manuscript/manuscript.Rmd
@@
 This hypothesis could be explored in more generality as it might apply to a wide variety of models with different optimal control strategies.
-In App. \ref{app:tbd} we provide some elementary support for this hypothesis by examining the performance of several ‘‘rule of thumb’’ policies based on conservative corrections to MSY.
+
+# Appendix A: collected time series data
+
+In this appendix we provide a wider selection of time series taken from the same data set as Fig. \ref{fig:timeseries}.
+Namely, Figs. \ref{fig:timeseries_msy}, \ref{fig:timeseries_tac} and \ref{fig:timeseries_ppo} display these time series.
+
+```{r} 
+#| label: timeseries_msy
+#| fig.cap: "Nine time series of the system controlled by the MSY constant mortality policy." 
+#| out.width: "6.5in"
+msy_tac_ppo |> 
+    group_by(rep) |>
+    filter(rep < 9, policy == "msy") |>
+    ggplot(aes(t, abundance, col = species)) + 
+        geom_line()  + 
+        scale_color_manual(values=c('palegreen3', 'coral1', 'deepskyblue4')) +
+        facet_wrap(~factor(rep)) +
+        labs(y = "abundance", x = "time") +
+        theme(aspect.ratio = 0.6)
+```
+
+```{r} 
+#| label: timeseries_tac
+#| fig.cap: "Nine time series of the system controlled by the policy using 80 percent of the MSY constant mortality." 
+#| out.width: "6.5in"
+msy_tac_ppo |> 
+    group_by(rep) |>
+    filter(rep < 9, policy == "tac") |>
+    ggplot(aes(t, abundance, col = species)) + 
+        geom_line()  + 
+        scale_color_manual(values=c('palegreen3', 'coral1', 'deepskyblue4')) +
+        facet_wrap(~factor(rep)) +
+        labs(y = "abundance", x = "time") +
+        theme(aspect.ratio = 0.6)
+```
+
+```{r} 
+#| label: timeseries_ppo
+#| fig.cap: "Nine time series of the system controlled by the trained DRL agent." 
+#| out.width: "6.5in"
+msy_tac_ppo |> 
+    group_by(rep) |>
+    filter(rep < 9, policy == "ppo") |>
+    ggplot(aes(t, abundance, col = species)) + 
+        geom_line()  + 
+        scale_color_manual(values=c('palegreen3', 'coral1', 'deepskyblue4')) +
+        facet_wrap(~factor(rep)) +
+        labs(y = "abundance", x = "time") +
+        theme(aspect.ratio = 0.6)
+```
+
+# Appendix B: Conservative scale-downs of MSY
+
+Here we show the performance of constant mortality policies with a fraction $\alpha$ of the MSY mortality rate (in the main text we focused on $\alpha=0.8$).
+Their performance is evaluated in Fig. \ref{fig:several-tac}.
+Notice the sharp transition in performance between $\alpha = 0.9$ and $0.95$.
+While the behaviour of the former closely mimics, e.g., the $\alpha=0.8$ policy, the performance of the latter is essentially the same as the optimal constant mortality ($\alpha=1$, see Fig.
+\ref{fig:rew_len_fig}).
+This points at the fact that the value of the parameter $\alpha$, although commonly chosen simply as a rule of thumb, lives on a ‘‘knife's edge,’’ small changes in this parameter may lead to drastic changes in the dynamics of the controlled system.
+What is worse: the particular scale at which this transition happens is generally dependent on the system's intrinsic dynamics.
+In reality, where fishery models and their respective parameters are subject to substantial uncertainty, this fact calls to question the purported additional safety obtained via these rules of thumb.
+
+```{r}
+#| label: several-tac
+#| fig.cap: "Performance of constant mortality policies with mortality $\\alpha M^*$ for several values of $\\alpha$.
+#| The top panel displays the episode lengths for each repetition, while the bottom pannel displays the episode harvest.
+#| The labels in the plots are the value of $\\alpha$ expressed as a percentage."
+#| out.width: "6.5in"
+N <- 30
+actions <- unique(msy_df$action)
+i85 <- which.min(abs(actions - best_action* 0.85))
+i90 <- which.min(abs(actions - best_action* 0.90))
+i95 <- which.min(abs(actions - best_action* 0.95))
+tac_85_sim <- msy_df |> 
+    filter(action == actions[[i85]]) |>
+    pivot_longer(c("X", "Y", "Z"),
+        names_to = "species", values_to = "abundance") |>
+    mutate(abundance = abundance + 1, policy = "85%")
+tac_90_sim <- msy_df |> 
+    filter(action == actions[[i90]]) |>
+    pivot_longer(c("X", "Y", "Z"),
+        names_to = "species", values_to = "abundance") |>
+    mutate(abundance = abundance + 1, policy = "90%")
+tac_95_sim <- msy_df |> 
+    filter(action == actions[[i95]]) |>
+    pivot_longer(c("X", "Y", "Z"),
+        names_to = "species", values_to = "abundance") |>
+    mutate(abundance = abundance + 1, policy = "95%")
+
+tac_80_sim <- tac_sim |> mutate(policy = "80%")
+
+tac4 <- bind_rows(tac_80_sim, tac_85_sim, tac_90_sim, tac_95_sim)
+
+p1 <- tac4 |> 
+    group_by(rep, policy) |>
+    filter(t==max(t), species == "X", rep < N) |>
+    ggplot(aes(rep, t)) + 
+      geom_col() +
+      facet_wrap(~factor(policy), ncol = 4) +
+      labs(y = "episode length", x = "repetition") +
+      theme(axis.text.x = element_blank())
+
+p2 <- tac4 |> 
+    group_by(rep, policy) |>
+    filter(t==max(t), species == "X", rep < N) |>
+    ggplot(aes(rep, reward)) + 
+      geom_col() +
+      facet_wrap(~factor(policy), ncol=4) +
+      labs(y = "episode harvest", x = "repetition") +
+      theme(axis.text.x = element_blank())
+
+(p1 / p2)
+```
 

### 2023-01-28 manuscript/manuscript.Rmd
cross refs

@@
 
-This time series is representative of the patterns observed in the data (see App. \ref{app:tbd}).
+This time series is representative of the patterns observed in the data (see App. A).
 The episode reward and episode length statistics are shown in Fig. \ref{fig:rew_len_fig}.
 There, we highlight two things:
-First, that the tendency towards early episode ends is fixed by the conservative approach of fishing at 80% of the optimal mortality rate.
-(In App. \ref{app:tbd} we discuss a warning about this approach, however: the arbitrarily chosen percentage itself lives on a ‘‘knife's edge.’’
-More concretely, choosing a higher percentage, say 90%, can switch the behaviour of the controlled system and give rise to the proliferation of early episode ends.)
+First, that the tendency towards early episode ends is fixed by the conservative approach of fishing at 80% of the MSY mortality rate.
+(In App. B we discuss and evaluate other policies using a fixed-percentage of the MSY mortality rate.)
 Second, that while the conservative 80% approach has a stable behavior, it consistently produces smaller harvests than a DRL generated policy.
@@
 Notice the anti-correlation between $X$ and $Y$ along the paths, this is especially prominent for the constant mortality plots.
-This anti-correlation is, furthermore, observed in the time series plots (see Fig. \ref{fig:timeseries} and App. \ref{app:tbd}).
+This anti-correlation is, furthermore, observed in the time series plots (see Fig. \ref{fig:timeseries} and App. A).
 
@@
 
-# Appendix A: collected time series data
+\appendix
+
+# Appendix
+
+## Collected time series data
 
@@
         labs(y = "abundance", x = "time") +
-        theme(aspect.ratio = 0.6)
+        theme(aspect.ratio = 0.5)
 ```
@@
         labs(y = "abundance", x = "time") +
-        theme(aspect.ratio = 0.6)
-```
-
-# Appendix B: Conservative scale-downs of MSY
+        theme(aspect.ratio = 0.5)
+```
+
+## Conservative scale-downs of MSY
 

### 2023-01-28 code/variable_tac.py
no var tac, getting too long

@@ -1,37 +1,0 @@
-import gym_fishing
-import gym
-import pandas as pd
-import numpy as np
-import ray
-
-@ray.remote
-def simulate(env, fixed_action, threshold_pop = 0.5):
-  df = []
-  for rep in range(50):
-    episode_reward = 0
-    observation = env.reset()
-    for t in range(env.Tmax):
-      action = fixed_action
-      if observation[0]+1 < threshold_pop:
-        action = action * (observation[0] + 1) / threshold_pop
-      df.append(np.append([t, rep, fixed_action, action, episode_reward], observation))
-      observation, reward, terminated, info = env.step(action)
-      episode_reward += reward
-      if terminated:
-        break
-  return(df)
-
-
-env = gym.make("threeFishing-v2")
-env.training = False
-actions = np.linspace(0,0.2,101)
-
-# define parllel loop and execute
-parallel = [simulate.remote(env, i) for i in actions]
-df = ray.get(parallel)
-
-# convert to data.frame & write to csv
-cols = ["t", "rep", "policy_action", "action", "reward", "X", "Y", "Z"]
-df2 = pd.DataFrame(np.vstack(df), columns = cols)
-df2.to_csv("data/var_tac.csv.xz", index=False)
-

### 2023-01-29 manuscript/manuscript.Rmd
fix pdf render issue (citation)
FMDP -> MDP

@@
 For the sake of clarity, we will only present here the class of POMDPs which are relevant to our work: *fully observable MDPs with a trivial emission function* (henceforth MDPs for short).
-An FMDP may be defined by the following data:
+An MDP may be defined by the following data:
 
@@
 
-At a time $t$, the FMDP agent observes the full state $s_t$ of the environment and chooses an action based on this observation according to a *policy function* $\pi(a_t | s_t)$.
+At a time $t$, the MDP agent observes the full state $s_t$ of the environment and chooses an action based on this observation according to a *policy function* $\pi(a_t | s_t)$.
 In return, it receives a discounted reward $\gamma^t r(a_t, s_t)$.

### 2023-01-29 manuscript/manuscript.Rmd
a sentence

@@
 This provides more flexibility to use RL in instances where the model of the system is not accurately known.
-In fact, it has been shown that model-free RL outperforms model-based alternatives in such instances \cite{Janner et al., 2019}.
+In fact, it has been shown that model-free RL outperforms model-based alternatives in such instances \cite{mbpo}.
 
@@
 
-Finally, in our study case, we have shown that the ‘‘rest periods’’ so prominent in both the DRL solution and in the optimal bang-bang policy -- periods of several time steps with no fishing at all -- may be important to maintain the sustainability of the fishery and to increase its long-term profits.
-This hypothesis could be explored in more generality as it might apply to a wide variety of models with different optimal control strategies.
+Finally, in our study case, we have shown that the ‘‘rest periods’’ -- periods of several time steps with no fishing at all -- prominent in both the DRL solution and in the optimal bang-bang policy, may be important to maintain the sustainability of the fishery and to increase its long-term profits.
+This hypothesis could be explored in more generality as it might apply to a wide variety of models.
 

### 2023-01-29 manuscript/manuscript.Rmd
mini

@@
 
-This time series is representative of the patterns observed in the data (see App. A).
+This time series is representative of the patterns observed in the data (see App. A.1).
 The episode reward and episode length statistics are shown in Fig. \ref{fig:rew_len_fig}.
@@
 First, that the tendency towards early episode ends is fixed by the conservative approach of fishing at 80% of the MSY mortality rate.
-(In App. B we discuss and evaluate other policies using a fixed-percentage of the MSY mortality rate.)
+(In App. A.2 we discuss and evaluate other policies using a fixed-percentage of the MSY mortality rate.)
 Second, that while the conservative 80% approach has a stable behavior, it consistently produces smaller harvests than a DRL generated policy.
@@
 Notice the anti-correlation between $X$ and $Y$ along the paths, this is especially prominent for the constant mortality plots.
-This anti-correlation is, furthermore, observed in the time series plots (see Fig. \ref{fig:timeseries} and App. A).
+This anti-correlation is, furthermore, observed in the time series plots (see Fig. \ref{fig:timeseries} and App. A.1).
 

### 2023-01-30 manuscript/manuscript.Rmd
hand-wavy footnote

@@
 
-One intuition for why a constant escapement policy performs so well in such a model comes from the anticorrelation between $X$ and $Y$ noted in the results section.
+One intuition for why a constant escapement policy performs so well in such a model comes from the anti-correlation between $X$ and $Y$ noted in the results section.
 We have observed this pattern arise in a variety of policies, both classical as well as some obtained by training DRL agents.
@@
 We hypothesize that this effectively leads to a dimensionality reduction: once the value of $X$ is known, along a path traced by the control system, then $Y$ may be roughly predicted.
-A similar, although weaker, anticorrelation is also observed between $X$ and $Z$.
+A similar, although weaker, anti-correlation is also observed between $X$ and $Z$.
 These two informal observations point at an effective simplification of the prediction process: the values of $Y$ and $Z$ are much less impactful to policy decisions than the value of $X$ itself.
-We have not yet pushed further to establish a more precise picture of this dynamical simplification.
+We have not yet pushed further to establish a more precise picture of this dynamical simplification.\footnote{
+A hand-waving way to understand this phenomenon is the following:
+The $X$-$Y$ anti-correlation can be associated to the Lotka-Volterra competition term between $X$ and $Y$, i.e. $c_{XY}XY$ in eq. \eqref{eq:3d model stoch}.
+Now, consider a fall in population $X$.
+Because of the aforementioned anti-correlation with $Y$, this is accompanied by a growth in $Y$.
+We may understand this, roughly, as a \emph{replacement} of $X$ for $Y$.
+But then, because the parameter $D = 1.1 > 1$, this increases the effective birth rate $Z$, i.e. the term $f(X+DY)Z$ in eq. \eqref{eq:3d model stoch}.
+This, in turn, leads to a growth of $Z$ and, thus, an anti-correlation between $Z$ and $X$.}
 

### 2023-01-30 manuscript/manuscript.Rmd
citations and model-based footnote made parenthesis

@@
 One particularly important aspect that has garnered much attention is the potential for the kind of highly non-linear functions that can support dynamics such as alternative stable states and hysteresis.
-A seminal example of such dynamics was introduced in \cite{may77}, using a one-dimensional model of a prey (resource) species under the pressure of a (fixed) predator:
+A seminal example of such dynamics was introduced in [@may77], using a one-dimensional model of a prey (resource) species under the pressure of a (fixed) predator:
 
@@
     \label{fig:may stable}
-    The fixed point diagram for the model \cite{may77} associated to $f_{\text{May}}$ as a function of varying the parameter $\beta H$.
+    The fixed point diagram for the model [@may77] associated to $f_{\text{May}}$ as a function of varying the parameter $\beta H$.
     That is, these are the zeroes of the function $f_{\text{May}}$. 
@@
 
-Reinforcement learning, on the other hand, can provide a model-free approach to control theory.\footnote{There are approaches to perform model-based reinforcement learning.
-While we will not focus on these in this paper they are discussed in \cite{tbd1, tbd2}.}
+Reinforcement learning, on the other hand, can provide a model-free approach to control theory.
 While a model is often used to generate training data, this model is not directly used by model-free RL algorithms.
 This provides more flexibility to use RL in instances where the model of the system is not accurately known.
-In fact, it has been shown that model-free RL outperforms model-based alternatives in such instances \cite{mbpo}.
+In fact, it has been shown that model-free RL outperforms model-based alternatives in such instances [@mbpo].
+(Note that there are approaches to perform model-based reinforcement learning. 
+While we will not focus on these in this paper they are surveyed in, e.g., [@moerland2023model; @polydoros2017survey].)
 

### 2023-01-30 manuscript/manuscript.Rmd
added stable_points.py

@@
 bibliography: bibliography.bib
-#biblio-style: spphys
+biblio-style: spphys
 
@@
 As can be expected, the dynamics of eq. \eqref{eq:3d model} are significantly more complex than those of eq. \eqref{eq:may}.
-There are a variety of fixed points for the equations---around 7 or 8 for wide ranges of parameter values.
+There are a variety of fixed points for the equations---around 8 or 9 for wide ranges of parameter values.
 Fig. \ref{fig:tbd} provides a rough idea of the fixed point structure by plotting the bifurcation diagram obtained from varying $D$ and fixing the other parameters.
 
-Throughout most of this paper we will fix the parameters of the model according to Table \ref{tab:tbd}.
-At these values, the fixed points of the model are:
+Throughout most of this paper we will fix the parameters of the model according to eq. \eqref{eq:param vals}.
+At these values, the fixed points of the model are:\footnote{
+We have included only three decimal digits for readability.
+These fixed points may be generated by running the ‘‘stable\_points.py’’ script in [@repo].
+}
 \begin{align*}
-  TBD
+  (X,Y,Z) \approx &(0.839, 0.054, 0.377), \quad 
+    (0.048, 0.774, 0.480), \quad
+    (0.359, 0.491, 0.509), \\
+    &(0.666, 0.666, 0.000),\quad
+    (0.899, 0.000, 0.303), \quad
+    (0.000, 0.818, 0.457), \\
+    &(1.000,0.000,0.000),\quad (0.000,1.000,0.000),\quad (0.000,0.000,0.000).
 \end{align*}
@@
 A fixed point might be stable against perturbations in one direction, but unstable against perturbations along another.
-While a detailed classification of these fixed points is beyond the scope of this work, we point out that the point $(TBD)$ is stable against perturbations in all directions.
+While a detailed classification of these fixed points is beyond the scope of this work, we point out that the point $(0.839, 0.054, 0.377)$ is stable against perturbations in all directions.
 We will thus use it as a starting point for our simulations.
@@
 Since our state and observation spaces are continuous, we will focus on deep reinforcement learning throughout this paper.
-Specifically, we parametrize $\pi$ using a neural network with two hidden layers of 64 neurons.
-
 We use the *proximal policy optimization (PPO)* algorithm to optimize $\pi$.
@@
 
-Within control theory, the usual setup is one where we use as much information from the model as possible in order to derive an optimal solution.
-**(Wax poetic a bit about Bellmann eq. approaches here?)**
-
-The classical sustainable fishery management approaches summarized in Sec. \ref{sec:classical} are model-based controls.
-As we saw in that section, these controls may run into trouble in the case where there are inaccuracies in the model parameter estimates.
-
-More generally, there are many situations in which the exact model of the system is not known or not tractable.
+Within control theory, the classical setup is one where we use as much information from the model as possible in order to derive an optimal solution.
+Here, one may find a vast literature on model-based methods to attain optimal, or near-optimal, control (see, e.g., [@zhang2019near; @sethi2019optimal; @anderson2007optimal]).
+
+The classical sustainable fishery management approaches summarized in Text Box 1, for instance, are model-based controls.
+As we saw there, these controls may run into trouble in the case where there are inaccuracies in the model parameter estimates.
+
+There are many situations, however, in which the exact model of the system is not known or not tractable.
 This is a standard situation in ecology: mathematical models capture the most prominent aspects of the ecosystem's dynamics, while ignoring or summarizing most of its complexity.
@@
 In fact, it has been shown that model-free RL outperforms model-based alternatives in such instances [@mbpo].
-(Note that there are approaches to perform model-based reinforcement learning. 
-While we will not focus on these in this paper they are surveyed in, e.g., [@moerland2023model; @polydoros2017survey].)
-
-**Mention CL here? Where solutions to other related problems in the curriculum are used to build solutions to the target problem.**
+(For recent surveys of model-based reinforcement learning, which we do not focus on here, see [@moerland2023model; @polydoros2017survey].)
 
@@
 The model's dynamic parameters were chosen as follows
-\begin{align*}
+\begin{align}
+\label{eq:param vals}
+\begin{split}
   r_X = K_X = r_Y = K_Y = 1, \quad \beta = 0.3, \quad c = 0.1\\
   c_{XY} = 0.5,\quad f = 0.15,\quad D = 1.1, \quad d_Z = 0.135.
-\end{align*}
+\end{split}
+\end{align}
 Moreover, the variances for the stochastic terms were chosen as
@@
 This material is based upon work supported by the National Science Foundation under Grant No. DBI-1942280.
-
-# References

### 2023-01-30 code/stable_points.py

@@ -1,0 +1,292 @@
+from scipy.optimize import fsolve
+from numpy.random import rand
+from pandas import DataFrame
+import numpy as np
+import time
+import sys
+
+def timer(func):
+    def wrapper(*args, **kwargs):
+        t = time.time()
+        r = func(*args, **kwargs)
+        print(f"Time spent on {func.__name__}: {time.time()-t}")
+        return r
+    return wrapper
+
+def L(v):
+    return v * (1 - v)
+
+def F(v,z,beta,c):
+    return beta * z * (v ** 2) / ((c ** 2) + (v ** 2))
+
+def X_eq(point):
+    X, Y, Z, beta, c, cXY, D, dZ, f, alpha = point
+    return L(X) - F(X, Z, beta, c) - cXY * X * Y
+
+def Y_eq(point):
+    X, Y, Z, beta, c, cXY, D, dZ, f, alpha = point
+    return L(Y) - D * F(Y, Z, beta, c) - cXY * X * Y
+
+def Z_eq(point):
+    X, Y, Z, beta, c, cXY, D, dZ, f, alpha = point
+    return alpha * (- dZ * Z + f * Z * (X + D * Y))
+
+def alt_Z_eq(point):
+    X, Y, Z, beta, c, cXY, D, dZ, f, alpha = point
+    return Z - 0.5
+    
+def set_param(param, param_value):
+    return param - param_value
+
+def eqs_params(point, param_vals):
+    """
+    Args: variables, and then parameters
+    
+    (vars)
+    X, Y, Z, 
+    
+    (params)
+    beta, c, cXY, D, dZ, f
+    """
+    X, Y, Z, beta, c, cXY, D, dZ, f, alpha = point
+    beta_val, c_val, cXY_val, D_val, dZ_val, f_val, alpha_val = param_vals
+    
+    return (
+        X_eq(point),
+        Y_eq(point),
+        Z_eq(point),  
+        # and now the eqs that fix the params:
+        set_param(beta, beta_val),
+        set_param(c, c_val),
+        set_param(cXY, cXY_val),
+        set_param(D, D_val),
+        set_param(dZ, dZ_val),
+        set_param(f, f_val),
+        set_param(alpha, alpha_val),
+    )
+
+def eqs(param_vals):
+    return lambda pt: eqs_params(pt, param_vals)
+    
+
+"""
+Find solutions
+"""
+
+def var_range(varname):
+    if varname == "cXY" or varname == "beta":
[…]

### 2023-01-30 code/stable_points.py
added fixed point plots

@@ -122,7 +122,7 @@ def fixed_points_functional(varname, var_range, param_inds, param_vals):
     
     
 
-def create_sol_list(param_vals, N = 100):
+def create_sol_list(param_vals, N = 150):
     sol_list = []
     for _ in range(N):
         eqs_at_params = eqs(param_vals)
@@ -231,19 +231,19 @@ def visualize(sol_list):
 def individual_plots(Xs, Ys, Zs, param_vals, varname):
     import matplotlib.pyplot as plt
     
-    plt.scatter(param_vals, Xs)
+    plt.scatter(param_vals, Xs, marker='.')
     plt.title(f"X coordinate of stable points")
     plt.xlabel(f"{varname}")
     plt.ylabel("X")
     plt.show()
 
-    plt.scatter(param_vals, Ys)
+    plt.scatter(param_vals, Ys, marker='.')
     plt.title(f"Y coordinate of stable points")
     plt.xlabel(f"{varname}")
     plt.ylabel("Y")
     plt.show()
 
-    plt.scatter(param_vals, Zs)
+    plt.scatter(param_vals, Zs, marker='.')
     plt.title(f"Z coordinate of stable points")
     plt.xlabel(f"{varname}")
     plt.ylabel("Z")
@@ -263,30 +263,30 @@ def choose_varying_parameter(name_list):
         return "cXY"
 
 def main() -> None:
-    # param_inds = {"beta":0, "c":1, "cXY":2, "D":3, "dZ":4, "f":5, "alpha":6}
-    # param_vals = (0.3, 0.1, 0.5, 1.1, 0.45, 0.5, 0.3)
-    # param_vals_list = list(param_vals)
+    param_vals = (0.3, 0.1, 0.5, 1.1, 0.45, 0.5, 0.3)
     param_vals_dict = {"beta":0.3, "c":0.1, "cXY":0.5, "D":1.1, "dZ":0.45, "f":0.5, "alpha":0.3}
     print(f"Fixed point for params {param_vals_dict}:")
+    sols = create_sol_list(param_vals)
     for sol in sols: print(sol);
 
     ## Uncomment to plot stable points as function of varying parameter:
+    param_inds = {"beta":0, "c":1, "cXY":2, "D":3, "dZ":4, "f":5, "alpha":6}
+    param_vals_list = list(param_vals)
 
-    sols = create_sol_list(param_vals)
-    # varname = "beta"
[…]

### 2023-01-30 manuscript/manuscript.Rmd
added fixed point figure

@@
 There are a variety of fixed points for the equations---around 8 or 9 for wide ranges of parameter values.
-Fig. \ref{fig:tbd} provides a rough idea of the fixed point structure by plotting the bifurcation diagram obtained from varying $D$ and fixing the other parameters.
+Fig. \ref{fig:tbd} provides a rough idea of the fixed point structure by plotting the bifurcation diagram obtained from varying $c_{XY}$ and fixing the other parameters.
+
+\begin{figure}
+  \centering
+  \begin{subfigure}[b]{0.32\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{figures/stX.png}
+    \caption{$X$ component of fixed points.}
+    \label{fig:stx}
+  \end{subfigure}
+  \hfill
+  \begin{subfigure}[b]{0.32\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{figures/stY.png}
+    \caption{$Y$ component of fixed points.}
+    \label{fig:sty}
+  \end{subfigure}
+  \hfill
+  \begin{subfigure}[b]{0.32\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{figures/stZ.png}
+    \caption{$Z$ component of fixed points.}
+    \label{fig:stz}
+  \end{subfigure}
+\end{figure}
 

### 2023-01-30 manuscript/manuscript.Rmd
typo

@@
 There are a variety of fixed points for the equations---around 8 or 9 for wide ranges of parameter values.
-Fig. \ref{fig:tbd} provides a rough idea of the fixed point structure by plotting the bifurcation diagram obtained from varying $c_{XY}$ and fixing the other parameters.
+Fig. \ref{fig:tbd} provides a rough idea of the fixed point structure by plotting the bifurcation diagram obtained from varying $D$ and fixing the other parameters.
 

### 2023-01-30 manuscript/manuscript.Rmd
crossref stable point fig

@@
 There are a variety of fixed points for the equations---around 8 or 9 for wide ranges of parameter values.
-Fig. \ref{fig:tbd} provides a rough idea of the fixed point structure by plotting the bifurcation diagram obtained from varying $D$ and fixing the other parameters.
+Fig. \ref{fig:fixed points} provides a rough idea of the fixed point structure by plotting the bifurcation diagram obtained from varying $D$ and fixing the other parameters.
 
@@
   \end{subfigure}
+  \caption{Numerically-computed fixed points of the three-species deterministic model, eq. \eqref{eq:3d model}.}
+  \label{fig:fixed points}
 \end{figure}

### 2023-01-30 manuscript/manuscript.Rmd
added abstract in

@@
 abstract: |
-  Write something here
+  TBD Change slightly!!!!!
+  Existing methods for optimal control methods struggle to deal with the
+  complexity commonly encountered in real-world systems, including
+  dimensionality, process error, model bias and heterogeneity.  Instead
+  of tackling these complexities directly, researchers have typically
+  sought to find exact optimal solutions to simplified models of the
+  processes in question. When is the optimal solution to a very
+  approximate, stylized model better than an approximate solution to a
+  more accurate model? This question has largely gone unanswered owing
+  to the difficulty of finding even approximate solutions in the case of
+  complex models.  Our approach draws on recent algorithmic and
+  computational advances in deep reinforcement learning. These methods
+  have hitherto focused on problems in games or robotic mechanics, which
+  operate under precisely known rules. We demonstrate the ability for
+  novel algorithms using deep neural networks to successfully
+  approximate such solutions (the ‘‘policy function’’ or control rule)
+  without knowing or ever attempting to infer a model for the process
+  itself. This powerful new technique lets us finally begin to answer
+  the question. We show that in many but not all cases, the optimal
+  policy for a carefully chosen over-simplified model can still
+  out-perform these novel algorithms trained to find approximate
+  solutions to simulations of a realistically complex system. Comparing
+  these two approaches can lead to insights in the importance of
+  real-world features of observation and process error, model biases,
+  and heterogeneity.
 
@@
 
-Eq. \eqref{eq:may} is an interesting study case of a *tipping point* (saddle-node bifurcation) (see Fig. \ref{fig:may stable}).
+Eq. \eqref{eq:may} is an interesting study case of a *tipping point* (saddle-node bifurcation) (see Fig. \ref{fig:may}).
 Holding the value of $\beta$ fixed, for intermediate values of $H$ there exist two stable fixed points for the state $\hat X_t$ of the system, these two attractors separated by an unstable fixed point.
@@
 
-\begin{figure}
-  \centering
-  \includegraphics[width = 0.8\linewidth]{figures/MayStablePoints.pdf}
-  \caption{
-    \label{fig:may stable}
-    The fixed point diagram for the model [@may77] associated to $f_{\text{May}}$ as a function of varying the parameter $\beta H$.
-    That is, these are the zeroes of the function $f_{\text{May}}$. 
-    Stable fixed points (also known as *attractors*) are plotted using a solid line, while the unstable fixed point is shown as a dotted line.
-  }
-\end{figure}
+```{r}
+#| label: may
+#| fig.cap: "The fixed point diagram for the model associated to $f_{\\text{May}}$ as a function of varying the parameter $\\beta H$.
+#| That is, these are the zeroes of the function $f_{\\text{May}}$. 
+#| Stable fixed points (also known as *attractors*) are plotted using a solid line, while the unstable fixed point is shown as a dotted line."
+#| out.width: "6.5in"
+
+knitr::include_graphics("figures/MayStablePoints.pdf")
+```
 

### 2023-01-30 code/stable_points.py
mini

@@ -5,14 +5,6 @@ import numpy as np
 import time
 import sys
 
-def timer(func):
-    def wrapper(*args, **kwargs):
-        t = time.time()
-        r = func(*args, **kwargs)
-        print(f"Time spent on {func.__name__}: {time.time()-t}")
-        return r
-    return wrapper
-
 def L(v):
     return v * (1 - v)
 
@@ -85,7 +77,6 @@ def var_range(varname):
     elif varname == "f":
         return np.linspace(0.05, 0.95, 150)
 
-@timer
 def fixed_points(varname, var_range, param_inds, param_vals):
     var_list = []
     sol_list = []
@@ -97,29 +88,6 @@ def fixed_points(varname, var_range, param_inds, param_vals):
         if s is not None:
             sol_list += s
     return var_list, sol_list
-
-@timer    
-def fixed_points_functional(varname, var_range, param_inds, param_vals):
-    """
-    var_range -> change_param_val {x} -> create_sol_list {x} -> flatten
-                                     \-> param_list {x} ------> flatten
-    """
-    def change_tuple_at_ind_generator(tup, ind):
-        return lambda x: reassign_tuple_entry(tup, ind, x)
-    change_param_val = change_tuple_at_ind_generator(param_vals, param_inds[varname])
-    def create_sol_list_and_repeated_param_val(param_vals, index):
-        sols = create_sol_list(param_vals)
-        return sols, len(sols)*[param_vals[index]] 
-    def univariate_create_sol_list_and_etc(index):
-        return lambda x: create_sol_list_and_repeated_param_val(x, index)
-    sol_list_and_repeated_params = univariate_create_sol_list_and_etc(param_inds[varname])
-    
-    param_vals_seq = map(change_param_val, var_range)
-    sol_param_seq = map(sol_list_and_repeated_params, param_vals_seq)
-    sol_param_seq_unzipped = list(zip(*sol_param_seq))[0]
-    sols, params = sol_param_seq_unzipped [0], sol_param_seq_unzipped[1]
-    return params, sols
-    
     
 
 def create_sol_list(param_vals, N = 150):

### 2023-01-30 manuscript/manuscript.Rmd
plots in bw

@@
 abstract: |
-  TBD Change slightly!!!!!
   Existing methods for optimal control methods struggle to deal with the
@@
   itself. This powerful new technique lets us finally begin to answer
-  the question. We show that in many but not all cases, the optimal
-  policy for a carefully chosen over-simplified model can still
-  out-perform these novel algorithms trained to find approximate
-  solutions to simulations of a realistically complex system. Comparing
-  these two approaches can lead to insights in the importance of
-  real-world features of observation and process error, model biases,
-  and heterogeneity.
+  the question. We showcase this approach by applying it to the control
+  of a highly non-linear three-variable model for a fishery. Here, the
+  reinforcement learning agent discovers an effective simplification of
+  the problem to obtain a simple, interpretable control rule. Moreover,
+  the policy obtained is both more profitable and more sustainable than
+  any constant mortality policy---the standard family of policies considered
+  in fishery management.
 
@@
   geom_vline(aes(xintercept=0.8*best_action), color="violetred1") +
-  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
+  theme_bw() + 
+  theme(panel.grid.major = element_blank(), 
+        panel.grid.minor = element_blank(),
+        aspect.ratio = 1)
 
@@
   xlim(0,2) +
-  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
-  xlab("X") + ylab("harvest")
+  xlab("X") + ylab("harvest") +
+  theme_bw() +
+  theme(panel.grid.major = element_blank(), 
+        panel.grid.minor = element_blank(), 
+        aspect.ratio = 1)
 
@@
         labs(y = "abundance", x = "time") +
+        theme_bw() +
         theme(aspect.ratio = 2/3)
@@
       labs(y = "episode length", x = "repetition") +
+      theme_bw() + 
       theme(axis.text.x = element_blank())
@@
       labs(y = "episode harvest", x = "repetition") +
+      theme_bw() + 
       theme(axis.text.x = element_blank())
@@
   geom_point() +
+  theme_bw() + 
   theme(aspect.ratio = 1/1) +
@@
   ylim(0,0.85) +
+  theme_bw() + 
   ggtitle("Opt. Const. Mortality")
@@
   ylim(0,0.85) +
+  theme_bw() + 
   ggtitle("80% Opt. Const. Mortality")
@@
   ylim(0,0.85) +
+  theme_bw() + 
   ggtitle("DRL")
@@
   ylim(0,0.85) +
+  theme_bw() + 
   ggtitle("Opt. Const. Escapement")
@@
         labs(y = "abundance", x = "time") +
+        theme_bw() + 
         theme(aspect.ratio = 0.5)

### 2023-01-30 code/stable_points.py
updated stable_points.py

@@ -251,10 +251,10 @@ def main() -> None:
     individual_plots(Xs, Ys, Zs, param_vals, varname)
     
     df = DataFrame(zip(param_vals,Xs,Ys,Zs), columns = [varname, "X", "Y", "Z"])
-    df.to_csv(f'FixedPonts_{varname}.csv', index = False)
+    df.to_csv(f'data/FixedPonts_{varname}.csv.xz', index = False)
 
     df_X = DataFrame(zip(param_vals, Xs), columns = [varname, "X"])
-    df.to_csv(f"FixedPoints_{varname}_X.csv", index=False)
+    df.to_csv(f"data/FixedPoints_{varname}_X.csv.xz", index=False)
 
 if __name__ == "__main__":
     main()

### 2023-01-30 manuscript/manuscript.Rmd
fig stable point caption
updated to google docs intro

@@
 
-<!--
-Existing methods for optimal control methods struggle to deal with the complexity commonly encountered in real-world systems, including dimensionality, process error, model bias and heterogeneity. 
-Instead of tackling these complexities directly, researchers have typically sought to find exact optimal solutions to simplified models of the processes in question. 
-When is the optimal solution to a very approximate, stylized model better than an approximate solution to a more accurate model? 
-This question has largely gone unanswered owing to the difficulty of finding even approximate solutions in the case of complex models.  
-Our approach draws on recent algorithmic and computational advances in deep reinforcement learning. 
-These methods have hitherto focused on problems in games or robotic mechanics, which operate under precisely known rules. 
-We demonstrate the ability for novel algorithms using deep neural networks to successfully approximate such solutions (the "policy function" or control rule) without knowing or ever attempting to infer a model for the process itself.
-This powerful new technique lets us finally begin to answer the question.
-We show that in many but not all cases, the optimal policy for a carefully chosen over-simplified model can still out-perform these novel algorithms trained to find approximate solutions to simulations of a realistically complex system.
-This illustrates the promise of combining complex, realistic simulations with emerging machine-learning techniques to improve our understanding and management of ecological systems.
--->
-
-Much effort has been spent grappling with the complexity of our natural world in contrast to relative simplicity of the models we use to understand it.
-Now, as we grapple with the manifold consequences of human impacts on that world, we are faced with difficult questions of figuring out how best to respond when our knowledge and models fall so far short of the complexity our actions have disrupted.
-Heroic amounts of data and computation are being brought to bear on developing better, more complex, more realistic models of our environments and ecosystems, yet these models remain disconnected or oversimplify the space of actions they seek to inform.
-In this paper we explore a different tack. 
-Here we explore the potential of a emerging data-driven machine learning technique, not to better capture the complexity of the natural processes, but rather, to embrace the complexity of the potential _policies_ that might be used to manage them.
+Much effort has been spent grappling with the complexity of our natural world in contrast to the relative simplicity of the models we use to understand it.
+Heroic amounts of data and computation are being brought to bear on developing better, more realistic models of our environments and ecosystems, in hopes of improving our capacity to address the many planetary crises.
+ But despite these efforts and advances, we remain faced with the difficult task of figuring out how best to respond to these crises.
+While simplified models have historically allowed for exploration of large decision spaces,, the new wave of rich models are applied to highly oversimplified descriptions of potential actions they seek to inform.
+For instance, Global Circulation Models (GCMs) such as HadOM3 [CITE] model earth's climate using 1.5M variables, while the comparably vast potential action space is modeled as one of six emissions scenarios from the IPCC [Raihi et al., 2017]
+
 Even as our research community develops simulations of the natural world that fit only in supercomputers, we analyze a space of policies that would fit on index cards.
-
-<!-- Why fisheries-focused examples -- maybe we can cut/merge this with paragraph above -->
-There are myriad examples from climate scenarios to protected areas, we will illustrate this with a specific example in the management of marine fisheries, which is both an important issue to itself and a frequent crucible in which methods of ecological management have been fired before extending to other fields, and where the space of possible policies is arguably already more richly developed than most.
-Fisheries are an essential natural resource that provide the primary source of protein for one in every four humans, and have faced widely documented declines due to over-fishing [@Worm2006; @Costello2016].
-Because of this importance, fisheries also represent some of the longest and best studied timeseries of population dynamics available to ecologists, as well as a correspondingly rich area for the development of theory and practice.
-Methods developed in the study and management of fisheries are frequently applied to other areas of ecological management, from concepts such as maximum sustainable yield [], optimal control under observation uncertainty [@Memarzadeh2019], to adaptive management [@Walters], management strategy evaluation [@Punt], to the studies of conservation policy and international negotiations [@].
-
-Mathematically, the ecological management may be understood as an optimal control problem - a manager must determine a sequence of actions (a "policy") to optimize some desired objective(s) and constraints (e.g.
-conservation goals, natural resource extraction, economic and social constraints.)
-Given an appropriate representation of the underlying mechanistic processes involved, such sequential decision problems may be posed more precisely as Markov Decision Process [MDP, @Marescot2013; @Memarzadeh2019].
-Such problems are notoriously difficult owing not only to the complexity of those underlying mechanisms alone, but to the exponentially branching consequences the decision-maker's actions have upon that system -- a game of chess is not won by thinking only one move ahead at a time.
-Thus, while MDP problems have been studied in a conservation context since they were first introduced more than six decades ago, their application has been highly constrained to simple scenarios.
-Meanwhile, a class of machine learning techniques designed for MDP problems known as reinforcement learning (RL), has cracked open the door for exploring complex strategies in more realistic settings [@rl_intro].
-
-- This is not a tutorial or a sales pitch for RL - which brings at least as many difficulties as advantages -- including being hard to use, easy to get stuck on far from optimal performance, which itself may be hard to diagnose, as we have already discussed in our tutorial.
-The method
-
-has through the use of clever algorithms, deep neural networks, and massive amounts of data,
-
-<!-- A sweeping summary of fisheries decision theory, don't bite me-->
-To understand the importance of a potential application of novel data-driven methods to the real world, it is most instructive to have a clearer picture of how models are currently used to inform ecological management in fisheries.
-In any issue as large and contentious as fisheries management, it may be far more difficult to provide a generalization  about how things actually work than to make sweeping recommendations about how it ought to.
-All the same, it is our firm belief that any researcher seeking to do the latter should first attempt to understand the former, so we beg the forbearance of any specialists in the practice as we seek to provide a concise overview of fisheries management today.
-Here, we focus specifically on practices in data rich, 'well-managed' fisheries, which, despite some issues, may be among the best examples of ecological management at continental scale that is largely driven by scientific processes, models and data.
-
-<!-- Fisheries population models are rich but decisions aren't, mostly constant mortality -->
-While fisheries stock assessment models are quite rich, the choice of possible strategies remains motivated by intution from much simpler, analytically tractable one-dimensional models.
-This fact is not always obvious at cursory inspection, and can be met with cynicism even among professional fisheries managers -- after all, typical stock assessment models have over 100 parameters (e.g.
-see the stock assessments referenced from in @RAMLegacy) and best practices such as management strategy evaluation (MSE) evaluate and select potential policies by comparing across large and diverse simulations [e.g.
-@Punt2016].
-However, the space of policies considered under such approaches is necessarily far smaller than the set of all possible action sequences.
-Instead, fisheries management approaches such as MSE rely on theory based on more tractable models to restrict the search space of possible policies.  
-Specifically, most fisheries policies are framed in terms of a *constant mortality* policy, (sometimes a simple piece-wise linear function, i.e.
-a single linear segment, is introduced as a precaution a very low stock sizes), where fishing 'mortality' is defined as the ratio of harvest (catch) to stock size, $F = H_t/x_t$.
-
-```{r}
-#| label: textbox
-#| out.width: "1\\linewidth"
-
-knitr::include_graphics("figures/textbox.png")
-```
-
-The intuition for a constant mortality strategy is simple and well established
-
-Constant mortality policies have a strong basis in theory as well as practical justifications.
-
-<!-- In contrast to constant mortality, the possible decision space is huge-->
-The management of any ecological system can be framed in terms of a Sequential Decision Process - 
-the manager observes the ecosystem state ($x_t$) and decides on action ($a_t$) to take (including potentially no action), potentially receiving any costs or benefits associated with that action and and state (reward $r_t$  $r_t = U(x_t,a_t)$),
-while the system evolves into a new state in response $x_{t+1} = f(x_t, a_t)$.
-When the probability of the future state can be determined by the current state and action, we say the system is Markovian and the process a Markov Decision Process (MDP).
-As in a game of chess, the space of possible sequences of actions a manager might choose over any realization of the Markov process is enormous.
-When the action and state space are discrete and sufficiently small, computational approaches such as dynamic programming may be used to guarantee an exhaustive search for a single optimal policy.
-Dynamic programming scales very poorly to larger state and action spaces.
-
-Deep reinforcement learning has opened up increasingly compelling alternative
-
-Effort $Q = E X_t$
-
-Fisheries policies are typically implemented in terms of *catch quotas*, i.e. a limit to the number of metric tonnes of fish that can be caught during a given season.
-
-and/or *effort controls*, restrictions on the fishing effort -- in numbers of boats, time of year, or days at sea
-
-# Figures brainstorm
-
-Figure 0: contrast 1D vs 3D strategies (see commented out diagram).
+Similar combinations of rich process models and highly simplified decision models (often not even given the status of 'model') are common [e.g., CITE].
+Modeling the potential action space as one of a handful of discrete scenarios is sometimes a well justified acknowledgement of the constraints faced by real-world decision-makers – particularly in the context of multilateral decisions –
+and may seem to reflect a division of responsibilities between 'scientists' modeling the 'natural processes' and policy-makers who make the decisions.
+But, more often, this simplification of decision choices is simply mathematically or conceptually convenient.
+<!-- there are questions of policy scale hiding here -- e.g.
+local regulators of a specific industry obviously face a different action space than national legislatures or international diplomats -->
+This simplification reflects trade-offs between tractablity and complexity at the basis of any mathematical modeling -- if we make both the state space and action space too realistic, the problem of finding the best sequence of actions quickly becomes intractable.
+However, emerging data-driven methods from machine learning offer a new choice -- algorithms that can find good strategies in such problems, but at the cost of opacity.
+
+In this paper, we focus on a well-developed application of model-based management of the natural world that has long illustrated the trade-offs between model complexity and policy complexity: the management of marine fisheries.
+Fisheries management is both an important issue to itself and a rich & frequent test-bed of ecological management more generally.
+Fisheries are an essential natural resource that provide the primary source of protein for one in every four humans, and have faced widely documented declines due to over-fishing [@Worm2006, @Costello2016].
+Fisheries management has  roots in both the fields of _ecosystem management_ and _natural resource economics_.
+Both fields might trace their origins to the notion of maximum sustainable yield (MSY), introduced independently by a fisheries ecologist @Schaefer1954 and the economist @Gordon1954 in the same year.
+From this shared origin, each field would depart from the simplifying assumptions of the Gordon-Schaefer model in divergent ways, leading to different techniques for deriving policies from models.The heart of the management problem is easily understood: a manager seeks to set quotas on fishing that will ensure the long-term profitability and sustainability of the industry.
+Mathematical approaches developed over the past century may be roughly divided between these two fields: (A), ecologists, focused on ever more realistic models of the biological processes of growth and recruitment of fish while considering relatively stylized suite of potential management strategies, and (B) economists, focused on far more stylized models of the ecology while exploring a far less constrained set of possible policies.
+The economist's approach can be characterized by the mathematics of a Markov decision process [MDP @Clark1973; @Clark1990; @Marescot2013], in which the decision-maker must observe the stock each year and recommend a possible action.
+In this approach, the policy space that must be searched is exponentially large -- for a management horizon of T decisions and a space of N actions, the number of possible policies is N^T.
+In contrast, fisheries ecologists and ecosystem management typically search a space of policies that does not scale with the time horizon.
+Under methods such as "Management Strategy Evaluation" (MSE, [@Punt2016]) a manager identifies a candidate set of "strategies" a priori, and then compares the performance of each strategy over a suite of simulations to determine which strategy gives the best outcome (i.e.
+best expected utility).
+This approach is far more amenable to complex simulations of fisheries dynamics and more closely corresponds to how most marine fishing quotas are managed today in the United States [see stock assessments documented in @RAMLegacyDB].
+<!-- Do we say here that these strategies retain the 'constant mortality' intuition of @Schaefer? -->
+
+Recent advances in machine learning may allow us to now once again bridge these approaches, while also bringing new challenges of their own.
+Novel data-driven methods have allowed these models to evolve into ever more complex and realistic simulations used in fisheries management, where models with over 100 parameters are not uncommon [@RamLegacyDB].
+Constrained by computational limits, MDP approaches have been intractable on suitably realistic models and largely confined to more academic applications [@Costello2016].
+However, advances _Deep Reinforcement Learning_, (DRL) a sub-field of machine learning, have recently demonstrated remarkable performance in a range of such MDP problems, from video games [@atari] to fusion reactions [@fusion] to the remarkable dialog abilities of ChatGPT [@chatgpt].
+RL methods also bring many challenges of their own: being notoriously difficult to train and evaluate, immense computational costs, and frequent challenges with transfer-ability or reproducibility.
+A review of all these issues is beyond our scope but can be found elsewhere [@rl_intro; @millie_AM-RL].
+Here, though, we will focus on the issue of opacity and interpretability raised by these methods.
+In contrast with optimization algorithms currently used in either ecosystem management or resource economics, RL algorithms have no guarantees of or metrics for convergence to an optimal solution.
+In general, one can only assess the performance of these black box methods relative to alternatives.
+
+<!--NEED some kind of map paragraph here saying what's ahead, what we show  -->
+In this article, we We compare the performance of a DRL solution to that of a management strategy evaluation across potential 'constant mortality' policies.\footnote{A repository with all the relevant code to reproduce our results may be found at [@our-repo] in the ‘‘code’’ subdirectory.
+The data used is found in the ‘‘data’’ subdirectory, but the user may use the code provided to generate new data sets.}
+We consider the problem of devising harvest strategies in a three species ecosystem (Figure 1B).
+We show that, somewhat surprisingly, the DRL policy closely resembles a *bang-bang* solution (see Text Box 1).
+Bang-bang solutions are known to be optimal for a variety of single-species models, however, there is no a-priori reason why they should be optimal for higher-dimensional systems.
+The DRL policy drives an oscillation process in the system and, in this way, maintains the long-term sustainability of fishing.
+In contrast, constant mortality policies overfish at low populations and underfish (relative to the DRL solution) at high populations.
+This leads to a tradeoff when leveraging constant mortality strategies: higher constant mortality rates lead to high profitability at the expense of a higher probability of overfishing to extinction, while lower constant mortality rates attain stability at the expense of profit.
+
+The DRL approach side-steps this tradeoff by optimizing over a more complex family of possible policies—policies parametrized by a neural network, as opposed to policies labeled by a single parameter (the mortality rate).
+We thus observe that the DRL solution leads to a higher profitability *and* stability as compared to constant mortality policies.
 
@@
 ```
-
-<!-- Reality __
-  |        \
-  v         \
-1-D Model   More expressive model
-  |                  |
-  v                  v
-Optimal        Suboptimal, but
-solution   well-performing solution -->
-
-Figure 1: 1-D and 3-D model conceptual figure. something about the objective / decision
-Figure 2: Stability / multistability in the 1D and 3D models.  state space + time views
-Figure 3: The 1-D optimal management solution.  'constant escapement' intuition etc
-
-RL figures:
-- schematic of RL (a-la previous marcus paper Fig 1)
-- Neural network optimization figure
-
-Results figures:
-- timeseries example of management under the RL policy.  probably compare to managing under 1-D solution / rule-of-thumb methods
-- state-space view of management doughnut 
-- visualization / encapsulation of the policy heatmaps, slices, policy vs position along ellipse 
-- reward plot over time (comparing methods)
 
@@
   \end{subfigure}
-  \caption{Numerically-computed fixed points of the three-species deterministic model, eq. \eqref{eq:3d model}.}
+  \caption{Numerically-computed fixed points of the three-species deterministic model, eq. \eqref{eq:3d model}.
+  Here, the parameter $D$ was varied between 0.75 and 1.25, and the three components of the fixed points are plotted separately, as a function of $D$.}
   \label{fig:fixed points}

### 2023-01-30 manuscript/manuscript.Rmd
mini intro messups with copying from google docs

@@
   - \usepackage{etoolbox}
+  - \usepackage{url}
   - \linenumbers
@@
 
-Recent advances in machine learning may allow us to now once again bridge these approaches, while also bringing new challenges of their own.
+Recent advances in machine learning may allow us to once again bridge these approaches, while also bringing new challenges of their own.
 Novel data-driven methods have allowed these models to evolve into ever more complex and realistic simulations used in fisheries management, where models with over 100 parameters are not uncommon [@RamLegacyDB].
@@
 <!--NEED some kind of map paragraph here saying what's ahead, what we show  -->
-In this article, we We compare the performance of a DRL solution to that of a management strategy evaluation across potential 'constant mortality' policies.\footnote{A repository with all the relevant code to reproduce our results may be found at [@our-repo] in the ‘‘code’’ subdirectory.
+In this article, we compare the performance of a DRL solution to that of a management strategy evaluation across potential 'constant mortality' policies.\footnote{A repository with all the relevant code to reproduce our results may be found at \url{https://github.com/boettiger-lab/approx-model-or-approx-soln} in the ‘‘code’’ subdirectory.
 The data used is found in the ‘‘data’’ subdirectory, but the user may use the code provided to generate new data sets.}
@@
 We have included only three decimal digits for readability.
-These fixed points may be generated by running the ‘‘stable\_points.py’’ script in [@repo].
+These fixed points may be generated by running the ‘‘stable\_points.py’’ script in \url{https://github.com/boettiger-lab/approx-model-or-approx-soln}.
 }
@@
 
-We used the Ray framework [@ray_package] for training, in particular we used a Ray PPOConfig object to build the algorithm using the default values for all hyper-parameters (cf.
-[@repo] for the code used).
+We used the Ray framework\footnote{\url{https://docs.ray.io/}} for training, in particular we used a Ray PPOConfig object to build the algorithm using the default values for all hyper-parameters.
 The agent was trained for 200 iterations of the PPO optimization step.

### 2023-01-30 manuscript/manuscript.Rmd
fig 1 caption

@@
 While simplified models have historically allowed for exploration of large decision spaces,, the new wave of rich models are applied to highly oversimplified descriptions of potential actions they seek to inform.
-For instance, Global Circulation Models (GCMs) such as HadOM3 [CITE] model earth's climate using 1.5M variables, while the comparably vast potential action space is modeled as one of six emissions scenarios from the IPCC [Raihi et al., 2017]
+For instance, Global Circulation Models (GCMs) such as HadOM3 [CITE] model earth's climate using 1.5M variables, while the comparably vast potential action space is modeled as one of six emissions scenarios from the IPCC [@riahi2017shared].
 
@@
 Recent advances in machine learning may allow us to once again bridge these approaches, while also bringing new challenges of their own.
-Novel data-driven methods have allowed these models to evolve into ever more complex and realistic simulations used in fisheries management, where models with over 100 parameters are not uncommon [@RamLegacyDB].
+Novel data-driven methods have allowed these models to evolve into ever more complex and realistic simulations used in fisheries management, where models with over 100 parameters are not uncommon [@RAMLegacyDB].
 Constrained by computational limits, MDP approaches have been intractable on suitably realistic models and largely confined to more academic applications [@Costello2016].
-However, advances _Deep Reinforcement Learning_, (DRL) a sub-field of machine learning, have recently demonstrated remarkable performance in a range of such MDP problems, from video games [@atari] to fusion reactions [@fusion] to the remarkable dialog abilities of ChatGPT [@chatgpt].
+However, advances _Deep Reinforcement Learning_, (DRL) a sub-field of machine learning, have recently demonstrated remarkable performance in a range of such MDP problems, from video games [@atari1; @atari2] to fusion reactions [@fusion1; @fusion2] to the remarkable dialog abilities of ChatGPT [@chatgpt].
 RL methods also bring many challenges of their own: being notoriously difficult to train and evaluate, immense computational costs, and frequent challenges with transfer-ability or reproducibility.
-A review of all these issues is beyond our scope but can be found elsewhere [@rl_intro; @millie_AM-RL].
+A review of all these issues is beyond our scope but can be found elsewhere [@rl_intro; @millie-1; @millie-2].
 Here, though, we will focus on the issue of opacity and interpretability raised by these methods.
@@
 #| fig.align: 'center'
-#| fig.cap:  '(A) Classical management strategies often have the common aspect of reducing the complex dynamics of the fishery ecosystem to a single equation governing the harvested population (say, $V1$).
-where the interaction between $V1$ and its environment is summarized to a set of parameters, such as the growth rate $r$, and the carrying capacity $K$.
+#| fig.cap:  '(A) Classical management strategies often have the common aspect of reducing the complex dynamics of the fishery ecosystem to a single equation governing the harvested population (say, $X$).
+where the interaction between $X$ and its environment is summarized to a set of parameters, such as the growth rate $r$, and the carrying capacity $K$.
 An advantage of one-dimensional approaches is that the optimal policy is often known exactly, and, moreover, is intuitive.
 For example, in the logistic equation shown, the maximal sustainable yield of the system is attained at K/2.
-#| (B) RL allows us to solve for quantitative decision strategies considering more complex ecosystem dynamics.
-For example, in this paper, we find harvest strategies for a three-species model using a model-free RL algorithm, PPO.
-#| (C) We show how the RL-based solution that considers all three species finds harvest policies with higher average reward and with (D) lower probability of harvest species collapse than the classical approaches, only considering the dynamics of V1.'
+#| (B) In contrast to this, we study the problem of managing a three species fishery, where we harvest species $X$.
+#| In addition to $X$, there is a species $Y$ which competes with $X$ for resources, and a species $Z$ which preys on $X$ and $Y$.
+#| We compare two approaches here. 
+#| The first (Harvest $B_{\mathrm{i}}$) is based on the intuition obtained from 1D control: we use simulated data to empirically determine which constant mortality policy is optimal for controlling this fishery.
+#| The second (Harvest $B_{\mathrm{ii}}$) is based on RL.
+#| Here, the policy function is expressed as a neural network and the algorithm uses a variant of \emph{gradient ascent} (namely, \emph{Proximal Policy Optimizaton (PPO)}) to optimize this network.
+#| Both approaches we compare are \emph{model-free:} the optimization algorithm treats the model as a black box, only using its input-output statistics.
+#| This way, the optimal policy is not computed from the values of the model parameters, but rather by comparing how policies perform in controlling the system.'
 #| out.width: '6.5in' 

### 2023-01-30 manuscript/manuscript.Rmd
updated paragraph in RL section saying what we used

@@
 -->
-Deep reinforcement learning uses this strategy, focusing on function families parameterized by neural networks. (See Fig. \ref{tbd}.)
+Deep reinforcement learning uses this strategy, focusing on function families parameterized by neural networks.
+(See Fig.
+1 and App.
+A in [@rl-intro] for a conceptual introduction to the use of reinforcement learning in the context of conservation decision making.)
+<!--(See Fig. \ref{tbd}.)-->
 <!-- The figure should kinda explain how the parametrization arises, and how gradient ascent is used. Maybe namedrop backpropagataion for the sake of usefulness to the reader? -->
@@
 Since our state and observation spaces are continuous, we will focus on deep reinforcement learning throughout this paper.
-We use the *proximal policy optimization (PPO)* algorithm to optimize $\pi$.
-Within the RL literature there is a wealth of algorithms from which to choose from, each with its pros and cons.
-Here we have focused on a single one of these for the sake of particularity---to draw a clear comparison between the RL-based and the classical fishery management approaches.
+Within the RL literature there is a wealth of algorithms from which to choose from to optimize $\pi$, each with its pros and cons.
+Most, if not all, of these are based on gradient ascent by using the technique of *back-propagation* to efficiently compute the gradient.
+Here we have used only one such algorithm (*proximal policy optimization (PPO)*) for the sake of particularity---to draw a clear comparison between the RL-based and the classical fishery management approaches.
 In practice, further improvements can be expected by a careful selection of the optimization algorithm. 
+(See, e.g., [@franccois2018introduction] for an overview of different optimization schemes used in DRL.)
 

### 2023-01-30 manuscript/manuscript.Rmd
mini

@@
 Fisheries management has  roots in both the fields of _ecosystem management_ and _natural resource economics_.
-Both fields might trace their origins to the notion of maximum sustainable yield (MSY), introduced independently by a fisheries ecologist @Schaefer1954 and the economist @Gordon1954 in the same year.
+Both fields might trace their origins to the notion of maximum sustainable yield (MSY), introduced independently by a fisheries ecologist [@Schaefer1954] and the economist [@Gordon1954] in the same year.
 From this shared origin, each field would depart from the simplifying assumptions of the Gordon-Schaefer model in divergent ways, leading to different techniques for deriving policies from models.The heart of the management problem is easily understood: a manager seeks to set quotas on fishing that will ensure the long-term profitability and sustainability of the industry.

### 2023-01-30 manuscript/manuscript.Rmd
tuning plot

@@
   theme(aspect.ratio = 1/1) + 
-  geom_vline(aes(xintercept=best_action), color="seagreen2") + 
-  geom_vline(aes(xintercept=0.8*best_action), color="violetred1") +
+  geom_vline(aes(xintercept=best_action), color="seagreen3", size=1) + 
+  geom_vline(aes(xintercept=0.8*best_action), color="violetred3", size=1) +
   theme_bw() + 
@@
 # p + stat_function(fun = linear_f_gen(0.5)) + xlim(0,2)
-lines <- p + stat_function(fun = linear_f_gen(0.01)) + 
-  stat_function(fun = linear_f_gen(0.02)) +
-  stat_function(fun = linear_f_gen(0.03)) +
-  stat_function(fun = linear_f_gen(0.8*best_action), col = "violetred") +
-  stat_function(fun = linear_f_gen(best_action), col = "seagreen2") +
-  stat_function(fun = linear_f_gen(0.055)) + 
-  stat_function(fun = linear_f_gen(0.065)) +
+lines <- p + stat_function(fun = linear_f_gen(0.009)) + 
+  stat_function(fun = linear_f_gen(0.018)) +
+  stat_function(fun = linear_f_gen(0.027)) +
+  stat_function(fun = linear_f_gen(0.8*best_action), col = "violetred3", size=1) +
+  stat_function(fun = linear_f_gen(best_action), col = "seagreen3", size=1) +
+  stat_function(fun = linear_f_gen(0.054)) + 
+  stat_function(fun = linear_f_gen(0.063)) +
+  stat_function(fun = linear_f_gen(0.072)) +
+  stat_function(fun = linear_f_gen(0.081)) +
+  stat_function(fun = linear_f_gen(0.090)) +
   xlim(0,2) +

### 2023-01-30 manuscript/manuscript.Rmd
updated fig 1 caption

@@
 #| fig.align: 'center'
-#| fig.cap:  '(A) Classical management strategies often have the common aspect of reducing the complex dynamics of the fishery ecosystem to a single equation governing the harvested population (say, $X$).
+#| fig.cap:  '(A) Classical management strategies are often derived by reducing the complex dynamics of the fishery ecosystem to a single equation governing the harvested population (say, $X$).
 where the interaction between $X$ and its environment is summarized to a set of parameters, such as the growth rate $r$, and the carrying capacity $K$.
@@
 #| In addition to $X$, there is a species $Y$ which competes with $X$ for resources, and a species $Z$ which preys on $X$ and $Y$.
-#| We compare two approaches here. 
+#| We compare two approaches solving harvest strategies in this system. 
 #| The first (Harvest $B_{\mathrm{i}}$) is based on the intuition obtained from 1D control: we use simulated data to empirically determine which constant mortality policy is optimal for controlling this fishery.
-#| The second (Harvest $B_{\mathrm{ii}}$) is based on RL.
-#| Here, the policy function is expressed as a neural network and the algorithm uses a variant of \emph{gradient ascent} (namely, \emph{Proximal Policy Optimizaton (PPO)}) to optimize this network.
+#| The second (Harvest $B_{\mathrm{ii}}$) is an RL policy, where the policy function is expressed as a neural network and the algorithm uses a variant of \emph{gradient ascent} (namely, \emph{Proximal Policy Optimizaton (PPO)}) to optimize this network.
 #| Both approaches we compare are \emph{model-free:} the optimization algorithm treats the model as a black box, only using its input-output statistics.

### 2023-01-30 manuscript/manuscript.Rmd
mini

@@
   theme_bw() +
-  theme(panel.grid.major = element_blank(), 
-        panel.grid.minor = element_blank(), 
+  theme(#panel.grid.major = element_blank(), 
+        #panel.grid.minor = element_blank(), 
         aspect.ratio = 1)
@@
 For example, tools such as *long short-term memories* [@lstm], *curriculum learning* [@curriculum-learning] and *open-ended learning* [@open-ended] can be incorporated when faced with parametric noise or drift, or generally uncertainties about the agent's environment.
-These techniques can be used to obtain agents who perform well over a variety of different environments [@open-ended].
+These techniques can be used to obtain agents who perform well over a variety of different environments.
 <!-- ML: Could also use LSTMs for parametric drift.

### 2023-01-30 manuscript/manuscript.Rmd
mini changes intro

@@
 abstract: |
-  Existing methods for optimal control methods struggle to deal with the
+  Existing methods for optimal control struggle to deal with the
   complexity commonly encountered in real-world systems, including
@@
  But despite these efforts and advances, we remain faced with the difficult task of figuring out how best to respond to these crises.
-While simplified models have historically allowed for exploration of large decision spaces,, the new wave of rich models are applied to highly oversimplified descriptions of potential actions they seek to inform.
+While simplified models have historically allowed for exploration of large decision spaces, the new wave of rich models are applied to highly oversimplified descriptions of potential actions they seek to inform.
 For instance, Global Circulation Models (GCMs) such as HadOM3 [CITE] model earth's climate using 1.5M variables, while the comparably vast potential action space is modeled as one of six emissions scenarios from the IPCC [@riahi2017shared].
@@
 From this shared origin, each field would depart from the simplifying assumptions of the Gordon-Schaefer model in divergent ways, leading to different techniques for deriving policies from models.The heart of the management problem is easily understood: a manager seeks to set quotas on fishing that will ensure the long-term profitability and sustainability of the industry.
-Mathematical approaches developed over the past century may be roughly divided between these two fields: (A), ecologists, focused on ever more realistic models of the biological processes of growth and recruitment of fish while considering relatively stylized suite of potential management strategies, and (B) economists, focused on far more stylized models of the ecology while exploring a far less constrained set of possible policies.
+Mathematical approaches developed over the past century may be roughly divided between these two fields: (A) ecologists, focused on ever more realistic models of the biological processes of growth and recruitment of fish while considering relatively stylized suite of potential management strategies, and (B) economists, focused on far more stylized models of the ecology while exploring a far less constrained set of possible policies.
 The economist's approach can be characterized by the mathematics of a Markov decision process [MDP @Clark1973; @Clark1990; @Marescot2013], in which the decision-maker must observe the stock each year and recommend a possible action.
-In this approach, the policy space that must be searched is exponentially large -- for a management horizon of T decisions and a space of N actions, the number of possible policies is N^T.
+In this approach, the policy space that must be searched is exponentially large -- for a management horizon of T decisions and a space of N actions, the number of possible policies is $N^T$.
 In contrast, fisheries ecologists and ecosystem management typically search a space of policies that does not scale with the time horizon.
@@
 However, advances _Deep Reinforcement Learning_, (DRL) a sub-field of machine learning, have recently demonstrated remarkable performance in a range of such MDP problems, from video games [@atari1; @atari2] to fusion reactions [@fusion1; @fusion2] to the remarkable dialog abilities of ChatGPT [@chatgpt].
-RL methods also bring many challenges of their own: being notoriously difficult to train and evaluate, immense computational costs, and frequent challenges with transfer-ability or reproducibility.
-A review of all these issues is beyond our scope but can be found elsewhere [@rl_intro; @millie-1; @millie-2].
+RL methods also bring many challenges of their own: being notoriously difficult to train and evaluate, requiring immense computational costs, and presenting frequent challenges with transfer-ability or reproducibility.
+A review of all these issues is beyond our scope but can be found elsewhere [@rl-intro; @millie-1; @millie-2].
 Here, though, we will focus on the issue of opacity and interpretability raised by these methods.
@@
 Bang-bang solutions are known to be optimal for a variety of single-species models, however, there is no a-priori reason why they should be optimal for higher-dimensional systems.
+This way, the DRL agent discovers and exploits, as it were, an effective simplification of the control problem.
+We provide some intuition as to how this simplification arises from the dynamics of the system, but leave a more rigorous characterization of this phenomenon for future work.
+
 The DRL policy drives an oscillation process in the system and, in this way, maintains the long-term sustainability of fishing.
@@
   \begin{split}
-    \Delta X_t &:= X_{t+1} - X_t = f_X(N_t) - M_X X_t\\
-    \Delta Y_t &:= X_{t+1} - Y_t = f_Y(N_t) - M_Y Y_t\\
+    \Delta X_t &= f_X(N_t) - M_X X_t\\
+    \Delta Y_t &= f_Y(N_t) - M_Y Y_t\\
     &\dots
@@
 \end{align}
-where $N_t = (X_t,\ Y_t,\ \dots) \in \mathbb{R}^{n}_+$ is a vector of populations and $f_i:\mathbb{R}^{n}\to\mathbb{R}$ are arbitrary functions.
+where $N_t = (X_t,\ Y_t,\ \dots) \in \mathbb{R}^{n}_+$ is a vector of populations, $\Delta X_t := X_{t+1} - X_t$, $\Delta Y_t := Y_{t+1} - Y_t$, and where $f_i:\mathbb{R}^{n}\to\mathbb{R}$ are arbitrary functions.
 Here, $M_i=M_i(N_t)\in\mathbb{R}_+$ is a state-dependent fish mortality arising from harvesting (sometimes referred to as *fishing effort*).
@@
 That is, the agent takes actions within the environment.
-As a result of an agent's action, the environment will transition to a new state and return a numerical *reward* to the agent.\footnote{See Fig.
-TBD in [@rl-intro-marcus] for a conceptual description of reinforcement learning algorithms.}
+As a result of an agent's action, the environment will transition to a new state and return a numerical *reward* to the agent.
+(See Fig.
+1 in [@rl-intro] for a conceptual description of reinforcement learning algorithms.)
 The rewards encode the agent's goal.
@@
 #| (B) Several constant mortality policies shown for illustrative purposes.
-Slopes $0.02$, $0.03$, $0.055$ and $0.065$ shown in black; the optimal linear policy $M^*X$ shown in green, while $0.8M^*X$ shown in violet.
-(TBD choose nicer and plentier  slopes)"
+Slopes $0.02$, $0.03$, $0.055$ and $0.065$ shown in black; the optimal linear policy $M^*X$ shown in green, while $0.8M^*X$ shown in violet."
 #| out.width : "6.5in"
@@
 
-In Fig. \ref{fig:state-space}, we perform a back to back visualization of all the policies evaluated up to now---MSY constant mortality, 80% MSY constant mortality, DRL and escapement.
+In Fig. \ref{fig:state-space}, we perform a back to back visualization of all the policies evaluated---MSY constant mortality, 80% MSY constant mortality, DRL and optimal constant escapement.
 The policies are computed at each state along the paths they traverse in the 50 evaluation rounds, and the values are projected on the $X$-$Z$ plane for visualization.

### 2023-01-31 code/plots.Rmd
added escapement appendix

@@
 We compare policies across constant effort (MSY-like policies), constant escapement, and the RL agent (PPO).
+
+<!-- Basic setup -->
+```{r setup}
+library(tidyverse)
+library(patchwork)
+library(RColorBrewer)
+```
+
+<!-- Data sources -->
+```{r}
+# for best const mortality and best const escapement
+msy_optimizer <- function(df) {
+  df |> 
+  group_by(action, rep) |>
+  filter(t == max(t)) |> 
+  group_by(action) |>
+  summarise(mean_reward = mean(reward), sd = sd(reward)) |> 
+  filter(mean_reward == max(mean_reward))
+}
+escapement_optimizer <- function(df) {
+  df |> 
+  group_by(escapement, rep) |>
+  filter(t == max(t)) |> 
+  group_by(escapement) |>
+  summarise(mean_reward = mean(reward), sd = sd(reward)) |> 
+  filter(mean_reward == max(mean_reward))
+}
+
+# Collect data (outsource to separate R script later TBD)
+msy_df <- read_csv("../data/msy.csv.xz")
+best_action <- msy_optimizer(msy_df)$action
+escapement_df <- read_csv("../data/escapement.csv.xz", show_col_types = FALSE)
+best_e <- escapement_optimizer(escapement_df)$escapement
+ppo_df <- read_csv("../data/PPO200.csv.xz")
+
+opt_escapement <- escapement_df |> filter(escapement == best_e)
+
+# set up for plots:
+
+msy_sim <- msy_df |>
+    filter(action == best_action) |> 
+    group_by(rep) |>
+    pivot_longer(c("X", "Y", "Z"),
+                names_to = "species", values_to = "abundance") |>
+    mutate(abundance = abundance + 1, policy = "msy") # natural units, policy name
+
+actions <- unique(msy_df$action)
+i <- which.min(abs(actions - best_action* 0.8))
+tac_sim <- msy_df |> 
+    filter(action == actions[[i]]) |>
+    pivot_longer(c("X", "Y", "Z"),
+        names_to = "species", values_to = "abundance") |>
+    mutate(abundance = abundance + 1, policy = "tac")
+
+ppo_sim <- ppo_df |> 
+    pivot_longer(c("X", "Y", "Z"),
+        names_to = "species", values_to = "abundance") |>
+    mutate(abundance = abundance + 1, policy = "ppo")
+
+esc_sim_df <- escapement_df |> 
+    pivot_longer(c("X", "Y", "Z"),
+        names_to = "species", values_to = "abundance") |>
+    mutate(abundance = abundance + 1)
+
+msy_tac_ppo <- bind_rows(msy_sim, tac_sim, ppo_sim)
+```
+
+```{r}
+#| label: tuning
+#| fig.cap : "Result of the tuning procedure for the constant mortality strategy. 
+#| (A) Mean episode reward (averaged over 50 episodes) shown as scatter data, and shaded in gray are two standard deviations around each scatter point.
+Vertical lines correspond to the optimal constant mortality rate $M^* = $ 0.045 (green) and $0.8 M^*=$ 0.036 (violet).
+#| (B) Several constant mortality policies shown for illustrative purposes.
+Slopes $0.02$, $0.03$, $0.055$ and $0.065$ shown in black; the optimal linear policy $M^*X$ shown in green, while $0.8M^*X$ shown in violet."
+#| out.width : "6.5in"
+msy_rewards <- msy_df |>
+  group_by(action, rep) |>
+  filter(t == max(t)) |> 
+  group_by(action) |>
+  summarise(mean_reward = mean(reward),
+            sd = sd(reward))
+
+rmsy <- msy_rewards |>
+  ggplot(aes(action, mean_reward, 
+             ymin = mean_reward - 2*sd,
+             ymax=mean_reward+2*sd)) +
+  geom_point() + geom_ribbon(alpha=0.4) +
+  labs(y = "episode reward", x = "constant mortality") +
+  theme(aspect.ratio = 1/1) + 
+  geom_vline(aes(xintercept=best_action), color="seagreen3", size=1) + 
+  geom_vline(aes(xintercept=0.8*best_action), color="violetred3", size=1) +
+  theme_bw() + 
+  theme(panel.grid.major = element_blank(), 
+        panel.grid.minor = element_blank(),
+        aspect.ratio = 1)
+
+# linear function blueprint
+linear_f <- function(slope, x) slope*x
+# generates linear function with given slope
+linear_f_gen <- function(slope) (function(x) slope*x)
+
+# dummy data set to keep ggplot happy
+p <- ggplot(data = data.frame(x = 0), mapping = aes(x = x))
+
+# p + stat_function(fun = linear_f_gen(0.5)) + xlim(0,2)
+lines <- p + stat_function(fun = linear_f_gen(0.009)) + 
+  stat_function(fun = linear_f_gen(0.018)) +
+  stat_function(fun = linear_f_gen(0.027)) +
+  stat_function(fun = linear_f_gen(0.8*best_action), col = "violetred3", size=1) +
+  stat_function(fun = linear_f_gen(best_action), col = "seagreen3", size=1) +
+  stat_function(fun = linear_f_gen(0.054)) + 
+  stat_function(fun = linear_f_gen(0.063)) +
+  stat_function(fun = linear_f_gen(0.072)) +
+  stat_function(fun = linear_f_gen(0.081)) +
+  stat_function(fun = linear_f_gen(0.090)) +
+  xlim(0,2) +
+  xlab("X") + ylab("harvest") +
+  theme_bw() +
+  theme(#panel.grid.major = element_blank(), 
+        #panel.grid.minor = element_blank(), 
+        aspect.ratio = 1)
+
+(rmsy | lines) + plot_annotation(tag_levels = c('A'))
+```
+
+<!-- Resuts figures -->
+
+```{r} 
+#| label: timeseries 
+#| fig.cap: "A sample time series for the three fish populations, with the system being controlled, respectively, by the MSY optimal mortality, by $80$ percent of the MSY optimal mortality, and by DRL."
+#| out.width: "6.5in"
+policy_labels <- list("msy" = "MSY Const. Mortality", "tac"= "80% MSY Const. Mortality", "ppo"= "DRL")
+policy_labeller <- function(variable,value){
+  return(policy_labels[value])
+}
+msy_tac_ppo |> 
+    group_by(rep, policy) |>
+    filter(rep == 6) |>
+    ggplot(aes(t, abundance, col = species)) + 
+        geom_line()  + 
+        scale_color_manual(values=c('palegreen3', 'coral1', 'deepskyblue4')) +
+        facet_wrap(~factor(policy, levels = c("msy", "tac", "ppo")), labeller = policy_labeller ) +
+        labs(y = "abundance", x = "time") +
+        theme_bw() +
+        theme(aspect.ratio = 2/3)
+```
+
+```{r} 
+#| label: rew_len_fig
+#| fig.cap: "Bar charts for the episode lengths (top panel) and episode total harvests (bottom panel) for 50 episodes." 
+#| out.width: "6.5in"
+p1 <- msy_tac_ppo |> 
+    group_by(rep, policy) |>
+    filter(t==max(t), species == "X", rep < 50) |>
+    ggplot(aes(rep, t)) + 
+      geom_col() +
+      facet_wrap(~factor(policy, levels = c("msy", "tac", "ppo")), labeller = policy_labeller ) +
+      labs(y = "episode length", x = "repetition") +
+      theme_bw() + 
+      theme(axis.text.x = element_blank())
+
+p2 <- msy_tac_ppo |> 
+    group_by(rep, policy) |>
+    filter(t==max(t), species == "X", rep < 50) |>
+    ggplot(aes(rep, reward)) + 
+      geom_col() +
+      facet_wrap(~factor(policy, levels = c("msy", "tac", "ppo")), labeller = policy_labeller ) +
+      labs(y = "episode harvest", x = "repetition") +
+      theme_bw() + 
+      theme(axis.text.x = element_blank())
+
+(p1 / p2) + plot_annotation(tag_levels = 'A')
+```
+
+```{r}
+#| label: action-X
+#| fig.cap: "(A-B) Harvest and escapement scatter data for the DRL policy together with the optimal constant escapement (pink) and constant mortality (green) policies for comparison. 
+#| (A) includes $Y$ dependence through color, whereas (B) includes $Z$ dependence similarly.
+#| (B-C) escapement (i.e. $X_t - h_t$) scatter data for the DRL policy. 
+#| Optimal constant escapement policy shown in pink for comparison."
+#| out.width: "6.5in"
+opt_escapement_harvest <- function(X) case_when(
+  X > best_e ~ X - best_e,
+  X <= best_e ~ 0
+)
+
+opt_escapement_escapement <- function(X) case_when(
+  X <= best_e ~ X,
+  X > best_e ~ best_e
+)
+
+msy_harvest <- function(X) best_action*X
+tac_harvest <- function(X) 0.8*msy_harvest(X)
+
+py <- ppo_df |>
+  mutate(X = X+1, Y = Y+1, Z = Z+1, harvest = action*X) |>
+  ggplot(aes(X, harvest, color = Y)) +
+  geom_point() +
+  theme_bw() + 
+  theme(aspect.ratio = 1/1) +
+  stat_function(fun = opt_escapement_harvest, col = 'deeppink', size = 0.7) + 
+  stat_function(fun = msy_harvest, col = 'seagreen2', size = 0.7)
+
+pz <- ppo_df |>
+  mutate(X = X+1, Y = Y+1, Z = Z+1, harvest = action*X) |>
+  ggplot(aes(X, harvest, color = Z)) +
+  geom_point() +
+  theme_bw() + 
+  theme(aspect.ratio = 1/1) +
+  stat_function(fun = opt_escapement_harvest, col = 'deeppink', size = 0.7) + 
+  stat_function(fun = msy_harvest, col = 'seagreen2', size = 0.7)
+
+escxy <- ppo_df |>
+  mutate(X = X+1, Y = Y+1, Z = Z+1, escapement = (1-action)*X) |>
+  ggplot(aes(X, escapement, color = Y)) +
+  geom_point() +
+  theme_bw() + 
+  theme(aspect.ratio = 1/1) +
+  stat_function(fun = opt_escapement_escapement, col = 'deeppink', size = 0.7)
+
+escxz <- ppo_df |>
+  mutate(X = X+1, Y = Y+1, Z = Z+1, escapement = (1-action)*X) |>
+  ggplot(aes(X, escapement, color = Z)) +
+  geom_point() +
+  theme_bw() + 
+  theme(aspect.ratio = 1/1) +
+  stat_function(fun = opt_escapement_escapement, col = 'deeppink', size = 0.7)
+
+((py | pz)/ (escxy | escxz)) + plot_annotation(tag_levels = c('A'))
+```
+```{r}
+#| label: state-space
+#| fig.cap: "Policy heatmaps.
+Each policy, say, $\\pi$, was evaluated at points $(x,y,z)$ visited during any of the 50 evaluation rounds.
+We projected this data down to the $X$-$Y$ plane---i.e.
+we plotted the heatmap $\\pi(x,y,z)$ as a function of $x$ and $y$."
+#| out.width: "6.5in"
+i_tac <- which.min(abs(actions - best_action * 0.8))
+
+# color pallette limits
+myPalette <- colorRampPalette(rev(brewer.pal(11, "Spectral")))
+sc <- scale_colour_gradientn(colours = myPalette(100), limits=c(0,0.25))
+
+pmsy <- msy_df |>
+  mutate(X = X+1, Y = Y+1, Z = Z+1, harvest = action * X) |>
+  filter(action == best_action) |>
+  ggplot(aes(X, Y, color = harvest)) +
+  geom_point() + sc +
+  xlim(0,1) +
+  ylim(0,0.85) +
+  theme_bw() + 
+  ggtitle("Opt. Const. Mortality")
+
+ptac <- msy_df |>
+  mutate(X = X+1, Y = Y+1, Z = Z+1, harvest = action * X) |>
+  filter(action == actions[[i_tac]]) |>
+  ggplot(aes(X, Y, color = harvest)) +
+  geom_point() + sc +
+  xlim(0,1) +
+  ylim(0,0.85) +
+  theme_bw() + 
+  ggtitle("80% Opt. Const. Mortality")
+
+pppo <- ppo_df |>
+  mutate(X = X+1, Y = Y+1, Z = Z+1, harvest = action * X) |>
+  ggplot(aes(X, Y, color = harvest)) +
+  geom_point() + sc +
+  xlim(0,1) +
+  ylim(0,0.85) +
+  theme_bw() + 
+  ggtitle("DRL")
+
+pesc <- opt_escapement |>
+  mutate(X = X+1, Y = Y+1, Z = Z+1, 
+        harvest = case_when(
+          X - escapement > 0 ~ X - escapement,
+          X - escapement <= 0 ~  0)
+         ) |> 
+  ggplot(aes(X, Y, color = harvest)) +
+  geom_point() + sc +
+  xlim(0,1) +
+  ylim(0,0.85) +
+  theme_bw() + 
+  ggtitle("Opt. Const. Escapement")
+
+(
+  (pmsy | ptac) / (pppo | pesc) +
+  plot_layout(guides = "collect") & theme(legend.position = 'right')
+) + 
+  plot_annotation(tag_levels = c('A'))
+```
+
+<!-- Appendices -->
+
+```{r} 
+#| label: timeseries_msy
+#| fig.cap: "Nine time series of the system controlled by the MSY constant mortality policy." 
+#| out.width: "6.5in"
+msy_tac_ppo |> 
+    group_by(rep) |>
+    filter(rep < 9, policy == "msy") |>
+    ggplot(aes(t, abundance, col = species)) + 
+        geom_line()  + 
+        scale_color_manual(values=c('palegreen3', 'coral1', 'deepskyblue4')) +
+        facet_wrap(~factor(rep)) +
+        labs(y = "abundance", x = "time") +
+        theme_bw() + 
+        theme(aspect.ratio = 0.5)
+```
+
+```{r} 
+#| label: timeseries_tac
+#| fig.cap: "Nine time series of the system controlled by the policy using 80 percent of the MSY constant mortality." 
+#| out.width: "6.5in"
+msy_tac_ppo |> 
+    group_by(rep) |>
+    filter(rep < 9, policy == "tac") |>
+    ggplot(aes(t, abundance, col = species)) + 
+        geom_line()  + 
+        scale_color_manual(values=c('palegreen3', 'coral1', 'deepskyblue4')) +
+        facet_wrap(~factor(rep)) +
+        labs(y = "abundance", x = "time") +
+        theme_bw() + 
+        theme(aspect.ratio = 0.5)
+```
+
+```{r} 
+#| label: timeseries_ppo
+#| fig.cap: "Nine time series of the system controlled by the trained DRL agent." 
+#| out.width: "6.5in"
+msy_tac_ppo |> 
+    group_by(rep) |>
+    filter(rep < 9, policy == "ppo") |>
+    ggplot(aes(t, abundance, col = species)) + 
+        geom_line()  + 
+        scale_color_manual(values=c('palegreen3', 'coral1', 'deepskyblue4')) +
+        facet_wrap(~factor(rep)) +
+        labs(y = "abundance", x = "time") +
+        theme_bw() + 
+        theme(aspect.ratio = 0.5)
+```
+
+<!-- Old code blocks: -->
 

### 2023-01-31 manuscript/manuscript.Rmd
added escapement appendix

@@
 
-```{r}
-msy_rewards <- msy_df |>
-  group_by(action, rep) |>
-  filter(t == max(t)) |> 
-  group_by(action) |>
-  summarise(mean_reward = mean(reward),
-            sd = sd(reward))
-
-esc_rewards <- escapement_df |>
-  group_by(escapement, rep) |>
-  filter(t == max(t)) |> 
-  group_by(escapement) |>
-  summarise(mean_reward = mean(reward),
-            sd = sd(reward))
-
-rmsy <- msy_rewards |>
-  ggplot(aes(action, mean_reward, 
-             ymin = mean_reward - 2*sd,
-             ymax=mean_reward+2*sd)) +
-  geom_point() + geom_ribbon(alpha=0.4) +
-  labs(y = "episode reward", x = "constant mortality") +
-  theme(aspect.ratio = 1/1) + 
-  geom_vline(aes(xintercept=0.0360), color="red") + 
-  geom_vline(aes(xintercept=0.0288), color="green")
-
-resc <- esc_rewards |>
-  ggplot(aes(esc_level, mean_reward, 
-             ymin = mean_reward - 2*sd,
-             ymax=mean_reward+2*sd)) +
-  geom_point() + geom_ribbon(alpha=0.4) +
-  labs(y = "episode reward", x = "constant escapement") +
-  theme(aspect.ratio = 1/1)
-
-# (rmsy | resc)
-```
-
 # Reinforcement learning {#sec:RL}
@@
 This hypothesis could be explored in more generality as it might apply to a wide variety of models.
+
+## Acknowledgements
+
+The title of this piece references a mathematical biology workshop at NIMBioS
+organized by Paul Armsworth, Alan Hastings, Megan Donahue, and Carl Towes in
+2011 which first sought to emphasize 'pretty darn good' control solutions to
+more realistic problems over optimal control to idealized ones. 
+This material is based upon work supported by the National Science Foundation under Grant No. DBI-1942280.
 
@@
 
-## Acknowledgements
-
-The title of this piece references a mathematical biology workshop at NIMBioS
-organized by Paul Armsworth, Alan Hastings, Megan Donahue, and Carl Towes in
-2011 which first sought to emphasize 'pretty darn good' control solutions to
-more realistic problems over optimal control to idealized ones. 
-This material is based upon work supported by the National Science Foundation under Grant No. DBI-1942280.
+## Optimal constant escapement policy evaluation
+
+Here we reproduce the plots from the main text for the optimal constant escapement policy to illustrate the similarity to the DRL-derived policy.
+
+Fig. \ref{fig:esc-tuning-eval} illustrates the tuning process of finding the optimal constant escapment policy (left panel, compare Fig. \ref{fig:tuning}).
+The same figure shows, in the middle panel, the reward histogram for the optimal constant escapement policy (compare Fig. \ref{fig:rew_len_fig}).
+Finally, the right panel displays the same histogram for the DRL policy (reproduced from Fig. \ref{fig:rew_len_fig}).
+We may see that the two policies lead to essentially the same reward distribution.
+
+```{r}
+#| label: esc-tuning-eval
+#| fig.cap: "(Left panel) Tuning results for constant escapement policies: mean episode reward scatter data (averaged over 50 episodes) together with two standard deviations shaded in grey. 
+#| The optimal escapement level (the one producing maximum mean harvest) is $X_{\text{esc.}}=0.62$.
+#| (Middle and right panels) Reward histograms for the optimal constant escapement policy (middle) and for the DRL policy (right).
+#| For both DRL and optimal constant escapement, no early episode ends are observed (i.e. all episodes last for 200 time steps)."
+#| out.width: "6.5in"
+#| fig.align: "center"
+esc_rewards <- escapement_df |>
+  group_by(escapement, rep) |>
+  filter(t == max(t)) |> 
+  group_by(escapement) |>
+  summarise(mean_reward = mean(reward),
+            sd = sd(reward))
+
+resc <- esc_rewards |>
+  ggplot(aes(escapement, mean_reward, 
+             ymin = mean_reward - 2*sd,
+             ymax=mean_reward+2*sd)) +
+  geom_point() + geom_ribbon(alpha=0.4) +
+  labs(y = "episode reward", x = "constant escapement") +
+  theme_bw() +
+  theme(aspect.ratio = 1/1)
+
+esc_len <- esc_sim_df |>
+  group_by(rep) |>
+  filter(t==max(t), species == "X", escapement == best_e, rep < 50) |>
+  ggplot(aes(rep, t)) + 
+      geom_col() +
+      labs(y = "episode length", x = "repetition") +
+      theme_bw() + 
+      theme(axis.text.x = element_blank())
+
+esc_rew <- esc_sim_df |>
+  group_by(rep) |>
+  filter(t==max(t), species == "X", escapement == best_e, rep < 50) |>
+  ggplot(aes(rep, reward)) + 
+      geom_col() +
+      labs(y = "episode harvest", x = "repetition", title = "Escapement") +
+      ylim(0,12.8) +
+      #ggtitle("Escapement") + 
+      theme_bw() + 
+      theme(axis.text.x = element_blank(), plot.title = element_text(hjust = 0.5))
+
+drl_rew <- msy_tac_ppo |> 
+    group_by(rep, policy) |>
+    filter(t==max(t), species == "X", policy == "ppo", rep < 50) |>
+    ggplot(aes(rep, reward)) + 
+      geom_col() +
+      labs(y = "episode harvest", x = "repetition", title = "DRL") +
+      ylim(0,12.8) +
+      #ggtitle("DRL") + 
+      theme_bw() + 
+      theme(axis.text.x = element_blank(), plot.title = element_text(hjust = 0.5))
+
+(resc | esc_rew | drl_rew)
+```
+
+Fig. \ref{fig:timeseries_esc} displays 9 time-series for episodes controlled with the optimal constant escapement policy.
+Again, this behaviour is essentially indistinguishable from the behavior of the system controlled by the DRL policy (see Fig. \ref{fig:timeseries_ppo}).
+
+This series of figures gives evidence that, in fact, the DRL agent we have trained has obtained a policy which, effectively, acts as an optimal constant escapement policy.
+Although in Fig.
+\ref{fig:action-X} we found some deviations from the constant escapement case (e.g.
+DRL was more conservative when fishing high populations), these do not seem to lead to a different dynamics.
+
+```{r} 
+#| label: timeseries_esc
+#| fig.cap: "Nine time series of the system controlled by the optimal constant escapement policy." 
+#| out.width: "6.5in"
+esc_sim_df |> 
+    group_by(rep) |>
+    filter(rep < 9, escapement == best_e) |>
+    ggplot(aes(t, abundance, col = species)) + 
+        geom_line()  + 
+        scale_color_manual(values=c('palegreen3', 'coral1', 'deepskyblue4')) +
+        facet_wrap(~factor(rep)) +
+        labs(y = "abundance", x = "time") +
+        theme_bw() + 
+        theme(aspect.ratio = 0.5)
+```
+# References

### 2023-01-31 manuscript/manuscript.Rmd
mini

@@
 
-We have shown that the DRL agent finds an approximately *bang-bang* type of solution (see Textbox 1 for the definition of bang-bang solutions and Figs.
+We have shown that the DRL agent finds an approximately *bang-bang* type of solution (see Text Box 1 for the definition of bang-bang solutions and Figs.
 \ref{fig:action-X} and \ref{fig:state-space} for the result).

### 2023-01-31 manuscript/manuscript.Rmd
mini
readded textbox

@@
 We thus observe that the DRL solution leads to a higher profitability *and* stability as compared to constant mortality policies.
+
+```{r}
+#| label: textbox
+#| out.width: "1\\linewidth"
+knitr::include_graphics("figures/textbox.png")
+```
 
@@
 #| fig.cap: "(Left panel) Tuning results for constant escapement policies: mean episode reward scatter data (averaged over 50 episodes) together with two standard deviations shaded in grey. 
-#| The optimal escapement level (the one producing maximum mean harvest) is $X_{\text{esc.}}=0.62$.
+#| The optimal escapement level (the one producing maximum mean harvest) is $X_{\\text{esc.}}=0.62$.
 #| (Middle and right panels) Reward histograms for the optimal constant escapement policy (middle) and for the DRL policy (right).

### 2023-01-31 manuscript/manuscript.Rmd
textbox and conceptual fig positioning
revised abstract

@@
   complexity commonly encountered in real-world systems, including
-  dimensionality, process error, model bias and heterogeneity.  Instead
-  of tackling these complexities directly, researchers have typically
-  sought to find exact optimal solutions to simplified models of the
-  processes in question. When is the optimal solution to a very
-  approximate, stylized model better than an approximate solution to a
-  more accurate model? This question has largely gone unanswered owing
-  to the difficulty of finding even approximate solutions in the case of
-  complex models.  Our approach draws on recent algorithmic and
-  computational advances in deep reinforcement learning. These methods
-  have hitherto focused on problems in games or robotic mechanics, which
-  operate under precisely known rules. We demonstrate the ability for
-  novel algorithms using deep neural networks to successfully
-  approximate such solutions (the ‘‘policy function’’ or control rule)
-  without knowing or ever attempting to infer a model for the process
-  itself. This powerful new technique lets us finally begin to answer
-  the question. We showcase this approach by applying it to the control
-  of a highly non-linear three-variable model for a fishery. Here, the
-  reinforcement learning agent discovers an effective simplification of
-  the problem to obtain a simple, interpretable control rule. Moreover,
-  the policy obtained is both more profitable and more sustainable than
-  any constant mortality policy---the standard family of policies considered
-  in fishery management.
+  dimensionality, process error, model bias and data heterogeneity.
+  Instead of tackling these system complexities directly, researchers 
+  have typically sought to simplify models to fit optimal control 
+  methods. But when is the optimal solution to an approximate, stylized 
+  model better than an approximate solution to a more accurate model? 
+  While this question has largely gone unanswered owing to the 
+  difficulty of finding even approximate solutions for complex models, 
+  recent algorithmic and computational advances in deep reinforcement 
+  learning (DRL) might finally allow us to address these questions. DRL 
+  methods have to date been applied primarily in the context of games 
+  or robotic mechanics, which operate under precisely known rules. Here, 
+  we demonstrate the ability for DRL algorithms using deep neural networks 
+  to successfully approximate solutions (the ‘‘policy function’’ or control 
+  rule) in a non-linear three-variable model for a fishery without knowing 
+  or ever attempting to infer a model for the process itself. We find that 
+  the reinforcement learning agent discovers an effective simplification of
+  the problem to obtain an interpretable control rule. We show that the 
+  policy obtained with DRL is both more profitable and more sustainable 
+  than any constant mortality policy---the standard family of policies 
+  considered in fishery management.
 
@@
 
+```{r}
+#| label: textbox
+#| out.width: "1\\linewidth"
+knitr::include_graphics("figures/textbox.png")
+```
+
 Recent advances in machine learning may allow us to once again bridge these approaches, while also bringing new challenges of their own.
@@
 In general, one can only assess the performance of these black box methods relative to alternatives.
-
-<!--NEED some kind of map paragraph here saying what's ahead, what we show  -->
-In this article, we compare the performance of a DRL solution to that of a management strategy evaluation across potential 'constant mortality' policies.\footnote{A repository with all the relevant code to reproduce our results may be found at \url{https://github.com/boettiger-lab/approx-model-or-approx-soln} in the ‘‘code’’ subdirectory.
-The data used is found in the ‘‘data’’ subdirectory, but the user may use the code provided to generate new data sets.}
-We consider the problem of devising harvest strategies in a three species ecosystem (Figure 1B).
-We show that, somewhat surprisingly, the DRL policy closely resembles a *bang-bang* solution (see Text Box 1).
-Bang-bang solutions are known to be optimal for a variety of single-species models, however, there is no a-priori reason why they should be optimal for higher-dimensional systems.
-This way, the DRL agent discovers and exploits, as it were, an effective simplification of the control problem.
-We provide some intuition as to how this simplification arises from the dynamics of the system, but leave a more rigorous characterization of this phenomenon for future work.
-
-The DRL policy drives an oscillation process in the system and, in this way, maintains the long-term sustainability of fishing.
-In contrast, constant mortality policies overfish at low populations and underfish (relative to the DRL solution) at high populations.
-This leads to a tradeoff when leveraging constant mortality strategies: higher constant mortality rates lead to high profitability at the expense of a higher probability of overfishing to extinction, while lower constant mortality rates attain stability at the expense of profit.
-
-The DRL approach side-steps this tradeoff by optimizing over a more complex family of possible policies—policies parametrized by a neural network, as opposed to policies labeled by a single parameter (the mortality rate).
-We thus observe that the DRL solution leads to a higher profitability *and* stability as compared to constant mortality policies.
-
-```{r}
-#| label: textbox
-#| out.width: "1\\linewidth"
-knitr::include_graphics("figures/textbox.png")
-```
 
@@
 ```
+
+<!--NEED some kind of map paragraph here saying what's ahead, what we show  -->
+In this article, we compare the performance of a DRL solution to that of a management strategy evaluation across potential 'constant mortality' policies.\footnote{A repository with all the relevant code to reproduce our results may be found at \url{https://github.com/boettiger-lab/approx-model-or-approx-soln} in the ‘‘code’’ subdirectory.
+The data used is found in the ‘‘data’’ subdirectory, but the user may use the code provided to generate new data sets.}
+We consider the problem of devising harvest strategies in a three species ecosystem (Figure 1B).
+We show that, somewhat surprisingly, the DRL policy closely resembles a *bang-bang* solution (see Text Box 1).
+Bang-bang solutions are known to be optimal for a variety of single-species models, however, there is no a-priori reason why they should be optimal for higher-dimensional systems.
+This way, the DRL agent discovers and exploits, as it were, an effective simplification of the control problem.
+We provide some intuition as to how this simplification arises from the dynamics of the system, but leave a more rigorous characterization of this phenomenon for future work.
+
+The DRL policy drives an oscillation process in the system and, in this way, maintains the long-term sustainability of fishing.
+In contrast, constant mortality policies overfish at low populations and underfish (relative to the DRL solution) at high populations.
+This leads to a tradeoff when leveraging constant mortality strategies: higher constant mortality rates lead to high profitability at the expense of a higher probability of overfishing to extinction, while lower constant mortality rates attain stability at the expense of profit.
+
+The DRL approach side-steps this tradeoff by optimizing over a more complex family of possible policies—policies parametrized by a neural network, as opposed to policies labeled by a single parameter (the mortality rate).
+We thus observe that the DRL solution leads to a higher profitability *and* stability as compared to constant mortality policies.
 

### 2023-01-31 manuscript/manuscript.Rmd
added hadcm3 gcm references

@@
 While simplified models have historically allowed for exploration of large decision spaces, the new wave of rich models are applied to highly oversimplified descriptions of potential actions they seek to inform.
-For instance, Global Circulation Models (GCMs) such as HadOM3 [CITE] model earth's climate using 1.5M variables, while the comparably vast potential action space is modeled as one of six emissions scenarios from the IPCC [@riahi2017shared].
+For instance, Global Circulation Models (GCMs) such as HadCM3 [@pope2000impact; @gordon2000simulation; @collins2001internal] model earth's climate using 1.5M variables, while the comparably vast potential action space is modeled as one of six emissions scenarios from the IPCC [@riahi2017shared].
 

### 2023-01-31 code/plots.Rmd
added all plots to plots.Rmd

@@
 
+<!-- time series -->
 ```{r} 
@@
         theme(aspect.ratio = 0.5)
+```
+
+```{r} 
+#| label: timeseries_esc
+#| fig.cap: "Nine time series of the system controlled by the optimal constant escapement policy." 
+#| out.width: "6.5in"
+esc_sim_df |> 
+    group_by(rep) |>
+    filter(rep < 9, escapement == best_e) |>
+    ggplot(aes(t, abundance, col = species)) + 
+        geom_line()  + 
+        scale_color_manual(values=c('palegreen3', 'coral1', 'deepskyblue4')) +
+        facet_wrap(~factor(rep)) +
+        labs(y = "abundance", x = "time") +
+        theme_bw() + 
+        theme(aspect.ratio = 0.5)
+```
+
+<!-- msy scale downs -->
+
+```{r}
+#| label: several-tac
+#| fig.cap: "Performance of constant mortality policies with mortality $\\alpha M^*$ for several values of $\\alpha$.
+#| The top panel displays the episode lengths for each repetition, while the bottom pannel displays the episode harvest.
+#| The labels in the plots are the value of $\\alpha$ expressed as a percentage."
+#| out.width: "6.5in"
+N <- 30
+actions <- unique(msy_df$action)
+i85 <- which.min(abs(actions - best_action* 0.85))
+i90 <- which.min(abs(actions - best_action* 0.90))
+i95 <- which.min(abs(actions - best_action* 0.95))
+tac_85_sim <- msy_df |> 
+    filter(action == actions[[i85]]) |>
+    pivot_longer(c("X", "Y", "Z"),
+        names_to = "species", values_to = "abundance") |>
+    mutate(abundance = abundance + 1, policy = "85%")
+tac_90_sim <- msy_df |> 
+    filter(action == actions[[i90]]) |>
+    pivot_longer(c("X", "Y", "Z"),
+        names_to = "species", values_to = "abundance") |>
+    mutate(abundance = abundance + 1, policy = "90%")
+tac_95_sim <- msy_df |> 
+    filter(action == actions[[i95]]) |>
+    pivot_longer(c("X", "Y", "Z"),
+        names_to = "species", values_to = "abundance") |>
+    mutate(abundance = abundance + 1, policy = "95%")
+
+tac_80_sim <- tac_sim |> mutate(policy = "80%")
+
+tac4 <- bind_rows(tac_80_sim, tac_85_sim, tac_90_sim, tac_95_sim)
+
+p1 <- tac4 |> 
+    group_by(rep, policy) |>
+    filter(t==max(t), species == "X", rep < N) |>
+    ggplot(aes(rep, t)) + 
+      geom_col() +
+      facet_wrap(~factor(policy), ncol = 4) +
+      labs(y = "episode length", x = "repetition") +
+      theme_bw() + 
+      theme(axis.text.x = element_blank())
+
+p2 <- tac4 |> 
+    group_by(rep, policy) |>
+    filter(t==max(t), species == "X", rep < N) |>
+    ggplot(aes(rep, reward)) + 
+      geom_col() +
+      facet_wrap(~factor(policy), ncol=4) +
+      labs(y = "episode harvest", x = "repetition") +
+      theme_bw() + 
+      theme(axis.text.x = element_blank())
+
+(p1 / p2)
+```
+
+<!-- escapement tuning and evaluation -->
+
+```{r}
+#| label: esc-tuning-eval
+#| fig.cap: "(Left panel) Tuning results for constant escapement policies: mean episode reward scatter data (averaged over 50 episodes) together with two standard deviations shaded in grey. 
+#| The optimal escapement level (the one producing maximum mean harvest) is $X_{\\text{esc.}}=0.62$.
+#| (Middle and right panels) Reward histograms for the optimal constant escapement policy (middle) and for the DRL policy (right).
+#| For both DRL and optimal constant escapement, no early episode ends are observed (i.e. all episodes last for 200 time steps)."
+#| out.width: "6.5in"
+#| fig.align: "center"
+esc_rewards <- escapement_df |>
+  group_by(escapement, rep) |>
+  filter(t == max(t)) |> 
+  group_by(escapement) |>
+  summarise(mean_reward = mean(reward),
+            sd = sd(reward))
+
+resc <- esc_rewards |>
+  ggplot(aes(escapement, mean_reward, 
+             ymin = mean_reward - 2*sd,
+             ymax=mean_reward+2*sd)) +
+  geom_point() + geom_ribbon(alpha=0.4) +
+  labs(y = "episode reward", x = "constant escapement") +
+  theme_bw() +
+  theme(aspect.ratio = 1/1)
+
+esc_len <- esc_sim_df |>
+  group_by(rep) |>
+  filter(t==max(t), species == "X", escapement == best_e, rep < 50) |>
+  ggplot(aes(rep, t)) + 
+      geom_col() +
+      labs(y = "episode length", x = "repetition") +
+      theme_bw() + 
+      theme(axis.text.x = element_blank())
+
+esc_rew <- esc_sim_df |>
+  group_by(rep) |>
+  filter(t==max(t), species == "X", escapement == best_e, rep < 50) |>
+  ggplot(aes(rep, reward)) + 
+      geom_col() +
+      labs(y = "episode harvest", x = "repetition", title = "Escapement") +
+      ylim(0,12.8) +
+      #ggtitle("Escapement") + 
+      theme_bw() + 
+      theme(axis.text.x = element_blank(), plot.title = element_text(hjust = 0.5))
+
+drl_rew <- msy_tac_ppo |> 
+    group_by(rep, policy) |>
+    filter(t==max(t), species == "X", policy == "ppo", rep < 50) |>
+    ggplot(aes(rep, reward)) + 
+      geom_col() +
+      labs(y = "episode harvest", x = "repetition", title = "DRL") +
+      ylim(0,12.8) +
+      #ggtitle("DRL") + 
+      theme_bw() + 
+      theme(axis.text.x = element_blank(), plot.title = element_text(hjust = 0.5))
+
+(resc | esc_rew | drl_rew)
 ```

### 2023-01-31 code/plots.Rmd
erased old code blocks in plots.Rmd

@@
 ```
-
-<!-- Old code blocks: -->
-
-```{r setup}
-library(tidyverse)
-
-best_reward <- function(df) {
-  df |> 
-  group_by(action, rep) |>
-  filter(t == max(t)) |> 
-  group_by(action) |>
-  summarise(mean_reward = mean(reward), sd = sd(reward)) |> 
-  filter(mean_reward == max(mean_reward))
-}
-```
-
-# Constant Mortality
-
-```{r}
-msy <- read_csv("../data/msy.csv.xz")
-best <- best_reward(msy)
-best
-```
-
-```{r}
-# we don't realize all states at each possible action during simulation
-msy |> mutate(X = X+1) |>
- # distinct(action, X) |>
-  filter(action == best$action) |>
-  mutate(harvest = action * X) |>
-  ggplot(aes(X, harvest, col=action, group=action)) + geom_point()
-
-expand_grid(mortality = seq(0,.1, length.out = 20),
-            state = seq(0,1, length.out=100)) |> 
-  mutate(case = "grid") |> 
-  bind_rows(tibble(mortality = best$action, 
-                   state= seq(0,1, length.out=100),
-                   case = "optimal")) |>
-  mutate(harvest = mortality * state) |>
-  ggplot(aes(state, harvest, col=case, group=mortality)) +
-  geom_line(show.legend = FALSE, lwd=1.5, alpa=0.8) + 
-  scale_color_manual(values=c("grey60", "darkblue"))
-```
-
-```{r}
-rewards <- msy |>
-  group_by(action, rep) |>
-  filter(t == max(t)) |> 
-  group_by(action) |>
-  summarise(mean_reward = mean(reward),
-            sd = sd(reward))
-
-rewards |>
-  ggplot(aes(action, mean_reward, 
-             ymin = mean_reward - 2*sd,
-             ymax=mean_reward+2*sd)) +
-  geom_point() + geom_ribbon(alpha=0.4)
-```
-
-```{r}
-msy_sim <- msy |>
-    filter(action == best$action) |> 
-    group_by(rep) |>
-    pivot_longer(c("X", "Y", "Z"),
-                names_to = "species", values_to = "abundance") |>
-    mutate(abundance = abundance + 1) # natural units
-
-msy_sim
-```
-
-```{r}
-msy_sim |>
-    filter(rep < 5) |>
-    ggplot(aes(t, abundance,  col=species, group =interaction(rep,species)))+ 
-        geom_line() + facet_wrap(~rep)
-
-```
-
-```{r}
-msy_sim |> 
-  group_by(rep) |>
-  filter(t==max(t), species=="X") |>
-  ggplot(aes(rep, reward)) + geom_col()
-```
-# TAC
-
-80% of  MSY
-
-```{r}
-# determine which action is 80% of MSY
-actions <- unique(msy$action)
-i <- which.min(abs(actions - best$action* 0.8))
-```
-
-```{r}
-tac_sim <- msy |> 
-  filter(action == actions[[i]]) |>
-  pivot_longer(c("X", "Y", "Z"),
-               names_to = "species", values_to = "abundance")
-tac_sim |>
-  filter(rep < 20) |>
-  ggplot(aes(t, abundance,  col=species)) + 
-    geom_line() + facet_wrap(~rep)
-
-```
-```{r}
-tac_sim |> 
-  group_by(rep) |>
-  filter(t==max(t), species=="X") |>
-  ggplot(aes(rep, reward)) + geom_col()
-```
-
-# PPO Agent
-
-```{r}
-ppo <- read_csv("../data/PPO200.csv.xz")
-```
-```{r}
-ppo_sim <- ppo |> 
-  mutate(escapement = (X+1) - action * (X+1),
-         effort = action) |>
-  pivot_longer(c("X", "Y", "Z"),
-               names_to = "species", values_to = "abundance") |>
-  mutate(abundance = abundance + 1)
-```
-
-```{r}
-ppo_sim |> 
-  group_by(rep) |>
-  filter(t==max(t), species=="X") |>
-  ggplot(aes(rep, reward)) + geom_col()
-
-ppo_sim |> 
-  group_by(rep) |>
-  filter(t==max(t), species=="X") |> 
-  ungroup() |>
-  summarise(mean = mean(reward))
-```
-
-```{r}
-ppo_sim |>
-  filter(rep < 4) |>
-  ggplot(aes(t, abundance, col=species)) +
-  geom_line(alpha=0.6) + facet_wrap(~rep)
-```
-
-```{r}
-ppo |>
-  mutate(X = X+1, Z = Z+1, Z = round(Z, 1)) |>
-  distinct(action, X, Z) |> 
-  mutate(harvest = action * X) |>
-  ggplot(aes(X, harvest, group=Z, color=Z)) +
-  geom_smooth(se=FALSE, method=lm,  formula = y ~ splines::ns(x, 6)) +
-  geom_point(alpha=0.6, shape="+")
-
-```
-
-```{r}
-ppo |>
-  mutate(X = X+1, Z = Z+1, Z = round(Z, 1)) |>
-  distinct(action, X, Z) |> 
-  mutate(harvest = action * X,
-         escapement = X - harvest) |>
-  ggplot(aes(X, escapement, group=Z, color=Z)) +
-  #geom_smooth(se=FALSE, method=lm,  formula = y ~ splines::ns(x, 6)) +
-  geom_point(alpha=0.6, shape="+")
-
-```
-
-```{r}
-overall_ppo <- 
-  ppo |> 
-  group_by(rep) |>
-  filter(t==max(t)) |> ungroup() |>
-  filter(reward > 0) |>
-  summarise(reward = mean(reward),
-            t = mean(t),
-            effort = mean(action)) |>
-  mutate(policy="ppo")
-overall_ppo
-```
-
-# Constant Escapement
-
-```{r}
-escapement <- read_csv("../data/escapement.csv.xz", show_col_types = FALSE)
-## Determine which level of constant escapement gives highest net reward
-best_e <- best_reward(escapement)
-best_e
-```
-
-```{r}
-esc_sims <- escapement |> 
-  filter(action == best_e$action) |>
-  pivot_longer(c("X", "Y", "Z"),
-               names_to = "species", values_to = "abundance") |>
-  mutate(abundance = abundance + 1) |>
-  rename(escapement = action)
-
-esc_sims |>
-  filter(rep < 4) |>
-  ggplot(aes(t, abundance, col=species)) + geom_line() +
-  geom_line(aes(t, escapement), col="grey50") +
-  facet_wrap(~rep)
-```
-
-```{r}
-rewards <- escapement |>
-  group_by(action, rep) |>
-  filter(t == max(t)) |> 
-  group_by(action) |>
-  summarise(mean_reward = mean(reward),
-            sd = sd(reward))
-
-rewards |>
-  ggplot(aes(action, mean_reward, 
-             ymin = mean_reward - 2*sd,
-             ymax=mean_reward+2*sd)) +
-  geom_point() + geom_ribbon(alpha=0.4)
-```

### 2023-01-31 manuscript/manuscript.Rmd
eliminated a [CITE] note
added ref

@@
 Even as our research community develops simulations of the natural world that fit only in supercomputers, we analyze a space of policies that would fit on index cards.
-Similar combinations of rich process models and highly simplified decision models (often not even given the status of 'model') are common [e.g., CITE].
+Similar combinations of rich process models and highly simplified decision models (often not even given the status of 'model') are common.
 Modeling the potential action space as one of a handful of discrete scenarios is sometimes a well justified acknowledgement of the constraints faced by real-world decision-makers – particularly in the context of multilateral decisions –
@@
 RL methods also bring many challenges of their own: being notoriously difficult to train and evaluate, requiring immense computational costs, and presenting frequent challenges with transfer-ability or reproducibility.
-A review of all these issues is beyond our scope but can be found elsewhere [@rl-intro; @millie-1; @millie-2].
+A review of all these issues is beyond our scope but can be found elsewhere [@rl-intro; @millie-3].
 Here, though, we will focus on the issue of opacity and interpretability raised by these methods.

### 2023-01-31 manuscript/manuscript.Rmd
mini
parameter explanation

@@
 Thus, the term $M_X X_t$ is the total $X$ harvest at time $t$.
-We will focus on systems where a single species, $X$ is harvested, i.e. $M_Y = M_Z = \dots = 0$.
+We will focus on systems where a single species, $X$, is harvested, i.e. $M_Y = M_Z = \dots = 0$.
 
@@
 
-The model has five parameters: the growth rate $r$ and carrying capacity $K$ for $X$, a constant population $H$ of a species which preys on $X$, the maximal rate of predation $\beta$, and the predation half-maximum biomass $c$.
+The model has five parameters: the growth rate $r$ and carrying capacity $K$ for $X$, a constant population $H$ of a species which preys on $X$, the maximal predation rate $\beta$, and the predation rate half-maximum biomass $c$.
 
@@
 A parameter $c_{XY}$ mediating a Lotka-Volterra competition between $X$ and $Y$.
-A maximum hunting intensity $\beta$ and half-maximum $c$ specifying how $Z$ forages on $X$ and $Y$.
+A maximum predation rate $\beta$ and a predation rate half-maximum biomass $c$ specifying how $Z$ forages on $X$ and $Y$.
 A parameter $D$ regulating a relative preference of $Z$ to prey on $Y$.

### 2023-01-31 manuscript/manuscript.Rmd
mini changes to math models section, notation update for beginning of RL section

@@
 Eq. \eqref{eq:may} is an interesting study case of a *tipping point* (saddle-node bifurcation) (see Fig. \ref{fig:may}).
-Holding the value of $\beta$ fixed, for intermediate values of $H$ there exist two stable fixed points for the state $\hat X_t$ of the system, these two attractors separated by an unstable fixed point.
+Holding the value of $\beta$ fixed, for intermediate values of $H$ there exist two stable fixed points for the state $X_t$ of the system, these two attractors separated by an unstable fixed point.
 At a certain threshold value of $H$, however, the top stable fixed point collides with the unstable fixed point and both are annihilated.
 For this value of $H$, and for higher values, only the lower fixed point remains.
-This also creates the phenomenon of *hysteresis*, where returning $H$ to its original value is not sufficient to restore $\hat X_t$ to the original stable state.
+This also creates the phenomenon of *hysteresis*, where returning $H$ to its original value is not sufficient to restore $X_t$ to the original stable state.
 
@@
 
-Consider the following three species *generalization* of eq. \eqref{eq:may}:
+Consider the following three species generalization of eq. \eqref{eq:may}:
 
@@
 A parameter $D$ regulating a relative preference of $Z$ to prey on $Y$.
-Finally, a death rate $d_Z$ and a parameter $f$ related to the birth rate of $Z$.
+Finally, a death rate $d_Z$ and a parameter $f$ scaling the birth rate of $Z$.
 
@@
   \caption{Numerically-computed fixed points of the three-species deterministic model, eq. \eqref{eq:3d model}.
-  Here, the parameter $D$ was varied between 0.75 and 1.25, and the three components of the fixed points are plotted separately, as a function of $D$.}
+  Here, the parameter $D$ was varied between 0.75 and 1.25, and the values of the other parameters are fixed as in eq. \eqref{eq:param vals}. 
+  The three components of the fixed points are plotted separately, as a function of $D$.}
   \label{fig:fixed points}
@@
 \begin{align*}
-  (X,Y,Z) \approx &(0.839, 0.054, 0.377), \quad 
+  (X,Y,Z) \approx 
+  &(0.839, 0.054, 0.377), \quad 
     (0.048, 0.774, 0.480), \quad
     (0.359, 0.491, 0.509), \\
-    &(0.666, 0.666, 0.000),\quad
-    (0.899, 0.000, 0.303), \quad
-    (0.000, 0.818, 0.457), \\
+    &(0.899, 0.000, 0.303), \quad
+    (0.000, 0.818, 0.457), \quad
+    (0.666, 0.666, 0.000), \\
     &(1.000,0.000,0.000),\quad (0.000,1.000,0.000),\quad (0.000,0.000,0.000).
@@
 In practice, we will consider a stochastic contribution to these dynamics.
-This way, instead of eq. \eqref{eq:may}, we consider the following stochastic single-species system
+This way, instead of eq. \eqref{eq:3d model}, we consider the problem of harvesting the following stochastic model:
+<!--
 \begin{align}
@@
 
-Similarly, we use the following dynamical model in place of eq. \eqref{eq:3d model}
+Similarly, we use the following dynamical model in place of eq. \eqref{eq:3d model}-->
 \begin{align}
@@
 \end{align}
-Here, again, $\{\eta_{i,t}\}$ is a set of independent Gaussian variables centered at zero with variances, respectively, $\sigma_X^2$, $\sigma_Y^2$ and $\sigma_Z^2$.
+Here, $\{\eta_{i,t}\}$ is a set of independent Gaussian variables centered at zero with variances, respectively, $\sigma_X^2$, $\sigma_Y^2$ and $\sigma_Z^2$.
 
@@
 
-For our use case, the environment will be a computational model of the population dynamics of a fishery, with the environment state being a vector of all the fish populations, $S=(V_1, V_2, H)$.
-At each time step, the agent harvests one of the populations, $V_1$. 
-This changes the state as
+For our use case, the environment will be a computational model of the population dynamics of a fishery, with the environment state at time $t$ being a vector of all the fish populations, $S_t=(X_t,Y_t,Z_t)$.
+At each time step, the agent observes the state $S$ and harvests $X$,
 \begin{align*}
-  (V_1,\ V_2,\ H) \mapsto ((1-q)V_1,\ V_2,\ H),
+  (X_t, Y_t, Z_t) \mapsto ((1-M_X)\cdot X_t, Y_t, Z_t),
 \end{align*}
-where $q$ is a fishing *quota* set by the agent. 
-This secures a reward of $qV_1$.
-Afterwards, the environment undergoes a timestep under its natural dynamics given by \eqref{eq:3d model}.
+where the fish mortality-from-harvest $M_X=M_X(S_t)$ is the agent's action at time $t$. 
+This secures a reward of $M_X(S_t)X$.
+Afterwards, the environment undergoes a time step under its natural dynamics given by \eqref{eq:3d model stoch}.
 

### 2023-01-31 manuscript/manuscript.Rmd
mini changes methods section, s_t -> N_t for states in the RL section

@@
 
-For our use case, the environment will be a computational model of the population dynamics of a fishery, with the environment state at time $t$ being a vector of all the fish populations, $S_t=(X_t,Y_t,Z_t)$.
+For our use case, the environment will be a computational model of the population dynamics of a fishery, with the environment state at time $t$ being a vector of all the fish populations, $N_t=(X_t,Y_t,Z_t)$.
 At each time step, the agent observes the state $S$ and harvests $X$,
@@
 \end{align*}
-where the fish mortality-from-harvest $M_X=M_X(S_t)$ is the agent's action at time $t$. 
-This secures a reward of $M_X(S_t)X$.
+where the fish mortality-from-harvest $M_X=M_X(N_t)$ is the agent's action at time $t$. 
+This secures a reward of $M_X(N_t)X$.
 Afterwards, the environment undergoes a time step under its natural dynamics given by \eqref{eq:3d model stoch}.
@@
 
-- $T(s_{t+1}|s_t, a_t)$: *transition operator*, a conditional distribution which describes the dynamics of the system,
-
-- $r(s_t, a_t)$: *reward function*, the reward obtained after performing action $a_t\in\mathcal{A}$ in state $s_t$,
-
-- $d(s_0)$: *initial state distribution*, the initial state of the environment is sampled from this distribution,
+- $T(N_{t+1}|N_t, a_t)$: *transition operator*, a conditional distribution which describes the dynamics of the system (where $N_i\in\mathcal{S}$ are states of the environment),
+
+- $r(N_t, a_t)$: *reward function*, the reward obtained after performing action $a_t\in\mathcal{A}$ in state $N_t$,
+
+- $d(N_0)$: *initial state distribution*, the initial state of the environment is sampled from this distribution,
 
@@
 
-At a time $t$, the MDP agent observes the full state $s_t$ of the environment and chooses an action based on this observation according to a *policy function* $\pi(a_t | s_t)$.
-In return, it receives a discounted reward $\gamma^t r(a_t, s_t)$.
+At a time $t$, the MDP agent observes the full state $s_t$ of the environment and chooses an action based on this observation according to a *policy function* $\pi(a_t | N_t)$.
+In return, it receives a discounted reward $\gamma^t r(a_t, N_t)$.
 The discount factor helps regularize the agent, helping the optimization algorithm find solutions which pay off within a timescale of $t \sim \log(\gamma^{-1})^{-1}$.
@@
 With any fixed policy function, the agent will traverse a path 
-$\tau=(s_0,\ a_0,\ s_1,\ a_1\ \dots,\ s_{t_{\text{fin.}}})$ 
+$\tau=(N_0,\ a_0,\ N_1,\ a_1\ \dots,\ N_{t_{\text{fin.}}})$ 
 sampled randomly from the distribution
@@
   p_\pi(\tau) = 
-  d(s_0) \prod_{t=0}^{ t_{\text{fin.}}-1 }
-  \pi( a_t | s_t ) T( s_{t+1} | s_t, a_t ).
+  d(N_0) \prod_{t=0}^{ t_{\text{fin.}}-1 }
+  \pi( a_t | N_t ) T( N_{t+1} | N_t, a_t ).
 \end{align*}
@@
 $$
-  R(\tau) = \sum_{t=0}^{ t_{\text{fin.}}-1 } \gamma^t r(a_t, s_t),
+  R(\tau) = \sum_{t=0}^{ t_{\text{fin.}}-1 } \gamma^t r(a_t, N_t),
 $$
@@
 This makes it unfeasible to directly optimize $\pi$.
-In practice, an alternative approach is used: $\pi$ is optimized over a much lower-dimensional parameterized family of functions.
+In practice, an alternative approach is used: $\pi$ is optimized over a much lower-dimensional parameterized family of functions.\footnote{
+Policies are, in general, functions from state space to policy space.
+In our case, these are real-valued functions on a cube $\pi:[0,2]^{\times 3}\to \mathbb{R}_+$.
+The space of all such functions is highly singular, spanning a \emph{non-separable Hilbert space}.
+Even restricting ourselves to continuous policy functions, we end up with a set of policies which span the infinite dimensional space $L^2([0,2]^{\times3})$.
+One way to avoid optimizing over an infinite dimensional ambient space is to discretize state space into a set of bins.
+This approach runs into tractability problems: First, the dimension of policy space scales exponentially with the number of species.
+Second, even for a fixed number of species (e.g., 3), the dimension optimized over can be prohibitively large---for example if one uses 1000 bins for each population in a three-species model, the overall number of parameters being optimized over is $10^9$.
+Neural networks with much smaller number of parameters, on the other hand, can be quite expressive and sufficient to find a rather good (if not optimal) policy function.
+}
 <!-- ML: I am not following this line of thought here.
@@
 
-Since our state and observation spaces are continuous, we will focus on deep reinforcement learning throughout this paper.
-Within the RL literature there is a wealth of algorithms from which to choose from to optimize $\pi$, each with its pros and cons.
+We will focus on deep reinforcement learning throughout this paper.
+Within the DRL literature there is a wealth of algorithms from which to choose from to optimize $\pi$, each with its pros and cons.
 Most, if not all, of these are based on gradient ascent by using the technique of *back-propagation* to efficiently compute the gradient.
@@
 We trained a DRL agent parametrized by a neural network $\theta$ with two hidden, 64-neuron, layers.
-The state space used was a cube $[0,2]^{\times 3}$, while the policies explored mapped states $N_t$ to a single number, $h_t = \pi_\theta(N_t)$, the harvest quota on species $X$ at that timestep.
+The state space used was a cube $[0,2]^{\times 3}$, while the policies explored mapped states $N_t$ to a single number, $M_X(N_t) = \pi_\theta(N_t)$, the harvest mortality on species $X$ at that timestep.
 Each episode in the training and in the evaluation stages was at most 200 time steps long.
@@
 $$
-where the higher value of $X_{\text{thresh.}}$ helped select for more sustainable fishing strategies (for both the DRL and constant mortality approaches).
+where the higher value of $X_{\text{thresh.}}$ helped select for more sustainable fishing strategies.
 

### 2023-03-02 code/envs/fish_tipping.py
install-free env

@@ -1,0 +1,132 @@
+# NB: It is typical to use float32 precision to benefit from enhanced GPU speeds
+
+
+import numpy as np
+import gymnasium as gym
+from gymnasium import spaces
+
+class fish_tipping(gym.Env):
+    """A 3-species ecosystem model"""
+    def __init__(self, config=None):
+        config = config or {}
+        parameters = {
+         "r_x": np.float32(0.1),
+         "r_y": np.float32(0.1),
+         "K_x": np.float32(1.0),
+         "K_y": np.float32(1.0),
+         "beta": np.float32(0.3),
+         "v0":  np.float32(0.1),
+         "D": np.float32(1.1),
+         "tau_yx": np.float32(0),
+         "tau_xy": np.float32(0),
+         "cV": np.float32(0.5), 
+         "f": np.float32(0.5), 
+         "dH": np.float32(0.45),
+         "alpha": np.float32(0.3),
+         "sigma_x": np.float32(0.05),
+         "sigma_y": np.float32(0.1),
+         "sigma_z": np.float32(0.1)
+        }
+        initial_state = np.array([0.8396102377828771, 
+                                    0.05489978383850558,
+                                    0.3773367609828674],
+                                    dtype=np.float32) 
+        ## these parameters may be specified in config                                  
+        self.Tmax = config.get("Tmax", 200)
+        self.threshhold = config.get("threshhold", np.float32(0.08))
+        self.training = config.get("training", True)
+        self.initial_state = config.get("initial_state", initial_state)
+        self.parameters = config.get("parameters", parameters)
+        
+        self.bound = 2 * self.parameters["K_x"]
+        
+        self.action_space = spaces.Box(
+            np.array([0], dtype=np.float32),
+            np.array([1], dtype=np.float32),
+            dtype = np.float32
+        )
+        self.observation_space = spaces.Box(
+            np.array([-1, -1, -1], dtype=np.float32),
+            np.array([1, 1, 1], dtype=np.float32),
+            dtype=np.float32,
+        )        
+        self.reset(seed=config.get("seed", None))
+
+
+    def reset(self, *, seed=None, options=None):
+        self.timestep = 0
+        self.state = self.initial_state
+        return [self.state], {}
+
+    def step(self, action):
+        action = np.clip(action, [0], [1])
+        pop = self.population() # current state in natural units
+        
[…]

### 2023-03-02 code/envs/fish_tipping.py
proof-of-principle :construction:

@@ -15,7 +15,7 @@ class fish_tipping(gym.Env):
          "K_x": np.float32(1.0),
          "K_y": np.float32(1.0),
          "beta": np.float32(0.3),
-         "v0":  np.float32(0.1),
+         "v0":  np.float32(0.5),
          "D": np.float32(1.1),
          "tau_yx": np.float32(0),
          "tau_xy": np.float32(0),
@@ -23,19 +23,20 @@ class fish_tipping(gym.Env):
          "f": np.float32(0.5), 
          "dH": np.float32(0.45),
          "alpha": np.float32(0.3),
-         "sigma_x": np.float32(0.05),
-         "sigma_y": np.float32(0.1),
-         "sigma_z": np.float32(0.1)
+         "sigma_x": 0, #np.float32(0.05),
+         "sigma_y": 0, #np.float32(0.1),
+         "sigma_z": 0 #np.float32(0.1)
         }
-        initial_state = np.array([0.8396102377828771, 
-                                    0.05489978383850558,
-                                    0.3773367609828674],
-                                    dtype=np.float32) 
+        initial_pop = np.array([0.8396102377828771, 
+                                0.05489978383850558,
+                                0.3773367609828674],
+                                dtype=np.float32)
+                                
         ## these parameters may be specified in config                                  
         self.Tmax = config.get("Tmax", 200)
-        self.threshhold = config.get("threshhold", np.float32(0.08))
+        self.threshold = config.get("threshold", np.float32(1e-4))
         self.training = config.get("training", True)
-        self.initial_state = config.get("initial_state", initial_state)
+        self.initial_pop = config.get("initial_pop", initial_pop)
         self.parameters = config.get("parameters", parameters)
         
         self.bound = 2 * self.parameters["K_x"]
@@ -55,8 +56,8 @@ class fish_tipping(gym.Env):
 
     def reset(self, *, seed=None, options=None):
         self.timestep = 0
-        self.state = self.initial_state
-        return [self.state], {}
+        self.state = self.update_state(self.initial_pop)
+        return self.state, {}
 
     def step(self, action):
         action = np.clip(action, [0], [1])
@@ -72,7 +73,7 @@ class fish_tipping(gym.Env):
         # in training mode only: punish for population collapse
         if any(pop <= self.threshold) and self.training:
             terminated = True
[…]

### 2023-03-02 code/envs/test_env.py

@@ -1,0 +1,28 @@
+env = fish_tipping(config = {"Tmax": 200})
+env.Tmax
+#env.threshold = 0.00
+env.training = False
+env.reset()
+env.state = np.array([-0.5,-1,-1], dtype=np.float32)
+
+env.state = np.array([-1,-0.5,-1], dtype=np.float32)
+
+env.state = np.array([-0.7,-1,-0.5], dtype=np.float32)
+
+env.population()
+
+df = []
+episode_reward = 0
+action = 0.01
+observation, _ = env.reset()
+for t in range(env.Tmax):
+  df.append(np.append([t, action, episode_reward], observation))
+  observation, reward, terminated, done, info = env.step(action)
+  episode_reward += reward
+  if terminated:
+    break
+  
+import pandas as pd
+cols = ["t", "action", "reward", "X", "Y", "Z"]
+df = pd.DataFrame(df, columns = cols)
+env.population()

### 2023-03-02 code/msy.py

@@ -1,4 +1,3 @@
-import gym_fishing
 import gym
 import pandas as pd
 import numpy as np
@@ -7,19 +6,18 @@ import ray
 @ray.remote
 def simulate(env, action):
   df = []
-  for rep in range(50):
+  for rep in range(5):
     episode_reward = 0
-    observation = env.reset()
+    observation, _ = env.reset()
     for t in range(env.Tmax):
       df.append(np.append([t, rep, action, episode_reward], observation))
-      observation, reward, terminated, info = env.step(action)
+      observation, reward, terminated, done, info = env.step(action)
       episode_reward += reward
       if terminated:
         break
   return(df)
 
-
-env = gym.make("threeFishing-v2")
+env = fish_tipping()
 env.training = False
 actions = np.linspace(0,0.1,101)
 

### 2023-03-02 code/train-agent.py

@@ -6,6 +6,8 @@ import pandas as pd
 import numpy as np
 import torch
 
+register_env("fish_tipping", fish_tipping)
+
 ## We could call env directly without this if only  our envs took a env_config dict argument
 register_env("threeFishing-v2", lambda config: forageVVHcont())
 
@@ -14,10 +16,10 @@ config.num_envs_per_worker=20
 config = config.resources(num_gpus=torch.cuda.device_count())
 config.framework_str="torch"
 config.create_env_on_local_worker = True
-agent = config.build(env="threeFishing-v2")
+agent = config.build(env="fish_tipping")
 
 
-iterations = 200
+iterations = 220
 checkpoint = ("cache/checkpoint_000{}".format(iterations))
 
 if not os.path.exists(checkpoint): # train only if no trained agent saved
@@ -53,8 +55,6 @@ for rep in range(50):
 cols = ["t", "rep", "action", "reward", "X", "Y", "Z"]
 df = pd.DataFrame(df, columns = cols)
 df.to_csv(f"data/PPO{iterations}.csv.xz", index = False)
-
-
 
 ## Plots ## 
 import plotnine

### 2023-03-02 code/plots.Rmd
wip

--- 2023-01-31 code/plots.Rmd
@@
 We compare policies across constant effort (MSY-like policies), constant escapement, and the RL agent (PPO).
-
-<!-- Basic setup -->
-```{r setup}
-library(tidyverse)
-library(patchwork)
-library(RColorBrewer)
-```
-
-<!-- Data sources -->
-```{r}
-# for best const mortality and best const escapement
-msy_optimizer <- function(df) {
-  df |> 
-  group_by(action, rep) |>
-  filter(t == max(t)) |> 
-  group_by(action) |>
-  summarise(mean_reward = mean(reward), sd = sd(reward)) |> 
-  filter(mean_reward == max(mean_reward))
-}
-escapement_optimizer <- function(df) {
-  df |> 
-  group_by(escapement, rep) |>
-  filter(t == max(t)) |> 
-  group_by(escapement) |>
-  summarise(mean_reward = mean(reward), sd = sd(reward)) |> 
-  filter(mean_reward == max(mean_reward))
-}
-
-# Collect data (outsource to separate R script later TBD)
-msy_df <- read_csv("../data/msy.csv.xz")
-best_action <- msy_optimizer(msy_df)$action
-escapement_df <- read_csv("../data/escapement.csv.xz", show_col_types = FALSE)
-best_e <- escapement_optimizer(escapement_df)$escapement
-ppo_df <- read_csv("../data/PPO200.csv.xz")
-
-opt_escapement <- escapement_df |> filter(escapement == best_e)
-
-# set up for plots:
-
-msy_sim <- msy_df |>
-    filter(action == best_action) |> 
-    group_by(rep) |>
-    pivot_longer(c("X", "Y", "Z"),
-                names_to = "species", values_to = "abundance") |>
-    mutate(abundance = abundance + 1, policy = "msy") # natural units, policy name
-
-actions <- unique(msy_df$action)
-i <- which.min(abs(actions - best_action* 0.8))
-tac_sim <- msy_df |> 
-    filter(action == actions[[i]]) |>
-    pivot_longer(c("X", "Y", "Z"),
-        names_to = "species", values_to = "abundance") |>
-    mutate(abundance = abundance + 1, policy = "tac")
-
-ppo_sim <- ppo_df |> 
-    pivot_longer(c("X", "Y", "Z"),
-        names_to = "species", values_to = "abundance") |>
-    mutate(abundance = abundance + 1, policy = "ppo")
-
-esc_sim_df <- escapement_df |> 
-    pivot_longer(c("X", "Y", "Z"),
-        names_to = "species", values_to = "abundance") |>
-    mutate(abundance = abundance + 1)
-
-msy_tac_ppo <- bind_rows(msy_sim, tac_sim, ppo_sim)
-```
-
-```{r}
-#| label: tuning
-#| fig.cap : "Result of the tuning procedure for the constant mortality strategy. 
-#| (A) Mean episode reward (averaged over 50 episodes) shown as scatter data, and shaded in gray are two standard deviations around each scatter point.
-Vertical lines correspond to the optimal constant mortality rate $M^* = $ 0.045 (green) and $0.8 M^*=$ 0.036 (violet).
-#| (B) Several constant mortality policies shown for illustrative purposes.
-Slopes $0.02$, $0.03$, $0.055$ and $0.065$ shown in black; the optimal linear policy $M^*X$ shown in green, while $0.8M^*X$ shown in violet."
-#| out.width : "6.5in"
-msy_rewards <- msy_df |>
-  group_by(action, rep) |>
-  filter(t == max(t)) |> 
-  group_by(action) |>
-  summarise(mean_reward = mean(reward),
-            sd = sd(reward))
-
-rmsy <- msy_rewards |>
-  ggplot(aes(action, mean_reward, 
-             ymin = mean_reward - 2*sd,
-             ymax=mean_reward+2*sd)) +
-  geom_point() + geom_ribbon(alpha=0.4) +
-  labs(y = "episode reward", x = "constant mortality") +
-  theme(aspect.ratio = 1/1) + 
-  geom_vline(aes(xintercept=best_action), color="seagreen3", size=1) + 
-  geom_vline(aes(xintercept=0.8*best_action), color="violetred3", size=1) +
-  theme_bw() + 
-  theme(panel.grid.major = element_blank(), 
-        panel.grid.minor = element_blank(),
-        aspect.ratio = 1)
-
-# linear function blueprint
-linear_f <- function(slope, x) slope*x
-# generates linear function with given slope
-linear_f_gen <- function(slope) (function(x) slope*x)
-
-# dummy data set to keep ggplot happy
-p <- ggplot(data = data.frame(x = 0), mapping = aes(x = x))
-
-# p + stat_function(fun = linear_f_gen(0.5)) + xlim(0,2)
-lines <- p + stat_function(fun = linear_f_gen(0.009)) + 
-  stat_function(fun = linear_f_gen(0.018)) +
-  stat_function(fun = linear_f_gen(0.027)) +
-  stat_function(fun = linear_f_gen(0.8*best_action), col = "violetred3", size=1) +
-  stat_function(fun = linear_f_gen(best_action), col = "seagreen3", size=1) +
-  stat_function(fun = linear_f_gen(0.054)) + 
-  stat_function(fun = linear_f_gen(0.063)) +
-  stat_function(fun = linear_f_gen(0.072)) +
-  stat_function(fun = linear_f_gen(0.081)) +
-  stat_function(fun = linear_f_gen(0.090)) +
-  xlim(0,2) +
-  xlab("X") + ylab("harvest") +
-  theme_bw() +
-  theme(#panel.grid.major = element_blank(), 
-        #panel.grid.minor = element_blank(), 
-        aspect.ratio = 1)
-
-(rmsy | lines) + plot_annotation(tag_levels = c('A'))
-```
-
-<!-- Resuts figures -->
-
-```{r} 
-#| label: timeseries 
-#| fig.cap: "A sample time series for the three fish populations, with the system being controlled, respectively, by the MSY optimal mortality, by $80$ percent of the MSY optimal mortality, and by DRL."
-#| out.width: "6.5in"
-policy_labels <- list("msy" = "MSY Const. Mortality", "tac"= "80% MSY Const. Mortality", "ppo"= "DRL")
-policy_labeller <- function(variable,value){
-  return(policy_labels[value])
-}
-msy_tac_ppo |> 
-    group_by(rep, policy) |>
-    filter(rep == 6) |>
-    ggplot(aes(t, abundance, col = species)) + 
-        geom_line()  + 
-        scale_color_manual(values=c('palegreen3', 'coral1', 'deepskyblue4')) +
-        facet_wrap(~factor(policy, levels = c("msy", "tac", "ppo")), labeller = policy_labeller ) +
-        labs(y = "abundance", x = "time") +
-        theme_bw() +
-        theme(aspect.ratio = 2/3)
-```
-
-```{r} 
-#| label: rew_len_fig
-#| fig.cap: "Bar charts for the episode lengths (top panel) and episode total harvests (bottom panel) for 50 episodes." 
-#| out.width: "6.5in"
-p1 <- msy_tac_ppo |> 
-    group_by(rep, policy) |>
-    filter(t==max(t), species == "X", rep < 50) |>
-    ggplot(aes(rep, t)) + 
-      geom_col() +
-      facet_wrap(~factor(policy, levels = c("msy", "tac", "ppo")), labeller = policy_labeller ) +
-      labs(y = "episode length", x = "repetition") +
-      theme_bw() + 
-      theme(axis.text.x = element_blank())
-
-p2 <- msy_tac_ppo |> 
-    group_by(rep, policy) |>
-    filter(t==max(t), species == "X", rep < 50) |>
-    ggplot(aes(rep, reward)) + 
-      geom_col() +
-      facet_wrap(~factor(policy, levels = c("msy", "tac", "ppo")), labeller = policy_labeller ) +
-      labs(y = "episode harvest", x = "repetition") +
-      theme_bw() + 
-      theme(axis.text.x = element_blank())
-
-(p1 / p2) + plot_annotation(tag_levels = 'A')
-```
-
-```{r}
-#| label: action-X
-#| fig.cap: "(A-B) Harvest and escapement scatter data for the DRL policy together with the optimal constant escapement (pink) and constant mortality (green) policies for comparison. 
-#| (A) includes $Y$ dependence through color, whereas (B) includes $Z$ dependence similarly.
-#| (B-C) escapement (i.e. $X_t - h_t$) scatter data for the DRL policy. 
-#| Optimal constant escapement policy shown in pink for comparison."
-#| out.width: "6.5in"
-opt_escapement_harvest <- function(X) case_when(
-  X > best_e ~ X - best_e,
-  X <= best_e ~ 0
-)
-
-opt_escapement_escapement <- function(X) case_when(
-  X <= best_e ~ X,
-  X > best_e ~ best_e
-)
-
-msy_harvest <- function(X) best_action*X
-tac_harvest <- function(X) 0.8*msy_harvest(X)
-
-py <- ppo_df |>
-  mutate(X = X+1, Y = Y+1, Z = Z+1, harvest = action*X) |>
-  ggplot(aes(X, harvest, color = Y)) +
-  geom_point() +
-  theme_bw() + 
-  theme(aspect.ratio = 1/1) +
-  stat_function(fun = opt_escapement_harvest, col = 'deeppink', size = 0.7) + 
-  stat_function(fun = msy_harvest, col = 'seagreen2', size = 0.7)
-
-pz <- ppo_df |>
-  mutate(X = X+1, Y = Y+1, Z = Z+1, harvest = action*X) |>
-  ggplot(aes(X, harvest, color = Z)) +
-  geom_point() +
-  theme_bw() + 
-  theme(aspect.ratio = 1/1) +
-  stat_function(fun = opt_escapement_harvest, col = 'deeppink', size = 0.7) + 
-  stat_function(fun = msy_harvest, col = 'seagreen2', size = 0.7)
-
-escxy <- ppo_df |>
-  mutate(X = X+1, Y = Y+1, Z = Z+1, escapement = (1-action)*X) |>
-  ggplot(aes(X, escapement, color = Y)) +
-  geom_point() +
-  theme_bw() + 
-  theme(aspect.ratio = 1/1) +
-  stat_function(fun = opt_escapement_escapement, col = 'deeppink', size = 0.7)
-
-escxz <- ppo_df |>
-  mutate(X = X+1, Y = Y+1, Z = Z+1, escapement = (1-action)*X) |>
-  ggplot(aes(X, escapement, color = Z)) +
-  geom_point() +
-  theme_bw() + 
-  theme(aspect.ratio = 1/1) +
-  stat_function(fun = opt_escapement_escapement, col = 'deeppink', size = 0.7)
-
-((py | pz)/ (escxy | escxz)) + plot_annotation(tag_levels = c('A'))
-```
-```{r}
-#| label: state-space
-#| fig.cap: "Policy heatmaps.
-Each policy, say, $\\pi$, was evaluated at points $(x,y,z)$ visited during any of the 50 evaluation rounds.
-We projected this data down to the $X$-$Y$ plane---i.e.
-we plotted the heatmap $\\pi(x,y,z)$ as a function of $x$ and $y$."
-#| out.width: "6.5in"
-i_tac <- which.min(abs(actions - best_action * 0.8))
-
-# color pallette limits
-myPalette <- colorRampPalette(rev(brewer.pal(11, "Spectral")))
-sc <- scale_colour_gradientn(colours = myPalette(100), limits=c(0,0.25))
-
-pmsy <- msy_df |>
-  mutate(X = X+1, Y = Y+1, Z = Z+1, harvest = action * X) |>
-  filter(action == best_action) |>
-  ggplot(aes(X, Y, color = harvest)) +
-  geom_point() + sc +
-  xlim(0,1) +
-  ylim(0,0.85) +
-  theme_bw() + 
-  ggtitle("Opt. Const. Mortality")
-
-ptac <- msy_df |>
-  mutate(X = X+1, Y = Y+1, Z = Z+1, harvest = action * X) |>
-  filter(action == actions[[i_tac]]) |>
-  ggplot(aes(X, Y, color = harvest)) +
-  geom_point() + sc +
-  xlim(0,1) +
-  ylim(0,0.85) +
-  theme_bw() + 
-  ggtitle("80% Opt. Const. Mortality")
-
-pppo <- ppo_df |>
-  mutate(X = X+1, Y = Y+1, Z = Z+1, harvest = action * X) |>
-  ggplot(aes(X, Y, color = harvest)) +
-  geom_point() + sc +
-  xlim(0,1) +
-  ylim(0,0.85) +
-  theme_bw() + 
-  ggtitle("DRL")
-
-pesc <- opt_escapement |>
-  mutate(X = X+1, Y = Y+1, Z = Z+1, 
-        harvest = case_when(
-          X - escapement > 0 ~ X - escapement,
-          X - escapement <= 0 ~  0)
-         ) |> 
-  ggplot(aes(X, Y, color = harvest)) +
-  geom_point() + sc +
-  xlim(0,1) +
-  ylim(0,0.85) +
-  theme_bw() + 
-  ggtitle("Opt. Const. Escapement")
-
-(
-  (pmsy | ptac) / (pppo | pesc) +
-  plot_layout(guides = "collect") & theme(legend.position = 'right')
-) + 
-  plot_annotation(tag_levels = c('A'))
-```
-
-<!-- Appendices -->
-
-```{r} 
-#| label: timeseries_msy
-#| fig.cap: "Nine time series of the system controlled by the MSY constant mortality policy." 
-#| out.width: "6.5in"
-msy_tac_ppo |> 
-    group_by(rep) |>
-    filter(rep < 9, policy == "msy") |>
-    ggplot(aes(t, abundance, col = species)) + 
-        geom_line()  + 
-        scale_color_manual(values=c('palegreen3', 'coral1', 'deepskyblue4')) +
-        facet_wrap(~factor(rep)) +
-        labs(y = "abundance", x = "time") +
-        theme_bw() + 
-        theme(aspect.ratio = 0.5)
-```
-
-```{r} 
-#| label: timeseries_tac
-#| fig.cap: "Nine time series of the system controlled by the policy using 80 percent of the MSY constant mortality." 
-#| out.width: "6.5in"
-msy_tac_ppo |> 
-    group_by(rep) |>
-    filter(rep < 9, policy == "tac") |>
-    ggplot(aes(t, abundance, col = species)) + 
-        geom_line()  + 
-        scale_color_manual(values=c('palegreen3', 'coral1', 'deepskyblue4')) +
-        facet_wrap(~factor(rep)) +
-        labs(y = "abundance", x = "time") +
-        theme_bw() + 
-        theme(aspect.ratio = 0.5)
-```
-
-```{r} 
-#| label: timeseries_ppo
-#| fig.cap: "Nine time series of the system controlled by the trained DRL agent." 
-#| out.width: "6.5in"
-msy_tac_ppo |> 
-    group_by(rep) |>
-    filter(rep < 9, policy == "ppo") |>
-    ggplot(aes(t, abundance, col = species)) + 
-        geom_line()  + 
-        scale_color_manual(values=c('palegreen3', 'coral1', 'deepskyblue4')) +
-        facet_wrap(~factor(rep)) +
-        labs(y = "abundance", x = "time") +
-        theme_bw() + 
-        theme(aspect.ratio = 0.5)
-```
-
-<!-- Old code blocks: -->
 
@@
   group_by(action) |>
-  summarise(mean_reward = mean(reward), sd = sd(reward)) |> 
-  filter(mean_reward == max(mean_reward))
+  summarise(reward = mean(reward)) |> 
+  filter(reward == max(reward))
 }
@@
 
-# Constant Mortality
+# Constant Effort
 
 ```{r}
-msy <- read_csv("../data/msy.csv.xz")
+msy <- read_csv("../data/msy.csv.gz")
 best <- best_reward(msy)
@@
 ```{r}
-# we don't realize all states at each possible action during simulation
-msy |> mutate(X = X+1) |>
- # distinct(action, X) |>
-  filter(action == best$action) |>
-  mutate(harvest = action * X) |>
-  ggplot(aes(X, harvest, col=action, group=action)) + geom_point()
+msy$action |> max()
 
-expand_grid(mortality = seq(0,.1, length.out = 20),
-            state = seq(0,1, length.out=100)) |> 
-  mutate(case = "grid") |> 
-  bind_rows(tibble(mortality = best$action, 
-                   state= seq(0,1, length.out=100),
-                   case = "optimal")) |>
-  mutate(harvest = mortality * state) |>
-  ggplot(aes(state, harvest, col=case, group=mortality)) +
-  geom_line(show.legend = FALSE, lwd=1.5, alpa=0.8) + 
-  scale_color_manual(values=c("grey60", "darkblue"))
-```
-
-```{r}
 rewards <- msy |>
@@
   filter(action == best$action) |> 
-    group_by(rep) |>
-    pivot_longer(c("X", "Y", "Z"),
+  pivot_longer(c("sp1", "sp2", "sp3"),
                names_to = "species", values_to = "abundance") |>
@@
 
-msy_sim
-```
-
-```{r}
 msy_sim |>
-    filter(rep < 5) |>
-    ggplot(aes(t, abundance,  col=species, group =interaction(rep,species)))+ 
+  filter(rep < 4) |>
+  ggplot(aes(t, abundance,  col=species)) + 
     geom_line() + facet_wrap(~rep)
 
+msy_sim |> 
+  filter(rep < 30, species == "sp1") |>
+  ggplot(aes(t, abundance, group = interaction(rep,species),  col=species)) + 
+    geom_line(alpha=0.5)
+
 ```
-
 ```{r}
@@
   group_by(rep) |>
-  filter(t==max(t), species=="X") |>
+  filter(t==max(t), species=="sp1") |>
   ggplot(aes(rep, reward)) + geom_col()
@@
 ```{r}
-# determine which action is 80% of MSY
 actions <- unique(msy$action)
 i <- which.min(abs(actions - best$action* 0.8))
-```
-
-```{r}
 tac_sim <- msy |> 
   filter(action == actions[[i]]) |>
-  pivot_longer(c("X", "Y", "Z"),
+  pivot_longer(c("sp1", "sp2", "sp3"),
                names_to = "species", values_to = "abundance")
 tac_sim |>
-  filter(rep < 20) |>
+  filter(rep < 4) |>
   ggplot(aes(t, abundance,  col=species)) + 
@@
 ```{r}
-ppo <- read_csv("../data/PPO200.csv.xz")
-```
-```{r}
-ppo_sim <- ppo |> 
-  mutate(escapement = (X+1) - action * (X+1),
-         effort = action) |>
-  pivot_longer(c("X", "Y", "Z"),
-               names_to = "species", values_to = "abundance") |>
-  mutate(abundance = abundance + 1)
+ppo <- read_csv("../data/PPO250.csv.gz")
+
 ```
@@
 ```{r}
-ppo_sim |> 
-  group_by(rep) |>
-  filter(t==max(t), species=="X") |>
-  ggplot(aes(rep, reward)) + geom_col()
+ppo_sims <- ppo |> 
+  mutate(escapement = (sp1+1) - action * (sp1+1),
+         effort = action) |>
+  pivot_longer(c("sp1", "sp2", "sp3"),
+               names_to = "species", values_to = "abundance") |>
+  mutate(abundance = abundance + 1)
 
-ppo_sim |> 
-  group_by(rep) |>
-  filter(t==max(t), species=="X") |> 
-  ungroup() |>
-  summarise(mean = mean(reward))
-```
-
-```{r}
-ppo_sim |>
+ppo_sims |>
   filter(rep < 4) |>
@@
 ```{r}
-ppo |>
-  mutate(X = X+1, Z = Z+1, Z = round(Z, 1)) |>
-  distinct(action, X, Z) |> 
-  mutate(harvest = action * X) |>
-  ggplot(aes(X, harvest, group=Z, color=Z)) +
-  geom_smooth(se=FALSE, method=lm,  formula = y ~ splines::ns(x, 6)) +
-  geom_point(alpha=0.6, shape="+")
-
-```
-
-```{r}
-ppo |>
-  mutate(X = X+1, Z = Z+1, Z = round(Z, 1)) |>
-  distinct(action, X, Z) |> 
-  mutate(harvest = action * X,
-         escapement = X - harvest) |>
-  ggplot(aes(X, escapement, group=Z, color=Z)) +
-  #geom_smooth(se=FALSE, method=lm,  formula = y ~ splines::ns(x, 6)) +
-  geom_point(alpha=0.6, shape="+")
-
-```
-
-```{r}
-overall_ppo <- 
-  ppo |> 
-  group_by(rep) |>
-  filter(t==max(t)) |> ungroup() |>
-  filter(reward > 0) |>
-  summarise(reward = mean(reward),
-            t = mean(t),
-            effort = mean(action)) |>
-  mutate(policy="ppo")
-overall_ppo
+ppo_sims |>
+  filter(species == "sp1") |>
+  group_by(t) |>
+  summarise(abundance = mean(abundance), 
+            effort = mean(effort),
+            escapement = mean(escapement)) |> 
+  pivot_longer(-t, names_to = "metric", values_to = "population") |>
+  ggplot(aes(t, population, col=metric)) + geom_line()
 ```
@@
 ```{r}
-escapement <- read_csv("../data/escapement.csv.xz", show_col_types = FALSE)
+escapement <- read_csv("../data/escapement.csv.gz", show_col_types = FALSE)
+
 ## Determine which level of constant escapement gives highest net reward
@@
   filter(action == best_e$action) |>
-  pivot_longer(c("X", "Y", "Z"),
+  pivot_longer(c("sp1", "sp2", "sp3"),
                names_to = "species", values_to = "abundance") |>
   mutate(abundance = abundance + 1) |>
-  rename(escapement = action)
+  rename(escapement = action) |>
+  mutate(effort = pmax(c(abundance - escapement),0) / abundance)
 
@@
 ```
+
+```{r}
+
+ppo |> 
+  group_by(rep) |>
+  filter(t==max(t)) |> ungroup() |>
+  summarise(reward = mean(reward),
+            t = mean(t),
+            effort = mean(action)) |>
+  mutate(policy="ppo")
+
+msy_sim |> 
+  group_by(rep) |>
+  filter(t==max(t)) |> ungroup() |>
+  summarise(reward = mean(reward),
+            t = mean(t),
+            effort = mean(action))|>
+  mutate(policy="msy")
+
+esc_sims |> 
+  group_by(rep) |>
+  filter(t==max(t)) |> ungroup() |>
+  summarise(reward = mean(reward),
+            t = mean(t),
+            effort = mean(effort))|>
+  mutate(policy="escapement")
+
+```

### 2023-03-02 src/explore.Rmd
clean

--- 2023-01-31 manuscript/manuscript.Rmd
@@
 ---
-title: "Pretty darn good control: when are approximate solutions better than approximate models"
-titlerunning: Pretty darn good control
-thanks: | 
-    This material is based upon work supported by the National Science
-    Foundation under Grant No. DBI-1942280.
-
-authors: 
-- name: Felipe Montealegre-Mora
-  address: Department of Environmental Science, Policy, and Management, University of California, Berkeley
-- name: Marcus Lapeyrolerie
-  address: Department of Environmental Science, Policy, and Management, University of California, Berkeley
-- name: Melissa Chapman
-  address: Department of Environmental Science, Policy, and Management, University of California, Berkeley
-- name: Abigail G. Keller
-  address: Department of Environmental Science, Policy, and Management, University of California, Berkeley
-- name: Carl Boettiger
-  address: Department of Environmental Science, Policy, and Management, University of California, Berkeley
-  email: cboettig@berkeley.edu
-
-keywords:
-- Optimal Control
-- Reinforcement Learning
-- Uncertainty
-- Decision Theory
-
-#PACS: 
-#- PAC1
-#- superPAC
-
-#MSC:
-#- MSC code 1
-#- MSC code 2
-
-abstract: |
-  Existing methods for optimal control struggle to deal with the
-  complexity commonly encountered in real-world systems, including
-  dimensionality, process error, model bias and data heterogeneity.
-  Instead of tackling these system complexities directly, researchers 
-  have typically sought to simplify models to fit optimal control 
-  methods. But when is the optimal solution to an approximate, stylized 
-  model better than an approximate solution to a more accurate model? 
-  While this question has largely gone unanswered owing to the 
-  difficulty of finding even approximate solutions for complex models, 
-  recent algorithmic and computational advances in deep reinforcement 
-  learning (DRL) might finally allow us to address these questions. DRL 
-  methods have to date been applied primarily in the context of games 
-  or robotic mechanics, which operate under precisely known rules. Here, 
-  we demonstrate the ability for DRL algorithms using deep neural networks 
-  to successfully approximate solutions (the ‘‘policy function’’ or control 
-  rule) in a non-linear three-variable model for a fishery without knowing 
-  or ever attempting to infer a model for the process itself. We find that 
-  the reinforcement learning agent discovers an effective simplification of
-  the problem to obtain an interpretable control rule. We show that the 
-  policy obtained with DRL is both more profitable and more sustainable 
-  than any constant mortality policy---the standard family of policies 
-  considered in fishery management.
-
-bibliography: bibliography.bib
-biblio-style: spphys
-
-header-includes:
-  - \usepackage{amsmath}
-  - \usepackage{amssymb}
-  - \usepackage{lineno}
-  - \usepackage{setspace}
-  - \usepackage{bookmark}
-  - \usepackage{float}
-  - \usepackage{graphicx}
-  - \usepackage{caption}
-  - \usepackage{subcaption}
-  - \usepackage{tikz-cd}
-  - \usepackage{tcolorbox}
-  - \usepackage{etoolbox}
-  - \usepackage{url}
-  - \linenumbers
-  - \doublespacing
-
-# bibstyle options spbasic(default), spphys, spmpsci
-# output: rticles::springer_article
-
-output: 
-  rticles::arxiv_article:
-    keep_tex: true
-
-editor_options: 
-  markdown: 
-    wrap: sentence
-
+output: github_document
 ---
 
-```{r knit_global, include=FALSE}
-knitr::opts_chunk$set(cache = TRUE, echo = FALSE, message = FALSE, 
-                      warning = FALSE, fig.height=5, fig.width=7.5)
+```{r setup}
 library(tidyverse)
 library(patchwork)
+library(RColorBrewer)
 ```
 
+<!-- Data sources -->
 ```{r}
@@
 best_e <- escapement_optimizer(escapement_df)$escapement
-ppo_df <- read_csv("../data/PPO200.csv.xz")
+
+ppo_df <- read_csv("../data/PPO220.csv.xz")
 
 opt_escapement <- escapement_df |> filter(escapement == best_e)
+```
 
+```{r}
 # set up for plots:
@@
 ```
-
-<!--
- Special Issue Description:
-
-This special issue highlights the development of novel data-driven methods, including statistics, machine learning, parameter estimation, and uncertainty quantification, and combinations thereof, towards modeling biological systems.
-These newly developed methods will tackle challenges that are commonly encountered when modeling real-world experimental, field, pre-clinical, or clinical data.
-Examples of such challenges include high dimensionality, computational complexity, observation or process error, model bias, and intra- or inter-individual heterogeneity.
-Contributions to this special issue require validation of new methods with real-world data or simulated data sets that contain features of real-world data that exemplify an outlined modeling challenge.
-Papers should include a discussion justifying why the developed method is novel and not an application of previously developed methods, as well as how the method may be broadly applicable across different areas of biology, including medical, ecological, genetics, and epidemiological applications.
--->
-
-# Introduction {#intro}
-
-Much effort has been spent grappling with the complexity of our natural world in contrast to the relative simplicity of the models we use to understand it.
-Heroic amounts of data and computation are being brought to bear on developing better, more realistic models of our environments and ecosystems, in hopes of improving our capacity to address the many planetary crises.
- But despite these efforts and advances, we remain faced with the difficult task of figuring out how best to respond to these crises.
-While simplified models have historically allowed for exploration of large decision spaces, the new wave of rich models are applied to highly oversimplified descriptions of potential actions they seek to inform.
-For instance, Global Circulation Models (GCMs) such as HadCM3 [@pope2000impact; @gordon2000simulation; @collins2001internal] model earth's climate using 1.5M variables, while the comparably vast potential action space is modeled as one of six emissions scenarios from the IPCC [@riahi2017shared].
-
-Even as our research community develops simulations of the natural world that fit only in supercomputers, we analyze a space of policies that would fit on index cards.
-Similar combinations of rich process models and highly simplified decision models (often not even given the status of 'model') are common.
-Modeling the potential action space as one of a handful of discrete scenarios is sometimes a well justified acknowledgement of the constraints faced by real-world decision-makers – particularly in the context of multilateral decisions –
-and may seem to reflect a division of responsibilities between 'scientists' modeling the 'natural processes' and policy-makers who make the decisions.
-But, more often, this simplification of decision choices is simply mathematically or conceptually convenient.
-<!-- there are questions of policy scale hiding here -- e.g.
-local regulators of a specific industry obviously face a different action space than national legislatures or international diplomats -->
-This simplification reflects trade-offs between tractablity and complexity at the basis of any mathematical modeling -- if we make both the state space and action space too realistic, the problem of finding the best sequence of actions quickly becomes intractable.
-However, emerging data-driven methods from machine learning offer a new choice -- algorithms that can find good strategies in such problems, but at the cost of opacity.
-
-In this paper, we focus on a well-developed application of model-based management of the natural world that has long illustrated the trade-offs between model complexity and policy complexity: the management of marine fisheries.
-Fisheries management is both an important issue to itself and a rich & frequent test-bed of ecological management more generally.
-Fisheries are an essential natural resource that provide the primary source of protein for one in every four humans, and have faced widely documented declines due to over-fishing [@Worm2006, @Costello2016].
-Fisheries management has  roots in both the fields of _ecosystem management_ and _natural resource economics_.
-Both fields might trace their origins to the notion of maximum sustainable yield (MSY), introduced independently by a fisheries ecologist [@Schaefer1954] and the economist [@Gordon1954] in the same year.
-From this shared origin, each field would depart from the simplifying assumptions of the Gordon-Schaefer model in divergent ways, leading to different techniques for deriving policies from models.The heart of the management problem is easily understood: a manager seeks to set quotas on fishing that will ensure the long-term profitability and sustainability of the industry.
-Mathematical approaches developed over the past century may be roughly divided between these two fields: (A) ecologists, focused on ever more realistic models of the biological processes of growth and recruitment of fish while considering relatively stylized suite of potential management strategies, and (B) economists, focused on far more stylized models of the ecology while exploring a far less constrained set of possible policies.
-The economist's approach can be characterized by the mathematics of a Markov decision process [MDP @Clark1973; @Clark1990; @Marescot2013], in which the decision-maker must observe the stock each year and recommend a possible action.
-In this approach, the policy space that must be searched is exponentially large -- for a management horizon of T decisions and a space of N actions, the number of possible policies is $N^T$.
-In contrast, fisheries ecologists and ecosystem management typically search a space of policies that does not scale with the time horizon.
-Under methods such as "Management Strategy Evaluation" (MSE, [@Punt2016]) a manager identifies a candidate set of "strategies" a priori, and then compares the performance of each strategy over a suite of simulations to determine which strategy gives the best outcome (i.e.
-best expected utility).
-This approach is far more amenable to complex simulations of fisheries dynamics and more closely corresponds to how most marine fishing quotas are managed today in the United States [see stock assessments documented in @RAMLegacyDB].
-<!-- Do we say here that these strategies retain the 'constant mortality' intuition of @Schaefer? -->
-
-```{r}
-#| label: textbox
-#| out.width: "1\\linewidth"
-knitr::include_graphics("figures/textbox.png")
-```
-
-Recent advances in machine learning may allow us to once again bridge these approaches, while also bringing new challenges of their own.
-Novel data-driven methods have allowed these models to evolve into ever more complex and realistic simulations used in fisheries management, where models with over 100 parameters are not uncommon [@RAMLegacyDB].
-Constrained by computational limits, MDP approaches have been intractable on suitably realistic models and largely confined to more academic applications [@Costello2016].
-However, advances _Deep Reinforcement Learning_, (DRL) a sub-field of machine learning, have recently demonstrated remarkable performance in a range of such MDP problems, from video games [@atari1; @atari2] to fusion reactions [@fusion1; @fusion2] to the remarkable dialog abilities of ChatGPT [@chatgpt].
-RL methods also bring many challenges of their own: being notoriously difficult to train and evaluate, requiring immense computational costs, and presenting frequent challenges with transfer-ability or reproducibility.
-A review of all these issues is beyond our scope but can be found elsewhere [@rl-intro; @millie-3].
-Here, though, we will focus on the issue of opacity and interpretability raised by these methods.
-In contrast with optimization algorithms currently used in either ecosystem management or resource economics, RL algorithms have no guarantees of or metrics for convergence to an optimal solution.
-In general, one can only assess the performance of these black box methods relative to alternatives.
-
-```{r}
-#| label: conceptual
-#| echo: FALSE
-#| message: FALSE
-#| fig.align: 'center'
-#| fig.cap:  '(A) Classical management strategies are often derived by reducing the complex dynamics of the fishery ecosystem to a single equation governing the harvested population (say, $X$).
-where the interaction between $X$ and its environment is summarized to a set of parameters, such as the growth rate $r$, and the carrying capacity $K$.
-An advantage of one-dimensional approaches is that the optimal policy is often known exactly, and, moreover, is intuitive.
-For example, in the logistic equation shown, the maximal sustainable yield of the system is attained at K/2.
-#| (B) In contrast to this, we study the problem of managing a three species fishery, where we harvest species $X$.
-#| In addition to $X$, there is a species $Y$ which competes with $X$ for resources, and a species $Z$ which preys on $X$ and $Y$.
-#| We compare two approaches solving harvest strategies in this system. 
-#| The first (Harvest $B_{\mathrm{i}}$) is based on the intuition obtained from 1D control: we use simulated data to empirically determine which constant mortality policy is optimal for controlling this fishery.
-#| The second (Harvest $B_{\mathrm{ii}}$) is an RL policy, where the policy function is expressed as a neural network and the algorithm uses a variant of \emph{gradient ascent} (namely, \emph{Proximal Policy Optimizaton (PPO)}) to optimize this network.
-#| Both approaches we compare are \emph{model-free:} the optimization algorithm treats the model as a black box, only using its input-output statistics.
-#| This way, the optimal policy is not computed from the values of the model parameters, but rather by comparing how policies perform in controlling the system.'
-#| out.width: '6.5in' 
-#| fig.pos: 'H'
-
-knitr::include_graphics("figures/conceptual.png")
-```
-
-<!--NEED some kind of map paragraph here saying what's ahead, what we show  -->
-In this article, we compare the performance of a DRL solution to that of a management strategy evaluation across potential 'constant mortality' policies.\footnote{A repository with all the relevant code to reproduce our results may be found at \url{https://github.com/boettiger-lab/approx-model-or-approx-soln} in the ‘‘code’’ subdirectory.
-The data used is found in the ‘‘data’’ subdirectory, but the user may use the code provided to generate new data sets.}
-We consider the problem of devising harvest strategies in a three species ecosystem (Figure 1B).
-We show that, somewhat surprisingly, the DRL policy closely resembles a *bang-bang* solution (see Text Box 1).
-Bang-bang solutions are known to be optimal for a variety of single-species models, however, there is no a-priori reason why they should be optimal for higher-dimensional systems.
-This way, the DRL agent discovers and exploits, as it were, an effective simplification of the control problem.
-We provide some intuition as to how this simplification arises from the dynamics of the system, but leave a more rigorous characterization of this phenomenon for future work.
-
-The DRL policy drives an oscillation process in the system and, in this way, maintains the long-term sustainability of fishing.
-In contrast, constant mortality policies overfish at low populations and underfish (relative to the DRL solution) at high populations.
-This leads to a tradeoff when leveraging constant mortality strategies: higher constant mortality rates lead to high profitability at the expense of a higher probability of overfishing to extinction, while lower constant mortality rates attain stability at the expense of profit.
-
-The DRL approach side-steps this tradeoff by optimizing over a more complex family of possible policies—policies parametrized by a neural network, as opposed to policies labeled by a single parameter (the mortality rate).
-We thus observe that the DRL solution leads to a higher profitability *and* stability as compared to constant mortality policies.
-
-# Mathematical models of fisheries
-
-In this section we introduce some models describing the population dynamics of fisheries.
-In general, the class of models that appear in this context are *first order finite difference equations* (potentially stochastic difference equations).
-For $n$ species, these models have the general form
-\begin{align}
-  \label{eq:general model}
-  \begin{split}
-    \Delta X_t &= f_X(N_t) - M_X X_t\\
-    \Delta Y_t &= f_Y(N_t) - M_Y Y_t\\
-    &\dots
-  \end{split}
-\end{align}
-where $N_t = (X_t,\ Y_t,\ \dots) \in \mathbb{R}^{n}_+$ is a vector of populations, $\Delta X_t := X_{t+1} - X_t$, $\Delta Y_t := Y_{t+1} - Y_t$, and where $f_i:\mathbb{R}^{n}\to\mathbb{R}$ are arbitrary functions.
-Here, $M_i=M_i(N_t)\in\mathbb{R}_+$ is a state-dependent fish mortality arising from harvesting (sometimes referred to as *fishing effort*).
-Thus, the term $M_X X_t$ is the total $X$ harvest at time $t$.
-We will focus on systems where a single species, $X$, is harvested, i.e. $M_Y = M_Z = \dots = 0$.
-
-## A single species model
-
-Optimal control policies for fisheries are frequently based on 1-dimensional models, $n=1$, as described in $Text$ $Box$ $1$.
-The most familiar model of $f(X)$ is that of *logistic growth*, for which
-
-\begin{align}
-  \label{eq:logistic}
-  f(X_t) = r X_t\big(1 - X_t / K \big) =: L(X_t;\ r, K).
-\end{align}
-
-Real world ecological systems are obviously far more complicated than this simple model suggests.
-One particularly important aspect that has garnered much attention is the potential for the kind of highly non-linear functions that can support dynamics such as alternative stable states and hysteresis.
-A seminal example of such dynamics was introduced in [@may77], using a one-dimensional model of a prey (resource) species under the pressure of a (fixed) predator:
-
-\begin{align}
-  \label{eq:may}
-  f_{\text{May}}(X_t)
-  := L(X_t;\ r, K) - F(X_t, H; \beta,c),
-\end{align}
-where,
-\begin{align*}
-  F(X_t,H;\ \beta,c) := \frac{\beta H X_t^2}{c^2 + X_t^2}.
-\end{align*}
-
-The model has five parameters: the growth rate $r$ and carrying capacity $K$ for $X$, a constant population $H$ of a species which preys on $X$, the maximal predation rate $\beta$, and the predation rate half-maximum biomass $c$.
-
-Eq. \eqref{eq:may} is an interesting study case of a *tipping point* (saddle-node bifurcation) (see Fig. \ref{fig:may}).
-Holding the value of $\beta$ fixed, for intermediate values of $H$ there exist two stable fixed points for the state $\hat X_t$ of the system, these two attractors separated by an unstable fixed point.
-At a certain threshold value of $H$, however, the top stable fixed point collides with the unstable fixed point and both are annihilated.
-For this value of $H$, and for higher values, only the lower fixed point remains.
-This also creates the phenomenon of *hysteresis*, where returning $H$ to its original value is not sufficient to restore $\hat X_t$ to the original stable state.
-
-```{r}
-#| label: may
-#| fig.cap: "The fixed point diagram for the model associated to $f_{\\text{May}}$ as a function of varying the parameter $\\beta H$.
-#| That is, these are the zeroes of the function $f_{\\text{May}}$. 
-#| Stable fixed points (also known as *attractors*) are plotted using a solid line, while the unstable fixed point is shown as a dotted line."
-#| out.width: "6.5in"
-
-knitr::include_graphics("figures/MayStablePoints.pdf")
-```
-
-This structure implies two things.
-First, that a drift in $H$ could lead to catastrophic consequences, with the population $X_t$ plummeting to the lower fixed stable point.
-Second, that if the evolution of $X_t$ is *stochastic*, then, even at values of $H$ below the threshold point, the system runs a sizeable danger of tipping over towards the lower stable point.
-
-<!-- Discuss control solutions of this model(?) -->
-<!-- Briefly motivate reason to consider 3d model, adding competition and predation processes -->
-
-## A three species model
-
-Consider the following three species *generalization* of eq. \eqref{eq:may}:
-
-\begin{align}
-\label{eq:3d model}
-\begin{split}
-  \Delta X_t &= L(X_t;\ r_X, K_X) - F(X_t, Z_t;\ \beta, c) - c_{XY}X_tY_t - h_t,\\
-  \Delta Y_t &= L(Y_t;\ r_Y, K_Y) - D F(Y_t, Z_t;\ \beta, c) - c_{XY}X_tY_t,\\
-  \Delta Z_t &= \left(f(X_t + DY_t\right) - d_Z)Z_t.
-\end{split}
-\end{align}
-The model is conceptualized in Fig. \ref{fig:conceptual}, panel B.
-It contains three populations, the state at time $t$ is $(X_t, Y_t, Z_t)$. 
-Species $Z$ preys on both $X$ and $Y$, while the latter two compete for resources.
-There are ten parameters in this model: 
-The growth rate and carrying capacity, $r_X$, $K_X$, $r_Y$ and $K_Y$, of $X$ and $Y$.
-A parameter $c_{XY}$ mediating a Lotka-Volterra competition between $X$ and $Y$.
-A maximum predation rate $\beta$ and a predation rate half-maximum biomass $c$ specifying how $Z$ forages on $X$ and $Y$.
-A parameter $D$ regulating a relative preference of $Z$ to prey on $Y$.
-Finally, a death rate $d_Z$ and a parameter $f$ related to the birth rate of $Z$.
-
-This model generalizes eq. \eqref{eq:may} in the following sense:
-If $f=d_Z=c_{XY}=0$, and we let $Z_{t=0}=H$, then eq. \eqref{eq:3d model} becomes
-\begin{align*}
-  \Delta X_t &= L(X_t;\ r_X, K_X) - F(X_t, H;\ \beta, c) - h_t,\\
-  \Delta Y_t &= L(Y_t;\ r_Y, K_Y) - D F(Y_t, H;\ \beta, c),\\
-  Z_t &= H.
-\end{align*}
-The equations for $X_t$ and $Y_t$ thus become decoupled, with $X_t$ evolving according to \eqref{eq:may}.
-
-As can be expected, the dynamics of eq. \eqref{eq:3d model} are significantly more complex than those of eq. \eqref{eq:may}.
-There are a variety of fixed points for the equations---around 8 or 9 for wide ranges of parameter values.
-Fig. \ref{fig:fixed points} provides a rough idea of the fixed point structure by plotting the bifurcation diagram obtained from varying $D$ and fixing the other parameters.
-
-\begin{figure}
-  \centering
-  \begin{subfigure}[b]{0.32\textwidth}
-    \centering
-    \includegraphics[width=\textwidth]{figures/stX.png}
-    \caption{$X$ component of fixed points.}
-    \label{fig:stx}
-  \end{subfigure}
-  \hfill
-  \begin{subfigure}[b]{0.32\textwidth}
-    \centering
-    \includegraphics[width=\textwidth]{figures/stY.png}
-    \caption{$Y$ component of fixed points.}
-    \label{fig:sty}
-  \end{subfigure}
-  \hfill
-  \begin{subfigure}[b]{0.32\textwidth}
-    \centering
-    \includegraphics[width=\textwidth]{figures/stZ.png}
-    \caption{$Z$ component of fixed points.}
-    \label{fig:stz}
-  \end{subfigure}
-  \caption{Numerically-computed fixed points of the three-species deterministic model, eq. \eqref{eq:3d model}.
-  Here, the parameter $D$ was varied between 0.75 and 1.25, and the three components of the fixed points are plotted separately, as a function of $D$.}
-  \label{fig:fixed points}
-\end{figure}
-
-Throughout most of this paper we will fix the parameters of the model according to eq. \eqref{eq:param vals}.
-At these values, the fixed points of the model are:\footnote{
-We have included only three decimal digits for readability.
-These fixed points may be generated by running the ‘‘stable\_points.py’’ script in \url{https://github.com/boettiger-lab/approx-model-or-approx-soln}.
-}
-\begin{align*}
-  (X,Y,Z) \approx &(0.839, 0.054, 0.377), \quad 
-    (0.048, 0.774, 0.480), \quad
-    (0.359, 0.491, 0.509), \\
-    &(0.666, 0.666, 0.000),\quad
-    (0.899, 0.000, 0.303), \quad
-    (0.000, 0.818, 0.457), \\
-    &(1.000,0.000,0.000),\quad (0.000,1.000,0.000),\quad (0.000,0.000,0.000).
-\end{align*}
-The classification of fixed points for multi-dimensional systems is generally more complicated than in one dimension.
-A fixed point might be stable against perturbations in one direction, but unstable against perturbations along another.
-While a detailed classification of these fixed points is beyond the scope of this work, we point out that the point $(0.839, 0.054, 0.377)$ is stable against perturbations in all directions.
-We will thus use it as a starting point for our simulations.
-
-## Stochastic dynamics
-
-We have, up to now, presented eqs. \eqref{eq:may} and \eqref{eq:3d model} as *deterministic* dynamical models for the sake of clarity.
-In practice, we will consider a stochastic contribution to these dynamics.
-This way, instead of eq. \eqref{eq:may}, we consider the following stochastic single-species system
-\begin{align}
-  \label{eq:may stoch}
-  \Delta X_t
-  = L(X_t;\ r, K) - F(X_t, H; \beta,c) - h_t + \eta_t,
-\end{align}
-where $\{\eta_t\}_t$ are independent Gaussian variables with zero mean.
-The variance $\sigma^2$ of $\eta_t$ is an additional parameter of the model.
-
-Similarly, we use the following dynamical model in place of eq. \eqref{eq:3d model}
-\begin{align}
-  \label{eq:3d model stoch}
-  \begin{split}
-    \Delta X_t &= L(X_t;\ r_X, K_X) - F(X_t, Z_t;\ \beta, c) - c_{XY}X_tY_t - h_t + \eta_{X,t},\\
-    \Delta Y_t &= L(Y_t;\ r_Y, K_Y) - D F(Y_t, Z_t;\ \beta, c) - c_{XY}X_tY_t + \eta_{Y,t},\\
-    \Delta Z_t &= \left(f(X_t + DY_t\right) - d_Z)Z_t + \eta_{Z,t}.
-\end{split}
-\end{align}
-Here, again, $\{\eta_{i,t}\}$ is a set of independent Gaussian variables centered at zero with variances, respectively, $\sigma_X^2$, $\sigma_Y^2$ and $\sigma_Z^2$.
-
-# Reinforcement learning {#sec:RL}
-
-Reinforcement learning (RL) is a way of approaching *sequential decision problems* through machine learning.
-All applications of RL can be conceptually separated into two parts: an *agent* and an *environment* which the agent interacts with.
-That is, the agent takes actions within the environment.
-As a result of an agent's action, the environment will transition to a new state and return a numerical *reward* to the agent.
-(See Fig.
-1 in [@rl-intro] for a conceptual description of reinforcement learning algorithms.)
-The rewards encode the agent's goal.
-The main task of any RL algorithm is then to maximize the cumulative reward received. 
-This objective is achieved by aggregating experience in what is called the *training* period and learning from such experience.
-<!-- ML: We could cite our deep RL paper here instead of going into further details. Not sure what we are expecting as the audience. -->.
-
-<!-- The *environment* is commonly a computer simulation, although it sometimes can be a real world system.
-The *agent*, on the other hand, is a computer program which interacts with the environment. -->
-
-For our use case, the environment will be a computational model of the population dynamics of a fishery, with the environment state being a vector of all the fish populations, $S=(V_1, V_2, H)$.
-At each time step, the agent harvests one of the populations, $V_1$. 
-This changes the state as
-\begin{align*}
-  (V_1,\ V_2,\ H) \mapsto ((1-q)V_1,\ V_2,\ H),
-\end{align*}
-where $q$ is a fishing *quota* set by the agent. 
-This secures a reward of $qV_1$.
-Afterwards, the environment undergoes a timestep under its natural dynamics given by \eqref{eq:3d model}.
-
-## Mathematical framework for RL
-
-The RL environment can be formally described as a discrete time *partially observable Markov decision process (POMDP)*.
-This formalization is rather flexible and allows one, e.g., to account for situations where the agent may not fully observe the environment state, or where the only observations available to the agent are certain functions of the underlying state.
-For the sake of clarity, we will only present here the class of POMDPs which are relevant to our work: *fully observable MDPs with a trivial emission function* (henceforth MDPs for short).
-An MDP may be defined by the following data:
-
-- $\mathcal{S}$: *state space*, the set of states of the environment,
-
-- $\mathcal{A}$: *action space*, the set of actions which the agent may choose from,
-
-- $T(s_{t+1}|s_t, a_t)$: *transition operator*, a conditional distribution which describes the dynamics of the system,
-
-- $r(s_t, a_t)$: *reward function*, the reward obtained after performing action $a_t\in\mathcal{A}$ in state $s_t$,
-
-- $d(s_0)$: *initial state distribution*, the initial state of the environment is sampled from this distribution,
-
-- $\gamma\in[0,1]$: *discount factor*.
-
-At a time $t$, the MDP agent observes the full state $s_t$ of the environment and chooses an action based on this observation according to a *policy function* $\pi(a_t | s_t)$.
-In return, it receives a discounted reward $\gamma^t r(a_t, s_t)$.
-The discount factor helps regularize the agent, helping the optimization algorithm find solutions which pay off within a timescale of $t \sim \log(\gamma^{-1})^{-1}$.
-
-With any fixed policy function, the agent will traverse a path 
-$\tau=(s_0,\ a_0,\ s_1,\ a_1\ \dots,\ s_{t_{\text{fin.}}})$ 
-sampled randomly from the distribution
-\begin{align*}
-  p_\pi(\tau) = 
-  d(s_0) \prod_{t=0}^{ t_{\text{fin.}}-1 }
-  \pi( a_t | s_t ) T( s_{t+1} | s_t, a_t ).
-\end{align*}
-Reinforcement learning seeks to optimize $\pi$ such that the expected rewards are maximal,
-$$  
-  \pi^* = \mathrm{argmax}\ \mathbb{E}_{\tau\sim p_\pi}[R(\tau)],
-$$
-where,
-$$
-  R(\tau) = \sum_{t=0}^{ t_{\text{fin.}}-1 } \gamma^t r(a_t, s_t),
-$$
-is the cumulative reward of path $\tau$.
-The function $J(\pi):=\mathbb{E}_{\tau\sim p_\pi}[R(\tau)]$ is called the *expected return*.
-
-## Deep Reinforcement Learning
-
-The policy function $\pi$ often lives in a high or even infinite-dimensional space.
-This makes it unfeasible to directly optimize $\pi$.
-In practice, an alternative approach is used: $\pi$ is optimized over a much lower-dimensional parameterized family of functions.
-<!-- ML: I am not following this line of thought here.
-I see the main reason why NN's are used as being that NN's are universal function approximators, and we don't know what the policy and/or value function will look like.
-
-FM: my take here is that the space of functions from states to actions is in principle infinite-dimensional or at least very high dimensional.
-(e.g.
-exponentially high dimensional in terms of the number of degrees of freedom in state space, for the case of discretized state spaces).
-By restricting our ansatz to neural networks of a fixed depth/breadth, we're restricting ourselves to a finite dimensional (or, in the discrete case, potentially much lower dimensional) manifold of policies over which gradient descent may be performed much more efficiently.
-E.g.
-the dimensionality of the space we're searching for is determined by the size of the neural network, it's finite dimensional for continuous state spaces, and its dimension is not necessarily exponential in the number of species.
--->
-Deep reinforcement learning uses this strategy, focusing on function families parameterized by neural networks.
-(See Fig.
-1 and App.
-A in [@rl-intro] for a conceptual introduction to the use of reinforcement learning in the context of conservation decision making.)
-<!--(See Fig. \ref{tbd}.)-->
-<!-- The figure should kinda explain how the parametrization arises, and how gradient ascent is used. Maybe namedrop backpropagataion for the sake of usefulness to the reader? -->
-
-Since our state and observation spaces are continuous, we will focus on deep reinforcement learning throughout this paper.
-Within the RL literature there is a wealth of algorithms from which to choose from to optimize $\pi$, each with its pros and cons.
-Most, if not all, of these are based on gradient ascent by using the technique of *back-propagation* to efficiently compute the gradient.
-Here we have used only one such algorithm (*proximal policy optimization (PPO)*) for the sake of particularity---to draw a clear comparison between the RL-based and the classical fishery management approaches.
-In practice, further improvements can be expected by a careful selection of the optimization algorithm. 
-(See, e.g., [@franccois2018introduction] for an overview of different optimization schemes used in DRL.)
-
-## Model-free reinforcement learning
-
-Within control theory, the classical setup is one where we use as much information from the model as possible in order to derive an optimal solution.
-Here, one may find a vast literature on model-based methods to attain optimal, or near-optimal, control (see, e.g., [@zhang2019near; @sethi2019optimal; @anderson2007optimal]).
-
-The classical sustainable fishery management approaches summarized in Text Box 1, for instance, are model-based controls.
-As we saw there, these controls may run into trouble in the case where there are inaccuracies in the model parameter estimates.
-
-There are many situations, however, in which the exact model of the system is not known or not tractable.
-This is a standard situation in ecology: mathematical models capture the most prominent aspects of the ecosystem's dynamics, while ignoring or summarizing most of its complexity.
-In this case, it is clear, model-based controls run a grave danger of mismanaging the system.
-
-Reinforcement learning, on the other hand, can provide a model-free approach to control theory.
-While a model is often used to generate training data, this model is not directly used by model-free RL algorithms.
-This provides more flexibility to use RL in instances where the model of the system is not accurately known.
-In fact, it has been shown that model-free RL outperforms model-based alternatives in such instances [@mbpo].
-(For recent surveys of model-based reinforcement learning, which we do not focus on here, see [@moerland2023model; @polydoros2017survey].)
-
-This context provides a motivation for this paper. 
-Indeed, models for ecosystem dynamics are only ever approximate and incomplete descriptions of reality. 
-This way, it is plausible that model-free RL controls outperform currently used model-based controls in ecological management problems.
-
-# Methods
-
-\textbf{Training a DRL agent.}
-We trained a DRL agent parametrized by a neural network $\theta$ with two hidden, 64-neuron, layers.
-The state space used was a cube $[0,2]^{\times 3}$, while the policies explored mapped states $N_t$ to a single number, $h_t = \pi_\theta(N_t)$, the harvest quota on species $X$ at that timestep.
-Each episode in the training and in the evaluation stages was at most 200 time steps long.
-Episodes were ended early in case of a ‘‘near-extinction event,’’ a moment in which any of the populations dipped below a threshold value, i.e.,
-\begin{align}
-  \label{eq:thresh}
-  X_t \leq X_{\text{thresh.}},\quad
-  Y_t \leq Y_{\text{thresh.}},\quad
-  \text{or,}\quad
-  Z \leq Z_{\text{thresh.}}.
-\end{align}
-The reward function had two componenets:
-First, at each time step the harvest $h_t$ was received as a reward.
-Second, if the episode ended early (say, at time-step $t$) because of a near-extinction event, a negative reward of $-50/t$ was awarded.
-The constant 50 was chosen so as to select against highly extractive policies in the early stages of training, and thus guide the agent towards sustainable policies.
-This technique, falling under the ‘‘umbrella’’ category of *reward shaping*, is commonly used -- alongside hyper-parameter tuning and state space normalization -- to improve the speed of training by avoiding badly-performing local optima in the gradient ascent algorithm.
-
-Thus, we normalized state space (to be contained in a cube of length 2) and engaged in reward shaping in order to optimize our agent's learning.
-We did not need to tune hyper-parameters for this goal, however.
-
-We used the Ray framework\footnote{\url{https://docs.ray.io/}} for training, in particular we used a Ray PPOConfig object to build the algorithm using the default values for all hyper-parameters.
-The agent was trained for 200 iterations of the PPO optimization step.
-
-\textbf{Tuning the constant mortality strategy.}
-A grid of 100 mortality rates was laid in the interval $[0, 0.1]$.
-For each one of these rates, say $M$, we simulated 50 episodes based on \eqref{eq:3d model stoch}:
-At each time step the state $(X_t, Y_t, Z_t)$ was observed, a harvest of $MX_t$ was collected and then the system evolved according to its natural dynamics \eqref{eq:3d model stoch}.
-The optimal mortality rate was the value $M^*$ for which the mean episode reward was maximal.
-We additionally evaluated a more conservative strategy where a mortality of $0.8 M^*$ instead is used.
-
-The result of the tuning procedure is shown in Fig. \ref{fig:tuning}. 
-There, we see that the optimal value $M^*$ lies in an area with high episode reward variance.
-We will see in Fig. \ref{fig:rew_len_fig} that a likely candidate for this variance is the presence of episodes with early endings.
-Thus, the value $0.8 M^*$ supports a smaller variance, and correspondingly -- as Fig. \ref{fig:rew_len_fig} shows -- this policy leads to a much smaller probability of a near-extinction event.
 
@@
 
-\textbf{Parameter values used.}
-The model's dynamic parameters were chosen as follows
-\begin{align}
-\label{eq:param vals}
-\begin{split}
-  r_X = K_X = r_Y = K_Y = 1, \quad \beta = 0.3, \quad c = 0.1\\
-  c_{XY} = 0.5,\quad f = 0.15,\quad D = 1.1, \quad d_Z = 0.135.
-\end{split}
-\end{align}
-Moreover, the variances for the stochastic terms were chosen as
-$$
-  \sigma^2(\eta_{X,t}) = \sigma^2(\eta_{Y,t}) = \sigma^2(\eta_{Z,t}) = 0.05.
-$$
-The threshold values used to determine an early episode end were chosen as
-$$
-  X_{\text{thresh.}} = 0.08,\quad Y_{\text{thresh.}} = Z_{\text{thresh.}} = 10^{-4},
-$$
-where the higher value of $X_{\text{thresh.}}$ helped select for more sustainable fishing strategies (for both the DRL and constant mortality approaches).
-
-# Results
-
-We evaluated each of the three control approaches (MSY constant mortality, 80% MSY constant mortality and DRL) by running 50 episodes using the respective control rule.
-In Fig. \ref{fig:timeseries} we show the time series of one such repetition.
-The left panel uses an MSY constant mortality control $M^*X$, the middle panel uses $0.8 M^* X$, while the right panel is controlled by the trained DRL agent.
-We see that that optimal constant mortality control leads to a near-extinction event at around $t=150$.
-The constant mortality policy at $0.8 M^*$ is already conservative enough to support a sustainable path.
-There, the population levels are maintained roughly in a stationary state.
-Finally, the DRL time series shows an oscillatory behavior, with a slight dephase between the $Y$ and $Z$ curves, while the $X$ oscillations are roughly anticorrelated with $Z$.
+<!-- Resuts figures -->
 
@@
 ```
-
-This time series is representative of the patterns observed in the data (see App. A.1).
-The episode reward and episode length statistics are shown in Fig. \ref{fig:rew_len_fig}.
-There, we highlight two things:
-First, that the tendency towards early episode ends is fixed by the conservative approach of fishing at 80% of the MSY mortality rate.
-(In App. A.2 we discuss and evaluate other policies using a fixed-percentage of the MSY mortality rate.)
-Second, that while the conservative 80% approach has a stable behavior, it consistently produces smaller harvests than a DRL generated policy.
 
@@
 ```
-
-To understand the DRL-derived policy, Fig. \ref{fig:action-X} depicts the harvest as a function of $X$.
-Specifically, for each point $(x,y,z)$ visited by one of the aforementioned 50 evaluation episodes, we evaluate the harvest collected by the DRL agent and we subsequently plot the scatter data as a function of the projection $x$ onto the $X$ axis.
-In panel A we have used color to add the a $Y$-dependence, while in panel B we did the same with $Z$-dependence.
-This shows that the $Z$ and $Y$ dependence of the policy is rather weak and in fact it has roughly the same shape as an escapement policy.
-
-Fig. \ref{fig:action-X} additionally shows (pink line) the optimal constant escapement policy and the MSY constant mortality policy (green line).
-The optimal constant escapement was computed in a manner similar to how the MSY constant mortality was computed: for each escapement in a grid of 100 values between 0 and 1, 50 episodes were run using the corresponding escapement policy.
-The value with the best mean episode reward was chosen as the optimal escapement value.
-We observe that the DRL policy approximates the optimal constant escapement policy, with the difference of being more conservative in its harvest at high $X$ values.
-This pattern can be seen in panels (C-D), where at high $X$ values, the DRL escapement mildly increases in a non-linear fashion rather than staying constant.
-Finally, relative to these two policies, we observe that MSY constant mortality underfishes at high $X$ values and overfishes at low $X$ values.
 
@@
 ```
-
-In Fig. \ref{fig:state-space}, we perform a back to back visualization of all the policies evaluated---MSY constant mortality, 80% MSY constant mortality, DRL and optimal constant escapement.
-The policies are computed at each state along the paths they traverse in the 50 evaluation rounds, and the values are projected on the $X$-$Z$ plane for visualization.
-This figure highlights the similarity between the policy found by the DRL agent and the optimal escapement policy.
-
-Notice the anti-correlation between $X$ and $Y$ along the paths, this is especially prominent for the constant mortality plots.
-This anti-correlation is, furthermore, observed in the time series plots (see Fig. \ref{fig:timeseries} and App. A.1).
-
 ```{r}
@@
 # color pallette limits
-library(RColorBrewer)
 myPalette <- colorRampPalette(rev(brewer.pal(11, "Spectral")))
@@
 
-# Discussion
+<!-- Appendices -->
 
-Fisheries are complex ecosystems, with interactions between species leading to highly non-linear dynamics.
-While current models for population estimation include many important aspects of this complexity, it is still common to use simplified dynamical models in order to guide decision making.
-This provides very interpretable solutions, such as constant mortality policies.
-There is a drawback here, however: due to the simplicity of these dynamical models, the policies might not respond effectively in many situations---situations where the predictions of these simple models deviate considerably from reality.
-Because of this, policies such as MSY have faced pushback and are believed to have contributed to the depletion of fish stocks [@Worm2006; @Costello2016].
-
-We propose an alternative approach to the problem of fishery control: to use a more expressive -- albeit more complicated -- dynamical model to guide decision making.
-Furthermore, rather than computing the optimal control policy for the model (something that is impossible in practice for complex systems) we use deep reinforcement learning to obtain a ‘‘pretty darn good’’ policy.
-This policy is estimated in a *model-free* setting, i.e., the agent treats the dynamical model (e.g. eq. \eqref{eq:3d model stoch}) as a black box of input-output pairs.
-By not relying on precise knowledge of the model's parameter values, but rather just relying on input-output statistics, model-free approaches have gained traction in a variety of control theory settings (see, e.g., [@sato2019; @ramirez2022; @zeng2019]).
-
-We compare deep reinforcement learning-based policies against optimal constant mortality policies. 
-While the latter are MSY-inspired, they are optimized in a model-free way as well: the optimal constant mortality rate is computed empirically from simulated data.
-
-Because of the simplicity of the constant mortality ansatz, the optimal such policy may be easily estimated (see Figure \ref{fig:tuning}).
-In contrast, DRL optimizes over the more expressive -- and more complicated -- family of policies parametrized by a neural network.
-Because of this, our DRL approach uses a variant of *gradient ascent* (the proximal policy optimization algorithm) to obtain a good candidate policy.
-
-We have shown that the DRL agent finds an approximately *bang-bang* type of solution (see Text Box 1 for the definition of bang-bang solutions and Figs.
-\ref{fig:action-X} and \ref{fig:state-space} for the result).
-While bang-bang solutions are known to be optimal for a wide variety of single-species fishery control problems [@hening2019; @hening2021], there is no such proof of optimality for general multi-species models.
-In this sense this example is rather surprising and reassuring: while no general guarantee points at the optimality of bang-bang solutions for multi-species model, the existence of one such example provides a benchmark for the use of DRL in fishery control.
-
-One intuition for why a constant escapement policy performs so well in such a model comes from the anti-correlation between $X$ and $Y$ noted in the results section.
-We have observed this pattern arise in a variety of policies, both classical as well as some obtained by training DRL agents.
-In fact, ‘‘crashes’’ in the $X$ population are almost invariably shortly preceeded by dramatic increases in $Y$.
-We hypothesize that this effectively leads to a dimensionality reduction: once the value of $X$ is known, along a path traced by the control system, then $Y$ may be roughly predicted.
-A similar, although weaker, anti-correlation is also observed between $X$ and $Z$.
-These two informal observations point at an effective simplification of the prediction process: the values of $Y$ and $Z$ are much less impactful to policy decisions than the value of $X$ itself.
-We have not yet pushed further to establish a more precise picture of this dynamical simplification.\footnote{
-A hand-waving way to understand this phenomenon is the following:
-The $X$-$Y$ anti-correlation can be associated to the Lotka-Volterra competition term between $X$ and $Y$, i.e. $c_{XY}XY$ in eq. \eqref{eq:3d model stoch}.
-Now, consider a fall in population $X$.
-Because of the aforementioned anti-correlation with $Y$, this is accompanied by a growth in $Y$.
-We may understand this, roughly, as a \emph{replacement} of $X$ for $Y$.
-But then, because the parameter $D = 1.1 > 1$, this increases the effective birth rate $Z$, i.e. the term $f(X+DY)Z$ in eq. \eqref{eq:3d model stoch}.
-This, in turn, leads to a growth of $Z$ and, thus, an anti-correlation between $Z$ and $X$.}
-
-This manuscript functions as a first step in the process of incorporating DRL into fishery decision making.
-We have shown that highly non-linear systems may give rise to simple, good policies, and that these may be found via reinforcement learning.
-Moreover, the framework allows for the possibility that these simplifications do not in fact happen, and that good harvest policies are highly non-linear, multi-parameter functions.
-
-In this regard, DRL provides a framework flexible enough to capture the intricate population dynamics that are relevant in fishery management.
-Importantly, DRL is equipped with tools to deal with complex models for which simple policies might yield unsatisfactory results (and thus more refined techniques are needed to control them).
-For example, tools such as *long short-term memories* [@lstm], *curriculum learning* [@curriculum-learning] and *open-ended learning* [@open-ended] can be incorporated when faced with parametric noise or drift, or generally uncertainties about the agent's environment.
-These techniques can be used to obtain agents who perform well over a variety of different environments.
-<!-- ML: Could also use LSTMs for parametric drift.
-Might be better to use a different selling point for curriculum learning like if you want to have a bit more generalizability in your RL agent.
--->
-
-Finally, in our study case, we have shown that the ‘‘rest periods’’ -- periods of several time steps with no fishing at all -- prominent in both the DRL solution and in the optimal bang-bang policy, may be important to maintain the sustainability of the fishery and to increase its long-term profits.
-This hypothesis could be explored in more generality as it might apply to a wide variety of models.
-
-## Acknowledgements
-
-The title of this piece references a mathematical biology workshop at NIMBioS
-organized by Paul Armsworth, Alan Hastings, Megan Donahue, and Carl Towes in
-2011 which first sought to emphasize 'pretty darn good' control solutions to
-more realistic problems over optimal control to idealized ones. 
-This material is based upon work supported by the National Science Foundation under Grant No. DBI-1942280.
-
-\appendix
-
-# Appendix
-
-## Collected time series data
-
-In this appendix we provide a wider selection of time series taken from the same data set as Fig. \ref{fig:timeseries}.
-Namely, Figs. \ref{fig:timeseries_msy}, \ref{fig:timeseries_tac} and \ref{fig:timeseries_ppo} display these time series.
-
+<!-- time series -->
 ```{r} 
@@
 
-## Conservative scale-downs of MSY
+```{r} 
+#| label: timeseries_esc
+#| fig.cap: "Nine time series of the system controlled by the optimal constant escapement policy." 
+#| out.width: "6.5in"
+esc_sim_df |> 
+    group_by(rep) |>
+    filter(rep < 9, escapement == best_e) |>
+    ggplot(aes(t, abundance, col = species)) + 
+        geom_line()  + 
+        scale_color_manual(values=c('palegreen3', 'coral1', 'deepskyblue4')) +
+        facet_wrap(~factor(rep)) +
+        labs(y = "abundance", x = "time") +
+        theme_bw() + 
+        theme(aspect.ratio = 0.5)
+```
 
-Here we show the performance of constant mortality policies with a fraction $\alpha$ of the MSY mortality rate (in the main text we focused on $\alpha=0.8$).
-Their performance is evaluated in Fig. \ref{fig:several-tac}.
-Notice the sharp transition in performance between $\alpha = 0.9$ and $0.95$.
-While the behaviour of the former closely mimics, e.g., the $\alpha=0.8$ policy, the performance of the latter is essentially the same as the optimal constant mortality ($\alpha=1$, see Fig.
-\ref{fig:rew_len_fig}).
-This points at the fact that the value of the parameter $\alpha$, although commonly chosen simply as a rule of thumb, lives on a ‘‘knife's edge,’’ small changes in this parameter may lead to drastic changes in the dynamics of the controlled system.
-What is worse: the particular scale at which this transition happens is generally dependent on the system's intrinsic dynamics.
-In reality, where fishery models and their respective parameters are subject to substantial uncertainty, this fact calls to question the purported additional safety obtained via these rules of thumb.
+<!-- msy scale downs -->
 
@@
 
-## Optimal constant escapement policy evaluation
-
-Here we reproduce the plots from the main text for the optimal constant escapement policy to illustrate the similarity to the DRL-derived policy.
-
-Fig. \ref{fig:esc-tuning-eval} illustrates the tuning process of finding the optimal constant escapment policy (left panel, compare Fig. \ref{fig:tuning}).
-The same figure shows, in the middle panel, the reward histogram for the optimal constant escapement policy (compare Fig. \ref{fig:rew_len_fig}).
-Finally, the right panel displays the same histogram for the DRL policy (reproduced from Fig. \ref{fig:rew_len_fig}).
-We may see that the two policies lead to essentially the same reward distribution.
+<!-- escapement tuning and evaluation -->
 
@@
 ```
-
-Fig. \ref{fig:timeseries_esc} displays 9 time-series for episodes controlled with the optimal constant escapement policy.
-Again, this behaviour is essentially indistinguishable from the behavior of the system controlled by the DRL policy (see Fig. \ref{fig:timeseries_ppo}).
-
-This series of figures gives evidence that, in fact, the DRL agent we have trained has obtained a policy which, effectively, acts as an optimal constant escapement policy.
-Although in Fig.
-\ref{fig:action-X} we found some deviations from the constant escapement case (e.g.
-DRL was more conservative when fishing high populations), these do not seem to lead to a different dynamics.
-
-```{r} 
-#| label: timeseries_esc
-#| fig.cap: "Nine time series of the system controlled by the optimal constant escapement policy." 
-#| out.width: "6.5in"
-esc_sim_df |> 
-    group_by(rep) |>
-    filter(rep < 9, escapement == best_e) |>
-    ggplot(aes(t, abundance, col = species)) + 
-        geom_line()  + 
-        scale_color_manual(values=c('palegreen3', 'coral1', 'deepskyblue4')) +
-        facet_wrap(~factor(rep)) +
-        labs(y = "abundance", x = "time") +
-        theme_bw() + 
-        theme(aspect.ratio = 0.5)
-```
-# References

### 2023-03-02 src/envs/fish_tipping.py

@@ -5,7 +5,7 @@ import numpy as np
 import gymnasium as gym
 from gymnasium import spaces
 
-class fish_tipping(gym.Env):
+class three_sp(gym.Env):
     """A 3-species ecosystem model"""
     def __init__(self, config=None):
         config = config or {}
@@ -23,9 +23,9 @@ class fish_tipping(gym.Env):
          "f": np.float32(0.5), 
          "dH": np.float32(0.45),
          "alpha": np.float32(0.3),
-         "sigma_x": 0, #np.float32(0.05),
-         "sigma_y": 0, #np.float32(0.1),
-         "sigma_z": 0 #np.float32(0.1)
+         "sigma_x": np.float32(0.05),
+         "sigma_y": np.float32(0.1),
+         "sigma_z": np.float32(0.1)
         }
         initial_pop = np.array([0.8396102377828771, 
                                 0.05489978383850558,

### 2023-03-02 src/escapement.py

@@ -1,4 +1,7 @@
-import gym_fishing
+import sys
+sys.path.insert(0, ".") # rstudio / repl_python doesn't add cwd to path
+from src.envs import fish_tipping
+
 import gym
 import pandas as pd
 import numpy as np
@@ -9,18 +12,20 @@ def simulate(env, esc_level):
   df = []
   for rep in range(50):
     episode_reward = 0
-    observation = env.reset()
+    observation, _ = env.reset()
     for t in range(env.Tmax):
+      population = env.population()
       df.append(np.append([t, rep, esc_level, episode_reward], observation))
-      sp1_pop = (observation[0] + 1 ) # natural state-space
-      action = np.max([1 - esc_level / sp1_pop, 0]) # action is a mortality rate. 
+      action = np.max([1 - esc_level / population[0], 0]) # action is a mortality rate. 
+      observation, reward, terminated, done, info = env.step(action)
       episode_reward += reward
       if terminated:
         break
   return(df)
 
-env = gym.make("threeFishing-v2")
-env.training = False
+env = fish_tipping.three_sp()
+env.training = True
+env.threshold = 0
 esc_choices = np.linspace(0,1,101)
 
 # define parllel loop and execute

### 2023-03-02 src/msy.py

@@ -1,3 +1,7 @@
+import sys
+sys.path.insert(0, ".") # rstudio / repl_python doesn't add cwd to path
+from src.envs import fish_tipping
+
 import gym
 import pandas as pd
 import numpy as np
@@ -6,7 +10,7 @@ import ray
 @ray.remote
 def simulate(env, action):
   df = []
-  for rep in range(5):
+  for rep in range(30):
     episode_reward = 0
     observation, _ = env.reset()
     for t in range(env.Tmax):
@@ -17,7 +21,7 @@ def simulate(env, action):
         break
   return(df)
 
-env = fish_tipping()
+env = fish_tipping.three_sp()
 env.training = False
 actions = np.linspace(0,0.1,101)
 

### 2023-03-02 src/train-agent.py

@@ -1,15 +1,15 @@
-from gym_fishing.envs import forageVVHcont
 from ray.rllib.algorithms import ppo
 from ray.tune import register_env
 import os
 import pandas as pd
 import numpy as np
 import torch
+from src.envs import fish_tipping
 
-register_env("fish_tipping", fish_tipping)
+
+register_env("fish_tipping", lambda config: fish_tipping.three_sp())
 
 ## We could call env directly without this if only  our envs took a env_config dict argument
-register_env("threeFishing-v2", lambda config: forageVVHcont())
 
 config = ppo.PPOConfig()
 config.num_envs_per_worker=20
@@ -17,7 +17,6 @@ config = config.resources(num_gpus=torch.cuda.device_count())
 config.framework_str="torch"
 config.create_env_on_local_worker = True
 agent = config.build(env="fish_tipping")
-
 
 iterations = 220
 checkpoint = ("cache/checkpoint_000{}".format(iterations))
@@ -36,18 +35,18 @@ stats = agent.evaluate() # built-in method to evaluate agent on eval env
 
 # Initialize saved copy of eval environment:
 config = agent.evaluation_config.env_config
-config.update({'seed': 41})
+config.update({'seed': 42})
 env = agent.env_creator(config)
 
 env.training = False
 df = []
 for rep in range(50):
   episode_reward = 0
-  observation = env.reset()
+  observation, _ = env.reset()
   for t in range(env.Tmax):
     action = agent.compute_single_action(observation)
     df.append(np.append([t, rep, action[0], episode_reward], observation))
-    observation, reward, terminated, info = env.step(action)
+    observation, reward, terminated, done, info = env.step(action)
     episode_reward += reward
     if terminated:
       break
@@ -72,7 +71,7 @@ ggplot(df2, aes("t", "value", color="variable")) + geom_line()
 
 ## quick policy plot
 df = []
-obs = env.reset()
+obs, _ = env.reset()
 states = np.linspace(-1,0.5,101)
 episode_reward = 0
 for rep in range(5):

### 2023-03-03 code/escapement.py
missing line

@@ -14,6 +14,7 @@ def simulate(env, esc_level):
       df.append(np.append([t, rep, esc_level, episode_reward], observation))
       sp1_pop = (observation[0] + 1 ) # natural state-space
       action = np.max([1 - esc_level / sp1_pop, 0]) # action is a mortality rate. 
+      observation, reward, terminated, info = env.step(action)
       episode_reward += reward
       if terminated:
         break

### 2023-03-03 src/envs/fish_tipping.py
:construction:

@@ -10,12 +10,12 @@ class three_sp(gym.Env):
     def __init__(self, config=None):
         config = config or {}
         parameters = {
-         "r_x": np.float32(0.1),
-         "r_y": np.float32(0.1),
+         "r_x": np.float32(1.0),
+         "r_y": np.float32(1.0),
          "K_x": np.float32(1.0),
          "K_y": np.float32(1.0),
          "beta": np.float32(0.3),
-         "v0":  np.float32(0.5),
+         "v0":  np.float32(0.1),
          "D": np.float32(1.1),
          "tau_yx": np.float32(0),
          "tau_xy": np.float32(0),
@@ -24,8 +24,8 @@ class three_sp(gym.Env):
          "dH": np.float32(0.45),
          "alpha": np.float32(0.3),
          "sigma_x": np.float32(0.05),
-         "sigma_y": np.float32(0.1),
-         "sigma_z": np.float32(0.1)
+         "sigma_y": np.float32(0.05),
+         "sigma_z": np.float32(0.05)
         }
         initial_pop = np.array([0.8396102377828771, 
                                 0.05489978383850558,
@@ -35,6 +35,7 @@ class three_sp(gym.Env):
         ## these parameters may be specified in config                                  
         self.Tmax = config.get("Tmax", 200)
         self.threshold = config.get("threshold", np.float32(1e-4))
+        self.init_sigma = config.get("init_sigma", np.float32(1e-4))
         self.training = config.get("training", True)
         self.initial_pop = config.get("initial_pop", initial_pop)
         self.parameters = config.get("parameters", parameters)
@@ -57,7 +58,9 @@ class three_sp(gym.Env):
     def reset(self, *, seed=None, options=None):
         self.timestep = 0
         self.state = self.update_state(self.initial_pop)
-        return self.state, {}
+        self.state += np.float32(self.init_sigma * np.random.normal(size=3) )
+        info = {}
+        return self.state, info
 
     def step(self, action):
         action = np.clip(action, [0], [1])

### 2023-03-03 src/envs/test_env.py

@@ -1,28 +1,29 @@
-env = fish_tipping(config = {"Tmax": 200})
-env.Tmax
-#env.threshold = 0.00
+from envs import fish_tipping
+
+import gym
+import pandas as pd
+import numpy as np
+import ray
+env = fish_tipping.three_sp()
 env.training = False
-env.reset()
-env.state = np.array([-0.5,-1,-1], dtype=np.float32)
-
-env.state = np.array([-1,-0.5,-1], dtype=np.float32)
-
-env.state = np.array([-0.7,-1,-0.5], dtype=np.float32)
-
-env.population()
-
 df = []
 episode_reward = 0
-action = 0.01
 observation, _ = env.reset()
+esc_level = 0.6
 for t in range(env.Tmax):
-  df.append(np.append([t, action, episode_reward], observation))
+  population = env.population()
+  df.append(np.append([t, esc_level, episode_reward], observation))
+  action = np.max([1 - esc_level / population[0], 0]) # action is a mortality rate. 
   observation, reward, terminated, done, info = env.step(action)
   episode_reward += reward
-  if terminated:
-    break
-  
-import pandas as pd
-cols = ["t", "action", "reward", "X", "Y", "Z"]
-df = pd.DataFrame(df, columns = cols)
-env.population()
+
+print(episode_reward)
+
+cols = ["t", "escapement", "reward", "X", "Y", "Z"]
+df2 = pd.DataFrame(df,columns=cols)
+
+from plotnine import ggplot, geom_point, aes, geom_line, facet_wrap, geom_path
+
+ggplot(df2, aes("t", "X")) + geom_line()
+
+

### 2023-03-03 src/train-agent.py

@@ -1,13 +1,13 @@
+from envs import fish_tipping # setwd to source file location if using repl_python()
 from ray.rllib.algorithms import ppo
 from ray.tune import register_env
 import os
 import pandas as pd
 import numpy as np
 import torch
-from src.envs import fish_tipping
 
 
-register_env("fish_tipping", lambda config: fish_tipping.three_sp())
+register_env("fish_tipping",fish_tipping.three_sp)
 
 ## We could call env directly without this if only  our envs took a env_config dict argument
 
@@ -18,7 +18,7 @@ config.framework_str="torch"
 config.create_env_on_local_worker = True
 agent = config.build(env="fish_tipping")
 
-iterations = 220
+iterations = 200
 checkpoint = ("cache/checkpoint_000{}".format(iterations))
 
 if not os.path.exists(checkpoint): # train only if no trained agent saved
@@ -40,7 +40,7 @@ env = agent.env_creator(config)
 
 env.training = False
 df = []
-for rep in range(50):
+for rep in range(20):
   episode_reward = 0
   observation, _ = env.reset()
   for t in range(env.Tmax):

### 2023-03-04 src/envs/fish_tipping.py
WIP

@@ -23,7 +23,7 @@ class three_sp(gym.Env):
          "f": np.float32(0.5), 
          "dH": np.float32(0.45),
          "alpha": np.float32(0.3),
-         "sigma_x": np.float32(0.05),
+         "sigma_x": np.float32(0.1),
          "sigma_y": np.float32(0.05),
          "sigma_z": np.float32(0.05)
         }
@@ -35,7 +35,7 @@ class three_sp(gym.Env):
         ## these parameters may be specified in config                                  
         self.Tmax = config.get("Tmax", 200)
         self.threshold = config.get("threshold", np.float32(1e-4))
-        self.init_sigma = config.get("init_sigma", np.float32(1e-4))
+        self.init_sigma = config.get("init_sigma", np.float32(1e-3))
         self.training = config.get("training", True)
         self.initial_pop = config.get("initial_pop", initial_pop)
         self.parameters = config.get("parameters", parameters)

### 2023-03-04 src/msy.py

@@ -1,6 +1,4 @@
-import sys
-sys.path.insert(0, ".") # rstudio / repl_python doesn't add cwd to path
-from src.envs import fish_tipping
+from envs import fish_tipping
 
 import gym
 import pandas as pd
@@ -23,7 +21,7 @@ def simulate(env, action):
 
 env = fish_tipping.three_sp()
 env.training = False
-actions = np.linspace(0,0.1,101)
+actions = np.linspace(0,0.2,101)
 
 # define parllel loop and execute
 parallel = [simulate.remote(env, i) for i in actions]

### 2023-03-04 src/train-agent.py

@@ -1,4 +1,5 @@
-from envs import fish_tipping # setwd to source file location if using repl_python()
+
+from src.envs import fish_tipping
 from ray.rllib.algorithms import ppo
 from ray.tune import register_env
 import os
@@ -12,13 +13,14 @@ register_env("fish_tipping",fish_tipping.three_sp)
 ## We could call env directly without this if only  our envs took a env_config dict argument
 
 config = ppo.PPOConfig()
+config.training(vf_clip_param = 50.0)
 config.num_envs_per_worker=20
 config = config.resources(num_gpus=torch.cuda.device_count())
 config.framework_str="torch"
 config.create_env_on_local_worker = True
 agent = config.build(env="fish_tipping")
 
-iterations = 200
+iterations = 150
 checkpoint = ("cache/checkpoint_000{}".format(iterations))
 
 if not os.path.exists(checkpoint): # train only if no trained agent saved
@@ -40,7 +42,7 @@ env = agent.env_creator(config)
 
 env.training = False
 df = []
-for rep in range(20):
+for rep in range(50):
   episode_reward = 0
   observation, _ = env.reset()
   for t in range(env.Tmax):
@@ -71,18 +73,16 @@ ggplot(df2, aes("t", "value", color="variable")) + geom_line()
 
 ## quick policy plot
 df = []
-obs, _ = env.reset()
 states = np.linspace(-1,0.5,101)
-episode_reward = 0
-for rep in range(5):
-  for observation in states:
-      obs[0] = observation
+for rep in range(10):
+  obs, _ = env.reset()
+  for state in states:
+      obs[0] = state
       action = agent.compute_single_action(obs)
-      df.append([observation, action[0]])
-df2 = pd.DataFrame(df, columns =  ["observation","action"])
-
-# transform to natural space & compute escapement
-df2["escapement"] = df2["observation"] + 1 - df2["action"]
-ggplot(df2, aes("observation", "escapement")) + geom_point()
+      escapement = max(state + 1 - action[0], 0)
+      df.append([state+1, escapement, action[0], rep])
+      
+df2 = pd.DataFrame(df, columns=["observation","escapement","action","rep"])
+ggplot(df2, aes("observation", "escapement", color = "rep")) + geom_point(shape=".")
 
 

### 2023-06-13 src/envs/two_three_fishing.py
msy, esc performance

@@ -20,7 +20,7 @@ class twoThreeFishing(gym.Env):
       "growth_fn", 
       lambda x: np.array(x * (1 - x), dtype = np.float32)
     )
-    self.num_speces = 3
+    self.num_species = 3
     #
     self.initial_pop = config.get("initial_pop", _DEFAULT_INIT_POP)
     self.Tmax = config.get("Tmax", 200)
@@ -111,6 +111,7 @@ class oneThreeFishing(gym.Env):
       "growth_fn", 
       lambda x: np.array(x * (1 - x), dtype = np.float32)
     )
+    self.num_species = 3
     #
     self.initial_pop = config.get("initial_pop", _DEFAULT_INIT_POP)
     self.Tmax = config.get("Tmax", 200)

### 2023-06-13 src/esc_heatmap.py

@@ -1,0 +1,55 @@
+import pandas as pd
+import os
+from eval_util import (
+  generate_episodes, episode_plots, state_policy_plot, values_at_max_t,
+  GaussianProcessPolicy, generate_gpp_episodes
+)
+from parameters import parameters
+from envs import two_three_fishing
+from two_fishery_esc_fns import esc_grid_heatmap, esc_grid_ep_ends_heatmap
+from train_fns import create_agent
+
+DATACODE = "RXDRIFT"
+ENVCODE = "2FISHERY"
+DATAPATH = os.path.join("../data/results_data", ENVCODE, DATACODE)
+os.makedirs(DATAPATH, exist_ok=True)
+
+# define problem
+env_class = two_three_fishing.twoThreeFishing
+parameters_obj = parameters()
+
+# create agent
+agent = create_agent(
+  env_class = env_class,
+  parameters_obj=parameters_obj,
+  datacode = DATACODE, 
+  env_name="twoThreeFishing", 
+  fluctuating=True, 
+  training=True,
+)
+
+# create associated env
+env_config = agent.evaluation_config.env_config
+env_config.update({'seed': 42})
+env = agent.env_creator(env_config)
+
+#esc_grid_heatmap(
+esc_grid_ep_ends_heatmap(
+  env,
+  path = DATAPATH,
+  filename ="esc_heatmap.png",
+  grid_nr = 51,
+  repetitions = 100,
+)
+
+
+
+"""
+esc = pd.read_csv(os.path.join(DATAPATH, "esc.csv.xz"))
+esc_max_t = values_at_max_t(esc)
+
+esc_max_t["reward_std"] = esc_max_t["reward"]
+result = esc_max_t.agg({"reward":"mean", "reward_std":"std"})
+
+print(result)
+"""

### 2023-06-13 src/eval_util.py

@@ -117,9 +117,10 @@ def evaluate_policy(
   bound = 4,
   ):
   """ ranges in large [0,bound] 'population space' (0,4 in practice) """
+  nr_gridpoints_to_eval = 51
   # X - policy
   x_policy = []
-  for x in np.linspace(0,1,100):
+  for x in np.linspace(0,1,nr_gridpoints_to_eval):
     for yz in it.product(
       np.linspace(*range_Y,5), 
       np.linspace(*range_Z,5), 
@@ -136,7 +137,7 @@ def evaluate_policy(
   
   # Y - policy
   y_policy = []
-  for y in np.linspace(0,1,100):
+  for y in np.linspace(0,1,nr_gridpoints_to_eval):
     for xz in it.product(
       np.linspace(*range_X,5), 
       np.linspace(*range_Z,5), 
@@ -154,7 +155,7 @@ def evaluate_policy(
       
   # Z - policy
   z_policy = []
-  for z in np.linspace(0,1,100):
+  for z in np.linspace(0,1,nr_gridpoints_to_eval):
     for xy in it.product(
       np.linspace(*range_X,5), 
       np.linspace(*range_Y,5), 
@@ -199,6 +200,7 @@ def state_policy_plot(agent, env, path_and_filename = None):
   
   return pd.concat([x_policy_df, y_policy_df, z_policy_df])
 
+
 # Interpolate policy
 
 from sklearn.gaussian_process import GaussianProcessRegressor
@@ -216,10 +218,12 @@ def GaussianProcessPolicy(policy_df, length_scale=10, noise_level=0.1):
     1.0 * RBF(length_scale = length_scale) 
     + WhiteKernel(noise_level=noise_level)
     )
+  print("Fitting Gaussian Process...")
   gpp = (
     GaussianProcessRegressor(kernel=kernel, random_state=0)
     .fit(predictors, targets)
     )
+  print("Done fitting Gaussian Process...")
   return gpp
 
 
@@ -479,7 +483,7 @@ def episode_plots_1species(df, *, path_and_filename = "path_plots.png"):
     df2.loc[
       (df2.variable == "reward")
     ].groupby(["rep"]), 
-    aes("t", "value")
+    aes("t", "value", color="variable")
   ) + geom_line()
   
   
[…]

### 2023-06-13 src/msy.py

@@ -1,70 +1,52 @@
-from envs import fish_tipping, growth_functions
+from envs import oneSpecies, two_three_fishing, growth_functions
+from msy_fns import find_msy_1fish
+from train_fns import create_env
+from parameters import parameters_oneSp, parameters
+from eval_util import values_at_max_t
 
+from plotnine import ggplot, geom_bar, geom_point, geom_line, aes
 import gymnasium as gym
 import pandas as pd
 import numpy as np
 import ray
+import os
 
-@ray.remote
-def simulate(env, action):
-  df = []
-  for rep in range(30):
-    episode_reward = 0
-    observation, _ = env.reset()
-    for t in range(env.Tmax):
-      df.append(np.append([t, rep, action, episode_reward], observation))
-      observation, reward, terminated, done, info = env.step(action)
-      episode_reward += reward
-      if terminated:
-        break
-  return(df)
+REPS = 100
+MSY_GRID_SIZE = 51
+ENVCODE = "1FISHERY"
+DATACODE = "DEFAULT"
+FLUCTUATING = False
+DATAPATH = os.path.join("../data/results_data", ENVCODE, DATACODE)
+ENV_CLASS = two_three_fishing.oneThreeFishing #oneSpecies.singleSp
 
-config = {}
-config["growth_fn"] = growth_functions.v0_drift_growth
-config["fluctuating"] = True
-env = fish_tipping.three_sp(
-    config = config
+parameter_obj = parameters() #parameters_oneSp()
+
+env = create_env(
+  env_class = ENV_CLASS, 
+  parameter_obj=parameter_obj, 
+  datacode = "DEFAULT", 
+  fluctuating=False, 
+  training=True,
 )
 
-
-_DATACODE = "V0DRIFT"
-env.training = False
-actions = np.linspace(0,0.2,101)
-
-# define parllel loop and execute
-parallel = [simulate.remote(env, i) for i in actions]
-df = ray.get(parallel)
-
-# convert to data.frame & write to csv
-cols = ["t", "rep", "action", "reward", "X", "Y", "Z"]
-df = pd.DataFrame(np.vstack(df), columns = cols)
-
-df2 = (
-  df
-  .loc[df.t == df.t.max()]
-  .groupby(["action"])
-  .agg({"reward": "mean"})
[…]

### 2023-06-13 src/msy_fns.py

@@ -1,0 +1,142 @@
+import ray
+import pandas as pd
+import numpy as np
+import itertools
+
+@ray.remote
+def msy_max_t_1fish(env, mortality, repetitions):
+  x = []
+  path = []
+  for rep in range(repetitions):
+    episode_reward = 0
+    observation, _ = env.reset()
+    T = 0
+    for t in range(env.Tmax):
+      population = env.population()
+      act= mortality
+      action = np.array([act], dtype = np.float32)
+      path.append([t, rep, mortality, act, episode_reward, *population])
+      observation, reward, terminated, done, info = env.step(action)
+      episode_reward += reward
+      if terminated:
+        T = t
+        break
+      else:
+        T = env.Tmax
+    x.append([T, rep, mortality, act, episode_reward, *population])
+  return(x, path)
+
+@ray.remote
+def msy_max_t(env, mortality_x, mortality_y, repetitions):
+  x = []
+  path = []
+  for rep in range(repetitions):
+    episode_reward = 0
+    observation, _ = env.reset()
+    T = 0
+    for t in range(env.Tmax):
+      population = env.population()
+      act_x, act_y= mortality_x, mortality_y
+      action = np.array([act_x, act_y], dtype = np.float32)
+      path.append([t, rep, mortality_x, mortality_y, act_x, act_y, episode_reward, *population])
+      observation, reward, terminated, done, info = env.step(action)
+      episode_reward += reward
+      if terminated:
+        T = t
+        break
+      else:
+        T = env.Tmax
+    x.append([T, rep, mortality_x, mortality_y, act_x, act_y, episode_reward, *population])
+  return(x, path)
+
+def generate_msy_episodes(env, grid_nr=101, repetitions=100, only_max_times = False):
+  mortality_choices = np.linspace(0,0.5,grid_nr)
+
+  # define parllel loop and execute
+  parallel = [msy_max_t_1fish.remote(env, i, repetitions) for i in mortality_choices]
+  x = ray.get(parallel) # list of tuples of final point, history
+  
+  cols = ["t", "rep", "mortality", "act", "reward", "X", "Y", "Z"] # assume 3 species as default
+  if env.num_species == 1:
[…]

### 2023-06-13 src/parameters.py

@@ -65,3 +65,5 @@ class parameters_oneSp(parameters):
   
   def init_state(self):
     return np.array([0.7], dtype=np.float32)
+  
+  

### 2023-06-13 src/train_fns.py

@@ -50,7 +50,7 @@ def create_env(
 	config["growth_fn"] = _datacode_to_growth_fn[datacode]
 	config["fluctuating"] = fluctuating
 	config["training"] = training
-	config["initial_pop"] = parameters.init_state()
+	config["initial_pop"] = parameter_obj.init_state()
 	env = env_class(
 	    config = config
 	)

### 2023-06-13 src/workflow.py

@@ -10,12 +10,16 @@ from eval_util import (
   GaussianProcessPolicy, generate_gpp_episodes,
   generate_episodes_1fish, episode_plots_1fish, state_policy_plot_1fish,
   GaussianProcessPolicy_1fish, generate_gpp_episodes_1fish, 
+  gpp_policy_plot_1fish, gpp_policy_plot_2fish,
+  
 )
+from msy_fns import find_msy_2fish, find_msy_1fish
 
 import matplotlib.pyplot as plt
 import os
 import seaborn as sns
 import pandas as pd
+import numpy as np
 from plotnine import (
   ggplot, geom_point, aes, geom_line, facet_wrap, geom_path, geom_bar, geom_histogram
 )
@@ -30,14 +34,14 @@ Data-code list:
 '''
 
 # globals
-NAME_MODIFIER = ""
-ITERATIONS = 200
+NAME_MODIFIER = "V1_REPLICATE"
+ITERATIONS = 300
 REPS = 100
 ESC_GRID_SIZE = 51
-DATACODE = "KLIMIT_RXDRIFT"
-FLUCTUATING = True
-ENVCODE = "2FISHERY"
-DATAPATH = os.path.join("../data/results_data", ENVCODE, DATACODE)
+DATACODE = "DEFAULT"
+FLUCTUATING = False
+ENVCODE = "1FISHERY"
+DATAPATH = os.path.join("../data/results_data", ENVCODE, DATACODE, NAME_MODIFIER)
 ENVCODE_TO_ENV = {
   "1FISHERY": two_three_fishing.oneThreeFishing,
   "2FISHERY": two_three_fishing.twoThreeFishing,
@@ -45,11 +49,6 @@ ENVCODE_TO_ENV = {
 ENV_CLASS = ENVCODE_TO_ENV[ENVCODE]
 
 os.makedirs(DATAPATH, exist_ok=True)
-
-def what_to_do(workflow_choice):
-  _ = ray.get(
-    [workflow_choice.remote() for _ in range(1)]
-  )
 
 
 # escapement
@@ -76,12 +75,18 @@ def workflow_twoFishery():
   
   # uncontrolled
   generate_uncontrolled_timeseries_plot(
-    env, path_and_filename=f"{DATAPATH}/uncontrolled_timeseries.png", T=200, reps=1
+    env, path_and_filename=os.path.join(DATAPATH, "uncontrolled_timeseries.png"), T=200, reps=1
   )
[…]

### 2023-06-13 src/workflow_oneSp.py

@@ -1,0 +1,191 @@
+from parameters import parameters_oneSp
+from envs import oneSpecies
+from eval_util import (
+  generate_episodes_1fish, episode_plots_1species,
+  state_policy_plot_1species,
+  GaussianProcessPolicy_1species, generate_gpp_episodes_1fish,
+  values_at_max_t, gpp_policy_plot_1species,
+)
+from train_fns import create_agent, train_agent
+from uncontrolled_fns import generate_uncontrolled_timeseries_plot_oneSp
+from one_fishery_esc_fns import find_best_esc_1fish
+from msy_fns import find_msy_1fish
+
+import matplotlib.pyplot as plt
+import os
+import seaborn as sns
+import pandas as pd
+from plotnine import (
+  ggplot, geom_point, aes, geom_line, facet_wrap, geom_path, geom_bar, geom_histogram
+)
+import ray
+
+# globals
+ITERATIONS = 80
+REPS = 100
+ESC_GRID_SIZE = 101
+DATACODE = "ONESP"
+FLUCTUATING = False
+DATAPATH = os.path.join("../data/results_data", DATACODE)
+ENV_CLASS = oneSpecies.singleSp
+
+os.makedirs(DATAPATH, exist_ok=True)
+
+@ray.remote(num_gpus=1, num_cpus=20)
+def workflow_oneSpecies():
+  # define problem
+  parameters_obj = parameters_oneSp()
+  
+  # create agent
+  agent = create_agent(
+    env_class = ENV_CLASS,
+    parameters_obj=parameters_obj,
+    datacode = DATACODE, 
+    env_name="oneSpFishing", 
+    fluctuating=FLUCTUATING, 
+    training=True,
+  )
+  
+  # create associated env
+  env_config = agent.evaluation_config.env_config
+  env_config.update({'seed': 42})
+  env = agent.env_creator(env_config)
+  
+  # uncontrolled
+  generate_uncontrolled_timeseries_plot_oneSp(
+    env, path_and_filename=f"{DATAPATH}/uncontrolled_timeseries.png", T=200, reps=5
+  )
+  
+  # escapement
+  esc_filename = f"{DATAPATH}/esc_{REPS}.csv.xz"
[…]

### 2023-06-14 manuscript/figures/classical_fisheries_textbox.Rmd
modified textbox

@@
 
-There are several existing approaches that have been used to manage fisheries, including \textbf{escapement}, control rules associated with \textbf{maximum sustainable yield (MSY)}, among others.
+There are several existing approaches that have been used to manage fisheries, most prominently including \textbf{constant escapement policies} and \textbf{constant mortality policies}. 
 We collectively refer to these as $classical$, and will compare their performance to RL-based management strategies. 
@@
 In the equation above, $h_t$ is the $harvest$ at time step $t$.
-The goal is to choose the harvest policy $h:X_t\mapsto h_t$, such that long-term profits are maximized.
+The goal is to choose the harvest policy $h:X_t\mapsto h_t$, such that long-term catch is maximized.
 
@@
 The optimizer is an \textbf{escapement} policy, which corresponds to a harvest, $h_t$, that keeps the system at its optimal growth rate as much as possible:
-
 \begin{equation*}
   h_t = \begin{cases}
-    X_t - K/2, \qquad &\text{if } X_t > K/2\\
-    0, \qquad &\text{else.}
+    X_t - a, \qquad &\text{if } X_t > a\\
+    0, \qquad &\text{else,}
   \end{cases}
 \end{equation*}
+where $a$ is the stock size at which growth is maximized.
+For example, in the logistic growth example above, $a=K/2$.
 
-Escapement policies, or more generally $bang$ $bang$ policies, tend to be the optimal solution for these types of control problems.
+This type of \emph{bang-bang} policy tends to be the optimal solution for these types of control problems. 
 A drawback of these solutions in the fishery context, is the possible presence of several time steps with zero harvest.
@@
 
-One ubiquitous solution, often called \textbf{maximum sustainable yield (MSY)}, provides biological reference points that can be simply calculated from the surplus production model parameters.
-The population size at the maximum growth rate, $X = K/2$, is approached asymptotically from any positive initial state $X_0 \in (0,K)$, under a constant mortality $M_{MSY}=r/2$ policy.
-At equilibrium, this produces the maximum sustainable yield:
-
+One ubiquitous solution is based on a constant mortality policy:
+\begin{equation*}
+  h_t = aX_t,
+\end{equation*}
+for some constant $a$.
+The policy with optimal value of $a$ is known as a \emph{maximum sustainable yield (MSY)} policy.
+In the logistic example above, this optimum mortality rate is $a=rK/4$. 
+Under this policy, the stock size at the maximum growth rate $X_{MSY} = K/2$ is approached asymptotically from any positive initial state $X_0 \in (0,K)$.
+Thus, in equilibrium the MSY policy tends to match the results of constant escapement, namely, having the harvest rate be the maximum sustainable yield of the model: 
 \begin{equation*}
@@
 \end{equation*}
+That is, at the MSY biomass, $X_{MSY}$, the logistic growth of $X$ is cancelled exactly by the harvest.
 
-That is, at the MSY biomass, $X_{MSY}$, the logistic growth of $X$ is cancelled exactly by the harvest.
-This \textbf{MSY constant mortality} rule fixes the drawback in the escapement policy by having $h(X)>0$ for all $X>0$.
+This MSY policy fixes the drawback of the escapement policy---$h(X)>0$ for all $X>0$.
 It, however, has its own drawbacks, as it is particularly sensitive to misestimates of the parameter $r$.
-
 Due to this, similar but more conservative policies are often applied where the constant rate of fishing mortality is $< M_{MSY}$.

### 2023-06-14 manuscript/manuscript.Rmd
a bit of intro

@@
 While simplified models have historically allowed for exploration of large decision spaces, the new wave of rich models are applied to highly oversimplified descriptions of potential actions they seek to inform.
-For instance, Global Circulation Models (GCMs) such as HadCM3 [@pope2000impact; @gordon2000simulation; @collins2001internal] model earth's climate using 1.5M variables, while the comparably vast potential action space is modeled as one of six emissions scenarios from the IPCC [@riahi2017shared].
+For instance, Global Circulation Models (GCMs) such as HadCM3 [@pope2000impact; @gordon2000simulation; @collins2001internal] model earth's climate using 1.5M variables, while the comparably vast potential action space is modeled as one of seven emissions marker scenarios from the IPCC [@riahi2017shared].
 
@@
 <!--NEED some kind of map paragraph here saying what's ahead, what we show  -->
-In this article, we compare the performance of a DRL solution to that of a management strategy evaluation across potential 'constant mortality' policies.\footnote{A repository with all the relevant code to reproduce our results may be found at \url{https://github.com/boettiger-lab/approx-model-or-approx-soln} in the ‘‘code’’ subdirectory.
+In this article, we compare the performance of a DRL solution to that of classical management strategies (see Textbox 1).\footnote{A repository with all the relevant code to reproduce our results may be found at \url{https://github.com/boettiger-lab/approx-model-or-approx-soln} in the ‘‘src’’ subdirectory.
 The data used is found in the ‘‘data’’ subdirectory, but the user may use the code provided to generate new data sets.} 
-We consider the problem of devising harvest strategies in a three species ecosystem (Figure 1B).
+We consider the problem of devising harvest strategies in a for a series of ecosystem models (Figure **TBD**):
+*1) one species, one fishery:* a simple single-species recruitment model based on [@may77];
+*2) three species, one fishery:* a three-species generalization of model *1)*, where one of the species is harvested;
+*3) three species, two fisheries:* the same three-species model as above but with two harvested species;\footnote{
+As we prepared this manuscript we became aware of [@nicolalove] which considers an ecological control problem with a similar mathematical formulation, but using an analytical calculus-based approach.
+}
+*4) three species, two fisheries, parameter variation:* a three-species model of which two are harvested, as above, with a time-varying parameter.
+This last model is meant to be a toy model of climate change's effect on the system.
+
 We show that, somewhat surprisingly, the DRL policy closely resembles a *bang-bang* solution (see Text Box 1).

### 2023-06-14 manuscript/manuscript.Rmd
intro

@@
 This last model is meant to be a toy model of climate change's effect on the system.
+Across all of these scenarios, two goals are balanced in the decision process: maximizing long-term catch and preventing stock sizes to fall below some a-priori threshold.
+
+Regarding control for these 4 models, we show that: *1)* while all approaches match in performance in the single species deterministic model, even mild stochasticity plummets the performance of MSY relative to constant escapement.
+Moreover, DRL recovers the optimal constant escapement policy.
+*2)* In the three species, single fishery model, all management strategies perform similarly.
+*3)* In the three species, two fisheries model DRL outperforms both classical strategies, with MSY surprisingly performing significantly better than constant escapement.
+In particular, we observe that DRL strategies are more sensitive to stochastic variations of the system which allows it to adaptively manage the system to prevent population collapses below the threshold.
+*4)* In the three species, two fisheries model with parametric time-variation, the performance gap between DRL and MSY is broadened.
+In this case, as in *3)*, constant escapement strategies perform significantly worse than DRL and MSY.
+
+Our findings paint a picture of how a single-species optimal management strategy may loose performance rather dramatically when controlling a more complex ecosystem.
+Moreover, rather unintuitively, management strategies, such as MSY, known to be suboptimal and unsustainable in the single-species can turn out to even outperform the aforementioned single-species optimal strategy (e.g.
+constant escapement).
+Finally, within this regime of complex, possibly varying, ecosystems, we show that DRL consistently finds a policy which effectively either matches the best classical strategy, or outperforms it.
+We strengthen this result with a stability investigation: we show that random perturbations in the model's parameter values used do not significantly vary the conclusion that DRL outperforms constant escapement.
 
@@
 
-The DRL policy drives an oscillation process in the system and, in this way, maintains the long-term sustainability of fishing.
-In contrast, constant mortality policies overfish at low populations and underfish (relative to the DRL solution) at high populations.
-This leads to a tradeoff when leveraging constant mortality strategies: higher constant mortality rates lead to high profitability at the expense of a higher probability of overfishing to extinction, while lower constant mortality rates attain stability at the expense of profit.
-
+We notice that both classical management strategies are faced with a tradeoff---optimal classical strategies accept a rather large fraction of episodes ending with a population crash, whose negative reward is counteracted with a higher economic output.\footnote{
+In our mathematical formulation of the decision problem, we have assumed for simplicity that the fishing effort cost is zero and that fish price is stable over time.
+This way, we equate economic output with harvested biomass.
+}
+Conversely, more conservative suboptimal classical strategies (e.g. with a higher constant escapement value) lead to lower total reward on average.
 The DRL approach side-steps this tradeoff by optimizing over a more complex family of possible policies—policies parametrized by a neural network, as opposed to policies labeled by a single parameter (the mortality rate).

### 2023-06-14 manuscript/manuscript.Rmd
intro

@@
 We strengthen this result with a stability investigation: we show that random perturbations in the model's parameter values used do not significantly vary the conclusion that DRL outperforms constant escapement.
-
-We show that, somewhat surprisingly, the DRL policy closely resembles a *bang-bang* solution (see Text Box 1).
-Bang-bang solutions are known to be optimal for a variety of single-species models, however, there is no a-priori reason why they should be optimal for higher-dimensional systems.
-This way, the DRL agent discovers and exploits, as it were, an effective simplification of the control problem.
-We provide some intuition as to how this simplification arises from the dynamics of the system, but leave a more rigorous characterization of this phenomenon for future work.
 

### 2023-06-14 manuscript/manuscript.Rmd
changes in math model

@@
 Thus, the term $M_X X_t$ is the total $X$ harvest at time $t$.
-We will focus on systems where a single species, $X$, is harvested, i.e. $M_Y = M_Z = \dots = 0$.
+This formulation of stock recruitment as a discrete finite difference process is common among fisheries, as opposed to continuous time formulations which involve instantaneous growth rates.
+This growth rate simplifies e.g.
+the possibly seasonal nature of reproduction (which would need to be accounted for in a realistic continuous-time model) by simply considering the total recruitment experienced by the population over a full year.
 
@@
 #| That is, these are the zeroes of the function $f_{\\text{May}}$. 
-#| Stable fixed points (also known as *attractors*) are plotted using a solid line, while the unstable fixed point is shown as a dotted line."
+#| Stable fixed points (also known as attractors) are plotted using a solid line, while the unstable fixed point is shown as a dotted line."
 #| out.width: "6.5in"
@@
 
-Consider the following three species generalization of eq. \eqref{eq:may}:
-
+Consider the following three species recruitment model which generalizes eq. \eqref{eq:may}:
 \begin{align}
@@
 \begin{split}
-  \Delta X_t &= L(X_t;\ r_X, K_X) - F(X_t, Z_t;\ \beta, c) - c_{XY}X_tY_t - h_t,\\
+  \Delta X_t &= L(X_t;\ r_X, K_X) - F(X_t, Z_t;\ \beta, c) - c_{XY}X_tY_t,\\
   \Delta Y_t &= L(Y_t;\ r_Y, K_Y) - D F(Y_t, Z_t;\ \beta, c) - c_{XY}X_tY_t,\\
@@
 \end{align}
-The model is conceptualized in Fig. \ref{fig:conceptual}, panel B.
 It contains three populations, the state at time $t$ is $(X_t, Y_t, Z_t)$. 

### 2023-06-14 manuscript/manuscript.Rmd
no more 3D stable points, added r_X(t) part

@@
 As can be expected, the dynamics of eq. \eqref{eq:3d model} are significantly more complex than those of eq. \eqref{eq:may}.
-There are a variety of fixed points for the equations---around 8 or 9 for wide ranges of parameter values.
-Fig. \ref{fig:fixed points} provides a rough idea of the fixed point structure by plotting the bifurcation diagram obtained from varying $D$ and fixing the other parameters.
-
-\begin{figure}
-  \centering
-  \begin{subfigure}[b]{0.32\textwidth}
-    \centering
-    \includegraphics[width=\textwidth]{figures/stX.png}
-    \caption{$X$ component of fixed points.}
-    \label{fig:stx}
-  \end{subfigure}
-  \hfill
-  \begin{subfigure}[b]{0.32\textwidth}
-    \centering
-    \includegraphics[width=\textwidth]{figures/stY.png}
-    \caption{$Y$ component of fixed points.}
-    \label{fig:sty}
-  \end{subfigure}
-  \hfill
-  \begin{subfigure}[b]{0.32\textwidth}
-    \centering
-    \includegraphics[width=\textwidth]{figures/stZ.png}
-    \caption{$Z$ component of fixed points.}
-    \label{fig:stz}
-  \end{subfigure}
-  \caption{Numerically-computed fixed points of the three-species deterministic model, eq. \eqref{eq:3d model}.
-  Here, the parameter $D$ was varied between 0.75 and 1.25, and the values of the other parameters are fixed as in eq. \eqref{eq:param vals}. 
-  The three components of the fixed points are plotted separately, as a function of $D$.}
-  \label{fig:fixed points}
-\end{figure}
-
-Throughout most of this paper we will fix the parameters of the model according to eq. \eqref{eq:param vals}.
-At these values, the fixed points of the model are:\footnote{
-We have included only three decimal digits for readability.
-These fixed points may be generated by running the ‘‘stable\_points.py’’ script in \url{https://github.com/boettiger-lab/approx-model-or-approx-soln}.
-}
-\begin{align*}
-  (X,Y,Z) \approx 
-  &(0.839, 0.054, 0.377), \quad 
-    (0.048, 0.774, 0.480), \quad
-    (0.359, 0.491, 0.509), \\
-    &(0.899, 0.000, 0.303), \quad
-    (0.000, 0.818, 0.457), \quad
-    (0.666, 0.666, 0.000), \\
-    &(1.000,0.000,0.000),\quad (0.000,1.000,0.000),\quad (0.000,0.000,0.000).
-\end{align*}
-The classification of fixed points for multi-dimensional systems is generally more complicated than in one dimension.
-A fixed point might be stable against perturbations in one direction, but unstable against perturbations along another.
-While a detailed classification of these fixed points is beyond the scope of this work, we point out that the point $(0.839, 0.054, 0.377)$ is stable against perturbations in all directions.
-We will thus use it as a starting point for our simulations.
+Here, a fixed point analysis provides a less intuitive picture of the dynamics---it no longer necessarily alternates in a tipping manner between two stable points.
+
+For most of our paper, parameters used in this model will be given as in eq. \eqref{eq:param vals}.
+In this paper we do not set out to provide a realistic model of any given fishery, but rather to show that the DRL approach may be applicable to a broad variety of plausible fish recruitment models.
+Because of this, we also provide a *stability analysis*, where we perturb the parameter values randomly.
+Finally, in the time-varying model, we use the following drift on the growth rate of $X$:
+\begin{equation*}
+  r_X = r_X(t) = \max(1-t/200, 1/2).
+\end{equation*}
 

### 2023-06-14 manuscript/manuscript.Rmd
streamlined 3D math a bit

@@
 \begin{split}
-  \Delta X_t &= L(X_t;\ r_X, K_X) - F(X_t, Z_t;\ \beta, c) - c_{XY}X_tY_t,\\
-  \Delta Y_t &= L(Y_t;\ r_Y, K_Y) - D F(Y_t, Z_t;\ \beta, c) - c_{XY}X_tY_t,\\
+  \Delta X_t &= L(X_t;\ r_X, K_X) - F(X_t, Z_t;\ \beta, c) - c_{XY}X_tY_t - h_{X,t},\\
+  \Delta Y_t &= L(Y_t;\ r_Y, K_Y) - D F(Y_t, Z_t;\ \beta, c) - c_{XY}X_tY_t - h_{Y,t},\\
   \Delta Z_t &= \left(f(X_t + DY_t\right) - d_Z)Z_t.
@@
 
+As mentioned in the introduction, we consider two harvesting scenarios: one where only $X$ is harvested (i.e. $h_{Y,t}=0$), and one where both $X$ and $Y$ are harvested.
+For most of our paper, parameters used in this model will be given as in eq. \eqref{eq:param vals}.
+In the time-varying model, however, we use the following drift on the growth rate of $X$:
+\begin{equation*}
+  r_X = r_X(t) = \max(1-t/200, 1/2).
+\end{equation*}
+
 This model generalizes eq. \eqref{eq:may} in the following sense:
@@
 \begin{align*}
-  \Delta X_t &= L(X_t;\ r_X, K_X) - F(X_t, H;\ \beta, c) - h_t,\\
+  \Delta X_t &= L(X_t;\ r_X, K_X) - F(X_t, H;\ \beta, c),\\
   \Delta Y_t &= L(Y_t;\ r_Y, K_Y) - D F(Y_t, H;\ \beta, c),\\
@@
 Here, a fixed point analysis provides a less intuitive picture of the dynamics---it no longer necessarily alternates in a tipping manner between two stable points.
-
-For most of our paper, parameters used in this model will be given as in eq. \eqref{eq:param vals}.
-In this paper we do not set out to provide a realistic model of any given fishery, but rather to show that the DRL approach may be applicable to a broad variety of plausible fish recruitment models.
-Because of this, we also provide a *stability analysis*, where we perturb the parameter values randomly.
-Finally, in the time-varying model, we use the following drift on the growth rate of $X$:
-\begin{equation*}
-  r_X = r_X(t) = \max(1-t/200, 1/2).
-\end{equation*}
 

### 2023-06-14 manuscript/manuscript.Rmd
RL discussion update

@@
 
-For our use case, the environment will be a computational model of the population dynamics of a fishery, with the environment state at time $t$ being a vector of all the fish populations, $N_t=(X_t,Y_t,Z_t)$.
-At each time step, the agent observes the state $S$ and harvests $X$,
-\begin{align*}
-  (X_t, Y_t, Z_t) \mapsto ((1-M_X)\cdot X_t, Y_t, Z_t),
-\end{align*}
-where the fish mortality-from-harvest $M_X=M_X(N_t)$ is the agent's action at time $t$. 
-This secures a reward of $M_X(N_t)X$.
-Afterwards, the environment undergoes a time step under its natural dynamics given by \eqref{eq:3d model stoch}.
+For our use case, the environment will be a computational model of the population dynamics of a fishery, with the environment state at time $t$ being a vector of all the fish populations, e.g.
+$N_t=(X_t,Y_t,Z_t)$ in the three species case.
+At each time step, the agent observes the state $S$ and enacts some harvest---reducing $X_t$ to $X_t - M_X(N_t)\cdot X_t$, and, in the two-fishery scenario, also reducing $Y_t$ to $Y_t - M_Y(N_t)\cdot Y_t$.
+Here the fish mortality-rates-from-harvest (i.e. $M_X=M_X(N_t)$ and $M_Y=M_Y(N_t)$), are the agent's action at time $t$. 
+This secures a reward of $M_X(N_t)X$ in the single fishery case, and in the two-fishery case the reward is $M_X(N_t)X + M_Y(N_t)Y$.
+After this harvest portion of the time step, the environment changes according to its natural dynamics given by \eqref{eq:3d model stoch}.
+
+As mentioned, this discretized time framework allows a simplification of the possibly seasonal mating behavioral patterns of the species involved. 
+This approximation is commonly used in fisheries for species with annual reproductive cycles (see e.g. [@mangel2006theoretical], Chap. 6).
 

### 2023-06-14 manuscript/manuscript.Rmd
some curriculum learning comments, a tiny bit on methods

@@
 Policies are, in general, functions from state space to policy space.
-In our case, these are real-valued functions on a cube $\pi:[0,2]^{\times 3}\to \mathbb{R}_+$.
+In our paper, these are $\pi:[0,1]^{\times 3}\to \mathbb{R}_+$ for the single fishery case, and $\pi:[0,1]^{\times 3}\to \mathbb{R}_+^2$ for two fisheries.
 The space of all such functions is highly singular, spanning a \emph{non-separable Hilbert space}.
-Even restricting ourselves to continuous policy functions, we end up with a set of policies which span the infinite dimensional space $L^2([0,2]^{\times3})$.
+Even restricting ourselves to continuous policy functions, we end up with a set of policies which span the infinite dimensional space $L^2([0,1]^{\times3})$.
 One way to avoid optimizing over an infinite dimensional ambient space is to discretize state space into a set of bins.
@@
 
+Model-free DRL provides, moreover, a framework within which agents can be trained to be generally competent over a *variety* of different models.
+This could more faithfully capture the ubiquitous uncertainty around ecosystem models.
+The aforementioned framework---known as *curriculum learning*---is considerably more intensive on computational resources than the ‘‘vanilla’’ DRL methods we have used in this paper.\footnote{
+All our agents were trained in a local server with two commercial GPUs. 
+The training time was between 30 minutes and one hour in each case.
+}
+Due to the increased computational requirements of this framework, we have left the exploration in this direction for future work.
+
 # Methods
@@
 We trained a DRL agent parametrized by a neural network $\theta$ with two hidden, 64-neuron, layers.
-The state space used was a cube $[0,2]^{\times 3}$, while the policies explored mapped states $N_t$ to a single number, $M_X(N_t) = \pi_\theta(N_t)$, the harvest mortality on species $X$ at that timestep.
+The state space used was a cube $[0,1]^{\times 3}$, while the policies explored mapped states $N_t$ to a single number, $M_X(N_t) = \pi_\theta(N_t)$, the harvest mortality on species $X$ at that timestep.
 Each episode in the training and in the evaluation stages was at most 200 time steps long.
@@
 First, at each time step the harvest $h_t$ was received as a reward.
-Second, if the episode ended early (say, at time-step $t$) because of a near-extinction event, a negative reward of $-50/t$ was awarded.
-The constant 50 was chosen so as to select against highly extractive policies in the early stages of training, and thus guide the agent towards sustainable policies.
+Second, if the episode ended early (say, at time-step $t$) because of a near-extinction event, a negative reward of $-100/t$ was awarded.
+The constant 100 was chosen so as to select against highly extractive policies in the early stages of training, and thus guide the agent towards sustainable policies.
 This technique, falling under the ‘‘umbrella’’ category of *reward shaping*, is commonly used -- alongside hyper-parameter tuning and state space normalization -- to improve the speed of training by avoiding badly-performing local optima in the gradient ascent algorithm.
 
-Thus, we normalized state space (to be contained in a cube of length 2) and engaged in reward shaping in order to optimize our agent's learning.
+Thus, we normalized state space (to be contained in a cube of length 1) and engaged in reward shaping in order to optimize our agent's learning.
 We did not need to tune hyper-parameters for this goal, however.
@@
 We used the Ray framework\footnote{\url{https://docs.ray.io/}} for training, in particular we used a Ray PPOConfig object to build the algorithm using the default values for all hyper-parameters.
-The agent was trained for 200 iterations of the PPO optimization step.
+The agent was trained for 300 iterations of the PPO optimization step.
 
 \textbf{Tuning the constant mortality strategy.}
-A grid of 100 mortality rates was laid in the interval $[0, 0.1]$.
+A grid of 100 mortality rates was laid in the interval $[0, 0.5]$.
 For each one of these rates, say $M$, we simulated 50 episodes based on \eqref{eq:3d model stoch}:

### 2023-06-14 manuscript/manuscript.Rmd
methods restructured a bit

@@
 
-\textbf{Training a DRL agent.}
-We trained a DRL agent parametrized by a neural network $\theta$ with two hidden, 64-neuron, layers.
-The state space used was a cube $[0,1]^{\times 3}$, while the policies explored mapped states $N_t$ to a single number, $M_X(N_t) = \pi_\theta(N_t)$, the harvest mortality on species $X$ at that timestep.
+\textbf{The Environment.}
+As seen in Fig.
+**TBD CONCEPTUAL**, we consider four different models of the ecosystem with which the fishery interacts, each of this is called an *environment*, alluding to the environment in an RL setting.
+To recap, these four environments are a single-species growth model \eqref{eq:may77} for a harvested species; a three-species model \eqref{eq:3d} with a single harvested species; the same three-species model but with *two* harvested species; and, finally, a three-species model with a time-varying parameter and two harvested species.
+
+The policies explored were functions from states $N_t$ to a single number, $M_X(N_t) = \pi_\theta(N_t)$, in the single fishery case.
+In the two-fishery case, policies lead to two-dimensional actions: $(M_X(N_t), M_Y(N_t)) = \pi_\theta(N_t)$.
 Each episode in the training and in the evaluation stages was at most 200 time steps long.
@@
 Second, if the episode ended early (say, at time-step $t$) because of a near-extinction event, a negative reward of $-100/t$ was awarded.
-The constant 100 was chosen so as to select against highly extractive policies in the early stages of training, and thus guide the agent towards sustainable policies.
-This technique, falling under the ‘‘umbrella’’ category of *reward shaping*, is commonly used -- alongside hyper-parameter tuning and state space normalization -- to improve the speed of training by avoiding badly-performing local optima in the gradient ascent algorithm.
-
-Thus, we normalized state space (to be contained in a cube of length 1) and engaged in reward shaping in order to optimize our agent's learning.
-We did not need to tune hyper-parameters for this goal, however.
-
+This reward function balances the extractive motivation of the fishery with conservation goals that go beyond the scope of long-term sustainable harvests.
+
+\textbf{Training a DRL agent.}
+We trained a DRL agent parametrized by a neural network $\theta$ with two hidden, 64-neuron, layers, on a local server with 2 commercial GPUs.
 We used the Ray framework\footnote{\url{https://docs.ray.io/}} for training, in particular we used a Ray PPOConfig object to build the algorithm using the default values for all hyper-parameters.
+In particular, no hyperparameter tuning was performed.
 The agent was trained for 300 iterations of the PPO optimization step.
+The total training time was on the order of 30 minutes to 1 hour.
+
+The state space used was normalized case-by-case as follows: 
+*1)* for the single species model, a line segment $[0,1]$, *2)* for all three-species models, a cute $[0,1]^{\times 3}$.
+We used simulation data to derive a bound on the population sizes typically observed, and thus be able to normalize states to a finite volume.
 

### 2023-06-15 manuscript/manuscript.Rmd
threshold comment

@@
 \end{align}
+For a realistic fishery model, this threshold value could represent a population value beyond which drastic ecological consequences may ensue, e.g. a tipping point.
+In our setting we have chosen $X_{\text{thresh.}}=Y_{\text{thresh.}}=Z_{\text{thresh.}}=0.05$ as a rule of thumb---given that under natural (unharvested) dynamics the populations range within values of 0.5 to 1, this would represent on the order of a 90% population decrease from its ‘‘natural’’ level.
+
 The reward function had two componenets:

### 2023-06-15 data/results_data/get_avg_performances.py
added performance summaries

@@ -1,0 +1,30 @@
+import pandas as pd
+import os
+
+fname_list = [os.path.join(*x) for x in 
+  [
+    ["ONESP", "high_beta", "comparison_100.csv.xz"],
+    ["1FISHERY", "DEFAULT", "comparison_300.csv.xz"],
+    ["2FISHERY", "DEFAULT", "comparison_300.csv.xz"],
+    ["2FISHERY", "RXDRIFT", "comparison_300.csv.xz"],
+  ]
+]
+
+print(fname_list)
+
+for fname in fname_list:
+  df = pd.read_csv(fname)
+  print(
+    "\n###########################################\n", 
+    fname, 
+    "\n###########################################\n",
+  )
+  for strategy in ['CMort', 'CEsc', 'PPO', 'PPO+GP']:
+    print(f"strategy = {strategy}")
+    df_strat = df.loc[df.strategy == strategy]
+    avg = df_strat.agg({'reward':'mean'})
+    df_maxt = df_strat.loc[df.t == 199]
+    frac = len(df_maxt.index)/len(df_strat)
+    print(avg)
+    print("fraction max t=199:", frac, "\n")
+  

### 2023-06-15 src/envs/growth_functions.py

@@ -8,7 +8,7 @@ import numpy as np
 def single_species_growth(pop, parameters):
   X = pop[0]
   p = parameters
-  pop[0] += p["r"] * X * (1 - X / p["K"]) - p["betaZ"] * X / (X**2 + p["v0"]**2)
+  pop[0] += p["r"] * X * (1 - X / p["K"]) - p["betaZ"] * X**2 / (X**2 + p["v0"]**2) + p["sigma"] * np.random.normal() * X
   pop = pop.astype(np.float32)
   return pop
 

### 2023-06-15 src/parameters.py

@@ -60,6 +60,7 @@ class parameters_oneSp(parameters):
     "K": np.float32(1.0),
     "betaZ": np.float32(0.1),
     "v0": np.float32(0.3),
+    "sigma": np.float32(0.1),
     }
     self.reset_params = self.params
   

### 2023-06-15 manuscript/manuscript.Rmd
threshold

@@
 For a realistic fishery model, this threshold value could represent a population value beyond which drastic ecological consequences may ensue, e.g. a tipping point.
-In our setting we have chosen $X_{\text{thresh.}}=Y_{\text{thresh.}}=Z_{\text{thresh.}}=0.05$ as a rule of thumb---given that under natural (unharvested) dynamics the populations range within values of 0.5 to 1, this would represent on the order of a 90% population decrease from its ‘‘natural’’ level.
+In our setting we have chosen $X_{\text{thresh.}}=Y_{\text{thresh.}}=Z_{\text{thresh.}}=0.05$ as a rule of thumb---given that under natural (unharvested) dynamics the populations range within values of 0.5 to 1, this would represent on the order of a 90-95% population decrease from its ‘‘natural’’ level.
+This roughly represents a near-extinction event.
 

### 2023-06-15 manuscript/manuscript.Rmd
a bit on constant mortality tuning section

@@
 **TBD CONCEPTUAL**, we consider four different models of the ecosystem with which the fishery interacts, each of this is called an *environment*, alluding to the environment in an RL setting.
-To recap, these four environments are a single-species growth model \eqref{eq:may77} for a harvested species; a three-species model \eqref{eq:3d} with a single harvested species; the same three-species model but with *two* harvested species; and, finally, a three-species model with a time-varying parameter and two harvested species.
+To recap, these four environments are a single-species growth model \eqref{eq:may77} for a harvested species; a three-species model \eqref{eq:3d model stoch} with a single harvested species; the same three-species model but with *two* harvested species; and, finally, a three-species model with a time-varying parameter and two harvested species.
 
@@
 \textbf{Training a DRL agent.}
-We trained a DRL agent parametrized by a neural network $\theta$ with two hidden, 64-neuron, layers, on a local server with 2 commercial GPUs.
+We trained a DRL agent parametrized by a neural network with two hidden, 64-neuron, layers, on a local server with 2 commercial GPUs.
 We used the Ray framework\footnote{\url{https://docs.ray.io/}} for training, in particular we used a Ray PPOConfig object to build the algorithm using the default values for all hyper-parameters.
@@
 \textbf{Tuning the constant mortality strategy.}
-A grid of 100 mortality rates was laid in the interval $[0, 0.5]$.
-For each one of these rates, say $M$, we simulated 50 episodes based on \eqref{eq:3d model stoch}:
-At each time step the state $(X_t, Y_t, Z_t)$ was observed, a harvest of $MX_t$ was collected and then the system evolved according to its natural dynamics \eqref{eq:3d model stoch}.
-The optimal mortality rate was the value $M^*$ for which the mean episode reward was maximal.
-We additionally evaluated a more conservative strategy where a mortality of $0.8 M^*$ instead is used.
-
+A grid of 101 mortality rates was laid in the interval $[0, 0.5]$ as possible mortality values in the single-fishery cases.
+Due to the high memory usage in the two-fishery case, a slightly coarser $51\times51$ grid was set in the square $[0,0.5]^2$ instead.
+Since the approach for tuning is completely analogous for all scenarios considered, here we focus on the ‘‘3 species, 1 fishery’’ case.
+For each one of these choices of a constant mortality rate, say $M_X$, we simulated 100 episodes based on \eqref{eq:3d model stoch}:
+At each time step the state $(X_t, Y_t, Z_t)$ was observed, a harvest of $M_X X_t$ was collected and then the system evolved according to its natural dynamics \eqref{eq:3d model stoch}.
+The optimal mortality rate was the value $M_X^*$ for which the mean episode reward was maximal.
 The result of the tuning procedure is shown in Fig. \ref{fig:tuning}.
-There, we see that the optimal value $M^*$ lies in an area with high episode reward variance.
-We will see in Fig. \ref{fig:rew_len_fig} that a likely candidate for this variance is the presence of episodes with early endings.
-Thus, the value $0.8 M^*$ supports a smaller variance, and correspondingly -- as Fig. \ref{fig:rew_len_fig} shows -- this policy leads to a much smaller probability of a near-extinction event.
 

### 2023-06-15 manuscript/manuscript.Rmd
added gaussian process stuff, started constant escapement, added stability analysis

@@
 For most of our paper, parameters used in this model will be given as in eq. \eqref{eq:param vals}.
-In the time-varying model, however, we use the following drift on the growth rate of $X$:
-\begin{equation*}
+In the time-varying model, moreover, we use the following drift on the growth rate of $X$:
+\begin{equation}
+  \label{eq:rx(t)}
   r_X = r_X(t) = \max(1-t/200, 1/2).
-\end{equation*}
+\end{equation}
 
@@
 
+Policies obtained from the PPO algorithm can be ‘‘noisy’’ as their optimization algorithm is randomized (see, e.g. App. \textbf{TBD} for visualizations of these policies).
+We perform some interpolation using a Gaussian process regressor in order to de-noise these policies.
+Specifically, we use the *GaussianProcessRegressor* object of the *sklearn* Python library with a kernel given by
+\begin{align*}
+  \text{RBF}(\text{length scale = 10}) + \text{WhiteNosie}(\text{noise level = 0.1}).
+\end{align*}
+The length scale and noise level values of this kernel were chosen arbitrarily---no hyperparameter tuning was needed to produce satisfactory interpolation, as will be shown in the results section.
+
 \textbf{Tuning the constant mortality strategy.}
@@
 
+\textbf{Tuning the constant escapement strategy.}
+This tuning procedure was analogous to that of the constant mortality strategy just summarized.
+Namely, \textbf{TBD}.
+
 \textbf{Parameter values used.}
-The model's dynamic parameters were chosen as follows
+The single-species model's (eq. \eqref{eq:may77}) dynamic parameters were chosen as
+\begin{align}
+\label{eq:1d param vals}
+r = K = 1, \beta = 0.25, c = 0.1.
+\end{align}
+Here, the values of $\beta$ and $c$ are chosen as to make the system be roughly close to its tipping point.
+
+The three-species model's (eq. \eqref{eq:3d model stoch}) dynamic parameters were chosen as follows
 \begin{align}
@@
 \begin{split}
-  r_X = K_X = r_Y = K_Y = 1, \quad \beta = 0.3, \quad c = 0.1\\
-  c_{XY} = 0.5,\quad f = 0.15,\quad D = 1.1, \quad d_Z = 0.135.
+  r_X = K_X = r_Y = K_Y = 1, \quad \beta = 0.3, \quad c = 0.3\\
+  c_{XY} = 0.1,\quad f = 0.1,\quad D = 1.1, \quad d_Z = 0.1.
 \end{split}
@@
 $$
-The threshold values used to determine an early episode end were chosen as
-$$
-  X_{\text{thresh.}} = 0.08,\quad Y_{\text{thresh.}} = Z_{\text{thresh.}} = 10^{-4},
-$$
-where the higher value of $X_{\text{thresh.}}$ helped select for more sustainable fishing strategies.
+In the case with a time-varying $r_X$ parameter, we used $r_X(t)$ given as in eq. \eqref{eq:rx(t)} and all other parameters as given in eq. \eqref{eq:param vals}.
+
+The values of $c$ and $\beta$ in the three-species model are slightly different than their values in the one-species model.
+These values were chosen heuristically: 
+We observed that choosing the lower value of $c=0.1$ in this case would lead to quick near-extinction events even without a harvest.
+Moreover, $\beta$ was slightly increased simply to put more predation pressure on the $X$ and $Y$ populations and make them slightly more fragile.
+
+\textbf{Stability analysis.}
+To ensure that our results do not strongly depend on our parameter choices, we performed a stability analysis.
+Here, we perturbed parameters randomly and measured the difference in performance between our DRL-based methods and the constant escapement strategy.
+We observed that the difference in performance is maintained for even relatively high noise strengths.
+For simplicity we only perform stability analysis on the ‘‘3 species, 2 fisheries and a time varying parameter’’ case.
+
+More specifically, for each parametric noise strength $\sigma_{\text{param.}}$ choice in $[0.04, 0.08, \dots, 0.2]$, we executed the following procedure:
+Sample 100 choices of perturbed parameter values, where the perturbation is as follows---each parameter $P$ is perturbed to $(1+g_P)P$ where $g_P$ is a Gaussian random variable with variance $\sigma_{\text{param.}}^2$.
+For each of these parameter choices, tune the constant escapement strategy and train the DRL model on the corresponding model.
+Measure the average reward difference between strategies for each of these parameter choice samples.
+Average this difference over the 100 samples.
+The stability results are displayed in Fig. \textbf{TBD}.
 

### 2023-06-15 src/workflow.py
added a bit of guidance in workflow.py

@@ -24,16 +24,62 @@ from plotnine import (
 )
 import ray
 
-'''
-Data-code list:
+"""
+Script used to produce and evaluate policies on a three-species system.
+Global variables (in all caps at the top of the code) can be used to set the
+environment we desire for the control problem in broad strokes:
+  
+  - ENVCODE ( = '1FISHERY' or '2FISHERY') is used to encode how many fisheries
+      our problem has. 1FISHERY defaults to population X being harvested, while
+      2FISHERY defaults to populations X and Y being harvested.
+  - DATACODE ( = to any option displayed below in the Data-code list) is used
+      to choose which growth function is used as the system's 'natural' (i.e. unharvested) 
+      dynamics
+  - NAME_MODIFIER: optionally may be set to modify the subdirectory in which data and
+      plots are saved
+  - ITERATIONS: number of iterations over which the DRL agent is trained
+  - REPS: number of replicate samples used to optimize and evaluate constant escapement
+      and constant mortality strategies. Also number of replicate samples used to
+      evaluate DRL policies.
+  - ESC_GRID_SIZE:
+      if ENVCODE == 1FISHERY: number of choices of constant escapement values 
+        on [0, 1] used to evaluate and optimize the strategy. (Resp., number
+        of choices of const. mortality values on [0, 0.5] used to evaluate
+        and optimize the strategy.)
+      if ENVCODE == 2FISHERY: similar to above, but now a 2D grid of 
+        ESC_GRID_SIZE x ESC_GRID_SIZE points is laid on [0, 1]^2 (resp.
+        [0, 0.5]^2). These are the possible choices of X-escapement and Y-escapement
+        (resp. X-mortality and Y-mortality) used to optimize and evaluate the
+        cosntant escapement (resp. constant mortality) strategy.
+  - DATAPATH: not to be changed. It is the directory where data and plots are saved.
+        
+"""
 
-"DEFAULT","RXDRIFT","V0DRIFT","DDRIFT","KLIMIT",
-"KLIMIT_RXDRIFT","BETADRIFT","CVDRIFT","YABIOTIC",
-"ZABIOTIC","COUPLFLUC",
+'''
+Data-code list. These codes are searchable in the envs/growth_fns.py document in 
+order to find the source code for the growth function.
+
[…]

### 2023-06-15 src/train_fns.py
moved old code out of the way

@@ -4,7 +4,6 @@ from envs import two_three_fishing
 from envs import growth_functions
 from parameters import parameters
 from util import dict_pretty_print
-import callback_fn
 from ray.rllib.algorithms import ppo
 from ray.rllib.agents.ppo import PPOTrainer
 from ray.tune import register_env

### 2023-06-15 src/workflow.py
more guidance

@@ -59,23 +59,23 @@ environment we desire for the control problem in broad strokes:
 Data-code list. These codes are searchable in the envs/growth_fns.py document in 
 order to find the source code for the growth function.
 
-"DEFAULT", Default 3-species model as presented in the manuscript (no time-varying 
-    parameters).
-"RXDRIFT", As DEFAULT, but with a time-varying r_X parameter
-"V0DRIFT", As DEFAULT, but with a time-varying v0 parameter (this parameter is 
-    called 'c' in manuscript, sorry)
-"DDRIFT", As DEFAULT, but with a time-varying D parameter
-"KLIMIT", Alternative non-time-varying model. Here DEFAULT's X-Y Lotka-Volterra competition 
-    term is replaced by reduction on X's and Y'x carrying capacity
-"KLIMIT_RXDRIFT", As KLIMIT, but with a time-varying r_X parameter
-"BETADRIFT", As DEFAULT, but with a time-varying beta parameter
-"CVDRIFT", As DEFAULT, but with a time-varying X-Y Lotka-Volterra competition parameter, 
-    c_V (this parameter is called 'c_XY' in manuscript, sorry). 
-"YABIOTIC", Similar to DEFAULT, but Y population is set to vary sinusoidally irrespective
-    of the X and Z values.
-"ZABIOTIC", Similar to YABIOTIC, but it is Z varying sinusoidally now.
-"COUPLFLUC", As DEFAULT but with sinusoidally varying v_0 parameter (this parameter
-    is called 'c' in manuscript, sorry)
+DEFAULT        - Default 3-species model as presented in the manuscript (no time-varying 
+                 parameters).
+RXDRIFT        - As DEFAULT, but with a time-varying r_X parameter
+V0DRIFT        - As DEFAULT, but with a time-varying v0 parameter (this parameter is 
+                 called 'c' in manuscript, sorry)
+DDRIFT         - As DEFAULT, but with a time-varying D parameter
+KLIMIT         - Alternative non-time-varying model. Here DEFAULT's X-Y Lotka-Volterra competition 
+                 term is replaced by reduction on X's and Y'x carrying capacity
+KLIMIT_RXDRIFT - As KLIMIT, but with a time-varying r_X parameter
+BETADRIFT      - As DEFAULT, but with a time-varying beta parameter
[…]

### 2023-06-15 src/workflow.py
mini

@@ -382,7 +382,7 @@ def workflow_oneFishery():
   gpp_max_t['strategy']='PPO+GP'
   esc_max_t['strategy']='CEsc'
   msy_max_t['strategy']='CMort'
-  tot_max_t = pd.concat([msy_max_t, esc_max_t, ppo_max_t, gpp_max_t])
+  tot_max_t = pd.concat([esc_max_t, msy_max_t, ppo_max_t, gpp_max_t])
   
   tot_max_t.to_csv(os.path.join(DATAPATH,f"comparison_{ITERATIONS}.csv.xz"))
   print("Done with data generation, finishing up on plots...")

### 2023-06-15 src/msy_fns.py
fractional_msy_script.py

@@ -3,6 +3,8 @@ import pandas as pd
 import numpy as np
 import itertools
 
+""" Simulate """
+
 @ray.remote
 def msy_max_t_1fish(env, mortality, repetitions):
   x = []
@@ -101,6 +103,8 @@ def generate_msy_episodes_1fish(env, grid_nr=101, repetitions=100, only_max_time
     return df_max_times
   return df_max_times, df
 
+""" optimize """
+
 def find_msy_1fish(env, grid_nr=101, repetitions=100):
   df_max_times, df = generate_msy_episodes_1fish(env, grid_nr=grid_nr, repetitions=repetitions)
   tmp = (
@@ -129,6 +133,29 @@ def find_msy_2fish(env, grid_nr=51, repetitions=100):
   		(df.mortality_y == best.mortality_y.values[0])
   	]
 	)
+	
+""" fraction-of-msy const mortality simulation """
+
+def frac_msy_1fish(env, msy, *, fraction=0.8, repetitions = 100):
+  return msy_max_t_1fish(env, fraction * msy, repetitions=repetitions)
+
+def frac_msy_2fish(env, msy_x, msy_y, *, fraction=0.8, repetitions = 100):
+  return msy_max_t_1fish(env, fraction * msy_x, fraction * msy_y, repetitions=repetitions)
+
+# in practice I'll have to access the files I already made with the saved data
+def csv_to_frac_msy_1fish(env, fname, fraction=0.8, repetitions = 100):
+  df = pd.read_csv(fname)
+  msy = df.mortality[0]
+  return frac_msy_1fish(env, msy, fraction=fraction, repetitions=repetitions)
+
+def csv_to_frac_msy_2fish(env, fname, fraction=0.8, repetitions = 100):
+  df = pd.read_csv(fname)
+  msy_x = df.mortality_x[0]
+  msy_y = df.mortality_y[0]
+  return frac_msy_2fish(env, msy_x, msy_y, fraction=fraction, repetitions=repetitions)
+  
+
+""" misc """
 
 def msy_performances_1fish(env, grid_nr = 101, repetitions = 100):
   df_max_times = generate_msy_episodes_1fish(
@@ -140,3 +167,5 @@ def msy_performances_1fish(env, grid_nr = 101, repetitions = 100):
     .groupby(['act'], as_index=False)
     .agg({'reward': 'mean', 'reward_std':'std'})
   )
+  
+

### 2023-06-15 src/fractional_msy_script.py
ok now i actually added fractional_msy

@@ -1,0 +1,16 @@
+import pandas as pd
+import seaborn as sns
+from plotnine import ggplot, aes, geom_bar, geom_point, geom_line
+import os
+
+from msy_fns import csv_to_frac_msy_1fish, csv_to_frac_msy_2fish
+
+#globals
+1sp1fish = ["data", "results_data", "ONESP", "high_beta", "msy_100.csv.xz"]
+3sp1fish = ["data", "results_data", "1FISHERY", "DEFAULT", "msy_100.csv.xz"]
+3sp2fish = ["data", "results_data", "2FISHERY", "DEFAULT", "msy_100.csv.xz"]
+3sp2fish_timevar = ["data", "results_data", "2FISHERY", "RXDRIFT", "msy_100.csv.xz"]
+
+fname = os.path.join(
+  *1sp1fish, 
+)

### 2023-06-15 data/results_data/2FISHERY/DEFAULT/approx_1D1D/params.py
added a bunch of new 1fishery and 2fishery data and plots

@@ -1,0 +1,18 @@
+parameters_obj.params = {
+    "r_x": np.float32(1.0),
+    "r_y": np.float32(1.0),
+    "K_x": np.float32(1.0),
+    "K_y": np.float32(1.0),
+    "beta": np.float32(0.2),
+    "v0":  np.float32(0.1),
+    "D": np.float32(1.1),
+    "tau_yx": np.float32(0.0),
+    "tau_xy": np.float32(0.0),
+    "cV": np.float32(0.05),
+    "f": np.float32(0.1),
+    "dH": np.float32(0.1),
+    "alpha": np.float32(0.05),
+    "sigma_x": np.float32(parameters_obj.sigma),
+    "sigma_y": np.float32(parameters_obj.sigma),
+    "sigma_z": np.float32(parameters_obj.sigma)
+    }

### 2023-06-15 src/fractional_msy_script.py
a bit more fractional msy script

@@ -3,6 +3,9 @@ import seaborn as sns
 from plotnine import ggplot, aes, geom_bar, geom_point, geom_line
 import os
 
+from envs import two_three_fishing, singleSp
+from train_fns import create_env
+from parameters import parameters, parameters_oneSp
 from msy_fns import csv_to_frac_msy_1fish, csv_to_frac_msy_2fish
 
 #globals
@@ -11,6 +14,19 @@ from msy_fns import csv_to_frac_msy_1fish, csv_to_frac_msy_2fish
 3sp2fish = ["data", "results_data", "2FISHERY", "DEFAULT", "msy_100.csv.xz"]
 3sp2fish_timevar = ["data", "results_data", "2FISHERY", "RXDRIFT", "msy_100.csv.xz"]
 
-fname = os.path.join(
-  *1sp1fish, 
+env_class = two_three_fishing
+
+parameter_obj_3sp = parameters()
+parameter_obj_1sp = parameters_oneSp()
+
+env = create_env(
+  env_class = , 
+  parameter_obj, 
+  datacode = "DEFAULT", 
+  fluctuating=True, 
+  training=True
 )
+
+fname = os.path.join(*1sp1fish)
+
+msy_frac_df = csv_to_frac_msy_1fish()

### 2023-06-15 src/envs/oneSpecies.py
added oneSpecies class

@@ -1,0 +1,91 @@
+import numpy as np
+import gymnasium as gym
+from gymnasium import spaces
+from parameters import parameters
+
+def dummy_growth(pop,parameters):
+  return pop * (1 - pop)
+
+_DEFAULT_INIT_POP = np.array([0.7], dtype=np.float32)
+
+def default_penalty(t):
+  return - 100/(t+1)
+
+class singleSp(gym.Env):
+  """ Single species model. """
+  def __init__(self, config=None):
+    self.config = config or {}
+    #
+    self.parameters = config.get("parameters", None)
+    if not self.parameters:
+      print("the config dict argument needs a parameters component!")
+    self.growth_fn = config.get("growth_fn", dummy_growth)
+    self.initial_pop = config.get("initial_pop", _DEFAULT_INIT_POP)
+    self.Tmax = config.get("Tmax", 200)
+    self.threshold = config.get("threshold", np.float32(5e-2))
+    self.init_sigma = config.get("init_sigma", np.float32(5e-3))
+    self.num_species = 1
+    #
+    self.bound = 2 * self.config.get("bound", 2)
+    self.fluctuating = self.config.get("fluctuating", False)
+    self.training = self.config.get("training", True)
+    self.early_end_penalty = self.config.get("early_end_penalty", default_penalty)
+    self.timestep = 0
+    #
+    self.action_space = spaces.Box(
+      np.array([0], dtype=np.float32),
+      np.array([1], dtype=np.float32),
+      dtype = np.float32
+    )
+    self.observation_space = spaces.Box(
+      np.array([0], dtype=np.float32),
+      np.array([1], dtype=np.float32),
+      dtype=np.float32,
+    )
+    self.reset(seed=config.get("seed", None))
+    #################################################################
+  
+  def reset(self, *, seed = None, options = None):
+    # "*" forces keyword args to be named on calls if values are provided
+    self.timestep = 0
+    self.state = self.update_state(self.initial_pop)
+    self.state += np.float32(self.init_sigma * np.random.normal(size=1) )
+    info = {}
+    return self.state, info
+  
+  def step(self, action):
+    action = np.clip(action, [0], [1])
+    # action *= 0.5
+    pop = self.population() # current state in natural units
+    
+    # harvest & recruitment
+    pop, reward = self.harvest(pop, action)
+    if self.fluctuating:
+      pop = self.growth_fn(pop, self.parameters, self.timestep)
+    else:
[…]

### 2023-06-15 src/fractional_msy_script.py
preliminary-finished frac_msy script

@@ -3,30 +3,49 @@ import seaborn as sns
 from plotnine import ggplot, aes, geom_bar, geom_point, geom_line
 import os
 
-from envs import two_three_fishing, singleSp
+from envs import two_three_fishing, oneSpecies
 from train_fns import create_env
 from parameters import parameters, parameters_oneSp
 from msy_fns import csv_to_frac_msy_1fish, csv_to_frac_msy_2fish
 
-#globals
-1sp1fish = ["data", "results_data", "ONESP", "high_beta", "msy_100.csv.xz"]
-3sp1fish = ["data", "results_data", "1FISHERY", "DEFAULT", "msy_100.csv.xz"]
-3sp2fish = ["data", "results_data", "2FISHERY", "DEFAULT", "msy_100.csv.xz"]
-3sp2fish_timevar = ["data", "results_data", "2FISHERY", "RXDRIFT", "msy_100.csv.xz"]
-
-env_class = two_three_fishing
+1sp1fish = ["data", "results_data", "ONESP", "high_beta"]
+3sp1fish = ["data", "results_data", "1FISHERY", "DEFAULT"]
+3sp2fish = ["data", "results_data", "2FISHERY", "DEFAULT"]
+3sp2fish_timevar = ["data", "results_data", "2FISHERY", "RXDRIFT"]
 
 parameter_obj_3sp = parameters()
 parameter_obj_1sp = parameters_oneSp()
 
+# choosing env to evaluate / corresponding data location
+scenario = 1sp1fish
+env_class = oneSpecies.singleSp
+fname = os.path.join(*scenario, "msy_100.csv.xz")
+DATACODE = "ONESP"
+
 env = create_env(
-  env_class = , 
+  env_class = env_class, 
   parameter_obj, 
-  datacode = "DEFAULT", 
-  fluctuating=True, 
+  datacode = DATACODE, 
+  fluctuating=False, 
   training=True
 )
 
-fname = os.path.join(*1sp1fish)
 
-msy_frac_df = csv_to_frac_msy_1fish()
+msy_frac_df, _ = csv_to_frac_msy_1fish(
+  env, fname, fraction=0.8, repetitions=100
+)
+del _
+
+msy_frac_df.to_csv(
+  os.path.join(*scenario, "frac_0-8_msy.csv.xz")
+)
+
+# plots
+
+rew_hist = (
[…]

### 2023-06-15 src/fractional_msy_script.py
mini

@@ -8,10 +8,10 @@ from train_fns import create_env
 from parameters import parameters, parameters_oneSp
 from msy_fns import csv_to_frac_msy_1fish, csv_to_frac_msy_2fish
 
-1sp1fish = ["data", "results_data", "ONESP", "high_beta"]
-3sp1fish = ["data", "results_data", "1FISHERY", "DEFAULT"]
-3sp2fish = ["data", "results_data", "2FISHERY", "DEFAULT"]
-3sp2fish_timevar = ["data", "results_data", "2FISHERY", "RXDRIFT"]
+_1sp1fish = ["data", "results_data", "ONESP", "high_beta"]
+_3sp1fish = ["data", "results_data", "1FISHERY", "DEFAULT"]
+_3sp2fish = ["data", "results_data", "2FISHERY", "DEFAULT"]
+_3sp2fish_timevar = ["data", "results_data", "2FISHERY", "RXDRIFT"]
 
 parameter_obj_3sp = parameters()
 parameter_obj_1sp = parameters_oneSp()

### 2023-06-15 src/fractional_msy_script.py
mini

@@ -17,7 +17,7 @@ parameter_obj_3sp = parameters()
 parameter_obj_1sp = parameters_oneSp()
 
 # choosing env to evaluate / corresponding data location
-scenario = 1sp1fish
+scenario = _1sp1fish
 env_class = oneSpecies.singleSp
 fname = os.path.join(*scenario, "msy_100.csv.xz")
 DATACODE = "ONESP"

### 2023-06-15 src/fractional_msy_script.py
mini

@@ -23,11 +23,11 @@ fname = os.path.join(*scenario, "msy_100.csv.xz")
 DATACODE = "ONESP"
 
 env = create_env(
-  env_class = env_class, 
+  env_class, 
   parameter_obj, 
   datacode = DATACODE, 
   fluctuating=False, 
-  training=True
+  training=True,
 )
 
 

### 2023-06-15 src/fractional_msy_script.py
mini

@@ -18,6 +18,7 @@ parameter_obj_1sp = parameters_oneSp()
 
 # choosing env to evaluate / corresponding data location
 scenario = _1sp1fish
+parameter_obj = parameter_obj_1sp
 env_class = oneSpecies.singleSp
 fname = os.path.join(*scenario, "msy_100.csv.xz")
 DATACODE = "ONESP"

### 2023-06-15 src/fractional_msy_script.py
mini

@@ -8,9 +8,9 @@ from train_fns import create_env
 from parameters import parameters, parameters_oneSp
 from msy_fns import csv_to_frac_msy_1fish, csv_to_frac_msy_2fish
 
-_1sp1fish = ["data", "results_data", "ONESP", "high_beta"]
-_3sp1fish = ["data", "results_data", "1FISHERY", "DEFAULT"]
-_3sp2fish = ["data", "results_data", "2FISHERY", "DEFAULT"]
+_1sp1fish = ["..", "data", "results_data", "ONESP", "high_beta"]
+_3sp1fish = ["..", "data", "results_data", "1FISHERY", "DEFAULT"]
+_3sp2fish = ["..", "data", "results_data", "2FISHERY", "DEFAULT"]
 _3sp2fish_timevar = ["data", "results_data", "2FISHERY", "RXDRIFT"]
 
 parameter_obj_3sp = parameters()

### 2023-06-15 src/msy_fns.py
mini

@@ -137,10 +137,10 @@ def find_msy_2fish(env, grid_nr=51, repetitions=100):
 """ fraction-of-msy const mortality simulation """
 
 def frac_msy_1fish(env, msy, *, fraction=0.8, repetitions = 100):
-  return msy_max_t_1fish(env, fraction * msy, repetitions=repetitions)
+  return ray.get(msy_max_t_1fish.remote(env, fraction * msy, repetitions=repetitions))
 
 def frac_msy_2fish(env, msy_x, msy_y, *, fraction=0.8, repetitions = 100):
-  return msy_max_t_1fish(env, fraction * msy_x, fraction * msy_y, repetitions=repetitions)
+  return ray.get(msy_max_t_1fish.remote(env, fraction * msy_x, fraction * msy_y, repetitions=repetitions))
 
 # in practice I'll have to access the files I already made with the saved data
 def csv_to_frac_msy_1fish(env, fname, fraction=0.8, repetitions = 100):

### 2023-06-15 src/fractional_msy_script.py
mini

@@ -37,6 +37,12 @@ msy_frac_df, _ = csv_to_frac_msy_1fish(
 )
 del _
 
+cols = ["t", "rep", "mortality", "act", "reward", "X", "Y", "Z"] # assume 1 species as default
+if env.num_species == 1:
+  cols = ["t", "rep", "mortality", "act", "reward", "X"]
+  
+msy_frac_df = pd.DataFrame(msy_frac_df, columns=cols)
+
 msy_frac_df.to_csv(
   os.path.join(*scenario, "frac_0-8_msy.csv.xz")
 )

### 2023-06-15 src/fractional_msy_script.py
mini (have been debugging fractional_msy_Script

@@ -53,6 +53,6 @@ rew_hist = (
   ggplot(data=msy_frac_df, mapping=aes(x='rep',weight='reward')) 
   +geom_bar()
 )
-msy_frac_df.save(
+rew_hist.save(
   os.path.join(*scenario, "frac_0-8_msy_rewards.png")
 )

### 2023-06-15 src/fractional_msy_script.py
a bit debugging, more needed

@@ -17,11 +17,11 @@ parameter_obj_3sp = parameters()
 parameter_obj_1sp = parameters_oneSp()
 
 # choosing env to evaluate / corresponding data location
-scenario = _1sp1fish
-parameter_obj = parameter_obj_1sp
-env_class = oneSpecies.singleSp
+scenario = _3sp2fish
+parameter_obj = parameter_obj_3sp
+env_class = two_three_fishing.twoThreeFishing #oneSpecies.singleSp
 fname = os.path.join(*scenario, "msy_100.csv.xz")
-DATACODE = "ONESP"
+DATACODE = "DEFAULT"
 
 env = create_env(
   env_class, 
@@ -32,14 +32,17 @@ env = create_env(
 )
 
 
-msy_frac_df, _ = csv_to_frac_msy_1fish(
-  env, fname, fraction=0.8, repetitions=100
+msy_frac_df, _ = csv_to_frac_msy_2fish(
+  env, fname, fraction=0.8, repetitions=100,
 )
 del _
 
-cols = ["t", "rep", "mortality", "act", "reward", "X", "Y", "Z"] # assume 1 species as default
-if env.num_species == 1:
+cols = ["t", "rep", "mortality_x", "mortality_y", "act_x", "act_y", "reward", "X", "Y", "Z"] # assume _3sp2fish by default
+if scenario == _1sp1fish:
   cols = ["t", "rep", "mortality", "act", "reward", "X"]
+if scenario == _3sp1fish:
+  cols = ["t", "rep", "mortality", "act", "reward", "X", "Y", "Z"]
+  
   
 msy_frac_df = pd.DataFrame(msy_frac_df, columns=cols)
 

### 2023-06-15 src/msy_fns.py

@@ -152,6 +152,7 @@ def csv_to_frac_msy_2fish(env, fname, fraction=0.8, repetitions = 100):
   df = pd.read_csv(fname)
   msy_x = df.mortality_x[0]
   msy_y = df.mortality_y[0]
+  print(msy_x, "\n\n")
   return frac_msy_2fish(env, msy_x, msy_y, fraction=fraction, repetitions=repetitions)
   
 

### 2023-06-16 src/msy_fns.py
msy_fns mini

@@ -152,7 +152,6 @@ def csv_to_frac_msy_2fish(env, fname, fraction=0.8, repetitions = 100):
   df = pd.read_csv(fname)
   msy_x = df.mortality_x[0]
   msy_y = df.mortality_y[0]
-  print(msy_x, "\n\n")
   return frac_msy_2fish(env, msy_x, msy_y, fraction=fraction, repetitions=repetitions)
   
 

### 2023-06-16 src/fractional_msy_script.py
fixed bugs on fractional msy script, added generated data on 3sp2fish constant parameters generated

@@ -31,31 +31,42 @@ env = create_env(
   training=True,
 )
 
-
-msy_frac_df, _ = csv_to_frac_msy_2fish(
-  env, fname, fraction=0.8, repetitions=100,
-)
-del _
-
-cols = ["t", "rep", "mortality_x", "mortality_y", "act_x", "act_y", "reward", "X", "Y", "Z"] # assume _3sp2fish by default
-if scenario == _1sp1fish:
-  cols = ["t", "rep", "mortality", "act", "reward", "X"]
-if scenario == _3sp1fish:
-  cols = ["t", "rep", "mortality", "act", "reward", "X", "Y", "Z"]
+for fraction in [0.8, 0.9, 0.95, 1]:
+  msy_frac_df, _ = csv_to_frac_msy_2fish(
+    env, fname, fraction=fraction, repetitions=100,
+  )
+  del _
   
+  cols = ["t", "rep", "mortality_x", "mortality_y", "act_x", "act_y", "reward", "X", "Y", "Z"] # assume _3sp2fish by default
+  if scenario == _1sp1fish:
+    cols = ["t", "rep", "mortality", "act", "reward", "X"]
+  if scenario == _3sp1fish:
+    cols = ["t", "rep", "mortality", "act", "reward", "X", "Y", "Z"]
+    
+    
+  msy_frac_df = pd.DataFrame(msy_frac_df, columns=cols)
+  msy_frac_df['reward_std'] = msy_frac_df['reward']
+  df_finished_eps = msy_frac_df[msy_frac_df.t == 200]
+  frac_max_t = len(df_finished_eps.index)/len(msy_frac_df.index)
   
-msy_frac_df = pd.DataFrame(msy_frac_df, columns=cols)
-
-msy_frac_df.to_csv(
-  os.path.join(*scenario, "frac_0-8_msy.csv.xz")
-)
-
-# plots
-
-rew_hist = (
-  ggplot(data=msy_frac_df, mapping=aes(x='rep',weight='reward')) 
-  +geom_bar()
-)
-rew_hist.save(
-  os.path.join(*scenario, "frac_0-8_msy_rewards.png")
-)
+  print(f"""fraction = {fraction}
+  
+  {msy_frac_df.agg({'reward':'mean', 'reward_std':'std'})}
+  
+  fraction ep len == 200: {frac_max_t}
+  
+  """)
+  
+  msy_frac_df.to_csv(
[…]

### 2023-06-16 src/msy_fns.py

@@ -136,11 +136,11 @@ def find_msy_2fish(env, grid_nr=51, repetitions=100):
 	
 """ fraction-of-msy const mortality simulation """
 
-def frac_msy_1fish(env, msy, *, fraction=0.8, repetitions = 100):
+def frac_msy_1fish(env, msy, fraction=0.8, repetitions = 100):
   return ray.get(msy_max_t_1fish.remote(env, fraction * msy, repetitions=repetitions))
 
-def frac_msy_2fish(env, msy_x, msy_y, *, fraction=0.8, repetitions = 100):
-  return ray.get(msy_max_t_1fish.remote(env, fraction * msy_x, fraction * msy_y, repetitions=repetitions))
+def frac_msy_2fish(env, msy_x, msy_y, fraction=0.8, repetitions = 100):
+  return ray.get(msy_max_t.remote(env, fraction * msy_x, fraction * msy_y, repetitions=repetitions))
 
 # in practice I'll have to access the files I already made with the saved data
 def csv_to_frac_msy_1fish(env, fname, fraction=0.8, repetitions = 100):

### 2023-06-16 src/fractional_msy_script.py
fixed a final minibug, generated fractional msy data for all 4 scenarios

@@ -11,28 +11,34 @@ from msy_fns import csv_to_frac_msy_1fish, csv_to_frac_msy_2fish
 _1sp1fish = ["..", "data", "results_data", "ONESP", "high_beta"]
 _3sp1fish = ["..", "data", "results_data", "1FISHERY", "DEFAULT"]
 _3sp2fish = ["..", "data", "results_data", "2FISHERY", "DEFAULT"]
-_3sp2fish_timevar = ["data", "results_data", "2FISHERY", "RXDRIFT"]
+_3sp2fish_timevar = ["..", "data", "results_data", "2FISHERY", "RXDRIFT"]
 
 parameter_obj_3sp = parameters()
 parameter_obj_1sp = parameters_oneSp()
 
+""" envs: two_three_fishing.twoThreeFishing two_three_fishing.oneThreeFishing oneSpecies.singleSp"""
+""" fns: csv_to_frac_msy_1fish csv_to_frac_msy_2fish """
+
 # choosing env to evaluate / corresponding data location
-scenario = _3sp2fish
+scenario = _3sp2fish_timevar
 parameter_obj = parameter_obj_3sp
-env_class = two_three_fishing.twoThreeFishing #oneSpecies.singleSp
+env_class = two_three_fishing.twoThreeFishing
+DATACODE = "RXDRIFT"
+FLUCTUATING = True
+function_to_use = csv_to_frac_msy_2fish
+
 fname = os.path.join(*scenario, "msy_100.csv.xz")
-DATACODE = "DEFAULT"
 
 env = create_env(
   env_class, 
   parameter_obj, 
   datacode = DATACODE, 
-  fluctuating=False, 
+  fluctuating=FLUCTUATING, 
   training=True,
 )
 
 for fraction in [0.8, 0.9, 0.95, 1]:
-  msy_frac_df, _ = csv_to_frac_msy_2fish(
+  msy_frac_df, _ = function_to_use(
     env, fname, fraction=fraction, repetitions=100,
   )
   del _

### 2023-06-16 manuscript/manuscript.Rmd
mini
improved the methods section a bit

@@
 **TBD CONCEPTUAL**, we consider four different models of the ecosystem with which the fishery interacts, each of this is called an *environment*, alluding to the environment in an RL setting.
-To recap, these four environments are a single-species growth model \eqref{eq:may77} for a harvested species; a three-species model \eqref{eq:3d model stoch} with a single harvested species; the same three-species model but with *two* harvested species; and, finally, a three-species model with a time-varying parameter and two harvested species.
-
-The policies explored were functions from states $N_t$ to a single number, $M_X(N_t) = \pi_\theta(N_t)$, in the single fishery case.
-In the two-fishery case, policies lead to two-dimensional actions: $(M_X(N_t), M_Y(N_t)) = \pi_\theta(N_t)$.
-Each episode in the training and in the evaluation stages was at most 200 time steps long.
-Episodes were ended early in case of a ‘‘near-extinction event,’’ a moment in which any of the populations dipped below a threshold value, i.e.,
+To recap, these four environments are a single-species growth model \eqref{eq:may} for a harvested species; a three-species model \eqref{eq:3d model stoch} with a single harvested species; the same three-species model but with *two* harvested species; and, finally, a three-species model with a time-varying parameter and two harvested species.
+
+The policies explored were functions from states $N_t$ to either a single number $M_X(N_t) = \pi_\theta(N_t)$, or a pair of numbers $(M_X(N_t), M_Y(N_t)) = \pi_\theta(N_t)$ depending on the environment considered.
+\textbf{Probably a table showing the different combinations of species and fishery numbers would be good, possibly giving these ‘scenarios’ names.}
+
+Our goal is to evaluate the performance of different policy strategies over a specified window of time.
+We have chosen this window to be 200 timesteps, where each discrete time-step represents the dynamical evolution of the system over a year.
+_Episodes_ begin at timestep zero and are controlled by the policy and the system's natural evolution until either the maximum timespan is reached, or until a ‘‘near-extinction event’’ occurs---that is, until any of the populations go below a given threshold,
 \begin{align}
@@
 \end{align}
-For a realistic fishery model, this threshold value could represent a population value beyond which drastic ecological consequences may ensue, e.g. a tipping point.
 In our setting we have chosen $X_{\text{thresh.}}=Y_{\text{thresh.}}=Z_{\text{thresh.}}=0.05$ as a rule of thumb---given that under natural (unharvested) dynamics the populations range within values of 0.5 to 1, this would represent on the order of a 90-95% population decrease from its ‘‘natural’’ level.
-This roughly represents a near-extinction event.
-
-The reward function had two componenets:
-First, at each time step the harvest $h_t$ was received as a reward.
-Second, if the episode ended early (say, at time-step $t$) because of a near-extinction event, a negative reward of $-100/t$ was awarded.
+
+The reward function defining our policy optimization problem has two componenets:
+The first is economical: the total biomass harvested over an episode.
+The second seeks to reflect conservation goals: if a near-extinction event occurs, the episode is ended early and a negative reward of $-100/t$ was awarded.
 This reward function balances the extractive motivation of the fishery with conservation goals that go beyond the scope of long-term sustainable harvests.
@@
 In particular, no hyperparameter tuning was performed.
-The agent was trained for 300 iterations of the PPO optimization step.
+The agent was trained for 300 iterations of the PPO optimization step for the three-species cases.
 The total training time was on the order of 30 minutes to 1 hour.
+For the single-species model, the training iterations were scaled down to 100 and the training time was around 10 minutes.
 

### 2023-06-16 manuscript/manuscript.Rmd
typos

@@
 
-The policies explored were functions from states $N_t$ to either a single number $M_X(N_t) = \pi_\theta(N_t)$, or a pair of numbers $(M_X(N_t), M_Y(N_t)) = \pi_\theta(N_t)$ depending on the environment considered.
+The policies explored were functions from states $N_t$ to either a single number $M_X(N_t) = \pi(N_t)$, or a pair of numbers $(M_X(N_t), M_Y(N_t)) = \pi(N_t)$ depending on the environment considered.
 \textbf{Probably a table showing the different combinations of species and fishery numbers would be good, possibly giving these ‘scenarios’ names.}
@@
 Our goal is to evaluate the performance of different policy strategies over a specified window of time.
-We have chosen this window to be 200 timesteps, where each discrete time-step represents the dynamical evolution of the system over a year.
-_Episodes_ begin at timestep zero and are controlled by the policy and the system's natural evolution until either the maximum timespan is reached, or until a ‘‘near-extinction event’’ occurs---that is, until any of the populations go below a given threshold,
+We have chosen this window to be 200 time-steps, where each discrete time-step represents the dynamical evolution of the system over a year.
+_Episodes_ begin at time-step zero and are controlled by the policy and the system's natural evolution until either the maximum time-span is reached, or until a ‘‘near-extinction event’’ occurs---that is, until any of the populations go below a given threshold,
 \begin{align}
@@
 
-The reward function defining our policy optimization problem has two componenets:
+The reward function defining our policy optimization problem has two components.
 The first is economical: the total biomass harvested over an episode.
-The second seeks to reflect conservation goals: if a near-extinction event occurs, the episode is ended early and a negative reward of $-100/t$ was awarded.
+The second seeks to reflect conservation goals: if a near-extinction event occurs at time $t$, the episode is ended early and a negative reward of $-100/t$ was awarded.
 This reward function balances the extractive motivation of the fishery with conservation goals that go beyond the scope of long-term sustainable harvests.
@@
 We trained a DRL agent parametrized by a neural network with two hidden, 64-neuron, layers, on a local server with 2 commercial GPUs.
-We used the Ray framework\footnote{\url{https://docs.ray.io/}} for training, in particular we used a Ray PPOConfig object to build the algorithm using the default values for all hyper-parameters.
+We used the Ray framework\footnote{\url{https://docs.ray.io/}} for training, specifically we used the Ray PPOConfig class to build the policy optimization algorithm using the default values for all hyper-parameters.
 In particular, no hyperparameter tuning was performed.
@@
 \textbf{Parameter values used.}
-The single-species model's (eq. \eqref{eq:may77}) dynamic parameters were chosen as
+The single-species model's (eq. \eqref{eq:may}) dynamic parameters were chosen as
 \begin{align}

### 2023-06-16 manuscript/manuscript.Rmd
started results section

@@
 For each of these parameter choices, tune the constant escapement strategy and train the DRL model on the corresponding model.
-Measure the average reward difference between strategies for each of these parameter choice samples.
-Average this difference over the 100 samples.
+Measure the average reward difference between strategies for each of these parameter choice samples using 100 replicate evaluation episodes.
+Take the mean of this average difference over the 100 perturbed parameter samples.
 The stability results are displayed in Fig. \textbf{TBD}.
@@
 
-We evaluated each of the three control approaches (MSY constant mortality, 80% MSY constant mortality and DRL) by running 50 episodes using the respective control rule.
+We evaluated each of the four control approaches considered -- MSY constant mortality, optimal constant escapement, PPO, and Gaussian process interpolation of the latter -- on each of the four scenarios examined (see Sec.
+TBD).
+Furthermore, we characterized the trade-off between economic output and sustainability faced by constant mortality policies.
+This was done by evaluating constant mortality policies with a fraction of the optimal mortality rate (specifically, 80%, 90% and 95%).
+All evaluations were based on 100 replicate episodes for each
+
+We will visualize the results concerning the last scenario considered (‘‘3 species, 2 of which harvested, and a time-varying parameter’’) in this section, leaving the other cases for the appendix.
+This is the most complex scenario considered and where our results show the most compelling advantage of DRL methods with respect to classical strategies.
+
+Our main result is summarized in Fig. **TBD** which displays the total reward distributions for the policy obtained through each strategy.
+Here we note two surprising things:
+First, optimal constant escapement performs considerably worse than not only DRL-based methods, but also than MSY.
+This conceptually shows how one-dimensional intuitions on ecological control may loose validity for complex ecosystems. **(this might be unwelcome by the reviewers if not discussed properly)**
+Second, we see that classical strategies have relatively spread out reward distributions as compared to DRL methods.
+
+## Old results (to be copy-pasted above)
+
 In Fig. \ref{fig:timeseries} we show the time series of one such repetition.

### 2023-06-17 manuscript/manuscript.Rmd
finished setting up the skeleton of the results section. now need to fill it out with plots

@@
 Second, we see that classical strategies have relatively spread out reward distributions as compared to DRL methods.
+
+To assess what is the culprit for the classical strategies' low performance with respect to the DRL-based strategies, we plot the duration of each of the 100 evaluation episodes, for each of the strategies.
+We see that early episode ends are a rather likely explanation for the wideness of constant escapement's wide reward distribution.
+The constant mortality strategy is also observed to lead to near-extinction events rather often, but in this case the episode lengths tend to be concentrated at higher values relative to the constant escapement case.
+
+We examine the trade-off between profit and sustainability faced by classical strategies in Fig. **TBD**.
+Here, two quantities are plotted: the fraction of evaluation episodes of maximal length (i.e. episodes with no near-extinction events) is and the average reward obtained.
+On the x-axis we have several sub-optimal constant mortality strategies that err on the conservative side: e.g. the policy labeled ‘‘80% MSY’’ has the form
+\begin{align*}
+  \pi: (X_t, Y_t, Z_t)\mapsto (0.8 M_X^*.\ 0.8 M_Y^*),
+\end{align*}
+where $(M_X^*,\ M_Y^*)$ is the optimal constat mortality strategy.
+We see that sufficiently conservative policies attain high sustainability, but only at a high price in terms of profit.\footnote{
+As noted before, here we equate economic profit with biomass caught.
+This is done as an approximation to convey the conceptual message more clearly, and we do not expect our results to significantly change if, e.g., ‘‘effort cost’’ is included in the reward function.
+When we refer to ‘‘large differences’’ in profit, or ‘‘paying dearly,’’ we mean that the ratio between average rewards is considerable---e.g. a 15\% loss in profit.
+}
+
+One problem that is often encountered when using machine learning methods is the interpretability of the output of these methods.
+In our case, the output is a policy function parametrized by a neural network $\theta$:
+\begin{align*}
+  \pi_\theta: N_t \mapsto (M_X(N_t), M_Y(N_t)),
+\end{align*}
+where $N_t=(X_t,Y_t,Z_t)$ is the state of the system at time $t$, and where $M_X$ and $M_Y$ are the mortalities due to harvest during that time-step.
+Here we visualize this policy function and provide an interpretation for it.
+Given its high dimension, it is not possible to fully display how the policy function obtained ‘‘looks like’’---we thus project it down to certain relevant axes.
+The result of this procedure is shown in Fig. **TBD**.
+In that figure, the shape of the optimal escapement strategy is provided for comparison.
+
+Moreover, in an attempt to further understand the relationship between the different management strategies considered, Fig.
+**TBD** presents the time-series of a representative episode for each strategy.
+Here we see a pattern that arises throughout the entire dataset---early episode ends are largely due to near-extinctions of $Z$.
 

### 2023-06-19 manuscript/manuscript.Rmd
added stability analysis subsection, still only a couple of sentences in it
change presentation rX(t)
started reframing mathematical models section

@@
 
-# Mathematical models of fisheries
-
-In this section we introduce some models describing the population dynamics of fisheries.
+# Mathematical models of fisheries considered
+
+Here we introduce the four mathematical models of fisheries for which we compare different management strategies.
+In general, the class of models that appear in this context are *stochastic, first order, finite difference equations*.
+For $n$ species, these models have the general form
+\begin{align}
+  \label{eq:general model}
+  \begin{split}
+    \Delta X_t &= f_X(N_t) + \eta_{X,t} - M_X X_t\\
+    \Delta Y_t &= f_Y(N_t) + \eta_{Y,t} - M_Y Y_t\\
+    \Delta Z_t &= f_Z(N_t) + \eta_{Z,t} - M_Z Z_t\\
+    &\dots
+  \end{split}
+\end{align}
+where $N_t = (X_t,\ Y_t,\ Z_t,\ \dots) \in \mathbb{R}^{n}_+$ is a vector of populations, $\Delta X_t := X_{t+1} - X_t$, $f_i:\mathbb{R}^{n}\to\mathbb{R}$ are arbitrary functions, and where $\eta_{i,t}$ are Gaussian random variables.
+Here, $M_i=M_i(N_t)\in\mathbb{R}_+$ is a state-dependent fish mortality arising from harvesting the $i$-th species (sometimes this is referred to as *fishing effort*).
+
+The term $M_X X_t$ is the total $X$ harvest at time $t$.
+This formulation of stock recruitment as a discrete finite difference process is common among fisheries, as opposed to continuous time formulations which involve instantaneous growth rates.
+This growth rate simplifies e.g.
+the possibly seasonal nature of reproduction (which would need to be accounted for in a realistic continuous-time model) by simply considering the total recruitment experienced by the population over a full year.
+
+One of the models we consider is a single-species classical model of ecological tipping points.
+The other three models are all three-species models with similar dynamics---differing by the amount of species that are harvested, and whether the parameters of the model are fixed or time-dependent.
+Following this logic, the first subsection will be dedicated to the single-species model and the second
+
+# Mathematical models of fisheries (old)
+
+In this section we introduce the models describing the population dynamics of fisheries.
 In general, the class of models that appear in this context are *first order finite difference equations* (potentially stochastic difference equations).
@@
     \Delta Y_t &= f_Y(N_t) - M_Y Y_t\\
+    \Delta Z_t &= f_Z(N_t) - M_Z Z_t\\
     &\dots
@@
 \end{align}
-where $N_t = (X_t,\ Y_t,\ \dots) \in \mathbb{R}^{n}_+$ is a vector of populations, $\Delta X_t := X_{t+1} - X_t$, $\Delta Y_t := Y_{t+1} - Y_t$, and where $f_i:\mathbb{R}^{n}\to\mathbb{R}$ are arbitrary functions.
-Here, $M_i=M_i(N_t)\in\mathbb{R}_+$ is a state-dependent fish mortality arising from harvesting (sometimes referred to as *fishing effort*).
-Thus, the term $M_X X_t$ is the total $X$ harvest at time $t$.
+where $N_t = (X_t,\ Y_t,\ Z_t,\ \dots) \in \mathbb{R}^{n}_+$ is a vector of populations, $\Delta X_t := X_{t+1} - X_t$, and where $f_i:\mathbb{R}^{n}\to\mathbb{R}$ are arbitrary functions.
+Here, $M_i=M_i(N_t)\in\mathbb{R}_+$ is a state-dependent fish mortality arising from harvesting the $i$-th species (sometimes this is referred to as *fishing effort*).
+
+The term $M_X X_t$ is the total $X$ harvest at time $t$.
 This formulation of stock recruitment as a discrete finite difference process is common among fisheries, as opposed to continuous time formulations which involve instantaneous growth rates.
@@
 the possibly seasonal nature of reproduction (which would need to be accounted for in a realistic continuous-time model) by simply considering the total recruitment experienced by the population over a full year.
+
+In the rest of this section we will focus our discussion the ‘‘natural’’ dynamics of the system, i.e. the dynamics of the system under zero fishing effort.
 
@@
 A seminal example of such dynamics was introduced in [@may77], using a one-dimensional model of a prey (resource) species under the pressure of a (fixed) predator:
-
 \begin{align}
@@
   f_{\text{May}}(X_t)
-  := L(X_t;\ r, K) - F(X_t, H; \beta,c),
+  := L(X_t;\ r, K) - \frac{\beta H X_t^2}{c^2 + X_t^2}.
 \end{align}
-where,
+In the following, we will denote
 \begin{align*}
@@
   \label{eq:rx(t)}
-  r_X = r_X(t) = \max(1-t/200, 1/2).
+  r_X = r_X(t) = 
+  \begin{cases}
+    1-t/200, \quad & t \leq 100,\\
+    1/2, \quad & t > 100
+  \end{cases}
 \end{equation}
@@
 Here we see a pattern that arises throughout the entire dataset---early episode ends are largely due to near-extinctions of $Z$.
+
+## Stability analysis
+
+In this section we present results intended to show that the effects that we observe in this paper are not the result of a careful selection of parameter values, but rather arise for a wide variety of parameter values.
 

### 2023-06-19 manuscript/manuscript.Rmd
new math section getting nice

@@
 
+The fishing efforts are the *control variables* of our problem---these may be set by the manager at each time-step to specified values.
+We make two further simplifying assumptions on the control problem:
+*1. Full observation:* the manager is able to accurately measure $N_t$ and use that measurement to inform their decision.
+*2. Perfect execution:* the action chosen by the manager is implemented perfectly (i.e., there is no noise affecting the actual value of the fishing efforts).
+In future work, we intend to relax these assumptions to get more realistic management scenarios.
+
 One of the models we consider is a single-species classical model of ecological tipping points.
 The other three models are all three-species models with similar dynamics---differing by the amount of species that are harvested, and whether the parameters of the model are fixed or time-dependent.
-Following this logic, the first subsection will be dedicated to the single-species model and the second
-
-# Mathematical models of fisheries (old)
-
-In this section we introduce the models describing the population dynamics of fisheries.
-In general, the class of models that appear in this context are *first order finite difference equations* (potentially stochastic difference equations).
-For $n$ species, these models have the general form
-\begin{align}
-  \label{eq:general model}
-  \begin{split}
-    \Delta X_t &= f_X(N_t) - M_X X_t\\
-    \Delta Y_t &= f_Y(N_t) - M_Y Y_t\\
-    \Delta Z_t &= f_Z(N_t) - M_Z Z_t\\
-    &\dots
-  \end{split}
-\end{align}
-where $N_t = (X_t,\ Y_t,\ Z_t,\ \dots) \in \mathbb{R}^{n}_+$ is a vector of populations, $\Delta X_t := X_{t+1} - X_t$, and where $f_i:\mathbb{R}^{n}\to\mathbb{R}$ are arbitrary functions.
-Here, $M_i=M_i(N_t)\in\mathbb{R}_+$ is a state-dependent fish mortality arising from harvesting the $i$-th species (sometimes this is referred to as *fishing effort*).
-
-The term $M_X X_t$ is the total $X$ harvest at time $t$.
-This formulation of stock recruitment as a discrete finite difference process is common among fisheries, as opposed to continuous time formulations which involve instantaneous growth rates.
-This growth rate simplifies e.g.
-the possibly seasonal nature of reproduction (which would need to be accounted for in a realistic continuous-time model) by simply considering the total recruitment experienced by the population over a full year.
-
-In the rest of this section we will focus our discussion the ‘‘natural’’ dynamics of the system, i.e. the dynamics of the system under zero fishing effort.
-
-## A single species model
+Following this logic, the first subsection will be dedicated to the single-species model and the second will focus on the three-species models.
+
+## The single species model
 
@@
 The most familiar model of $f(X)$ is that of *logistic growth*, for which
-
 \begin{align}
@@
   \label{eq:may}
-  f_{\text{May}}(X_t)
-  := L(X_t;\ r, K) - \frac{\beta H X_t^2}{c^2 + X_t^2}.
+  \Delta X_t
+  = L(X_t;\ r, K) - \frac{\beta H X_t^2}{c^2 + X_t^2} + \eta_{X,t} - M_X X_t.
 \end{align}
@@
 
-The model has five parameters: the growth rate $r$ and carrying capacity $K$ for $X$, a constant population $H$ of a species which preys on $X$, the maximal predation rate $\beta$, and the predation rate half-maximum biomass $c$.
+The model has six parameters: the growth rate $r$ and carrying capacity $K$ for $X$, a constant population $H$ of a species which preys on $X$, the maximal predation rate $\beta$, the predation rate half-maximum biomass $c$, and the variance $\sigma_X^2$ of $\eta_{X,t}$.
+(Here and in the following we will center our random variables at zero.)
 
@@
 #| label: may
-#| fig.cap: "The fixed point diagram for the model associated to $f_{\\text{May}}$ as a function of varying the parameter $\\beta H$.
-#| That is, these are the zeroes of the function $f_{\\text{May}}$. 
+#| fig.cap: "The fixed point diagram for model (TBD) as a function of varying the parameter $\\beta H$, assuming zero noise.
 #| Stable fixed points (also known as attractors) are plotted using a solid line, while the unstable fixed point is shown as a dotted line."
@@
 
-## A three species model
-
-Consider the following three species recruitment model which generalizes eq. \eqref{eq:may}:
+## The three species models
+
+We consider three models of a three-species fishery ecosystem.
+These are all closely related---in fact, their natural dynamics (i.e. dynamics under zero harvest) is given by the same equations:
 \begin{align}
@@
 \begin{split}
-  \Delta X_t &= L(X_t;\ r_X, K_X) - F(X_t, Z_t;\ \beta, c) - c_{XY}X_tY_t - h_{X,t},\\
-  \Delta Y_t &= L(Y_t;\ r_Y, K_Y) - D F(Y_t, Z_t;\ \beta, c) - c_{XY}X_tY_t - h_{Y,t},\\
-  \Delta Z_t &= \left(f(X_t + DY_t\right) - d_Z)Z_t.
+  \Delta X_t &= L(X_t;\ r_X, K_X) - F(X_t, Z_t;\ \beta, c) - c_{XY}X_tY_t + \eta_{X,t},\\
+  \Delta Y_t &= L(Y_t;\ r_Y, K_Y) - D F(Y_t, Z_t;\ \beta, c) - c_{XY}X_tY_t + \eta_{Y,t},\\
+  \Delta Z_t &= \left(f(X_t + DY_t\right) - d_Z)Z_t + \eta_{Z,t}.
 \end{split}
 \end{align}
-It contains three populations, the state at time $t$ is $(X_t, Y_t, Z_t)$. 
+The three species modelled are $X$, $Y$ and $Z$.
 Species $Z$ preys on both $X$ and $Y$, while the latter two compete for resources.
@@
 
-As mentioned in the introduction, we consider two harvesting scenarios: one where only $X$ is harvested (i.e. $h_{Y,t}=0$), and one where both $X$ and $Y$ are harvested.
-For most of our paper, parameters used in this model will be given as in eq. \eqref{eq:param vals}.
-In the time-varying model, moreover, we use the following drift on the growth rate of $X$:
+The three models will branch off of eq. \eqref{eq:3d model} in the following way.
+*1. Three species, one fishery:* in this first model, only $X$ is harvested, that is we fix $M_Y=M_Z=0$ and leave $M_X$ as a control variable.
+All parameters here are constant.
+*2. Three species, two fisheries:* as the previous case but with $X$ and $Y$ being harvested. 
+In other words, we set $M_Z=0$ and leave $M_X$ and $M_Y$ as control variables.
+Finally, *3. Three species, two fiseries, time-variation:* in this model we allow for the harvest of $X$ and $Y$, as in the previous model, but now we include a time dependent parameter:
 \begin{equation}
@@
     1-t/200, \quad & t \leq 100,\\
-    1/2, \quad & t > 100
+    1/2, \quad & t > 100.
   \end{cases}
 \end{equation}
+This is intended to reflect in a simple manner a possible effect of climate change: where the reproductive rate of $X$ is reduced linearly over time until it stabilizes.
+In future work we intend to capture more realistic models of the effects of climate change, but for this exploratory work we have chosen this simple expression for the sake of conceptual clarity.
+
+## Old 3 species text (to be copy-pased)
+
+As mentioned in the introduction, we consider two harvesting scenarios: one where only $X$ is harvested (i.e. $h_{Y,t}=0$), and one where both $X$ and $Y$ are harvested.
+For most of our paper, parameters used in this model will be given as in eq. \eqref{eq:param vals}.
+In the time-varying model, moreover, we use the following drift on the growth rate of $X$:
 

### 2023-06-20 manuscript/manuscript.Rmd
updated methods and math sections

@@
 Species $Z$ preys on both $X$ and $Y$, while the latter two compete for resources.
-There are ten parameters in this model: 
+There are thirteen parameters in this model: 
 The growth rate and carrying capacity, $r_X$, $K_X$, $r_Y$ and $K_Y$, of $X$ and $Y$.
@@
 A parameter $D$ regulating a relative preference of $Z$ to prey on $Y$.
-Finally, a death rate $d_Z$ and a parameter $f$ scaling the birth rate of $Z$.
+A death rate $d_Z$ and a parameter $f$ scaling the birth rate of $Z$.
+Finally, the noise variances $\sigma_X$, $\sigma_Y$ and $\sigma_Z$.
+
+Eqs. \eqref{eq:3d model} are a generalization of the single-species model \eqref{eq:may}: this can be seen by setting 
+\begin{align*}
+  c_{XY} = d = f = \sigma_Z = 0.
+\end{align*}
+In our simulations, we use $c_{XY} = d = f = \sigma_Z = 0.1$ so, in broad strokes, our three-species model is a slight perturbation of the single-species model.
+Put differently, we are in a regime where, just by examining parameters, it would seem that the single-species model is a reasonable approximation.
+**Is this clear? Should I maybe erase these last sentences?**
 
 The three models will branch off of eq. \eqref{eq:3d model} in the following way.
-*1. Three species, one fishery:* in this first model, only $X$ is harvested, that is we fix $M_Y=M_Z=0$ and leave $M_X$ as a control variable.
+*1. Three species, one fishery:* in this first model, only $X$ is harvested, that is, in the notation of eq. \eqref{eq:general model}, we fix $M_Y=M_Z=0$ and leave $M_X$ as a control variable.
 All parameters here are constant.
@@
 
-## Old 3 species text (to be copy-pased)
-
-As mentioned in the introduction, we consider two harvesting scenarios: one where only $X$ is harvested (i.e. $h_{Y,t}=0$), and one where both $X$ and $Y$ are harvested.
-For most of our paper, parameters used in this model will be given as in eq. \eqref{eq:param vals}.
-In the time-varying model, moreover, we use the following drift on the growth rate of $X$:
-
-This model generalizes eq. \eqref{eq:may} in the following sense:
-If $f=d_Z=c_{XY}=0$, and we let $Z_{t=0}=H$, then eq. \eqref{eq:3d model} becomes
-\begin{align*}
-  \Delta X_t &= L(X_t;\ r_X, K_X) - F(X_t, H;\ \beta, c),\\
-  \Delta Y_t &= L(Y_t;\ r_Y, K_Y) - D F(Y_t, H;\ \beta, c),\\
-  Z_t &= H.
-\end{align*}
-The equations for $X_t$ and $Y_t$ thus become decoupled, with $X_t$ evolving according to \eqref{eq:may}.
-
-As can be expected, the dynamics of eq. \eqref{eq:3d model} are significantly more complex than those of eq. \eqref{eq:may}.
-Here, a fixed point analysis provides a less intuitive picture of the dynamics---it no longer necessarily alternates in a tipping manner between two stable points.
-
-## Stochastic dynamics
-
-We have, up to now, presented eqs. \eqref{eq:may} and \eqref{eq:3d model} as *deterministic* dynamical models for the sake of clarity.
-In practice, we will consider a stochastic contribution to these dynamics.
-This way, instead of eq. \eqref{eq:3d model}, we consider the problem of harvesting the following stochastic model:
-<!--
-\begin{align}
-  \label{eq:may stoch}
-  \Delta X_t
-  = L(X_t;\ r, K) - F(X_t, H; \beta,c) - h_t + \eta_t,
-\end{align}
-where $\{\eta_t\}_t$ are independent Gaussian variables with zero mean.
-The variance $\sigma^2$ of $\eta_t$ is an additional parameter of the model.
-
-Similarly, we use the following dynamical model in place of eq. \eqref{eq:3d model}-->
-\begin{align}
-  \label{eq:3d model stoch}
-  \begin{split}
-    \Delta X_t &= L(X_t;\ r_X, K_X) - F(X_t, Z_t;\ \beta, c) - c_{XY}X_tY_t - h_t + \eta_{X,t},\\
-    \Delta Y_t &= L(Y_t;\ r_Y, K_Y) - D F(Y_t, Z_t;\ \beta, c) - c_{XY}X_tY_t + \eta_{Y,t},\\
-    \Delta Z_t &= \left(f(X_t + DY_t\right) - d_Z)Z_t + \eta_{Z,t}.
-\end{split}
-\end{align}
-Here, $\{\eta_{i,t}\}$ is a set of independent Gaussian variables centered at zero with variances, respectively, $\sigma_X^2$, $\sigma_Y^2$ and $\sigma_Z^2$.
-
 # Reinforcement learning {#sec:RL}
@@
 This secures a reward of $M_X(N_t)X$ in the single fishery case, and in the two-fishery case the reward is $M_X(N_t)X + M_Y(N_t)Y$.
-After this harvest portion of the time step, the environment changes according to its natural dynamics given by \eqref{eq:3d model stoch}.
+After this harvest portion of the time step, the environment changes according to its natural dynamics given by \eqref{eq:3d model}.
 
@@
 This approximation is commonly used in fisheries for species with annual reproductive cycles (see e.g. [@mangel2006theoretical], Chap. 6).
+The division of the time-step into two parts (harvest and dynamics) happening sequentially is performed for all control methods evaluated, not only RL-based approaches.
 
@@
 This tuning procedure was analogous to that of the constant mortality strategy just summarized.
-Namely, \textbf{TBD}.
+Namely:
+A grid of 101 constant escapement values was laid out on the interval $[0,1]$ for the single-fishery models (resp. a $51\times 51$ grid on $[0,1]^2$ for two-fishery models).
+These grids were the possible constant escapement values optimized over (in the two fishery case, these were $(\text{esc}_X,\ \text{esc}_Y)$ pairs).
+That is, each grid point represents a constant escapement policy.
+We used each of these policies to manage 100 replicate episodes.
+The optimal policy was the policy with the highest average reward obtained.
 
@@
 \label{eq:1d param vals}
-r = K = 1, \beta = 0.25, c = 0.1.
+r = K = 1,\quad \beta = 0.25,\quad c = 0.1.
 \end{align}

### 2023-06-20 manuscript/manuscript.Rmd
mini

@@
 \end{align*}
-In our simulations, we use $c_{XY} = d = f = \sigma_Z = 0.1$ so, in broad strokes, our three-species model is a slight perturbation of the single-species model.
+In our simulations, we use $c_{XY} = d = f = 0.1$ and $\sigma_Z=0.05$ so, in broad strokes, our three-species model is a slight perturbation of the single-species model.
 Put differently, we are in a regime where, just by examining parameters, it would seem that the single-species model is a reasonable approximation.

### 2023-06-20 manuscript/manuscript.Rmd
stability subsection, recovering c esc in 1D subsection

@@
 This was done by evaluating constant mortality policies with a fraction of the optimal mortality rate (specifically, 80%, 90% and 95%).
-All evaluations were based on 100 replicate episodes for each
+All evaluations were based on 100 replicate episodes.
 
@@
 
-To assess what is the culprit for the classical strategies' low performance with respect to the DRL-based strategies, we plot the duration of each of the 100 evaluation episodes, for each of the strategies.
+To assess what is the culprit for the classical strategies' low performance with respect to the DRL-based strategies, we plot the duration of each of the 100 evaluation episodes of each strategy.
 We see that early episode ends are a rather likely explanation for the wideness of constant escapement's wide reward distribution.
@@
 
+## Recovering constant escapement for a single species
+
+While the optimal control for our single-species model \eqref{eq:may} can not be easily proven to be constant escapement (since the right-hand side of that equation is not concave), from experience we can expect for constant escapement to either be optimal or near-optimal.
+We give evidence for this intuition by showing that our DRL method recovers a constant escapement solution when trained.
+These results are shown in Fig. TBD.
+
 ## Stability analysis
@@
 In this section we present results intended to show that the effects that we observe in this paper are not the result of a careful selection of parameter values, but rather arise for a wide variety of parameter values.
+
+Our main result in this respect is Fig. **TBD**.
+There, we plot the *median episode reward difference* between the two DRL-based methods we considered, and the optimal constant escapement strategy.
+This figure shows that, for a wide range of parameter values, DRL-based strategies can have a considerable advantage over an optimized constant escapement policy (the single-species optimal solution).
 

### 2023-06-21 manuscript/manuscript.Rmd
added discussion on drl policy and time-series

@@
 
-Moreover, in an attempt to further understand the relationship between the different management strategies considered, Fig.
-**TBD** presents the time-series of a representative episode for each strategy.
-Here we see a pattern that arises throughout the entire dataset---early episode ends are largely due to near-extinctions of $Z$.
+We notice that the DRL-derived policy has similarities to a constant escapement policy.
+Here, the key difference is that the escapement value for each of the fished species is sensitive to variations in the other populations. 
+This can be seen as a color gradients in the plots of $(X, M_X)$ and $(Y, M_Y)$, where the gradient corresponds to differing values of $Z$.
+Moreover, this can be seen as an anti-correlation in the plot of $(X, M_Y)$---for optimal constant escapement, $M_Y$ is uncorrelated to $X$.
+
+The sensitivity of the policy to, e.g., the values of $Z$ can be seen in the sample time series displayed in Fig. **TBD**.
+Here, we can see that species $Z$ becomes endangered due to harvesting for all management strategies.
+The DRL-based strategy, however, is sensitive to the values of $Z$ and can respond accordingly by scaling the fishing effort with the value of $Z$.
+In particular, the policy responds to periods of low values of $Z$ by restricting fishing on $X$ and $Y$, thus promoting $Z$'s growth.
+This pattern is rather common among the whole dataset---early episode ends are largely due to near-extinctions of $Z$ for all management strategies.
 

### 2023-06-21 manuscript/manuscript.Rmd
erased old plots not to be used

@@
 ```{r}
-# for best const mortality and best const escapement
-msy_optimizer <- function(df) {
-  df |> 
-  group_by(action, rep) |>
-  filter(t == max(t)) |> 
-  group_by(action) |>
-  summarise(mean_reward = mean(reward), sd = sd(reward)) |> 
-  filter(mean_reward == max(mean_reward))
-}
-escapement_optimizer <- function(df) {
-  df |> 
-  group_by(escapement, rep) |>
-  filter(t == max(t)) |> 
-  group_by(escapement) |>
-  summarise(mean_reward = mean(reward), sd = sd(reward)) |> 
-  filter(mean_reward == max(mean_reward))
-}
 
-# Collect data (outsource to separate R script later TBD)
-msy_df <- read_csv("../data/msy.csv.xz")
-best_action <- msy_optimizer(msy_df)$action
-escapement_df <- read_csv("../data/escapement.csv.xz", show_col_types = FALSE)
-best_e <- escapement_optimizer(escapement_df)$escapement
-ppo_df <- read_csv("../data/PPO200.csv.xz")
-
-opt_escapement <- escapement_df |> filter(escapement == best_e)
-
-# set up for plots:
-
-msy_sim <- msy_df |>
-    filter(action == best_action) |> 
-    group_by(rep) |>
-    pivot_longer(c("X", "Y", "Z"),
-                names_to = "species", values_to = "abundance") |>
-    mutate(abundance = abundance + 1, policy = "msy") # natural units, policy name
-
-actions <- unique(msy_df$action)
-i <- which.min(abs(actions - best_action* 0.8))
-tac_sim <- msy_df |> 
-    filter(action == actions[[i]]) |>
-    pivot_longer(c("X", "Y", "Z"),
-        names_to = "species", values_to = "abundance") |>
-    mutate(abundance = abundance + 1, policy = "tac")
-
-ppo_sim <- ppo_df |> 
-    pivot_longer(c("X", "Y", "Z"),
-        names_to = "species", values_to = "abundance") |>
-    mutate(abundance = abundance + 1, policy = "ppo")
-
-esc_sim_df <- escapement_df |> 
-    pivot_longer(c("X", "Y", "Z"),
-        names_to = "species", values_to = "abundance") |>
-    mutate(abundance = abundance + 1)
-
-msy_tac_ppo <- bind_rows(msy_sim, tac_sim, ppo_sim)
 ```
@@
 
-```{r}
-#| label: tuning
-#| fig.cap : "Result of the tuning procedure for the constant mortality strategy. 
-#| (A) Mean episode reward (averaged over 50 episodes) shown as scatter data, and shaded in gray are two standard deviations around each scatter point.
-Vertical lines correspond to the optimal constant mortality rate $M^* = $ 0.045 (green) and $0.8 M^*=$ 0.036 (violet).
-#| (B) Several constant mortality policies shown for illustrative purposes.
-Slopes $0.02$, $0.03$, $0.055$ and $0.065$ shown in black; the optimal linear policy $M^*X$ shown in green, while $0.8M^*X$ shown in violet."
-#| out.width : "6.5in"
-msy_rewards <- msy_df |>
-  group_by(action, rep) |>
-  filter(t == max(t)) |> 
-  group_by(action) |>
-  summarise(mean_reward = mean(reward),
-            sd = sd(reward))
-
-rmsy <- msy_rewards |>
-  ggplot(aes(action, mean_reward, 
-             ymin = mean_reward - 2*sd,
-             ymax=mean_reward+2*sd)) +
-  geom_point() + geom_ribbon(alpha=0.4) +
-  labs(y = "episode reward", x = "constant mortality") +
-  theme(aspect.ratio = 1/1) + 
-  geom_vline(aes(xintercept=best_action), color="seagreen3", size=1) + 
-  geom_vline(aes(xintercept=0.8*best_action), color="violetred3", size=1) +
-  theme_bw() + 
-  theme(panel.grid.major = element_blank(), 
-        panel.grid.minor = element_blank(),
-        aspect.ratio = 1)
-
-# linear function blueprint
-linear_f <- function(slope, x) slope*x
-# generates linear function with given slope
-linear_f_gen <- function(slope) (function(x) slope*x)
-
-# dummy data set to keep ggplot happy
-p <- ggplot(data = data.frame(x = 0), mapping = aes(x = x))
-
-# p + stat_function(fun = linear_f_gen(0.5)) + xlim(0,2)
-lines <- p + stat_function(fun = linear_f_gen(0.009)) + 
-  stat_function(fun = linear_f_gen(0.018)) +
-  stat_function(fun = linear_f_gen(0.027)) +
-  stat_function(fun = linear_f_gen(0.8*best_action), col = "violetred3", size=1) +
-  stat_function(fun = linear_f_gen(best_action), col = "seagreen3", size=1) +
-  stat_function(fun = linear_f_gen(0.054)) + 
-  stat_function(fun = linear_f_gen(0.063)) +
-  stat_function(fun = linear_f_gen(0.072)) +
-  stat_function(fun = linear_f_gen(0.081)) +
-  stat_function(fun = linear_f_gen(0.090)) +
-  xlim(0,2) +
-  xlab("X") + ylab("harvest") +
-  theme_bw() +
-  theme(#panel.grid.major = element_blank(), 
-        #panel.grid.minor = element_blank(), 
-        aspect.ratio = 1)
-
-(rmsy | lines) + plot_annotation(tag_levels = c('A'))
-```
-
 \textbf{Tuning the constant escapement strategy.}
@@
 
-```{r} 
-#| label: timeseries 
-#| fig.cap: "A sample time series for the three fish populations, with the system being controlled, respectively, by the MSY optimal mortality, by $80$ percent of the MSY optimal mortality, and by DRL."
-#| out.width: "6.5in"
-policy_labels <- list("msy" = "MSY Const. Mortality", "tac"= "80% MSY Const. Mortality", "ppo"= "DRL")
-policy_labeller <- function(variable,value){
-  return(policy_labels[value])
-}
-msy_tac_ppo |> 
-    group_by(rep, policy) |>
-    filter(rep == 6) |>
-    ggplot(aes(t, abundance, col = species)) + 
-        geom_line()  + 
-        scale_color_manual(values=c('palegreen3', 'coral1', 'deepskyblue4')) +
-        facet_wrap(~factor(policy, levels = c("msy", "tac", "ppo")), labeller = policy_labeller ) +
-        labs(y = "abundance", x = "time") +
-        theme_bw() +
-        theme(aspect.ratio = 2/3)
-```
-
 This time series is representative of the patterns observed in the data (see App. A.1).
@@
 Second, that while the conservative 80% approach has a stable behavior, it consistently produces smaller harvests than a DRL generated policy.
-
-```{r} 
-#| label: rew_len_fig
-#| fig.cap: "Bar charts for the episode lengths (top panel) and episode total harvests (bottom panel) for 50 episodes." 
-#| out.width: "6.5in"
-p1 <- msy_tac_ppo |> 
-    group_by(rep, policy) |>
-    filter(t==max(t), species == "X", rep < 50) |>
-    ggplot(aes(rep, t)) + 
-      geom_col() +
-      facet_wrap(~factor(policy, levels = c("msy", "tac", "ppo")), labeller = policy_labeller ) +
-      labs(y = "episode length", x = "repetition") +
-      theme_bw() + 
-      theme(axis.text.x = element_blank())
-
-p2 <- msy_tac_ppo |> 
-    group_by(rep, policy) |>
-    filter(t==max(t), species == "X", rep < 50) |>
-    ggplot(aes(rep, reward)) + 
-      geom_col() +
-      facet_wrap(~factor(policy, levels = c("msy", "tac", "ppo")), labeller = policy_labeller ) +
-      labs(y = "episode harvest", x = "repetition") +
-      theme_bw() + 
-      theme(axis.text.x = element_blank())
-
-(p1 / p2) + plot_annotation(tag_levels = 'A')
-```
 
@@
 
-```{r}
-#| label: action-X
-#| fig.cap: "(A-B) Harvest and escapement scatter data for the DRL policy together with the optimal constant escapement (pink) and constant mortality (green) policies for comparison. 
-#| (A) includes $Y$ dependence through color, whereas (B) includes $Z$ dependence similarly.
-#| (B-C) escapement (i.e. $X_t - h_t$) scatter data for the DRL policy. 
-#| Optimal constant escapement policy shown in pink for comparison."
-#| out.width: "6.5in"
-opt_escapement_harvest <- function(X) case_when(
-  X > best_e ~ X - best_e,
-  X <= best_e ~ 0
-)
-
-opt_escapement_escapement <- function(X) case_when(
-  X <= best_e ~ X,
-  X > best_e ~ best_e
-)
-
-msy_harvest <- function(X) best_action*X
-tac_harvest <- function(X) 0.8*msy_harvest(X)
-
-py <- ppo_df |>
-  mutate(X = X+1, Y = Y+1, Z = Z+1, harvest = action*X) |>
-  ggplot(aes(X, harvest, color = Y)) +
-  geom_point() +
-  theme_bw() + 
-  theme(aspect.ratio = 1/1) +
-  stat_function(fun = opt_escapement_harvest, col = 'deeppink', size = 0.7) + 
-  stat_function(fun = msy_harvest, col = 'seagreen2', size = 0.7)
-
-pz <- ppo_df |>
-  mutate(X = X+1, Y = Y+1, Z = Z+1, harvest = action*X) |>
-  ggplot(aes(X, harvest, color = Z)) +
-  geom_point() +
-  theme_bw() + 
-  theme(aspect.ratio = 1/1) +
-  stat_function(fun = opt_escapement_harvest, col = 'deeppink', size = 0.7) + 
-  stat_function(fun = msy_harvest, col = 'seagreen2', size = 0.7)
-
-escxy <- ppo_df |>
-  mutate(X = X+1, Y = Y+1, Z = Z+1, escapement = (1-action)*X) |>
-  ggplot(aes(X, escapement, color = Y)) +
-  geom_point() +
-  theme_bw() + 
-  theme(aspect.ratio = 1/1) +
-  stat_function(fun = opt_escapement_escapement, col = 'deeppink', size = 0.7)
-
-escxz <- ppo_df |>
-  mutate(X = X+1, Y = Y+1, Z = Z+1, escapement = (1-action)*X) |>
-  ggplot(aes(X, escapement, color = Z)) +
-  geom_point() +
-  theme_bw() + 
-  theme(aspect.ratio = 1/1) +
-  stat_function(fun = opt_escapement_escapement, col = 'deeppink', size = 0.7)
-
-((py | pz)/ (escxy | escxz)) + plot_annotation(tag_levels = c('A'))
-```
-
 In Fig. \ref{fig:state-space}, we perform a back to back visualization of all the policies evaluated---MSY constant mortality, 80% MSY constant mortality, DRL and optimal constant escapement.
@@
 This anti-correlation is, furthermore, observed in the time series plots (see Fig. \ref{fig:timeseries} and App. A.1).
-
-```{r}
-#| label: state-space
-#| fig.cap: "Policy heatmaps.
-Each policy, say, $\\pi$, was evaluated at points $(x,y,z)$ visited during any of the 50 evaluation rounds.
-We projected this data down to the $X$-$Y$ plane---i.e.
-we plotted the heatmap $\\pi(x,y,z)$ as a function of $x$ and $y$."
-#| out.width: "6.5in"
-i_tac <- which.min(abs(actions - best_action * 0.8))
-
-# color pallette limits
-library(RColorBrewer)
-myPalette <- colorRampPalette(rev(brewer.pal(11, "Spectral")))
-sc <- scale_colour_gradientn(colours = myPalette(100), limits=c(0,0.25))
-
-pmsy <- msy_df |>
-  mutate(X = X+1, Y = Y+1, Z = Z+1, harvest = action * X) |>
-  filter(action == best_action) |>
-  ggplot(aes(X, Y, color = harvest)) +
-  geom_point() + sc +
-  xlim(0,1) +
-  ylim(0,0.85) +
-  theme_bw() + 
-  ggtitle("Opt. Const. Mortality")
-
-ptac <- msy_df |>
-  mutate(X = X+1, Y = Y+1, Z = Z+1, harvest = action * X) |>
-  filter(action == actions[[i_tac]]) |>
-  ggplot(aes(X, Y, color = harvest)) +
-  geom_point() + sc +
-  xlim(0,1) +
-  ylim(0,0.85) +
-  theme_bw() + 
-  ggtitle("80% Opt. Const. Mortality")
-
-pppo <- ppo_df |>
-  mutate(X = X+1, Y = Y+1, Z = Z+1, harvest = action * X) |>
-  ggplot(aes(X, Y, color = harvest)) +
-  geom_point() + sc +
-  xlim(0,1) +
-  ylim(0,0.85) +
-  theme_bw() + 
-  ggtitle("DRL")
-
-pesc <- opt_escapement |>
-  mutate(X = X+1, Y = Y+1, Z = Z+1, 
-        harvest = case_when(
-          X - escapement > 0 ~ X - escapement,
-          X - escapement <= 0 ~  0)
-         ) |> 
-  ggplot(aes(X, Y, color = harvest)) +
-  geom_point() + sc +
-  xlim(0,1) +
-  ylim(0,0.85) +
-  theme_bw() + 
-  ggtitle("Opt. Const. Escapement")
-
-(
-  (pmsy | ptac) / (pppo | pesc) +
-  plot_layout(guides = "collect") & theme(legend.position = 'right')
-) + 
-  plot_annotation(tag_levels = c('A'))
-```
 
@@
 
-```{r} 
-#| label: timeseries_msy
-#| fig.cap: "Nine time series of the system controlled by the MSY constant mortality policy." 
-#| out.width: "6.5in"
-msy_tac_ppo |> 
-    group_by(rep) |>
-    filter(rep < 9, policy == "msy") |>
-    ggplot(aes(t, abundance, col = species)) + 
-        geom_line()  + 
-        scale_color_manual(values=c('palegreen3', 'coral1', 'deepskyblue4')) +
-        facet_wrap(~factor(rep)) +
-        labs(y = "abundance", x = "time") +
-        theme_bw() + 
-        theme(aspect.ratio = 0.5)
-```
-
-```{r} 
-#| label: timeseries_tac
-#| fig.cap: "Nine time series of the system controlled by the policy using 80 percent of the MSY constant mortality." 
-#| out.width: "6.5in"
-msy_tac_ppo |> 
-    group_by(rep) |>
-    filter(rep < 9, policy == "tac") |>
-    ggplot(aes(t, abundance, col = species)) + 
-        geom_line()  + 
-        scale_color_manual(values=c('palegreen3', 'coral1', 'deepskyblue4')) +
-        facet_wrap(~factor(rep)) +
-        labs(y = "abundance", x = "time") +
-        theme_bw() + 
-        theme(aspect.ratio = 0.5)
-```
-
-```{r} 
-#| label: timeseries_ppo
-#| fig.cap: "Nine time series of the system controlled by the trained DRL agent." 
-#| out.width: "6.5in"
-msy_tac_ppo |> 
-    group_by(rep) |>
-    filter(rep < 9, policy == "ppo") |>
-    ggplot(aes(t, abundance, col = species)) + 
-        geom_line()  + 
-        scale_color_manual(values=c('palegreen3', 'coral1', 'deepskyblue4')) +
-        facet_wrap(~factor(rep)) +
-        labs(y = "abundance", x = "time") +
-        theme_bw() + 
-        theme(aspect.ratio = 0.5)
-```
-
 ## Conservative scale-downs of MSY
@@
 
-```{r}
-#| label: several-tac
-#| fig.cap: "Performance of constant mortality policies with mortality $\\alpha M^*$ for several values of $\\alpha$.
-#| The top panel displays the episode lengths for each repetition, while the bottom pannel displays the episode harvest.
-#| The labels in the plots are the value of $\\alpha$ expressed as a percentage."
-#| out.width: "6.5in"
-N <- 30
-actions <- unique(msy_df$action)
-i85 <- which.min(abs(actions - best_action* 0.85))
-i90 <- which.min(abs(actions - best_action* 0.90))
-i95 <- which.min(abs(actions - best_action* 0.95))
-tac_85_sim <- msy_df |> 
-    filter(action == actions[[i85]]) |>
-    pivot_longer(c("X", "Y", "Z"),
-        names_to = "species", values_to = "abundance") |>
-    mutate(abundance = abundance + 1, policy = "85%")
-tac_90_sim <- msy_df |> 
-    filter(action == actions[[i90]]) |>
-    pivot_longer(c("X", "Y", "Z"),
-        names_to = "species", values_to = "abundance") |>
-    mutate(abundance = abundance + 1, policy = "90%")
-tac_95_sim <- msy_df |> 
-    filter(action == actions[[i95]]) |>
-    pivot_longer(c("X", "Y", "Z"),
-        names_to = "species", values_to = "abundance") |>
-    mutate(abundance = abundance + 1, policy = "95%")
-
-tac_80_sim <- tac_sim |> mutate(policy = "80%")
-
-tac4 <- bind_rows(tac_80_sim, tac_85_sim, tac_90_sim, tac_95_sim)
-
-p1 <- tac4 |> 
-    group_by(rep, policy) |>
-    filter(t==max(t), species == "X", rep < N) |>
-    ggplot(aes(rep, t)) + 
-      geom_col() +
-      facet_wrap(~factor(policy), ncol = 4) +
-      labs(y = "episode length", x = "repetition") +
-      theme_bw() + 
-      theme(axis.text.x = element_blank())
-
-p2 <- tac4 |> 
-    group_by(rep, policy) |>
-    filter(t==max(t), species == "X", rep < N) |>
-    ggplot(aes(rep, reward)) + 
-      geom_col() +
-      facet_wrap(~factor(policy), ncol=4) +
-      labs(y = "episode harvest", x = "repetition") +
-      theme_bw() + 
-      theme(axis.text.x = element_blank())
-
-(p1 / p2)
-```
-
 ## Optimal constant escapement policy evaluation
@@
 
-```{r}
-#| label: esc-tuning-eval
-#| fig.cap: "(Left panel) Tuning results for constant escapement policies: mean episode reward scatter data (averaged over 50 episodes) together with two standard deviations shaded in grey. 
-#| The optimal escapement level (the one producing maximum mean harvest) is $X_{\\text{esc.}}=0.62$.
-#| (Middle and right panels) Reward histograms for the optimal constant escapement policy (middle) and for the DRL policy (right).
-#| For both DRL and optimal constant escapement, no early episode ends are observed (i.e. all episodes last for 200 time steps)."
-#| out.width: "6.5in"
-#| fig.align: "center"
-esc_rewards <- escapement_df |>
-  group_by(escapement, rep) |>
-  filter(t == max(t)) |> 
-  group_by(escapement) |>
-  summarise(mean_reward = mean(reward),
-            sd = sd(reward))
-
-resc <- esc_rewards |>
-  ggplot(aes(escapement, mean_reward, 
-             ymin = mean_reward - 2*sd,
-             ymax=mean_reward+2*sd)) +
-  geom_point() + geom_ribbon(alpha=0.4) +
-  labs(y = "episode reward", x = "constant escapement") +
-  theme_bw() +
-  theme(aspect.ratio = 1/1)
-
-esc_len <- esc_sim_df |>
-  group_by(rep) |>
-  filter(t==max(t), species == "X", escapement == best_e, rep < 50) |>
-  ggplot(aes(rep, t)) + 
-      geom_col() +
-      labs(y = "episode length", x = "repetition") +
-      theme_bw() + 
-      theme(axis.text.x = element_blank())
-
-esc_rew <- esc_sim_df |>
-  group_by(rep) |>
-  filter(t==max(t), species == "X", escapement == best_e, rep < 50) |>
-  ggplot(aes(rep, reward)) + 
-      geom_col() +
-      labs(y = "episode harvest", x = "repetition", title = "Escapement") +
-      ylim(0,12.8) +
-      #ggtitle("Escapement") + 
-      theme_bw() + 
-      theme(axis.text.x = element_blank(), plot.title = element_text(hjust = 0.5))
-
-drl_rew <- msy_tac_ppo |> 
-    group_by(rep, policy) |>
-    filter(t==max(t), species == "X", policy == "ppo", rep < 50) |>
-    ggplot(aes(rep, reward)) + 
-      geom_col() +
-      labs(y = "episode harvest", x = "repetition", title = "DRL") +
-      ylim(0,12.8) +
-      #ggtitle("DRL") + 
-      theme_bw() + 
-      theme(axis.text.x = element_blank(), plot.title = element_text(hjust = 0.5))
-
-(resc | esc_rew | drl_rew)
-```
-
 Fig. \ref{fig:timeseries_esc} displays 9 time-series for episodes controlled with the optimal constant escapement policy.
@@
 
-```{r} 
-#| label: timeseries_esc
-#| fig.cap: "Nine time series of the system controlled by the optimal constant escapement policy." 
-#| out.width: "6.5in"
-esc_sim_df |> 
-    group_by(rep) |>
-    filter(rep < 9, escapement == best_e) |>
-    ggplot(aes(t, abundance, col = species)) + 
-        geom_line()  + 
-        scale_color_manual(values=c('palegreen3', 'coral1', 'deepskyblue4')) +
-        facet_wrap(~factor(rep)) +
-        labs(y = "abundance", x = "time") +
-        theme_bw() + 
-        theme(aspect.ratio = 0.5)
-```
 # References

### 2023-06-21 manuscript/manuscript.Rmd
comment on PPO policy evaluation on grid and how the interpolation is done

@@
 \end{align*}
+This interpolation method is applied to scatter data of the PPO policy evaluated on 3 different grids:
+One $51\times 51 \times 5$, one $51\times 5 \times 51$ and one $5\times 51 \times 51$ grid on $(X,Y,Z)$ states.\footnote{
+The raw dataset is found at \textbf{TBD} in the repository with the source code and data linked above.
+Scatter plots visualizing this policy are shown in App. \textbf{TBD}.
+}
 The length scale and noise level values of this kernel were chosen arbitrarily---no hyperparameter tuning was needed to produce satisfactory interpolation, as will be shown in the results section.

### 2023-06-21 manuscript/manuscript.Rmd
created summary data for the plots, moved some of the relevant data for main text plots to manuscript_data
added main result violin plot to paper

@@
 library(patchwork)
+library(ggplot2)
 ```
@@
 ```{r}
+# read data
+#frac_msy_8 <- read_csv("manuscript_data/0.8msy.csv.xz")
+#frac_msy_9 <- read_csv("manuscript_data/0.9msy.csv.xz")
+#frac_msy_95 <- read_csv("manuscript_data/0.95msy.csv.xz")
+#frac_msy_100 <- read_csv("manuscript_data/1msy.csv.xz")
+frac_msy <- read_csv("manuscript_data/frac_msy_summaries.csv.xz")
 
+main_summaries <- read_csv("manuscript_data/comparison_3-2-var_300.csv.xz") %>%
+  mutate(strategy=recode(strategy, "PPO+GP"="PPO_GP")) %>%
+  arrange(strategy)
+
+esc_ends <- read_csv("manuscript_data/esc_ends_100.csv.xz")
+msy_ends <- read_csv("manuscript_data/msy_ends_100.csv.xz")
+ppo_ends <- read_csv("manuscript_data/ppo300_ends.csv.xz")
+gpp_ends <- read_csv("manuscript_data/gpp300_ends.csv.xz")
 ```
@@
 
+```{r}
+#| label: main
+#| fig.cap: "Reward distributions for the four strategies considered.
+These are based on 100 evaluation episodes.
+We denote CEsc for constant escapement, CMort for constant mortality, PPO for the output policy of the PPO optimization algorithm, and PPO GP for the Gaussian process interpolation of the PPO policy."
+#| out.width: "6in"
+main_summaries %>% 
+  ggplot() +
+  geom_violin(mapping = aes(x=reward,y=strategy,fill = strategy)) +
+  geom_boxplot(aes(x=reward,y=strategy, fill = strategy), width = 0.1, color = 'black', alpha=0.2)+
+  theme(
+    legend.position="none"
+  )
+```
+
 To assess what is the culprit for the classical strategies' low performance with respect to the DRL-based strategies, we plot the duration of each of the 100 evaluation episodes of each strategy.
@@
 \begin{align*}
-  \pi: (X_t, Y_t, Z_t)\mapsto (0.8 M_X^*.\ 0.8 M_Y^*),
+  \pi: (X_t, Y_t, Z_t)\mapsto (0.8 M_X^*,\ 0.8 M_Y^*),
 \end{align*}

### 2023-06-22 manuscript/manuscript.Rmd
added histogram and fractional MSY tradeoff

@@
 We evaluated each of the four control approaches considered -- MSY constant mortality, optimal constant escapement, PPO, and Gaussian process interpolation of the latter -- on each of the four scenarios examined (see Sec.
-TBD).
+3).
 Furthermore, we characterized the trade-off between economic output and sustainability faced by constant mortality policies.
@@
 
-Our main result is summarized in Fig. **TBD** which displays the total reward distributions for the policy obtained through each strategy.
+Our main result is summarized in Fig. \ref{fig:main} which displays the total reward distributions for the policy obtained through each strategy.
 Here we note two surprising things:
@@
 
-To assess what is the culprit for the classical strategies' low performance with respect to the DRL-based strategies, we plot the duration of each of the 100 evaluation episodes of each strategy.
+To assess what is the culprit for the classical strategies' low performance with respect to the DRL-based strategies, we plot the duration of each of the 100 evaluation episodes of each strategy in Fig.
+\ref{fig:ephistograms}.
 We see that early episode ends are a rather likely explanation for the wideness of constant escapement's wide reward distribution.
@@
 
-We examine the trade-off between profit and sustainability faced by classical strategies in Fig. **TBD**.
+```{r}
+#| label: ephistograms
+#| fig.cap: "Hisotgrams of episode lengths and rewards for the four different management strategies considered.
+Only the first 50 evaluation episodes (from a total of 100) were included, for ease of visualization."
+#| out.width: "6in"
+esc_len <- esc_ends %>%
+  filter(rep<50) %>%
+  ggplot(
+    mapping = aes(x = rep, y = t)
+  ) +
+  geom_col() +
+  labs(x="repetition", y="ep. duration")
+msy_len <- msy_ends %>%
+  filter(rep<50) %>%
+  ggplot(
+    mapping = aes(x = rep, y = t)
+  ) +
+  geom_col() +
+  labs(x="repetition", y="ep. duration")
+ppo_len <- ppo_ends %>%
+  filter(rep<50) %>%
+  ggplot(mapping = aes(x = rep, y = t)) +
+  geom_col() +
+  labs(x="repetition", y="ep. duration")
+gpp_len <- gpp_ends %>%
+  filter(rep<50) %>%
+  ggplot(
+    mapping = aes(x = rep, y = t)
+  ) +
+  geom_col() +
+  labs(x="repetition", y="ep. reward")
+
+esc_rew <- esc_ends %>%
+  filter(rep<50) %>%
+  ggplot(
+    mapping = aes(x = rep, y = reward)
+  ) +
+  geom_col() +
+  labs(x="repetition", y="ep. reward")
+msy_rew <- msy_ends %>%
+  filter(rep<50) %>%
+  ggplot(
+    mapping = aes(x = rep, y = reward)
+  ) +
+  geom_col() +
+  labs(x="repetition", y="ep. reward")
+ppo_rew <- ppo_ends %>%
+  filter(rep<50) %>%
+  ggplot(mapping = aes(x = rep, y = reward)) +
+  geom_col() +
+  labs(x="repetition", y="ep. reward")
+gpp_rew <- gpp_ends %>%
+  filter(rep<50) %>%
+  ggplot(
+    mapping = aes(x = rep, y = reward)
+  ) +
+  geom_col() +
+  labs(x="repetition", y="ep. duration")
+
+(esc_len | msy_len | ppo_len | gpp_len) / (esc_rew | msy_rew | ppo_rew | gpp_rew)
+```
+
+We examine the trade-off between profit and sustainability faced by the constant mortality strategy in Fig. \ref{fig:fracmsy}.
+We expect a similar and, possibly, more pronounced effect for the constant escapement strategy but do not analyze this case here.
 Here, two quantities are plotted: the fraction of evaluation episodes of maximal length (i.e. episodes with no near-extinction events) is and the average reward obtained.
@@
 }
+
+```{r}
+#| label: fracmsy
+#| fig.cap: "TBD"
+#| out.width: "6in"
+frac_msy %>%
+  ggplot(aes(x=msy_fraction)) + 
+  geom_line(aes(y=reward)) +
+  geom_errorbar(aes(ymin=reward-reward_std, ymax=reward+reward_std), width=.01,
+                 position=position_dodge(0.05)) +
+  geom_line(aes(y=num_full_eps)) +
+  geom_point(aes(y=reward)) + 
+  geom_point(aes(y=num_full_eps)) + 
+  scale_y_continuous(
+
+    # Features of the first axis
+    name = "% full-length eps",
+
+    # Add a second axis and specify its features
+    sec.axis = sec_axis(~.+20, name="Mean ep. reward")
+  )+
+  labs(x="% MSY constant mortality")
+```
 

### 2023-06-22 manuscript/manuscript.Rmd
finished frac msy tradeoff plot

@@
 #| label: fracmsy
-#| fig.cap: "TBD"
+#| fig.cap: "Trade-off between reward and probability of a near-extinction event for constant mortality policies.
+We evaluated policies at the full optimal constant escapement value, and also at 0.8, 0.9, 0.95 of the latter.
+Each evaluation is based on 100 episodes.
+On the left we plot the percentage of episodes which last their maximum time window, i.e.
+that do not see a near-extinction event.
+On the right, we plot the mean episdoe reward and standard deviation for each policy."
 #| out.width: "6in"
-frac_msy %>%
-  ggplot(aes(x=msy_fraction)) + 
-  geom_line(aes(y=reward)) +
+lengths <- frac_msy %>%
+  ggplot(aes(x=msy_fraction, y=num_full_eps)) +
+  geom_col(fill="darksalmon")+
+  labs(x="% MSY constant mortality", y="% full-length eps.") + 
+  coord_cartesian(ylim=c(70,101))
+
+rewards <- frac_msy %>%
+  ggplot(aes(x=msy_fraction, y=reward)) +
+  geom_line()+
+  geom_point()+
   geom_errorbar(aes(ymin=reward-reward_std, ymax=reward+reward_std), width=.01,
                  position=position_dodge(0.05)) +
-  geom_line(aes(y=num_full_eps)) +
-  geom_point(aes(y=reward)) + 
-  geom_point(aes(y=num_full_eps)) + 
-  scale_y_continuous(
+  labs(x="% MSY constant mortality", y="Mean ep. reward")
 
-    # Features of the first axis
-    name = "% full-length eps",
-
-    # Add a second axis and specify its features
-    sec.axis = sec_axis(~.+20, name="Mean ep. reward")
-  )+
-  labs(x="% MSY constant mortality")
+(lengths | rewards)
 ```

### 2023-06-22 manuscript/manuscript.Rmd
added policy plot

@@
 where $N_t=(X_t,Y_t,Z_t)$ is the state of the system at time $t$, and where $M_X$ and $M_Y$ are the mortalities due to harvest during that time-step.
-Here we visualize this policy function and provide an interpretation for it.
+Moreover, in the case of our ‘‘PPO+GP’’ management strategy, we sample $\pi_\theta$ on a grid of points $G={(X_i, Y_i, Z_i)}_i$ and use a Gaussian process to interpolate $\pi_\theta$, i.e.
+\begin{align}
+  \pi_{\text{PPO+GP}} = \text{Gaussian process interpolation of } \pi_\theta|_G.
+\end{align}
+This process was explained in detail in the Methods section.
+
+Here we visualize the resulting $\pi_{\text{PPO+GP}}$ policy and provide an interpretation for it.\footnote{
+A similar characterization of $\pi_\theta$ is found on the appendix. 
+Here we focus on the Gaussian process interpolation as it is smoother and, thus, easier to interpret.
+}
 Given its high dimension, it is not possible to fully display how the policy function obtained ‘‘looks like’’---we thus project it down to certain relevant axes.
-The result of this procedure is shown in Fig. **TBD**.
+The result of this procedure is shown in Fig. \ref{fig:gpp-policy}.
 In that figure, the shape of the optimal escapement strategy is provided for comparison.
+
+\begin{figure}
+  \centering
+  \includegraphics[scale=0.25]{figures/gpp-policy-pdf.pdf}
+  \caption{Plots of the PPO+GP policy $\pi_{\text{PPO+GP}}$ along several relevant axes.
+Here $M_X$ and $M_Y$ are the $X$ and $Y$ components of the policy function.
+The values of the plots are generated in the following way: For each variable $X$, $Y$, and $Z$, the times-eries of evaluation episodes are used to generate a window of typical values that variable attains when controlled by $\pi_{\text{PPO+GP}}$.
+Then, for each plot either $X$ or $Y$ was varied on $[0,1]$ along the $x$ axis, while the other variables (resp.
+$Y$ and $Z$, or $X$ and $Z$) were varied within the typical window computed before.
+The value of one of the latter two variables were visualized as color.}
+  \label{fig:gpp-policy}
+\end{figure}
 

### 2023-06-22 manuscript/manuscript.Rmd
included timeseries figure

@@
 
-The sensitivity of the policy to, e.g., the values of $Z$ can be seen in the sample time series displayed in Fig. **TBD**.
+This sensitivity of the policy to, for instance, the values of $Z$ can be seen in the sample time series displayed in Fig. \ref{fig:gpp-ep}.
 Here, we can see that species $Z$ becomes endangered due to harvesting for all management strategies.
 The DRL-based strategy, however, is sensitive to the values of $Z$ and can respond accordingly by scaling the fishing effort with the value of $Z$.
-In particular, the policy responds to periods of low values of $Z$ by restricting fishing on $X$ and $Y$, thus promoting $Z$'s growth.
+In particular, the policy responds to the period of diminishing values of $Z$ near the end of the episode, by restricting fishing on $X$ and $Y$, thus promoting $Z$'s growth.
 This pattern is rather common among the whole dataset---early episode ends are largely due to near-extinctions of $Z$ for all management strategies.
+
+\begin{figure}
+  \centering
+  \includegraphics[scale=0.33]{figures/timeseries-gpp.jpeg}
+  \caption{Time-series of an episode managed with $\pi_{\text{PPO+GP}}$.
+Here we plot the state of the system on the bottom panel, and the actions taken (fishing efforts on $X$ and $Y$---respectively $M_X$ and $M_Y$) on the top panel.}
+  \label{fig:gpp-ep}
+\end{figure}
 

### 2023-06-22 manuscript/manuscript.Rmd
mini

@@
 where $N_t=(X_t,Y_t,Z_t)$ is the state of the system at time $t$, and where $M_X$ and $M_Y$ are the mortalities due to harvest during that time-step.
-Moreover, in the case of our ‘‘PPO+GP’’ management strategy, we sample $\pi_\theta$ on a grid of points $G={(X_i, Y_i, Z_i)}_i$ and use a Gaussian process to interpolate $\pi_\theta$, i.e.
+Moreover, in the case of our ‘‘PPO+GP’’ management strategy, we sample $\pi_\theta$ on a grid of points $G=\{(X_i, Y_i, Z_i)\}_i$ and use a Gaussian process to interpolate $\pi_\theta$, i.e.
 \begin{align}

### 2023-06-22 manuscript/manuscript.Rmd
added single species policy, and parametric noise plots
caption for stability result

@@
 We give evidence for this intuition by showing that our DRL method recovers a constant escapement solution when trained.
-These results are shown in Fig. TBD.
+These results are shown in Fig. \ref{fig:1sp-pol}
+Here we show both the output PPO policy, and its Gaussian process interpolation.
+This helps build an intuition about the relationship between our ‘‘PPO’’ and ‘‘PPO+GP’’ management strategies.
+
+\begin{figure}
+  \centering
+  \includegraphics[scale=0.25]{figures/1sp-policy.jpeg}
+  \caption{Left panel: the policy obtained from 100 training iterations of the PPO algorithm on the ‘‘single species, single fishery’’ model.
+Right panel: the Gaussian process interpolation of the left panel.
+We plot both as scatter data evaluated on a 101-point grid on $[0,1]$, but these policies may of course be evaluated continuously---on any possible value of $X$.}
+  \label{fig:1sp-pol}
+\end{figure}
 
@@
 
-Our main result in this respect is Fig. **TBD**.
+Our main result in this respect is Fig. \ref{fig:stability}.
 There, we plot the *median episode reward difference* between the two DRL-based methods we considered, and the optimal constant escapement strategy.
 This figure shows that, for a wide range of parameter values, DRL-based strategies can have a considerable advantage over an optimized constant escapement policy (the single-species optimal solution).
+
+\begin{figure}
+  \centering
+  \includegraphics{figures/jiggle_v1.png}
+  \caption{Mean reward difference between DRL methods (resp.
+our ‘‘PPO’’ and ‘‘PPO+GP’’ strategies), on the one hand, and the optimal constant escapement policy on the other.
+The dynamic parameters in eq.
+\eqref{eq:3d model} were randomly perturbed from the values given in \eqref{eq:param vals} according to the procedure detailed in Sec.
+4.
+The noise strength of this perturbation is plotted as the x-axis.
+For each noise strength, 100 parameter perturbations were sampled---each one giving rise to a \emph{realization} of the model.
+For each such realization, we optimized a constant escapement policy and trained a PPO agent.
+Moreover, we interpolated the PPO policy using a Gaussian process, as detailed in Sec 4.
+Then, for each realization we compared the performance of these policies: we measured the mean reward difference betwen PPO and constant escapement, and between PPO+GP and constant escapement.
+The plot represents the distribution of reward differences observed at a given noise strength---we plot the mean and the standard deviation of the mean reward differences observed.
+In an equation, we plot the means $\mathbb{E}_{P}[\mu_P^{\text{DRL}}-\mu_P^{\text{CEsc}}]$, where $P$ are the paramter values, $\mu_P^{\text{DRL}}$ is the mean reward for a DRL policy trained on the problem with parameter values $P$, and, similarly, $\mu_P^{\text{CEsc}}$ is the mean reward of the optimal constant escpament policy for parameter values $P$.}
+  \label{fig:stability}
+\end{figure}
 

### 2023-06-22 manuscript/manuscript.Rmd
erased old results section

@@
 
-## Old results (to be copy-pasted above)
-
-In Fig. \ref{fig:timeseries} we show the time series of one such repetition.
-The left panel uses an MSY constant mortality control $M^*X$, the middle panel uses $0.8 M^* X$, while the right panel is controlled by the trained DRL agent.
-We see that that optimal constant mortality control leads to a near-extinction event at around $t=150$.
-The constant mortality policy at $0.8 M^*$ is already conservative enough to support a sustainable path.
-There, the population levels are maintained roughly in a stationary state.
-Finally, the DRL time series shows an oscillatory behavior, with a slight dephase between the $Y$ and $Z$ curves, while the $X$ oscillations are roughly anticorrelated with $Z$.
-
-This time series is representative of the patterns observed in the data (see App. A.1).
-The episode reward and episode length statistics are shown in Fig. \ref{fig:rew_len_fig}.
-There, we highlight two things:
-First, that the tendency towards early episode ends is fixed by the conservative approach of fishing at 80% of the MSY mortality rate.
-(In App. A.2 we discuss and evaluate other policies using a fixed-percentage of the MSY mortality rate.)
-Second, that while the conservative 80% approach has a stable behavior, it consistently produces smaller harvests than a DRL generated policy.
-
-To understand the DRL-derived policy, Fig. \ref{fig:action-X} depicts the harvest as a function of $X$.
-Specifically, for each point $(x,y,z)$ visited by one of the aforementioned 50 evaluation episodes, we evaluate the harvest collected by the DRL agent and we subsequently plot the scatter data as a function of the projection $x$ onto the $X$ axis.
-In panel A we have used color to add the a $Y$-dependence, while in panel B we did the same with $Z$-dependence.
-This shows that the $Z$ and $Y$ dependence of the policy is rather weak and in fact it has roughly the same shape as an escapement policy.
-
-Fig. \ref{fig:action-X} additionally shows (pink line) the optimal constant escapement policy and the MSY constant mortality policy (green line).
-The optimal constant escapement was computed in a manner similar to how the MSY constant mortality was computed: for each escapement in a grid of 100 values between 0 and 1, 50 episodes were run using the corresponding escapement policy.
-The value with the best mean episode reward was chosen as the optimal escapement value.
-We observe that the DRL policy approximates the optimal constant escapement policy, with the difference of being more conservative in its harvest at high $X$ values.
-This pattern can be seen in panels (C-D), where at high $X$ values, the DRL escapement mildly increases in a non-linear fashion rather than staying constant.
-Finally, relative to these two policies, we observe that MSY constant mortality underfishes at high $X$ values and overfishes at low $X$ values.
-
-In Fig. \ref{fig:state-space}, we perform a back to back visualization of all the policies evaluated---MSY constant mortality, 80% MSY constant mortality, DRL and optimal constant escapement.
-The policies are computed at each state along the paths they traverse in the 50 evaluation rounds, and the values are projected on the $X$-$Z$ plane for visualization.
-This figure highlights the similarity between the policy found by the DRL agent and the optimal escapement policy.
-
-Notice the anti-correlation between $X$ and $Y$ along the paths, this is especially prominent for the constant mortality plots.
-This anti-correlation is, furthermore, observed in the time series plots (see Fig. \ref{fig:timeseries} and App. A.1).
-
 # Discussion

### 2023-06-22 manuscript/manuscript.Rmd
added table of models

@@
 **TBD CONCEPTUAL**, we consider four different models of the ecosystem with which the fishery interacts, each of this is called an *environment*, alluding to the environment in an RL setting.
+We summarize these models in Table \ref{tab:models}.
 To recap, these four environments are a single-species growth model \eqref{eq:may} for a harvested species; a three-species model \eqref{eq:3d model stoch} with a single harvested species; the same three-species model but with *two* harvested species; and, finally, a three-species model with a time-varying parameter and two harvested species.
@@
 The policies explored were functions from states $N_t$ to either a single number $M_X(N_t) = \pi(N_t)$, or a pair of numbers $(M_X(N_t), M_Y(N_t)) = \pi(N_t)$ depending on the environment considered.
-\textbf{Probably a table showing the different combinations of species and fishery numbers would be good, possibly giving these ‘scenarios’ names.}
+
+\begin{table}
+  \begin{center}
+    \begin{tabular}{|l|c|c|c|c|}
+      \hline
+      \textbf{Model name} & \textbf{Model eqs.} & \textbf{N. Sp.} & \textbf{Harv. Sp.} & \textbf{Stationary?}\\
+      \hline
+      1 species, 1 fishery & \eqref{eq:may} & 1 & $X$ & Yes \\
+      3 species, 1 fishery & \eqref{eq:3d model} & 3 & $X$ & Yes \\
+      3 species, 2 fishery & \eqref{eq:3d model} & 3 & $X$ and $Y$ & Yes \\
+      3 species, 2 fishery, time-varying parameters & \eqref{eq:3d model} & 3 & $X$ and $Y$ & No \\
+      \hline
+    \end{tabular}
+    \caption{Table of models considered in this paper.
+Here, \textbf{N.
+Sp.} is the \emph{number of species} in the model, \textbf{Harve.
+Sp.} is the species of the model which are harvested, and \textbf{Stationary?} refers to whether the parameters of the model have fixed values (or, on the contrary, if they vary in time).
+The only non-stationary case presented in the paper is where $r_X$ drifts linearly with time.
+In the code repository associated to the paper, we consider other possible choices of non-stationarity.}
+    \label{tab:models}
+  \end{center}
+\end{table}
 

### 2023-06-22 manuscript/manuscript.Rmd
added new conceptual fig (missing the values of performances)

@@
 
+<!--
 ```{r}
@@
 ```
+-->
+
+```{r}
+#| label: conceptual-v2
+#| echo: FALSE
+#| message: FALSE
+#| fig.align: 'center'
+#| fig.cap:  'An experimental-design type of visualization of the management scenarios considered in this paper. 
+#| On the x-axis are four different fishery management problems: a single species model, and three three-species models.
+#| From left to right, the latter are: three species out of which one, X, is harvested, and all parameters are constant in time; 
+#| three species out of which two, X and Y, are harvested and all parameters are constant in time;
+#| and finally three species out of which two, X and Y, are harvested and one parameter depends on time. 
+#| We represent the latter’s time-dependence with a clock next to the X variable, and we intend to use it as an example of a possible simplified model for the effects of climate change.
+#| On the y-axis we have different management strategies with which one may control each of the models. 
+#| On the bottom we have the constant mortality strategy (CMort), where one optimizes over constant fishing effort strategies.
+#| Above that is the constant escapement strategy (CEsc), based on calling off all fishing below a certain threshold population value.
+#| Finally, on top we have DRL-based strategies where policies are in general functions of the full state of the system, and they are parametrized by a neural network.'
+#| out.width: '6.5in' 
+#| fig.pos: 'H'
+
+knitr::include_graphics("figures/conceptual-v2.jpeg")
+```
 
@@
 The data used is found in the ‘‘data’’ subdirectory, but the user may use the code provided to generate new data sets.} 
-We consider the problem of devising harvest strategies in a for a series of ecosystem models (Figure **TBD**):
+We consider the problem of devising harvest strategies in a for a series of ecosystem models (Figure \ref{fig:conceptual-v2}):
 *1) one species, one fishery:* a simple single-species recruitment model based on [@may77];

### 2023-06-22 manuscript/manuscript.Rmd
parameter values and rx(t)-climate change comments

@@
 \end{align*}
-In our simulations, we use $c_{XY} = d = f = 0.1$ and $\sigma_Z=0.05$ so, in broad strokes, our three-species model is a slight perturbation of the single-species model.
+Throughout most of our paper, we use the parameter values given in eq. \eqref{eq:param vals}.
+In particular, $c_{XY} = d = f = 0.1$ and $\sigma_Z=0.05$ so, in broad strokes, our three-species model is a slight perturbation of the single-species model.
 Put differently, we are in a regime where, just by examining parameters, it would seem that the single-species model is a reasonable approximation.
-**Is this clear? Should I maybe erase these last sentences?**
+**Is this actually true? Should I maybe erase these last sentences?**
 
@@
 *1. Three species, one fishery:* in this first model, only $X$ is harvested, that is, in the notation of eq. \eqref{eq:general model}, we fix $M_Y=M_Z=0$ and leave $M_X$ as a control variable.
-All parameters here are constant.
+All parameters here are constant, with values given by \eqref{eq:param vals}.
 *2. Three species, two fisheries:* as the previous case but with $X$ and $Y$ being harvested. 
@@
 \end{equation}
-This is intended to reflect in a simple manner a possible effect of climate change: where the reproductive rate of $X$ is reduced linearly over time until it stabilizes.
-In future work we intend to capture more realistic models of the effects of climate change, but for this exploratory work we have chosen this simple expression for the sake of conceptual clarity.
+All other parameters are fixed by eq. \eqref{eq:param vals}.
+
+Eq. \eqref{eq:rx(t)} is intended to reflect in a simple manner a possible effect of climate change: where the reproductive rate of $X$ is reduced linearly over time until it stabilizes.
+In future work we intend to test the DRL framework in realistic models of particular fisheries.
+As a part of that effort, we will attempt capture more detailed models of the effects of climate change.
+For this exploratory work we have chosen this simple expression, however, for the sake of conceptual clarity.
 

### 2023-06-22 manuscript/manuscript.Rmd
mini

@@
 As seen in Fig.
-**TBD CONCEPTUAL**, we consider four different models of the ecosystem with which the fishery interacts, each of this is called an *environment*, alluding to the environment in an RL setting.
+\ref{fig:conceptual-v2}, we consider four different models of the ecosystem with which the fishery interacts, each of this is called an *environment*, alluding to the environment in an RL setting.
 We summarize these models in Table \ref{tab:models}.
-To recap, these four environments are a single-species growth model \eqref{eq:may} for a harvested species; a three-species model \eqref{eq:3d model stoch} with a single harvested species; the same three-species model but with *two* harvested species; and, finally, a three-species model with a time-varying parameter and two harvested species.
+To recap, these four environments are a single-species growth model \eqref{eq:may} for a harvested species; a three-species model \eqref{eq:3d model} with a single harvested species; the same three-species model but with *two* harvested species; and, finally, a three-species model with a time-varying parameter and two harvested species.
 
@@
 Here, \textbf{N.
-Sp.} is the \emph{number of species} in the model, \textbf{Harve.
+Sp.} is the \emph{number of species} in the model, \textbf{Harv.
 Sp.} is the species of the model which are harvested, and \textbf{Stationary?} refers to whether the parameters of the model have fixed values (or, on the contrary, if they vary in time).

### 2023-06-22 manuscript/manuscript.Rmd
mini

@@
 
-The three-species model's (eq. \eqref{eq:3d model stoch}) dynamic parameters were chosen as follows
+The three-species model's (eq. \eqref{eq:3d model}) dynamic parameters were chosen as follows
 \begin{align}
@@
   \centering
-  \includegraphics{figures/jiggle_v1.png}
+  \includegraphics[scale=0.8]{figures/jiggle_v1.png}
   \caption{Mean reward difference between DRL methods (resp.
-our ‘‘PPO’’ and ‘‘PPO+GP’’ strategies), on the one hand, and the optimal constant escapement policy on the other.
+our ‘‘PPO’’ and ‘‘PPO+GP’’ strategies), on the one hand, and the optimal constant escapement policy (‘‘CEsc’’) on the other.
 The dynamic parameters in eq.

### 2023-06-22 manuscript/manuscript.Rmd
single species comments on jitter

@@
 
+There is a presence of certain high-mortality points at low $X$ values in the PPO policy (which in turn generates a rising fishing mortality for $X$ values below a certain threshold in the PPO+GP policy).
+This is likely due to experience of near-extinctions early on in the training process---where, given an impending extinction, it is more rewardful to fish intensively. 
+These ‘‘jitters’’ are likely not fully erased through the optimization algorithm since near-extinction events become extremely rare after only a few training iterations.
+This way, the agent does not further explore that region of state space to generate new experience.
+We believe the most important aspect of the constant escapement policy reproduced by PPO is the fact that there is some sufficiently-wide window below the threshold of the policy (i.e.
+below the optimal escapement value), on which no fishing is performed.
+That is, there exists some sufficiently large $\varepsilon$ such that, if $X_{\text{thresh.}}$ is the optimal escapement value of the system, then $\pi_{PPO}(X)=0$ for all $X\in[X_{\text{thresh.}}-\varepsilon,\ X_{\text{thresh.}}]$.
+
 \begin{figure}

### 2023-06-22 manuscript/manuscript.Rmd
mini

@@
 Here $M_X$ and $M_Y$ are the $X$ and $Y$ components of the policy function.
-The values of the plots are generated in the following way: For each variable $X$, $Y$, and $Z$, the times-eries of evaluation episodes are used to generate a window of typical values that variable attains when controlled by $\pi_{\text{PPO+GP}}$.
+The values of the plots are generated in the following way: For each variable $X$, $Y$, and $Z$, the time-series of evaluation episodes are used to generate a window of typical values that variable attains when controlled by $\pi_{\text{PPO+GP}}$.
 Then, for each plot either $X$ or $Y$ was varied on $[0,1]$ along the $x$ axis, while the other variables (resp.

### 2023-06-23 manuscript/manuscript.Rmd
started discussion

@@
 Furthermore, rather than computing the optimal control policy for the model (something that is impossible in practice for complex systems) we use deep reinforcement learning to obtain a ‘‘pretty darn good’’ policy.
-This policy is estimated in a *model-free* setting, i.e., the agent treats the dynamical model (e.g. eq. \eqref{eq:3d model stoch}) as a black box of input-output pairs.
+This policy is estimated in a *model-free* setting, i.e., the agent treats the dynamical model (e.g. eq. \eqref{eq:3d model}) as a black box of input-output pairs.
 By not relying on precise knowledge of the model's parameter values, but rather just relying on input-output statistics, model-free approaches have gained traction in a variety of control theory settings (see, e.g., [@sato2019; @ramirez2022; @zeng2019]).
 
-We compare deep reinforcement learning-based policies against optimal constant mortality policies. 
-While the latter are MSY-inspired, they are optimized in a model-free way as well: the optimal constant mortality rate is computed empirically from simulated data.
-
-Because of the simplicity of the constant mortality ansatz, the optimal such policy may be easily estimated (see Figure \ref{fig:tuning}).
+We compare deep reinforcement learning-based policies against classical management strategies (optimal constant mortality and optimal constant escapement policies policies). 
+While the latter are inspired by the shape of optimal solutions in the single-species setting, they are optimized in a model-free way as well: e.g.
+the optimal constant mortality rate is computed empirically from simulated data.
+
+Because of the simplicity of the classical policy functions, the optimal such policy may be easily estimated through a grid-search.
+This is the case since these policy functions are only specified by one or two parameters (in the single fishery, or two fishery cases).
 In contrast, DRL optimizes over the more expressive -- and more complicated -- family of policies parametrized by a neural network.
 Because of this, our DRL approach uses a variant of *gradient ascent* (the proximal policy optimization algorithm) to obtain a good candidate policy.
+
+We showed that for sufficiently complex management scenarios---with multiple harvested species and, possibly, time-varying parameters---, optimal constant escapement can perform poorly both in terms of reward (which may be used as a proxy for economic performance), as in terms of sustainability (i.e.
+probability of a near-extinction).
+In the case we investigated in detail, the optimal constant mortality policy surprisingly performs much better than constant escapement.
+We nonetheless observe that the constant mortality strategy is faced with a trade-off: higher sustainability is achieved for sub-optimal constant mortality rates, but only at a significant decrease in the mean episode reward.
+We expect that with increasing complexity this phenomenon might become more pronounced.
+This is because of the rigidity of classical strategies: the simplicity of their expressions, depending only on a few parameters, means that policy optimization is constrained to a rather reduced subset of the space of possible policies.
+
+For the most complicated ecosystem models we considered (those with three species, two of which are harvested) we saw a significant improvement in both average reward and sustainability when using DRL-based strategies.
+In particular, there was a stark difference in performance between both DRL-based strategies tested, and constant escapement---the optimal solution for a wide array of single-species models.
+This, in addition, within a regime of parameter space where species interaction terms about an order of magnitude smaller than the single-species logistic growth---i.e., in a regime where logistic growth could be considered a reasonable approximation of the system.
+
+# Old discussion
 

### 2023-06-26 manuscript/manuscript.Rmd
mini
added a bit discussion, fixed some typos

--- 2023-06-22 manuscript/manuscript.Rmd
@@
 Take the mean of this average difference over the 100 perturbed parameter samples.
-The stability results are displayed in Fig. \textbf{TBD}.
 
@@
 Furthermore, rather than computing the optimal control policy for the model (something that is impossible in practice for complex systems) we use deep reinforcement learning to obtain a ‘‘pretty darn good’’ policy.
-This policy is estimated in a *model-free* setting, i.e., the agent treats the dynamical model (e.g. eq. \eqref{eq:3d model stoch}) as a black box of input-output pairs.
+This policy is estimated in a *model-free* setting, i.e., the agent treats the dynamical model (e.g. eq. \eqref{eq:3d model}) as a black box of input-output pairs.
 By not relying on precise knowledge of the model's parameter values, but rather just relying on input-output statistics, model-free approaches have gained traction in a variety of control theory settings (see, e.g., [@sato2019; @ramirez2022; @zeng2019]).
 
-We compare deep reinforcement learning-based policies against optimal constant mortality policies. 
-While the latter are MSY-inspired, they are optimized in a model-free way as well: the optimal constant mortality rate is computed empirically from simulated data.
-
-Because of the simplicity of the constant mortality ansatz, the optimal such policy may be easily estimated (see Figure \ref{fig:tuning}).
+We compare deep reinforcement learning-based policies against classical management strategies (optimal constant mortality and optimal constant escapement policies policies). 
+While the latter are inspired by the shape of optimal solutions in the single-species setting, they are optimized in a model-free way as well: e.g.
+the optimal constant mortality rate is computed empirically from simulated data.
+
+Because of the simplicity of the classical policy functions, the optimal such policy may be easily estimated through a grid-search.
+This is the case since these policy functions are only specified by one or two parameters (respectively in the single fishery, or two fishery cases).
 In contrast, DRL optimizes over the more expressive -- and more complicated -- family of policies parametrized by a neural network.
 Because of this, our DRL approach uses a variant of *gradient ascent* (the proximal policy optimization algorithm) to obtain a good candidate policy.
+
+We showed that for sufficiently complex management scenarios -- with multiple harvested species and, possibly, time-varying parameters -- DRL-based management strategies perform significantly better than constant escapement.
+This with respect to both average rewards received and conservation goals.
+In this sense, an approximate solution to a more complicated and expressive model, can outperform the optimal solution of the single-species problem---even when the parameters of the single-species solution are empirically optimized.
+
+What is more, this DRL advantage happened within a regime of parameter space where species interaction terms were much smaller than the single-species logistic growth terms.
+Indeed, interaction parameters in \eqref{eq:3d model} are about an order of magnitude smaller than $r_X$ and $r_Y$.
+In this situation logistic growth is a reasonable approximation of the system, and thus a constant escapement solution is a plausible control strategy.
+
+We found that the optimal constant mortality policy surprisingly performs much better than constant escapement (Fig. \ref{fig:main}).
+However, it can be observed in Fig.
+\ref{fig:fracmsy} that the constant mortality strategy faces a trade-off: high sustainability is achieved for sub-optimal constant mortality rates, but only at a significant decrease in the mean episode reward.
+We expect that with increasing ecosystem complexity this phenomenon might become more pronounced.
+We can understand this as a consequence of the rigidity of classical strategies: the simplicity of their expressions, depending only on a few parameters, means that policy optimization is constrained to a rather reduced subset of the space of possible policies.
+
+Finally, we performed a stability analysis to ensure that the advantage of DRL-based techniques over constant escapement is a ubiquitous phenomenon and not a result of a lucky selection of parameter values.
+We found (in Fig. \ref{fig:stability}) that an advantage can be observed even for relatively high-noise perturbations of the parameters---noise with a variance of 20% of the parameter values.
+The aforementioned Figure summarizes the statistics of 100 parameter perturbations so that we may expect perturbations of up to 60% (i.e.
+three sigmas) in the parameter values to appear in these statistics.
+
+# Old discussion
 

### 2023-06-26 manuscript/manuscript.Rmd
erased old discussion

@@
 
-# Old discussion
-
-We have shown that the DRL agent finds an approximately *bang-bang* type of solution (see Text Box 1 for the definition of bang-bang solutions and Figs.
-\ref{fig:action-X} and \ref{fig:state-space} for the result).
-While bang-bang solutions are known to be optimal for a wide variety of single-species fishery control problems [@hening2019; @hening2021], there is no such proof of optimality for general multi-species models.
-In this sense this example is rather surprising and reassuring: while no general guarantee points at the optimality of bang-bang solutions for multi-species model, the existence of one such example provides a benchmark for the use of DRL in fishery control.
-
-One intuition for why a constant escapement policy performs so well in such a model comes from the anti-correlation between $X$ and $Y$ noted in the results section.
-We have observed this pattern arise in a variety of policies, both classical as well as some obtained by training DRL agents.
-In fact, ‘‘crashes’’ in the $X$ population are almost invariably shortly preceeded by dramatic increases in $Y$.
-We hypothesize that this effectively leads to a dimensionality reduction: once the value of $X$ is known, along a path traced by the control system, then $Y$ may be roughly predicted.
-A similar, although weaker, anti-correlation is also observed between $X$ and $Z$.
-These two informal observations point at an effective simplification of the prediction process: the values of $Y$ and $Z$ are much less impactful to policy decisions than the value of $X$ itself.
-We have not yet pushed further to establish a more precise picture of this dynamical simplification.\footnote{
-A hand-waving way to understand this phenomenon is the following:
-The $X$-$Y$ anti-correlation can be associated to the Lotka-Volterra competition term between $X$ and $Y$, i.e. $c_{XY}XY$ in eq. \eqref{eq:3d model stoch}.
-Now, consider a fall in population $X$.
-Because of the aforementioned anti-correlation with $Y$, this is accompanied by a growth in $Y$.
-We may understand this, roughly, as a \emph{replacement} of $X$ for $Y$.
-But then, because the parameter $D = 1.1 > 1$, this increases the effective birth rate $Z$, i.e. the term $f(X+DY)Z$ in eq. \eqref{eq:3d model stoch}.
-This, in turn, leads to a growth of $Z$ and, thus, an anti-correlation between $Z$ and $X$.}
-
-This manuscript functions as a first step in the process of incorporating DRL into fishery decision making.
-We have shown that highly non-linear systems may give rise to simple, good policies, and that these may be found via reinforcement learning.
-Moreover, the framework allows for the possibility that these simplifications do not in fact happen, and that good harvest policies are highly non-linear, multi-parameter functions.
-
-In this regard, DRL provides a framework flexible enough to capture the intricate population dynamics that are relevant in fishery management.
-Importantly, DRL is equipped with tools to deal with complex models for which simple policies might yield unsatisfactory results (and thus more refined techniques are needed to control them).
-For example, tools such as *long short-term memories* [@lstm], *curriculum learning* [@curriculum-learning] and *open-ended learning* [@open-ended] can be incorporated when faced with parametric noise or drift, or generally uncertainties about the agent's environment.
-These techniques can be used to obtain agents who perform well over a variety of different environments.
-<!-- ML: Could also use LSTMs for parametric drift.
-Might be better to use a different selling point for curriculum learning like if you want to have a bit more generalizability in your RL agent.
--->
-
-Finally, in our study case, we have shown that the ‘‘rest periods’’ -- periods of several time steps with no fishing at all -- prominent in both the DRL solution and in the optimal bang-bang policy, may be important to maintain the sustainability of the fishery and to increase its long-term profits.
-This hypothesis could be explored in more generality as it might apply to a wide variety of models.
-
 ## Acknowledgements

### 2023-06-26 manuscript/manuscript.Rmd
added esc_heatmap

@@
 For each one of these choices of a constant mortality rate, say $M_X$, we simulated 100 episodes based on \eqref{eq:3d model stoch}:
-At each time step the state $(X_t, Y_t, Z_t)$ was observed, a harvest of $M_X X_t$ was collected and then the system evolved according to its natural dynamics \eqref{eq:3d model stoch}.
+At each time step the state $(X_t, Y_t, Z_t)$ was observed, a harvest of $M_X X_t$ was collected and then the system evolved according to its natural dynamics \eqref{eq:3d model}.
 The optimal mortality rate was the value $M_X^*$ for which the mean episode reward was maximal.
-The result of the tuning procedure is shown in Fig. \ref{fig:tuning}.
 
@@
 The optimal policy was the policy with the highest average reward obtained.
+A visualization of the tuning procedure for the ‘‘three species, two fisheries, time-varying parameter’’ model is shown in Fig. \ref{fig:tuning}.
+
+\begin{figure}
+  \centering
+  \includegraphics[scale=0.5]{figures/esc_heatmap.png}
+  \caption{Visualization of the cosntant escapement strategy tuning procedure for the ‘‘three-species, two-fishery, time-varying parameter’’ model.
+There was a certain multiplicity in this tuning strategy: a ‘‘ridge of optimality’’ where policies had essentially equivalent behavior.
+Throughout our investigation, we tuned constant escapment in several occasions and, on each occassion, a different optimal policy along the ridge was found.
+The results for different policies along the ridge were in practice equivalent, with no discernible difference in performance.
+We highlighted the ridge with a white dotted line.}
+  \label{fig:tuning}
+\end{figure}
 

### 2023-06-26 manuscript/manuscript.Rmd
mini

@@
       3 species, 2 fishery & \eqref{eq:3d model} & 3 & $X$ and $Y$ & Yes \\
-      3 species, 2 fishery, time-varying parameters & \eqref{eq:3d model} & 3 & $X$ and $Y$ & No \\
+      3 species, 2 fishery, non-stationary & \eqref{eq:3d model} & 3 & $X$ and $Y$ & No \\
       \hline
@@
 Since the approach for tuning is completely analogous for all scenarios considered, here we focus on the ‘‘3 species, 1 fishery’’ case.
-For each one of these choices of a constant mortality rate, say $M_X$, we simulated 100 episodes based on \eqref{eq:3d model stoch}:
+For each one of these choices of a constant mortality rate, say $M_X$, we simulated 100 episodes based on \eqref{eq:3d model}:
 At each time step the state $(X_t, Y_t, Z_t)$ was observed, a harvest of $M_X X_t$ was collected and then the system evolved according to its natural dynamics \eqref{eq:3d model}.
@@
 The optimal policy was the policy with the highest average reward obtained.
-A visualization of the tuning procedure for the ‘‘three species, two fisheries, time-varying parameter’’ model is shown in Fig. \ref{fig:tuning}.
+A visualization of the tuning procedure for the ‘‘three species, two fisheries, non-stationary’’ model is shown in Fig. \ref{fig:tuning}.
 
@@
   \includegraphics[scale=0.5]{figures/esc_heatmap.png}
-  \caption{Visualization of the cosntant escapement strategy tuning procedure for the ‘‘three-species, two-fishery, time-varying parameter’’ model.
+  \caption{Visualization of the cosntant escapement strategy tuning procedure for the ‘‘three-species, two-fishery, non-stationary’’ model.
 There was a certain multiplicity in this tuning strategy: a ‘‘ridge of optimality’’ where policies had essentially equivalent behavior.

### 2023-06-26 manuscript/manuscript.Rmd
mini

@@
 The state space used was normalized case-by-case as follows: 
-*1)* for the single species model, a line segment $[0,1]$, *2)* for all three-species models, a cute $[0,1]^{\times 3}$.
+*1)* for the single species model, a line segment $[0,1]$, *2)* for all three-species models, a cube $[0,1]^{\times 3}$.
 We used simulation data to derive a bound on the population sizes typically observed, and thus be able to normalize states to a finite volume.

### 2023-06-26 manuscript/manuscript.Rmd
added other cases considered as appendix

@@
 
+## Results for stationary models
+
+In the main text we focused on the non-stationary model (‘‘three species, two fisheries, non-stationary’’ in Table 1) for the sake of space and because our results were most compelling there.
+Here we present the reward distributions for the other models considered.
+```{r}
+summary_32 <- read_csv("manuscript_data/comparison_3-2-const_300.csv.xz") %>%
+  mutate(strategy=recode(strategy, "PPO+GP"="PPO_GP")) %>%
+  arrange(strategy)
+summary_31 <- read_csv("manuscript_data/comparison_3-1_300.csv.xz") %>%
+  mutate(strategy=recode(strategy, "PPO+GP"="PPO_GP")) %>%
+  arrange(strategy)
+summary_11 <- read_csv("manuscript_data/comparison_1-1_100.csv.xz") %>%
+  mutate(strategy=recode(strategy, "PPO+GP"="PPO_GP")) %>%
+  arrange(strategy)
+```
+
+```{r}
+#| label: main_32
+#| fig.cap: "Reward distributions for the four strategies considered.
+These are based on 100 evaluation episodes of the ‘‘three species, two fisheries’’ model in Table 1.
+We denote CEsc for constant escapement, CMort for constant mortality, PPO for the output policy of the PPO optimization algorithm, and PPO GP for the Gaussian process interpolation of the PPO policy."
+#| out.width: "6in"
+summary_32 %>% 
+  ggplot() +
+  geom_violin(mapping = aes(x=reward,y=strategy,fill = strategy)) +
+  geom_boxplot(aes(x=reward,y=strategy, fill = strategy), width = 0.1, color = 'black', alpha=0.2)+
+  theme(
+    legend.position="none"
+  )
+```
+
+```{r}
+#| label: main_31
+#| fig.cap: "Reward distributions for the four strategies considered.
+These are based on 100 evaluation episodes of the ‘‘three species, one fishery’’ model in Table 1.
+We denote CEsc for constant escapement, CMort for constant mortality, PPO for the output policy of the PPO optimization algorithm, and PPO GP for the Gaussian process interpolation of the PPO policy."
+#| out.width: "6in"
+summary_31 %>% 
+  ggplot() +
+  geom_violin(mapping = aes(x=reward,y=strategy,fill = strategy)) +
+  geom_boxplot(aes(x=reward,y=strategy, fill = strategy), width = 0.1, color = 'black', alpha=0.2)+
+  theme(
+    legend.position="none"
+  )
+```
+
+```{r}
+#| label: main_11
+#| fig.cap: "Reward distributions for the four strategies considered.
+These are based on 100 evaluation episodes of the ‘‘one species, one fishery’’ model in Table 1.
+We denote CEsc for constant escapement, CMort for constant mortality, PPO for the output policy of the PPO optimization algorithm, and PPO GP for the Gaussian process interpolation of the PPO policy."
+#| out.width: "6in"
+summary_11 %>% 
+  ggplot() +
+  geom_violin(mapping = aes(x=reward,y=strategy,fill = strategy)) +
+  geom_boxplot(aes(x=reward,y=strategy, fill = strategy), width = 0.1, color = 'black', alpha=0.2)+
+  theme(
+    legend.position="none"
+  )
+```
+
+## Old appendices:
+
 ## Collected time series data

### 2023-06-26 manuscript/manuscript.Rmd
added ppo policy appendix

@@
 
-Policies obtained from the PPO algorithm can be ‘‘noisy’’ as their optimization algorithm is randomized (see, e.g. App. \textbf{TBD} for visualizations of these policies).
+Policies obtained from the PPO algorithm can be ‘‘noisy’’ as their optimization algorithm is randomized (see, e.g.
+App.
+B for visualizations of the PPO policy in the ‘‘3 species, 2 fisheries, non-stationary’’ case).
 We perform some interpolation using a Gaussian process regressor in order to de-noise these policies.
@@
 
-We will visualize the results concerning the last scenario considered (‘‘3 species, 2 of which harvested, and a time-varying parameter’’) in this section, leaving the other cases for the appendix.
+We will visualize the results concerning the last scenario considered (‘‘3 species, 2 of which harvested, and a time-varying parameter’’) in this section, leaving the other cases for App. A.
 This is the most complex scenario considered and where our results show the most compelling advantage of DRL methods with respect to classical strategies.
@@
 In the main text we focused on the non-stationary model (‘‘three species, two fisheries, non-stationary’’ in Table 1) for the sake of space and because our results were most compelling there.
-Here we present the reward distributions for the other models considered.
+Here we present the reward distributions for the other models considered---the three stationary models, lines 1-3 in Table 1.
+
 ```{r}
@@
 
+## PPO policy function for non-stationary model
+
+In the main text, Fig. \ref{fig:gpp-policy}, we presented a visualization of the PPO+GP policy function obtained for the ‘‘three species, two fisheries, non-stationary’’ model.
+This policy function is a Gaussian process regression of scatter data of the PPO policy function.
+In Fig. \ref{fig:ppo-policy} we present a representation of this scatter data in a similar format as Fig. \ref{fig:gpp-policy}.
+
+\begin{figure}
+  \centering
+  \includegraphics[scale=0.35]{figures/ppo-policy.jpeg}
+  \caption{Plots of the PPO policy $\pi_{\text{PPO}}$ along several relevant axes.
+Here $M_X$ and $M_Y$ are the $X$ and $Y$ components of the policy function.
+The values of the plots are generated in the following way: For each variable $X$, $Y$, and $Z$, the time-series of evaluation episodes are used to generate a window of typical values that variable attains when controlled by $\pi_{\text{PPO}}$.
+Then, for each plot either $X$ or $Y$ was evaluated on 100 values in $[0,1]$ along the $x$ axis, while the other variables (resp.
+$Y$ and $Z$, or $X$ and $Z$) were varied within the typical window computed before.
+Within this window, 5 values are used.
+The value of one of the latter two variables were visualized as color.
+This scatter data was used as an input to generate $\pi_{\text{PPO+GP}}$, visualized in the main text.}
+  \label{fig:ppo-policy}
+\end{figure}
+
 ## Old appendices:

### 2023-06-26 manuscript/manuscript.Rmd
erased old appendices

@@
 One $51\times 51 \times 5$, one $51\times 5 \times 51$ and one $5\times 51 \times 51$ grid on $(X,Y,Z)$ states.\footnote{
-The raw dataset is found at \textbf{TBD} in the repository with the source code and data linked above.
-Scatter plots visualizing this policy are shown in App. \textbf{TBD}.
+The raw dataset is found at the data/results\_data/2FISHERY/RXDRIFT subdirectory in the repository with the source code and data linked above.
+Scatter plots visualizing this policy are shown in App. B.
 }
@@
 
-## Old appendices:
-
-## Collected time series data
-
-In this appendix we provide a wider selection of time series taken from the same data set as Fig. \ref{fig:timeseries}.
-Namely, Figs. \ref{fig:timeseries_msy}, \ref{fig:timeseries_tac} and \ref{fig:timeseries_ppo} display these time series.
-
-## Conservative scale-downs of MSY
-
-Here we show the performance of constant mortality policies with a fraction $\alpha$ of the MSY mortality rate (in the main text we focused on $\alpha=0.8$).
-Their performance is evaluated in Fig. \ref{fig:several-tac}.
-Notice the sharp transition in performance between $\alpha = 0.9$ and $0.95$.
-While the behaviour of the former closely mimics, e.g., the $\alpha=0.8$ policy, the performance of the latter is essentially the same as the optimal constant mortality ($\alpha=1$, see Fig.
-\ref{fig:rew_len_fig}).
-This points at the fact that the value of the parameter $\alpha$, although commonly chosen simply as a rule of thumb, lives on a ‘‘knife's edge,’’ small changes in this parameter may lead to drastic changes in the dynamics of the controlled system.
-What is worse: the particular scale at which this transition happens is generally dependent on the system's intrinsic dynamics.
-In reality, where fishery models and their respective parameters are subject to substantial uncertainty, this fact calls to question the purported additional safety obtained via these rules of thumb.
-
-## Optimal constant escapement policy evaluation
-
-Here we reproduce the plots from the main text for the optimal constant escapement policy to illustrate the similarity to the DRL-derived policy.
-
-Fig. \ref{fig:esc-tuning-eval} illustrates the tuning process of finding the optimal constant escapment policy (left panel, compare Fig. \ref{fig:tuning}).
-The same figure shows, in the middle panel, the reward histogram for the optimal constant escapement policy (compare Fig. \ref{fig:rew_len_fig}).
-Finally, the right panel displays the same histogram for the DRL policy (reproduced from Fig. \ref{fig:rew_len_fig}).
-We may see that the two policies lead to essentially the same reward distribution.
-
-Fig. \ref{fig:timeseries_esc} displays 9 time-series for episodes controlled with the optimal constant escapement policy.
-Again, this behaviour is essentially indistinguishable from the behavior of the system controlled by the DRL policy (see Fig. \ref{fig:timeseries_ppo}).
-
-This series of figures gives evidence that, in fact, the DRL agent we have trained has obtained a policy which, effectively, acts as an optimal constant escapement policy.
-Although in Fig.
-\ref{fig:action-X} we found some deviations from the constant escapement case (e.g.
-DRL was more conservative when fishing high populations), these do not seem to lead to a different dynamics.
-
 # References

### 2023-06-26 manuscript/manuscript.Rmd
mini
abstract

@@
   or ever attempting to infer a model for the process itself. We find that
-  the reinforcement learning agent discovers an effective simplification of
-  the problem to obtain an interpretable control rule. We show that the 
-  policy obtained with DRL is both more profitable and more sustainable 
-  than any constant mortality policy---the standard family of policies 
-  considered in fishery management.
+  the reinforcement learning agent discovers a policy that outperforms both
+  constant escapement and constant mortality policies---the standard family of policies 
+  considered in fishery management. This DRL policy has 
+  the shape of a constant escapement policy whose escapement values depend
+  on the stock sizes of other species in the model.
 
@@
 Here we present the reward distributions for the other models considered---the three stationary models, lines 1-3 in Table 1.
+These results are shown in Figs. \ref{fig:main_32}, \ref{fig:main_31} and \ref{fig:main_11}.
 

### 2023-06-28 manuscript/manuscript.Rmd
reshuffled table and fig 1, started using Model i for model names

@@
 
+<!--NEED some kind of map paragraph here saying what's ahead, what we show  -->
+In this article, we compare the performance of a DRL solution to that of classical management strategies (see Textbox 1).\footnote{A repository with all the relevant code to reproduce our results may be found at \url{https://github.com/boettiger-lab/approx-model-or-approx-soln} in the ‘‘src’’ subdirectory.
+The data used is found in the ‘‘data’’ subdirectory, but the user may use the code provided to generate new data sets.} 
+We consider the problem of devising harvest strategies in a for a series of ecosystem models with increasing complexity (Table \ref{tab:models}):
+*1) one species, one fishery:* a simple single-species recruitment model based on [@may77];
+*2) three species, one fishery:* a three-species generalization of model *1)*, where one of the species is harvested;
+*3) three species, two fisheries:* the same three-species model as above but with two harvested species;\footnote{
+As we prepared this manuscript we became aware of [@nicolalove] which considers an ecological control problem with a similar mathematical formulation, but using an analytical calculus-based approach.
+}
+*4) three species, two fisheries, parameter variation:* a three-species model of which two are harvested, as above, with a time-varying parameter.
+This last model is meant to be a toy model of climate change's effect on the system.
+Across all of these scenarios, two goals are balanced in the decision process: maximizing long-term catch and preventing stock sizes to fall below some a-priori threshold.
+
+\begin{table}
+  \begin{center}
+    \begin{tabular}{|l|c|c|c|c|}
+      \hline
+      \textbf{Model name} & \textbf{Model eqs.} & \textbf{N. Sp.} & \textbf{Harv. Sp.} & \textbf{Stationary?}\\
+      \hline
+      Model 1 & \eqref{eq:may} & 1 & $X$ & Yes \\
+      Model 2 & \eqref{eq:3d model} & 3 & $X$ & Yes \\
+      Model 3 & \eqref{eq:3d model} & 3 & $X$ and $Y$ & Yes \\
+      Model 4 & \eqref{eq:3d model} & 3 & $X$ and $Y$ & No \\
+      \hline
+    \end{tabular}
+    \caption{Table of models considered in this paper.
+Here, \textbf{N.
+Sp.} is the \emph{number of species} in the model, \textbf{Harv.
+Sp.} is the species of the model which are harvested, and \textbf{Stationary?} refers to whether the parameters of the model have fixed values (or, on the contrary, if they vary in time).
+The only non-stationary case presented in the paper is where $r_X$ drifts linearly with time.
+In the code repository associated to the paper, we consider other possible choices of non-stationarity.}
+    \label{tab:models}
+  \end{center}
+\end{table}
+
+This way, we evaluate classical management strategies (CMort and CEsc) and DRL-based strategies on four different models.
+This experimental design is summrized in Fig.  \ref{fig:conceptual-v2}.
+
 ```{r}
@@
 
-<!--NEED some kind of map paragraph here saying what's ahead, what we show  -->
-In this article, we compare the performance of a DRL solution to that of classical management strategies (see Textbox 1).\footnote{A repository with all the relevant code to reproduce our results may be found at \url{https://github.com/boettiger-lab/approx-model-or-approx-soln} in the ‘‘src’’ subdirectory.
-The data used is found in the ‘‘data’’ subdirectory, but the user may use the code provided to generate new data sets.} 
-We consider the problem of devising harvest strategies in a for a series of ecosystem models (Figure \ref{fig:conceptual-v2}):
-*1) one species, one fishery:* a simple single-species recruitment model based on [@may77];
-*2) three species, one fishery:* a three-species generalization of model *1)*, where one of the species is harvested;
-*3) three species, two fisheries:* the same three-species model as above but with two harvested species;\footnote{
-As we prepared this manuscript we became aware of [@nicolalove] which considers an ecological control problem with a similar mathematical formulation, but using an analytical calculus-based approach.
+Regarding control for these 4 models, we show the following. 
+*Model 1:* DRL-based strategies recover the optimal CEsc policy function.  
+Constant mortality performs considerably worse than these three.\footnote{
+As will be explained later, all our models are stochastic. 
+If we set stochasticity to zero in Model 1, constant mortality matches the performance of the other management strategies.
 }
-*4) three species, two fisheries, parameter variation:* a three-species model of which two are harvested, as above, with a time-varying parameter.
-This last model is meant to be a toy model of climate change's effect on the system.
-Across all of these scenarios, two goals are balanced in the decision process: maximizing long-term catch and preventing stock sizes to fall below some a-priori threshold.
-
-Regarding control for these 4 models, we show that: *1)* while all approaches match in performance in the single species deterministic model, even mild stochasticity plummets the performance of MSY relative to constant escapement.
-Moreover, DRL recovers the optimal constant escapement policy.
-*2)* In the three species, single fishery model, all management strategies perform similarly.
-*3)* In the three species, two fisheries model DRL outperforms both classical strategies, with MSY surprisingly performing significantly better than constant escapement.
+*Model 2:* Here, all management strategies perform similarly.
+*Model 3:* For this model, DRL outperforms both classical strategies, with CMort surprisingly performing significantly better than constant escapement.
 In particular, we observe that DRL strategies are more sensitive to stochastic variations of the system which allows it to adaptively manage the system to prevent population collapses below the threshold.
-*4)* In the three species, two fisheries model with parametric time-variation, the performance gap between DRL and MSY is broadened.
-In this case, as in *3)*, constant escapement strategies perform significantly worse than DRL and MSY.
+*Model 4:* Here, the performance gap between DRL and classical management strategies is broadened.
+In this case, as in Model 3, constant CEsc performs significantly worse than DRL and CMort.
 
@@
 The policies explored were functions from states $N_t$ to either a single number $M_X(N_t) = \pi(N_t)$, or a pair of numbers $(M_X(N_t), M_Y(N_t)) = \pi(N_t)$ depending on the environment considered.
-
-\begin{table}
-  \begin{center}
-    \begin{tabular}{|l|c|c|c|c|}
-      \hline
-      \textbf{Model name} & \textbf{Model eqs.} & \textbf{N. Sp.} & \textbf{Harv. Sp.} & \textbf{Stationary?}\\
-      \hline
-      1 species, 1 fishery & \eqref{eq:may} & 1 & $X$ & Yes \\
-      3 species, 1 fishery & \eqref{eq:3d model} & 3 & $X$ & Yes \\
-      3 species, 2 fishery & \eqref{eq:3d model} & 3 & $X$ and $Y$ & Yes \\
-      3 species, 2 fishery, non-stationary & \eqref{eq:3d model} & 3 & $X$ and $Y$ & No \\
-      \hline
-    \end{tabular}
-    \caption{Table of models considered in this paper.
-Here, \textbf{N.
-Sp.} is the \emph{number of species} in the model, \textbf{Harv.
-Sp.} is the species of the model which are harvested, and \textbf{Stationary?} refers to whether the parameters of the model have fixed values (or, on the contrary, if they vary in time).
-The only non-stationary case presented in the paper is where $r_X$ drifts linearly with time.
-In the code repository associated to the paper, we consider other possible choices of non-stationarity.}
-    \label{tab:models}
-  \end{center}
-\end{table}
 

### 2023-06-28 manuscript/manuscript.Rmd
reshuffled intro a bit

@@
 
-Our findings paint a picture of how a single-species optimal management strategy may loose performance rather dramatically when controlling a more complex ecosystem.
-Moreover, rather unintuitively, management strategies, such as MSY, known to be suboptimal and unsustainable in the single-species can turn out to even outperform the aforementioned single-species optimal strategy (e.g.
-constant escapement).
-Finally, within this regime of complex, possibly varying, ecosystems, we show that DRL consistently finds a policy which effectively either matches the best classical strategy, or outperforms it.
-We strengthen this result with a stability investigation: we show that random perturbations in the model's parameter values used do not significantly vary the conclusion that DRL outperforms constant escapement.
-
-We notice that both classical management strategies are faced with a tradeoff---optimal classical strategies accept a rather large fraction of episodes ending with a population crash, whose negative reward is counteracted with a higher economic output.\footnote{
+We show that in the most complex scenario, Model 4, CMort is faced with a tradeoff---the optimal mortality rate leads a rather large fraction of episodes ending with a population crash, whose negative reward is counteracted with a higher economic output.\footnote{
 In our mathematical formulation of the decision problem, we have assumed for simplicity that the fishing effort cost is zero and that fish price is stable over time.
@@
 }
-Conversely, more conservative suboptimal classical strategies (e.g. with a higher constant escapement value) lead to lower total reward on average.
-The DRL approach side-steps this tradeoff by optimizing over a more complex family of possible policies—policies parametrized by a neural network, as opposed to policies labeled by a single parameter (the mortality rate).
-We thus observe that the DRL solution leads to a higher profitability *and* stability as compared to constant mortality policies.
+Conversely, more conservative, lower, mortality rates lead to lower total reward on average.
+The DRL approach side-steps this trade-off by optimizing over a more complex family of possible policies--—policies parametrized by a neural network, as opposed to policies labeled by a single parameter (the mortality rate).
+
+Our findings paint a picture of how a single-species optimal management strategy may loose performance rather dramatically when controlling a more complex ecosystem.
+Here, DRL performs better from both economical *and* conservation points of view.
+Moreover, rather unintuitively, CMort -- known to be suboptimal and unsustainable for single-species models -- can turn out to even outperform the CEsc -- the single-species optimal strategy -- for complex ecosystems.
+Finally, within this regime of complex, possibly varying, ecosystems, we show that DRL consistently finds a policy which effectively either matches the best classical strategy, or outperforms it.
+We strengthen this result with a stability investigation: we show that random perturbations in the model's parameter values used do not significantly vary the conclusion that DRL outperforms constant escapement.
 

### 2023-06-28 manuscript/manuscript.Rmd
mathematical models section

@@
 
-Here we introduce the four mathematical models of fisheries for which we compare different management strategies.
+Here we mathematically introduce the four fishery models for which we compare different management strategies.
 In general, the class of models that appear in this context are *stochastic, first order, finite difference equations*.
@@
 *2. Perfect execution:* the action chosen by the manager is implemented perfectly (i.e., there is no noise affecting the actual value of the fishing efforts).
-In future work, we intend to relax these assumptions to get more realistic management scenarios.
-
-One of the models we consider is a single-species classical model of ecological tipping points.
-The other three models are all three-species models with similar dynamics---differing by the amount of species that are harvested, and whether the parameters of the model are fixed or time-dependent.
+
+Model 1 is a single-species classical model of ecological tipping points.
+Models 2-4 are all three-species models with similar dynamics.
 Following this logic, the first subsection will be dedicated to the single-species model and the second will focus on the three-species models.
@@
 One particularly important aspect that has garnered much attention is the potential for the kind of highly non-linear functions that can support dynamics such as alternative stable states and hysteresis.
-A seminal example of such dynamics was introduced in [@may77], using a one-dimensional model of a prey (resource) species under the pressure of a (fixed) predator:
+A seminal example of such dynamics was introduced in [@may77], using a one-dimensional model of a prey (resource) species under the pressure of a (fixed) predator.
+In the notation of eq. \eqref{eq:general model},
 \begin{align}
   \label{eq:may}
-  \Delta X_t
-  = L(X_t;\ r, K) - \frac{\beta H X_t^2}{c^2 + X_t^2} + \eta_{X,t} - M_X X_t.
+  f_X(X_t)
+  = L(X_t;\ r, K) - \frac{\beta H X_t^2}{c^2 + X_t^2}.
 \end{align}
@@
 
-The model has six parameters: the growth rate $r$ and carrying capacity $K$ for $X$, a constant population $H$ of a species which preys on $X$, the maximal predation rate $\beta$, the predation rate half-maximum biomass $c$, and the variance $\sigma_X^2$ of $\eta_{X,t}$.
-(Here and in the following we will center our random variables at zero.)
+The model has six parameters: the growth rate $r$ and carrying capacity $K$ for $X$, a constant population $H$ of a species which preys on $X$, the maximal predation rate $\beta$, the predation rate half-maximum biomass $c$, and the variance $\sigma_X^2$ of the stochastic term $\eta_{X,t}$.
+(Here and in the following we will center all random variables at zero.)
 
@@
 
-We consider three models of a three-species fishery ecosystem.
-These are all closely related---in fact, their natural dynamics (i.e. dynamics under zero harvest) is given by the same equations:
+Models 2-4 are three-species models and they are all closely related---in fact, their natural dynamics (i.e. dynamics under zero harvest) is essentially given by the same equations:
 \begin{align}
@@
 \begin{split}
-  \Delta X_t &= L(X_t;\ r_X, K_X) - F(X_t, Z_t;\ \beta, c) - c_{XY}X_tY_t + \eta_{X,t},\\
-  \Delta Y_t &= L(Y_t;\ r_Y, K_Y) - D F(Y_t, Z_t;\ \beta, c) - c_{XY}X_tY_t + \eta_{Y,t},\\
-  \Delta Z_t &= \left(f(X_t + DY_t\right) - d_Z)Z_t + \eta_{Z,t}.
+  f_X(N_t) &= L(X_t;\ r_X, K_X) - F(X_t, Z_t;\ \beta, c) - c_{XY}X_tY_t, \\
+  f_Y(N_t) &= L(Y_t;\ r_Y, K_Y) - D F(Y_t, Z_t;\ \beta, c) - c_{XY}X_tY_t,\\
+  f_Z(N_t) &= \left(f(X_t + DY_t\right) - d_Z)Z_t.
 \end{split}
@@
 
-Eqs. \eqref{eq:3d model} are a generalization of the single-species model \eqref{eq:may}: this can be seen by setting 
-\begin{align*}
-  c_{XY} = d = f = \sigma_Z = 0.
-\end{align*}
-Throughout most of our paper, we use the parameter values given in eq. \eqref{eq:param vals}.
-In particular, $c_{XY} = d = f = 0.1$ and $\sigma_Z=0.05$ so, in broad strokes, our three-species model is a slight perturbation of the single-species model.
-Put differently, we are in a regime where, just by examining parameters, it would seem that the single-species model is a reasonable approximation.
-**Is this actually true? Should I maybe erase these last sentences?**
-
 The three models will branch off of eq. \eqref{eq:3d model} in the following way.
-*1. Three species, one fishery:* in this first model, only $X$ is harvested, that is, in the notation of eq. \eqref{eq:general model}, we fix $M_Y=M_Z=0$ and leave $M_X$ as a control variable.
-All parameters here are constant, with values given by \eqref{eq:param vals}.
-*2. Three species, two fisheries:* as the previous case but with $X$ and $Y$ being harvested. 
+*Model 1:* here, only $X$ is harvested, that is, in the notation of eq. \eqref{eq:general model}, we fix $M_Y=M_Z=0$ and leave $M_X$ as a control variable.
+All parameters here are constant.
+*Model 2:* as Model 1, but with $X$ and $Y$ being harvested. 
 In other words, we set $M_Z=0$ and leave $M_X$ and $M_Y$ as control variables.
-Finally, *3. Three species, two fiseries, time-variation:* in this model we allow for the harvest of $X$ and $Y$, as in the previous model, but now we include a time dependent parameter:
+*Model 3:* here $X$ and $Y$ are harvested, but now we include a non-stationary parameter:
 \begin{equation}
@@
 \end{equation}
-All other parameters are fixed by eq. \eqref{eq:param vals}.
+All other parameters are constant.
 
@@
 
+## Pasted to-be-discussed
+
+Eqs. \eqref{eq:3d model} are a generalization of the single-species model \eqref{eq:may}: this can be seen by setting 
+\begin{align*}
+  c_{XY} = d = f = \sigma_Z = 0.
+\end{align*}
+Throughout most of our paper, we use the parameter values given in eq. \eqref{eq:param vals}.
+In particular, $c_{XY} = d = f = 0.1$ and $\sigma_Z=0.05$ so, in broad strokes, our three-species model is a slight perturbation of the single-species model.
+
+## Discussion
+
 Fisheries are complex ecosystems, with interactions between species leading to highly non-linear dynamics.

### 2023-06-28 manuscript/manuscript.Rmd
mini

@@
 The three models will branch off of eq. \eqref{eq:3d model} in the following way.
-*Model 1:* here, only $X$ is harvested, that is, in the notation of eq. \eqref{eq:general model}, we fix $M_Y=M_Z=0$ and leave $M_X$ as a control variable.
+*Model 2:* here, only $X$ is harvested, that is, in the notation of eq. \eqref{eq:general model}, we fix $M_Y=M_Z=0$ and leave $M_X$ as a control variable.
 All parameters here are constant.
-*Model 2:* as Model 1, but with $X$ and $Y$ being harvested. 
+*Model 3:* as Model 1, but with $X$ and $Y$ being harvested. 
 In other words, we set $M_Z=0$ and leave $M_X$ and $M_Y$ as control variables.
-*Model 3:* here $X$ and $Y$ are harvested, but now we include a non-stationary parameter:
+*Model 4:* here $X$ and $Y$ are harvested, but now we include a non-stationary parameter:
 \begin{equation}
@@
 All other parameters are constant.
-
 Eq. \eqref{eq:rx(t)} is intended to reflect in a simple manner a possible effect of climate change: where the reproductive rate of $X$ is reduced linearly over time until it stabilizes.
-In future work we intend to test the DRL framework in realistic models of particular fisheries.
-As a part of that effort, we will attempt capture more detailed models of the effects of climate change.
-For this exploratory work we have chosen this simple expression, however, for the sake of conceptual clarity.
 

### 2023-06-28 manuscript/manuscript.Rmd
RL, erased niki footnote

@@
 *2) three species, one fishery:* a three-species generalization of model *1)*, where one of the species is harvested;
-*3) three species, two fisheries:* the same three-species model as above but with two harvested species;\footnote{
-As we prepared this manuscript we became aware of [@nicolalove] which considers an ecological control problem with a similar mathematical formulation, but using an analytical calculus-based approach.
-}
+*3) three species, two fisheries:* the same three-species model as above but with two harvested species;
 *4) three species, two fisheries, parameter variation:* a three-species model of which two are harvested, as above, with a time-varying parameter.
@@
 
-For our use case, the environment will be a computational model of the population dynamics of a fishery, with the environment state at time $t$ being a vector of all the fish populations, e.g.
-$N_t=(X_t,Y_t,Z_t)$ in the three species case.
-At each time step, the agent observes the state $S$ and enacts some harvest---reducing $X_t$ to $X_t - M_X(N_t)\cdot X_t$, and, in the two-fishery scenario, also reducing $Y_t$ to $Y_t - M_Y(N_t)\cdot Y_t$.
+For this paper, we will consider four environments corresponding to each of the models considered (Table \ref{tab:models}).
+At each time step, the agent observes the state $S$ and enacts some harvest---reducing $X_t$ to $X_t - M_X(N_t)\cdot X_t$, and, for Models 3 and 4, also reducing $Y_t$ to $Y_t - M_Y(N_t)\cdot Y_t$.
 Here the fish mortality-rates-from-harvest (i.e. $M_X=M_X(N_t)$ and $M_Y=M_Y(N_t)$), are the agent's action at time $t$. 
-This secures a reward of $M_X(N_t)X$ in the single fishery case, and in the two-fishery case the reward is $M_X(N_t)X + M_Y(N_t)Y$.
-After this harvest portion of the time step, the environment changes according to its natural dynamics given by \eqref{eq:3d model}.
-
-As mentioned, this discretized time framework allows a simplification of the possibly seasonal mating behavioral patterns of the species involved. 
+This secures a reward of $M_X(N_t)X$ for Models 1 and 2, and, similarly, a reward of $M_X(N_t)X + M_Y(N_t)Y$ for Models 3 and 4.
+After this harvest portion of the time step, the environment changes according to its natural dynamics (see eq. \eqref{eq:3d model}).
+
+As mentioned, discretising time allows a simplification of the possibly seasonal mating behavioral patterns of the species involved. 
 This approximation is commonly used in fisheries for species with annual reproductive cycles (see e.g. [@mangel2006theoretical], Chap. 6).

### 2023-06-28 manuscript/manuscript.Rmd
moved gaussian process stuff to appendix, updated more stuff on methods section

@@
 \textbf{The Environment.}
-As seen in Fig.
-\ref{fig:conceptual-v2}, we consider four different models of the ecosystem with which the fishery interacts, each of this is called an *environment*, alluding to the environment in an RL setting.
-We summarize these models in Table \ref{tab:models}.
+We considered the problem of managing four increasingly complex models (see Table \ref{tab:models}).
 To recap, these four environments are a single-species growth model \eqref{eq:may} for a harvested species; a three-species model \eqref{eq:3d model} with a single harvested species; the same three-species model but with *two* harvested species; and, finally, a three-species model with a time-varying parameter and two harvested species.
 
-The policies explored were functions from states $N_t$ to either a single number $M_X(N_t) = \pi(N_t)$, or a pair of numbers $(M_X(N_t), M_Y(N_t)) = \pi(N_t)$ depending on the environment considered.
-
-Our goal is to evaluate the performance of different policy strategies over a specified window of time.
-We have chosen this window to be 200 time-steps, where each discrete time-step represents the dynamical evolution of the system over a year.
-_Episodes_ begin at time-step zero and are controlled by the policy and the system's natural evolution until either the maximum time-span is reached, or until a ‘‘near-extinction event’’ occurs---that is, until any of the populations go below a given threshold,
+The policies explored were functions from states $N_t$ to either a single number $M_X(N_t) = \pi(N_t)$ (for Models 1 and 2), or a pair of numbers $(M_X(N_t), M_Y(N_t)) = \pi(N_t)$ (for Models 3 and 4).
+
+Our goal was to evaluate the performance of different policy strategies over a specified window of time.
+We chose this window to be 200 time-steps, where each discrete time-step represents the dynamical evolution of the system over a year.
+Each time-step was composed of two parts: 
+First, a *harvest period* where the harvest is collected from the system (e.g. the population $X$ is reduced to $X_t\mapsto X_t - M_X(N_t) X_t$).
+Second, a *recruitment and interaction period*, where the system's state evolves according to its natural dynamics.
+
+Training and evaluating management strategies was done by simulating *episodes.* 
+An episode begins at a fixed initial state and the system is controlled with a management policy until $t=200$, or until a ‘‘near-extinction event’’ occurs---that is, until any of the populations go below a given threshold,
 \begin{align}
@@
 \end{align}
-In our setting we have chosen $X_{\text{thresh.}}=Y_{\text{thresh.}}=Z_{\text{thresh.}}=0.05$ as a rule of thumb---given that under natural (unharvested) dynamics the populations range within values of 0.5 to 1, this would represent on the order of a 90-95% population decrease from its ‘‘natural’’ level.
-
-The reward function defining our policy optimization problem has two components.
-The first is economical: the total biomass harvested over an episode.
-The second seeks to reflect conservation goals: if a near-extinction event occurs at time $t$, the episode is ended early and a negative reward of $-100/t$ was awarded.
-This reward function balances the extractive motivation of the fishery with conservation goals that go beyond the scope of long-term sustainable harvests.
+In our setting we have chosen $X_{\text{thresh.}}=Y_{\text{thresh.}}=Z_{\text{thresh.}}=0.05$ as a rule of thumb---given that under natural (unharvested) dynamics the populations range within values of 0.5 to 1, this would represent on the order of a 90-95% population decrease from their ‘‘natural’’ level.
+
+The reward function defining our policy optimization problem had two components.
+The first was economical: the total biomass harvested over an episode.
+The second seeked to reflect conservation goals: if a near-extinction event occurred at time $t$, the episode was ended early and a negative reward of $-100/t$ was awarded.
+This reward function balanced the extractive motivation of the fishery with conservation goals which went beyond the scope of long-term sustainable harvests.
 
@@
 The state space used was normalized case-by-case as follows: 
-*1)* for the single species model, a line segment $[0,1]$, *2)* for all three-species models, a cube $[0,1]^{\times 3}$.
-We used simulation data to derive a bound on the population sizes typically observed, and thus be able to normalize states to a finite volume.
-
-Policies obtained from the PPO algorithm can be ‘‘noisy’’ as their optimization algorithm is randomized (see, e.g.
-App.
-B for visualizations of the PPO policy in the ‘‘3 species, 2 fisheries, non-stationary’’ case).
-We perform some interpolation using a Gaussian process regressor in order to de-noise these policies.
-Specifically, we use the *GaussianProcessRegressor* object of the *sklearn* Python library with a kernel given by
-\begin{align*}
-  \text{RBF}(\text{length scale = 10}) + \text{WhiteNosie}(\text{noise level = 0.1}).
-\end{align*}
-This interpolation method is applied to scatter data of the PPO policy evaluated on 3 different grids:
-One $51\times 51 \times 5$, one $51\times 5 \times 51$ and one $5\times 51 \times 51$ grid on $(X,Y,Z)$ states.\footnote{
-The raw dataset is found at the data/results\_data/2FISHERY/RXDRIFT subdirectory in the repository with the source code and data linked above.
-Scatter plots visualizing this policy are shown in App. B.
-}
-The length scale and noise level values of this kernel were chosen arbitrarily---no hyperparameter tuning was needed to produce satisfactory interpolation, as will be shown in the results section.
+*Model 1:* a line segment $[0,1]$, 
+*Models 2-4:* a cube $[0,1]^{\times 3}$.
+We used simulated data to derive a bound on the population sizes typically observed, and thus be able to normalize states to a finite volume.
+
+Policies obtained from the PPO algorithm can be ‘‘noisy’’ as their optimization algorithm is randomized (see, e.g. App. B for a visualization of the PPO policy obtained for Model 4).
+We smoothed this policy out using a Gaussian process regressor interpolation. 
+Details for this interpolation process can be found in App. C.
 
@@
 
+## Gaussian process interpolation
+
+Here we summarize the procedure used to interpolate the PPO policy (visualized in Fig. \ref{fig:ppo-policy}).
+We use the *GaussianProcessRegressor* object of the *sklearn* Python library with a kernel given by
+\begin{align*}
+  \text{RBF}(\text{length scale = 10}) + \text{WhiteNosie}(\text{noise level = 0.1}).
+\end{align*}
+This interpolation method is applied to scatter data of the PPO policy evaluated on 3 different grids on $(X,Y,Z)$ states:
+$G_X$, a $51\times 5 \times 5$ grid; $G_Y$, a $5\times 51 \times 5$ grid; and $G_Z$, a $5\times 5 \times 51$ grid.
+This combination of grids was used instead of a single dense grid in order to reduce the computational intensity of the interpolation procedure.
+For $G_X$, the 5 values for $Y$ and $Z$ were varied in a ‘‘popular window,’’ i.e. episode time-series data was used to determine windows of $Y$ and $Z$ values which were most likely. 
+The grids $G_Y$ and $G_Z$ were generated in a similar fashion, *mutatis mutandis*.\footnote{
+The raw dataset is found at the data/results\_data/2FISHERY/RXDRIFT sub-directory in the repository with the source code and data linked above.
+Scatter plots visualizing this policy are shown in App. B.
+}
+The length scale and noise level values of this kernel were chosen arbitrarily---no hyperparameter tuning was needed to produce satisfactory interpolation, as will be shown in the results section.
+
 # References

### 2023-06-28 manuscript/manuscript.Rmd
finished going over methods, started modifying results

@@
 
-\textbf{Tuning the constant mortality strategy.}
-A grid of 101 mortality rates was laid in the interval $[0, 0.5]$ as possible mortality values in the single-fishery cases.
-Due to the high memory usage in the two-fishery case, a slightly coarser $51\times51$ grid was set in the square $[0,0.5]^2$ instead.
-Since the approach for tuning is completely analogous for all scenarios considered, here we focus on the ‘‘3 species, 1 fishery’’ case.
-For each one of these choices of a constant mortality rate, say $M_X$, we simulated 100 episodes based on \eqref{eq:3d model}:
-At each time step the state $(X_t, Y_t, Z_t)$ was observed, a harvest of $M_X X_t$ was collected and then the system evolved according to its natural dynamics \eqref{eq:3d model}.
-The optimal mortality rate was the value $M_X^*$ for which the mean episode reward was maximal.
-
-\textbf{Tuning the constant escapement strategy.}
+\textbf{Tuning the CMort strategy.}
+In order to estimate the optimum mortality rate, we optimized over a grid of possible mortality rates.
+Namely, for Models 1 and 2, grid of 101 mortality rates was laid in the interval $[0, 0.5]$;
+for Models 3 and 4, a $51\times51$ grid was set in the square $[0,0.5]^2$.
+The latter had a slightly coarser grid due to the high memory cost of using the denser grid.
+Since the approach for tuning was completely analogous for all models evaluated, here we discuss only Model 4.
+For each one of these choices of mortality rates, say $(M_X, M_Y)$, we simulated 100 episodes based on \eqref{eq:3d model}:
+At each time step the state $(X_t, Y_t, Z_t)$ was observed, a harvest of $M_X X_t+M_YY_t$ was collected and then the system evolved according to its natural dynamics \eqref{eq:3d model}.
+The optimal mortality rate was the value $(M_X^*,\ M_Y^*)$ for which the mean episode reward was maximal.
+
+\textbf{Tuning the CEsc strategy.}
 This tuning procedure was analogous to that of the constant mortality strategy just summarized.
 Namely:
-A grid of 101 constant escapement values was laid out on the interval $[0,1]$ for the single-fishery models (resp. a $51\times 51$ grid on $[0,1]^2$ for two-fishery models).
-These grids were the possible constant escapement values optimized over (in the two fishery case, these were $(\text{esc}_X,\ \text{esc}_Y)$ pairs).
-That is, each grid point represents a constant escapement policy.
+A grid of 101 escapement values was laid out on the interval $[0,1]$ for Models 1 and 2; and a $51\times 51$ grid on $[0,1]^2$ was laid out for Models 3 and 4.
+Each grid point represented a CEsc policy.
 We used each of these policies to manage 100 replicate episodes.
 The optimal policy was the policy with the highest average reward obtained.
-A visualization of the tuning procedure for the ‘‘three species, two fisheries, non-stationary’’ model is shown in Fig. \ref{fig:tuning}.
+A visualization of the tuning outcome for Model 4 is shown in Fig. \ref{fig:tuning}.
 
@@
   \includegraphics[scale=0.5]{figures/esc_heatmap.png}
-  \caption{Visualization of the cosntant escapement strategy tuning procedure for the ‘‘three-species, two-fishery, non-stationary’’ model.
+  \caption{Visualization of the cosntant escapement strategy tuning procedure for Model 4.
 There was a certain multiplicity in this tuning strategy: a ‘‘ridge of optimality’’ where policies had essentially equivalent behavior.
@@
 \end{align}
-Here, the values of $\beta$ and $c$ are chosen as to make the system be roughly close to its tipping point.
-
-The three-species model's (eq. \eqref{eq:3d model}) dynamic parameters were chosen as follows
+Here, the values of $\beta$ and $c$ were chosen as to make the system be roughly close to its tipping point.
+
+For Models 2 and 3, their dynamic parameters (in eq. \eqref{eq:3d model}) were chosen as follows
 \begin{align}
@@
 \end{align}
-Moreover, the variances for the stochastic terms were chosen as
+Moreover, the variances for the stochastic terms were chosenas
 $$
@@
 $$
-In the case with a time-varying $r_X$ parameter, we used $r_X(t)$ given as in eq. \eqref{eq:rx(t)} and all other parameters as given in eq. \eqref{eq:param vals}.
-
-The values of $c$ and $\beta$ in the three-species model are slightly different than their values in the one-species model.
+For Model 4, we used $r_X(t)$ given as in eq. \eqref{eq:rx(t)} and all other parameters were given as in eq. \eqref{eq:param vals}.
+
+The values of $c$ and $\beta$ in the three-species model were slightly different than their values in the one-species model.
 These values were chosen heuristically: 
@@
 We observed that the difference in performance is maintained for even relatively high noise strengths.
-For simplicity we only perform stability analysis on the ‘‘3 species, 2 fisheries and a time varying parameter’’ case.
-
-More specifically, for each parametric noise strength $\sigma_{\text{param.}}$ choice in $[0.04, 0.08, \dots, 0.2]$, we executed the following procedure:
-Sample 100 choices of perturbed parameter values, where the perturbation is as follows---each parameter $P$ is perturbed to $(1+g_P)P$ where $g_P$ is a Gaussian random variable with variance $\sigma_{\text{param.}}^2$.
-For each of these parameter choices, tune the constant escapement strategy and train the DRL model on the corresponding model.
-Measure the average reward difference between strategies for each of these parameter choice samples using 100 replicate evaluation episodes.
-Take the mean of this average difference over the 100 perturbed parameter samples.
+We only considered the most complex case here, Model 4.
+
+For each value of parametric noise strength $\sigma_{\text{param.}}$, we executed the following procedure:
+We sampled 100 choices of perturbed parameter values, where the perturbation was as follows---each parameter $P$ was perturbed to $(1+g_P)P$ where $g_P$ was a Gaussian random variable with variance $\sigma_{\text{param.}}^2$.
+For each of these sample parameter sets, we tuned CEsc and trained the DRL agent.
+We measured the average reward difference between these two strategies for each of sample (this was done using 100 replicate evaluation episodes).
+Finally, we took the mean of this average difference over the 100 perturbed parameter samples.
+
+The parametric noise strength values used were $[0.04, 0.08, \dots, 0.2]$.
 
@@
 
-We evaluated each of the four control approaches considered -- MSY constant mortality, optimal constant escapement, PPO, and Gaussian process interpolation of the latter -- on each of the four scenarios examined (see Sec.
-3).
+We evaluated each of the four management strategies considered on Models 1-4.
+To recap, the management strategies were CEsc, CMort, PPO and a Gaussian process interpolation of PPO (‘‘PPO+G’’).
 Furthermore, we characterized the trade-off between economic output and sustainability faced by constant mortality policies.
@@
 
-We will visualize the results concerning the last scenario considered (‘‘3 species, 2 of which harvested, and a time-varying parameter’’) in this section, leaving the other cases for App. A.
+We will visualize the results concerning Model 4 in this section, leaving the other models for App. A.
 This is the most complex scenario considered and where our results show the most compelling advantage of DRL methods with respect to classical strategies.
@@
 Our main result is summarized in Fig. \ref{fig:main} which displays the total reward distributions for the policy obtained through each strategy.
-Here we note two surprising things:
-First, optimal constant escapement performs considerably worse than not only DRL-based methods, but also than MSY.
-This conceptually shows how one-dimensional intuitions on ecological control may loose validity for complex ecosystems. **(this might be unwelcome by the reviewers if not discussed properly)**
-Second, we see that classical strategies have relatively spread out reward distributions as compared to DRL methods.
+Here we see that CEsc has a long-tailed distribution of rewards, and its average reward is much lower than other management strategies.
+CMort has a shorter-tailed distribution and a much higher average reward.
+Finally, both DRL-based strategies have a more concentrated reward distribution and a higher average reward than CMort.
 
@@
 \ref{fig:ephistograms}.
-We see that early episode ends are a rather likely explanation for the wideness of constant escapement's wide reward distribution.
-The constant mortality strategy is also observed to lead to near-extinction events rather often, but in this case the episode lengths tend to be concentrated at higher values relative to the constant escapement case.
+We see that early episode ends are prevalent in classical strategies and rare for DRL-based strategies.
+Early episode ends tend to happen at lower $t$ values for CEsc than CMort.
+Thus, distribution of episode durations seems to be widest for CEsc, followed by CMort, DRL and DRL+GP.
 
@@
   geom_col() +
-  labs(x="repetition", y="ep. duration")
+  labs(x="", y="ep. duration")
 msy_len <- msy_ends %>%
@@
   geom_col() +
-  labs(x="repetition", y="ep. duration")
+  labs(x="", y="")
 ppo_len <- ppo_ends %>%
@@
   geom_col() +
-  labs(x="repetition", y="ep. duration")
+  labs(x="", y="")
 gpp_len <- gpp_ends %>%
@@
   geom_col() +
-  labs(x="repetition", y="ep. reward")
+  labs(x="", y="")
 
@@
   geom_col() +
-  labs(x="repetition", y="ep. reward")
+  labs(x="repetition", y="")
 ppo_rew <- ppo_ends %>%
@@
   geom_col() +
-  labs(x="repetition", y="ep. reward")
+  labs(x="repetition", y="")
 gpp_rew <- gpp_ends %>%
@@
   geom_col() +
-  labs(x="repetition", y="ep. duration")
+  labs(x="repetition", y="")
 
@@
 
-We examine the trade-off between profit and sustainability faced by the constant mortality strategy in Fig. \ref{fig:fracmsy}.
-We expect a similar and, possibly, more pronounced effect for the constant escapement strategy but do not analyze this case here.
-Here, two quantities are plotted: the fraction of evaluation episodes of maximal length (i.e. episodes with no near-extinction events) is and the average reward obtained.
+We examine the trade-off between profit and sustainability faced by the CMort strategy in Fig. \ref{fig:fracmsy}.
+We expect a similar and, possibly, more pronounced effect for the CEsc strategy but do not analyze this case here.
+Here, two quantities are plotted: the fraction of evaluation episodes with maximal length (i.e. episodes with no near-extinction events), and the average reward.
 On the x-axis we have several sub-optimal constant mortality strategies that err on the conservative side: e.g. the policy labeled ‘‘80% MSY’’ has the form
@@
 
+Here we note two surprising things:
+First, optimal constant escapement performs considerably worse than not only DRL-based methods, but also than MSY.
+This conceptually shows how one-dimensional intuitions on ecological control may loose validity for complex ecosystems. **(this might be unwelcome by the reviewers if not discussed properly)**
+Second, we see that classical strategies have relatively spread out reward distributions as compared to DRL methods.
+
 ## Discussion

### 2023-06-28 manuscript/manuscript.Rmd
got all the way up to policy visualization

@@
 We examine the trade-off between profit and sustainability faced by the CMort strategy in Fig. \ref{fig:fracmsy}.
-We expect a similar and, possibly, more pronounced effect for the CEsc strategy but do not analyze this case here.
-Here, two quantities are plotted: the fraction of evaluation episodes with maximal length (i.e. episodes with no near-extinction events), and the average reward.
-On the x-axis we have several sub-optimal constant mortality strategies that err on the conservative side: e.g. the policy labeled ‘‘80% MSY’’ has the form
+Two quantities are plotted: the fraction of evaluation episodes with maximal length (i.e. episodes with no near-extinction events), and the average reward.
+On the x-axis we have several sub-optimal mortality rates that err on the conservative side: e.g. the policy labeled ‘‘80% Opt. CMort’’ has the form
 \begin{align*}
@@
 \end{align*}
-where $(M_X^*,\ M_Y^*)$ is the optimal constat mortality strategy.
+where $(M_X^*,\ M_Y^*)$ is the optimal CMort strategy.
 We see that sufficiently conservative policies attain high sustainability, but only at a high price in terms of profit.\footnote{
@@
 }
+We expect a similar and, possibly, more pronounced effect for the CEsc strategy but do not analyze this case here.
 
@@
   geom_col(fill="darksalmon")+
-  labs(x="% MSY constant mortality", y="% full-length eps.") + 
+  labs(x="% Opt. CMort", y="% full-length eps.") + 
   coord_cartesian(ylim=c(70,101))
@@
                  position=position_dodge(0.05)) +
-  labs(x="% MSY constant mortality", y="Mean ep. reward")
+  labs(x="% Opt. CMort", y="Mean ep. reward")
 
@@
 One problem that is often encountered when using machine learning methods is the interpretability of the output of these methods.
-In our case, the output is a policy function parametrized by a neural network $\theta$:
+For our PPO strategy, the output is a policy function parametrized by a neural network $\theta$:
 \begin{align*}
@@
 where $N_t=(X_t,Y_t,Z_t)$ is the state of the system at time $t$, and where $M_X$ and $M_Y$ are the mortalities due to harvest during that time-step.
-Moreover, in the case of our ‘‘PPO+GP’’ management strategy, we sample $\pi_\theta$ on a grid of points $G=\{(X_i, Y_i, Z_i)\}_i$ and use a Gaussian process to interpolate $\pi_\theta$, i.e.
-\begin{align}
-  \pi_{\text{PPO+GP}} = \text{Gaussian process interpolation of } \pi_\theta|_G.
-\end{align}
-This process was explained in detail in the Methods section.
-
-Here we visualize the resulting $\pi_{\text{PPO+GP}}$ policy and provide an interpretation for it.\footnote{
-A similar characterization of $\pi_\theta$ is found on the appendix. 
-Here we focus on the Gaussian process interpolation as it is smoother and, thus, easier to interpret.
-}
+While the values of the neural network parameters are hard to interpret, the actual shape of the policy function is much more understandable.
+
+Here we visualize the PPO+GP policy function and provide an interpretation for it, as this function is smoother and less noisy than the PPO policy function.
+The PPO policy function is visualized similarly in App. B.
+
 Given its high dimension, it is not possible to fully display how the policy function obtained ‘‘looks like’’---we thus project it down to certain relevant axes.

### 2023-06-29 manuscript/manuscript.Rmd
First pass at edits

--- 2023-06-28 manuscript/manuscript.Rmd
@@
 While simplified models have historically allowed for exploration of large decision spaces, the new wave of rich models are applied to highly oversimplified descriptions of potential actions they seek to inform.
+<!-- ML: Can we clarify that we are talking about dynamics/process models here?
+Model is a very overloaded term for me-->
 For instance, Global Circulation Models (GCMs) such as HadCM3 [@pope2000impact; @gordon2000simulation; @collins2001internal] model earth's climate using 1.5M variables, while the comparably vast potential action space is modeled as one of seven emissions marker scenarios from the IPCC [@riahi2017shared].
@@
 This simplification reflects trade-offs between tractablity and complexity at the basis of any mathematical modeling -- if we make both the state space and action space too realistic, the problem of finding the best sequence of actions quickly becomes intractable.
-However, emerging data-driven methods from machine learning offer a new choice -- algorithms that can find good strategies in such problems, but at the cost of opacity.
+However, emerging data-driven methods from machine learning offer a new choice -- algorithms that can find good strategies in previously intractable problems, but at the cost of opacity.
 
 In this paper, we focus on a well-developed application of model-based management of the natural world that has long illustrated the trade-offs between model complexity and policy complexity: the management of marine fisheries.
-Fisheries management is both an important issue to itself and a rich & frequent test-bed of ecological management more generally.
+Fisheries management is both an important issue to society and a rich & frequent test-bed of ecological management more generally.
 Fisheries are an essential natural resource that provide the primary source of protein for one in every four humans, and have faced widely documented declines due to over-fishing [@Worm2006, @Costello2016].
@@
 From this shared origin, each field would depart from the simplifying assumptions of the Gordon-Schaefer model in divergent ways, leading to different techniques for deriving policies from models.The heart of the management problem is easily understood: a manager seeks to set quotas on fishing that will ensure the long-term profitability and sustainability of the industry.
-Mathematical approaches developed over the past century may be roughly divided between these two fields: (A) ecologists, focused on ever more realistic models of the biological processes of growth and recruitment of fish while considering relatively stylized suite of potential management strategies, and (B) economists, focused on far more stylized models of the ecology while exploring a far less constrained set of possible policies.
+Mathematical approaches developed over the past century may be roughly divided between these two fields: (A) ecologists, focused on ever more realistic models of the biological processes of growth and recruitment of fish while considering a relatively stylized suite of potential management strategies, and (B) economists, focused on far more stylized models of the ecology while exploring a far less constrained set of possible policies.
 The economist's approach can be characterized by the mathematics of a Markov decision process [MDP @Clark1973; @Clark1990; @Marescot2013], in which the decision-maker must observe the stock each year and recommend a possible action.
@@
 Constrained by computational limits, MDP approaches have been intractable on suitably realistic models and largely confined to more academic applications [@Costello2016].
-However, advances _Deep Reinforcement Learning_, (DRL) a sub-field of machine learning, have recently demonstrated remarkable performance in a range of such MDP problems, from video games [@atari1; @atari2] to fusion reactions [@fusion1; @fusion2] to the remarkable dialog abilities of ChatGPT [@chatgpt].
-RL methods also bring many challenges of their own: being notoriously difficult to train and evaluate, requiring immense computational costs, and presenting frequent challenges with transfer-ability or reproducibility.
+However, advances in _Deep Reinforcement Learning_, (DRL) a sub-field of machine learning, have recently demonstrated remarkable performance in a range of such MDP problems, from video games [@atari1; @atari2] to fusion reactions [@fusion1; @fusion2] to the remarkable dialog abilities of ChatGPT [@chatgpt].
+RL methods also bring many challenges of their own: being notoriously difficult to train and evaluate, requiring immense computational costs, and presenting frequent challenges with reproducibility.
 A review of all these issues is beyond our scope but can be found elsewhere [@rl-intro; @millie-3].
@@
 
-<!--NEED some kind of map paragraph here saying what's ahead, what we show  -->
-In this article, we compare the performance of a DRL solution to that of classical management strategies (see Textbox 1).\footnote{A repository with all the relevant code to reproduce our results may be found at \url{https://github.com/boettiger-lab/approx-model-or-approx-soln} in the ‘‘src’’ subdirectory.
-The data used is found in the ‘‘data’’ subdirectory, but the user may use the code provided to generate new data sets.} 
-We consider the problem of devising harvest strategies in a for a series of ecosystem models with increasing complexity (Table \ref{tab:models}):
-*1) one species, one fishery:* a simple single-species recruitment model based on [@may77];
-*2) three species, one fishery:* a three-species generalization of model *1)*, where one of the species is harvested;
-*3) three species, two fisheries:* the same three-species model as above but with two harvested species;
-*4) three species, two fisheries, parameter variation:* a three-species model of which two are harvested, as above, with a time-varying parameter.
-This last model is meant to be a toy model of climate change's effect on the system.
-Across all of these scenarios, two goals are balanced in the decision process: maximizing long-term catch and preventing stock sizes to fall below some a-priori threshold.
-
-\begin{table}
-  \begin{center}
-    \begin{tabular}{|l|c|c|c|c|}
-      \hline
-      \textbf{Model name} & \textbf{Model eqs.} & \textbf{N. Sp.} & \textbf{Harv. Sp.} & \textbf{Stationary?}\\
-      \hline
-      Model 1 & \eqref{eq:may} & 1 & $X$ & Yes \\
-      Model 2 & \eqref{eq:3d model} & 3 & $X$ & Yes \\
-      Model 3 & \eqref{eq:3d model} & 3 & $X$ and $Y$ & Yes \\
-      Model 4 & \eqref{eq:3d model} & 3 & $X$ and $Y$ & No \\
-      \hline
-    \end{tabular}
-    \caption{Table of models considered in this paper.
-Here, \textbf{N.
-Sp.} is the \emph{number of species} in the model, \textbf{Harv.
-Sp.} is the species of the model which are harvested, and \textbf{Stationary?} refers to whether the parameters of the model have fixed values (or, on the contrary, if they vary in time).
-The only non-stationary case presented in the paper is where $r_X$ drifts linearly with time.
-In the code repository associated to the paper, we consider other possible choices of non-stationarity.}
-    \label{tab:models}
-  \end{center}
-\end{table}
-
-This way, we evaluate classical management strategies (CMort and CEsc) and DRL-based strategies on four different models.
-This experimental design is summrized in Fig.  \ref{fig:conceptual-v2}.
-
 ```{r}
@@
 
-Regarding control for these 4 models, we show the following. 
-*Model 1:* DRL-based strategies recover the optimal CEsc policy function.  
-Constant mortality performs considerably worse than these three.\footnote{
-As will be explained later, all our models are stochastic. 
-If we set stochasticity to zero in Model 1, constant mortality matches the performance of the other management strategies.
+<!--NEED some kind of map paragraph here saying what's ahead, what we show  -->
+In this article, we compare the performance of a DRL solution to that of classical management strategies (see Textbox 1).\footnote{A repository with all the relevant code to reproduce our results may be found at \url{https://github.com/boettiger-lab/approx-model-or-approx-soln} in the ‘‘src’’ subdirectory.
+The data used is found in the ‘‘data’’ subdirectory, but the user may use the code provided to generate new data sets.} 
+We consider the problem of devising harvest strategies for a series of ecosystem models (Figure \ref{fig:conceptual-v2}):
+*1) one species, one fishery:* a simple single-species recruitment model based on [@may77];
+*2) three species, one fishery:* a three-species generalization of model *1)*, where one of the species is harvested;
+*3) three species, two fisheries:* the same three-species model as above but with two harvested species;\footnote{
+As we prepared this manuscript we became aware of [@nicolalove] which considers an ecological control problem with a similar mathematical formulation, but using an analytical calculus-based approach.
 }
-*Model 2:* Here, all management strategies perform similarly.
-*Model 3:* For this model, DRL outperforms both classical strategies, with CMort surprisingly performing significantly better than constant escapement.
+*4) three species, two fisheries, parameter variation:* a three-species model of which two are harvested, as above, with a time-varying parameter.
+This last model is meant to be a toy model of climate change's effect on the system.
+Across all of these scenarios, two goals are balanced in the decision process: maximizing long-term catch and preventing stock sizes to fall below some a-priori threshold.
+
+Regarding control for these 4 models, we show that: *1)* while all approaches match in performance in the single species deterministic model, even mild stochasticity plummets the performance of MSY relative to constant escapement.
+Moreover, DRL recovers the optimal constant escapement policy.
+*2)* In the three species, single fishery model, all management strategies perform similarly.
+*3)* In the three species, two fisheries model, DRL outperforms both classical strategies, with MSY surprisingly performing significantly better than constant escapement.
 In particular, we observe that DRL strategies are more sensitive to stochastic variations of the system which allows it to adaptively manage the system to prevent population collapses below the threshold.
-*Model 4:* Here, the performance gap between DRL and classical management strategies is broadened.
-In this case, as in Model 3, constant CEsc performs significantly worse than DRL and CMort.
-
-We show that in the most complex scenario, Model 4, CMort is faced with a tradeoff---the optimal mortality rate leads a rather large fraction of episodes ending with a population crash, whose negative reward is counteracted with a higher economic output.\footnote{
+*4)* In the three species, two fisheries model with parametric time-variation, the performance gap between DRL and MSY is broadened.
+In this case, as in *3)*, constant escapement strategies perform significantly worse than DRL and MSY.
+
+Our findings paint a picture of how a single-species optimal management strategy may loose performance rather dramatically when controlling a more complex ecosystem.
+Moreover, rather unintuitively, management strategies, such as MSY, known to be suboptimal and unsustainable in the single-species can turn out to even outperform the aforementioned single-species optimal strategy (e.g.
+constant escapement).
+Finally, within this regime of complex, possibly varying, ecosystems, we show that DRL consistently finds a policy which effectively either matches the best classical strategy, or outperforms it.
+We strengthen this result with a stability investigation: we show that random perturbations in the model's parameter values used do not significantly vary the conclusion that DRL outperforms constant escapement.
+<!-- ML: I think would be clearest to specify "in the environment's/dynamics model's param..."-->
+
+We notice that both classical management strategies are faced with a tradeoff -- optimal classical strategies accept a rather large fraction of episodes ending with a population crash, whose negative reward is counteracted with a higher economic output.\footnote{
 In our mathematical formulation of the decision problem, we have assumed for simplicity that the fishing effort cost is zero and that fish price is stable over time.
@@
 }
-Conversely, more conservative, lower, mortality rates lead to lower total reward on average.
-The DRL approach side-steps this trade-off by optimizing over a more complex family of possible policies--—policies parametrized by a neural network, as opposed to policies labeled by a single parameter (the mortality rate).
-
-Our findings paint a picture of how a single-species optimal management strategy may loose performance rather dramatically when controlling a more complex ecosystem.
-Here, DRL performs better from both economical *and* conservation points of view.
-Moreover, rather unintuitively, CMort -- known to be suboptimal and unsustainable for single-species models -- can turn out to even outperform the CEsc -- the single-species optimal strategy -- for complex ecosystems.
-Finally, within this regime of complex, possibly varying, ecosystems, we show that DRL consistently finds a policy which effectively either matches the best classical strategy, or outperforms it.
-We strengthen this result with a stability investigation: we show that random perturbations in the model's parameter values used do not significantly vary the conclusion that DRL outperforms constant escapement.
+Conversely, more conservative suboptimal classical strategies (e.g. with a higher constant escapement value) lead to lower total reward on average.
+The DRL approach side-steps this tradeoff by optimizing over a more complex family of possible policies—policies parametrized by a neural network, as opposed to policies labeled by a single parameter (the mortality rate).
+We thus observe that the DRL solution leads to a higher profitability *and* stability as compared to constant mortality policies.
 
@@
 
-Here we mathematically introduce the four fishery models for which we compare different management strategies.
+Here we introduce the four mathematical models of fisheries for which we compare different management strategies.
 In general, the class of models that appear in this context are *stochastic, first order, finite difference equations*.
@@
 *2. Perfect execution:* the action chosen by the manager is implemented perfectly (i.e., there is no noise affecting the actual value of the fishing efforts).
-
-Model 1 is a single-species classical model of ecological tipping points.
-Models 2-4 are all three-species models with similar dynamics.
+In future work, we intend to relax these assumptions to get more realistic management scenarios.
+
+One of the models we consider is a single-species classical model of ecological tipping points.
+The other three models are all three-species models with similar dynamics---differing by the amount of species that are harvested, and whether the parameters of the model are fixed or time-dependent.
+<!--ML: Triple parenthetical dash on purpose or an accident?
+Noticing this has been done before -->
 Following this logic, the first subsection will be dedicated to the single-species model and the second will focus on the three-species models.
@@
 One particularly important aspect that has garnered much attention is the potential for the kind of highly non-linear functions that can support dynamics such as alternative stable states and hysteresis.
-A seminal example of such dynamics was introduced in [@may77], using a one-dimensional model of a prey (resource) species under the pressure of a (fixed) predator.
-In the notation of eq. \eqref{eq:general model},
+A seminal example of such dynamics was introduced in [@may77], using a one-dimensional model of a prey (resource) species under the pressure of a (fixed) predator:
 \begin{align}
   \label{eq:may}
-  f_X(X_t)
-  = L(X_t;\ r, K) - \frac{\beta H X_t^2}{c^2 + X_t^2}.
+  \Delta X_t
+  = L(X_t;\ r, K) - \frac{\beta H X_t^2}{c^2 + X_t^2} + \eta_{X,t} - M_X X_t.
 \end{align}
@@
 
-The model has six parameters: the growth rate $r$ and carrying capacity $K$ for $X$, a constant population $H$ of a species which preys on $X$, the maximal predation rate $\beta$, the predation rate half-maximum biomass $c$, and the variance $\sigma_X^2$ of the stochastic term $\eta_{X,t}$.
-(Here and in the following we will center all random variables at zero.)
+The model has six parameters: the growth rate $r$ and carrying capacity $K$ for $X$, a constant population $H$ of a species which preys on $X$, the maximal predation rate $\beta$, the predation rate half-maximum biomass $c$, and the variance $\sigma_X^2$ of $\eta_{X,t}$.
+(Here and in the following we will center our random variables at zero.)
 
@@
 
-Models 2-4 are three-species models and they are all closely related---in fact, their natural dynamics (i.e. dynamics under zero harvest) is essentially given by the same equations:
+We consider three models of a three-species fishery ecosystem.
+These are all closely related---in fact, their natural dynamics (i.e. dynamics under zero harvest) is given by the same equations:
 \begin{align}
@@
 \begin{split}
-  f_X(N_t) &= L(X_t;\ r_X, K_X) - F(X_t, Z_t;\ \beta, c) - c_{XY}X_tY_t, \\
-  f_Y(N_t) &= L(Y_t;\ r_Y, K_Y) - D F(Y_t, Z_t;\ \beta, c) - c_{XY}X_tY_t,\\
-  f_Z(N_t) &= \left(f(X_t + DY_t\right) - d_Z)Z_t.
+  \Delta X_t &= L(X_t;\ r_X, K_X) - F(X_t, Z_t;\ \beta, c) - c_{XY}X_tY_t + \eta_{X,t},\\
+  \Delta Y_t &= L(Y_t;\ r_Y, K_Y) - D F(Y_t, Z_t;\ \beta, c) - c_{XY}X_tY_t + \eta_{Y,t},\\
+  \Delta Z_t &= \left(f(X_t + DY_t\right) - d_Z)Z_t + \eta_{Z,t}.
 \end{split}
 \end{align}
-The three species modelled are $X$, $Y$ and $Z$.
+The three species modeled are $X$, $Y$ and $Z$.
 Species $Z$ preys on both $X$ and $Y$, while the latter two compete for resources.
@@
 
+Eqs. \eqref{eq:3d model} are a generalization of the single-species model \eqref{eq:may}: this can be seen by setting 
+\begin{align*}
+  c_{XY} = d = f = \sigma_Z = 0.
+\end{align*}
+Throughout most of our paper, we use the parameter values given in eq. \eqref{eq:param vals}.
+In particular, $c_{XY} = d = f = 0.1$ and $\sigma_Z=0.05$ so, in broad strokes, our three-species model is a slight perturbation of the single-species model.
+Put differently, we are in a regime where, just by examining parameters, it would seem that the single-species model is a reasonable approximation.
+**Is this actually true? Should I maybe erase these last sentences?** <!-- ML: cc Carl, but I think these sentences seem reasonable-->
+
 The three models will branch off of eq. \eqref{eq:3d model} in the following way.
-*Model 2:* here, only $X$ is harvested, that is, in the notation of eq. \eqref{eq:general model}, we fix $M_Y=M_Z=0$ and leave $M_X$ as a control variable.
-All parameters here are constant.
-*Model 3:* as Model 1, but with $X$ and $Y$ being harvested. 
+*1. Three species, one fishery:* in this first model, only $X$ is harvested, that is, in the notation of eq. \eqref{eq:general model}, we fix $M_Y=M_Z=0$ and leave $M_X$ as a control variable.
+All parameters here are constant, with values given by \eqref{eq:param vals}.
+*2. Three species, two fisheries:* as the previous case but with $X$ and $Y$ being harvested. 
 In other words, we set $M_Z=0$ and leave $M_X$ and $M_Y$ as control variables.
-*Model 4:* here $X$ and $Y$ are harvested, but now we include a non-stationary parameter:
+Finally, *3. Three species, two fisheries, time-variation:* in this model we allow for the harvest of $X$ and $Y$, as in the previous model, but now we include a time dependent parameter:
 \begin{equation}
@@
 \end{equation}
-All other parameters are constant.
+All other parameters are fixed by eq. \eqref{eq:param vals}.
+
 Eq. \eqref{eq:rx(t)} is intended to reflect in a simple manner a possible effect of climate change: where the reproductive rate of $X$ is reduced linearly over time until it stabilizes.
+In future work we intend to test the DRL framework in realistic models of particular fisheries.
+As a part of that effort, we will attempt to capture more detailed models of the effects of climate change.
+For this exploratory work we have chosen this simple expression, however, for the sake of conceptual clarity.
 
@@
 All applications of RL can be conceptually separated into two parts: an *agent* and an *environment* which the agent interacts with.
-That is, the agent takes actions within the environment.
+That is, the agent performs actions within the environment.
 As a result of an agent's action, the environment will transition to a new state and return a numerical *reward* to the agent.
@@
 
-For this paper, we will consider four environments corresponding to each of the models considered (Table \ref{tab:models}).
-At each time step, the agent observes the state $S$ and enacts some harvest---reducing $X_t$ to $X_t - M_X(N_t)\cdot X_t$, and, for Models 3 and 4, also reducing $Y_t$ to $Y_t - M_Y(N_t)\cdot Y_t$.
+For our use case, the environment will be a computational model of the population dynamics of a fishery, with the environment state at time $t$ being a vector of all the fish populations, e.g.
+$N_t=(X_t,Y_t,Z_t)$ in the three species case.
+At each time step, the agent observes the state $S$ and enacts some harvest---reducing $X_t$ to $X_t - M_X(N_t)\cdot X_t$, and, in the two-fishery scenario, also reducing $Y_t$ to $Y_t - M_Y(N_t)\cdot Y_t$.
 Here the fish mortality-rates-from-harvest (i.e. $M_X=M_X(N_t)$ and $M_Y=M_Y(N_t)$), are the agent's action at time $t$. 
-This secures a reward of $M_X(N_t)X$ for Models 1 and 2, and, similarly, a reward of $M_X(N_t)X + M_Y(N_t)Y$ for Models 3 and 4.
-After this harvest portion of the time step, the environment changes according to its natural dynamics (see eq. \eqref{eq:3d model}).
-
-As mentioned, discretising time allows a simplification of the possibly seasonal mating behavioral patterns of the species involved. 
+This secures a reward of $M_X(N_t)X$ in the single fishery case, and in the two-fishery case the reward is $M_X(N_t)X + M_Y(N_t)Y$.
+After this harvest portion of the time step, the environment changes according to its natural dynamics given by \eqref{eq:3d model}.
+
+As mentioned previously, this discretized time framework allows a simplification of the possibly seasonal mating behavioral patterns of the species involved. 
 This approximation is commonly used in fisheries for species with annual reproductive cycles (see e.g. [@mangel2006theoretical], Chap. 6).
@@
 This formalization is rather flexible and allows one, e.g., to account for situations where the agent may not fully observe the environment state, or where the only observations available to the agent are certain functions of the underlying state.
-For the sake of clarity, we will only present here the class of POMDPs which are relevant to our work: *fully observable MDPs with a trivial emission function* (henceforth MDPs for short).
+For the sake of clarity, we will only present here the subclass of POMDPs which are relevant to our work: *fully observable MDPs* (henceforth MDPs for short).
 An MDP may be defined by the following data:
@@
 
-The policy function $\pi$ often lives in a high or even infinite-dimensional space.
+The optimal policy function $\pi$ often lives in a high or even infinite-dimensional space.
 This makes it unfeasible to directly optimize $\pi$.
@@
 }
-<!-- ML: I am not following this line of thought here.
-I see the main reason why NN's are used as being that NN's are universal function approximators, and we don't know what the policy and/or value function will look like.
-
-FM: my take here is that the space of functions from states to actions is in principle infinite-dimensional or at least very high dimensional.
-(e.g.
-exponentially high dimensional in terms of the number of degrees of freedom in state space, for the case of discretized state spaces).
-By restricting our ansatz to neural networks of a fixed depth/breadth, we're restricting ourselves to a finite dimensional (or, in the discrete case, potentially much lower dimensional) manifold of policies over which gradient descent may be performed much more efficiently.
-E.g.
-the dimensionality of the space we're searching for is determined by the size of the neural network, it's finite dimensional for continuous state spaces, and its dimension is not necessarily exponential in the number of species.
--->
 Deep reinforcement learning uses this strategy, focusing on function families parameterized by neural networks.
@@
 Within the DRL literature there is a wealth of algorithms from which to choose from to optimize $\pi$, each with its pros and cons.
-Most, if not all, of these are based on gradient ascent by using the technique of *back-propagation* to efficiently compute the gradient.
-Here we have used only one such algorithm (*proximal policy optimization (PPO)*) for the sake of particularity---to draw a clear comparison between the RL-based and the classical fishery management approaches.
+Most of these are based on gradient ascent by using the technique of *back-propagation* to efficiently compute the gradient.
+Here we have used only one such algorithm (*proximal policy optimization (PPO)*) to draw a clear comparison between the RL-based and the classical fishery management approaches.
 In practice, further improvements can be expected by a careful selection of the optimization algorithm. 
@@
 Indeed, models for ecosystem dynamics are only ever approximate and incomplete descriptions of reality. 
-This way, it is plausible that model-free RL controls outperform currently used model-based controls in ecological management problems.
+This way, it is plausible that model-free RL controls could outperform currently used model-based controls in ecological management problems.
 
@@
 \textbf{The Environment.}
-We considered the problem of managing four increasingly complex models (see Table \ref{tab:models}).
+As seen in Fig.
+\ref{fig:conceptual-v2}, we consider four different models of the ecosystem with which the fishery interacts, each of these is called an *environment*, alluding to the environment in an RL setting.
+We summarize these models in Table \ref{tab:models}.
 To recap, these four environments are a single-species growth model \eqref{eq:may} for a harvested species; a three-species model \eqref{eq:3d model} with a single harvested species; the same three-species model but with *two* harvested species; and, finally, a three-species model with a time-varying parameter and two harvested species.
 
-The policies explored were functions from states $N_t$ to either a single number $M_X(N_t) = \pi(N_t)$ (for Models 1 and 2), or a pair of numbers $(M_X(N_t), M_Y(N_t)) = \pi(N_t)$ (for Models 3 and 4).
-
-Our goal was to evaluate the performance of different policy strategies over a specified window of time.
-We chose this window to be 200 time-steps, where each discrete time-step represents the dynamical evolution of the system over a year.
-Each time-step was composed of two parts: 
-First, a *harvest period* where the harvest is collected from the system (e.g. the population $X$ is reduced to $X_t\mapsto X_t - M_X(N_t) X_t$).
-Second, a *recruitment and interaction period*, where the system's state evolves according to its natural dynamics.
-
-Training and evaluating management strategies was done by simulating *episodes.* 
-An episode begins at a fixed initial state and the system is controlled with a management policy until $t=200$, or until a ‘‘near-extinction event’’ occurs---that is, until any of the populations go below a given threshold,
+The policies explored were functions from states $N_t$ to either a single number $M_X(N_t) = \pi(N_t)$, or a pair of numbers $(M_X(N_t), M_Y(N_t)) = \pi(N_t)$ depending on the environment considered.
+
+\begin{table}
+  \begin{center}
+    \begin{tabular}{|l|c|c|c|c|}
+      \hline
+      \textbf{Model name} & \textbf{Model eqs.} & \textbf{N. Sp.} & \textbf{Harv. Sp.} & \textbf{Stationary?}\\
+      \hline
+      1 species, 1 fishery & \eqref{eq:may} & 1 & $X$ & Yes \\
+      3 species, 1 fishery & \eqref{eq:3d model} & 3 & $X$ & Yes \\
+      3 species, 2 fishery & \eqref{eq:3d model} & 3 & $X$ and $Y$ & Yes \\
+      3 species, 2 fishery, non-stationary & \eqref{eq:3d model} & 3 & $X$ and $Y$ & No \\
+      \hline
+    \end{tabular}
+    \caption{Table of models considered in this paper.
+Here, \textbf{N.
+Sp.} is the \emph{number of species} in the model, \textbf{Harv.
+Sp.} is the species of the model which are harvested, and \textbf{Stationary?} refers to whether the parameters of the model have fixed values (or, on the contrary, if they vary in time).
+The only non-stationary case presented in the paper is where $r_X$ drifts linearly with time.
+In the code repository associated to the paper, we consider other possible choices of non-stationarity.}
+    \label{tab:models}
+  \end{center}
+\end{table}
+
+Our goal is to evaluate the performance of different policies over a specified window of time.
+We have chosen this window to be 200 time-steps, where each discrete time-step represents the dynamical evolution of the system over a year.
+_Episodes_ begin at time-step zero and are controlled by the policy and the system's natural evolution until either the maximum time-span is reached, or until a ‘‘near-extinction event’’ occurs---that is, until any of the populations go below a given threshold,
 \begin{align}
@@
 \end{align}
-In our setting we have chosen $X_{\text{thresh.}}=Y_{\text{thresh.}}=Z_{\text{thresh.}}=0.05$ as a rule of thumb---given that under natural (unharvested) dynamics the populations range within values of 0.5 to 1, this would represent on the order of a 90-95% population decrease from their ‘‘natural’’ level.
-
-The reward function defining our policy optimization problem had two components.
-The first was economical: the total biomass harvested over an episode.
-The second seeked to reflect conservation goals: if a near-extinction event occurred at time $t$, the episode was ended early and a negative reward of $-100/t$ was awarded.
-This reward function balanced the extractive motivation of the fishery with conservation goals which went beyond the scope of long-term sustainable harvests.
+In our setting we have chosen $X_{\text{thresh.}}=Y_{\text{thresh.}}=Z_{\text{thresh.}}=0.05$ as a rule of thumb---given that under natural (unharvested) dynamics the populations range within values of 0.5 to 1, this would represent on the order of a 90-95% population decrease from its ‘‘natural’’ level.
+<!-- ML: Should we talk about units somewhere, seeing "0.5 to 1" might be startling to some people here-->
+
+The reward function defining our policy optimization problem has two components.
+The first is economical: the total biomass harvested over an episode.
+The second seeks to reflect conservation goals: if a near-extinction event occurs at time $t$, the episode terminates early and a negative reward of $-100/t$ is received.
+This reward function balances the extractive motivation of the fishery with conservation goals that go beyond the scope of long-term sustainable harvests.
 
@@
 The state space used was normalized case-by-case as follows: 
-*Model 1:* a line segment $[0,1]$, 
-*Models 2-4:* a cube $[0,1]^{\times 3}$.
-We used simulated data to derive a bound on the population sizes typically observed, and thus be able to normalize states to a finite volume.
-
-Policies obtained from the PPO algorithm can be ‘‘noisy’’ as their optimization algorithm is randomized (see, e.g. App. B for a visualization of the PPO policy obtained for Model 4).
-We smoothed this policy out using a Gaussian process regressor interpolation. 
-Details for this interpolation process can be found in App. C.
+*1)* for the single species model, a line segment $[0,1]$, *2)* for all three-species models, a cube $[0,1]^{\times 3}$.
+We used simulation data to derive a bound on the population sizes typically observed, and thus be able to normalize states to a finite volume.
+
+Policies obtained from the PPO algorithm can be ‘‘noisy’’ as their optimization algorithm is randomized (see, e.g.
+App.
+B for visualizations of the PPO policy in the ‘‘3 species, 2 fisheries, non-stationary’’ case).
+We perform some interpolation using a Gaussian process regressor in order to de-noise these policies.
+Specifically, we use the *GaussianProcessRegressor* object of the *sklearn* Python library with a kernel given by
+\begin{align*}
+  \text{RBF}(\text{length scale = 10}) + \text{WhiteNosie}(\text{noise level = 0.1}).
+\end{align*}
+This interpolation method is applied to scatter data of the PPO policy evaluated on 3 different grids:
+One $51\times 51 \times 5$, one $51\times 5 \times 51$ and one $5\times 51 \times 51$ grid on $(X,Y,Z)$ states.\footnote{
+The raw dataset is found at the data/results\_data/2FISHERY/RXDRIFT subdirectory in the repository with the source code and data linked above.
+Scatter plots visualizing this policy are shown in App. B.
+}
+The length scale and noise level values of this kernel were chosen arbitrarily---no hyperparameter tuning was needed to produce satisfactory interpolation, as will be shown in the results section.
 
@@
 First, optimal constant escapement performs considerably worse than not only DRL-based methods, but also than MSY.
-This conceptually shows how one-dimensional intuitions on ecological control may loose validity for complex ecosystems. **(this might be unwelcome by the reviewers if not discussed properly)**
+This conceptually shows how one-dimensional intuitions on ecological control may lose validity for complex ecosystems. **(this might be unwelcome by the reviewers if not discussed properly)**
 Second, we see that classical strategies have relatively spread out reward distributions as compared to DRL methods.
@@
 #| label: ephistograms
-#| fig.cap: "Hisotgrams of episode lengths and rewards for the four different management strategies considered.
+#| fig.cap: "Histograms of episode lengths and rewards for the four different management strategies considered.
 Only the first 50 evaluation episodes (from a total of 100) were included, for ease of visualization."
@@
 We expect a similar and, possibly, more pronounced effect for the constant escapement strategy but do not analyze this case here.
-Here, two quantities are plotted: the fraction of evaluation episodes of maximal length (i.e. episodes with no near-extinction events) is and the average reward obtained.
+Here, two quantities are plotted: the fraction of evaluation episodes of maximal length (i.e. episodes with no near-extinction events) and the average reward obtained.
 On the x-axis we have several sub-optimal constant mortality strategies that err on the conservative side: e.g. the policy labeled ‘‘80% MSY’’ has the form
@@
 \end{align*}
-where $(M_X^*,\ M_Y^*)$ is the optimal constat mortality strategy.
+where $(M_X^*,\ M_Y^*)$ is the optimal constant mortality strategy.
 We see that sufficiently conservative policies attain high sustainability, but only at a high price in terms of profit.\footnote{
@@
 Here, the key difference is that the escapement value for each of the fished species is sensitive to variations in the other populations. 
-This can be seen as a color gradients in the plots of $(X, M_X)$ and $(Y, M_Y)$, where the gradient corresponds to differing values of $Z$.
+This can be seen as color gradients in the plots of $(X, M_X)$ and $(Y, M_Y)$, where the gradient corresponds to differing values of $Z$.
 Moreover, this can be seen as an anti-correlation in the plot of $(X, M_Y)$---for optimal constant escapement, $M_Y$ is uncorrelated to $X$.
@@
 
-While the optimal control for our single-species model \eqref{eq:may} can not be easily proven to be constant escapement (since the right-hand side of that equation is not concave), from experience we can expect for constant escapement to either be optimal or near-optimal.
+While the optimal control for our single-species model \eqref{eq:may} can not be easily proven to be constant escapement (since the right-hand side of that equation is not concave), from experience we can expect constant escapement to either be optimal or near-optimal.
 We give evidence for this intuition by showing that our DRL method recovers a constant escapement solution when trained.
@@
 There is a presence of certain high-mortality points at low $X$ values in the PPO policy (which in turn generates a rising fishing mortality for $X$ values below a certain threshold in the PPO+GP policy).
-This is likely due to experience of near-extinctions early on in the training process---where, given an impending extinction, it is more rewardful to fish intensively. 
+This is likely due to experience of near-extinctions early on in the training process---where, given an impending extinction, there is a higher reward for intensive fishing. 
 These ‘‘jitters’’ are likely not fully erased through the optimization algorithm since near-extinction events become extremely rare after only a few training iterations.
@@
 below the optimal escapement value), on which no fishing is performed.
-That is, there exists some sufficiently large $\varepsilon$ such that, if $X_{\text{thresh.}}$ is the optimal escapement value of the system, then $\pi_{PPO}(X)=0$ for all $X\in[X_{\text{thresh.}}-\varepsilon,\ X_{\text{thresh.}}]$.
+That is, there exists some sufficiently large $\varepsilon$ such that if $X_{\text{thresh.}}$ is the optimal escapement value of the system, then $\pi_{PPO}(X)=0$ for all $X\in[X_{\text{thresh.}}-\varepsilon,\ X_{\text{thresh.}}]$.
 
@@
 In this section we present results intended to show that the effects that we observe in this paper are not the result of a careful selection of parameter values, but rather arise for a wide variety of parameter values.
+<!--ML: For clarity, I think it would be good to specify that you are referring to parameters of the dynamics model.-->
 
@@
 
-## Pasted to-be-discussed
-
-Eqs. \eqref{eq:3d model} are a generalization of the single-species model \eqref{eq:may}: this can be seen by setting 
-\begin{align*}
-  c_{XY} = d = f = \sigma_Z = 0.
-\end{align*}
-Throughout most of our paper, we use the parameter values given in eq. \eqref{eq:param vals}.
-In particular, $c_{XY} = d = f = 0.1$ and $\sigma_Z=0.05$ so, in broad strokes, our three-species model is a slight perturbation of the single-species model.
-
-## Discussion
-
 Fisheries are complex ecosystems, with interactions between species leading to highly non-linear dynamics.
 While current models for population estimation include many important aspects of this complexity, it is still common to use simplified dynamical models in order to guide decision making.
-This provides very interpretable solutions, such as constant mortality policies.
+<!-- ML: First half of this sentence comes off as vague, esp.
+without a citation -->
+This provides easily interpretable solutions, such as constant mortality policies.
 There is a drawback here, however: due to the simplicity of these dynamical models, the policies might not respond effectively in many situations---situations where the predictions of these simple models deviate considerably from reality.
@@
 We propose an alternative approach to the problem of fishery control: to use a more expressive -- albeit more complicated -- dynamical model to guide decision making.
-Furthermore, rather than computing the optimal control policy for the model (something that is impossible in practice for complex systems) we use deep reinforcement learning to obtain a ‘‘pretty darn good’’ policy.
+Furthermore, rather than computing the optimal control policy for the model (something that is impossible in practice for complex systems), we use deep reinforcement learning to obtain a ‘‘pretty darn good’’ policy.
 This policy is estimated in a *model-free* setting, i.e., the agent treats the dynamical model (e.g. eq. \eqref{eq:3d model}) as a black box of input-output pairs.
@@
 This is the case since these policy functions are only specified by one or two parameters (respectively in the single fishery, or two fishery cases).
-In contrast, DRL optimizes over the more expressive -- and more complicated -- family of policies parametrized by a neural network.
+In contrast, DRL optimizes over the more expressive -- and more complicated -- family of policies parameterized by neural networks.
 Because of this, our DRL approach uses a variant of *gradient ascent* (the proximal policy optimization algorithm) to obtain a good candidate policy.
+<!-- ML: I don't think this sentence is the most helpful here.
+Might just be better to say something like how NN's do well in approximating just about any function empirically-->
 
 We showed that for sufficiently complex management scenarios -- with multiple harvested species and, possibly, time-varying parameters -- DRL-based management strategies perform significantly better than constant escapement.
-This with respect to both average rewards received and conservation goals.
+This is with respect to both average rewards received and conservation goals.
 In this sense, an approximate solution to a more complicated and expressive model, can outperform the optimal solution of the single-species problem---even when the parameters of the single-species solution are empirically optimized.
@@
 three sigmas) in the parameter values to appear in these statistics.
+<!--ML: Great to have this paragraph in here, but it doesn't strike me as the most satisfying one to end on.
+Might be better to end on a paragraph that points to future directions or gives a high-level review/takeaway.
+-->
 
@@
 
-## Gaussian process interpolation
-
-Here we summarize the procedure used to interpolate the PPO policy (visualized in Fig. \ref{fig:ppo-policy}).
-We use the *GaussianProcessRegressor* object of the *sklearn* Python library with a kernel given by
-\begin{align*}
-  \text{RBF}(\text{length scale = 10}) + \text{WhiteNosie}(\text{noise level = 0.1}).
-\end{align*}
-This interpolation method is applied to scatter data of the PPO policy evaluated on 3 different grids on $(X,Y,Z)$ states:
-$G_X$, a $51\times 5 \times 5$ grid; $G_Y$, a $5\times 51 \times 5$ grid; and $G_Z$, a $5\times 5 \times 51$ grid.
-This combination of grids was used instead of a single dense grid in order to reduce the computational intensity of the interpolation procedure.
-For $G_X$, the 5 values for $Y$ and $Z$ were varied in a ‘‘popular window,’’ i.e. episode time-series data was used to determine windows of $Y$ and $Z$ values which were most likely. 
-The grids $G_Y$ and $G_Z$ were generated in a similar fashion, *mutatis mutandis*.\footnote{
-The raw dataset is found at the data/results\_data/2FISHERY/RXDRIFT sub-directory in the repository with the source code and data linked above.
-Scatter plots visualizing this policy are shown in App. B.
-}
-The length scale and noise level values of this kernel were chosen arbitrarily---no hyperparameter tuning was needed to produce satisfactory interpolation, as will be shown in the results section.
-
 # References

### 2023-06-29 manuscript/manuscript.Rmd

--- 2023-06-28 manuscript/manuscript.Rmd
@@
 \end{align}
-The three species modelled are $X$, $Y$ and $Z$.
+The three species modeled are $X$, $Y$ and $Z$.
 Species $Z$ preys on both $X$ and $Y$, while the latter two compete for resources.
@@
 
-As mentioned, discretising time allows a simplification of the possibly seasonal mating behavioral patterns of the species involved. 
+As mentioned previously, discretising time allows a simplification of the possibly seasonal mating behavioral patterns of the species involved. 
 This approximation is commonly used in fisheries for species with annual reproductive cycles (see e.g. [@mangel2006theoretical], Chap. 6).
@@
 
-Training and evaluating management strategies was done by simulating *episodes.* 
+Training and evaluating management strategies were performed by simulating *episodes.* 
 An episode begins at a fixed initial state and the system is controlled with a management policy until $t=200$, or until a ‘‘near-extinction event’’ occurs---that is, until any of the populations go below a given threshold,
@@
 In our setting we have chosen $X_{\text{thresh.}}=Y_{\text{thresh.}}=Z_{\text{thresh.}}=0.05$ as a rule of thumb---given that under natural (unharvested) dynamics the populations range within values of 0.5 to 1, this would represent on the order of a 90-95% population decrease from their ‘‘natural’’ level.
+<!-- ML: Should we talk about units somewhere, seeing "0.5 to 1" might be startling to some people here-->
 
@@
 While current models for population estimation include many important aspects of this complexity, it is still common to use simplified dynamical models in order to guide decision making.
-This provides very interpretable solutions, such as constant mortality policies.
+<!-- ML: First half of this sentence comes off as vague, esp.
+without a citation -->
+This provides easily interpretable solutions, such as constant mortality policies.
 There is a drawback here, however: due to the simplicity of these dynamical models, the policies might not respond effectively in many situations---situations where the predictions of these simple models deviate considerably from reality.

### 2023-06-29 manuscript/manuscript.Rmd
started future directions

--- 2023-06-28 manuscript/manuscript.Rmd
@@
 
-Our findings paint a picture of how a single-species optimal management strategy may loose performance rather dramatically when controlling a more complex ecosystem.
+Our findings paint a picture of how a single-species optimal management strategy may lose performance rather dramatically when controlling a more complex ecosystem.
 Here, DRL performs better from both economical *and* conservation points of view.
@@
 
-## Pasted to-be-discussed
-
-Eqs. \eqref{eq:3d model} are a generalization of the single-species model \eqref{eq:may}: this can be seen by setting 
-\begin{align*}
-  c_{XY} = d = f = \sigma_Z = 0.
-\end{align*}
-Throughout most of our paper, we use the parameter values given in eq. \eqref{eq:param vals}.
-In particular, $c_{XY} = d = f = 0.1$ and $\sigma_Z=0.05$ so, in broad strokes, our three-species model is a slight perturbation of the single-species model.
-
-Here we note two surprising things:
-First, optimal constant escapement performs considerably worse than not only DRL-based methods, but also than MSY.
-This conceptually shows how one-dimensional intuitions on ecological control may loose validity for complex ecosystems. **(this might be unwelcome by the reviewers if not discussed properly)**
-Second, we see that classical strategies have relatively spread out reward distributions as compared to DRL methods.
-
 ## Discussion
@@
 In contrast, DRL optimizes over the more expressive -- and more complicated -- family of policies parametrized by a neural network.
-Because of this, our DRL approach uses a variant of *gradient ascent* (the proximal policy optimization algorithm) to obtain a good candidate policy.
-
-We showed that for sufficiently complex management scenarios -- with multiple harvested species and, possibly, time-varying parameters -- DRL-based management strategies perform significantly better than constant escapement.
+Neural networks are often used as flexible function approximators that can be efficiently optimized over.
+
+We showed that for sufficiently complex management scenarios -- Models 3 and 4 -- DRL-based management strategies perform significantly better than CEsc.
 This with respect to both average rewards received and conservation goals.
@@
 
-What is more, this DRL advantage happened within a regime of parameter space where species interaction terms were much smaller than the single-species logistic growth terms.
-Indeed, interaction parameters in \eqref{eq:3d model} are about an order of magnitude smaller than $r_X$ and $r_Y$.
-In this situation logistic growth is a reasonable approximation of the system, and thus a constant escapement solution is a plausible control strategy.
-
-We found that the optimal constant mortality policy surprisingly performs much better than constant escapement (Fig. \ref{fig:main}).
+We found that the optimal CMort policy surprisingly performs much better than constant escapement (Fig. \ref{fig:main}).
 However, it can be observed in Fig.
@@
 
+The question of when multi-species models are well-approximated by single-species models was studied in detail in [@burgess].
+Here our approach is dual to that of the aforementioned paper.
+Rather than first optimizing a single-species model to approximate a more complex model and then finding the MSY value for the single-species model, we used simulated data to optimize CMort and CEsc directly on the three-species model.
+We do not investigate further whether our three-species models are well approximated by a single-species model in the sense of [@burgess].
+However:
+*1)* Because the interaction terms in \eqref{eq:3d model} are about an order of magnitude smaller than $r_X$ and $r_Y$, Models 2-4 are ‘‘close’’ in parameter-space to a single-species model.
+*2)* The fact that for Model 2 all strategies match in performance suggests that \eqref{eq:3d model} might be well-approximated by a single species model.
+This in turn suggests that the reason that DRL outperforms both single-species strategies for Models 3 and 4 is not due a lack of a single-species approximation for either $X$ or $Y$, but due to the complexity of having two harvested species.
+Moreover, non-stationarity exacerbated the advantage of DRL over CEsc and CMort, which can be expected as non-stationary parameters have been observed to introduce bias into single-species reference point estimation [@burgess].
+
 Finally, we performed a stability analysis to ensure that the advantage of DRL-based techniques over constant escapement is a ubiquitous phenomenon and not a result of a lucky selection of parameter values.
@@
 three sigmas) in the parameter values to appear in these statistics.
+
+## Future directions
+
+There are a number of interesting directions that we intend to explore in future work.
+
+First, it is to benchmark our results on increasingly more compex and realistic fishery models.
+This would include non-stationarities in the dynamical parameters to accurately reflect the effects of climate change on the ecosystem.
+This added complexity would likely pose a computational challenge---in future work we will likely need to test several different DRL training algorithms (see e.g.
+[@rl-intro] for a non-exhaustive list), and it is very likely that hyperparameter tuning will need to be performed.
+Moreover, it may be that larger neural networks than the one we used in this research will be needed for the policy function.
+This all will mean that considerable technical work will be needed in order to make this next step computationally feasible (e.g. we might need to make more extensive use of GPUs and parallelization).
 
@@
 
-# Appendix
-
-## Results for stationary models
+# Appendix: Results for stationary models
 
@@
 
-## PPO policy function for non-stationary model
+# Appendix: PPO policy function for non-stationary model
 
@@
 
-## Gaussian process interpolation
+# Appendix: Gaussian process interpolation
 

### 2023-06-30 manuscript/manuscript.Rmd
draft of future directions

@@
 
-There are a number of interesting directions that we intend to explore in future work.
-
-First, it is to benchmark our results on increasingly more compex and realistic fishery models.
+There are a number of interesting directions that would be interseting to explore in future work.
+
+First, benchmarking our results against increasingly more compex and realistic fishery models.
 This would include non-stationarities in the dynamical parameters to accurately reflect the effects of climate change on the ecosystem.
@@
 This all will mean that considerable technical work will be needed in order to make this next step computationally feasible (e.g. we might need to make more extensive use of GPUs and parallelization).
+
+Second, to account for noisy estimates of the system's state and imperfect policy implementation.
+This could be done straightforwardly, albeit it might increase the training time before DRL approaches converge, as well as introducing the need for hyperparameter tuning.
+
+Third, to account for the systematic uncertainties behind the dynamics of the ecosystem---that is, to account for model biases with respect to reality.
+Here, one can employ tools from curriculum learning in order to train an agent that is *generally capable* of good management over a range of different dynamical models.
+This way, one can incorporate different models---expressing different aspects of the ecosystem---into the learning process of the agent.
+We believe that this step will likely be necessary if DRL algorithms are to be applied successfully in the fishery management problem.
+Curriculum learning is rather expensive computationally, however, and involves a non-trivial *curriculum design* which will guide the agent in its learning process.
+This way, considerable technical work would be needed for this direction.
 

### 2023-06-30 manuscript/manuscript.Rmd
mini

@@
 #| fig.cap: "Reward distributions for the four strategies considered.
-These are based on 100 evaluation episodes of the ‘‘three species, two fisheries’’ model in Table 1.
+These are based on 100 evaluation episodes of Model 3 in Table 1.
 We denote CEsc for constant escapement, CMort for constant mortality, PPO for the output policy of the PPO optimization algorithm, and PPO GP for the Gaussian process interpolation of the PPO policy."
@@
 #| fig.cap: "Reward distributions for the four strategies considered.
-These are based on 100 evaluation episodes of the ‘‘three species, one fishery’’ model in Table 1.
+These are based on 100 evaluation episodes of Model 2 in Table 1.
 We denote CEsc for constant escapement, CMort for constant mortality, PPO for the output policy of the PPO optimization algorithm, and PPO GP for the Gaussian process interpolation of the PPO policy."
@@
 #| fig.cap: "Reward distributions for the four strategies considered.
-These are based on 100 evaluation episodes of the ‘‘one species, one fishery’’ model in Table 1.
+These are based on 100 evaluation episodes of Model 1 in Table 1.
 We denote CEsc for constant escapement, CMort for constant mortality, PPO for the output policy of the PPO optimization algorithm, and PPO GP for the Gaussian process interpolation of the PPO policy."

### 2023-06-30 manuscript/manuscript.Rmd
abbys comments

@@
 
-- $T(N_{t+1}|N_t, a_t)$: *transition operator*, a conditional distribution which describes the dynamics of the system (where $N_i\in\mathcal{S}$ are states of the environment),
-
-- $r(N_t, a_t)$: *reward function*, the reward obtained after performing action $a_t\in\mathcal{A}$ in state $N_t$,
+- $T(N_{t+1}|N_t, a_t, t)$: *transition operator*, a conditional distribution which describes the dynamics of the system (where $N_i\in\mathcal{S}$ are states of the environment),\footnote{
+Transition operators are commonly discussed without having a direct time-dependence for simplicity, but the inclusion of $t$ as an argument to $T$ does not alter the structure of the learning problem appreciably.
+}
+
+- $r(N_t, a_t, t)$: *reward function*, the reward obtained after performing action $a_t\in\mathcal{A}$ in state $N_t$,
 
@@
   d(N_0) \prod_{t=0}^{ t_{\text{fin.}}-1 }
-  \pi( a_t | N_t ) T( N_{t+1} | N_t, a_t ).
+  \pi( a_t | N_t ) T( N_{t+1} | N_t, a_t, t ).
 \end{align*}
@@
 $$
-  R(\tau) = \sum_{t=0}^{ t_{\text{fin.}}-1 } \gamma^t r(a_t, N_t),
+  R(\tau) = \sum_{t=0}^{ t_{\text{fin.}}-1 } \gamma^t r(a_t, N_t, t),
 $$
@@
 The second seeked to reflect conservation goals: if a near-extinction event occurred at time $t$, the episode was ended early and a negative reward of $-100/t$ was awarded.
-This reward function balanced the extractive motivation of the fishery with conservation goals which went beyond the scope of long-term sustainable harvests.
+This reward function balanced the extractive motivation of the fishery with conservation goals which went beyond the scope of long-term sustainable harvests commonly used in fishery management.
 
@@
 #| fig.cap: "Hisotgrams of episode lengths and rewards for the four different management strategies considered.
-Only the first 50 evaluation episodes (from a total of 100) were included, for ease of visualization."
+Only the first 50 evaluation episodes (from a total of 100) were included, for ease of visualization.
+From left to right, the four management strategies compared are CEsc, CMort, PPO, and PPO+GP."
 #| out.width: "6in"
@@
   geom_col() +
-  labs(x="", y="ep. duration")
+  labs(x="", y="ep. duration") +
+  theme(plot.title = element_text(hjust = 0.5))+
+  ggtitle("CEsc")
 msy_len <- msy_ends %>%
@@
   geom_col() +
-  labs(x="", y="")
+  labs(x="", y="") +
+  theme(plot.title = element_text(hjust = 0.5))+
+  ggtitle("CMort")
 ppo_len <- ppo_ends %>%
@@
   geom_col() +
-  labs(x="", y="")
+  labs(x="", y="")+
+  theme(plot.title = element_text(hjust = 0.5))+
+  ggtitle("PPO")
 gpp_len <- gpp_ends %>%
@@
   geom_col() +
-  labs(x="", y="")
+  labs(x="", y="")+
+  theme(plot.title = element_text(hjust = 0.5))+
+  ggtitle("PPO+GP")
 
@@
 \begin{align*}
-  \text{RBF}(\text{length scale = 10}) + \text{WhiteNosie}(\text{noise level = 0.1}).
+  \text{RBF}(\text{length scale = 10}) + \text{WhiteNoise}(\text{noise level = 0.1}).
 \end{align*}

### 2023-06-30 manuscript/manuscript.Rmd

@@
 But despite these efforts and advances, we remain faced with the difficult task of figuring out how best to respond to these crises.
-While simplified models have historically allowed for exploration of large decision spaces, the new wave of rich models are applied to highly oversimplified descriptions of potential actions they seek to inform.
+While simplified process models for the population dynamics have historically allowed for exploration of large decision spaces, the new wave of rich models are applied to highly oversimplified descriptions of potential actions they seek to inform.
 For instance, Global Circulation Models (GCMs) such as HadCM3 [@pope2000impact; @gordon2000simulation; @collins2001internal] model earth's climate using 1.5M variables, while the comparably vast potential action space is modeled as one of seven emissions marker scenarios from the IPCC [@riahi2017shared].

### 2023-06-30 src/OldCode/explore.Rmd

--- 2023-03-02 src/explore.Rmd
@@
 
-ppo_df <- read_csv("../data/PPO220.csv.xz")
+ppo_df <- read_csv("../data/PPO150.csv.xz")
 
+opt_msy <- msy_df |> filter(action == best_action)
 opt_escapement <- escapement_df |> filter(escapement == best_e)
+```
+Overall mean reward
+
+```{r}
+list(
+  ppo_df |> mutate(method="PPO"),
+  opt_msy |> mutate(method="MSY"),
+  opt_escapement |> mutate(method="escapement")
+  ) |>
+  purrr::list_rbind() |>
+  group_by(rep, method) |> filter(t == max(t)) |> 
+  group_by(method) |> summarise(mean = mean(reward), sd = sd(reward))
+
 ```
@@
 ```{r}
-#| label: tuning
-#| fig.cap : "Result of the tuning procedure for the constant mortality strategy. 
-#| (A) Mean episode reward (averaged over 50 episodes) shown as scatter data, and shaded in gray are two standard deviations around each scatter point.
-Vertical lines correspond to the optimal constant mortality rate $M^* = $ 0.045 (green) and $0.8 M^*=$ 0.036 (violet).
-#| (B) Several constant mortality policies shown for illustrative purposes.
-Slopes $0.02$, $0.03$, $0.055$ and $0.065$ shown in black; the optimal linear policy $M^*X$ shown in green, while $0.8M^*X$ shown in violet."
-#| out.width : "6.5in"
 msy_rewards <- msy_df |>
@@
 
-rmsy <- msy_rewards |>
+msy_rewards |>
   ggplot(aes(action, mean_reward, 
@@
         aspect.ratio = 1) 
-
-# linear function blueprint
-linear_f <- function(slope, x) slope*x
-# generates linear function with given slope
-linear_f_gen <- function(slope) (function(x) slope*x)
-
-# dummy data set to keep ggplot happy
-p <- ggplot(data = data.frame(x = 0), mapping = aes(x = x))
-
-# p + stat_function(fun = linear_f_gen(0.5)) + xlim(0,2)
-lines <- p + stat_function(fun = linear_f_gen(0.009)) + 
-  stat_function(fun = linear_f_gen(0.018)) +
-  stat_function(fun = linear_f_gen(0.027)) +
-  stat_function(fun = linear_f_gen(0.8*best_action), col = "violetred3", size=1) +
-  stat_function(fun = linear_f_gen(best_action), col = "seagreen3", size=1) +
-  stat_function(fun = linear_f_gen(0.054)) + 
-  stat_function(fun = linear_f_gen(0.063)) +
-  stat_function(fun = linear_f_gen(0.072)) +
-  stat_function(fun = linear_f_gen(0.081)) +
-  stat_function(fun = linear_f_gen(0.090)) +
-  xlim(0,2) +
-  xlab("X") + ylab("harvest") +
-  theme_bw() +
-  theme(#panel.grid.major = element_blank(), 
-        #panel.grid.minor = element_blank(), 
-        aspect.ratio = 1)
-
-(rmsy | lines) + plot_annotation(tag_levels = c('A'))
 ```
@@
 ```{r} 
-#| label: timeseries 
-#| fig.cap: "A sample time series for the three fish populations, with the system being controlled, respectively, by the MSY optimal mortality, by $80$ percent of the MSY optimal mortality, and by DRL."
-#| out.width: "6.5in"
 policy_labels <- list("msy" = "MSY Const. Mortality", "tac"= "80% MSY Const. Mortality", "ppo"= "DRL")
@@
 ```{r} 
-#| label: rew_len_fig
-#| fig.cap: "Bar charts for the episode lengths (top panel) and episode total harvests (bottom panel) for 50 episodes." 
-#| out.width: "6.5in"
 p1 <- msy_tac_ppo |> 
     group_by(rep, policy) |>
-    filter(t==max(t), species == "X", rep < 50) |>
+    filter(t==max(t), species == "X", rep < 30) |>
     ggplot(aes(rep, t)) + 
@@
     group_by(rep, policy) |>
-    filter(t==max(t), species == "X", rep < 50) |>
+    filter(t==max(t), species == "X", rep < 30) |>
     ggplot(aes(rep, reward)) + 
@@
 ```{r}
-#| label: action-X
-#| fig.cap: "(A-B) Harvest and escapement scatter data for the DRL policy together with the optimal constant escapement (pink) and constant mortality (green) policies for comparison. 
-#| (A) includes $Y$ dependence through color, whereas (B) includes $Z$ dependence similarly.
-#| (B-C) escapement (i.e. $X_t - h_t$) scatter data for the DRL policy. 
-#| Optimal constant escapement policy shown in pink for comparison."
-#| out.width: "6.5in"
 opt_escapement_harvest <- function(X) case_when(
@@
 ```
-```{r}
-#| label: state-space
-#| fig.cap: "Policy heatmaps.
-Each policy, say, $\\pi$, was evaluated at points $(x,y,z)$ visited during any of the 50 evaluation rounds.
-We projected this data down to the $X$-$Y$ plane---i.e.
-we plotted the heatmap $\\pi(x,y,z)$ as a function of $x$ and $y$."
-#| out.width: "6.5in"
-i_tac <- which.min(abs(actions - best_action * 0.8))
-
-# color pallette limits
-myPalette <- colorRampPalette(rev(brewer.pal(11, "Spectral")))
-sc <- scale_colour_gradientn(colours = myPalette(100), limits=c(0,0.25))
-
-pmsy <- msy_df |>
-  mutate(X = X+1, Y = Y+1, Z = Z+1, harvest = action * X) |>
-  filter(action == best_action) |>
-  ggplot(aes(X, Y, color = harvest)) +
-  geom_point() + sc +
-  xlim(0,1) +
-  ylim(0,0.85) +
-  theme_bw() + 
-  ggtitle("Opt. Const. Mortality")
-
-ptac <- msy_df |>
-  mutate(X = X+1, Y = Y+1, Z = Z+1, harvest = action * X) |>
-  filter(action == actions[[i_tac]]) |>
-  ggplot(aes(X, Y, color = harvest)) +
-  geom_point() + sc +
-  xlim(0,1) +
-  ylim(0,0.85) +
-  theme_bw() + 
-  ggtitle("80% Opt. Const. Mortality")
-
-pppo <- ppo_df |>
-  mutate(X = X+1, Y = Y+1, Z = Z+1, harvest = action * X) |>
-  ggplot(aes(X, Y, color = harvest)) +
-  geom_point() + sc +
-  xlim(0,1) +
-  ylim(0,0.85) +
-  theme_bw() + 
-  ggtitle("DRL")
-
-pesc <- opt_escapement |>
-  mutate(X = X+1, Y = Y+1, Z = Z+1, 
-        harvest = case_when(
-          X - escapement > 0 ~ X - escapement,
-          X - escapement <= 0 ~  0)
-         ) |> 
-  ggplot(aes(X, Y, color = harvest)) +
-  geom_point() + sc +
-  xlim(0,1) +
-  ylim(0,0.85) +
-  theme_bw() + 
-  ggtitle("Opt. Const. Escapement")
-
-(
-  (pmsy | ptac) / (pppo | pesc) +
-  plot_layout(guides = "collect") & theme(legend.position = 'right')
-) + 
-  plot_annotation(tag_levels = c('A'))
-```
-
-<!-- Appendices -->
 
@@
 ```{r} 
-#| label: timeseries_msy
-#| fig.cap: "Nine time series of the system controlled by the MSY constant mortality policy." 
-#| out.width: "6.5in"
 msy_tac_ppo |> 
@@
 ```{r} 
-#| label: timeseries_tac
-#| fig.cap: "Nine time series of the system controlled by the policy using 80 percent of the MSY constant mortality." 
-#| out.width: "6.5in"
 msy_tac_ppo |> 
@@
 ```{r} 
-#| label: timeseries_ppo
-#| fig.cap: "Nine time series of the system controlled by the trained DRL agent." 
-#| out.width: "6.5in"
 msy_tac_ppo |> 
@@
 ```{r} 
-#| label: timeseries_esc
-#| fig.cap: "Nine time series of the system controlled by the optimal constant escapement policy." 
-#| out.width: "6.5in"
 esc_sim_df |> 
@@
 ```
-
-<!-- msy scale downs -->
-
-```{r}
-#| label: several-tac
-#| fig.cap: "Performance of constant mortality policies with mortality $\\alpha M^*$ for several values of $\\alpha$.
-#| The top panel displays the episode lengths for each repetition, while the bottom pannel displays the episode harvest.
-#| The labels in the plots are the value of $\\alpha$ expressed as a percentage."
-#| out.width: "6.5in"
-N <- 30
-actions <- unique(msy_df$action)
-i85 <- which.min(abs(actions - best_action* 0.85))
-i90 <- which.min(abs(actions - best_action* 0.90))
-i95 <- which.min(abs(actions - best_action* 0.95))
-tac_85_sim <- msy_df |> 
-    filter(action == actions[[i85]]) |>
-    pivot_longer(c("X", "Y", "Z"),
-        names_to = "species", values_to = "abundance") |>
-    mutate(abundance = abundance + 1, policy = "85%")
-tac_90_sim <- msy_df |> 
-    filter(action == actions[[i90]]) |>
-    pivot_longer(c("X", "Y", "Z"),
-        names_to = "species", values_to = "abundance") |>
-    mutate(abundance = abundance + 1, policy = "90%")
-tac_95_sim <- msy_df |> 
-    filter(action == actions[[i95]]) |>
-    pivot_longer(c("X", "Y", "Z"),
-        names_to = "species", values_to = "abundance") |>
-    mutate(abundance = abundance + 1, policy = "95%")
-
-tac_80_sim <- tac_sim |> mutate(policy = "80%")
-
-tac4 <- bind_rows(tac_80_sim, tac_85_sim, tac_90_sim, tac_95_sim)
-
-p1 <- tac4 |> 
-    group_by(rep, policy) |>
-    filter(t==max(t), species == "X", rep < N) |>
-    ggplot(aes(rep, t)) + 
-      geom_col() +
-      facet_wrap(~factor(policy), ncol = 4) +
-      labs(y = "episode length", x = "repetition") +
-      theme_bw() + 
-      theme(axis.text.x = element_blank())
-
-p2 <- tac4 |> 
-    group_by(rep, policy) |>
-    filter(t==max(t), species == "X", rep < N) |>
-    ggplot(aes(rep, reward)) + 
-      geom_col() +
-      facet_wrap(~factor(policy), ncol=4) +
-      labs(y = "episode harvest", x = "repetition") +
-      theme_bw() + 
-      theme(axis.text.x = element_blank())
-
-(p1 / p2)
-```
-
 <!-- escapement tuning and evaluation -->
@@
 ```{r}
-#| label: esc-tuning-eval
-#| fig.cap: "(Left panel) Tuning results for constant escapement policies: mean episode reward scatter data (averaged over 50 episodes) together with two standard deviations shaded in grey. 
-#| The optimal escapement level (the one producing maximum mean harvest) is $X_{\\text{esc.}}=0.62$.
-#| (Middle and right panels) Reward histograms for the optimal constant escapement policy (middle) and for the DRL policy (right).
-#| For both DRL and optimal constant escapement, no early episode ends are observed (i.e. all episodes last for 200 time steps)."
-#| out.width: "6.5in"
-#| fig.align: "center"
 esc_rewards <- escapement_df |>

### 2023-06-30 code/escapement.py
removed escapement

@@ -1,33 +1,0 @@
-import gym_fishing
-import gym
-import pandas as pd
-import numpy as np
-import ray
-
-@ray.remote
-def simulate(env, esc_level):
-  df = []
-  for rep in range(50):
-    episode_reward = 0
-    observation = env.reset()
-    for t in range(env.Tmax):
-      df.append(np.append([t, rep, esc_level, episode_reward], observation))
-      sp1_pop = (observation[0] + 1 ) # natural state-space
-      action = np.max([1 - esc_level / sp1_pop, 0]) # action is a mortality rate. 
-      observation, reward, terminated, info = env.step(action)
-      episode_reward += reward
-      if terminated:
-        break
-  return(df)
-
-env = gym.make("threeFishing-v2")
-env.training = False
-esc_choices = np.linspace(0,1,101)
-
-# define parllel loop and execute
-parallel = [simulate.remote(env, i) for i in esc_choices]
-df = ray.get(parallel)
-
-cols = ["t", "rep", "escapement", "reward", "X", "Y", "Z"]
-df2 = pd.DataFrame(np.vstack(df), columns = cols)
-df2.to_csv("data/escapement.csv.xz", index=False)

### 2023-07-01 manuscript/manuscript.Rmd
CMort CEsc throughout, burgess reference fixed

@@
 Regarding control for these 4 models, we show the following. 
-*Model 1:* DRL-based strategies recover the optimal CEsc policy function.  
-Constant mortality performs considerably worse than these three.\footnote{
+*Model 1:* DRL-based strategies recover the optimal constant escapement (CEsc) policy function.  
+Constant mortality (CMort) performs considerably worse than these three.\footnote{
 As will be explained later, all our models are stochastic. 
-If we set stochasticity to zero in Model 1, constant mortality matches the performance of the other management strategies.
+If we set stochasticity to zero in Model 1, CMort matches the performance of the other management strategies.
 }
 *Model 2:* Here, all management strategies perform similarly.
-*Model 3:* For this model, DRL outperforms both classical strategies, with CMort surprisingly performing significantly better than constant escapement.
+*Model 3:* For this model, DRL outperforms both classical strategies, with CMort surprisingly performing significantly better than CEsc.
 In particular, we observe that DRL strategies are more sensitive to stochastic variations of the system which allows it to adaptively manage the system to prevent population collapses below the threshold.
@@
 Finally, within this regime of complex, possibly varying, ecosystems, we show that DRL consistently finds a policy which effectively either matches the best classical strategy, or outperforms it.
-We strengthen this result with a stability investigation: we show that random perturbations in the model's parameter values used do not significantly vary the conclusion that DRL outperforms constant escapement.
+We strengthen this result with a stability investigation: we show that random perturbations in the model's parameter values used do not significantly vary the conclusion that DRL outperforms CEsc.
 
@@
 #| label: may
-#| fig.cap: "The fixed point diagram for model (TBD) as a function of varying the parameter $\\beta H$, assuming zero noise.
+#| fig.cap: "The fixed point diagram for the unharvested dynamics of Model 1 as a function of varying the parameter $\\beta H$, assuming zero noise.
 #| Stable fixed points (also known as attractors) are plotted using a solid line, while the unstable fixed point is shown as a dotted line."
@@
 \textbf{Tuning the CEsc strategy.}
-This tuning procedure was analogous to that of the constant mortality strategy just summarized.
+This tuning procedure was analogous to that of the CMort strategy just summarized.
 Namely:
@@
 To ensure that our results do not strongly depend on our parameter choices, we performed a stability analysis.
-Here, we perturbed parameters randomly and measured the difference in performance between our DRL-based methods and the constant escapement strategy.
+Here, we perturbed parameters randomly and measured the difference in performance between our DRL-based methods and the CEsc strategy.
 We observed that the difference in performance is maintained for even relatively high noise strengths.
@@
 To recap, the management strategies were CEsc, CMort, PPO and a Gaussian process interpolation of PPO (‘‘PPO+G’’).
-Furthermore, we characterized the trade-off between economic output and sustainability faced by constant mortality policies.
-This was done by evaluating constant mortality policies with a fraction of the optimal mortality rate (specifically, 80%, 90% and 95%).
+Furthermore, we characterized the trade-off between economic output and sustainability faced by CMort policies.
+This was done by evaluating CMort policies with a fraction of the optimal mortality rate (specifically, 80%, 90% and 95%).
 All evaluations were based on 100 replicate episodes.
@@
 #| label: fracmsy
-#| fig.cap: "Trade-off between reward and probability of a near-extinction event for constant mortality policies.
+#| fig.cap: "Trade-off between reward and probability of a near-extinction event for CMort policies.
 We evaluated policies at the full optimal constant escapement value, and also at 0.8, 0.9, 0.95 of the latter.
@@
 
-We notice that the DRL-derived policy has similarities to a constant escapement policy.
+We notice that the DRL-derived policy has similarities to a CEsc policy.
 Here, the key difference is that the escapement value for each of the fished species is sensitive to variations in the other populations. 
 This can be seen as color gradients in the plots of $(X, M_X)$ and $(Y, M_Y)$, where the gradient corresponds to differing values of $Z$.
-Moreover, this can be seen as an anti-correlation in the plot of $(X, M_Y)$---for optimal constant escapement, $M_Y$ is uncorrelated to $X$.
+Moreover, this can be seen as an anti-correlation in the plot of $(X, M_Y)$---for optimal CEsc, $M_Y$ is uncorrelated to $X$.
 
@@
 
-While the optimal control for our single-species model \eqref{eq:may} can not be easily proven to be constant escapement (since the right-hand side of that equation is not concave), from experience we can expect constant escapement to either be optimal or near-optimal.
-We give evidence for this intuition by showing that our DRL method recovers a constant escapement solution when trained.
+While the optimal control for our single-species model \eqref{eq:may} can not be easily proven to be CEsc (since the right-hand side of that equation is not concave), from experience we can expect CEsc to either be optimal or near-optimal.
+We give evidence for this intuition by showing that our DRL method recovers a CEsc solution when trained.
 These results are shown in Fig. \ref{fig:1sp-pol}
@@
 This way, the agent does not further explore that region of state space to generate new experience.
-We believe the most important aspect of the constant escapement policy reproduced by PPO is the fact that there is some sufficiently-wide window below the threshold of the policy (i.e.
+We believe the most important aspect of the CEsc policy reproduced by PPO is the fact that there is some sufficiently-wide window below the threshold of the policy (i.e.
 below the optimal escapement value), on which no fishing is performed.
@@
 Our main result in this respect is Fig. \ref{fig:stability}.
-There, we plot the *median episode reward difference* between the two DRL-based methods we considered, and the optimal constant escapement strategy.
-This figure shows that, for a wide range of parameter values, DRL-based strategies can have a considerable advantage over an optimized constant escapement policy (the single-species optimal solution).
+There, we plot the *median episode reward difference* between the two DRL-based methods we considered, and the optimal CEsc strategy.
+This figure shows that, for a wide range of parameter values, DRL-based strategies can have a considerable advantage over an optimized CEsc policy (the single-species optimal solution).
 
@@
 For each noise strength, 100 parameter perturbations were sampled---each one giving rise to a \emph{realization} of the model.
-For each such realization, we optimized a constant escapement policy and trained a PPO agent.
+For each such realization, we optimized a CEsc policy and trained a PPO agent.
 Moreover, we interpolated the PPO policy using a Gaussian process, as detailed in Sec 4.
-Then, for each realization we compared the performance of these policies: we measured the mean reward difference betwen PPO and constant escapement, and between PPO+GP and constant escapement.
+Then, for each realization we compared the performance of these policies: we measured the mean reward difference betwen PPO and CEsc, and between PPO+GP and CEsc.
 The plot represents the distribution of reward differences observed at a given noise strength---we plot the mean and the standard deviation of the mean reward differences observed.
@@
 without a citation -->
-This provides easily interpretable solutions, such as constant mortality policies.
+This provides easily interpretable solutions, such as CMort policies.
 There is a drawback here, however: due to the simplicity of these dynamical models, the policies might not respond effectively in many situations---situations where the predictions of these simple models deviate considerably from reality.
@@
 
-We compare deep reinforcement learning-based policies against classical management strategies (optimal constant mortality and optimal constant escapement policies policies). 
+We compare deep reinforcement learning-based policies against classical management strategies (optimal CMort and optimal CEsc policies policies). 
 While the latter are inspired by the shape of optimal solutions in the single-species setting, they are optimized in a model-free way as well: e.g.
-the optimal constant mortality rate is computed empirically from simulated data.
+the optimal mortality rate is computed empirically from simulated data.
 
@@
 
-We found that the optimal CMort policy surprisingly performs much better than constant escapement (Fig. \ref{fig:main}).
+We found that the optimal CMort policy surprisingly performs much better than CEsc (Fig. \ref{fig:main}).
 However, it can be observed in Fig.
-\ref{fig:fracmsy} that the constant mortality strategy faces a trade-off: high sustainability is achieved for sub-optimal constant mortality rates, but only at a significant decrease in the mean episode reward.
+\ref{fig:fracmsy} that the CMort strategy faces a trade-off: high sustainability is achieved for sub-optimal mortality rates, but only at a significant decrease in the mean episode reward.
 We expect that with increasing ecosystem complexity this phenomenon might become more pronounced.
@@
 
-The question of when multi-species models are well-approximated by single-species models was studied in detail in [@burgess].
+The question of when multi-species models are well-approximated by single-species models was studied in detail in [@burgess2017describing].
 Here our approach is dual to that of the aforementioned paper.
 Rather than first optimizing a single-species model to approximate a more complex model and then finding the MSY value for the single-species model, we used simulated data to optimize CMort and CEsc directly on the three-species model.
-We do not investigate further whether our three-species models are well approximated by a single-species model in the sense of [@burgess].
+We do not investigate further whether our three-species models are well approximated by a single-species model in the sense of [@burgess2017describing].
 However:
@@
 This in turn suggests that the reason that DRL outperforms both single-species strategies for Models 3 and 4 is not due a lack of a single-species approximation for either $X$ or $Y$, but due to the complexity of having two harvested species.
-Moreover, non-stationarity exacerbated the advantage of DRL over CEsc and CMort, which can be expected as non-stationary parameters have been observed to introduce bias into single-species reference point estimation [@burgess].
-
-Finally, we performed a stability analysis to ensure that the advantage of DRL-based techniques over constant escapement is a ubiquitous phenomenon and not a result of a lucky selection of parameter values.
+Moreover, non-stationarity exacerbated the advantage of DRL over CEsc and CMort, which can be expected as non-stationary parameters have been observed to introduce bias into single-species reference point estimation [@burgess2017describing].
+
+Finally, we performed a stability analysis to ensure that the advantage of DRL-based techniques over CEsc is a ubiquitous phenomenon and not a result of a lucky selection of parameter values.
 We found (in Fig. \ref{fig:stability}) that an advantage can be observed even for relatively high-noise perturbations of the parameters---noise with a variance of 20% of the parameter values.

### 2023-07-03 manuscript/manuscript.Rmd
added new conceptual figure

@@
 #| fig.cap:  'An experimental-design type of visualization of the management scenarios considered in this paper. 
-#| On the x-axis are four different fishery management problems: a single species model, and three three-species models.
-#| From left to right, the latter are: three species out of which one, X, is harvested, and all parameters are constant in time; 
-#| three species out of which two, X and Y, are harvested and all parameters are constant in time;
-#| and finally three species out of which two, X and Y, are harvested and one parameter depends on time. 
-#| We represent the latter’s time-dependence with a clock next to the X variable, and we intend to use it as an example of a possible simplified model for the effects of climate change.
+#| On the x-axis are four different fishery management problems (Table 1).
+#| We represent the Model 4’s non-stationarity with a clock next to the X variable, and we intend to use it as an example of a possible simplified model for the effects of climate change.
 #| On the y-axis we have different management strategies with which one may control each of the models. 
-#| On the bottom we have the constant mortality strategy (CMort), where one optimizes over constant fishing effort strategies.
-#| Above that is the constant escapement strategy (CEsc), based on calling off all fishing below a certain threshold population value.
-#| Finally, on top we have DRL-based strategies where policies are in general functions of the full state of the system, and they are parametrized by a neural network.'
+#| On the bottom we have the constant escapement strategy (CEsc), based on calling off all fishing below a certain threshold population value.
+#| Above that is the constant mortality strategy (CMort), where one optimizes over constant fishing effort strategies.
+#| Finally, on top we have DRL-based strategies where policies are in general functions of the full state of the system, and they are parametrized by a neural network.
+#| The specific DRL-based strategy is referred to as PPO+GP in the main text, due to the algorithm used to produce the policy.
+#| The results plotted are the average reward obtained by the strategy over 100 episodes, and the fraction of those episodes which do not end with a near-extinction event.
+#| We have normalized to the highest reward in each column in order to enhance the comparison between strategies.
+#| For illustrative purposes we have color-coded the results using a two-dimensional color legend displayed on the bottom left.'
 #| out.width: '6.5in' 
@@
 
-knitr::include_graphics("figures/conceptual-v2.jpeg")
+knitr::include_graphics("figures/conceptual-v3.jpeg")
 ```
@@
 In particular, we observe that DRL strategies are more sensitive to stochastic variations of the system which allows it to adaptively manage the system to prevent population collapses below the threshold.
-*Model 4:* Here, the performance gap between DRL and classical management strategies is broadened.
-In this case, as in Model 3, constant CEsc performs significantly worse than DRL and CMort.
+*Model 4:* Here, the performance gap between DRL and CEsc is broadened, and the gap between DRL and CMort is maintained.
 

### 2023-07-03 manuscript/manuscript.Rmd
mini

@@
 #| The specific DRL-based strategy is referred to as PPO+GP in the main text, due to the algorithm used to produce the policy.
-#| The results plotted are the average reward obtained by the strategy over 100 episodes, and the fraction of those episodes which do not end with a near-extinction event.
+#| The results plotted are the average reward obtained by the strategy over 100 episodes, and the fraction of those episodes which do not end with a near-extinction event (denoted 'Perc' for 'Percentage').
 #| We have normalized to the highest reward in each column in order to enhance the comparison between strategies.
@@
 In particular, we observe that DRL strategies are more sensitive to stochastic variations of the system which allows it to adaptively manage the system to prevent population collapses below the threshold.
-*Model 4:* Here, the performance gap between DRL and CEsc is broadened, and the gap between DRL and CMort is maintained.
+*Model 4:* Here, the performance gap between DRL and both classical strategies is maintained.
 

### 2023-07-03 manuscript/manuscript.Rmd
mini

@@
 #| The specific DRL-based strategy is referred to as PPO+GP in the main text, due to the algorithm used to produce the policy.
-#| The results plotted are the average reward obtained by the strategy over 100 episodes, and the fraction of those episodes which do not end with a near-extinction event (denoted 'Perc' for 'Percentage').
+#| The results plotted are the average reward obtained by the strategy over 100 episodes, and the fraction of those episodes which do not end with a near-extinction event (denoted Perc for Percentage).
 #| We have normalized to the highest reward in each column in order to enhance the comparison between strategies.

### 2023-07-04 manuscript/manuscript.Rmd
paragraph on RL envs as models

@@
 While simplified process models for the population dynamics have historically allowed for exploration of large decision spaces, the new wave of rich models are applied to highly oversimplified descriptions of potential actions they seek to inform.
-For instance, Global Circulation Models (GCMs) such as HadCM3 [@pope2000impact; @gordon2000simulation; @collins2001internal] model earth's climate using 1.5M variables, while the comparably vast potential action space is modeled as one of seven emissions marker scenarios from the IPCC [@riahi2017shared].
+For instance, Global Circulation Models (GCMs) such as HadCM3 [@pope2000impact; @gordon2000simulation; @collins2001internal] model earth's climate using 1.5M variables, while the comparably vast potential action space is modeled much more minimalistically, with 5 SSP socioeconomic storylines and 7 SSP-RCP marker scenarios summarizing the action space at the IPCC [@riahi2017shared].
 
@@
   f_Y(N_t) &= L(Y_t;\ r_Y, K_Y) - D F(Y_t, Z_t;\ \beta, c) - c_{XY}X_tY_t,\\
-  f_Z(N_t) &= \left(f(X_t + DY_t\right) - d_Z)Z_t.
+  f_Z(N_t) &= \left(b(X_t + DY_t\right) - d_Z)Z_t.
 \end{split}
@@
 A parameter $D$ regulating a relative preference of $Z$ to prey on $Y$.
-A death rate $d_Z$ and a parameter $f$ scaling the birth rate of $Z$.
+A death rate $d_Z$ and a parameter $b$ scaling the birth rate of $Z$.
 Finally, the noise variances $\sigma_X$, $\sigma_Y$ and $\sigma_Z$.
@@
 That is, the agent performs actions within the environment.
-As a result of an agent's action, the environment will transition to a new state and return a numerical *reward* to the agent.
+
+After the agent takes an action, the environment will transition to a new state and return a numerical *reward* to the agent.
 (See Fig.
@@
 This objective is achieved by aggregating experience in what is called the *training* period and learning from such experience.
-<!-- ML: We could cite our deep RL paper here instead of going into further details. Not sure what we are expecting as the audience. -->.
+<!-- ML: We could cite our deep RL paper here instead of going into further details. Not sure what we are expecting as the audience. -->
 
@@
 
-For this paper, we will consider four environments corresponding to each of the models considered (Table \ref{tab:models}).
+The *environment* is commonly a computer simulation.
+It is important to note here the role that real time-series data of stock sizes can play in this process.
+This data is not used directly to train the RL agent, but rather to estimate the model defining the environment.
+This environment is subsequently used to train the agent.
+In this paper, we focus on the second step---we take the estimated model of reality as a given, and train an RL agent on it.\footnote{
+In this sense, it is important to note that the classical management strategies we compare against have a similar flow of information.
+Namely, data is used to estimate a dynamical model, and this model is used to generate a policy function.
+The difference to our approach is located in the process of *how* the model is used to optimize a policy.
+Because of this difference, RL-based approaches can produce good heuristic solutions for complex problems.
+}
+
+Specifically, we consider four environments corresponding to each of the models considered (Table \ref{tab:models}).
 At each time step, the agent observes the state $S$ and enacts some harvest---reducing $X_t$ to $X_t - M_X(N_t)\cdot X_t$, and, for Models 3 and 4, also reducing $Y_t$ to $Y_t - M_Y(N_t)\cdot Y_t$.
@@
 This secures a reward of $M_X(N_t)X$ for Models 1 and 2, and, similarly, a reward of $M_X(N_t)X + M_Y(N_t)Y$ for Models 3 and 4.
-After this harvest portion of the time step, the environment changes according to its natural dynamics (see eq. \eqref{eq:3d model}).
+After this harvest portion of the time step, the environment evolves naturally according to eqs. \eqref{eq:may} and \eqref{eq:3d model} (Sec. 2).
 
@@
 This approximation is commonly used in fisheries for species with annual reproductive cycles (see e.g. [@mangel2006theoretical], Chap. 6).
-The division of the time-step into two parts (harvest and dynamics) happening sequentially is performed for all control methods evaluated, not only RL-based approaches.
+Moreover, the separation of each time-step into a harvest period and a natural growth period assumes that harvest has little disruption on the reproductive process.
+A detailed model which includes such a disruption is outside of the scope of this work.
 
@@
 We will focus on deep reinforcement learning throughout this paper.
-Within the DRL literature there is a wealth of algorithms from which to choose from to optimize $\pi$, each with its pros and cons.
+Within the DRL literature there is a wealth of algorithms from which to choose to optimize $\pi$, each with its pros and cons.
 Most of these are based on gradient ascent by using the technique of *back-propagation* to efficiently compute the gradient.
@@
   r_X = K_X = r_Y = K_Y = 1, \quad \beta = 0.3, \quad c = 0.3\\
-  c_{XY} = 0.1,\quad f = 0.1,\quad D = 1.1, \quad d_Z = 0.1.
+  c_{XY} = 0.1,\quad b = 0.1,\quad D = 1.1, \quad d_Z = 0.1.
 \end{split}
@@
 Our main result in this respect is Fig. \ref{fig:stability}.
-There, we plot the *median episode reward difference* between the two DRL-based methods we considered, and the optimal CEsc strategy.
+There, we plot the *average episode reward difference* between the two DRL-based methods we considered, and the optimal CEsc strategy.
 This figure shows that, for a wide range of parameter values, DRL-based strategies can have a considerable advantage over an optimized CEsc policy (the single-species optimal solution).
@@
 This in turn suggests that the reason that DRL outperforms both single-species strategies for Models 3 and 4 is not due a lack of a single-species approximation for either $X$ or $Y$, but due to the complexity of having two harvested species.
-Moreover, non-stationarity exacerbated the advantage of DRL over CEsc and CMort, which can be expected as non-stationary parameters have been observed to introduce bias into single-species reference point estimation [@burgess2017describing].
+Moreover, the non-stationarity in Model 4 maintained the advantage of DRL over CEsc and CMort. 
+Here, one may have expected an exacerbation of that advantage due to non-stationarities introducing biases to single-species approximate models [@burgess2017describing].
+We did, however, measure a decrease in the sustainability of CEsc in Model 4 with respect to Model 3.
 

### 2023-07-04 manuscript/manuscript.Rmd
intro now includes paragraph on ABC and TAC

@@
 Fisheries are an essential natural resource that provide the primary source of protein for one in every four humans, and have faced widely documented declines due to over-fishing [@Worm2006, @Costello2016].
+Fisheries management centers around the process of sampling populations to determine fishing quotas based on population estimates.
+This decision is often guided by a model of the dynamics of the system.
+Our paper focuses on the decision side of this problem rather than the measurement step.
+
 Fisheries management has roots in both the fields of _ecosystem management_ and _natural resource economics_.
@@
 <!--NEED some kind of map paragraph here saying what's ahead, what we show  -->
-In this article, we compare the performance of a DRL solution to that of classical management strategies (see Textbox 1).\footnote{A repository with all the relevant code to reproduce our results may be found at \url{https://github.com/boettiger-lab/approx-model-or-approx-soln} in the ‘‘src’’ subdirectory.
-The data used is found in the ‘‘data’’ subdirectory, but the user may use the code provided to generate new data sets.} 
+
+In most US fisheries, the mortality policy is often piecewise linear (often with one constant and one linear piece), and the allowable biological catch (ABC) or total allowable catch (TAC) is set at some heuristic (e.g.
+80%) below the ‘overfishing limit’, $F_{MSY}$, i.e.
+the highest (constant) mortality that can be sustained indefinitely (in the model – reality of course does not permit such definitions).
+This fixed mortality management can be seen, for instance, in most of the fisheries listed in the widely used R.A. Myers Legacy Stock Assessment Database.
+Here, we have focused on purely constant mortality policies, rather than piecewise linear mortality funcitons, for simplicity.
+Escapement-based management is less common, except in salmonoids, as it requires closing a fishery whenever the measured biomass falls below $B_{MSY}$.
+
+In this article, we compare against two common methods: constant mortality (CMort) and constant escapement (CEsc), introduced in Text Box 1.\footnote{
+A repository with all the relevant code to reproduce our results may be found at \url{https://github.com/boettiger-lab/approx-model-or-approx-soln} in the ‘‘src’’ subdirectory. 
+The data used is found in the ‘‘data’’ subdirectory, but the user may use the code provided to generate new data sets.
+}
+
 We consider the problem of devising harvest strategies in a for a series of ecosystem models with increasing complexity (Table \ref{tab:models}):

### 2023-07-05 manuscript/manuscript.Rmd
minitypos

@@
 In this paper, we focus on a well-developed application of model-based management of the natural world that has long illustrated the trade-offs between model complexity and policy complexity: the management of marine fisheries.
-Fisheries management is both an important issue to society and a rich & frequent test-bed of ecological management more generally.
+Fisheries management is both an important issue to society as well as a rich and frequent test-bed of ecological management more generally.
 Fisheries are an essential natural resource that provide the primary source of protein for one in every four humans, and have faced widely documented declines due to over-fishing [@Worm2006, @Costello2016].
@@
 
-## Discussion
-
 Fisheries are complex ecosystems, with interactions between species leading to highly non-linear dynamics.
@@
 
-We compare deep reinforcement learning-based policies against classical management strategies (optimal CMort and optimal CEsc policies policies). 
+We compare deep reinforcement learning-based policies against classical management strategies (CMort and CEsc). 
 While the latter are inspired by the shape of optimal solutions in the single-species setting, they are optimized in a model-free way as well: e.g.

### 2023-07-05 manuscript/manuscript.tex
added tex

--- 2022-12-13 manuscript/manuscript.tex
@@
-% !TeX program = pdfLaTeX
-\documentclass[smallextended]{svjour3}       % onecolumn (second format)
-%\documentclass[twocolumn]{svjour3}          % twocolumn
-%
-\smartqed  % flush right qed marks, e.g. at end of proof
-%
-\usepackage{amsmath,amssymb}
+\documentclass{article}
+
+\usepackage{arxiv}
+
+\usepackage[utf8]{inputenc} % allow utf-8 input
+\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
+\usepackage{lmodern}        % https://github.com/rstudio/rticles/issues/343
+\usepackage{hyperref}       % hyperlinks
+\usepackage{url}            % simple URL typesetting
+\usepackage{booktabs}       % professional-quality tables
+\usepackage{amsfonts}       % blackboard math symbols
+\usepackage{nicefrac}       % compact symbols for 1/2, etc.
+\usepackage{microtype}      % microtypography
 \usepackage{graphicx}
-\usepackage[utf8]{inputenc}
 
-\usepackage[hyphens]{url} % not crucial - just used below for the URL
-\usepackage{hyperref}
+\title{Pretty darn good control: when are approximate solutions better
+than approximate models}
 
-% common operators
-\newcommand*{\der}[2]{\frac{\mathrm{d}#1}{\mathrm{d}#2}}
-
-% structural markers
-\def\div{\;|\;}
-\def\bigdiv{\;\big|\;}
-\def\Bigdiv{\;\Big|\;}
-\def\Biggdiv{\;\Bigg|\;}
-
-%
-% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
-%
-% insert here the call for the packages your document requires
-%\usepackage{latexsym}
-% etc.
-%
-% please place your own definitions here and don't use \def but
-% \newcommand{}{}
-%
-% Insert the name of "your journal" with
-% \journalname{myjournal}
-%
-
-%% load any required packages here
+\author{
+    Felipe Montealegre-Mora
+   \\
+     \\
+   \\
+  \texttt{} \\
+   \And
+    Marcus Lapeyrolerie
+   \\
+     \\
+   \\
+  \texttt{} \\
+   \And
+    Melissa Chapman
+   \\
+     \\
+   \\
+  \texttt{} \\
+   \And
+    Abigail G. Keller
+   \\
+     \\
+   \\
+  \texttt{} \\
+   \And
+    Carl Boettiger
+   \\
+     \\
+   \\
+  \texttt{\href{mailto:cboettig@berkeley.edu}{\nolinkurl{cboettig@berkeley.edu}}} \\
+  }
 
@@
 
+% Pandoc citation processing
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+\newlength{\csllabelwidth}
+\setlength{\csllabelwidth}{3em}
+\newlength{\cslentryspacingunit} % times entry-spacing
+\setlength{\cslentryspacingunit}{\parskip}
+% for Pandoc 2.8 to 2.10.1
+\newenvironment{cslreferences}%
+  {}%
+  {\par}
+% For Pandoc 2.11+
+\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
+ {% don't indent paragraphs
+  \setlength{\parindent}{0pt}
+  % turn on hanging indent if param 1 is 1
+  \ifodd #1
+  \let\oldpar\par
+  \def\par{\hangindent=\cslhangindent\oldpar}
+  \fi
+  % set entry spacing
+  \setlength{\parskip}{#2\cslentryspacingunit}
+ }%
+ {}
+\usepackage{calc}
+\newcommand{\CSLBlock}[1]{#1\hfill\break}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
+\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
+
+\usepackage{amsmath}
+\usepackage{amssymb}
+\usepackage{lineno}
+\usepackage{setspace}
+\usepackage{bookmark}
+\usepackage{float}
+\usepackage{graphicx}
+\usepackage{caption}
+\usepackage{subcaption}
+\usepackage{tikz-cd}
+\usepackage{tcolorbox}
+\usepackage{etoolbox}
+\usepackage{url}
+\linenumbers
+\doublespacing
 \begin{document}
-
-\title{Title here \thanks{Grants or other notes about the article that
-should go on the front page should be placed here. General
-acknowledgments should be placed at the end of the article.} }
- \subtitle{Do you have a subtitle? If so, write it here}
-
-    \titlerunning{Short form of title (if too long for head)}
-
-\author{  Äüthör 1 \and  Âuthóř 2 \and  }
-
-    \authorrunning{ Short form of author list if too long for running
-head }
-
-\institute{
-        Äüthör 1 \at
-     Department of YYY, University of XXX \\
-     \email{\href{mailto:abc@def}{\nolinkurl{abc@def}}}  %  \\
-%             \emph{Present address:} of F. Author  %  if needed
-    \and
-        Âuthóř 2 \at
-     Department of ZZZ, University of WWW \\
-     \email{\href{mailto:djf@wef}{\nolinkurl{djf@wef}}}  %  \\
-%             \emph{Present address:} of F. Author  %  if needed
-    \and
-    }
-
-\date{Received: date / Accepted: date}
-% The correct dates will be entered by the editor
-
 \maketitle
@@
 \begin{abstract}
-The text of your abstract. 150 -- 250 words.
-\\
-\keywords{
-        key \and
-        dictionary \and
-        word \and
-    }
-
-    \subclass{
-                    MSC code 1 \and
-                    MSC code 2 \and
-            }
-
+Existing methods for optimal control struggle to deal with the
+complexity commonly encountered in real-world systems, including
+dimensionality, process error, model bias and data heterogeneity.
+Instead of tackling these system complexities directly, researchers have
+typically sought to simplify models to fit optimal control methods. But
+when is the optimal solution to an approximate, stylized model better
+than an approximate solution to a more accurate model? While this
+question has largely gone unanswered owing to the difficulty of finding
+even approximate solutions for complex models, recent algorithmic and
+computational advances in deep reinforcement learning (DRL) might
+finally allow us to address these questions. DRL methods have to date
+been applied primarily in the context of games or robotic mechanics,
+which operate under precisely known rules. Here, we demonstrate the
+ability for DRL algorithms using deep neural networks to successfully
+approximate solutions (the ``policy function'' or control rule) in a
+non-linear three-variable model for a fishery without knowing or ever
+attempting to infer a model for the process itself. We find that the
+reinforcement learning agent discovers a policy that outperforms both
+constant escapement and constant mortality policies---the standard
+family of policies considered in fishery management. This DRL policy has
+the shape of a constant escapement policy whose escapement values depend
+on the stock sizes of other species in the model.
 \end{abstract}
 
-\def\spacingset#1{\renewcommand{\baselinestretch}%
-{#1}\small\normalsize} \spacingset{1}
+\keywords{
+    Optimal Control
+   \and
+    Reinforcement Learning
+   \and
+    Uncertainty
+   \and
+    Decision Theory
+  }
 
@@
 
-Your text comes here. Separate text sections with \cite{Mislevy06Cog}.
+Much effort has been spent grappling with the complexity of our natural
+world in contrast to the relative simplicity of the models we use to
+understand it. Heroic amounts of data and computation are being brought
+to bear on developing better, more realistic models of our environments
+and ecosystems, in hopes of improving our capacity to address the many
+planetary crises. But despite these efforts and advances, we remain
+faced with the difficult task of figuring out how best to respond to
+these crises. While simplified process models for the population
+dynamics have historically allowed for exploration of large decision
+spaces, the new wave of rich models are applied to highly oversimplified
+descriptions of potential actions they seek to inform. For instance,
+Global Circulation Models (GCMs) such as HadCM3 (Pope et al. 2000; C.
+Gordon et al. 2000; Collins, Tett, and Cooper 2001) model earth's
+climate using 1.5M variables, while the comparably vast potential action
+space is modeled much more minimalistically, with 5 SSP socioeconomic
+storylines and 7 SSP-RCP marker scenarios summarizing the action space
+at the IPCC (Riahi et al. 2017).
 
-\hypertarget{sec:1}{%
-\section{Background}\label{sec:background}}
+Even as our research community develops simulations of the natural world
+that fit only in supercomputers, we analyze a space of policies that
+would fit on index cards. Similar combinations of rich process models
+and highly simplified decision models (often not even given the status
+of `model') are common. Modeling the potential action space as one of a
+handful of discrete scenarios is sometimes a well justified
+acknowledgement of the constraints faced by real-world decision-makers
+-- particularly in the context of multilateral decisions -- and may seem
+to reflect a division of responsibilities between `scientists' modeling
+the `natural processes' and policy-makers who make the decisions. But,
+more often, this simplification of decision choices is simply
+mathematically or conceptually convenient. This simplification reflects
+trade-offs between tractablity and complexity at the basis of any
+mathematical modeling -- if we make both the state space and action
+space too realistic, the problem of finding the best sequence of actions
+quickly becomes intractable. However, emerging data-driven methods from
+machine learning offer a new choice -- algorithms that can find good
+strategies in previously intractable problems, but at the cost of
+opacity.
 
-% Text with citations by \cite{Galyardt14mmm}.
-In this section we first review two well-established techniques commonly used in sustainable fishery management. 
-These are the maximum sustainable yield (MSY) and the constant escapement (CE) approaches. 
-After this, deep reinforcement learning is briefly reviewed
+In this paper, we focus on a well-developed application of model-based
+management of the natural world that has long illustrated the trade-offs
+between model complexity and policy complexity: the management of marine
+fisheries. Fisheries management is both an important issue to society as
+well as a rich and frequent test-bed of ecological management more
+generally. Fisheries are an essential natural resource that provide the
+primary source of protein for one in every four humans, and have faced
+widely documented declines due to over-fishing Costello et al. (2016).
+Fisheries management centers around the process of sampling populations
+to determine fishing quotas based on population estimates. This decision
+is often guided by a model of the dynamics of the system. Our paper
+focuses on the decision side of this problem rather than the measurement
+step.
 
-\hypertarget{sec:2}{%
-\subsection{Fishery management}\label{sec:fishery}}
+Fisheries management has roots in both the fields of \emph{ecosystem
+management} and \emph{natural resource economics}. Both fields might
+trace their origins to the notion of maximum sustainable yield (MSY),
+introduced independently by a fisheries ecologist (Schaefer 1954) and
+the economist (H. S. Gordon and Press 1954) in the same year. From this
+shared origin, each field would depart from the simplifying assumptions
+of the Gordon-Schaefer model in divergent ways, leading to different
+techniques for deriving policies from models.The heart of the management
+problem is easily understood: a manager seeks to set quotas on fishing
+that will ensure the long-term profitability and sustainability of the
+industry. Mathematical approaches developed over the past century may be
+roughly divided between these two fields: (A) ecologists, focused on
+ever more realistic models of the biological processes of growth and
+recruitment of fish while considering a relatively stylized suite of
+potential management strategies, and (B) economists, focused on far more
+stylized models of the ecology while exploring a far less constrained
+set of possible policies. The economist's approach can be characterized
+by the mathematics of a Markov decision process (MDP Colin W. Clark
+1973; Colin W. Clark 1990; Marescot et al. 2013), in which the
+decision-maker must observe the stock each year and recommend a possible
+action. In this approach, the policy space that must be searched is
+exponentially large -- for a management horizon of T decisions and a
+space of N actions, the number of possible policies is \(N^T\). In
+contrast, fisheries ecologists and ecosystem management typically search
+a space of policies that does not scale with the time horizon. Under
+methods such as ``Management Strategy Evaluation'' (MSE, (Punt et al.
+2016)) a manager identifies a candidate set of ``strategies'' a priori,
+and then compares the performance of each strategy over a suite of
+simulations to determine which strategy gives the best outcome
+(i.e.~best expected utility). This approach is far more amenable to
+complex simulations of fisheries dynamics and more closely corresponds
+to how most marine fishing quotas are managed today in the United States
+(see stock assessments documented in RAM Legacy Stock Assessment
+Database 2020).
 
-\hypertarget{sec:3}{%
-\subsection{Deep reinforcement learning}\label{sec:rl}}
+\includegraphics[width=1\linewidth]{figures/textbox}
 
-\hypertarget{sec:4}{%
-\section{Dynamical models used}\label{sec:model}}
+Recent advances in machine learning may allow us to once again bridge
+these approaches, while also bringing new challenges of their own. Novel
+data-driven methods have allowed these models to evolve into ever more
+complex and realistic simulations used in fisheries management, where
+models with over 100 parameters are not uncommon (RAM Legacy Stock
+Assessment Database 2020). Constrained by computational limits, MDP
+approaches have been intractable on suitably realistic models and
+largely confined to more academic applications (Costello et al. 2016).
+However, advances in \emph{Deep Reinforcement Learning}, (DRL) a
+sub-field of machine learning, have recently demonstrated remarkable
+performance in a range of such MDP problems, from video games (Bellemare
+et al. 2013; Mnih et al. 2013) to fusion reactions (Degrave et al. 2022;
+Seo et al. 2022) to the remarkable dialog abilities of ChatGPT (OpenAI
+2022). RL methods also bring many challenges of their own: being
+notoriously difficult to train and evaluate, requiring immense
+computational costs, and presenting frequent challenges with
+reproducibility. A review of all these issues is beyond our scope but
+can be found elsewhere (Lapeyrolerie et al. 2022; Chapman et al. 2023).
+Here, though, we will focus on the issue of opacity and interpretability
+raised by these methods. In contrast with optimization algorithms
+currently used in either ecosystem management or resource economics, RL
+algorithms have no guarantees of or metrics for convergence to an
+optimal solution. In general, one can only assess the performance of
+these black box methods relative to alternatives.
 
-In this section we present three models of increasing complexity which plausibly describe the population dynamics of a marine ecosystem.
-These models will form the test beds for the comparison between classical fishery management strategies and DRL.
+In most US fisheries, the mortality policy is often piecewise linear
+(often with one constant and one linear piece), and the allowable
+biological catch (ABC) or total allowable catch (TAC) is set at some
+heuristic (e.g.~80\%) below the `overfishing limit', \(F_{MSY}\),
+i.e.~the highest (constant) mortality that can be sustained indefinitely
+(in the model -- reality of course does not permit such definitions).
+This fixed mortality management can be seen, for instance, in most of
+the fisheries listed in the widely used R.A. Myers Legacy Stock
+Assessment Database. Here, we have focused on purely constant mortality
+policies, rather than piecewise linear mortality funcitons, for
+simplicity. Escapement-based management is less common, except in
+salmonoids, as it requires closing a fishery whenever the measured
+biomass falls below \(B_{MSY}\).
 
-\hypertarget{sec:4.1}{%
-\subsection{A one-dimensional tipping point model.}\label{p:may}}
+In this article, we compare against two common methods: constant
+mortality (CMort) and constant escapement (CEsc), introduced in Text Box
+1.\footnote{
+A repository with all the relevant code to reproduce our results may be found at \url{https://github.com/boettiger-lab/approx-model-or-approx-soln} in the ‘‘src’’ subdirectory. 
+The data used is found in the ‘‘data’’ subdirectory, but the user may use the code provided to generate new data sets.
+}
 
-Consider a population $V$ whose dynamics is given by
-\begin{align}
-  \label{eq:may77}
-  \der{V}{t} = rV\left(
-    1-V/K
-    \right) 
-    -
-    \frac{\beta H V^2}{V_0^2 + V^2}. 
+We consider the problem of devising harvest strategies in a for a series
+of ecosystem models with increasing complexity (Table \ref{tab:models}):
+\emph{1) one species, one fishery:} a simple single-species recruitment
+model based on (May 1977); \emph{2) three species, one fishery:} a
+three-species generalization of model \emph{1)}, where one of the
+species is harvested; \emph{3) three species, two fisheries:} the same
+three-species model as above but with two harvested species; \emph{4)
+three species, two fisheries, parameter variation:} a three-species
+model of which two are harvested, as above, with a time-varying
+parameter. This last model is meant to be a toy model of climate
+change's effect on the system. Across all of these scenarios, two goals
+are balanced in the decision process: maximizing long-term catch and
+preventing stock sizes to fall below some a-priori threshold.
+
+\begin{table}
+  \begin{center}
+    \begin{tabular}{|l|c|c|c|c|}
+      \hline
+      \textbf{Model name} & \textbf{Model eqs.} & \textbf{N. Sp.} & \textbf{Harv. Sp.} & \textbf{Stationary?}\\
+      \hline
+      Model 1 & \eqref{eq:may} & 1 & $X$ & Yes \\
+      Model 2 & \eqref{eq:3d model} & 3 & $X$ & Yes \\
+      Model 3 & \eqref{eq:3d model} & 3 & $X$ and $Y$ & Yes \\
+      Model 4 & \eqref{eq:3d model} & 3 & $X$ and $Y$ & No \\
+      \hline
+    \end{tabular}
+    \caption{Table of models considered in this paper.
+Here, \textbf{N.
+Sp.} is the \emph{number of species} in the model, \textbf{Harv.
+Sp.} is the species of the model which are harvested, and \textbf{Stationary?} refers to whether the parameters of the model have fixed values (or, on the contrary, if they vary in time).
+The only non-stationary case presented in the paper is where $r_X$ drifts linearly with time.
+In the code repository associated to the paper, we consider other possible choices of non-stationarity.}
+    \label{tab:models}
+  \end{center}
+\end{table}
+
+This way, we evaluate classical management strategies (CMort and CEsc)
+and DRL-based strategies on four different models. This experimental
+design is summrized in Fig. \ref{fig:conceptual-v2}.
+
+\begin{figure}[H]
+
+{\centering \includegraphics[width=6.5in]{figures/conceptual-v3}
+
+}
+
+\caption{An experimental-design type of visualization of the management scenarios considered in this paper.
+On the x-axis are four different fishery management problems (Table 1).
+We represent the Model 4’s non-stationarity with a clock next to the X variable, and we intend to use it as an example of a possible simplified model for the effects of climate change.
+On the y-axis we have different management strategies with which one may control each of the models.
+On the bottom we have the constant escapement strategy (CEsc), based on calling off all fishing below a certain threshold population value.
+Above that is the constant mortality strategy (CMort), where one optimizes over constant fishing effort strategies.
+Finally, on top we have DRL-based strategies where policies are in general functions of the full state of the system, and they are parametrized by a neural network.
+The specific DRL-based strategy is referred to as PPO+GP in the main text, due to the algorithm used to produce the policy.
+The results plotted are the average reward obtained by the strategy over 100 episodes, and the fraction of those episodes which do not end with a near-extinction event (denoted Perc for Percentage).
+We have normalized to the highest reward in each column in order to enhance the comparison between strategies.
+For illustrative purposes we have color-coded the results using a two-dimensional color legend displayed on the bottom left.}\label{fig:conceptual-v2}
+\end{figure}
+
+Regarding control for these 4 models, we show the following. \emph{Model
+1:} DRL-based strategies recover the optimal constant escapement (CEsc)
+policy function. Constant mortality (CMort) performs considerably worse
+than these three.\footnote{
+As will be explained later, all our models are stochastic. 
+If we set stochasticity to zero in Model 1, CMort matches the performance of the other management strategies.
+} \emph{Model 2:} Here, all management strategies perform similarly.
+\emph{Model 3:} For this model, DRL outperforms both classical
+strategies, with CMort surprisingly performing significantly better than
+CEsc. In particular, we observe that DRL strategies are more sensitive
+to stochastic variations of the system which allows it to adaptively
+manage the system to prevent population collapses below the threshold.
+\emph{Model 4:} Here, the performance gap between DRL and both classical
+strategies is maintained.
+
+We show that in the most complex scenario, Model 4, CMort is faced with
+a tradeoff---the optimal mortality rate leads a rather large fraction of
+episodes ending with a population crash, whose negative reward is
+counteracted with a higher economic output.\footnote{
+In our mathematical formulation of the decision problem, we have assumed for simplicity that the fishing effort cost is zero and that fish price is stable over time.
+This way, we equate economic output with harvested biomass.
+} Conversely, more conservative, lower, mortality rates lead to lower
+total reward on average. The DRL approach side-steps this trade-off by
+optimizing over a more complex family of possible policies-----policies
+parametrized by a neural network, as opposed to policies labeled by a
+single parameter (the mortality rate).
+
+Our findings paint a picture of how a single-species optimal management
+strategy may lose performance rather dramatically when controlling a
+more complex ecosystem. Here, DRL performs better from both economical
+\emph{and} conservation points of view. Moreover, rather unintuitively,
+CMort -- known to be suboptimal and unsustainable for single-species
+models -- can turn out to even outperform the CEsc -- the single-species
+optimal strategy -- for complex ecosystems. Finally, within this regime
+of complex, possibly varying, ecosystems, we show that DRL consistently
+finds a policy which effectively either matches the best classical
+strategy, or outperforms it. We strengthen this result with a stability
+investigation: we show that random perturbations in the model's
+parameter values used do not significantly vary the conclusion that DRL
+outperforms CEsc.
+
+\hypertarget{mathematical-models-of-fisheries-considered}{%
+\section{Mathematical models of fisheries
+considered}\label{mathematical-models-of-fisheries-considered}}
+
+Here we mathematically introduce the four fishery models for which we
+compare different management strategies. In general, the class of models
+that appear in this context are \emph{stochastic, first order, finite
+difference equations}. For \(n\) species, these models have the general
+form \begin{align}
+  \label{eq:general model}
+  \begin{split}
+    \Delta X_t &= f_X(N_t) + \eta_{X,t} - M_X X_t\\
+    \Delta Y_t &= f_Y(N_t) + \eta_{Y,t} - M_Y Y_t\\
+    \Delta Z_t &= f_Z(N_t) + \eta_{Z,t} - M_Z Z_t\\
+    &\dots
+  \end{split}
+\end{align} where
+\(N_t = (X_t,\ Y_t,\ Z_t,\ \dots) \in \mathbb{R}^{n}_+\) is a vector of
+populations, \(\Delta X_t := X_{t+1} - X_t\),
+\(f_i:\mathbb{R}^{n}\to\mathbb{R}\) are arbitrary functions, and where
+\(\eta_{i,t}\) are Gaussian random variables. Here,
+\(M_i=M_i(N_t)\in\mathbb{R}_+\) is a state-dependent fish mortality
+arising from harvesting the \(i\)-th species (sometimes this is referred
+to as \emph{fishing effort}).
+
+The term \(M_X X_t\) is the total \(X\) harvest at time \(t\). This
+formulation of stock recruitment as a discrete finite difference process
+is common among fisheries, as opposed to continuous time formulations
+which involve instantaneous growth rates. This growth rate simplifies
+e.g.~the possibly seasonal nature of reproduction (which would need to
+be accounted for in a realistic continuous-time model) by simply
+considering the total recruitment experienced by the population over a
+full year.
+
+The fishing efforts are the \emph{control variables} of our
+problem---these may be set by the manager at each time-step to specified
+values. We make two further simplifying assumptions on the control
+problem: \emph{1. Full observation:} the manager is able to accurately
+measure \(N_t\) and use that measurement to inform their decision.
+\emph{2. Perfect execution:} the action chosen by the manager is
+implemented perfectly (i.e., there is no noise affecting the actual
+value of the fishing efforts).
+
+Model 1 is a single-species classical model of ecological tipping
+points. Models 2-4 are all three-species models with similar dynamics.
+Following this logic, the first subsection will be dedicated to the
+single-species model and the second will focus on the three-species
+models.
+
+\hypertarget{the-single-species-model}{%
+\subsection{The single species model}\label{the-single-species-model}}
+
+Optimal control policies for fisheries are frequently based on
+1-dimensional models, \(n=1\), as described in \(Text\) \(Box\) \(1\).
+The most familiar model of \(f(X)\) is that of \emph{logistic growth},
+for which \begin{align}
+  \label{eq:logistic}
+  f(X_t) = r X_t\big(1 - X_t / K \big) =: L(X_t;\ r, K).
 \end{align}
-This model has been used in~\cite{may77} to describe a grazing ecosystem, where a species $V$ of vegetation is harvested by a constant herbivore population $H$. 
-% Here the only dynamic degree of freedom is $V$, while $r,\ K,\ \beta,\ V_0,$ and $H > 0$ are fixed positive constants.
 
-In~\eqref{eq:may77}, a population $V$ grows logistically with rate $r$ up to carrying capacity $K$.
-This is expressed by the first term in the equation,
+Real world ecological systems are obviously far more complicated than
+this simple model suggests. One particularly important aspect that has
+garnered much attention is the potential for the kind of highly
+non-linear functions that can support dynamics such as alternative
+stable states and hysteresis. A seminal example of such dynamics was
+introduced in (May 1977), using a one-dimensional model of a prey
+(resource) species under the pressure of a (fixed) predator. In the
+notation of eq. \eqref{eq:general model}, \begin{align}
+  \label{eq:may}
+  f_X(X_t)
+  = L(X_t;\ r, K) - \frac{\beta H X_t^2}{c^2 + X_t^2}.
+\end{align} In the following, we will denote \begin{align*}
+  F(X_t,H;\ \beta,c) := \frac{\beta H X_t^2}{c^2 + X_t^2}.
+\end{align*}
+
+The model has six parameters: the growth rate \(r\) and carrying
+capacity \(K\) for \(X\), a constant population \(H\) of a species which
+preys on \(X\), the maximal predation rate \(\beta\), the predation rate
+half-maximum biomass \(c\), and the variance \(\sigma_X^2\) of the
+stochastic term \(\eta_{X,t}\). (Here and in the following we will
+center all random variables at zero.)
+
+Eq. \eqref{eq:may} is an interesting study case of a \emph{tipping
+point} (saddle-node bifurcation) (see Fig. \ref{fig:may}). Holding the
+value of \(\beta\) fixed, for intermediate values of \(H\) there exist
+two stable fixed points for the state \(X_t\) of the system, these two
+attractors separated by an unstable fixed point. At a certain threshold
+value of \(H\), however, the top stable fixed point collides with the
+unstable fixed point and both are annihilated. For this value of \(H\),
+and for higher values, only the lower fixed point remains. This also
+creates the phenomenon of \emph{hysteresis}, where returning \(H\) to
+its original value is not sufficient to restore \(X_t\) to the original
+stable state.
+
+\begin{figure}
+\includegraphics[width=6.5in]{figures/MayStablePoints} \caption{The fixed point diagram for the unharvested dynamics of Model 1 as a function of varying the parameter $\beta H$, assuming zero noise.
+Stable fixed points (also known as attractors) are plotted using a solid line, while the unstable fixed point is shown as a dotted line.}\label{fig:may}
+\end{figure}
+
+This structure implies two things. First, that a drift in \(H\) could
+lead to catastrophic consequences, with the population \(X_t\)
+plummeting to the lower fixed stable point. Second, that if the
+evolution of \(X_t\) is \emph{stochastic}, then, even at values of \(H\)
+below the threshold point, the system runs a sizeable danger of tipping
+over towards the lower stable point.
+
+\hypertarget{the-three-species-models}{%
+\subsection{The three species models}\label{the-three-species-models}}
+
+Models 2-4 are three-species models and they are all closely
+related---in fact, their natural dynamics (i.e.~dynamics under zero
+harvest) is essentially given by the same equations: \begin{align}
+\label{eq:3d model}
+\begin{split}
+  f_X(N_t) &= L(X_t;\ r_X, K_X) - F(X_t, Z_t;\ \beta, c) - c_{XY}X_tY_t, \\
+  f_Y(N_t) &= L(Y_t;\ r_Y, K_Y) - D F(Y_t, Z_t;\ \beta, c) - c_{XY}X_tY_t,\\
+  f_Z(N_t) &= \left(b(X_t + DY_t\right) - d_Z)Z_t.
+\end{split}
+\end{align} The three species modeled are \(X\), \(Y\) and \(Z\).
+Species \(Z\) preys on both \(X\) and \(Y\), while the latter two
+compete for resources. There are thirteen parameters in this model: The
+growth rate and carrying capacity, \(r_X\), \(K_X\), \(r_Y\) and
+\(K_Y\), of \(X\) and \(Y\). A parameter \(c_{XY}\) mediating a
+Lotka-Volterra competition between \(X\) and \(Y\). A maximum predation
+rate \(\beta\) and a predation rate half-maximum biomass \(c\)
+specifying how \(Z\) forages on \(X\) and \(Y\). A parameter \(D\)
+regulating a relative preference of \(Z\) to prey on \(Y\). A death rate
+\(d_Z\) and a parameter \(b\) scaling the birth rate of \(Z\). Finally,
+the noise variances \(\sigma_X\), \(\sigma_Y\) and \(\sigma_Z\).
+
+The three models will branch off of eq. \eqref{eq:3d model} in the
+following way. \emph{Model 2:} here, only \(X\) is harvested, that is,
+in the notation of eq. \eqref{eq:general model}, we fix \(M_Y=M_Z=0\)
+and leave \(M_X\) as a control variable. All parameters here are
+constant. \emph{Model 3:} as Model 1, but with \(X\) and \(Y\) being
+harvested. In other words, we set \(M_Z=0\) and leave \(M_X\) and
+\(M_Y\) as control variables. \emph{Model 4:} here \(X\) and \(Y\) are
+harvested, but now we include a non-stationary parameter:
+\begin{equation}
+  \label{eq:rx(t)}
+  r_X = r_X(t) = 
+  \begin{cases}
+    1-t/200, \quad & t \leq 100,\\
+    1/2, \quad & t > 100.
+  \end{cases}
+\end{equation} All other parameters are constant. Eq. \eqref{eq:rx(t)}
+is intended to reflect in a simple manner a possible effect of climate
+change: where the reproductive rate of \(X\) is reduced linearly over
+time until it stabilizes.
+
+\hypertarget{sec:RL}{%
+\section{Reinforcement learning}\label{sec:RL}}
+
+Reinforcement learning (RL) is a way of approaching \emph{sequential
+decision problems} through machine learning. All applications of RL can
+be conceptually separated into two parts: an \emph{agent} and an
+\emph{environment} which the agent interacts with. That is, the agent
+performs actions within the environment.
+
+After the agent takes an action, the environment will transition to a
+new state and return a numerical \emph{reward} to the agent. (See Fig. 1
+in (Lapeyrolerie et al. 2022) for a conceptual description of
+reinforcement learning algorithms.) The rewards encode the agent's goal.
+The main task of any RL algorithm is then to maximize the cumulative
+reward received. This objective is achieved by aggregating experience in
+what is called the \emph{training} period and learning from such
+experience.
+
+The \emph{environment} is commonly a computer simulation. It is
+important to note here the role that real time-series data of stock
+sizes can play in this process. This data is not used directly to train
+the RL agent, but rather to estimate the model defining the environment.
+This environment is subsequently used to train the agent. In this paper,
+we focus on the second step---we take the estimated model of reality as
+a given, and train an RL agent on it.\footnote{
+In this sense, it is important to note that the classical management strategies we compare against have a similar flow of information.
+Namely, data is used to estimate a dynamical model, and this model is used to generate a policy function.
+The difference to our approach is located in the process of *how* the model is used to optimize a policy.
+Because of this difference, RL-based approaches can produce good heuristic solutions for complex problems.
+}
+
+Specifically, we consider four environments corresponding to each of the
+models considered (Table \ref{tab:models}). At each time step, the agent
+observes the state \(S\) and enacts some harvest---reducing \(X_t\) to
+\(X_t - M_X(N_t)\cdot X_t\), and, for Models 3 and 4, also reducing
+\(Y_t\) to \(Y_t - M_Y(N_t)\cdot Y_t\). Here the fish
+mortality-rates-from-harvest (i.e.~\(M_X=M_X(N_t)\) and
+\(M_Y=M_Y(N_t)\)), are the agent's action at time \(t\). This secures a
+reward of \(M_X(N_t)X\) for Models 1 and 2, and, similarly, a reward of
+\(M_X(N_t)X + M_Y(N_t)Y\) for Models 3 and 4. After this harvest portion
+of the time step, the environment evolves naturally according to eqs.
+\eqref{eq:may} and \eqref{eq:3d model} (Sec. 2).
+
+As mentioned previously, discretising time allows a simplification of
+the possibly seasonal mating behavioral patterns of the species
+involved. This approximation is commonly used in fisheries for species
+with annual reproductive cycles (see e.g. (Mangel 2006), Chap. 6).
+Moreover, the separation of each time-step into a harvest period and a
+natural growth period assumes that harvest has little disruption on the
+reproductive process. A detailed model which includes such a disruption
+is outside of the scope of this work.
+
+\hypertarget{mathematical-framework-for-rl}{%
+\subsection{Mathematical framework for
+RL}\label{mathematical-framework-for-rl}}
+
+The RL environment can be formally described as a discrete time
+\emph{partially observable Markov decision process (POMDP)}. This
+formalization is rather flexible and allows one, e.g., to account for
+situations where the agent may not fully observe the environment state,
+or where the only observations available to the agent are certain
+functions of the underlying state. For the sake of clarity, we will only
+present here the subclass of POMDPs which are relevant to our work:
+\emph{fully observable MDPs} (henceforth MDPs for short). An MDP may be
+defined by the following data:
+
+\begin{itemize}
+\item
+  \(\mathcal{S}\): \emph{state space}, the set of states of the
+  environment,
+\item
+  \(\mathcal{A}\): \emph{action space}, the set of actions which the
+  agent may choose from,
+\item
+  \(T(N_{t+1}|N_t, a_t, t)\): \emph{transition operator}, a conditional
+  distribution which describes the dynamics of the system (where
+  \(N_i\in\mathcal{S}\) are states of the environment),\footnote{
+  Transition operators are commonly discussed without having a direct time-dependence for simplicity, but the inclusion of $t$ as an argument to $T$ does not alter the structure of the learning problem appreciably.
+  }
+\item
+  \(r(N_t, a_t, t)\): \emph{reward function}, the reward obtained after
+  performing action \(a_t\in\mathcal{A}\) in state \(N_t\),
+\item
+  \(d(N_0)\): \emph{initial state distribution}, the initial state of
+  the environment is sampled from this distribution,
+\item
+  \(\gamma\in[0,1]\): \emph{discount factor}.
+\end{itemize}
+
+At a time \(t\), the MDP agent observes the full state \(s_t\) of the
+environment and chooses an action based on this observation according to
+a \emph{policy function} \(\pi(a_t | N_t)\). In return, it receives a
+discounted reward \(\gamma^t r(a_t, N_t)\). The discount factor helps
+regularize the agent, helping the optimization algorithm find solutions
+which pay off within a timescale of \(t \sim \log(\gamma^{-1})^{-1}\).
+
+With any fixed policy function, the agent will traverse a path
+\(\tau=(N_0,\ a_0,\ N_1,\ a_1\ \dots,\ N_{t_{\text{fin.}}})\) sampled
+randomly from the distribution \begin{align*}
+  p_\pi(\tau) = 
+  d(N_0) \prod_{t=0}^{ t_{\text{fin.}}-1 }
+  \pi( a_t | N_t ) T( N_{t+1} | N_t, a_t, t ).
+\end{align*} Reinforcement learning seeks to optimize \(\pi\) such that
+the expected rewards are maximal, \[  
+  \pi^* = \mathrm{argmax}\ \mathbb{E}_{\tau\sim p_\pi}[R(\tau)],
+\] where, \[
+  R(\tau) = \sum_{t=0}^{ t_{\text{fin.}}-1 } \gamma^t r(a_t, N_t, t),
+\] is the cumulative reward of path \(\tau\). The function
+\(J(\pi):=\mathbb{E}_{\tau\sim p_\pi}[R(\tau)]\) is called the
+\emph{expected return}.
+
+\hypertarget{deep-reinforcement-learning}{%
+\subsection{Deep Reinforcement
+Learning}\label{deep-reinforcement-learning}}
+
+The optimal policy function \(\pi\) often lives in a high or even
+infinite-dimensional space. This makes it unfeasible to directly
+optimize \(\pi\). In practice, an alternative approach is used: \(\pi\)
+is optimized over a much lower-dimensional parameterized family of
+functions.\footnote{
+Policies are, in general, functions from state space to policy space.
+In our paper, these are $\pi:[0,1]^{\times 3}\to \mathbb{R}_+$ for the single fishery case, and $\pi:[0,1]^{\times 3}\to \mathbb{R}_+^2$ for two fisheries.
+The space of all such functions is highly singular, spanning a \emph{non-separable Hilbert space}.
+Even restricting ourselves to continuous policy functions, we end up with a set of policies which span the infinite dimensional space $L^2([0,1]^{\times3})$.
+One way to avoid optimizing over an infinite dimensional ambient space is to discretize state space into a set of bins.
+This approach runs into tractability problems: First, the dimension of policy space scales exponentially with the number of species.
+Second, even for a fixed number of species (e.g., 3), the dimension optimized over can be prohibitively large---for example if one uses 1000 bins for each population in a three-species model, the overall number of parameters being optimized over is $10^9$.
+Neural networks with much smaller number of parameters, on the other hand, can be quite expressive and sufficient to find a rather good (if not optimal) policy function.
+} Deep reinforcement learning uses this strategy, focusing on function
+families parameterized by neural networks. (See Fig. 1 and App. A in
+(Lapeyrolerie et al. 2022) for a conceptual introduction to the use of
+reinforcement learning in the context of conservation decision making.)
+
+We will focus on deep reinforcement learning throughout this paper.
+Within the DRL literature there is a wealth of algorithms from which to
+choose to optimize \(\pi\), each with its pros and cons. Most of these
+are based on gradient ascent by using the technique of
+\emph{back-propagation} to efficiently compute the gradient. Here we
+have used only one such algorithm (\emph{proximal policy optimization
+(PPO)}) to draw a clear comparison between the RL-based and the
+classical fishery management approaches. In practice, further
+improvements can be expected by a careful selection of the optimization
+algorithm. (See, e.g., (François-Lavet et al. 2018) for an overview of
+different optimization schemes used in DRL.)
+
+\hypertarget{model-free-reinforcement-learning}{%
+\subsection{Model-free reinforcement
+learning}\label{model-free-reinforcement-learning}}
+
+Within control theory, the classical setup is one where we use as much
+information from the model as possible in order to derive an optimal
+solution. Here, one may find a vast literature on model-based methods to
+attain optimal, or near-optimal, control (see, e.g., (Zhang, Li, and
+Liao 2019; Sethi and Sethi 2019; Anderson and Moore 2007)).
+
+The classical sustainable fishery management approaches summarized in
+Text Box 1, for instance, are model-based controls. As we saw there,
+these controls may run into trouble in the case where there are
+inaccuracies in the model parameter estimates.
+
+There are many situations, however, in which the exact model of the
+system is not known or not tractable. This is a standard situation in
+ecology: mathematical models capture the most prominent aspects of the
+ecosystem's dynamics, while ignoring or summarizing most of its
+complexity. In this case, it is clear, model-based controls run a grave
+danger of mismanaging the system.
+
+Reinforcement learning, on the other hand, can provide a model-free
+approach to control theory. While a model is often used to generate
+training data, this model is not directly used by model-free RL
+algorithms. This provides more flexibility to use RL in instances where
+the model of the system is not accurately known. In fact, it has been
+shown that model-free RL outperforms model-based alternatives in such
+instances (Janner et al. 2019). (For recent surveys of model-based
+reinforcement learning, which we do not focus on here, see (Moerland et
+al. 2023; Polydoros and Nalpantidis 2017).)
+
+This context provides a motivation for this paper. Indeed, models for
+ecosystem dynamics are only ever approximate and incomplete descriptions
+of reality. This way, it is plausible that model-free RL controls could
+outperform currently used model-based controls in ecological management
+problems.
+
+Model-free DRL provides, moreover, a framework within which agents can
+be trained to be generally competent over a \emph{variety} of different
+models. This could more faithfully capture the ubiquitous uncertainty
+around ecosystem models. The aforementioned framework---known as
+\emph{curriculum learning}---is considerably more intensive on
+computational resources than the ``vanilla'' DRL methods we have used in
+this paper.\footnote{
+All our agents were trained in a local server with two commercial GPUs. 
+The training time was between 30 minutes and one hour in each case.
+} Due to the increased computational requirements of this framework, we
+have left the exploration in this direction for future work.
+
+\hypertarget{methods}{%
+\section{Methods}\label{methods}}
+
+\textbf{The Environment.} We considered the problem of managing four
+increasingly complex models (see Table \ref{tab:models}). To recap,
+these four environments are a single-species growth model \eqref{eq:may}
+for a harvested species; a three-species model \eqref{eq:3d model} with
+a single harvested species; the same three-species model but with
+\emph{two} harvested species; and, finally, a three-species model with a
+time-varying parameter and two harvested species.
+
+The policies explored were functions from states \(N_t\) to either a
+single number \(M_X(N_t) = \pi(N_t)\) (for Models 1 and 2), or a pair of
+numbers \((M_X(N_t), M_Y(N_t)) = \pi(N_t)\) (for Models 3 and 4).
+
+Our goal was to evaluate the performance of different policy strategies
+over a specified window of time. We chose this window to be 200
+time-steps, where each discrete time-step represents the dynamical
+evolution of the system over a year. Each time-step was composed of two
+parts: First, a \emph{harvest period} where the harvest is collected
+from the system (e.g.~the population \(X\) is reduced to
+\(X_t\mapsto X_t - M_X(N_t) X_t\)). Second, a \emph{recruitment and
+interaction period}, where the system's state evolves according to its
+natural dynamics.
+
+Training and evaluating management strategies were performed by
+simulating \emph{episodes.} An episode begins at a fixed initial state
+and the system is controlled with a management policy until \(t=200\),
+or until a ``near-extinction event'' occurs---that is, until any of the
+populations go below a given threshold, \begin{align}
+  \label{eq:thresh}
+  X_t \leq X_{\text{thresh.}},\quad
+  Y_t \leq Y_{\text{thresh.}},\quad
+  \text{or,}\quad
+  Z \leq Z_{\text{thresh.}}.
+\end{align} In our setting we have chosen
+\(X_{\text{thresh.}}=Y_{\text{thresh.}}=Z_{\text{thresh.}}=0.05\) as a
+rule of thumb---given that under natural (unharvested) dynamics the
+populations range within values of 0.5 to 1, this would represent on the
+order of a 90-95\% population decrease from their ``natural'' level.
+
+The reward function defining our policy optimization problem had two
+components. The first was economical: the total biomass harvested over
+an episode. The second seeked to reflect conservation goals: if a
+near-extinction event occurred at time \(t\), the episode was ended
+early and a negative reward of \(-100/t\) was awarded. This reward
+function balanced the extractive motivation of the fishery with
+conservation goals which went beyond the scope of long-term sustainable
+harvests commonly used in fishery management.
+
+\textbf{Training a DRL agent.} We trained a DRL agent parametrized by a
+neural network with two hidden, 64-neuron, layers, on a local server
+with 2 commercial GPUs. We used the Ray
+framework\footnote{\url{https://docs.ray.io/}} for training,
+specifically we used the Ray PPOConfig class to build the policy
+optimization algorithm using the default values for all
+hyper-parameters. In particular, no hyperparameter tuning was performed.
+The agent was trained for 300 iterations of the PPO optimization step
+for the three-species cases. The total training time was on the order of
+30 minutes to 1 hour. For the single-species model, the training
+iterations were scaled down to 100 and the training time was around 10
+minutes.
+
+The state space used was normalized case-by-case as follows: \emph{Model
+1:} a line segment \([0,1]\), \emph{Models 2-4:} a cube
+\([0,1]^{\times 3}\). We used simulated data to derive a bound on the
+population sizes typically observed, and thus be able to normalize
+states to a finite volume.
+
+Policies obtained from the PPO algorithm can be ``noisy'' as their
+optimization algorithm is randomized (see, e.g.~App. B for a
+visualization of the PPO policy obtained for Model 4). We smoothed this
+policy out using a Gaussian process regressor interpolation. Details for
+this interpolation process can be found in App. C.
+
+\textbf{Tuning the CMort strategy.} In order to estimate the optimum
+mortality rate, we optimized over a grid of possible mortality rates.
+Namely, for Models 1 and 2, grid of 101 mortality rates was laid in the
+interval \([0, 0.5]\); for Models 3 and 4, a \(51\times51\) grid was set
+in the square \([0,0.5]^2\). The latter had a slightly coarser grid due
+to the high memory cost of using the denser grid. Since the approach for
+tuning was completely analogous for all models evaluated, here we
+discuss only Model 4. For each one of these choices of mortality rates,
+say \((M_X, M_Y)\), we simulated 100 episodes based on
+\eqref{eq:3d model}: At each time step the state \((X_t, Y_t, Z_t)\) was
+observed, a harvest of \(M_X X_t+M_YY_t\) was collected and then the
+system evolved according to its natural dynamics \eqref{eq:3d model}.
+The optimal mortality rate was the value \((M_X^*,\ M_Y^*)\) for which
+the mean episode reward was maximal.
+
+\textbf{Tuning the CEsc strategy.} This tuning procedure was analogous
+to that of the CMort strategy just summarized. Namely: A grid of 101
+escapement values was laid out on the interval \([0,1]\) for Models 1
+and 2; and a \(51\times 51\) grid on \([0,1]^2\) was laid out for Models
+3 and 4. Each grid point represented a CEsc policy. We used each of
+these policies to manage 100 replicate episodes. The optimal policy was
+the policy with the highest average reward obtained. A visualization of
+the tuning outcome for Model 4 is shown in Fig. \ref{fig:tuning}.
+
+\begin{figure}
+  \centering
+  \includegraphics[scale=0.5]{figures/esc_heatmap.png}
+  \caption{Visualization of the cosntant escapement strategy tuning procedure for Model 4.
+There was a certain multiplicity in this tuning strategy: a ‘‘ridge of optimality’’ where policies had essentially equivalent behavior.
+Throughout our investigation, we tuned constant escapment in several occasions and, on each occassion, a different optimal policy along the ridge was found.
+The results for different policies along the ridge were in practice equivalent, with no discernible difference in performance.
+We highlighted the ridge with a white dotted line.}
+  \label{fig:tuning}
+\end{figure}
+
+\textbf{Parameter values used.} The single-species model's (eq.
+\eqref{eq:may}) dynamic parameters were chosen as \begin{align}
+\label{eq:1d param vals}
+r = K = 1,\quad \beta = 0.25,\quad c = 0.1.
+\end{align} Here, the values of \(\beta\) and \(c\) were chosen as to
+make the system be roughly close to its tipping point.
+
+For Models 2 and 3, their dynamic parameters (in eq.
+\eqref{eq:3d model}) were chosen as follows \begin{align}
+\label{eq:param vals}
+\begin{split}
+  r_X = K_X = r_Y = K_Y = 1, \quad \beta = 0.3, \quad c = 0.3\\
+  c_{XY} = 0.1,\quad b = 0.1,\quad D = 1.1, \quad d_Z = 0.1.
+\end{split}
+\end{align} Moreover, the variances for the stochastic terms were
+chosenas \[
+  \sigma^2(\eta_{X,t}) = \sigma^2(\eta_{Y,t}) = \sigma^2(\eta_{Z,t}) = 0.05.
+\] For Model 4, we used \(r_X(t)\) given as in eq. \eqref{eq:rx(t)} and
+all other parameters were given as in eq. \eqref{eq:param vals}.
+
+The values of \(c\) and \(\beta\) in the three-species model were
+slightly different than their values in the one-species model. These
+values were chosen heuristically: We observed that choosing the lower
+value of \(c=0.1\) in this case would lead to quick near-extinction
+events even without a harvest. Moreover, \(\beta\) was slightly
+increased simply to put more predation pressure on the \(X\) and \(Y\)
+populations and make them slightly more fragile.
+
+\textbf{Stability analysis.} To ensure that our results do not strongly
+depend on our parameter choices, we performed a stability analysis.
+Here, we perturbed parameters randomly and measured the difference in
+performance between our DRL-based methods and the CEsc strategy. We
+observed that the difference in performance is maintained for even
+relatively high noise strengths. We only considered the most complex
+case here, Model 4.
+
+For each value of parametric noise strength \(\sigma_{\text{param.}}\),
+we executed the following procedure: We sampled 100 choices of perturbed
+parameter values, where the perturbation was as follows---each parameter
+\(P\) was perturbed to \((1+g_P)P\) where \(g_P\) was a Gaussian random
+variable with variance \(\sigma_{\text{param.}}^2\). For each of these
+sample parameter sets, we tuned CEsc and trained the DRL agent. We
+measured the average reward difference between these two strategies for
+each of sample (this was done using 100 replicate evaluation episodes).
+Finally, we took the mean of this average difference over the 100
+perturbed parameter samples.
+
+The parametric noise strength values used were
+\([0.04, 0.08, \dots, 0.2]\).
+
+\hypertarget{results}{%
+\section{Results}\label{results}}
+
+We evaluated each of the four management strategies considered on Models
+1-4. To recap, the management strategies were CEsc, CMort, PPO and a
+Gaussian process interpolation of PPO (``PPO+G''). Furthermore, we
+characterized the trade-off between economic output and sustainability
+faced by CMort policies. This was done by evaluating CMort policies with
+a fraction of the optimal mortality rate (specifically, 80\%, 90\% and
+95\%). All evaluations were based on 100 replicate episodes.
+
+We will visualize the results concerning Model 4 in this section,
+leaving the other models for App. A. This is the most complex scenario
+considered and where our results show the most compelling advantage of
+DRL methods with respect to classical strategies.
+
+Our main result is summarized in Fig. \ref{fig:main} which displays the
+total reward distributions for the policy obtained through each
+strategy. Here we see that CEsc has a long-tailed distribution of
+rewards, and its average reward is much lower than other management
+strategies. CMort has a shorter-tailed distribution and a much higher
+average reward. Finally, both DRL-based strategies have a more
+concentrated reward distribution and a higher average reward than CMort.
+
+\begin{figure}
+\includegraphics[width=6in]{manuscript_files/figure-latex/main-1} \caption{Reward distributions for the four strategies considered.
+These are based on 100 evaluation episodes.
+We denote CEsc for constant escapement, CMort for constant mortality, PPO for the output policy of the PPO optimization algorithm, and PPO GP for the Gaussian process interpolation of the PPO policy.}\label{fig:main}
+\end{figure}
+
+To assess what is the culprit for the classical strategies' low
+performance with respect to the DRL-based strategies, we plot the
+duration of each of the 100 evaluation episodes of each strategy in Fig.
+\ref{fig:ephistograms}. We see that early episode ends are prevalent in
+classical strategies and rare for DRL-based strategies. Early episode
+ends tend to happen at lower \(t\) values for CEsc than CMort. Thus,
+distribution of episode durations seems to be widest for CEsc, followed
+by CMort, DRL and DRL+GP.
+
+\begin{figure}
+\includegraphics[width=6in]{manuscript_files/figure-latex/ephistograms-1} \caption{Hisotgrams of episode lengths and rewards for the four different management strategies considered.
+Only the first 50 evaluation episodes (from a total of 100) were included, for ease of visualization.
+From left to right, the four management strategies compared are CEsc, CMort, PPO, and PPO+GP.}\label{fig:ephistograms}
+\end{figure}
+
+We examine the trade-off between profit and sustainability faced by the
+CMort strategy in Fig. \ref{fig:fracmsy}. Two quantities are plotted:
+the fraction of evaluation episodes with maximal length (i.e.~episodes
+with no near-extinction events), and the average reward. On the x-axis
+we have several sub-optimal mortality rates that err on the conservative
+side: e.g.~the policy labeled ``80\% Opt. CMort'' has the form
 \begin{align*}
-  L(V \div r,\ K) := rV\left(1 - V / K\right).
-\end{align*}
-Moreover, $V$ is predated on by a (constant) population $H$, as can be seen from the negative term
-\begin{align*}
-  F(V,\ H \div \beta,\ V_0) := \frac{\beta H V^2}{V_0^2 + V^2}
-\end{align*}
-which saturates to $\beta H$ as $V\to\infty$, and whose half maximum is $V_0$, i.e.\ $F(V=V_0,\ H; \beta,\ V_0)=\beta H/2$.
+  \pi: (X_t, Y_t, Z_t)\mapsto (0.8 M_X^*,\ 0.8 M_Y^*),
+\end{align*} where \((M_X^*,\ M_Y^*)\) is the optimal CMort strategy. We
+see that sufficiently conservative policies attain high sustainability,
+but only at a high price in terms of profit.\footnote{
+As noted before, here we equate economic profit with biomass caught.
+This is done as an approximation to convey the conceptual message more clearly, and we do not expect our results to significantly change if, e.g., ‘‘effort cost’’ is included in the reward function.
+When we refer to ‘‘large differences’’ in profit, or ‘‘paying dearly,’’ we mean that the ratio between average rewards is considerable---e.g. a 15\% loss in profit.
+} We expect a similar and, possibly, more pronounced effect for the CEsc
+strategy but do not analyze this case here.
 
-Ref.~\cite{may77} studies the fixed points of~\eqref{eq:may77} in order to show that in certain parameter regimes, its dynamics can undergo a \emph{catastrophe}.
-A catastrophe is a sudden change in the state of the system from one stable state to another---often, the final state is ecologically detrimental, possibly associated with extinction or near-extinction events.
+\begin{figure}
+\includegraphics[width=6in]{manuscript_files/figure-latex/fracmsy-1} \caption{Trade-off between reward and probability of a near-extinction event for CMort policies.
+We evaluated policies at the full optimal constant escapement value, and also at 0.8, 0.9, 0.95 of the latter.
+Each evaluation is based on 100 episodes.
+On the left we plot the percentage of episodes which last their maximum time window, i.e.
+that do not see a near-extinction event.
+On the right, we plot the mean episdoe reward and standard deviation for each policy.}\label{fig:fracmsy}
+\end{figure}
 
-Fig.~\ref{fig:tbd} shows the stable $V$ populations for differing values of $H$. 
-Here one sees that as $H\to T_2$, the top stable state is annihilated with the unstable fixed point.
-This way, if, e.g.\ $H$ were to slowly drift until $H=T_2$, the system would collapse to the low stable state, leading to a near extinction of $V$.
+One problem that is often encountered when using machine learning
+methods is the interpretability of the output of these methods. For our
+PPO strategy, the output is a policy function parametrized by a neural
+network \(\theta\): \begin{align*}
+  \pi_\theta: N_t \mapsto (M_X(N_t), M_Y(N_t)),
+\end{align*} where \(N_t=(X_t,Y_t,Z_t)\) is the state of the system at
+time \(t\), and where \(M_X\) and \(M_Y\) are the mortalities due to
+harvest during that time-step. While the values of the neural network
+parameters are hard to interpret, the actual shape of the policy
+function is much more understandable.
 
-% For fixed values of $r,\ K,\ V_0$ and $\beta$, varying $H$ may produce a \emph{catastrophe} (see Fig.~\ref{fig:tbd})---a sudden disappearance of a stable fixed point of the system which .
+Here we visualize the PPO+GP policy function and provide an
+interpretation for it, as this function is smoother and less noisy than
+the PPO policy function. The PPO policy function is visualized similarly
+in App. B.
 
-\hypertarget{sec:5}{%
-\section{Results}\label{sec:results}}
+Given its high dimension, it is not possible to fully display how the
+policy function obtained ``looks like''---we thus project it down to
+certain relevant axes. The result of this procedure is shown in Fig.
+\ref{fig:gpp-policy}. In that figure, the shape of the optimal
+escapement strategy is provided for comparison.
 
-\hypertarget{sec:6}{%
-\section{Discussion}\label{sec:discussion}}
+\begin{figure}
+  \centering
+  \includegraphics[scale=0.25]{figures/gpp-policy-pdf.pdf}
+  \caption{Plots of the PPO+GP policy $\pi_{\text{PPO+GP}}$ along several relevant axes.
+Here $M_X$ and $M_Y$ are the $X$ and $Y$ components of the policy function.
+The values of the plots are generated in the following way: For each variable $X$, $Y$, and $Z$, the time-series of evaluation episodes are used to generate a window of typical values that variable attains when controlled by $\pi_{\text{PPO+GP}}$.
+Then, for each plot either $X$ or $Y$ was varied on $[0,1]$ along the $x$ axis, while the other variables (resp.
+$Y$ and $Z$, or $X$ and $Z$) were varied within the typical window computed before.
+The value of one of the latter two variables were visualized as color.}
+  \label{fig:gpp-policy}
+\end{figure}
 
-\hypertarget{paragraph-headings}{%
-\paragraph{Paragraph headings}\label{paragraph-headings}}
+We notice that the DRL-derived policy has similarities to a CEsc policy.
+Here, the key difference is that the escapement value for each of the
+fished species is sensitive to variations in the other populations. This
+can be seen as color gradients in the plots of \((X, M_X)\) and
+\((Y, M_Y)\), where the gradient corresponds to differing values of
+\(Z\). Moreover, this can be seen as an anti-correlation in the plot of
+\((X, M_Y)\)---for optimal CEsc, \(M_Y\) is uncorrelated to \(X\).
 
-Use paragraph headings as needed. 
-Use paragraph headings as needed. 
-Use paragraph headings as needed. 
-Use paragraph headings as needed. 
-Use paragraph headings as needed. 
-Use paragraph headings as needed. 
-Use paragraph headings as needed.
+This sensitivity of the policy to, for instance, the values of \(Z\) can
+be seen in the sample time series displayed in Fig. \ref{fig:gpp-ep}.
+Here, we can see that species \(Z\) becomes endangered due to harvesting
+for all management strategies. The DRL-based strategy, however, is
+sensitive to the values of \(Z\) and can respond accordingly by scaling
+the fishing effort with the value of \(Z\). In particular, the policy
+responds to the period of diminishing values of \(Z\) near the end of
+the episode, by restricting fishing on \(X\) and \(Y\), thus promoting
+\(Z\)'s growth. This pattern is rather common among the whole
+dataset---early episode ends are largely due to near-extinctions of
+\(Z\) for all management strategies.
 
-\begin{align}
-a^2+b^2=c^2
-\end{align}
+\begin{figure}
+  \centering
+  \includegraphics[scale=0.33]{figures/timeseries-gpp.jpeg}
+  \caption{Time-series of an episode managed with $\pi_{\text{PPO+GP}}$.
+Here we plot the state of the system on the bottom panel, and the actions taken (fishing efforts on $X$ and $Y$---respectively $M_X$ and $M_Y$) on the top panel.}
+  \label{fig:gpp-ep}
+\end{figure}
+
+\hypertarget{recovering-constant-escapement-for-a-single-species}{%
+\subsection{Recovering constant escapement for a single
+species}\label{recovering-constant-escapement-for-a-single-species}}
+
+While the optimal control for our single-species model \eqref{eq:may}
+can not be easily proven to be CEsc (since the right-hand side of that
+equation is not concave), from experience we can expect CEsc to either
+be optimal or near-optimal. We give evidence for this intuition by
+showing that our DRL method recovers a CEsc solution when trained. These
+results are shown in Fig. \ref{fig:1sp-pol} Here we show both the output
+PPO policy, and its Gaussian process interpolation. This helps build an
+intuition about the relationship between our ``PPO'' and ``PPO+GP''
+management strategies.
+
+There is a presence of certain high-mortality points at low \(X\) values
+in the PPO policy (which in turn generates a rising fishing mortality
+for \(X\) values below a certain threshold in the PPO+GP policy). This
+is likely due to experience of near-extinctions early on in the training
+process---where, given an impending extinction, there is a higher reward
+for intensive fishing. These ``jitters'' are likely not fully erased
+through the optimization algorithm since near-extinction events become
+extremely rare after only a few training iterations. This way, the agent
+does not further explore that region of state space to generate new
+experience. We believe the most important aspect of the CEsc policy
+reproduced by PPO is the fact that there is some sufficiently-wide
+window below the threshold of the policy (i.e.~below the optimal
+escapement value), on which no fishing is performed. That is, there
+exists some sufficiently large \(\varepsilon\) such that if
+\(X_{\text{thresh.}}\) is the optimal escapement value of the system,
+then \(\pi_{PPO}(X)=0\) for all
+\(X\in[X_{\text{thresh.}}-\varepsilon,\ X_{\text{thresh.}}]\).
+
+\begin{figure}
+  \centering
+  \includegraphics[scale=0.25]{figures/1sp-policy.jpeg}
+  \caption{Left panel: the policy obtained from 100 training iterations of the PPO algorithm on the ‘‘single species, single fishery’’ model.
+Right panel: the Gaussian process interpolation of the left panel.
+We plot both as scatter data evaluated on a 101-point grid on $[0,1]$, but these policies may of course be evaluated continuously---on any possible value of $X$.}
+  \label{fig:1sp-pol}
+\end{figure}
+
+\hypertarget{stability-analysis}{%
+\subsection{Stability analysis}\label{stability-analysis}}
+
+In this section we present results intended to show that the effects
+that we observe in this paper are not the result of a careful selection
+of parameter values, but rather arise for a wide variety of parameter
+values.
+
+Our main result in this respect is Fig. \ref{fig:stability}. There, we
+plot the \emph{average episode reward difference} between the two
+DRL-based methods we considered, and the optimal CEsc strategy. This
+figure shows that, for a wide range of parameter values, DRL-based
+strategies can have a considerable advantage over an optimized CEsc
+policy (the single-species optimal solution).
+
+\begin{figure}
+  \centering
+  \includegraphics[scale=0.8]{figures/jiggle_v1.png}
+  \caption{Mean reward difference between DRL methods (resp.
+our ‘‘PPO’’ and ‘‘PPO+GP’’ strategies), on the one hand, and the optimal constant escapement policy (‘‘CEsc’’) on the other.
+The dynamic parameters in eq.
+\eqref{eq:3d model} were randomly perturbed from the values given in \eqref{eq:param vals} according to the procedure detailed in Sec.
+4.
+The noise strength of this perturbation is plotted as the x-axis.
+For each noise strength, 100 parameter perturbations were sampled---each one giving rise to a \emph{realization} of the model.
+For each such realization, we optimized a CEsc policy and trained a PPO agent.
+Moreover, we interpolated the PPO policy using a Gaussian process, as detailed in Sec 4.
+Then, for each realization we compared the performance of these policies: we measured the mean reward difference betwen PPO and CEsc, and between PPO+GP and CEsc.
+The plot represents the distribution of reward differences observed at a given noise strength---we plot the mean and the standard deviation of the mean reward differences observed.
+In an equation, we plot the means $\mathbb{E}_{P}[\mu_P^{\text{DRL}}-\mu_P^{\text{CEsc}}]$, where $P$ are the paramter values, $\mu_P^{\text{DRL}}$ is the mean reward for a DRL policy trained on the problem with parameter values $P$, and, similarly, $\mu_P^{\text{CEsc}}$ is the mean reward of the optimal constant escpament policy for parameter values $P$.}
+  \label{fig:stability}
+\end{figure}
+
+\hypertarget{discussion}{%
+\section{Discussion}\label{discussion}}
+
+Fisheries are complex ecosystems, with interactions between species
+leading to highly non-linear dynamics. While current models for
+population estimation include many important aspects of this complexity,
+it is still common to use simplified dynamical models in order to guide
+decision making. This provides easily interpretable solutions, such as
+CMort policies. There is a drawback here, however: due to the simplicity
+of these dynamical models, the policies might not respond effectively in
+many situations---situations where the predictions of these simple
+models deviate considerably from reality. Because of this, policies such
+as MSY have faced pushback and are believed to have contributed to the
+depletion of fish stocks (Worm et al. 2006; Costello et al. 2016).
+
+We propose an alternative approach to the problem of fishery control: to
+use a more expressive -- albeit more complicated -- dynamical model to
+guide decision making. Furthermore, rather than computing the optimal
+control policy for the model (something that is impossible in practice
+for complex systems), we use deep reinforcement learning to obtain a
+``pretty darn good'' policy. This policy is estimated in a
+\emph{model-free} setting, i.e., the agent treats the dynamical model
+(e.g.~eq. \eqref{eq:3d model}) as a black box of input-output pairs. By
+not relying on precise knowledge of the model's parameter values, but
+rather just relying on input-output statistics, model-free approaches
+have gained traction in a variety of control theory settings (see, e.g.,
+(Sato 2019; Ramirez, Yu, and Perrusquia 2022; Zeng et al. 2019)).
+
+We compare deep reinforcement learning-based policies against classical
+management strategies (CMort and CEsc). While the latter are inspired by
+the shape of optimal solutions in the single-species setting, they are
+optimized in a model-free way as well: e.g.~the optimal mortality rate
+is computed empirically from simulated data.
+
+Because of the simplicity of the classical policy functions, the optimal
+such policy may be easily estimated through a grid-search. This is the
+case since these policy functions are only specified by one or two
+parameters (respectively in the single fishery, or two fishery cases).
+In contrast, DRL optimizes over the more expressive -- and more
+complicated -- family of policies parametrized by a neural network.
+Neural networks are often used as flexible function approximators that
+can be efficiently optimized over.
+
+We showed that for sufficiently complex management scenarios -- Models 3
+and 4 -- DRL-based management strategies perform significantly better
+than CEsc. This with respect to both average rewards received and
+conservation goals. In this sense, an approximate solution to a more
+complicated and expressive model, can outperform the optimal solution of
+the single-species problem---even when the parameters of the
+single-species solution are empirically optimized.
+
+We found that the optimal CMort policy surprisingly performs much better
+than CEsc (Fig. \ref{fig:main}). However, it can be observed in Fig.
+\ref{fig:fracmsy} that the CMort strategy faces a trade-off: high
+sustainability is achieved for sub-optimal mortality rates, but only at
+a significant decrease in the mean episode reward. We expect that with
+increasing ecosystem complexity this phenomenon might become more
+pronounced. We can understand this as a consequence of the rigidity of
+classical strategies: the simplicity of their expressions, depending
+only on a few parameters, means that policy optimization is constrained
+to a rather reduced subset of the space of possible policies.
+
+The question of when multi-species models are well-approximated by
+single-species models was studied in detail in (Burgess et al. 2017).
+Here our approach is dual to that of the aforementioned paper. Rather
+than first optimizing a single-species model to approximate a more
+complex model and then finding the MSY value for the single-species
+model, we used simulated data to optimize CMort and CEsc directly on the
+three-species model. We do not investigate further whether our
+three-species models are well approximated by a single-species model in
+the sense of (Burgess et al. 2017). However: \emph{1)} Because the
+interaction terms in \eqref{eq:3d model} are about an order of magnitude
+smaller than \(r_X\) and \(r_Y\), Models 2-4 are ``close'' in
+parameter-space to a single-species model. \emph{2)} The fact that for
+Model 2 all strategies match in performance suggests that
+\eqref{eq:3d model} might be well-approximated by a single species
+model. This in turn suggests that the reason that DRL outperforms both
+single-species strategies for Models 3 and 4 is not due a lack of a
+single-species approximation for either \(X\) or \(Y\), but due to the
+complexity of having two harvested species. Moreover, the
+non-stationarity in Model 4 maintained the advantage of DRL over CEsc
+and CMort. Here, one may have expected an exacerbation of that advantage
+due to non-stationarities introducing biases to single-species
+approximate models (Burgess et al. 2017). We did, however, measure a
+decrease in the sustainability of CEsc in Model 4 with respect to Model
+3.
+
+Finally, we performed a stability analysis to ensure that the advantage
+of DRL-based techniques over CEsc is a ubiquitous phenomenon and not a
+result of a lucky selection of parameter values. We found (in Fig.
+\ref{fig:stability}) that an advantage can be observed even for
+relatively high-noise perturbations of the parameters---noise with a
+variance of 20\% of the parameter values. The aforementioned Figure
+summarizes the statistics of 100 parameter perturbations so that we may
+expect perturbations of up to 60\% (i.e.~three sigmas) in the parameter
+values to appear in these statistics.
+
+\hypertarget{future-directions}{%
+\subsection{Future directions}\label{future-directions}}
+
+There are a number of interesting directions that would be interseting
+to explore in future work.
+
+First, benchmarking our results against increasingly more compex and
+realistic fishery models. This would include non-stationarities in the
+dynamical parameters to accurately reflect the effects of climate change
+on the ecosystem. This added complexity would likely pose a
+computational challenge---in future work we will likely need to test
+several different DRL training algorithms (see e.g. (Lapeyrolerie et al.
+2022) for a non-exhaustive list), and it is very likely that
+hyperparameter tuning will need to be performed. Moreover, it may be
+that larger neural networks than the one we used in this research will
+be needed for the policy function. This all will mean that considerable
+technical work will be needed in order to make this next step
+computationally feasible (e.g.~we might need to make more extensive use
+of GPUs and parallelization).
+
+Second, to account for noisy estimates of the system's state and
+imperfect policy implementation. This could be done straightforwardly,
+albeit it might increase the training time before DRL approaches
+converge, as well as introducing the need for hyperparameter tuning.
+
+Third, to account for the systematic uncertainties behind the dynamics
+of the ecosystem---that is, to account for model biases with respect to
+reality. Here, one can employ tools from curriculum learning in order to
+train an agent that is \emph{generally capable} of good management over
+a range of different dynamical models. This way, one can incorporate
+different models---expressing different aspects of the ecosystem---into
+the learning process of the agent. We believe that this step will likely
+be necessary if DRL algorithms are to be applied successfully in the
+fishery management problem. Curriculum learning is rather expensive
+computationally, however, and involves a non-trivial \emph{curriculum
+design} which will guide the agent in its learning process. This way,
+considerable technical work would be needed for this direction.
+
+\hypertarget{acknowledgements}{%
+\subsection{Acknowledgements}\label{acknowledgements}}
+
+The title of this piece references a mathematical biology workshop at
+NIMBioS organized by Paul Armsworth, Alan Hastings, Megan Donahue, and
+Carl Towes in 2011 which first sought to emphasize `pretty darn good'
+control solutions to more realistic problems over optimal control to
+idealized ones. This material is based upon work supported by the
+National Science Foundation under Grant No.~DBI-1942280.
+
+\appendix
+
+\hypertarget{appendix-results-for-stationary-models}{%
+\section{Appendix: Results for stationary
+models}\label{appendix-results-for-stationary-models}}
+
+In the main text we focused on the non-stationary model (``three
+species, two fisheries, non-stationary'' in Table 1) for the sake of
+space and because our results were most compelling there. Here we
+present the reward distributions for the other models considered---the
+three stationary models, lines 1-3 in Table 1. These results are shown
+in Figs. \ref{fig:main_32}, \ref{fig:main_31} and \ref{fig:main_11}.
+
+\begin{figure}
+\includegraphics[width=6in]{manuscript_files/figure-latex/main_32-1} \caption{Reward distributions for the four strategies considered.
+These are based on 100 evaluation episodes of Model 3 in Table 1.
+We denote CEsc for constant escapement, CMort for constant mortality, PPO for the output policy of the PPO optimization algorithm, and PPO GP for the Gaussian process interpolation of the PPO policy.}\label{fig:main_32}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=6in]{manuscript_files/figure-latex/main_31-1} \caption{Reward distributions for the four strategies considered.
+These are based on 100 evaluation episodes of Model 2 in Table 1.
+We denote CEsc for constant escapement, CMort for constant mortality, PPO for the output policy of the PPO optimization algorithm, and PPO GP for the Gaussian process interpolation of the PPO policy.}\label{fig:main_31}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=6in]{manuscript_files/figure-latex/main_11-1} \caption{Reward distributions for the four strategies considered.
+These are based on 100 evaluation episodes of Model 1 in Table 1.
+We denote CEsc for constant escapement, CMort for constant mortality, PPO for the output policy of the PPO optimization algorithm, and PPO GP for the Gaussian process interpolation of the PPO policy.}\label{fig:main_11}
+\end{figure}
+
+\hypertarget{appendix-ppo-policy-function-for-non-stationary-model}{%
+\section{Appendix: PPO policy function for non-stationary
+model}\label{appendix-ppo-policy-function-for-non-stationary-model}}
+
+In the main text, Fig. \ref{fig:gpp-policy}, we presented a
+visualization of the PPO+GP policy function obtained for the ``three
+species, two fisheries, non-stationary'' model. This policy function is
+a Gaussian process regression of scatter data of the PPO policy
+function. In Fig. \ref{fig:ppo-policy} we present a representation of
+this scatter data in a similar format as Fig. \ref{fig:gpp-policy}.
+
+\begin{figure}
+  \centering
+  \includegraphics[scale=0.35]{figures/ppo-policy.jpeg}
+  \caption{Plots of the PPO policy $\pi_{\text{PPO}}$ along several relevant axes.
+Here $M_X$ and $M_Y$ are the $X$ and $Y$ components of the policy function.
+The values of the plots are generated in the following way: For each variable $X$, $Y$, and $Z$, the time-series of evaluation episodes are used to generate a window of typical values that variable attains when controlled by $\pi_{\text{PPO}}$.
+Then, for each plot either $X$ or $Y$ was evaluated on 100 values in $[0,1]$ along the $x$ axis, while the other variables (resp.
+$Y$ and $Z$, or $X$ and $Z$) were varied within the typical window computed before.
+Within this window, 5 values are used.
+The value of one of the latter two variables were visualized as color.
+This scatter data was used as an input to generate $\pi_{\text{PPO+GP}}$, visualized in the main text.}
+  \label{fig:ppo-policy}
+\end{figure}
+
+\hypertarget{appendix-gaussian-process-interpolation}{%
+\section{Appendix: Gaussian process
+interpolation}\label{appendix-gaussian-process-interpolation}}
+
+Here we summarize the procedure used to interpolate the PPO policy
+(visualized in Fig. \ref{fig:ppo-policy}). We use the
+\emph{GaussianProcessRegressor} object of the \emph{sklearn} Python
+library with a kernel given by \begin{align*}
+  \text{RBF}(\text{length scale = 10}) + \text{WhiteNoise}(\text{noise level = 0.1}).
+\end{align*} This interpolation method is applied to scatter data of the
+PPO policy evaluated on 3 different grids on \((X,Y,Z)\) states:
+\(G_X\), a \(51\times 5 \times 5\) grid; \(G_Y\), a
+\(5\times 51 \times 5\) grid; and \(G_Z\), a \(5\times 5 \times 51\)
+grid. This combination of grids was used instead of a single dense grid
+in order to reduce the computational intensity of the interpolation
+procedure. For \(G_X\), the 5 values for \(Y\) and \(Z\) were varied in
+a ``popular window,'' i.e.~episode time-series data was used to
+determine windows of \(Y\) and \(Z\) values which were most likely. The
+grids \(G_Y\) and \(G_Z\) were generated in a similar fashion,
+\emph{mutatis mutandis}.\footnote{
+The raw dataset is found at the data/results\_data/2FISHERY/RXDRIFT sub-directory in the repository with the source code and data linked above.
+Scatter plots visualizing this policy are shown in App. B.
+} The length scale and noise level values of this kernel were chosen
+arbitrarily---no hyperparameter tuning was needed to produce
+satisfactory interpolation, as will be shown in the results section.
+
+\hypertarget{references}{%
+\section*{References}\label{references}}
+\addcontentsline{toc}{section}{References}
+
+\hypertarget{refs}{}
+\begin{CSLReferences}{1}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-anderson2007optimal}{}}%
+Anderson, Brian DO, and John B Moore. 2007. \emph{Optimal Control:
+Linear Quadratic Methods}. Courier Corporation.
+
+\leavevmode\vadjust pre{\hypertarget{ref-atari1}{}}%
+Bellemare, Marc G, Yavar Naddaf, Joel Veness, and Michael Bowling. 2013.
+{``The Arcade Learning Environment: An Evaluation Platform for General
+Agents.''} \emph{Journal of Artificial Intelligence Research} 47:
+253--79.
+
+\leavevmode\vadjust pre{\hypertarget{ref-burgess2017describing}{}}%
+Burgess, Matthew G, Henrique C Giacomini, Cody S Szuwalski, Christopher
+Costello, and Steven D Gaines. 2017. {``Describing Ecosystem Contexts
+with Single-Species Models: A Theoretical Synthesis for Fisheries.''}
+\emph{Fish and Fisheries} 18 (2): 264--84.
+
+\leavevmode\vadjust pre{\hypertarget{ref-millie-3}{}}%
+Chapman, Melissa, Lily Xu, Marcus Lapeyrolerie, and Carl Boettiger.
+2023. {``Bridging Adaptive Management and Reinforcement Learning for
+More Robust Decisions.''} \emph{Philosophical Transactions of the Royal
+Society B (Accepted)}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Clark1990}{}}%
+Clark, Colin W. 1990. \emph{{Mathematical Bioeconomics: The Optimal
+Management of Renewable Resources, 2nd Edition}}. Wiley-Interscience.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Clark1973}{}}%
+Clark, Colin W. 1973. {``Profit Maximization and the Extinction of
+Animal Species.''} \emph{Journal of Political Economy} 81 (4): 950--61.
+\url{https://doi.org/10.1086/260090}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-collins2001internal}{}}%
+Collins, MSFB, SFB Tett, and C Cooper. 2001. {``The Internal Climate
+Variability of HadCM3, a Version of the Hadley Centre Coupled Model
+Without Flux Adjustments.''} \emph{Climate Dynamics} 17: 61--81.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Costello2016}{}}%
+Costello, Christopher, Daniel Ovando, Tyler Clavelle, C. Kent Strauss,
+Ray Hilborn, Michael C. Melnychuk, Trevor A Branch, et al. 2016.
+{``{Global fishery prospects under contrasting management regimes}.''}
+\emph{Proceedings of the National Academy of Sciences} 113 (18):
+5125--29. \url{https://doi.org/10.1073/pnas.1520420113}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-fusion1}{}}%
+Degrave, Jonas, Federico Felici, Jonas Buchli, Michael Neunert, Brendan
+Tracey, Francesco Carpanese, Timo Ewalds, et al. 2022. {``Magnetic
+Control of Tokamak Plasmas Through Deep Reinforcement Learning.''}
+\emph{Nature} 602 (7897): 414--19.
+
+\leavevmode\vadjust pre{\hypertarget{ref-franccois2018introduction}{}}%
+François-Lavet, Vincent, Peter Henderson, Riashat Islam, Marc G
+Bellemare, Joelle Pineau, et al. 2018. {``An Introduction to Deep
+Reinforcement Learning.''} \emph{Foundations and Trends{\textregistered}
+in Machine Learning} 11 (3-4): 219--354.
+
+\leavevmode\vadjust pre{\hypertarget{ref-gordon2000simulation}{}}%
+Gordon, Chris, Claire Cooper, Catherine A Senior, Helene Banks, Jonathan
+M Gregory, Timonthy C Johns, John FB Mitchell, and Richard A Wood. 2000.
+{``The Simulation of SST, Sea Ice Extents and Ocean Heat Transports in a
+Version of the Hadley Centre Coupled Model Without Flux Adjustments.''}
+\emph{Climate Dynamics} 16: 147--68.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Gordon1954}{}}%
+Gordon, H. Scott, and Chicago Press. 1954. {``{The Economic Theory of a
+Common-Property Resource: The Fishery}.''} \emph{Journal of Political
+Economy} 62 (2): 124--42. \url{https://doi.org/10.1086/257497}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-mbpo}{}}%
+Janner, Michael, Justin Fu, Marvin Zhang, and Sergey Levine. 2019.
+{``When to {Trust} {Your} {Model}: {Model}-{Based} {Policy}
+{Optimization}.''} \emph{arXiv:1906.08253 {[}Cs, Stat{]}}, November.
+\url{http://arxiv.org/abs/1906.08253}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-rl-intro}{}}%
+Lapeyrolerie, Marcus, Melissa S Chapman, Kari EA Norman, and Carl
+Boettiger. 2022. {``Deep Reinforcement Learning for Conservation
+Decisions.''} \emph{Methods in Ecology and Evolution} 13 (11): 2649--62.
+
+\leavevmode\vadjust pre{\hypertarget{ref-mangel2006theoretical}{}}%
+Mangel, Marc. 2006. \emph{The Theoretical Biologist's Toolbox:
+Quantitative Methods for Ecology and Evolutionary Biology}. Cambridge
+University Press.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Marescot2013}{}}%
+Marescot, Lucile, Guillaume Chapron, Iadine Chadès, Paul L. Fackler,
+Christophe Duchamp, Eric Marboutin, and Olivier Gimenez. 2013.
+{``Complex Decisions Made Simple: A Primer on Stochastic Dynamic
+Programming.''} \emph{Methods in Ecology and Evolution} 4 (9): 872--84.
+\url{https://doi.org/10.1111/2041-210X.12082}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-may77}{}}%
+May, Robert M. 1977. {``Thresholds and Breakpoints in Ecosystems with a
+Multiplicity of Stable States.''} \emph{Nature} 269 (5628): 471--77.
+
+\leavevmode\vadjust pre{\hypertarget{ref-atari2}{}}%
+Mnih, Volodymyr, Koray Kavukcuoglu, David Silver, Alex Graves, Ioannis
+Antonoglou, Daan Wierstra, and Martin Riedmiller. 2013. {``Playing Atari
+with Deep Reinforcement Learning.''} \emph{arXiv Preprint
+arXiv:1312.5602}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-moerland2023model}{}}%
+Moerland, Thomas M, Joost Broekens, Aske Plaat, Catholijn M Jonker, et
+al. 2023. {``Model-Based Reinforcement Learning: A Survey.''}
+\emph{Foundations and Trends{\textregistered} in Machine Learning} 16
+(1): 1--118.
+
+\leavevmode\vadjust pre{\hypertarget{ref-chatgpt}{}}%
+OpenAI. 2022. {``{C}hat{G}{P}{T}: {O}ptimizing {L}anguage {M}odels for
+{D}ialogue.''} \url{https://openai.com/blog/chatgpt/}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-polydoros2017survey}{}}%
+Polydoros, Athanasios S, and Lazaros Nalpantidis. 2017. {``Survey of
+Model-Based Reinforcement Learning: Applications on Robotics.''}
+\emph{Journal of Intelligent \& Robotic Systems} 86 (2): 153--73.
+
+\leavevmode\vadjust pre{\hypertarget{ref-pope2000impact}{}}%
+Pope, VD, ML Gallani, PR Rowntree, and RA Stratton. 2000. {``The Impact
+of New Physical Parametrizations in the Hadley Centre Climate Model:
+HadAM3.''} \emph{Climate Dynamics} 16: 123--46.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Punt2016}{}}%
+Punt, André E, Doug S Butterworth, Carryn L de Moor, José A A De
+Oliveira, and Malcolm Haddon. 2016. {``{Management strategy evaluation:
+best practices}.''} \emph{Fish and Fisheries} 17 (2): 303--34.
+\url{https://doi.org/10.1111/faf.12104}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-RAMLegacyDB}{}}%
+RAM Legacy Stock Assessment Database. 2020. {``RAM Legacy Stock
+Assessment Database V4.491.''}
+\url{https://doi.org/10.5281/zenodo.3676088}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-ramirez2022}{}}%
+Ramirez, Jorge, Wen Yu, and Adolfo Perrusquia. 2022. {``Model-Free
+Reinforcement Learning from Expert Demonstrations: A Survey.''}
+\emph{Artificial Intelligence Review}, 1--29.
+
+\leavevmode\vadjust pre{\hypertarget{ref-riahi2017shared}{}}%
+Riahi, Keywan, Detlef P Van Vuuren, Elmar Kriegler, Jae Edmonds, Brian C
+O'neill, Shinichiro Fujimori, Nico Bauer, et al. 2017. {``The Shared
+Socioeconomic Pathways and Their Energy, Land Use, and Greenhouse Gas
+Emissions Implications: An Overview.''} \emph{Global Environmental
+Change} 42: 153--68.
+
+\leavevmode\vadjust pre{\hypertarget{ref-sato2019}{}}%
+Sato, Yoshiharu. 2019. {``Model-Free Reinforcement Learning for
+Financial Portfolios: A Brief Survey.''} \emph{arXiv Preprint
+arXiv:1904.04973}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Schaefer1954}{}}%
+Schaefer, Milner B. 1954. {``{Some aspects of the dynamics of
+populations important to the management of the commercial marine
+fisheries}.''} \emph{Bulletin of the Inter-American Tropical Tuna
+Commission} 1 (2): 27--56. \url{https://doi.org/10.1007/BF02464432}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-fusion2}{}}%
+Seo, J, Y-S Na, B Kim, CY Lee, MS Park, SJ Park, and YH Lee. 2022.
+{``Development of an Operation Trajectory Design Algorithm for Control
+of Multiple 0D Parameters Using Deep Reinforcement Learning in KSTAR.''}
+\emph{Nuclear Fusion} 62 (8): 086049.
+
+\leavevmode\vadjust pre{\hypertarget{ref-sethi2019optimal}{}}%
+Sethi, Suresh P, and Suresh P Sethi. 2019. \emph{What Is Optimal Control
+Theory?} Springer.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Worm2006}{}}%
+Worm, Boris, Edward B Barbier, Nicola Beaumont, J Emmett Duffy, Carl
+Folke, Benjamin S Halpern, Jeremy B C Jackson, et al. 2006. {``{Impacts
+of biodiversity loss on ocean ecosystem services.}''} \emph{Science (New
+York, N.Y.)} 314 (5800): 787--90.
+\url{https://doi.org/10.1126/science.1132294}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-zeng2019}{}}%
+Zeng, Deze, Lin Gu, Shengli Pan, Jingjing Cai, and Song Guo. 2019.
+{``Resource Management at the Network Edge: A Deep Reinforcement
+Learning Approach.''} \emph{IEEE Network} 33 (3): 26--33.
+
+\leavevmode\vadjust pre{\hypertarget{ref-zhang2019near}{}}%
+Zhang, Yinyan, Shuai Li, and Liefa Liao. 2019. {``Near-Optimal Control
+of Nonlinear Dynamical Systems: A Brief Survey.''} \emph{Annual Reviews
+in Control} 47: 71--80.
+
+\end{CSLReferences}
 
